### 2021-04-05 doc/mc-notes.tex
Background information on model counting

\documentclass{llncs}

\usepackage{latexsym}
\usepackage{times}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{marvosym}
%% Colored hyperlink 
\newcommand{\cref}[2]{\href{#1}{\color{blue}#2}}
%% Colored hyperlink showing link in TT font
% \newcommand{\chref}[1]{\href{#1}{\small\tt \color{blue}#1}}
\newcommand{\hcref}[1]{\cref{#1}{\small\tt #1}}

\bibliographystyle{splncs04}

\newcommand{\one}{\mbox{\bf 1}}
\newcommand{\zero}{\mbox{\bf 0}}
\newcommand{\leafone}{L_1}
\newcommand{\leafzero}{L_0}
\newcommand{\booland}{\land}
\newcommand{\boolor}{\lor}
\newcommand{\boolxor}{\oplus}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\ite}{\mbox{\it ITE}}

\newcommand{\opname}[1]{\mbox{\sc #1}}
\newcommand{\andop}{\opname{And}}
\newcommand{\implyop}{\opname{Imply}}

\newcommand{\turnstile}{\vdash}

\newcommand{\fname}[1]{\mbox{\small\sf #1}}

\newcommand{\lo}{\fname{Lo}}
\newcommand{\hi}{\fname{Hi}}
\newcommand{\var}{\fname{Var}}
\newcommand{\val}{\fname{Val}}

\newcommand{\interp}{\sigma}
\newcommand{\interpset}[1]{\Sigma_{#1}}
\newcommand{\mcount}{\mu}

\newcommand{\ifarg}{\textbf{I}}
\newcommand{\thenarg}{\textbf{T}}
\newcommand{\elsearg}{\textbf{E}}
\newcommand{\depend}{{\it D}}

\newcommand{\subs}[2]{[#1/#2]}
\newcommand{\substrue}[1]{\subs{#1}{\tautology}}
\newcommand{\subsfalse}[1]{\subs{#1}{\false}}
\newcommand{\subsflip}[1]{\subs{#1}{\obar{#1}}}

\title{Notes on Validated Model Counting}

\author{Randal E. Bryant}

\institute{
Computer Science Department \\
Carnegie Mellon University, Pittsburgh, PA, United States
}

\date{\today}

\begin{document}

\maketitle

\section{Notation}

Consider Boolean formulas over a set of variables $X$.  An
assignment $\interp$ is a function mapping each variable to 
truth value $\tautology$ (true) or $\nil$ (false).  We can extend
$\interp$ to Boolean formulas in the normal way, such that
$\interp(F)$ will be $\tautology$ (respectively, $\nil$) if the
evaluation of formula $F$ yields $\tautology$ (resp., $\nil$) when its
variables are assigned values according to $\interp$.

For Boolean formula $F$, we define its set of {\em models} $\interpset{F}$ as
\begin{eqnarray}
\interpset{F} & = & \{ \interp | \interp(F) = \tautology \}
\end{eqnarray}
The task of model counting is, given formula $F$, to determine the size of its set of models
$\mcount(F) = |\interpset{F}|$.  Ideally, this should be done without actually enumerating the set.

For interpretation $\interp$ and a Boolean formula $E$ over $X$, we
use the notation $\interp\subs{x}{E}$ to denote the intepretation
$\interp'$, such that $\interp'(y) = \interp(y)$ for all $y \not = x$
and $\interp'(x) = \interp(E)$.  In particular, the notation
$\interp\subsflip{x}$ indicates the interpretation in which the value
assigned to $x$ is complemented, while others remain unchanged.

A Boolean formula $F$ is said to be {\em independent} of variable $x$
if every $\interp \in \interpset{F}$ has $\interp\subsflip{x} \in
\interpset{F}$.

\begin{lemma}
\label{lemma:independent}
if Boolean formula $F$ is independent of variable $x$, then:
\begin{equation}
|\interpset{F}| \; = \; 2\cdot |\interpset{F\land x}| = \; 2\cdot |\interpset{F\land \obar{x}}| \label{eqn:independent}
\end{equation}  
\end{lemma}  
That is, the set of interpretations of $F$ are split evenly between
those assigning $\tautology$ to $x$ and those assigning $\nil$ to $x$.

\section{ITE Operation}

We consider a single Boolean operation, known as
``If-Then-Else,'' or simply ``ITE.''  For Boolean values $a, b, c$,
the operation is defined as $\ite(a,b,c) = (a \land b) \lor (\obar{a}
\land c)$.  This single operation can be used to express several common Boolean operations:
\begin{eqnarray*}
\obar{a} & = & \ite(a, \nil, \tautology) \\
a \land b & = & \ite(a, b, \nil) \\
a \lor b & = & \ite(a, \tautology, b) \\
a \rightarrow b & = & \ite(a, b, \tautology) \\
\end{eqnarray*}

For variable $x$ and Boolean formulas $F$ and $G$, the set of models
for $\ite(x, F, G)$ is given by the formula:
\begin{eqnarray}
  \interpset{\ite(x, F, G)} & = & \{ \interp \in \interpset{F} | \interp(x) = \tautology \}
  \cup \{ \interp \in \interpset{G} | \interp(x) = \nil \} \label{eqn:ite:interp}
\end{eqnarray}

\section{ITE Graphs}
Model counting is especially simple for a class of formulas we
call ``free ITE graphs.''  This representation can be obtained
directly from an Ordered Binary Decision Diagram (OBDD) representation
of a Boolean function~\cite{Bryant:1986}.

An {\em ITE graph} is defined to be a directed acyclic graph,
consisting of three node types:
\begin{description}
\item[Constant:] Corresponds to value $\tautology$ or $\nil$.  A constant node has no incoming arcs.
\item[Input:] Corresponds to one of the variables in $X$.   A variable node has no incoming arcs.
\item[Operator:] Represents an application of the $\ite$ operation.  An operator node has three incoming arcs, labeled \ifarg{}, \thenarg{}, and \elsearg{}, corresponding to the three arguments of the $\ite$ operation.
\end{description}

We consider only graphs having unique root nodes, i.e., nodes with no outgoing arcs.

A {\em proper} ITE graph is one for which the incoming \ifarg{} arc for every operator node is
from an input node.  That is, the result of one $\ite$ operation can
serve as the ``then'' or ``else'' argument to another $\ite$
operation, but not for the ``if'' argument.

When describing proper ITE graphs, we refer to constant and input
nodes by their associated value or variable.  An operator node is described by an expression of the form
$\ite(x, v_t, v_e)$, where $v_t$ and $v_e$ are the
nodes corresponding to the \thenarg{} and \elsearg{} inputs, respectively.

We can define the function $\depend$ mapping each node
$v$ to the set of variables on which it logically depends.  This can be expressed recursively as
\begin{eqnarray}
\depend(\tautology) & = & \emptyset \label{eqn:depend:true}\\
\depend(\nil) & = & \emptyset \label{eqn:depend:false}\\
\depend(x) & = & \{x \} \label{eqn:depend:variable}\\
\depend(\ite(x, v_t, v_e)) & = & \{ x \} \cup \depend(v_t) \cup \depend(v_e) \label{eqn:depend:ite}
\end{eqnarray}

\begin{lemma}
\label{lemma:free}
Any node $v$ in an ITE graph is independent of any variable $y \in X$ such that $y \not \in \depend(v)$.
\end{lemma}  
\noindent
{\em Proof:} Follows by the recursive definition of $\depend$ (\ref{eqn:depend:true}--\ref{eqn:depend:ite}) and by (\ref{eqn:ite:interp}).

A {\em free} ITE graph is a proper ITE graph satisfying the added
restriction that for every chain of nodes from an input or constant
node to the root node, and for every variable $x \in X$, only a single
operator node can have the input node for $x$ as its \ifarg{} input.
More formally, for every operator node $v = \ite(x, v_t, v_e)$ in a
free ITE graph, its incoming \thenarg{} and \elsearg{} arcs must be
from nodes such that $x \not \in \depend(v_t)$ and $x \not \in
\depend(v_e)$.

\section{Model Counting Free ITE Graphs}

While model counting is a difficult problem for arbitrary
formulas, it readily be computed when $F$ takes the form of a free
ITE graph.

Letting $n = |X|$, we can use the following recursive computation:
\begin{eqnarray}
\mcount(\tautology) & = & 2^{n} \label{eqn:count:true}\\
\mcount(\nil) & = & 0 \label{eqn:count:false}\\
\mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
\mcount(\ite(x, v_t, v_e)) & = & \frac{\mcount(v_t) + \mcount(v_e)}{2} \label{eqn:count:ite}
\end{eqnarray}

\begin{theorem}
Equations (\ref{eqn:count:true}--\ref{eqn:count:ite}) correctly hold for any node $v$ in a free ITE graph.
\end{theorem}  
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemma~\ref{lemma:independent}, (\ref{eqn:ite:interp}),
Lemma~\ref{lemma:free}, and (\ref{eqn:depend:ite}).  That is, both
$v_t$ and $v_e$ are independent of $x$.  That implies that half of the
satisfying assignments $\interp$ for $v_t$ have have $\interp(x) =
\tautology$.  Similarly, half of the satisfying assignments $\interp$
for $v_e$ have $\interp(x) = \nil$.  Equation (\ref{eqn:count:ite}) then follows by (\ref{eqn:ite:interp}).

\bibliography{references}

\end{document}

### 2021-04-05 doc/mc-notes.tex
Minor updates

@@
 \begin{lemma}
-\label{lemma:independent}
+\label{lemma:independent:split}
 if Boolean formula $F$ is independent of variable $x$, then:
@@
 
-We consider only graphs having unique root nodes, i.e., nodes with no outgoing arcs.
-
 A {\em proper} ITE graph is one for which the incoming \ifarg{} arc for every operator node is
@@
 We can define the function $\depend$ mapping each node
-$v$ to the set of variables on which it logically depends.  This can be expressed recursively as
+$v$ in a proper ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
 \begin{eqnarray}
@@
 \begin{lemma}
-\label{lemma:free}
+\label{lemma:independent:dset}
 Any node $v$ in an ITE graph is independent of any variable $y \in X$ such that $y \not \in \depend(v)$.
@@
 A {\em free} ITE graph is a proper ITE graph satisfying the added
-restriction that for every chain of nodes from an input or constant
-node to the root node, and for every variable $x \in X$, only a single
-operator node can have the input node for $x$ as its \ifarg{} input.
-More formally, for every operator node $v = \ite(x, v_t, v_e)$ in a
-free ITE graph, its incoming \thenarg{} and \elsearg{} arcs must be
-from nodes such that $x \not \in \depend(v_t)$ and $x \not \in
-\depend(v_e)$.
+property that for every operator node $v = \ite(x, v_t, v_e)$, its
+incoming \thenarg{} and \elsearg{} arcs must be from nodes such that
+$x \not \in \depend(v_t)$ and $x \not \in \depend(v_e)$.
+
+Free ITE graphs correspond directly to a class of binary decision diagrams known as {\em free} BDDs (FBDDs)~\cite{gergov:ieeetc:1994}.  Ordered BDDs are a subclass of FBDDs.
 
@@
 {\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
-from Lemma~\ref{lemma:independent}, (\ref{eqn:ite:interp}),
-Lemma~\ref{lemma:free}, and (\ref{eqn:depend:ite}).  That is, both
+from Lemma~\ref{lemma:independent:split}, (\ref{eqn:ite:interp}),
+Lemma~\ref{lemma:independent:dset}, and (\ref{eqn:depend:ite}).  That is, both
 $v_t$ and $v_e$ are independent of $x$.  That implies that half of the

### 2021-04-05 doc/mc-notes.tex
Improved theorem presentation
Minor edits

@@
 \label{lemma:independent:split}
-if Boolean formula $F$ is independent of variable $x$, then:
+If Boolean formula $F$ is independent of variable $x$, then:
 \begin{equation}
@@
 \end{lemma}  
-That is, the set of interpretations of $F$ are split evenly between
+That is, the set of models of $F$ is split evenly between
 those assigning $\tautology$ to $x$ and those assigning $\nil$ to $x$.
@@
 
-Letting $n = |X|$, we can use the following recursive computation:
+\begin{theorem}
+\label{theorem:count}
+Letting $n = |X|$, the following recursive formula holds for any node in a free ITE graph
 \begin{eqnarray}
@@
 \end{eqnarray}  
-
-\begin{theorem}
-Equations (\ref{eqn:count:true}--\ref{eqn:count:ite}) correctly hold for any node $v$ in a free ITE graph.
 \end{theorem}

### 2021-04-15 doc/mc-notes.tex
New notation
Added

@@
 \newcommand{\boolnot}{\neg}
-\newcommand{\tautology}{\top}
-\newcommand{\nil}{\bot}
+%\newcommand{\tautology}{\top}
+%\newcommand{\nil}{\bot}
+\newcommand{\tautology}{1}
+\newcommand{\nil}{0}
 \newcommand{\obar}[1]{\overline{#1}}
@@
 
-\newcommand{\subs}[2]{[#1/#2]}
+\newcommand{\subs}[2]{[#2/#1]}
 \newcommand{\substrue}[1]{\subs{#1}{\tautology}}
@@
 
-A {\em proper} ITE graph is one for which the incoming \ifarg{} arc for every operator node is
+An {\em input-controlled} ITE graph is one for which the incoming \ifarg{} arc for every operator node is
 from an input node.  That is, the result of one $\ite$ operation can
@@
 
-When describing proper ITE graphs, we refer to constant and input
+When describing input-controlled ITE graphs, we refer to constant and input
 nodes by their associated value or variable.  An operator node is described by an expression of the form
@@
 We can define the function $\depend$ mapping each node
-$v$ in a proper ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
+$v$ in an input-controlled ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
 \begin{eqnarray}
@@
 
-A {\em free} ITE graph is a proper ITE graph satisfying the added
+A {\em free} ITE graph is an input-controlled ITE graph satisfying the added
 property that for every operator node $v = \ite(x, v_t, v_e)$, its

### 2022-03-21 doc/mc-notes.tex
Added benchmarks
Generalize graphs

--- 2021-04-05 doc/mc-notes.tex
@@
 \documentclass{llncs}
-
 \usepackage{latexsym}
@@
 \bibliographystyle{splncs04}
+%\bibliographystyle{abbrv}
 
@@
 \newcommand{\nil}{\bot}
+%\newcommand{\tautology}{1}
+%\newcommand{\nil}{0}
 \newcommand{\obar}[1]{\overline{#1}}
@@
 
-\newcommand{\interp}{\sigma}
-\newcommand{\interpset}[1]{\Sigma_{#1}}
+\newcommand{\interp}{\alpha}
+\newcommand{\interpset}[1]{{\cal M}(#1)}
 \newcommand{\mcount}{\mu}
@@
 
-\title{Notes on Validated Model Counting}
+\newcommand{\subspace}{{\cal S}}
+
+\title{Notes on Validated Model Counting \\ Version of \today}
 
 \author{Randal E. Bryant}
+
+\authorrunning{R. E. Bryant}
+
+\titlerunning{Validated Model Counting}
 
@@
 }
-
-\date{\today}
 
@@
 
-For interpretation $\interp$ and a Boolean formula $E$ over $X$, we
-use the notation $\interp\subs{x}{E}$ to denote the intepretation
+For assignment $\interp$ and a Boolean formula $E$ over $X$, we
+use the notation $\interp\subs{x}{E}$ to denote the assignment
 $\interp'$, such that $\interp'(y) = \interp(y)$ for all $y \not = x$
 and $\interp'(x) = \interp(E)$.  In particular, the notation
-$\interp\subsflip{x}$ indicates the interpretation in which the value
+$\interp\subsflip{x}$ indicates the assignment in which the value
 assigned to $x$ is complemented, while others remain unchanged.
@@
 
+The following lemmas describe the primary abstractions used to perform
+model counting without enumerating all solutions.
+
 \begin{lemma}
-\label{lemma:independent:split}
-If Boolean formula $F$ is independent of variable $x$, then:
-\begin{equation}
-|\interpset{F}| \; = \; 2\cdot |\interpset{F\land x}| = \; 2\cdot |\interpset{F\land \obar{x}}| \label{eqn:independent}
-\end{equation}  
+  \label{lemma:negation}
+  For any formula $F$:
+  \begin{eqnarray*}
+    \mcount(\neg F) & =  & 2^n - \mcount(F)
+  \end{eqnarray*}
 \end{lemma}
-That is, the set of models of $F$ is split evenly between
-those assigning $\tautology$ to $x$ and those assigning $\nil$ to $x$.
+
+\begin{lemma}
+  \label{lemma:product}
+  For a partitioning of the variables $X$ into disjoint sets $X_1$ and $X_2$, and for formulas $F_1$ defined over $X_1$ and $F_2$ defined over $X_2$:
+  \begin{eqnarray*}
+    \mcount(F_1 \land F_2) &=& \mcount(F_1) \cdot \mcount(F_2)
+  \end{eqnarray*}
+\end{lemma}
+
+\begin{lemma}
+  \label{lemma:sum}
+  For formulas $F_1$ and $F_2$ such that $\interpset{F_1} \cap \interpset{F_2} = \emptyset$:
+  \begin{eqnarray*}
+    \mcount(F_1 \lor F_2) &=& \mcount(F_1) + \mcount(F_2)
+  \end{eqnarray*}
+\end{lemma}
 
@@
 ``If-Then-Else,'' or simply ``ITE.''  For Boolean values $a, b, c$,
-the operation is defined as $\ite(a,b,c) = (a \land b) \lor (\obar{a}
+the operation is defined as $\ite(a,b,c) = (a \land b) \lor (\neg a
 \land c)$.  This single operation can be used to express several common Boolean operations:
 \begin{eqnarray*}
-\obar{a} & = & \ite(a, \nil, \tautology) \\
+\neg a & = & \ite(a, \nil, \tautology) \\
 a \land b & = & \ite(a, b, \nil) \\
@@
 \section{ITE Graphs}
-Model counting is especially simple for a class of formulas we
-call ``free ITE graphs.''  This representation can be obtained
-directly from an Ordered Binary Decision Diagram (OBDD) representation
-of a Boolean function~\cite{Bryant:1986}.
 
-An {\em ITE graph} is defined to be a directed acyclic graph,
-consisting of three node types:
+An {\em ITE graph} describes a combinational logic circuit constructed from two-input multiplexors.
+It is defined as a directed acyclic graph
+with three node types:
 \begin{description}
@@
 
-A {\em proper} ITE graph is one for which the incoming \ifarg{} arc for every operator node is
-from an input node.  That is, the result of one $\ite$ operation can
-serve as the ``then'' or ``else'' argument to another $\ite$
-operation, but not for the ``if'' argument.
-
-When describing proper ITE graphs, we refer to constant and input
+When describing ITE graphs, we refer to constant and input
 nodes by their associated value or variable.  An operator node is described by an expression of the form
-$\ite(x, v_t, v_e)$, where $v_t$ and $v_e$ are the
-nodes corresponding to the \thenarg{} and \elsearg{} inputs, respectively.
+$\ite(v_i, v_t, v_e)$, where $v_i$, $v_t$, and $v_e$ are the
+nodes corresponding to the \ifarg{}, \thenarg{}, and \elsearg{} inputs, respectively.
 
 We can define the function $\depend$ mapping each node
-$v$ in a proper ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
+$v$ in an ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
 \begin{eqnarray}
@@
 \depend(x) & = & \{x \} \label{eqn:depend:variable}\\
-\depend(\ite(x, v_t, v_e)) & = & \{ x \} \cup \depend(v_t) \cup \depend(v_e) \label{eqn:depend:ite}
+\depend(\ite(x, v_t, v_e)) & = & \depend(v_i) \cup \depend(v_t) \cup \depend(v_e) \label{eqn:depend:ite}
 \end{eqnarray}
@@
 
-A {\em free} ITE graph is a proper ITE graph satisfying the added
-property that for every operator node $v = \ite(x, v_t, v_e)$, its
-incoming \thenarg{} and \elsearg{} arcs must be from nodes such that
-$x \not \in \depend(v_t)$ and $x \not \in \depend(v_e)$.
+A {\em variable-partitioned} ITE (VPITE) graph is one satisfying the property
+that for every operator node $v = \ite(v_i, v_t, v_e)$, the dependency
+set of its \ifarg{} input most be disjoint from those of its
+\thenarg{} and \elsearg{} inputs.  That is, both $\depend(v_i) \cap \depend(v_t) = \emptyset$ and
+$\depend(v_i) \cap \depend(v_e) = \emptyset$.
 
-Free ITE graphs correspond directly to a class of binary decision diagrams known as {\em free} BDDs (FBDDs)~\cite{gergov:ieeetc:1994}.  Ordered BDDs are a subclass of FBDDs.
-
-\section{Model Counting Free ITE Graphs}
+\section{Model Counting VPITE Graphs}
 
 While model counting is a difficult problem for arbitrary
-formulas, it readily be computed when $F$ takes the form of a free
-ITE graph.
+formulas, it readily be computed when $F$ takes the form of a VPITE graph.
 
@@
 \label{theorem:count}
-Letting $n = |X|$, the following recursive formula holds for any node in a free ITE graph
+Letting $n = |X|$, the following recursive formula holds for any node in a VPITE graph:
 \begin{eqnarray}
@@
 \mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
-\mcount(\ite(x, v_t, v_e)) & = & \frac{\mcount(v_t) + \mcount(v_e)}{2} \label{eqn:count:ite}
+\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2} \label{eqn:count:ite}
 \end{eqnarray}  
@@
 {\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
-from Lemma~\ref{lemma:independent:split}, (\ref{eqn:ite:interp}),
-Lemma~\ref{lemma:independent:dset}, and (\ref{eqn:depend:ite}).  That is, both
-$v_t$ and $v_e$ are independent of $x$.  That implies that half of the
-satisfying assignments $\interp$ for $v_t$ have have $\interp(x) =
-\tautology$.  Similarly, half of the satisfying assignments $\interp$
-for $v_e$ have $\interp(x) = \nil$.  Equation (\ref{eqn:count:ite}) then follows by (\ref{eqn:ite:interp}).
+from Lemmas~\ref{lemma:negation}--\ref{lemma:sum}, along with the definition of the ITE operator.
+
+\section{Relation to Other Representations}
+
+In general, there is a trade-off between the compactness of different
+Boolean function representations versus the ease with which different
+properties can readily be computed.  For example, Boolean formulas and
+Boolean circuits can be very compact, but basic properties of these
+representations, such as satisfiability, equivalence testing, and
+model counting are NP-hard.  On the other hand all of these properties
+can be computed in polynomial time for reduced, orderd binary decision
+diagrams (ROBDDs)~\cite{Bryant:1986}, but they can grow exponentially
+in relation to the representation of the function as a formula or
+circuit.  ITE graphs lie somewhere between these two extremes.  Here
+we examine how they relate to other representations and the implications
+of these relations.
+
+\subsection{Binary Decision Diagrams}
+
+Binary Decision Diagrams~\cite{akers:ieeetc:1978} also represent Boolean formulas
+as directed graphs consisting of ITE operators.  They can be seen as
+equivalent to a class of graphs we will call {\em input-controlled} ITE
+graphs.  These are ones for which the \ifarg{} input to every operator
+node is a variable.
+
+If we combine the properties of variable partitioned and input
+control, we arrive at the representation known as {\em
+  Free}~BDDs~\cite{gergov:ieeetc:1994}.  If we further restrict the
+variables in a variable-controlled, input-partitioned ITE graph to
+obey the same ordering along all paths formed in the graph, we arrive
+at OBDDs.  An OBDD can be reduced by a simple set of rules to give an
+ROBDD representation, and this form is canonical~\cite{Bryant:1986}.
+
+\subsection{Sententential Decision Diagrams}
+
+Darwiche uses the Sentential Decision Diagram (SDD)
+representation~\cite{darwiche:ijcai:2011} as his framework for several
+model counting programs.  An SDD can be seen as a combination of 1) a
+generalization of VPITE graphs consisting of $k$-input multiplexors,
+and 2) an ordering restriction on the variables.
+That is, each operator node in an SDD is defined to
+have $k$ data inputs and $k$ control inputs, where the Boolean
+functions corresponding to the control inputs $p_1, p_2, \ldots, p_k$
+satisfy $p_i \land p_j = \nil$ for $1 \leq i < j \leq k$ and
+$\bigvee_{i=1,k} p_i = \tautology$.  The $k$ data inputs represent
+Boolean functions $s_1, s_2, \ldots, s_k$.  In addition, the
+dependency sets for the control and data inputs must be disjoint:
+$\depend(p_i) \cap \depend(s_j) = \emptyset$ for $1 \leq i,j \leq k$.
+Furthermore, the ordering of the variables in an SDD must obey a hierarchical decomposition
+defined by
+a tree structure known as a {\em vtree}.  Indeed, an SDD can be seen
+as equivalent to a generalization of BDDs devised by McMillan, which
+he termed {\em tree} BDDs~\cite{mcmillan:cav:1994}.
+
+The graph part of an SDD can be readily translated into a VPITE graph
+by expanding each $k$-input node into $k-1$ ITE nodes, giving an
+output value defined as:
+\begin{displaymath}
+\ite(p_1, s_1, \ite(p_2, s_2, \ite( \cdots \ite(p_{k-1}, s_{k-1}, s_k)\cdots))
+\end{displaymath}
+If we express the size of an SDD as the number of branches, this
+translation from SDD to ITE graph will expand the size by a constant
+factor.
+
+\section{Probabilistic Equivalence Checking}
+
+One further feature of VPITE graphs is that there is a randomized
+polynomial time (RP) algorithm for determining whether two graphs
+represent equivalent Boolean functions.  This algorithm generalizes
+the RP equivalence-testing algorithm for free BDDs described by Blum,
+Chandra, and Wegman~\cite{blum:ipl:1980}.
 
@@
 \end{document}
+
+\section{A Model Counting Calculus}
+
+Objective: Create a simple system for defining and reasoning about an
+abstraction of Boolean formulas that suffices for tracking the number
+of models.
+
+Define a {\em counting subspace} $\subspace{} = \langle V, \alpha, k\rangle$, with:
+\begin{itemize}
+\item $V$ is a set of variables
+\item $\alpha$ is a partial assignment to variables in $V$, represented as a set of literals.
+\item $k$ is an integer, indicating the number of models for
+  the formula within the specified subspace.  For $n = |V|$ and $m= |\var(\alpha)|$, it satisfies
+  $0 \leq k \leq 2^{n-m}$.
+\end{itemize}
+
+\subsection{Subspace Primitives}
+\begin{itemize}
+\item
+Universal subspace: $\langle \emptyset, \emptyset, 1 \rangle$.
+\item
+Null subspace: $\langle \emptyset, \emptyset, 0 \rangle$.
+\item
+  Positive literal: $\langle \{ x\}, \{x\}, 1\rangle$.
+\item
+  Negative literal: $\langle \{ x\}, \{\obar{x}\}, 1\rangle$.
+\item
+  Unused variable: $\langle \{ x\}, \emptyset, 2\rangle$.
+\end{itemize}
+
+\subsection{Basic Subspace Operations}
+
+Assume $\subspace_1 = \langle V_1, \alpha_1, k_1\rangle$ and
+$\subspace_2 = \langle V_2, \alpha_2, k_2\rangle$ are two subspaces
+
+Subspace product $\subspace_1 \times \subspace_2$
+\begin{itemize}
+\item Condition: $V_1 \cap V_2 = \emptyset$
+\item Result: $\langle V_1 \cup V_2, \alpha_1 \cup \alpha_2, k_1 \cdot k_2\rangle$.
+\end{itemize}
+
+Subspace Sum $\subspace_1 + \subspace_2$
+\begin{itemize}
+\item Conditions:
+  \begin{itemize}
+  \item $V_1 = V_2$.
+  \item For some literal $l$ of variable $x \in V_1$ and assignment $\alpha$, the argument assignments satisfy
+    $\alpha_1 = \alpha \cup \{l\}$ and $\alpha_2 = \alpha \cup \{ \obar{l} \}$.
+  \end{itemize}
+\item Result: $\langle V_1 - \{ x\}, \alpha, (k_1 + k_2)/2\rangle$.
+\end{itemize}
+
+\subsection{Derived Subspace Operations}
+
+These useful operations can be constructed by combining primitive values and basic operations.
+
+Expanding subspace $\subspace{} = \langle V, \alpha, k\rangle$:
+\begin{itemize}
+\item
+  Condition: $x \not \in V$
+\item
+  Results:
+  \begin{itemize}
+    \item Positive expansion: $\langle V \cup \{ x \}, \alpha \cup \{ x \}, k \rangle$.
+    \item Negative expansion: $\langle V \cup \{ x \}, \alpha \cup \{ \obar{x} \}, k \rangle$.
+    \item Don't-care expansion: $\langle V \cup \{ x \}, \alpha , 2\cdot k \rangle$.
+  \end{itemize}
+\end{itemize}
+
+A particularly useful expansion is to apply the don't-care expansion
+with a set of variables $U$ such that $U \cap V = \emptyset$.  This
+will give the subspace $\langle U \cup V, \alpha, 2^{|U|} \cdot
+k\rangle$.

### 2022-03-28 doc/mc-notes.tex
Fixed typo
New notation

@@
 \newcommand{\boolnot}{\neg}
-\newcommand{\tautology}{\top}
-\newcommand{\nil}{\bot}
-%\newcommand{\tautology}{1}
-%\newcommand{\nil}{0}
+%\newcommand{\tautology}{\top}
+%\newcommand{\nil}{\bottom}
+\newcommand{\tautology}{1}
+\newcommand{\nil}{0}
 \newcommand{\obar}[1]{\overline{#1}}
@@
 \newcommand{\interpset}[1]{{\cal M}(#1)}
+\newcommand{\ring}{{\cal Z}}
+\newcommand{\cost}{\sigma}
+\newcommand{\density}{\rho}
+\newcommand{\hashset}{{\cal H}}
+\newcommand{\fhash}{h}
 \newcommand{\mcount}{\mu}
@@
 \section{Notation}
+
+Let $X = \{x_1, x_2, \ldots, x_n\}$ be a set of Boolean variables.  An
+{\em assignment} is a function $\interp$ assigning Boolean values to
+the variables: $\interp:X \rightarrow \{\nil, \tautology\}$.  We can
+also view an assignment as a set of {\em literals} $\{l_1, l_2,
+\ldots, l_n\}$, where each literal $l_i$ is either $x_i$ or
+$\obar{x}_i$, corresponding to the assignments $\interp(x_i)$ = 1 or 0,
+respectively.
+
+\subsection{Boolean Functions}
+
+A {\em Boolean function} $f:2^X \rightarrow \{0,1\}$ can be
+characterized by the set of assignments for which the function
+evaluates to 1: $\interpset{f} = \{ \interp | f(\interp) = 1\}$.  Let
+$\one$ denote the Boolean function that assigns value 1 to every
+assignment, and $\zero$ denote the assignment that assigns value 0 to
+every assignment.  These are characterized by the universal and empty
+assignment sets, respectively.
+
+From
+this we can define the {\em negation} of function $f$ as the function
+$\boolnot f$ such that
+$\interpset{\boolnot f} = \{ \interp | f(\interp) = 0\}$.
+We can also define the conjunction and disjunction operations over functions $f_1$ and $f_2$ as characterized by the sets
+$\interpset{f_1 \booland f_2} = \interpset{f_1} \cap \interpset{f_2}$ and
+$\interpset{f_1 \boolor f_2} = \interpset{f_1} \cup \interpset{f_2}$.
+
+For assignment $\interp$ and a Boolean formula $E$ over $X$, we
+use the notation $\interp\subs{x_i}{E}$ to denote the assignment
+$\interp'$, such that $\interp'(x_j) = \interp(x_j)$ for all $j \not = i$
+and $\interp'(x_i) = \interp(E)$, where $\interp(E)$ is the value obtained by evaluating formula $E$ with each variable assigned the value given by $\interp$.
+In particular, the notation
+$\interp\subsflip{x_i}$ indicates the assignment in which the value
+assigned to $x_i$ is complemented, while others remain unchanged.
+
+A Boolean function $f$ is said to be {\em independent} of variable
+$x_i$ if every $\interp \in \interpset{f}$ has $\interp\subsflip{x_i}
+\in \interpset{f}$.  The {\em dependency set} of $f$, denoted
+$\depend(f)$ consists of all variables $x_i$ for which $f$ is {\em
+  not} independent.
+
+\subsection{Separable Cost Functions}
+
+Let $\ring$ denote the elements of a commutative ring.  A {\em
+  separable cost function} $\cost:X \rightarrow \ring$ assigns a value
+from the ring to each variable.  We extend this function by defining
+the cost of literal as $\cost(\obar{x}_i) = 1 - \cost(x_i)$, the cost
+of an assignment as $\cost(\interp) = \prod_{l_i \in \interp}
+\cost(l_i)$, and the cost of a function $f$ as $\cost(f) =
+\sum_{\interp \in \interpset{f}} \cost(\interp)$.
+
+{\bf Example 1}: Let $\ring$ be the set of rational numbers and
+$\cost(x_i) = 1/2$ for all variables $x_i$.  Then cost of every
+assignment will be $1/2^{n}$, and the cost of a function will be its
+{\em density}, denoted $\density(f)$.  That is $\density(f) \in [0,
+  1]$ will be the fraction of assignments for which the function
+evaluates to 1.  Given an ability to compute the density of
+function $f$, we can scale this by $2^n$ to compute $|\interpset{f}|$, the core task of model counting.
+Using density as our metric has the advantage that it
+does not vary when the function is embedded in a larger domain $X'
+\supseteq X$.
+
+{\bf Example 2}: Let $\ring$ be a field with $|\ring| > 2n$,
+and let $\hashset$ be the set functions
+mapping elements of $X$ to elements of $\ring$.  For
+two distinct functions $f_1$ and $f_2$ and a randomly chosen $\fhash
+\in \hashset$, the probability that $h(f_1) = h(f_2)$ will be at most
+$2^n/|\ring|< 1/2$.  Therefore, these functions can be used as part of a
+randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
+
+\subsection{Computing Cost Functions}
+
+Three key properties of separable cost functions makes it
+possible, in some cases, to compute the cost of a Boolean formula without enumerating all of its satisfying solutions.
+
+\begin{lemma}[Complementation]
+  For separable cost function $\cost$ and Boolean function $f$:
+  $\cost(\neg f) = 1 - \cost(f)$.
+\end{lemma}
+
+\begin{lemma}[Conjunction over Independent Domains]
+  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\depend(f_1) \cap \depend(f_2) = \emptyset$:
+    $\cost(f_1 \land f_2) = \cost(f_1) \cdot \cost(f_2)$.
+\end{lemma}
+
+\begin{lemma}[Disjunction over Disjoint Domains]
+  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\interpset{f_1} \cap \interpset{f_2} = \emptyset$:
+    $\cost(f_1 \lor f_2) = \cost(f_1) + \cost(f_2)$.
+\end{lemma}
+
+\bibliography{references}
+
+\end{document}
 
@@
 \mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
-\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2} \label{eqn:count:ite}
+\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2^n} \label{eqn:count:ite}
 \end{eqnarray}  

### 2022-04-08 doc/mc-notes.tex
New proof framework

@@
 \newcommand{\boolor}{\lor}
+\newcommand{\pand}{\mathbin{\land_{\rm pv}}}
+\newcommand{\por}{\mathbin{\lor_{\rm pa}}}
 \newcommand{\boolxor}{\oplus}
@@
 From
-this we can define the {\em negation} of function $f$ as the function
+this we can define the {\em complement} of function $f$ as the function
 $\boolnot f$ such that
@@
 from the ring to each variable.  We extend this function by defining
-the cost of literal as $\cost(\obar{x}_i) = 1 - \cost(x_i)$, the cost
+the cost of literal $\obar{x}_i$ as $\cost(\obar{x}_i) = 1 - \cost(x_i)$, the cost
 of an assignment as $\cost(\interp) = \prod_{l_i \in \interp}
@@
 {\bf Example 1}: Let $\ring$ be the set of rational numbers and
-$\cost(x_i) = 1/2$ for all variables $x_i$.  Then cost of every
+$\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
 assignment will be $1/2^{n}$, and the cost of a function will be its
@@
 does not vary when the function is embedded in a larger domain $X'
-\supseteq X$.
+\supseteq X$.  As a variant, we can support some forms of
+weighted model counting by assigning weights of the form
+$\cost(x_i) = w_i$, where $0 \leq w_i \leq 1$ for every variable $x_i$.
 
 {\bf Example 2}: Let $\ring$ be a field with $|\ring| > 2n$,
-and let $\hashset$ be the set functions
+and let $\hashset$ be the set of functions
 mapping elements of $X$ to elements of $\ring$.  For
@@
 
-Three key properties of separable cost functions makes it
-possible, in some cases, to compute the cost of a Boolean formula without enumerating all of its satisfying solutions.
+Three key properties of separable cost functions makes it possible, in
+some cases, to compute the cost of a Boolean formula without
+enumerating all of its satisfying solutions.
 
 \begin{lemma}[Complementation]
+\label{lemma:complementation}
   For separable cost function $\cost$ and Boolean function $f$:
@@
 
-\begin{lemma}[Conjunction over Independent Domains]
+\begin{lemma}[Variable-Partitioned Conjunction]
+\label{lemma:conjunction}
   For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\depend(f_1) \cap \depend(f_2) = \emptyset$:
@@
 \end{lemma}
+We use the notation $f_1 \pand f_2$ to denote the conjunction of $f_1$ and
+$f_2$ under the condition that $f_1$ and $f_2$ are defined over
+disjoint sets of variables.
 
-\begin{lemma}[Disjunction over Disjoint Domains]
+\begin{lemma}[Assignment-Partitioned Disjunction]
+\label{lemma:disjunction}
   For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\interpset{f_1} \cap \interpset{f_2} = \emptyset$:
@@
 \end{lemma}
+We use the notation $f_1 \por f_2$ to denote the disjunction of $f_1$ and $f_2$ under the
+condition that $f_1$ and $f_2$ hold for mutually exclusive assignments.
+
+\section{Proof Framework for Cost Functions}
+
+We introduce the CRAT clausal proof framework for generating a
+checkable proof of the valuation of a Boolean formula, given in
+conjunctive normal form, according to a specified cost function.  A
+CRAT proof provides two capabilities: 1) a schema for computing a cost
+function using the ring operations, and 2) a proof that this schema
+accurately characterizes the cost function for the input formula.
+A proof can be either {\em function independent}, meaning that it holds
+for any cost function, or {\em function dependent}, meaning that
+it holds for only for a specific cost function.
+
+The CRAT format draws its inspiration from the QRAT format for
+quantified Boolean formulas (QBF).  The following are its key properties:
+\begin{itemize}
+\item Clause addition is allowed, as long as either 1) the new clause is
+  blocked, or 2) it satisfies the RAT property with respect to a sequence
+  of earlier antecedent clauses.
+\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and {\em abstraction}
+\begin{itemize}
+\item The checker tracks the dependency set for every input and
+  extension variable.  When an extension variable is introduced based
+  on the $\pand$ operation, the dependency sets of its arguments must
+  be disjoint.
+\item When an extension variable is introduced based on the $\por$
+  operation, the proof step must cite earlier steps providing a RUP proof
+  that the two arguments are mutually exclusive.
+\item The {\em abstraction} operation allows an extension
+  variable to serve as an alias for one of more other 
+  variables.  Use of this rule renders the proof function dependent.
+\item Boolean complement is provided implicitly by allowing the
+  arguments of the extension operations to be literals and not just
+  variables.
+\end{itemize}
+\item Unlike DRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
+\item Unlike QRAT, it need not support universal quantification.
+\end{itemize}
+
+A CRAT proof follows the same general form as a QRAT satisfaction
+proof.  That is, starting with the set of input clauses it produces a
+sequence of steps that both add and delete clauses.  At the final
+step, there must be only a single unit clause consisting of some
+variable or its complement.  Except for trivial cases, the final
+variable will be an extension variable.  The sequence of extension
+operations define the schema for computing the cost function.  That
+is, a value for the cost function is computed for every literal
+according to the rules of Lemmas
+\ref{lemma:complementation}--\ref{lemma:disjunction}.  For every
+abstraction step, all variables having the indicated
+extension variable as their alias must evaluate to the same value.
 
@@
 \begin{lemma}
-  \label{lemma:negation}
+  \label{lemma:complement}
   For any formula $F$:
@@
 {\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
-from Lemmas~\ref{lemma:negation}--\ref{lemma:sum}, along with the definition of the ITE operator.
+from Lemmas~\ref{lemma:complement}--\ref{lemma:sum}, along with the definition of the ITE operator.
 

### 2022-04-11 doc/mc-notes.tex
Update CRAT representation

@@
 \usepackage{graphicx}
+\usepackage{tikz}
+\usepackage{pgfplots}
 \usepackage{cite}
@@
 
+\newcommand{\makenode}[1]{{\mathbf #1}}
+\newcommand{\nodeu}{\makenode{u}}
+
 \newcommand{\interp}{\alpha}
@@
 \newcommand{\subsfalse}[1]{\subs{#1}{\false}}
-\newcommand{\subsflip}[1]{\subs{#1}{\obar{#1}}}
 
@@
 In particular, the notation
-$\interp\subsflip{x_i}$ indicates the assignment in which the value
+$\interp(\subs{x_i}{\obar{x}_i})$
+indicates the assignment in which the value
 assigned to $x_i$ is complemented, while others remain unchanged.
@@
 A Boolean function $f$ is said to be {\em independent} of variable
-$x_i$ if every $\interp \in \interpset{f}$ has $\interp\subsflip{x_i}
+$x_i$ if every $\interp \in \interpset{f}$ has $\interp(\subs{x_i}{\obar{x}_i})
 \in \interpset{f}$.  The {\em dependency set} of $f$, denoted
@@
   1]$ will be the fraction of assignments for which the function
-evaluates to 1.  Given an ability to compute the density of
-function $f$, we can scale this by $2^n$ to compute $|\interpset{f}|$, the core task of model counting.
-Using density as our metric has the advantage that it
-does not vary when the function is embedded in a larger domain $X'
-\supseteq X$.  As a variant, we can support some forms of
-weighted model counting by assigning weights of the form
-$\cost(x_i) = w_i$, where $0 \leq w_i \leq 1$ for every variable $x_i$.
+evaluates to 1.  Given an ability to compute the density of function
+$f$, we can scale this by $2^n$ to compute the total number of models
+$|\interpset{f}|$.  This is the core task of model counting.  Using
+density as our metric has the advantage that it does not vary when the
+function is embedded in a larger domain $X' \supseteq X$.  As a
+variant, we can support some forms of weighted model counting by
+assigning weights of the form $\cost(x_i) = w_i$, where $0 \leq w_i
+\leq 1$ for every variable $x_i$.
 
@@
 
-Three key properties of separable cost functions makes it possible, in
+Three key properties of separable cost functions make it possible, in
 some cases, to compute the cost of a Boolean formula without
@@
 conjunctive normal form, according to a specified cost function.  A
-CRAT proof provides two capabilities: 1) a schema for computing a cost
-function using the ring operations, and 2) a proof that this schema
+CRAT proof provides two capabilities: 1) a {\em schema} for computing a cost
+function using the ring operations, and 2) a {\em proof} that this schema
 accurately characterizes the cost function for the input formula.
-A proof can be either {\em function independent}, meaning that it holds
+A schema can be either {\em function independent}, meaning that it holds
 for any cost function, or {\em function dependent}, meaning that
@@
 \item Clause addition is allowed, as long as either 1) the new clause is
-  blocked, or 2) it satisfies the RAT property with respect to a sequence
+  blocked, or 2) it satisfies the RUP/RAT property with respect to a sequence
   of earlier antecedent clauses.
@@
 \item When an extension variable is introduced based on the $\por$
-  operation, the proof step must cite earlier steps providing a RUP proof
+  operation, the step must cite earlier steps providing a RUP/RAT proof
   that the two arguments are mutually exclusive.
@@
   variable to serve as an alias for one of more other 
-  variables.  Use of this rule renders the proof function dependent.
+  variables.  Use of this rule renders the schema function dependent.
 \item Boolean complement is provided implicitly by allowing the
@@
 
-A CRAT proof follows the same general form as a QRAT satisfaction
-proof.  That is, starting with the set of input clauses it produces a
-sequence of steps that both add and delete clauses.  At the final
-step, there must be only a single unit clause consisting of some
-variable or its complement.  Except for trivial cases, the final
-variable will be an extension variable.  The sequence of extension
-operations define the schema for computing the cost function.  That
-is, a value for the cost function is computed for every literal
-according to the rules of Lemmas
-\ref{lemma:complementation}--\ref{lemma:disjunction}.  For every
-abstraction step, all variables having the indicated
-extension variable as their alias must evaluate to the same value.
+\subsection{Syntax}
+
+\begin{table}
+  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
+  \label{tab:crat:syntax}
+\centering{
+  \begin{tabular}{lllll}
+    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
+    \midrule
+    \makebox[5mm][l]{$C$} & \makebox[7mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
+    $C$ & {\tt ab} & $L^{+}$ {\tt 0} & ${\tt -}C^{*}$  {\tt 0}  & Add blocked clause \\
+    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RUP/RAT clause \\
+     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RUP/RAT clause \\
+    \midrule
+        & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
+        & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
+        & {\tt a} & $V \; L$         & $V^{+}$ {\tt 0}  & First application of abstraction \\ 
+        & {\tt a} & $V \; L$         &                  & Subsequent applications of abstraction \\ 
+  \end{tabular}
+  }
+\end{table}
+
+Table~\ref{tab:crat:syntax} shows the set of proof rules for the CRAT
+format.  As with other clausal proof formats, a variable is
+represented by a positive integer $v$, with the first ones being input
+variables and successive ones being extension variables.  Literal $l$
+is represented by a signed integer, with $-v$ being the complement of
+variable $v$.  Each clause is indicated by a positive integer
+identifier $C$, with the first ones being input clauses and successive
+ones being added clauses.  Clause identifiers must be totally ordered,
+such that clause $C$ can only reference clauses $C'$ such that $C' <
+C$.  However, clause identifiers need not be consecutive.
+
+The first set of proof rules are similar to those in other clausal
+proofs.  Our syntax optionally allows input clauses to be listed with
+a rule of type {\tt i}.  Clauses can be added via
+blocked-clause and RUP/RAT rules.  The hints portion of a
+blocked-clause addition lists all earlier clauses containing
+the negated version of the pivot literal, with the clause IDs negated.  The hints portion of a
+resolution addition must contain a sequence of clause IDs such that
+the added clause is RUP/RAT with respect to these clauses.  Clause
+deletion requires an explicit justification that the deleted clauses
+is RUP/RAT with respect to other clauses.
+
+The second set of proof rules is unique to the CRAT format.  Each of
+these indicates the addition of an extension variable.  For each case,
+the rule must be followed by a sequence of blocked-clause additions.
+A product rule of the form ${\tt p}\;v\;l_1\;l_2$ indicates that $v$
+will represent the product $l_1 \pand l_2$.  The blocked clause
+additions must encode the formula $v \leftrightarrow (l_1 \land l_2)$.
+Literals $l_1$ and $l_2$ must have disjoint dependency sets.
+
+A sum rule of the form ${\tt s}\;v\;l_1\;l_2$ indicates that $v$ will
+represent the disjunction $l_1 \por l_2$.  The blocked clause
+additions must encode the formula $v \leftrightarrow (l_1 \lor l_2)$.
+The rule also contains a sequence of clause IDs such that the clause
+$\obar{l}_1 \lor \obar{l}_2$ is RUP/RAT with respect to the sequence.
+
+An abstraction rule of the form ${\tt a}\;v\;l$ indicates that
+variable $v$ will be an alias for literal $l$.  The syntax differs
+according to whether or not this is the first use of $v$ as an
+alias.  The first instance must include a list of variables that is a
+(possibly improper) superset of the dependency set for $l$.  It must
+be followed by a sequence of blocked-clause additions encoding the
+formula $v \leftrightarrow l$.  Subsequent instances do not
+include an explicit dependency set, but $\depend(l)$ must be a subset of
+those declared for $v$.  Furthermore, there are no blocked-clause
+additions for these cases.
+
+\subsection{Semantics}
+
+A CRAT proof follows the same general form as a QRAT dual proof---one
+that ensures that each clause addition and each clause deletion preserves
+equivalence.  Starting with the set of input clauses, it produces a
+sequence of steps that both add and delete clauses.  Each addition
+must be truth preserving and each deletion must be falsehood
+prserving.  At the end, all input clauses must have been deleted, and
+among the remaining clauses there must be only a single unit clause consisting of some variable or
+its complement.  Except for trivial cases, the final variable will be
+an extension variable.  The sequence of extension operations define
+the schema for computing the cost function.  That is, a value for the
+cost function is computed for every literal according to the rules of
+Lemmas \ref{lemma:complementation}--\ref{lemma:disjunction}.  For
+every abstraction step, all literals having the indicated extension
+variable as their alias must evaluate to the same value.  The computed
+cost is then the value computed for the literal in the final unit
+clause.
+
+\subsection{Example}
+
+\begin{figure}
+  \centering{
+    \begin{minipage}{0.4\textwidth}
+    A) Logical Schema\\[2ex]
+    \centering{\input{dd/c3}}
+    \end{minipage}
+    \begin{minipage}{0.4\textwidth}
+    B) Evaluation Schema\\[2ex]
+    \centering{\input{dd/c3-eval}}
+    \end{minipage}
+  }
+\caption{Schema for Formula $x_1 \lor x_2 \lor x_3$}
+\label{fig:c3:schema}
+\end{figure}
+
+\begin{figure}
+  \centering{
+  \begin{tabular}{llllll}
+    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
+\midrule
+    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause}\\
+            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
+    {\tt 2} & {\tt ab}  & {\tt 4  2 -3 0}  & {\tt 0}    & & Defining clauses for $p_4$ \\ 
+    {\tt 3} & {\tt ab}  & {\tt -4 -2 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
+    {\tt 4} & {\tt ab}  & {\tt -4 3 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
+            & {\tt s}   & {\tt 5 2 4}   & {\tt 3 0}  & & Declare $s_5 = x_2 \por p_4$ \\
+    {\tt 5} & {\tt ab}  & {\tt -5 2 4 0}  & {\tt 0}    & & Defining clauses for $s_5$ \\ 
+    {\tt 6} & {\tt ab}  & {\tt  5 -2 0}    & {\tt -5 0} &  \\  %%Defining clause for $s_5$ \\ 
+    {\tt 7} & {\tt ab}  & {\tt  5 -4 0}    & {\tt -5 0} & &  \\  %%Defining clause for $s_5$ \\ 
+            & {\tt p}   & {\tt 6 -1 5}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
+    {\tt 8} & {\tt ab}  & {\tt 6  1 -5 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
+    {\tt 9} & {\tt ab}  & {\tt -6 -1 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
+    {\tt 10} & {\tt ab}  & {\tt -6 5 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
+             & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
+    {\tt 11} & {\tt ab}  & {\tt -7 1 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
+    {\tt 12} & {\tt ab}  & {\tt  7 -1 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
+    {\tt 13} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
+    {\tt 14} & {\tt ar}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
+             & {\tt dr}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
+  \end{tabular}
+  }
+  \caption{CRAT Proof Steps for Formula $x_1 \lor x_2 \lor x_3$}
+  \label{fig:c3:crat}
+\end{figure}
+
+As an illustration, consider the Boolean formula $x_1 \lor x_2 \lor
+x_3$, represented by a single clause.  We cannot directly use the
+$\por$ operation to form this disjunction, since the sets of assignments
+satisfying the individual literals are not disjoint.  Instead, we must
+decompose this formula into a sequence of operations, as shown by the
+logical schema of Figure~\ref{fig:c3:schema}A\@.  Circles in this schema
+correspond to $\pand$ operations and squares to $\por$ operations.
+The subscripts of the variables and the operator labels correspond to
+the numbers of the input and extension variables in the CRAT proof.
+
+The conjunction of $\obar{x}_2$ and $x_3$ can be computed as $p_4 =
+\obar{x}_2 \pand x_3$, since the literals have disjoint dependency
+sets. We can then express the disjunction $x_2 \lor x_3$ as
+as $s_5 = x_2 \por p_4$.  A similar process forms the
+disjunction $x_1 \lor x_2 \lor x_3$ by first forming the product $p_6
+= \obar{x}_1 \pand s_5$ and the final sum $s_7 = x_1 \por p_6$.
+
+The logical representation can readily be converted into a schema for
+computing the cost of the formula, given weight $w_i$ for each
+variable $x_i$ for $1 \leq i \leq 3$.  This is illustrated in
+Figure~\ref{fig:c3:schema}B\@.  This schema is valid for any cost
+function, since it contains no abstraction operations.
+
+Figure~\ref{fig:c3:crat} shows an annotated version of the CRAT proof
+for this example.  Clause \#1 corresponds to the input formula, and
+clauses \#2--\#13 are the defining clauses for the four operations.
+Clause \#13 adds the unit clause corresponding to sum $s_7$,
+indicating that the extension variable variable will evaluate to
+$\tautology$.  for any assigment that satisfies the input clause.  We
+can write this as $C \turnstile  s_7$ for input clause $C$.  The
+deletion step at the end turns this around, showing that
+$s_7 \turnstile C_I$, and therefore the input clause can be deleted,
+This completes a proof that extension variable $s_7$ is logically
+equivalent to the input formula.
+
+\section{What's known, and what's not}
+
+Given an arbitrary CNF formula, we can use BDD operations to generate
+a schematic representation.  The proof generation can follow the
+methods we have used for generating unsatisfiability proofs of Boolean
+formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
+Boolean formulas~\cite{bryant:cade:2021}.  The key idea of that work
+was to use extended resolution to encode the semantics of the BDD
+nodes as part of the proof.  Here we can further decompose each $\ite$
+(short for ``if-then-else'') operation representing a BDD node into
+two $\pand$ and one $\por$ operation.  That is, we can encode
+$\ite(x,A,B$ as $(x \pand A) \por (\obar{x} \pand B$, where $x$ is an
+input variable and $A$ and $B$ are subformulas.  In doing so, we must
+make sure that $x$ is not in the dependency set of either $A$ or $B$.
+For ordered BDDs,~\cite{Bryant:1986} this property holds, since the
+variables in $A$ and $B$ must be greater in the variable ordering than
+$x$.
+
+The BDD representations of many of the formulas occuring in
+model-counting problems are far to large for this approach to be
+practical.  One refinement of the approach, implemented by the ADDMC
+model counter, is to abstract subformulas, keeping track only of the
+number of models they can have, representing a number of subformulas
+with a single ADD leaf node.  Our abstraction operation is intended to
+support this capability.  Under controlled conditions, we can
+introduce a new variable the represents the cost function for some
+subformula.  If subsequent formulas are guaranteed to yield the same
+cost, then these can alias to the earlier variable.  However, it is
+not clear how to prove that this aliasing preserves equivalence.
+
+Other model counters use either top-down and bottom-up methods to
+generate representations of the formula that are similar to our
+schmas.  These exploit the key abstractions we have identified.  We
+must find ways to modify these model counters to also generate CRAT
+proofs.
 
@@
 \end{document}
-
-Consider Boolean formulas over a set of variables $X$.  An
-assignment $\interp$ is a function mapping each variable to 
-truth value $\tautology$ (true) or $\nil$ (false).  We can extend
-$\interp$ to Boolean formulas in the normal way, such that
-$\interp(F)$ will be $\tautology$ (respectively, $\nil$) if the
-evaluation of formula $F$ yields $\tautology$ (resp., $\nil$) when its
-variables are assigned values according to $\interp$.
-
-For Boolean formula $F$, we define its set of {\em models} $\interpset{F}$ as
-\begin{eqnarray}
-\interpset{F} & = & \{ \interp | \interp(F) = \tautology \}
-\end{eqnarray}
-The task of model counting is, given formula $F$, to determine the size of its set of models
-$\mcount(F) = |\interpset{F}|$.  Ideally, this should be done without actually enumerating the set.
-
-For assignment $\interp$ and a Boolean formula $E$ over $X$, we
-use the notation $\interp\subs{x}{E}$ to denote the assignment
-$\interp'$, such that $\interp'(y) = \interp(y)$ for all $y \not = x$
-and $\interp'(x) = \interp(E)$.  In particular, the notation
-$\interp\subsflip{x}$ indicates the assignment in which the value
-assigned to $x$ is complemented, while others remain unchanged.
-
-A Boolean formula $F$ is said to be {\em independent} of variable $x$
-if every $\interp \in \interpset{F}$ has $\interp\subsflip{x} \in
-\interpset{F}$.
-
-The following lemmas describe the primary abstractions used to perform
-model counting without enumerating all solutions.
-
-\begin{lemma}
-  \label{lemma:complement}
-  For any formula $F$:
-  \begin{eqnarray*}
-    \mcount(\neg F) & =  & 2^n - \mcount(F)
-  \end{eqnarray*}
-\end{lemma}
-
-\begin{lemma}
-  \label{lemma:product}
-  For a partitioning of the variables $X$ into disjoint sets $X_1$ and $X_2$, and for formulas $F_1$ defined over $X_1$ and $F_2$ defined over $X_2$:
-  \begin{eqnarray*}
-    \mcount(F_1 \land F_2) &=& \mcount(F_1) \cdot \mcount(F_2)
-  \end{eqnarray*}
-\end{lemma}
-
-\begin{lemma}
-  \label{lemma:sum}
-  For formulas $F_1$ and $F_2$ such that $\interpset{F_1} \cap \interpset{F_2} = \emptyset$:
-  \begin{eqnarray*}
-    \mcount(F_1 \lor F_2) &=& \mcount(F_1) + \mcount(F_2)
-  \end{eqnarray*}
-\end{lemma}
-
-\section{ITE Operation}
-
-We consider a single Boolean operation, known as
-``If-Then-Else,'' or simply ``ITE.''  For Boolean values $a, b, c$,
-the operation is defined as $\ite(a,b,c) = (a \land b) \lor (\neg a
-\land c)$.  This single operation can be used to express several common Boolean operations:
-\begin{eqnarray*}
-\neg a & = & \ite(a, \nil, \tautology) \\
-a \land b & = & \ite(a, b, \nil) \\
-a \lor b & = & \ite(a, \tautology, b) \\
-a \rightarrow b & = & \ite(a, b, \tautology) \\
-\end{eqnarray*}
-
-For variable $x$ and Boolean formulas $F$ and $G$, the set of models
-for $\ite(x, F, G)$ is given by the formula:
-\begin{eqnarray}
-  \interpset{\ite(x, F, G)} & = & \{ \interp \in \interpset{F} | \interp(x) = \tautology \}
-  \cup \{ \interp \in \interpset{G} | \interp(x) = \nil \} \label{eqn:ite:interp}
-\end{eqnarray}
-
-\section{ITE Graphs}
-
-An {\em ITE graph} describes a combinational logic circuit constructed from two-input multiplexors.
-It is defined as a directed acyclic graph
-with three node types:
-\begin{description}
-\item[Constant:] Corresponds to value $\tautology$ or $\nil$.  A constant node has no incoming arcs.
-\item[Input:] Corresponds to one of the variables in $X$.   A variable node has no incoming arcs.
-\item[Operator:] Represents an application of the $\ite$ operation.
-An operator node has three incoming arcs, labeled \ifarg{}, \thenarg{}, and \elsearg{}, corresponding to the three arguments of the $\ite$ operation.
-\end{description}
-
-When describing ITE graphs, we refer to constant and input
-nodes by their associated value or variable.  An operator node is described by an expression of the form
-$\ite(v_i, v_t, v_e)$, where $v_i$, $v_t$, and $v_e$ are the
-nodes corresponding to the \ifarg{}, \thenarg{}, and \elsearg{} inputs, respectively.
-
-We can define the function $\depend$ mapping each node
-$v$ in an ITE graph to the set of variables on which it logically depends.  This can be expressed recursively as
-\begin{eqnarray}
-\depend(\tautology) & = & \emptyset \label{eqn:depend:true}\\
-\depend(\nil) & = & \emptyset \label{eqn:depend:false}\\
-\depend(x) & = & \{x \} \label{eqn:depend:variable}\\
-\depend(\ite(x, v_t, v_e)) & = & \depend(v_i) \cup \depend(v_t) \cup \depend(v_e) \label{eqn:depend:ite}
-\end{eqnarray}
-
-\begin{lemma}
-\label{lemma:independent:dset}
-Any node $v$ in an ITE graph is independent of any variable $y \in X$ such that $y \not \in \depend(v)$.
-\end{lemma}  
-\noindent
-{\em Proof:} Follows by the recursive definition of $\depend$ (\ref{eqn:depend:true}--\ref{eqn:depend:ite}) and by (\ref{eqn:ite:interp}).
-
-A {\em variable-partitioned} ITE (VPITE) graph is one satisfying the property
-that for every operator node $v = \ite(v_i, v_t, v_e)$, the dependency
-set of its \ifarg{} input most be disjoint from those of its
-\thenarg{} and \elsearg{} inputs.  That is, both $\depend(v_i) \cap \depend(v_t) = \emptyset$ and
-$\depend(v_i) \cap \depend(v_e) = \emptyset$.
-
-\section{Model Counting VPITE Graphs}
-
-While model counting is a difficult problem for arbitrary
-formulas, it readily be computed when $F$ takes the form of a VPITE graph.
-
-\begin{theorem}
-\label{theorem:count}
-Letting $n = |X|$, the following recursive formula holds for any node in a VPITE graph:
-\begin{eqnarray}
-\mcount(\tautology) & = & 2^{n} \label{eqn:count:true}\\
-\mcount(\nil) & = & 0 \label{eqn:count:false}\\
-\mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
-\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2^n} \label{eqn:count:ite}
-\end{eqnarray}  
-\end{theorem}
-\noindent
-{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
-from Lemmas~\ref{lemma:complement}--\ref{lemma:sum}, along with the definition of the ITE operator.
-
-\section{Relation to Other Representations}
-
-In general, there is a trade-off between the compactness of different
-Boolean function representations versus the ease with which different
-properties can readily be computed.  For example, Boolean formulas and
-Boolean circuits can be very compact, but basic properties of these
-representations, such as satisfiability, equivalence testing, and
-model counting are NP-hard.  On the other hand all of these properties
-can be computed in polynomial time for reduced, orderd binary decision
-diagrams (ROBDDs)~\cite{Bryant:1986}, but they can grow exponentially
-in relation to the representation of the function as a formula or
-circuit.  ITE graphs lie somewhere between these two extremes.  Here
-we examine how they relate to other representations and the implications
-of these relations.
-
-\subsection{Binary Decision Diagrams}
-
-Binary Decision Diagrams~\cite{akers:ieeetc:1978} also represent Boolean formulas
-as directed graphs consisting of ITE operators.  They can be seen as
-equivalent to a class of graphs we will call {\em input-controlled} ITE
-graphs.  These are ones for which the \ifarg{} input to every operator
-node is a variable.
-
-If we combine the properties of variable partitioned and input
-control, we arrive at the representation known as {\em
-  Free}~BDDs~\cite{gergov:ieeetc:1994}.  If we further restrict the
-variables in a variable-controlled, input-partitioned ITE graph to
-obey the same ordering along all paths formed in the graph, we arrive
-at OBDDs.  An OBDD can be reduced by a simple set of rules to give an
-ROBDD representation, and this form is canonical~\cite{Bryant:1986}.
-
-\subsection{Sententential Decision Diagrams}
-
-Darwiche uses the Sentential Decision Diagram (SDD)
-representation~\cite{darwiche:ijcai:2011} as his framework for several
-model counting programs.  An SDD can be seen as a combination of 1) a
-generalization of VPITE graphs consisting of $k$-input multiplexors,
-and 2) an ordering restriction on the variables.
-That is, each operator node in an SDD is defined to
-have $k$ data inputs and $k$ control inputs, where the Boolean
-functions corresponding to the control inputs $p_1, p_2, \ldots, p_k$
-satisfy $p_i \land p_j = \nil$ for $1 \leq i < j \leq k$ and
-$\bigvee_{i=1,k} p_i = \tautology$.  The $k$ data inputs represent
-Boolean functions $s_1, s_2, \ldots, s_k$.  In addition, the
-dependency sets for the control and data inputs must be disjoint:
-$\depend(p_i) \cap \depend(s_j) = \emptyset$ for $1 \leq i,j \leq k$.
-Furthermore, the ordering of the variables in an SDD must obey a hierarchical decomposition
-defined by
-a tree structure known as a {\em vtree}.  Indeed, an SDD can be seen
-as equivalent to a generalization of BDDs devised by McMillan, which
-he termed {\em tree} BDDs~\cite{mcmillan:cav:1994}.
-
-The graph part of an SDD can be readily translated into a VPITE graph
-by expanding each $k$-input node into $k-1$ ITE nodes, giving an
-output value defined as:
-\begin{displaymath}
-\ite(p_1, s_1, \ite(p_2, s_2, \ite( \cdots \ite(p_{k-1}, s_{k-1}, s_k)\cdots))
-\end{displaymath}
-If we express the size of an SDD as the number of branches, this
-translation from SDD to ITE graph will expand the size by a constant
-factor.
-
-\section{Probabilistic Equivalence Checking}
-
-One further feature of VPITE graphs is that there is a randomized
-polynomial time (RP) algorithm for determining whether two graphs
-represent equivalent Boolean functions.  This algorithm generalizes
-the RP equivalence-testing algorithm for free BDDs described by Blum,
-Chandra, and Wegman~\cite{blum:ipl:1980}.
-
-\bibliography{references}
-
-\end{document}
-
-\section{A Model Counting Calculus}
-
-Objective: Create a simple system for defining and reasoning about an
-abstraction of Boolean formulas that suffices for tracking the number
-of models.
-
-Define a {\em counting subspace} $\subspace{} = \langle V, \alpha, k\rangle$, with:
-\begin{itemize}
-\item $V$ is a set of variables
-\item $\alpha$ is a partial assignment to variables in $V$, represented as a set of literals.
-\item $k$ is an integer, indicating the number of models for
-  the formula within the specified subspace.  For $n = |V|$ and $m= |\var(\alpha)|$, it satisfies
-  $0 \leq k \leq 2^{n-m}$.
-\end{itemize}
-
-\subsection{Subspace Primitives}
-\begin{itemize}
-\item
-Universal subspace: $\langle \emptyset, \emptyset, 1 \rangle$.
-\item
-Null subspace: $\langle \emptyset, \emptyset, 0 \rangle$.
-\item
-  Positive literal: $\langle \{ x\}, \{x\}, 1\rangle$.
-\item
-  Negative literal: $\langle \{ x\}, \{\obar{x}\}, 1\rangle$.
-\item
-  Unused variable: $\langle \{ x\}, \emptyset, 2\rangle$.
-\end{itemize}
-
-\subsection{Basic Subspace Operations}
-
-Assume $\subspace_1 = \langle V_1, \alpha_1, k_1\rangle$ and
-$\subspace_2 = \langle V_2, \alpha_2, k_2\rangle$ are two subspaces
-
-Subspace product $\subspace_1 \times \subspace_2$
-\begin{itemize}
-\item Condition: $V_1 \cap V_2 = \emptyset$
-\item Result: $\langle V_1 \cup V_2, \alpha_1 \cup \alpha_2, k_1 \cdot k_2\rangle$.
-\end{itemize}
-
-Subspace Sum $\subspace_1 + \subspace_2$
-\begin{itemize}
-\item Conditions:
-  \begin{itemize}
-  \item $V_1 = V_2$.
-  \item For some literal $l$ of variable $x \in V_1$ and assignment $\alpha$, the argument assignments satisfy
-    $\alpha_1 = \alpha \cup \{l\}$ and $\alpha_2 = \alpha \cup \{ \obar{l} \}$.
-  \end{itemize}
-\item Result: $\langle V_1 - \{ x\}, \alpha, (k_1 + k_2)/2\rangle$.
-\end{itemize}
-
-\subsection{Derived Subspace Operations}
-
-These useful operations can be constructed by combining primitive values and basic operations.
-
-Expanding subspace $\subspace{} = \langle V, \alpha, k\rangle$:
-\begin{itemize}
-\item
-  Condition: $x \not \in V$
-\item
-  Results:
-  \begin{itemize}
-    \item Positive expansion: $\langle V \cup \{ x \}, \alpha \cup \{ x \}, k \rangle$.
-    \item Negative expansion: $\langle V \cup \{ x \}, \alpha \cup \{ \obar{x} \}, k \rangle$.
-    \item Don't-care expansion: $\langle V \cup \{ x \}, \alpha , 2\cdot k \rangle$.
-  \end{itemize}
-\end{itemize}
-
-A particularly useful expansion is to apply the don't-care expansion
-with a set of variables $U$ such that $U \cap V = \emptyset$.  This
-will give the subspace $\langle U \cup V, \alpha, 2^{|U|} \cdot
-k\rangle$.

### 2022-04-11 doc/dd/c3-eval.tex
Ready for review

\begin{tikzpicture}
\definecolor{neutralcolor}{RGB}{0,0,0}
\definecolor{pathcolor}{RGB}{0,0,0}
\definecolor{fillcolor}{RGB}{255,255,255}
\definecolor{background}{RGB}{225,225,225}
\definecolor{highcolor}{RGB}{0,0,0}
\definecolor{lowcolor}{RGB}{0,0,0}
\draw (1.32,6.80) [thin,neutralcolor] -- (0.26,5.74);
\draw (1.32,6.80) [thin,neutralcolor] -- (1.68,5.03);
\draw (1.68,5.03) [thin,neutralcolor] -- (2.74,3.97);
\draw (1.68,5.03) [thin,neutralcolor] -- (1.32,3.27);
\draw (1.32,3.27) [thin,neutralcolor] -- (0.26,2.21);
\draw (1.32,3.27) [thin,neutralcolor] -- (1.68,1.50);
\draw (1.68,1.50) [thin,neutralcolor] -- (2.74,0.44);
\draw (1.68,1.50) [thin,neutralcolor] -- (0.62,0.44);
\draw [thin,fill=fillcolor,draw=fillcolor] (1.68,5.03) circle [radius=0.26];
\node at (1.68,5.03) {{\texttt{*}}};
\draw [thin,fill=fillcolor,draw=fillcolor] (1.68,1.50) circle [radius=0.26];
\node at (1.68,1.50) {{\texttt{*}}};
\draw [thin,fill=fillcolor,draw=fillcolor] (2.74,0.44) circle [radius=0.44];
\node at (2.74,0.44) {$1-w_2$};
\draw [thin,fill=fillcolor,draw=fillcolor] (1.32,6.80) circle [radius=0.26];
\node at (1.32,6.80) {{\texttt{+}}};
\draw [thin,fill=fillcolor,draw=fillcolor] (1.32,3.27) circle [radius=0.26];
\node at (1.32,3.27) {{\texttt{+}}};
\draw [thin,fill=fillcolor,draw=fillcolor] (2.74,3.97) circle [radius=0.44];
\node at (2.74,3.97) {$1-w_1$};
\draw [thin,fill=fillcolor,draw=fillcolor] (0.26,2.21) circle [radius=0.26];
\node at (0.26,2.21) {$w_2$};
\draw [thin,fill=fillcolor,draw=fillcolor] (0.62,0.44) circle [radius=0.26];
\node at (0.62,0.44) {$w_3$};
\draw [thin,fill=fillcolor,draw=fillcolor] (0.26,5.74) circle [radius=0.26];
\node at (0.26,5.74) {$w_1$};
\node [right] at (1.59,6.80) {$s_7$};
\node [left] at (1.41,5.03) {$p_6$};
\node [right] at (1.59,3.27) {$s_5$};
\node [left] at (1.41,1.50) {$p_4$};
\end{tikzpicture}%%

### 2022-04-11 doc/dd/c3.tex
Ready for review

\begin{tikzpicture}
\definecolor{neutralcolor}{RGB}{0,0,0}
\definecolor{pathcolor}{RGB}{0,0,0}
\definecolor{fillcolor}{RGB}{255,255,255}
\definecolor{background}{RGB}{225,225,225}
\definecolor{highcolor}{RGB}{0,0,0}
\definecolor{lowcolor}{RGB}{0,0,0}
\draw (1.32,6.62) [thin,neutralcolor] -- (0.26,5.56);
\draw (1.32,6.62) [thin,neutralcolor] -- (1.68,4.85);
\draw (1.68,4.85) [thin,neutralcolor] -- (2.74,3.79);
\draw (1.68,4.85) [thin,neutralcolor] -- (1.32,3.09);
\draw (1.32,3.09) [thin,neutralcolor] -- (0.26,2.03);
\draw (1.32,3.09) [thin,neutralcolor] -- (1.68,1.32);
\draw (1.68,1.32) [thin,neutralcolor] -- (2.74,0.26);
\draw (1.68,1.32) [thin,neutralcolor] -- (0.62,0.26);
\draw [thin,fill=fillcolor,draw=neutralcolor] (1.68,4.85) circle [radius=0.53];
\node at (1.68,4.85) {$\mathbin{\land_{\textrm{pv}}}$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (1.68,1.32) circle [radius=0.53];
\node at (1.68,1.32) {$\mathbin{\land_{\textrm{pv}}}$};
\draw [thin,fill=fillcolor,draw=fillcolor] (2.74,0.26) circle [radius=0.26];
\node at (2.74,0.26) {$\overline{x}_2$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (0.79,6.09) rectangle (1.85,7.15);
\node at (1.32,6.62) {$\mathbin{\lor_{\textrm{pa}}}$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (0.79,2.56) rectangle (1.85,3.62);
\node at (1.32,3.09) {$\mathbin{\lor_{\textrm{pa}}}$};
\draw [thin,fill=fillcolor,draw=fillcolor] (2.74,3.79) circle [radius=0.26];
\node at (2.74,3.79) {$\overline{x}_1$};
\draw [thin,fill=fillcolor,draw=fillcolor] (0.26,2.03) circle [radius=0.26];
\node at (0.26,2.03) {$x_2$};
\draw [thin,fill=fillcolor,draw=fillcolor] (0.62,0.26) circle [radius=0.26];
\node at (0.62,0.26) {$x_3$};
\draw [thin,fill=fillcolor,draw=fillcolor] (0.26,5.56) circle [radius=0.26];
\node at (0.26,5.56) {$x_1$};
\node [right] at (1.85,6.62) {$s_7$};
\node [left] at (1.15,4.85) {$p_6$};
\node [right] at (1.85,3.09) {$s_5$};
\node [left] at (1.15,1.32) {$p_4$};
\end{tikzpicture}%%

### 2022-04-11 doc/mc-notes.tex
Ready for review

@@
 $\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
-assignment will be $1/2^{n}$, and the cost of a function will be its
-{\em density}, denoted $\density(f)$.  That is $\density(f) \in [0,
-  1]$ will be the fraction of assignments for which the function
-evaluates to 1.  Given an ability to compute the density of function
-$f$, we can scale this by $2^n$ to compute the total number of models
+assignment is then $1/2^{n}$, and the cost of a function is its
+{\em density}, denoted $\density(f)$.  That is, the density of $f$ satisfies
+$0 \leq \density(f) \leq 1$.  It
+is the fraction of assignments for which the $f$
+evaluates to 1, with $\density(\zero) = 0$ and $\density(\one) = 1$.  The density of a function
+$f$ can be scaled by $2^n$ to compute the total number of models
 $|\interpset{f}|$.  This is the core task of model counting.  Using
-density as our metric has the advantage that it does not vary when the
+density as the metric, rather than the number of models, has the advantage that it does not vary when the
 function is embedded in a larger domain $X' \supseteq X$.  As a
-variant, we can support some forms of weighted model counting by
+variant, some other forms of weighted model counting can be supported by
 assigning weights of the form $\cost(x_i) = w_i$, where $0 \leq w_i
@@
 conjunctive normal form, according to a specified cost function.  A
-CRAT proof provides two capabilities: 1) a {\em schema} for computing a cost
+CRAT proof provides two forms of information: 1) a {\em schema} for computing a cost
 function using the ring operations, and 2) a {\em proof} that this schema
@@
 
-The CRAT format draws its inspiration from the QRAT format for
+The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
+QRAT format for
 quantified Boolean formulas (QBF).  The following are its key properties:
 \begin{itemize}
-\item Clause addition is allowed, as long as either 1) the new clause is
-  blocked, or 2) it satisfies the RUP/RAT property with respect to a sequence
+\item As with LRAT, a clause can be added as long as either 1) it is
+  blocked, or 2) it satisfies the RUP/RAT property with respect to a supplied sequence
   of earlier antecedent clauses.
-\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and {\em abstraction}
+\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and to an abstraction rule.
 \begin{itemize}
@@
 \end{itemize}
-\item Unlike DRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
+\item Unlike LRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
 \item Unlike QRAT, it need not support universal quantification.
@@
 must be truth preserving and each deletion must be falsehood
-prserving.  At the end, all input clauses must have been deleted, and
+preserving.  At the end, all input clauses must have been deleted, and
 among the remaining clauses there must be only a single unit clause consisting of some variable or
@@
     \begin{minipage}{0.4\textwidth}
-    A) Logical Schema\\[2ex]
+      A) Logical Form   \\[2ex]
     \centering{\input{dd/c3}}
@@
     \begin{minipage}{0.4\textwidth}
-    B) Evaluation Schema\\[2ex]
+    B) Evaluation Form\\[2ex]
     \centering{\input{dd/c3-eval}}
@@
 x_3$, represented by a single clause.  We cannot directly use the
-$\por$ operation to form this disjunction, since the sets of assignments
+$\por$ operation to form these disjunctions, since the sets of assignments
 satisfying the individual literals are not disjoint.  Instead, we must
 decompose this formula into a sequence of operations, as shown by the
-logical schema of Figure~\ref{fig:c3:schema}A\@.  Circles in this schema
+logical form of Figure~\ref{fig:c3:schema}A\@.  Circles in this schema
 correspond to $\pand$ operations and squares to $\por$ operations.
@@
 clauses \#2--\#13 are the defining clauses for the four operations.
-Clause \#13 adds the unit clause corresponding to sum $s_7$,
+Each of the two sum operations lists one of the earlier defining
+clauses as a proof that its arguments are mutually exclusive.
+Clause \#14 adds the unit clause corresponding to sum $s_7$,
 indicating that the extension variable variable will evaluate to
-$\tautology$.  for any assigment that satisfies the input clause.  We
+$\tautology$  for any assigment that satisfies the input clause.  We
 can write this as $C \turnstile s_7$ for input clause $C$.  The
-deletion step at the end turns this around, showing that
-$s_7 \turnstile C_I$, and therefore the input clause can be deleted,
-This completes a proof that extension variable $s_7$ is logically
+deletion step at the end turns this around, showing that $s_7
+\turnstile C_I$, and therefore the input clause can be deleted, This
+completes a proof that extension variable $s_7$ is logically
 equivalent to the input formula.
 
-\section{What's known, and what's not}
+\section{Looking Ahead}
+
+\subsection{Implementing Certified Counters}
 
@@
 formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
-Boolean formulas~\cite{bryant:cade:2021}.  The key idea of that work
-was to use extended resolution to encode the semantics of the BDD
+Boolean formulas~\cite{bryant:cade:2021}.  The key idea 
+is to use extended resolution to encode the semantics of the BDD
 nodes as part of the proof.  Here we can further decompose each $\ite$
@@
 two $\pand$ and one $\por$ operation.  That is, we can encode
-$\ite(x,A,B$ as $(x \pand A) \por (\obar{x} \pand B$, where $x$ is an
-input variable and $A$ and $B$ are subformulas.  In doing so, we must
+$\ite(x,A,B)$ as $(x \pand A) \por (\obar{x} \pand B)$, where $x$ is an
+input variable and $A$ and $B$ are subformulas.
+The sum operation trivially satisfies the disjoint assignment requirement, since $x$ has positive polarity in the first product and negative polarity in the second.
+We must also
 make sure that $x$ is not in the dependency set of either $A$ or $B$.
-For ordered BDDs,~\cite{Bryant:1986} this property holds, since the
-variables in $A$ and $B$ must be greater in the variable ordering than
+For ordered BDDs,~\cite{Bryant:1986} this property holds, because the
+variables in $A$ and $B$ will be greater in the variable ordering than
 $x$.
@@
 model counter, is to abstract subformulas, keeping track only of the
-number of models they can have, representing a number of subformulas
+number of models they can have and representing a number of subformulas
 with a single ADD leaf node.  Our abstraction operation is intended to
-support this capability.  Under controlled conditions, we can
-introduce a new variable the represents the cost function for some
+support this capability.  We can
+introduce a new variable to represent the cost function for some
 subformula.  If subsequent formulas are guaranteed to yield the same
@@
 
+\subsection{TO-DO List}
+
+\begin{itemize}
+\item Proof Framework
+  \begin{itemize}
+  \item Generalities and details of the format
+  \item What should be the final state?
+  \item When can blocked clauses be deleted?
+  \item How can abstraction be incorporated?
+  \end{itemize}
+
+\item Checker
+  \begin{itemize}
+  \item Working prototype
+  \item C/C++ (or Rust?)
+  \item Formally verified
+  \end{itemize}
+
+\item Counters
+\begin{itemize}
+\item BDD-based
+  \begin{itemize}
+  \item Prototype
+  \item C/C++
+  \end{itemize}
+
+\item CDCL-based
+  \begin{itemize}
+  \item Prototype
+  \item C/C++
+  \end{itemize}
+\end{itemize}
+
+\end{itemize}
+
 \bibliography{references}

### 2022-04-23 doc/mc-notes.tex
New example

@@
 \newcommand{\boolor}{\lor}
-\newcommand{\pand}{\mathbin{\land_{\rm pv}}}
-\newcommand{\por}{\mathbin{\lor_{\rm pa}}}
+\newcommand{\pand}{\mathbin{\land_{\sf v}}}
+\newcommand{\por}{\mathbin{\lor_{\sf a}}}
 \newcommand{\boolxor}{\oplus}
@@
 \newcommand{\ite}{\mbox{\it ITE}}
+\newcommand{\pite}{\mbox{\it ITE}_{\sf v}}
+\newcommand{\oneminus}{{\sim}}
 
@@
 \newcommand{\interp}{\alpha}
+\newcommand{\uinterp}{{\cal U}}
 \newcommand{\interpset}[1]{{\cal M}(#1)}
@@
 $\obar{x}_i$, corresponding to the assignments $\interp(x_i)$ = 1 or 0,
-respectively.
+respectively.  We denote the set of all assignments over these variables as $\uinterp$.
 
@@
 
-{\bf Example 1}: Let $\ring$ be the set of rational numbers and
-$\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
+For ring elemement $a$, we use the notation $\oneminus a$ to denote $1 - a$.
+
+{\bf Example 1}: Let $\ring$ be the set of rational numbers of the form $a\cdot 2^b$ where $a$ and $b$ are integers.
+Let $\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
 assignment is then $1/2^{n}$, and the cost of a function is its
@@
 density as the metric, rather than the number of models, has the advantage that it does not vary when the
-function is embedded in a larger domain $X' \supseteq X$.  As a
-variant, some other forms of weighted model counting can be supported by
-assigning weights of the form $\cost(x_i) = w_i$, where $0 \leq w_i
-\leq 1$ for every variable $x_i$.
+function is embedded in a larger domain $X' \supseteq X$.
 
-{\bf Example 2}: Let $\ring$ be a field with $|\ring| > 2n$,
+{\bf Example 2}: Let $\ring$ be the set of rational numbers.  Assign a
+{\em weight} $w_i$ for each variable $x_i$ such that $0 \leq w_i \leq
+1$ and let $\cost(x_i) = w_i$.  This can support a restricted version of
+{\em weighted} model counting.  The key restrictions are: 1) the
+weight of an assignment equals the product of the weights of its
+literals, and 2) the weight of a variable $x_i$ and its complement
+$\obar{x}_i$ sum to 1.
+
+{\bf Example 3}: Let $\ring$ be a field with $|\ring| \geq 2n$,
 and let $\hashset$ be the set of functions
@@
   For separable cost function $\cost$ and Boolean function $f$:
-  $\cost(\neg f) = 1 - \cost(f)$.
+  $\cost(\neg f) = \oneminus \cost(f)$.
 \end{lemma}
@@
 
+We will also find it useful to introduce a restricted version of the
+if-then-else operator, denoted $\pite$.  That is, for functions $f_i$,
+$f_t$, and $f_e$, we define $\pite(f_i, f_t, f_e) = (f_i \pand f_t)
+\por (\neg f_i \pand f_e)$.  The two arguments to the disjunction are
+guaranteed to satisfy the constraints for an assignment-partitioned
+disjunction, since first argument can only be satisfied by assignments
+for which $f_i$ evaluates to $\tautology$, while the second can only
+be satisfied by assignments for which $f_i$ evaluates to $\nil$.  The
+only restriction of the $\pite$ operator is that
+$\depend(f_i) \cap \depend(f_t) = \emptyset$ and
+$\depend(f_i) \cap \depend(f_e) = \emptyset$.
+
+\section{Separable Schemas}
+
+\begin{table}
+  \caption{Recursive Definition of Separable Schemas}
+  \label{tab:schema}
+A) Primitive Operations\\
+\begin{center}
+\begin{tabular}{cccc}
+\toprule
+  \makebox[15mm]{$S$}           & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$}\\
+\midrule
+  $\nil$  & None  & $\emptyset$    & $\emptyset$ \\
+  $\tautology$        & None  & $\emptyset$    & $\uinterp$ \\
+$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ \\
+$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ \\
+$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$ \\
+$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$ \\
+\bottomrule
+\end{tabular}
+\end{center}
+B) Derived Operations\\
+\begin{center}
+\begin{tabular}{cc}
+\toprule
+  $S$           & Definition \\
+\midrule
+\makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
+\bottomrule
+\end{tabular}
+\end{center}
+\end{table}
+
+A {\em separable schema} is a directed acyclic graph representing a
+Boolean formula where the only allowed operations are $\neg$, $\pand$,
+$\por$ and $\pite$.  We use a graph representation to allow sharing of
+subformulas.  The set of schemas over a set of variables $\{x_1, x_2,
+\ldots, x_n\}$ can be defined recursively, as is shown in Table
+\ref{tab:schema}.  Each schema $S$ has an associated dependency set
+$\depend(S)$ and an associated set of models $\interpset{S}$.  We
+divide the operations into {\em basic} ones that are fundamental to
+the representation, and {\em derived} ones that can be constructed
+from basic operations.  The latter category consists of just the $\pite$ operation.
+
+A key property of a Boolean formula represented by separable schema
+$S$ is that, for any separable cost function $\cost$, the cost of the
+formula $\cost(S)$ can be computed with a linear number of ring
+operations.
+
+\begin{table}
+  \caption{Normalization Rules}
+  \label{tab:normalize}
+  \begin{center}
+  \begin{tabular}{ccccccc}
+    \makebox[20mm]{$\neg \nil$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\tautology$} & \makebox[15mm]{} &
+    \makebox[20mm]{$\neg \tautology$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\nil$} \\
+    $\neg \neg S$ & $\rightarrow$ & $S$ \\
+    $S \pand \nil$ & $\rightarrow$ & $\nil$ && $\nil \pand S$ & $\rightarrow$ & $\nil$ \\
+    $S \pand \tautology$ & $\rightarrow$ & $S$ && $\tautology \pand S$ & $\rightarrow$ & $S$ \\
+    $S \por \nil$ & $\rightarrow$ & $S$ && $\nil \por S$ & $\rightarrow$ & $S$ \\
+    $S \por \tautology$ & $\rightarrow$ & $\tautology$ && $\tautology \por S$ & $\rightarrow$ & $\tautology$ \\
+    $\pite(\neg S_1, S_2, S_3)$ & $\rightarrow$ & $\pite(S_1, S_3, S_2)$ \\
+    $\pite(\tautology, S_2, S_3)$ & $\rightarrow$ & $S_2$ & &
+    $\pite(\nil, S_2, S_3)$ & $\rightarrow$ & $S_3$ \\
+    $\pite(S_1, \tautology, \nil)$ & $\rightarrow$ & $S_1$ &&
+    $\pite(S_1, \nil, \tautology)$ & $\rightarrow$ & $\neg S_1$ \\
+    $\pite(S_1, S_2, \nil)$ & $\rightarrow$ & $S_1 \pand S_2$ & &
+    $\pite(S_1, \nil, S_3)$ & $\rightarrow$ & $\neg S_1 \pand S_3$ \\
+    $\pite(S_1, S_2, \tautology)$ & $\rightarrow$ & $\neg [S_1 \pand \neg S_2]$ & &
+    $\pite(S_1, \tautology, S_3)$ & $\rightarrow$ & $\neg [\neg S_1 \pand \neg S_3]$ \\
+  \end{tabular}
+  \end{center}
+\end{table}
+
+Table~\ref{tab:normalize} shows a list of {\em normalizing}
+transformations to simplify a separable schema.  These are mostly
+straightforward---eliminating extra negations and removing constant
+terms.  The most interesting are the two bottom rules, were we make
+use of DeMorgan's Laws to simplify an $\pite$ operation when one of
+the arguments is $\tautology$.  We will make use of this
+transformation when using BDDs to convert a set of clauses into a
+separable schema.
+
 \section{Proof Framework for Cost Functions}
 
-We introduce the CRAT clausal proof framework for generating a
-checkable proof of the valuation of a Boolean formula, given in
-conjunctive normal form, according to a specified cost function.  A
-CRAT proof provides two forms of information: 1) a {\em schema} for computing a cost
-function using the ring operations, and 2) a {\em proof} that this schema
-accurately characterizes the cost function for the input formula.
-A schema can be either {\em function independent}, meaning that it holds
-for any cost function, or {\em function dependent}, meaning that
-it holds for only for a specific cost function.
+The CRAT clausal proof framework provides a means for creating a
+checkable proof that a Boolean formula, given in conjunctive normal
+form, is logically equivalent to some separable schema.  This provides
+a framework for adding proof generation to model counting programs.
 
@@
 \item As with LRAT, a clause can be added as long as either 1) it is
-  blocked, or 2) it satisfies the RUP/RAT property with respect to a supplied sequence
+  blocked, or 2) it satisfies the RAT property with respect to a supplied sequence
   of earlier antecedent clauses.
-\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and to an abstraction rule.
+\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and $\pite$.
 \begin{itemize}
@@
   extension variable.  When an extension variable is introduced based
-  on the $\pand$ operation, the dependency sets of its arguments must
+  on the $\pand$ or $\pite$ operation, the dependency sets of its arguments must
   be disjoint.
 \item When an extension variable is introduced based on the $\por$
-  operation, the step must cite earlier steps providing a RUP/RAT proof
+  operation, the step must cite earlier steps providing a RAT proof
   that the two arguments are mutually exclusive.
-\item The {\em abstraction} operation allows an extension
-  variable to serve as an alias for one of more other 
-  variables.  Use of this rule renders the schema function dependent.
 \item Boolean complement is provided implicitly by allowing the
@@
 \end{itemize}
-\item Unlike LRAT, the deletion of a clause requires showing that this clause is implied by one or more other clauses.
+\item A CRAT proof must show that the schema is logically equivalent
+  to the input formula, not just that they are equisatisfiable.
+  Therefore, each deletion step must also be show to be equivalence
+  preserving, either because the clause is blocked or it follows from
+  remaining clauses by the RAT property.
 \item Unlike QRAT, it need not support universal quantification.
@@
     \midrule
-    \makebox[5mm][l]{$C$} & \makebox[7mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
+    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
     $C$ & {\tt ab} & $L^{+}$ {\tt 0} & ${\tt -}C^{*}$  {\tt 0}  & Add blocked clause \\
-    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RUP/RAT clause \\
-     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RUP/RAT clause \\
+    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RAT clause \\
+     & {\tt db} & $C$             & ${\tt -}C^{+}$  {\tt 0} & Delete blocked clause \\
+     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RAT clause \\
     \midrule
@@
         & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
-        & {\tt a} & $V \; L$         & $V^{+}$ {\tt 0}  & First application of abstraction \\ 
-        & {\tt a} & $V \; L$         &                  & Subsequent applications of abstraction \\ 
+        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
   \end{tabular}
@@
 Table~\ref{tab:crat:syntax} shows the set of proof rules for the CRAT
-format.  As with other clausal proof formats, a variable is
+format.
+As with other clausal proof formats, a variable is
 represented by a positive integer $v$, with the first ones being input
@@
 proofs.  Our syntax optionally allows input clauses to be listed with
-a rule of type {\tt i}.  Clauses can be added via
-blocked-clause and RUP/RAT rules.  The hints portion of a
-blocked-clause addition lists all earlier clauses containing
-the negated version of the pivot literal, with the clause IDs negated.  The hints portion of a
-resolution addition must contain a sequence of clause IDs such that
-the added clause is RUP/RAT with respect to these clauses.  Clause
-deletion requires an explicit justification that the deleted clauses
-is RUP/RAT with respect to other clauses.
+a rule of type {\tt i}.  Clauses can be added via blocked-clause and
+RAT rules.  As described below, however, blocked clauses can only be added
+to define $\pand$, $\por$, and $\pite$ operations.  The hints portion
+of a blocked-clause addition lists all earlier clauses containing the
+negated version of the pivot literal, with the clause IDs negated.
+The hints portion of a RAT addition must contain a sequence of clause
+IDs such that the added clause is RAT with respect to these clauses.
+Clause deletion requires an explicit justification that the deleted
+clauses is RAT with respect to other clauses.
 
@@
 these indicates the addition of an extension variable.  For each case,
-the rule must be followed by a sequence of blocked-clause additions.
+the rule must be followed by a sequence of blocked-clause additions
+providing the defining clauses for the extension variable.
+
 A product rule of the form ${\tt p}\;v\;l_1\;l_2$ indicates that $v$
@@
 The rule also contains a sequence of clause IDs such that the clause
-$\obar{l}_1 \lor \obar{l}_2$ is RUP/RAT with respect to the sequence.
+$\obar{l}_1 \lor \obar{l}_2$ is RAT with respect to the sequence.
 
-An abstraction rule of the form ${\tt a}\;v\;l$ indicates that
-variable $v$ will be an alias for literal $l$.  The syntax differs
-according to whether or not this is the first use of $v$ as an
-alias.  The first instance must include a list of variables that is a
-(possibly improper) superset of the dependency set for $l$.  It must
-be followed by a sequence of blocked-clause additions encoding the
-formula $v \leftrightarrow l$.  Subsequent instances do not
-include an explicit dependency set, but $\depend(l)$ must be a subset of
-those declared for $v$.  Furthermore, there are no blocked-clause
-additions for these cases.
+An if-then-else rule of the form ${\tt ite}\;v\;l_1\;l_2\;l_3$
+indicates that $v$ will represent $\pite(l_1, l_2, l_3)$.  The blocked
+clause additions must encode the formula
+$v \leftrightarrow \ite((l_1,l_2, l_3)$.
+The dependency set for $l_1$ must be disjoint from those
+of $l_2$ and $l_3$.
 
@@
 among the remaining clauses there must be only a single unit clause consisting of some variable or
-its complement.  Except for trivial cases, the final variable will be
-an extension variable.  The sequence of extension operations define
-the schema for computing the cost function.  That is, a value for the
-cost function is computed for every literal according to the rules of
-Lemmas \ref{lemma:complementation}--\ref{lemma:disjunction}.  For
-every abstraction step, all literals having the indicated extension
-variable as their alias must evaluate to the same value.  The computed
-cost is then the value computed for the literal in the final unit
-clause.
+its complement.  Except for trivial cases, the final literal will be
+an extension variable or its complement.  That variable will indicate the root node of the schema.
+Working from that root backward, the schema can be extracted from the CRAT file.
 
-\subsection{Example}
+\subsection{Example 1}
 
@@
     \begin{minipage}{0.4\textwidth}
-      A) Logical Form   \\[2ex]
+      A) Schema   \\[2ex]
     \centering{\input{dd/c3}}
@@
     \begin{minipage}{0.4\textwidth}
-    B) Evaluation Form\\[2ex]
+    B) Cost Computation\\[2ex]
     \centering{\input{dd/c3-eval}}
@@
   }
-\caption{Schema for Formula $x_1 \lor x_2 \lor x_3$}
+\caption{Schema for Formula $x_1 \lor x_2 \lor x_3$ and its Cost Computation}
 \label{fig:c3:schema}
@@
 satisfying the individual literals are not disjoint.  Instead, we must
-decompose this formula into a sequence of operations, as shown by the
-logical form of Figure~\ref{fig:c3:schema}A\@.  Circles in this schema
+decompose this formula into a sequence of operations.  Figure~\ref{fig:c3:schema}A shows one such decomposition.
+Circles in this schema
 correspond to $\pand$ operations and squares to $\por$ operations.
@@
 equivalent to the input formula.
+
+\subsection{Example 2}
+
+\begin{figure}
+  \centering{
+  \begin{tabular}{llllll}
+    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
+\midrule
+    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause $C_1$}\\
+
+    {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
+            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
+    {\tt 3} & {\tt ab}  & {\tt 4 2 3 0} & {\tt 0}    & & Defining clauses for $p_4$ \\
+    {\tt 4} & {\tt ab}  & {\tt -4 -2 0} & {\tt -2 0}    & & \\
+    {\tt 5} & {\tt ab}  & {\tt -4 -3 0} & {\tt -2 0}    & & \\
+            & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
+    {\tt 6} & {\tt ab}  & {\tt 5 1 -4 0} & {\tt 0}    & & Defining clauses for $p_5$ \\
+    {\tt 7} & {\tt ab}  & {\tt -5 -1 0} & {\tt -6 0}    & & \\
+    {\tt 8} & {\tt ab}  & {\tt -5 4 0} & {\tt -6 0}    & & \\
+    {\tt 9} & {\tt ar}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
+            & {\tt dr}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
+            & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
+    {\tt 10} & {\tt ab}  & {\tt 6 -1 2 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
+    {\tt 11} & {\tt ab}  & {\tt -6 1 0}    & {\tt -10 0} & & \\
+    {\tt 12} & {\tt ab}  & {\tt -6 -2 0}    & {\tt -10 0} & & \\ 
+    {\tt 13} & {\tt ar} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
+            & {\tt dr}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
+            & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
+    {\tt 14} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
+    {\tt 15} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -5 0} &  \\  
+    {\tt 16} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -5 0} & &  \\
+    {\tt 17} & {\tt ar}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
+             & {\tt dr}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
+             & {\tt dr}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
+  \end{tabular}
+}  
+  \caption{CRAT Proof Steps for Formula $(x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2$}
+  \label{fig:p2:crat}
+\end{figure}
+
+As a more complex example, consider the Boolean formula given by the
+conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
+\obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
+of DeMorgan's Laws to provide a more direct encoding of the problem.
+The complement of the formula can be written in DNF as
+$\obar{x}_1\,\obar{x}_2\,\obar{x}_3 \lor x_1\,\obar{x}_2$.  Each of
+the two conjuncts can be formed using $\pand$ operations, and their
+sum can be formed using a $\por$ operation.  In this example, we first
+form a term $t_1$ that equals $C_1$ and a term $t_2$ that equals
+$C_2$.  These are asserted as unit clauses in proof lines 9 and 13,
+allowing the input clauses to be deleted.  Then we generate $t_3$ as
+the conjunction of $t_1$ and $t_2$ and assert it as a unit clause.
+Based on this, we can delete the unit clauses for terms $t_1$ and
+$t_2$.  Term $t_3$ then becomes the root of the schematic representation.
 

### 2022-04-24 doc/dd/dd.py
Added edge bubbles

@@ -27,9 +27,10 @@ import math
 #       Position options: above, below, left, right
 
 # Edge
-#    e FID TID value=neutral [only=XX]
+#    e FID TID value=neutral mark=none [only=XX]
 #       FID and TID are Ids of source and destination nodes
 #       value options: neutral, high, low, path
+#       mark options: none, bubble   
 
 # Graphic primitives
 #
@@ -330,10 +331,14 @@ class Node(Only):
 class EdgeType:
     neutral, high, low, path = list(range(4))
     names = ["neutral", "high", "low", "path"]
+    none, bubble = list(range(2))
+    markNames = ["none", "bubble"]
 
     def parse(self, prefix):
         return getIndex(prefix, self.names)
         
+    def parseMark(self, prefix):
+        return getIndex(prefix, self.markNames)
 
 class Edge(Only):
     fromId = None
@@ -341,6 +346,7 @@ class Edge(Only):
     toId = None
     toNode = None
     etype = None
+    mtype = EdgeType.none
     edgeSpacing = 12.0
     edgeFraction = 0.65
 
@@ -364,7 +370,7 @@ class Edge(Only):
                 raise ParseException(line, "Invalid key=value syntax")
             prefix, value = info
             try:
-                key = parseLabel(prefix, ["type", "only"])
+                key = parseLabel(prefix, ["type", "mark", "only"])
             except NoMatchException:
                 raise ParseException(line, "Invalid prefix '%s'" % prefix)
             except MultiMatchException:
@@ -376,6 +382,13 @@ class Edge(Only):
                     raise ParseException(line, "Invalid type '%s'" % value)
                 except MultiMatchException:
                     raise ParseException(line, "Ambiguous type '%s'" % value)
+            elif key == "mark":
+                try:
+                    self.mtype = EdgeType().parseMark(value)
+                except NoMatchException:
+                    raise ParseException(line, "Invalid mark '%s'" % value)
+                except MultiMatchException:
+                    raise ParseException(line, "Ambiguous mark '%s'" % value)
             elif key == "only":
                 self.setOnly(value)
         return self
@@ -401,6 +414,11 @@ class Edge(Only):
[…]

### 2022-04-24 doc/dd/c3-eval.tex
Added and revised examples

@@
 \definecolor{lowcolor}{RGB}{0,0,0}
-\draw (1.32,6.80) [thin,neutralcolor] -- (0.26,5.74);
-\draw (1.32,6.80) [thin,neutralcolor] -- (1.68,5.03);
-\draw (1.68,5.03) [thin,neutralcolor] -- (2.74,3.97);
-\draw (1.68,5.03) [thin,neutralcolor] -- (1.32,3.27);
-\draw (1.32,3.27) [thin,neutralcolor] -- (0.26,2.21);
-\draw (1.32,3.27) [thin,neutralcolor] -- (1.68,1.50);
-\draw (1.68,1.50) [thin,neutralcolor] -- (2.74,0.44);
-\draw (1.68,1.50) [thin,neutralcolor] -- (0.62,0.44);
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.68,5.03) circle [radius=0.26];
-\node at (1.68,5.03) {{\texttt{*}}};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.68,1.50) circle [radius=0.26];
-\node at (1.68,1.50) {{\texttt{*}}};
-\draw [thin,fill=fillcolor,draw=fillcolor] (2.74,0.44) circle [radius=0.44];
-\node at (2.74,0.44) {$1-w_2$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.32,6.80) circle [radius=0.26];
-\node at (1.32,6.80) {{\texttt{+}}};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.32,3.27) circle [radius=0.26];
-\node at (1.32,3.27) {{\texttt{+}}};
-\draw [thin,fill=fillcolor,draw=fillcolor] (2.74,3.97) circle [radius=0.44];
-\node at (2.74,3.97) {$1-w_1$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (0.26,2.21) circle [radius=0.26];
-\node at (0.26,2.21) {$w_2$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (0.62,0.44) circle [radius=0.26];
-\node at (0.62,0.44) {$w_3$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (0.26,5.74) circle [radius=0.26];
-\node at (0.26,5.74) {$w_1$};
-\node [right] at (1.59,6.80) {$s_7$};
-\node [left] at (1.41,5.03) {$p_6$};
-\node [right] at (1.59,3.27) {$s_5$};
-\node [left] at (1.41,1.50) {$p_4$};
+\draw (2.12,8.74) [thin,neutralcolor] -- (2.12,7.41);
+\draw (2.12,7.41) [thin,neutralcolor] -- (0.35,4.77);
+\draw (2.12,7.41) [thin,neutralcolor] -- (2.12,5.65);
+\draw (2.12,5.65) [thin,neutralcolor] -- (0.35,4.77);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.24,5.21) circle [radius=0.09];
+\draw (2.12,5.65) [thin,neutralcolor] -- (2.12,3.88);
+\draw (2.12,3.88) [thin,neutralcolor] -- (0.35,1.24);
+\draw (2.12,3.88) [thin,neutralcolor] -- (2.12,2.12);
+\draw (2.12,2.12) [thin,neutralcolor] -- (0.35,1.24);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.24,1.68) circle [radius=0.09];
+\draw (2.12,2.12) [thin,neutralcolor] -- (2.12,0.35);
+\draw (0.35,4.77) [thin,neutralcolor] -- (0.35,3.88);
+\draw (0.35,1.24) [thin,neutralcolor] -- (0.35,0.35);
+\draw [thin,fill=neutralcolor,draw=neutralcolor] (0.35,4.77) circle [radius=0.00];
+\node at (0.35,4.77) {};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.24,5.21) circle [radius=0.26];
+\node at (1.24,5.21) {$\sim$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.24,1.68) circle [radius=0.26];
+\node at (1.24,1.68) {$\sim$};
+\draw [thin,fill=neutralcolor,draw=neutralcolor] (0.35,1.24) circle [radius=0.00];
+\node at (0.35,1.24) {};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (2.12,5.65) circle [radius=0.44];
+\node at (2.12,5.65) {$\times$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (2.12,2.12) circle [radius=0.44];
+\node at (2.12,2.12) {$\times$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (2.12,7.41) circle [radius=0.44];
+\node at (2.12,7.41) {$+$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (2.12,3.88) circle [radius=0.44];
+\node at (2.12,3.88) {$+$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (2.12,8.74) circle [radius=0.35];
+\node at (2.12,8.74) {$\sigma(\phi_1)$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (0.35,0.35) circle [radius=0.35];
+\node at (0.35,0.35) {$w_2$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (2.12,0.35) circle [radius=0.35];
+\node at (2.12,0.35) {$w_3$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (0.35,3.88) circle [radius=0.35];
+\node at (0.35,3.88) {$w_1$};
+\node [right] at (2.56,7.41) {$s_7$};
+\node [right] at (2.56,5.65) {$p_6$};
+\node [right] at (2.56,3.88) {$s_5$};
+\node [right] at (2.56,2.12) {$p_4$};
 \end{tikzpicture}%%

### 2022-04-24 doc/dd/c3.tex
New example
Added and revised examples

@@
 \definecolor{lowcolor}{RGB}{0,0,0}
-\draw (1.32,6.62) [thin,neutralcolor] -- (0.26,5.56);
-\draw (1.32,6.62) [thin,neutralcolor] -- (1.68,4.85);
-\draw (1.68,4.85) [thin,neutralcolor] -- (2.74,3.79);
-\draw (1.68,4.85) [thin,neutralcolor] -- (1.32,3.09);
-\draw (1.32,3.09) [thin,neutralcolor] -- (0.26,2.03);
-\draw (1.32,3.09) [thin,neutralcolor] -- (1.68,1.32);
-\draw (1.68,1.32) [thin,neutralcolor] -- (2.74,0.26);
-\draw (1.68,1.32) [thin,neutralcolor] -- (0.62,0.26);
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.68,4.85) circle [radius=0.53];
-\node at (1.68,4.85) {$\mathbin{\land_{\textrm{pv}}}$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.68,1.32) circle [radius=0.53];
-\node at (1.68,1.32) {$\mathbin{\land_{\textrm{pv}}}$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (2.74,0.26) circle [radius=0.26];
-\node at (2.74,0.26) {$\overline{x}_2$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.79,6.09) rectangle (1.85,7.15);
-\node at (1.32,6.62) {$\mathbin{\lor_{\textrm{pa}}}$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.79,2.56) rectangle (1.85,3.62);
-\node at (1.32,3.09) {$\mathbin{\lor_{\textrm{pa}}}$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (2.74,3.79) circle [radius=0.26];
-\node at (2.74,3.79) {$\overline{x}_1$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (0.26,2.03) circle [radius=0.26];
-\node at (0.26,2.03) {$x_2$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (0.62,0.26) circle [radius=0.26];
-\node at (0.62,0.26) {$x_3$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (0.26,5.56) circle [radius=0.26];
-\node at (0.26,5.56) {$x_1$};
-\node [right] at (1.85,6.62) {$s_7$};
-\node [left] at (1.15,4.85) {$p_6$};
-\node [right] at (1.85,3.09) {$s_5$};
-\node [left] at (1.15,1.32) {$p_4$};
+\draw (2.12,8.74) [thin,neutralcolor] -- (2.12,7.41);
+\draw (2.12,7.41) [thin,neutralcolor] -- (0.35,4.77);
+\draw (2.12,7.41) [thin,neutralcolor] -- (2.12,5.65);
+\draw (2.12,5.65) [thin,neutralcolor] -- (0.35,4.77);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.24,5.21) circle [radius=0.09];
+\draw (2.12,5.65) [thin,neutralcolor] -- (2.12,3.88);
+\draw (2.12,3.88) [thin,neutralcolor] -- (0.35,1.24);
+\draw (2.12,3.88) [thin,neutralcolor] -- (2.12,2.12);
+\draw (2.12,2.12) [thin,neutralcolor] -- (0.35,1.24);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.24,1.68) circle [radius=0.09];
+\draw (2.12,2.12) [thin,neutralcolor] -- (2.12,0.35);
+\draw (0.35,4.77) [thin,neutralcolor] -- (0.35,3.88);
+\draw (0.35,1.24) [thin,neutralcolor] -- (0.35,0.35);
+\draw [thin,fill=neutralcolor,draw=neutralcolor] (0.35,4.77) circle [radius=0.00];
+\node at (0.35,4.77) {};
+\draw [thin,fill=neutralcolor,draw=neutralcolor] (0.35,1.24) circle [radius=0.00];
+\node at (0.35,1.24) {};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (2.12,5.65) circle [radius=0.44];
+\node at (2.12,5.65) {$\mathbin{\land_{\textrm{v}}}$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (2.12,2.12) circle [radius=0.44];
+\node at (2.12,2.12) {$\mathbin{\land_{\textrm{v}}}$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (2.12,7.41) circle [radius=0.44];
+\node at (2.12,7.41) {$\mathbin{\lor_{\textrm{a}}}$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (2.12,3.88) circle [radius=0.44];
+\node at (2.12,3.88) {$\mathbin{\lor_{\textrm{a}}}$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (2.12,8.74) circle [radius=0.35];
+\node at (2.12,8.74) {$t_1$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (0.35,0.35) circle [radius=0.35];
+\node at (0.35,0.35) {$x_2$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (2.12,0.35) circle [radius=0.35];
+\node at (2.12,0.35) {$x_3$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (0.35,3.88) circle [radius=0.35];
+\node at (0.35,3.88) {$x_1$};
+\node [right] at (2.56,7.41) {$s_7$};
+\node [right] at (2.56,5.65) {$p_6$};
+\node [right] at (2.56,3.88) {$s_5$};
+\node [right] at (2.56,2.12) {$p_4$};
 \end{tikzpicture}%%

### 2022-04-24 doc/mc-notes.tex
Added and revised examples

@@
 
-For assignment $\interp$ and a Boolean formula $E$ over $X$, we
-use the notation $\interp\subs{x_i}{E}$ to denote the assignment
+For assignment $\interp$ and a Boolean formula $\phi$ over $X$, we
+use the notation $\interp\subs{x_i}{\phi}$ to denote the assignment
 $\interp'$, such that $\interp'(x_j) = \interp(x_j)$ for all $j \not = i$
-and $\interp'(x_i) = \interp(E)$, where $\interp(E)$ is the value obtained by evaluating formula $E$ with each variable assigned the value given by $\interp$.
+and $\interp'(x_i) = \interp(\phi)$, where $\interp(\phi)$ is the value obtained by evaluating formula $E$ with each variable assigned the value given by $\interp$.
 In particular, the notation
@@
   }
-\caption{Schema for Formula $x_1 \lor x_2 \lor x_3$ and its Cost Computation}
+\caption{Schema for Formula $\phi_1 = x_1 \lor x_2 \lor x_3$ and its Cost Computation}
 \label{fig:c3:schema}
@@
   \centering{
+    \begin{minipage}{0.4\textwidth}
+      A) Schema   \\[2ex]
+    \centering{\input{dd/p2}}
+    \end{minipage}
+    \begin{minipage}{0.4\textwidth}
+    B) Cost Computation\\[2ex]
+    \centering{\input{dd/p2-eval}}
+    \end{minipage}
+  }
+\caption{Schema for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
+\label{fig:c3:schema}
+\end{figure}
+
+\begin{figure}
+  \centering{
   \begin{tabular}{llllll}
@@
 }  
-  \caption{CRAT Proof Steps for Formula $(x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2$}
+  \caption{CRAT Proof Steps for Formula $\phi = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
   \label{fig:p2:crat}

### 2022-04-25 doc/mc-notes.tex
Ready for discussion

@@
 assignment, and $\zero$ denote the assignment that assigns value 0 to
-every assignment.  These are characterized by the universal and empty
-assignment sets, respectively.
+every assignment.  These are characterized by the universal assignment set $\uinterp$ and the empty
+assignment set $\emptyset$, respectively.
 
 From
-this we can define the {\em complement} of function $f$ as the function
+this we can define the {\em negation} of function $f$ as the function
 $\boolnot f$ such that
-$\interpset{\boolnot f} = \{ \interp | f(\interp) = 0\}$.
+$\interpset{\boolnot f} = \uinterp - \interpset{f}$.
 We can also define the conjunction and disjunction operations over functions $f_1$ and $f_2$ as characterized by the sets
@@
 indicates the assignment in which the value
-assigned to $x_i$ is complemented, while others remain unchanged.
+assigned to $x_i$ is negated, while others remain unchanged.
 
 A Boolean function $f$ is said to be {\em independent} of variable
-$x_i$ if every $\interp \in \interpset{f}$ has $\interp(\subs{x_i}{\obar{x}_i})
-\in \interpset{f}$.  The {\em dependency set} of $f$, denoted
+$x_i$ if every $\interp \in \interpset{f}$ has
+$\interp(\subs{x_i}{\obar{x}_i}) \in \interpset{f}$.
+The {\em dependency set} of $f$, denoted
 $\depend(f)$ consists of all variables $x_i$ for which $f$ is {\em
@@
 the cost of literal $\obar{x}_i$ as $\cost(\obar{x}_i) = 1 - \cost(x_i)$, the cost
-of an assignment as $\cost(\interp) = \prod_{l_i \in \interp}
-\cost(l_i)$, and the cost of a function $f$ as $\cost(f) =
-\sum_{\interp \in \interpset{f}} \cost(\interp)$.
+of an assignment as
+$\cost(\interp) = \prod_{l_i \in \interp} \cost(l_i)$,
+and the cost of a function $f$ as
+$\cost(f) = \sum_{\interp \in \interpset{f}} \cost(\interp)$.
 
@@
 assignment is then $1/2^{n}$, and the cost of a function is its
-{\em density}, denoted $\density(f)$.  That is, the density of $f$ satisfies
-$0 \leq \density(f) \leq 1$.  It
-is the fraction of assignments for which the $f$
+{\em density}, denoted $\density(f)$.  That is, the density of $f$, for which
+$0 \leq \density(f) \leq 1$, 
+is the fraction of assignments for which $f$
 evaluates to 1, with $\density(\zero) = 0$ and $\density(\one) = 1$.  The density of a function
@@
 {\bf Example 2}: Let $\ring$ be the set of rational numbers.  Assign a
-{\em weight} $w_i$ for each variable $x_i$ such that $0 \leq w_i \leq
-1$ and let $\cost(x_i) = w_i$.  This can support a restricted version of
-{\em weighted} model counting.  The key restrictions are: 1) the
+{\em weight} $w_i$ to each variable $x_i$ such that $0 \leq w_i \leq
+1$ and let $\cost(x_i) = w_i$.  This implements {\em weighted model counting}
+under the restrictions that: 
+1) the
 weight of an assignment equals the product of the weights of its
@@
 
-\begin{lemma}[Complementation]
-\label{lemma:complementation}
+\begin{proposition}[Complementation]
+\label{prop:complementation}
   For separable cost function $\cost$ and Boolean function $f$:
-  $\cost(\neg f) = \oneminus \cost(f)$.
-\end{lemma}
+  $\cost(\neg f) = 1 - \cost(f) = \oneminus \cost(f)$.
+\end{proposition}
 
-\begin{lemma}[Variable-Partitioned Conjunction]
-\label{lemma:conjunction}
+\begin{proposition}[Variable-Partitioned Conjunction]
+\label{prop:conjunction}
   For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\depend(f_1) \cap \depend(f_2) = \emptyset$:
     $\cost(f_1 \land f_2) = \cost(f_1) \cdot \cost(f_2)$.
-\end{lemma}
+\end{proposition}
 We use the notation $f_1 \pand f_2$ to denote the conjunction of $f_1$ and
@@
 
-\begin{lemma}[Assignment-Partitioned Disjunction]
-\label{lemma:disjunction}
+\begin{proposition}[Assignment-Partitioned Disjunction]
+\label{prop:disjunction}
   For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\interpset{f_1} \cap \interpset{f_2} = \emptyset$:
     $\cost(f_1 \lor f_2) = \cost(f_1) + \cost(f_2)$.
-\end{lemma}
+\end{proposition}
 We use the notation $f_1 \por f_2$ to denote the disjunction of $f_1$ and $f_2$ under the
@@
 
-We will also find it useful to introduce a restricted version of the
-if-then-else operator, denoted $\pite$.  That is, for functions $f_i$,
-$f_t$, and $f_e$, we define $\pite(f_i, f_t, f_e) = (f_i \pand f_t)
-\por (\neg f_i \pand f_e)$.  The two arguments to the disjunction are
-guaranteed to satisfy the constraints for an assignment-partitioned
-disjunction, since first argument can only be satisfied by assignments
-for which $f_i$ evaluates to $\tautology$, while the second can only
-be satisfied by assignments for which $f_i$ evaluates to $\nil$.  The
+We also find it useful to introduce a restricted version of the
+if-then-else operator, denoted $\pite$.  For functions $f_1$,
+$f_2$, and $f_3$, we define $\pite(f_1, f_2, f_3) = (f_1 \pand f_2)
+\por (\neg f_1 \pand f_3)$.  The two arguments to the disjunction are
+guaranteed to satisfy the requirements for an assignment-partitioned
+disjunction, since the first argument can only be satisfied by assignments
+for which $f_1$ evaluates to $\tautology$, while the second can only
+be satisfied by assignments for which $f_1$ evaluates to $\nil$.  The
 only restriction of the $\pite$ operator is that
-$\depend(f_i) \cap \depend(f_t) = \emptyset$ and
-$\depend(f_i) \cap \depend(f_e) = \emptyset$.
+$\depend(f_1) \cap \depend(f_2) = \emptyset$ and
+$\depend(f_1) \cap \depend(f_3) = \emptyset$.
 
@@
   \label{tab:schema}
-A) Primitive Operations\\
+A) Basic Operations\\
 \begin{center}
@@
 \end{center}
-B) Derived Operations\\
+B) Composite Operations\\
 \begin{center}
@@
 $\por$ and $\pite$.  We use a graph representation to allow sharing of
-subformulas.  The set of schemas over a set of variables $\{x_1, x_2,
-\ldots, x_n\}$ can be defined recursively, as is shown in Table
+subformulas.  The set of schemas over a set of variables
+$\{x_1, x_2, \ldots, x_n\}$
+can be defined recursively, as is shown in Table
 \ref{tab:schema}.  Each schema $S$ has an associated dependency set
@@
 divide the operations into {\em basic} ones that are fundamental to
-the representation, and {\em derived} ones that can be constructed
+the representation, and {\em composite} ones that can be constructed
 from basic operations.  The latter category consists of just the $\pite$ operation.
@@
 straightforward---eliminating extra negations and removing constant
-terms.  The most interesting are the two bottom rules, were we make
-use of DeMorgan's Laws to simplify an $\pite$ operation when one of
-the arguments is $\tautology$.  We will make use of this
+terms.  The most interesting are the two bottom rules, which
+use DeMorgan's Laws to simplify an $\pite$ operation when one of
+the arguments is $\tautology$ into a $\pand$ operation.  We will make use of this
 transformation when using BDDs to convert a set of clauses into a
@@
   blocked, or 2) it satisfies the RAT property with respect to a supplied sequence
-  of earlier antecedent clauses.
+  of earlier antecedent clauses.  However, blocked clauses can only be added when defining a schema operation.
 \item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and $\pite$.
@@
   to the input formula, not just that they are equisatisfiable.
-  Therefore, each deletion step must also be show to be equivalence
+  Therefore, each deletion step must also be shown to be equivalence
   preserving, either because the clause is blocked or it follows from
@@
 variable $v$.  Each clause is indicated by a positive integer
-identifier $C$, with the first ones being input clauses and successive
-ones being added clauses.  Clause identifiers must be totally ordered,
+identifier $C$, with the first ones being the IDs of the input clauses and successive
+ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
 such that clause $C$ can only reference clauses $C'$ such that $C' <
-C$.  However, clause identifiers need not be consecutive.
+C$.  Clause identifiers need not be consecutive.
 
@@
 proofs.  Our syntax optionally allows input clauses to be listed with
-a rule of type {\tt i}.  Clauses can be added via blocked-clause and
-RAT rules.  As described below, however, blocked clauses can only be added
+a rule of type {\tt i}.  Clauses can be added via a blocked-clause addition (command {\tt ab})
+or a RAT addition (command {\tt dr}).
+As described below, however, blocked clauses can only be added
 to define $\pand$, $\por$, and $\pite$ operations.  The hints portion
@@
 IDs such that the added clause is RAT with respect to these clauses.
-Clause deletion requires an explicit justification that the deleted
-clauses is RAT with respect to other clauses.
+Similarly, a clause can be deleted if it blocked with respect to one
+of its literals (command {\tt db}), or because it is RAT (command {\tt
+  dr}) The hints portion of a RAT deletion operation must be an
+ordered list of clause IDs justifying the deletion.
 
@@
 A CRAT proof follows the same general form as a QRAT dual proof---one
-that ensures that each clause addition and each clause deletion preserves
-equivalence.  Starting with the set of input clauses, it produces a
-sequence of steps that both add and delete clauses.  Each addition
-must be truth preserving and each deletion must be falsehood
+that ensures that each clause addition and each clause deletion
+preserves equivalence.  Starting with the set of input clauses, it
+produces a sequence of steps that both add and delete clauses.  Each
+addition must be truth preserving and each deletion must be falsehood
 preserving.  At the end, all input clauses must have been deleted, and
-among the remaining clauses there must be only a single unit clause consisting of some variable or
-its complement.  Except for trivial cases, the final literal will be
-an extension variable or its complement.  That variable will indicate the root node of the schema.
-Working from that root backward, the schema can be extracted from the CRAT file.
+among the remaining clauses there must be only a single unit clause
+consisting of some variable or its complement.  Except for trivial
+cases, the final literal will be an extension variable or its
+complement.  That literal will indicate the root of the schema as the
+(possibly negated) outpuf of a schema operation.  Working from that
+root backward, the schema can be extracted from the CRAT file.
 
@@
 
-As an illustration, consider the Boolean formula $x_1 \lor x_2 \lor
-x_3$, represented by a single clause.  We cannot directly use the
+As an illustration, consider the Boolean formula
+$\phi_1 = x_1 \lor x_2 \lor x_3$,
+represented by a single clause.  We cannot directly use the
 $\por$ operation to form these disjunctions, since the sets of assignments
 satisfying the individual literals are not disjoint.  Instead, we must
-decompose this formula into a sequence of operations.  Figure~\ref{fig:c3:schema}A shows one such decomposition.
-Circles in this schema
-correspond to $\pand$ operations and squares to $\por$ operations.
+decompose this formula into a sequence of operations.
+Figure~\ref{fig:c3:schema}A shows one such decomposition.
 The subscripts of the variables and the operator labels correspond to
 the numbers of the input and extension variables in the CRAT proof.
+Edges marked with dots indicate Boolean negation.
 
@@
 
-The logical representation can readily be converted into a schema for
-computing the cost of the formula, given weight $w_i$ for each
-variable $x_i$ for $1 \leq i \leq 3$.  This is illustrated in
-Figure~\ref{fig:c3:schema}B\@.  This schema is valid for any cost
-function, since it contains no abstraction operations.
+The logical representation can readily be converted into a formula for
+computing $\cost(\phi_1)$, the cost of formula $\phi_1$, given a weight
+$w_i$ for each variable $x_i$ for $1 \leq i \leq 3$.  This is
+illustrated in Figure~\ref{fig:c3:schema}B\@.  Note how the Boolean
+negations become $\oneminus$ operations.  This formula is valid for
+any cost function.
 
 Figure~\ref{fig:c3:crat} shows an annotated version of the CRAT proof
-for this example.  Clause \#1 corresponds to the input formula, and
+for this example.  Clause \#1
+is the input clause, and
 clauses \#2--\#13 are the defining clauses for the four operations.
@@
 clauses as a proof that its arguments are mutually exclusive.
-Clause \#14 adds the unit clause corresponding to sum $s_7$,
-indicating that the extension variable variable will evaluate to
-$\tautology$  for any assigment that satisfies the input clause.  We
-can write this as $C \turnstile s_7$ for input clause $C$.  The
-deletion step at the end turns this around, showing that $s_7
-\turnstile C_I$, and therefore the input clause can be deleted, This
-completes a proof that extension variable $s_7$ is logically
+Proof clause \#14 adds the unit clause for the extension variable $s_7$.
+We refer to the literal representing formula $\phi_1$ as $t_1$,
+and we therefore have $t_1 = s_7$.
+The unit clause indicates that extension variable $s_7$ will evaluate to
+$\tautology$  for any assigment that satisfies the formula.  We
+can write this as $\phi_1 \turnstile t_1$.  The
+deletion step at the end turns this around, showing that
+$t_1 \turnstile \phi_1$,
+and therefore the input clause can be deleted, This
+completes a proof that $t_1$ is logically
 equivalent to the input formula.
@@
 \caption{Schema for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
-\label{fig:c3:schema}
+\label{fig:p2:schema}
 \end{figure}
@@
 \obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
-of DeMorgan's Laws to provide a more direct encoding of the problem.
+of DeMorgan's Laws to provide a more compact encoding of the formula.
 The complement of the formula can be written in DNF as
@@
 formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
-Boolean formulas~\cite{bryant:cade:2021}.  The key idea 
-is to use extended resolution to encode the semantics of the BDD
-nodes as part of the proof.  Here we can further decompose each $\ite$
-(short for ``if-then-else'') operation representing a BDD node into
-two $\pand$ and one $\por$ operation.  That is, we can encode
-$\ite(x,A,B)$ as $(x \pand A) \por (\obar{x} \pand B)$, where $x$ is an
-input variable and $A$ and $B$ are subformulas.
-The sum operation trivially satisfies the disjoint assignment requirement, since $x$ has positive polarity in the first product and negative polarity in the second.
-We must also
-make sure that $x$ is not in the dependency set of either $A$ or $B$.
-For ordered BDDs,~\cite{Bryant:1986} this property holds, because the
-variables in $A$ and $B$ will be greater in the variable ordering than
-$x$.
-
-The BDD representations of many of the formulas occuring in
-model-counting problems are far to large for this approach to be
-practical.  One refinement of the approach, implemented by the ADDMC
-model counter, is to abstract subformulas, keeping track only of the
-number of models they can have and representing a number of subformulas
-with a single ADD leaf node.  Our abstraction operation is intended to
-support this capability.  We can
-introduce a new variable to represent the cost function for some
-subformula.  If subsequent formulas are guaranteed to yield the same
-cost, then these can alias to the earlier variable.  However, it is
-not clear how to prove that this aliasing preserves equivalence.
+Boolean formulas~\cite{bryant:cade:2021}.  The key idea is to use
+extended resolution to encode the semantics of the BDD nodes as part
+of the proof.  Each BDD node defines an ITE operation.  As the
+transformations of Table~\ref{tab:normalize} indicate, each of these
+will be represented in the proof by a literal, an $\pand$ operation,
+or an $\pite$ operation.  The ITE operation for a node has an input
+variable $x_i$ as its first argument.  In an ordered BDD, the
+dependency sets of the two children will include only variables
+greater than $x_i$ in the ordering.  Therefore, these operations will
+satisfy the partitioned-variable requirements of the  $\pand$ and $\pite$ operations.
 
@@
   \item Generalities and details of the format
-  \item What should be the final state?
-  \item When can blocked clauses be deleted?
-  \item How can abstraction be incorporated?
+  \item Can some form of abstraction be incorporated?
+    \begin{itemize}
+    \item Want to abstract a subformula to consider only on its cost and dependency set
+    \item Could represent with fresh extension variable
+    \item But how to prove logical equivalence?
+    \end{itemize}
   \end{itemize}
@@
 
-\item CDCL-based
+  \item SDD-based
   \begin{itemize}
-  \item Prototype
-  \item C/C++
-  \end{itemize}
+  \item Bottom-up
+  \item Top-down
   \end{itemize}
 
+\item Others?
+\end{itemize}
 \end{itemize}

### 2022-04-27 doc/mc-notes.tex
Minor updates

@@
 weight of an assignment equals the product of the weights of its
-literals, and 2) the weight of a variable $x_i$ and its complement
+literals, and 2) the weight of a variable $x_i$ and its negation
 $\obar{x}_i$ sum to 1.
@@
 
-\begin{proposition}[Complementation]
-\label{prop:complementation}
+\begin{proposition}[Negation]
+\label{prop:negation}
   For separable cost function $\cost$ and Boolean function $f$:
@@
 
-We also find it useful to introduce a restricted version of the
-if-then-else operator, denoted $\pite$.  For functions $f_1$,
-$f_2$, and $f_3$, we define $\pite(f_1, f_2, f_3) = (f_1 \pand f_2)
-\por (\neg f_1 \pand f_3)$.  The two arguments to the disjunction are
-guaranteed to satisfy the requirements for an assignment-partitioned
-disjunction, since the first argument can only be satisfied by assignments
-for which $f_1$ evaluates to $\tautology$, while the second can only
-be satisfied by assignments for which $f_1$ evaluates to $\nil$.  The
-only restriction of the $\pite$ operator is that
-$\depend(f_1) \cap \depend(f_2) = \emptyset$ and
-$\depend(f_1) \cap \depend(f_3) = \emptyset$.
+%% We also find it useful to introduce a restricted version of the
+%% if-then-else operator, denoted $\pite$.  For functions $f_1$,
+%% $f_2$, and $f_3$, we define $\pite(f_1, f_2, f_3) = (f_1 \pand f_2)
+%% \por (\neg f_1 \pand f_3)$.  The two arguments to the disjunction are
+%% guaranteed to satisfy the requirements for an assignment-partitioned
+%% disjunction, since the first argument can only be satisfied by assignments
+%% for which $f_1$ evaluates to $\tautology$, while the second can only
+%% be satisfied by assignments for which $f_1$ evaluates to $\nil$.  The
+%% only restriction of the $\pite$ operator is that
+%% $\depend(f_1) \cap \depend(f_2) = \emptyset$ and
+%% $\depend(f_1) \cap \depend(f_3) = \emptyset$.
 
@@
   \label{tab:schema}
-A) Basic Operations\\
+%%A) Basic Operations\\
 \begin{center}
@@
 \end{center}
-B) Composite Operations\\
-\begin{center}
-\begin{tabular}{cc}
-\toprule
-  $S$           & Definition \\
-\midrule
-\makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
-\bottomrule
-\end{tabular}
-\end{center}
+%% B) Composite Operations\\
+%% \begin{center}
+%% \begin{tabular}{cc}
+%% \toprule
+%%   $S$           & Definition \\
+%% \midrule
+%% \makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
+%% \bottomrule
+%% \end{tabular}
+%% \end{center}
 \end{table}
@@
 Boolean formula where the only allowed operations are $\neg$, $\pand$,
-$\por$ and $\pite$.  We use a graph representation to allow sharing of
+$\por$.
+%% and $\pite$.
+We use a graph representation to allow sharing of
 subformulas.  The set of schemas over a set of variables
@@
 \ref{tab:schema}.  Each schema $S$ has an associated dependency set
-$\depend(S)$ and an associated set of models $\interpset{S}$.  We
-divide the operations into {\em basic} ones that are fundamental to
-the representation, and {\em composite} ones that can be constructed
-from basic operations.  The latter category consists of just the $\pite$ operation.
+$\depend(S)$ and an associated set of models $\interpset{S}$.
+%% We
+%% divide the operations into {\em basic} ones that are fundamental to
+%% the representation, and {\em composite} ones that can be constructed
+%% from basic operations.  The latter category consists of just the $\pite$ operation.
 
@@
     $S \por \tautology$ & $\rightarrow$ & $\tautology$ && $\tautology \por S$ & $\rightarrow$ & $\tautology$ \\
-    $\pite(\neg S_1, S_2, S_3)$ & $\rightarrow$ & $\pite(S_1, S_3, S_2)$ \\
-    $\pite(\tautology, S_2, S_3)$ & $\rightarrow$ & $S_2$ & &
-    $\pite(\nil, S_2, S_3)$ & $\rightarrow$ & $S_3$ \\
-    $\pite(S_1, \tautology, \nil)$ & $\rightarrow$ & $S_1$ &&
-    $\pite(S_1, \nil, \tautology)$ & $\rightarrow$ & $\neg S_1$ \\
-    $\pite(S_1, S_2, \nil)$ & $\rightarrow$ & $S_1 \pand S_2$ & &
-    $\pite(S_1, \nil, S_3)$ & $\rightarrow$ & $\neg S_1 \pand S_3$ \\
-    $\pite(S_1, S_2, \tautology)$ & $\rightarrow$ & $\neg [S_1 \pand \neg S_2]$ & &
-    $\pite(S_1, \tautology, S_3)$ & $\rightarrow$ & $\neg [\neg S_1 \pand \neg S_3]$ \\
+%%     $\pite(\neg S_1, S_2, S_3)$ & $\rightarrow$ & $\pite(S_1, S_3, S_2)$ \\
+%%     $\pite(\tautology, S_2, S_3)$ & $\rightarrow$ & $S_2$ & &
+%%     $\pite(\nil, S_2, S_3)$ & $\rightarrow$ & $S_3$ \\
+%%     $\pite(S_1, \tautology, \nil)$ & $\rightarrow$ & $S_1$ &&
+%%     $\pite(S_1, \nil, \tautology)$ & $\rightarrow$ & $\neg S_1$ \\
+%%     $\pite(S_1, S_2, \nil)$ & $\rightarrow$ & $S_1 \pand S_2$ & &
+%%     $\pite(S_1, \nil, S_3)$ & $\rightarrow$ & $\neg S_1 \pand S_3$ \\
+%%     $\pite(S_1, S_2, \tautology)$ & $\rightarrow$ & $\neg [S_1 \pand \neg S_2]$ & &
+%%     $\pite(S_1, \tautology, S_3)$ & $\rightarrow$ & $\neg [\neg S_1 \pand \neg S_3]$ \\
   \end{tabular}
@@
 Table~\ref{tab:normalize} shows a list of {\em normalizing}
-transformations to simplify a separable schema.  These are mostly
-straightforward---eliminating extra negations and removing constant
-terms.  The most interesting are the two bottom rules, which
-use DeMorgan's Laws to simplify an $\pite$ operation when one of
-the arguments is $\tautology$ into a $\pand$ operation.  We will make use of this
-transformation when using BDDs to convert a set of clauses into a
-separable schema.
+transformations to simplify a separable schema.  These 
+eliminate extra negations and remove constant
+terms, such that constants only occur in schemas when representing constant functions $\zero$ and $\one$.
+%% The most interesting are the two bottom rules, which
+%% use DeMorgan's Laws to simplify an $\pite$ operation when one of
+%% the arguments is $\tautology$ into a $\pand$ operation.  We will make use of this
+%% transformation when using BDDs to convert a set of clauses into a
+%% separable schema.
 
@@
 checkable proof that a Boolean formula, given in conjunctive normal
-form, is logically equivalent to some separable schema.  This provides
-a framework for adding proof generation to model counting programs.
+form, is logically equivalent to a separable schema.  Once this
+equivalence has been established, the schema can form the basis
+for computations enabled by the
+this representation, including trusted model counting
 
@@
         & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
-        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
+%%        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
   \end{tabular}
@@
 a rule of type {\tt i}.  Clauses can be added via a blocked-clause addition (command {\tt ab})
-or a RAT addition (command {\tt dr}).
+or a RAT addition (command {\tt ar}).
 As described below, however, blocked clauses can only be added
-to define $\pand$, $\por$, and $\pite$ operations.  The hints portion
+to define $\pand$ and $\por$
+%%, and $\pite$
+operations.  The hints portion
 of a blocked-clause addition lists all earlier clauses containing the
@@
 
-An if-then-else rule of the form ${\tt ite}\;v\;l_1\;l_2\;l_3$
-indicates that $v$ will represent $\pite(l_1, l_2, l_3)$.  The blocked
-clause additions must encode the formula
-$v \leftrightarrow \ite((l_1,l_2, l_3)$.
-The dependency set for $l_1$ must be disjoint from those
-of $l_2$ and $l_3$.
+%% An if-then-else rule of the form ${\tt ite}\;v\;l_1\;l_2\;l_3$
+%% indicates that $v$ will represent $\pite(l_1, l_2, l_3)$.  The blocked
+%% clause additions must encode the formula
+%% $v \leftrightarrow \ite((l_1,l_2, l_3)$.
+%% The dependency set for $l_1$ must be disjoint from those
+%% of $l_2$ and $l_3$.
 
@@
 
-A CRAT proof follows the same general form as a QRAT dual proof---one
+A CRAT proof follows the same general form as a QRAT dual proof~\cite{bryant:cade:2021}---one
 that ensures that each clause addition and each clause deletion
@@
 of DeMorgan's Laws to provide a more compact encoding of the formula.
+
 The complement of the formula can be written in DNF as
@@
 form a term $t_1$ that equals $C_1$ and a term $t_2$ that equals
-$C_2$.  These are asserted as unit clauses in proof lines 9 and 13,
+$C_2$.  (Note that $t_1$ in this case is logically equivalent to $t_1$
+in the example of Figure~\ref{fig:c3:schema}, but the use of
+negation enables it to be represented with two operations
+rather than four.)
+Terms $t_1$ and $t_2$ are asserted as unit clauses in proof lines 9 and 13,
 allowing the input clauses to be deleted.  Then we generate $t_3$ as
@@
 Based on this, we can delete the unit clauses for terms $t_1$ and
-$t_2$.  Term $t_3$ then becomes the root of the schematic representation.
+$t_2$.  Term $t_3$ then becomes the root of the schematic representation shown in Figure~\ref{fig:p2:schema}.
 
@@
 extended resolution to encode the semantics of the BDD nodes as part
-of the proof.  Each BDD node defines an ITE operation.  As the
-transformations of Table~\ref{tab:normalize} indicate, each of these
-will be represented in the proof by a literal, an $\pand$ operation,
-or an $\pite$ operation.  The ITE operation for a node has an input
-variable $x_i$ as its first argument.  In an ordered BDD, the
-dependency sets of the two children will include only variables
-greater than $x_i$ in the ordering.  Therefore, these operations will
-satisfy the partitioned-variable requirements of the  $\pand$ and $\pite$ operations.
+of the proof.  Depending on which, if any, of its children are
+terminal nodes, a BDD node will be represented in the schema as either
+1) a literal, 2) a single $\pand$ operation, or 3) two $\pand$
+operations and one $\por$ operation.
 
@@
 generate representations of the formula that are similar to our
-schmas.  These exploit the key abstractions we have identified.  We
+schemas.  These exploit the key abstractions we have identified.  We
 must find ways to modify these model counters to also generate CRAT

### 2022-05-02 doc/dd/p2-bdd-eval.tex
Renamed

\begin{tikzpicture}
\definecolor{neutralcolor}{RGB}{0,0,0}
\definecolor{pathcolor}{RGB}{0,0,0}
\definecolor{fillcolor}{RGB}{255,255,255}
\definecolor{background}{RGB}{225,225,225}
\definecolor{highcolor}{RGB}{0,0,0}
\definecolor{lowcolor}{RGB}{0,0,0}
\draw (1.76,7.41) [thin,neutralcolor] -- (1.76,5.65);
\draw (1.76,5.65) [thin,neutralcolor] -- (0.44,3.88);
\draw (1.76,5.65) [thin,neutralcolor] -- (3.09,3.88);
\draw (0.44,3.88) [thin,neutralcolor] -- (1.76,3.00);
\draw (0.44,3.88) [thin,neutralcolor] -- (0.44,2.12);
\draw (3.09,3.88) [thin,neutralcolor] -- (1.76,3.00);
\draw (3.09,3.88) [thin,neutralcolor] -- (3.09,2.12);
\draw (1.76,3.00) [thin,neutralcolor] -- (1.76,2.12);
\draw (0.44,2.12) [thin,neutralcolor] -- (0.44,0.35);
\draw (0.44,2.12) [thin,neutralcolor] -- (1.76,1.24);
\draw (3.09,2.12) [thin,neutralcolor] -- (1.76,1.24);
\draw (1.76,1.24) [thin,neutralcolor] -- (1.76,0.35);
\draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,3.00) circle [radius=0.00];
\node at (1.76,3.00) {};
\draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,1.24) circle [radius=0.00];
\node at (1.76,1.24) {};
\draw [thin,fill=fillcolor,draw=neutralcolor] (3.09,3.88) circle [radius=0.44];
\node at (3.09,3.88) {$\times$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,3.30) circle [radius=0.26];
\node at (1.32,3.30) {$\sim$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,2.12) circle [radius=0.44];
\node at (0.44,2.12) {$\times$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.88) circle [radius=0.44];
\node at (0.44,3.88) {$\times$};
\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,2.12) circle [radius=0.35];
\node at (1.76,2.12) {$w_1$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (2.21,1.54) circle [radius=0.26];
\node at (2.21,1.54) {$\sim$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.44];
\node at (1.76,5.65) {$+$};
\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,7.41) circle [radius=0.35];
\node at (1.76,7.41) {$\sigma(\phi_2)$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,1.24) circle [radius=0.26];
\node at (0.44,1.24) {$\sim$};
\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,6.53) circle [radius=0.26];
\node at (1.76,6.53) {$\sim$};
\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,0.35) circle [radius=0.35];
\node at (1.76,0.35) {$w_2$};
\draw [thin,fill=fillcolor,draw=fillcolor] (0.44,0.35) circle [radius=0.35];
\node at (0.44,0.35) {$w_3$};
\draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,2.12) circle [radius=0.00];
\node at (3.09,2.12) {};
\draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,1.54) circle [radius=0.26];
\node at (1.32,1.54) {$\sim$};
\end{tikzpicture}%%

### 2022-05-02 doc/dd/p2-bdd.tex
Renamed

--- 2022-05-02 doc/dd/p2-bdd-eval.tex
@@
 \definecolor{lowcolor}{RGB}{0,0,0}
+\draw (0.44,7.41) [thin,neutralcolor] -- (0.44,3.88);
+\draw [fill=neutralcolor,draw=neutralcolor] (0.44,5.65) circle [radius=0.09];
 \draw (1.76,7.41) [thin,neutralcolor] -- (1.76,5.65);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.76,6.53) circle [radius=0.09];
+\draw (3.09,7.41) [thin,neutralcolor] -- (3.09,3.88);
+\draw [fill=neutralcolor,draw=neutralcolor] (3.09,5.65) circle [radius=0.09];
 \draw (1.76,5.65) [thin,neutralcolor] -- (0.44,3.88);
@@
 \draw (0.44,3.88) [thin,neutralcolor] -- (1.76,3.00);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.10,3.44) circle [radius=0.09];
 \draw (0.44,3.88) [thin,neutralcolor] -- (0.44,2.12);
@@
 \draw (0.44,2.12) [thin,neutralcolor] -- (0.44,0.35);
+\draw [fill=neutralcolor,draw=neutralcolor] (0.44,1.24) circle [radius=0.09];
 \draw (0.44,2.12) [thin,neutralcolor] -- (1.76,1.24);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.10,1.68) circle [radius=0.09];
 \draw (3.09,2.12) [thin,neutralcolor] -- (1.76,1.24);
+\draw [fill=neutralcolor,draw=neutralcolor] (2.43,1.68) circle [radius=0.09];
 \draw (1.76,1.24) [thin,neutralcolor] -- (1.76,0.35);
@@
 \draw [thin,fill=fillcolor,draw=neutralcolor] (3.09,3.88) circle [radius=0.44];
-\node at (3.09,3.88) {$\times$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,3.30) circle [radius=0.26];
-\node at (1.32,3.30) {$\sim$};
+\node at (3.09,3.88) {$\land_{\textrm{v}}$};
 \draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,2.12) circle [radius=0.44];
-\node at (0.44,2.12) {$\times$};
+\node at (0.44,2.12) {$\land_{\textrm{v}}$};
 \draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.88) circle [radius=0.44];
-\node at (0.44,3.88) {$\times$};
+\node at (0.44,3.88) {$\land_{\textrm{v}}$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (3.09,7.41) circle [radius=0.35];
+\node at (3.09,7.41) {$t_2$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,2.12) circle [radius=0.35];
-\node at (1.76,2.12) {$w_1$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (2.21,1.54) circle [radius=0.26];
-\node at (2.21,1.54) {$\sim$};
+\node at (1.76,2.12) {$x_1$};
 \draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.44];
-\node at (1.76,5.65) {$+$};
+\node at (1.76,5.65) {$\lor_{\textrm{a}}$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,7.41) circle [radius=0.35];
-\node at (1.76,7.41) {$\sigma(\phi_2)$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,1.24) circle [radius=0.26];
-\node at (0.44,1.24) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,6.53) circle [radius=0.26];
-\node at (1.76,6.53) {$\sim$};
+\node at (1.76,7.41) {$t_3$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (0.44,7.41) circle [radius=0.35];
+\node at (0.44,7.41) {$t_1$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,0.35) circle [radius=0.35];
-\node at (1.76,0.35) {$w_2$};
+\node at (1.76,0.35) {$x_2$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (0.44,0.35) circle [radius=0.35];
-\node at (0.44,0.35) {$w_3$};
+\node at (0.44,0.35) {$x_3$};
 \draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,2.12) circle [radius=0.00];
 \node at (3.09,2.12) {};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,1.54) circle [radius=0.26];
-\node at (1.32,1.54) {$\sim$};
+\node [right] at (2.21,5.65) {$s_7$};
+\node [right] at (3.53,3.88) {$p_6$};
+\node [left] at (0.00,3.88) {$p_5$};
+\node [left] at (0.00,2.12) {$p_4$};
 \end{tikzpicture}%%

### 2022-05-02 doc/dd/p2-cdcl.tex
Incorporate top-down proof example

--- 2022-05-02 doc/dd/p2-bdd.tex
@@
 \definecolor{lowcolor}{RGB}{0,0,0}
-\draw (0.44,7.41) [thin,neutralcolor] -- (0.44,3.88);
-\draw [fill=neutralcolor,draw=neutralcolor] (0.44,5.65) circle [radius=0.09];
 \draw (1.76,7.41) [thin,neutralcolor] -- (1.76,5.65);
-\draw [fill=neutralcolor,draw=neutralcolor] (1.76,6.53) circle [radius=0.09];
-\draw (3.09,7.41) [thin,neutralcolor] -- (3.09,3.88);
-\draw [fill=neutralcolor,draw=neutralcolor] (3.09,5.65) circle [radius=0.09];
 \draw (1.76,5.65) [thin,neutralcolor] -- (0.44,3.88);
@@
 \draw (0.44,3.88) [thin,neutralcolor] -- (0.44,2.12);
+\draw [fill=neutralcolor,draw=neutralcolor] (0.44,3.00) circle [radius=0.09];
 \draw (3.09,3.88) [thin,neutralcolor] -- (1.76,3.00);
@@
 \draw (3.09,2.12) [thin,neutralcolor] -- (1.76,1.24);
-\draw [fill=neutralcolor,draw=neutralcolor] (2.43,1.68) circle [radius=0.09];
 \draw (1.76,1.24) [thin,neutralcolor] -- (1.76,0.35);
@@
 \node at (0.44,3.88) {$\land_{\textrm{v}}$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (3.09,7.41) circle [radius=0.35];
-\node at (3.09,7.41) {$t_2$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,2.12) circle [radius=0.35];
@@
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,7.41) circle [radius=0.35];
-\node at (1.76,7.41) {$t_3$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (0.44,7.41) circle [radius=0.35];
-\node at (0.44,7.41) {$t_1$};
+\node at (1.76,7.41) {$t$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,0.35) circle [radius=0.35];

### 2022-05-02 doc/mc-notes.tex
Change notation
Incorporate top-down proof example

@@
 
-\newcommand{\turnstile}{\vdash}
+\newcommand{\turnstile}{\models}
 
@@
 
+Computing the cost function for a Boolean formula becomes
+straightforward when the formula contains only the operations
+$\pand$, $\por$, and $\neg$, as is demonstrated by
+Propositions~\ref{prop:negation}--\ref{prop:disjunction}.  We define {\em separable schemas}
+as a direct-acyclic graph representation of such formulas.
+Representing formulas as a graph enables sharing subformulas, yielding a more compact representation.
+
+\subsection{Schema Definition}
+
 \begin{table}
@@
 
-A {\em separable schema} is a directed acyclic graph representing a
-Boolean formula where the only allowed operations are $\neg$, $\pand$,
-$\por$.
-%% and $\pite$.
-We use a graph representation to allow sharing of
-subformulas.  The set of schemas over a set of variables
+The set of schemas over a set of variables
 $\{x_1, x_2, \ldots, x_n\}$
@@
 operations.
+
+\subsection{Schema Normalization}
 
@@
 %% separable schema.
+
+\subsection{Encoding the ITE Operation}
+\label{sec:ite}
+
+The if-then-else (ITE) operation arises when converting the CNF
+representation of a formula into a separable schema, both for
+bottom-up approaches based on decision diagrams, and for top-down
+approaches based on CDCL\@.  For functions $f_1$, $f_2$, and $f_3$, we
+define
+$\ite(f_1, f_2, f_3) = (f_1 \land f_2) \lor (\neg f_1 \land f_3)$.
+Observe that the $\lor$ operation in this expression satisfies
+the requirements for $\por$, since the first argument can only yield
+$\tautology$ for assignments that yield $\tautology$ for $f_1$, while
+the second can only yield $\tautology$ for assignments that yield
+$\nil$ for $f_1$.  Therefore, the only condition imposed on an
+expansion of $\ite$ into the allowed schema operations is that the
+dependency set for $f_1$ must be disjoint from those of $f_2$ and
+$f_3$.
+
+\begin{table}
+  \caption{Encodings of the ITE Operation}
+  \label{tab:ite}
+  \begin{center}
+  \begin{tabular}{ccc}
+    ITE Form & & Encoding \\
+    \midrule
+     $\ite(S_1, S_2, S_3)$ & $$ & $(S_1 \pand S_2) \por (\neg S_1 \land S_3)$ \\
+     $\ite(\tautology, S_2, S_3)$ & $$ & $S_2$ \\
+     $\ite(\nil, S_2, S_3)$ & $$ & $S_3$ \\
+     $\ite(S_1, \tautology, \nil)$ & $$ & $S_1$ \\
+     $\ite(S_1, \nil, \tautology)$ & $$ & $\neg S_1$ \\
+     $\ite(S_1, S_2, \nil)$ & $$ & $S_1 \pand S_2$ \\
+     $\ite(S_1, \nil, S_3)$ & $$ & $\neg S_1 \pand S_3$ \\
+     $\ite(S_1, S_2, \tautology)$ & $$ & $\neg (S_1 \pand \neg S_2)$ \\
+     $\ite(S_1, \tautology, S_3)$ & $$ & $\neg (\neg S_1 \pand \neg S_3)$ \\
+  \end{tabular}
+  \end{center}
+\end{table}
+
+Table~\ref{tab:ite} shows different ways to encode an ITE operation
+into schema operations.  All of these require the dependency set of
+the argument $S_1$ to be disjoint from those of arguments $S_2$ and
+$S_3$.  The first row shows the most general case, require one $\por$
+and two $\pand$ operations.  The other rows show special cases, where
+one or more argument is a constant.  Of these, the final two rows are
+particularly noteworthy.  They make use of DeMorgan's Laws to convert
+disjunctions into conjunctions.  These conjunctions can then be
+encoded with $\pand$ operations, since their arguments will have
+disjoint dependency sets.
 
@@
 complement.  That literal will indicate the root of the schema as the
-(possibly negated) outpuf of a schema operation.  Working from that
+(possibly negated) output of a schema operation.  Working from that
 root backward, the schema can be extracted from the CRAT file.
@@
   }
-\caption{Schema for Formula $\phi_1 = x_1 \lor x_2 \lor x_3$ and its Cost Computation}
+\caption{Schema \#1 for Formula $\phi_1 = x_1 \lor x_2 \lor x_3$ and its Cost Computation}
 \label{fig:c3:schema}
@@
   }
-  \caption{CRAT Proof Steps for Formula $x_1 \lor x_2 \lor x_3$}
+  \caption{CRAT Proof \#1 for Formula $x_1 \lor x_2 \lor x_3$}
   \label{fig:c3:crat}
@@
 
-\subsection{Example 2}
+\subsection{Example 2A}
 
@@
       A) Schema   \\[2ex]
-    \centering{\input{dd/p2}}
+    \centering{\input{dd/p2-bdd}}
     \end{minipage}
@@
     B) Cost Computation\\[2ex]
-    \centering{\input{dd/p2-eval}}
+    \centering{\input{dd/p2-bdd-eval}}
     \end{minipage}
   }
-\caption{Schema for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
-\label{fig:p2:schema}
+  \caption{Schema \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
+\label{fig:p2-bdd:schema}
 \end{figure}
@@
     {\tt 14} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
-    {\tt 15} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -5 0} &  \\  
-    {\tt 16} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -5 0} & &  \\
+    {\tt 15} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -14 0} &  \\  
+    {\tt 16} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -14 0} & &  \\
     {\tt 17} & {\tt ar}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
@@
 }  
-  \caption{CRAT Proof Steps for Formula $\phi = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
-  \label{fig:p2:crat}
+  \caption{CRAT Proof \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
+  \label{fig:p2-bdd:crat}
 \end{figure}
 
-As a more complex example, consider the Boolean formula given by the
-conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
+As a more complex example, consider the Boolean formula $\phi_2$ given
+by the conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
 \obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
-of DeMorgan's Laws to provide a more compact encoding of the formula.
+of DeMorgan's Laws to provide a more compact encoding of the formula,
+similar to the use of these laws when encoding ITE operations in
+Table~\ref{tab:ite}.
 
-The complement of the formula can be written in DNF as
-$\obar{x}_1\,\obar{x}_2\,\obar{x}_3 \lor x_1\,\obar{x}_2$.  Each of
-the two conjuncts can be formed using $\pand$ operations, and their
-sum can be formed using a $\por$ operation.  In this example, we first
-form a term $t_1$ that equals $C_1$ and a term $t_2$ that equals
-$C_2$.  (Note that $t_1$ in this case is logically equivalent to $t_1$
-in the example of Figure~\ref{fig:c3:schema}, but the use of
-negation enables it to be represented with two operations
-rather than four.)
-Terms $t_1$ and $t_2$ are asserted as unit clauses in proof lines 9 and 13,
-allowing the input clauses to be deleted.  Then we generate $t_3$ as
-the conjunction of $t_1$ and $t_2$ and assert it as a unit clause.
-Based on this, we can delete the unit clauses for terms $t_1$ and
-$t_2$.  Term $t_3$ then becomes the root of the schematic representation shown in Figure~\ref{fig:p2:schema}.
+Figure~\ref{fig:p2-bdd:schema} shows a schema representing the formula,
+and Figure~\ref{fig:p2-bdd:crat} shows the associated CRAT proof.  This
+proof was generated via a bottom-up strategy, such as would be created
+using BDDs.  It creates schematic representations of the input
+clauses and then forms their conjunction.  In the proof, unit
+clauses are generated for the representations of the input clauses,
+and then the input clauses are deleted.  These intermediate unit
+clauses are used to justify a unit clause for the final root, and then
+they are deleted.
+
+Using DeMorgan's Laws, $C_1$ can be written as 
+$t_1 = \neg [\obar{x}_1\land \obar{x}_2\land \obar{x}_3]$, and this can be
+expressed using $\pand$ operations $s_4$ and $s_5$, shown in
+Figure~\ref{fig:p2-bdd:schema}A\@.  (Note that $t_1$ in this case is
+logically equivalent to root $t_1$ in the schema of
+Figure~\ref{fig:c3:schema}A, but the use of negation enables it to be
+represented with two operations rather than four.)  Similarly, $C_2$
+can be written as $t_2 = \neg [x_1\land \obar{x}_2]$, and this can be
+represented by $\pand$ operation $p_6$.
+Terms $t_1$ and $t_2$ are asserted as unit clauses on proof lines 9 and 13,
+allowing the input clauses to be deleted.
+The conjunction $t_1 \land t_2$
+can be written as $t_3 = \neg[\obar{t_1} \lor \obar{t_2}]$, represented by $\por$ operation $s_7$.
+Term $t_3$ is asserted as a unit clause on proof line 17.
+Based on this, the unit clauses for terms $t_1$ and
+$t_2$ can be deleted.  Term $t_3$ then becomes the unique root of the cost
+computation shown in Figure~\ref{fig:p2-bdd:schema}B\@.
+
+\subsection{Example 2B}
+
+\begin{figure}
+  \centering{
+    \begin{minipage}{0.4\textwidth}
+      A) Schema   \\[2ex]
+    \centering{\input{dd/p2-cdcl}}
+    \end{minipage}
+    \begin{minipage}{0.4\textwidth}
+    B) Cost Computation\\[2ex]
+    \centering{\input{dd/p2-cdcl-eval}}
+    \end{minipage}
+  }
+\caption{Schema \#2B for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
+\label{fig:p2-cdcl:schema}
+\end{figure}
+
+\begin{figure}
+  \centering{
+  \begin{tabular}{llllll}
+    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
+\midrule
+    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause $C_1$}\\
+
+    {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
+            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
+    {\tt 3} & {\tt ab}  & {\tt 4 2 3 0} & {\tt 0}    & & Defining clauses for $p_4$ \\
+    {\tt 4} & {\tt ab}  & {\tt -4 -2 0} & {\tt -2 0}    & & \\
+    {\tt 5} & {\tt ab}  & {\tt -4 -3 0} & {\tt -2 0}    & & \\
+            & {\tt p}   & {\tt 5 -1 -4}  &             & & Declare $p_5 = \obar{x}_1 \pand \obar{p_4}$ \\
+    {\tt 6} & {\tt ab}  & {\tt 5 1 4 0} & {\tt 0}    & & Defining clauses for $p_5$ \\
+    {\tt 7} & {\tt ab}  & {\tt -5 -1 0} & {\tt -6 0}    & & \\
+    {\tt 8} & {\tt ab}  & {\tt -5 -4 0} & {\tt -6 0}    & & \\
+            & {\tt p}   & {\tt 6 1 2} &              & & Declare $p_6 = x_1 \pand x_2$ \\    
+    {\tt 9} & {\tt ab}  & {\tt 6 -1 -2 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
+    {\tt 10} & {\tt ab}  & {\tt -6 1 0}    & {\tt -9 0} & & \\
+    {\tt 11} & {\tt ab}  & {\tt -6 2 0}    & {\tt -9 0} & & \\ 
+            & {\tt s}   & {\tt 7 5 6}   & {\tt 7 10 0}  & & Declare $s_7 = p_5 \por p_6$ \\
+    {\tt 12} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
+    {\tt 13} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -12 0} & & \\  
+    {\tt 14} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -12 0} & & \\
+
+    {\tt 15} & {\tt ar}  & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p_4}$ \\
+    {\tt 16} & {\tt ar}  & {\tt 1 7 0}     & {\tt 13 6 15 0} & & Justify $\obar{x}_1 \rightarrow s_7$ \\
+    {\tt 17} & {\tt ar}  & {\tt -1 7 0}    & {\tt 14 9 2 0} & & Justify $x_1 \rightarrow s_7$ \\
+    {\tt 18} & {\tt ar}  & {\tt 7 0}       & {\tt 16 17 0}  & & Justify unit clause $t = s_7$ \\
+             & {\tt dr}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
+             & {\tt dr}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\
+  \end{tabular}
+}  
+  \caption{CRAT Proof \#2B for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
+  \label{fig:p2-cdcl:crat}
+\end{figure}
+
+Figure~\ref{fig:p2-cdcl:schema} shows an alternate schema representing
+the same formula $\phi_2$ as in Example~2A, and
+Figure~\ref{fig:p2-cdcl:crat} shows the associated CRAT proof.  This
+proof was generated via a top-down strategy, such as would be created
+using a model counter based on CDCL\@.  It starts by splitting on
+variable $x_1$.  Clause $C_1$ is trivially satisfied when $x_1$ is
+assigned $\tautology$, and clause $C_2$ becomes the clause $C'_2 =
+x_2$.  On the other hand, clause $C_2$ is trivially satisfied when
+$x_1$ is assigned $\nil$, and clause $C_1$ becomes $C'_1 = x_2 \lor
+x_3$.  Clause $C'_1$ can be represented schematically as
+$\neg[\obar{x}_2 \pand \obar{x}_3]$, with this operation labeled $p_4$
+in Figure~\ref{fig:p2-cdcl:schema}A\@.  The splitting on variable
+$x_1$ can be rejoined as $\ite(x_1, x_2, \obar{p_4})$, and this ITE
+operation can be expressed using the product operations $p_5$ and
+$p_6$, joined by the sum operation $t = s_7$.
+
+The two schemas shown in Figures~\ref{fig:p2-bdd:schema}A and
+\ref{fig:p2-cdcl:schema}A have similar structure.  They have very
+different negation patterns, but they are logically equivalent.  Their
+associated cost computations (Figures~\ref{fig:p2-bdd:schema}B and
+\ref{fig:p2-cdcl:schema}B) also yield the same results for arbitrary
+weights $w_1$, $w_2$, and $w_3$.
+
+The CRAT proof for the top-down approach follows a different pattern than does the proof based on a bottom-up construction.
+It does
+not create any intermediate unit clauses.  Instead, it constructs a proof
+that the root $t = s_7$ holds as a unit clause by splitting into the
+two assignments for variable $x_1$.  The proof that
+$\obar{x}_1 \rightarrow s_7$ (proof line 16) builds on the proof that
+$\obar{x}_1 \rightarrow \obar{p}_4$ (proof line 15), which derives from clause $C_1$.
+The proof that $x_1 \rightarrow x_7$ (proof line 17) derives from clause $C_2$.
+These two are combined to yield the unit clause $s_7$ (proof line 18).
+Given this unit clause, the two input clauses can then be deleted.
 
@@
 formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
-Boolean formulas~\cite{bryant:cade:2021}.  The key idea is to use
-extended resolution to encode the semantics of the BDD nodes as part
-of the proof.  Depending on which, if any, of its children are
-terminal nodes, a BDD node will be represented in the schema as either
-1) a literal, 2) a single $\pand$ operation, or 3) two $\pand$
-operations and one $\por$ operation.
+Boolean formulas~\cite{bryant:cade:2021}.
+Each BDD node can be expressed as an ITE operation and
+make use of the encodings shown in Table~\ref{tab:ite}.
 
-Other model counters use either top-down and bottom-up methods to
-generate representations of the formula that are similar to our
-schemas.  These exploit the key abstractions we have identified.  We
-must find ways to modify these model counters to also generate CRAT
-proofs.
+A second class of model counters proceeds top-down, based on the CDCL
+framework.  These choose a splitting variable $x$ and recursively
+construct schemas $S_1$ and $S_0$ for the two assignments to the
+variable.  These are combined as $\ite(x, S_1, S_0)$, using one of the
+encodings of $\ite$ shown in Table~\ref{tab:ite}.  Like CDCL, the
+top-down algorithm can make use of unit propagation, conflict
+detection, and clause learning.  It can also make use of {\em
+  variable} partitioning.  That is, suppose for some partial
+assignment to the variables, the input clauses decompose into two or
+more sets over disjoint variables.  Then the schema for each of these
+partitions can be generated separately, and these are joined via the $\pand$
+operation.
 

### 2022-05-03 doc/mc-notes.tex
Improved explanations

@@
      $\ite(S_1, \nil, S_3)$ & $$ & $\neg S_1 \pand S_3$ \\
+     $\ite(S_1, \tautology, S_3)$ & $$ & $\neg (\neg S_1 \pand \neg S_3)$ \\
      $\ite(S_1, S_2, \tautology)$ & $$ & $\neg (S_1 \pand \neg S_2)$ \\
-     $\ite(S_1, \tautology, S_3)$ & $$ & $\neg (\neg S_1 \pand \neg S_3)$ \\
   \end{tabular}
@@
 the argument $S_1$ to be disjoint from those of arguments $S_2$ and
-$S_3$.  The first row shows the most general case, require one $\por$
-and two $\pand$ operations.  The other rows show special cases, where
-one or more argument is a constant.  Of these, the final two rows are
-particularly noteworthy.  They make use of DeMorgan's Laws to convert
-disjunctions into conjunctions.  These conjunctions can then be
+$S_3$.  The first row shows the most general case, requiring one
+$\por$ and two $\pand$ operations.  The other rows show special cases,
+where one or more argument is a constant.  Of these, the final two
+rows are particularly noteworthy.  They make use of DeMorgan's Laws to
+convert disjunctions into conjunctions.  In particular, for Boolean
+functions $f_1$ and $f_2$, we can write $\ite(f_1, \tautology, f_2)$
+as $f_1 \lor f_2$, and by DeMorgan's Laws as $\neg(\neg f_1 \land
+\neg f_2)$.  Similarly $\ite(f_1, f_2, \tautology) = \neg f_1 \lor f_2 = \neg(f_1 \land \neg f_2)$.
+These conjunctions can then be
 encoded with $\pand$ operations, since their arguments will have

### 2022-05-05 doc/mc-notes.tex
Fixed mistake in Blum derivation

@@
 
-{\bf Example 3}: Let $\ring$ be a field with $|\ring| \geq 2n$,
+{\bf Example 3}: Let $\ring$ be a finite field with $z = |\ring| \geq 2n$,
 and let $\hashset$ be the set of functions
@@
 two distinct functions $f_1$ and $f_2$ and a randomly chosen $\fhash
-\in \hashset$, the probability that $h(f_1) = h(f_2)$ will be at most
-$2^n/|\ring|< 1/2$.  Therefore, these functions can be used as part of a
+\in \hashset$, the probability that $h(f_1) \not = h(f_2)$ will be at least
+$(1-1/z)^n \geq (1-1/2n)^n >  1/2$.  Therefore, these functions can be used as part of a
 randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

### 2022-05-11 doc/mc-notes.tex
Fixed mistake in Blum derivation
Updated to reflect new CRAT format

@@
 \in \hashset$, the probability that $h(f_1) \not = h(f_2)$ will be at least
-$(1-1/z)^n \geq (1-1/2n)^n >  1/2$.  Therefore, these functions can be used as part of a
+$\left(1-\frac{1}{z}\right)^n \geq \left(1-\frac{1}{2n}\right)^n >  1/2$.  Therefore, these functions can be used as part of a
 randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
@@
 
-The CRAT clausal proof framework provides a means for creating a
+The CRAT clausal proof framework provides a means to express a
 checkable proof that a Boolean formula, given in conjunctive normal
@@
 for computations enabled by the
-this representation, including trusted model counting
+representation, including trusted model counting
 
@@
 \begin{itemize}
-\item As with LRAT, a clause can be added as long as either 1) it is
-  blocked, or 2) it satisfies the RAT property with respect to a supplied sequence
-  of earlier antecedent clauses.  However, blocked clauses can only be added when defining a schema operation.
-\item  Extension variables can be introduced only according to the operations $\pand$, $\por$, and $\pite$.
+  \item
+  In addition to explicit clause additions and deletions, the proof contains declarations of $\pand$ and $\por$ operations.
 \begin{itemize}
+\item 
+  These declarations implicitly add extension variables and their defining clauses to the proof.
+  This is the only means for generating extension variables or adding blocked clauses to the proof.
 \item The checker tracks the dependency set for every input and
   extension variable.  When an extension variable is introduced based
-  on the $\pand$ or $\pite$ operation, the dependency sets of its arguments must
-  be disjoint.
-\item When an extension variable is introduced based on the $\por$
-  operation, the step must cite earlier steps providing a RAT proof
-  that the two arguments are mutually exclusive.
+  on the $\pand$ operation, the dependency sets of its arguments must
+  be disjoint.  The resulting extension variable has a dependency set equal to the union of those of its arguments.
+\item Declaring a $\por$ operation requires a sequence of clauses
+  providing a RUP proof that the arguments are mutually exclusive.  This sequence can either be provided explicitly or inferred by the proof checker.
 \item Boolean complement is provided implicitly by allowing the
-  arguments of the extension operations to be literals and not just
+  arguments of the $\por$ and $\pand$ operations to be literals and not just
   variables.
 \end{itemize}
-\item A CRAT proof must show that the schema is logically equivalent
-  to the input formula, not just that they are equisatisfiable.
-  Therefore, each deletion step must also be shown to be equivalence
-  preserving, either because the clause is blocked or it follows from
-  remaining clauses by the RAT property.
-\item Unlike QRAT, it need not support universal quantification.
+\item
+  Clauses can also be added when they satisfy the RUP
+  property, with respect to a sequence of existing clauses.  This sequence can
+  be either supplied explicitly or inferred by the proof checker.
+\item
+  Deleting clauses requires proving that the resulting formula is not weaker.
+  \begin{itemize}
+    \item For an input clause or a clause declared by RUP addition, its
+      deletion must be accompanied by a sequence of remaining clauses
+      providing a RUP proof of the clause.  This sequence can either
+      be provided explicitly or inferred by the proof checker.
+    \item The clauses defining a $\pand$ or $\por$ operation are
+      implicitly deleted by deleting the operation.  This can only be
+      done when the only undeleted clauses containing
+      references to the associated extension variable are those
+      implicitly defined when the operation was declared.
+      This implies that there can be no undeleted operations having
+      the operation result as an argument.
+  \end{itemize}
 \end{itemize}
@@
     \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
-    $C$ & {\tt ab} & $L^{+}$ {\tt 0} & ${\tt -}C^{*}$  {\tt 0}  & Add blocked clause \\
-    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RAT clause \\
-     & {\tt db} & $C$             & ${\tt -}C^{+}$  {\tt 0} & Delete blocked clause \\
-     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RAT clause \\
+    $C$ & {\tt a} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RUP clause.  Explicit hint \\
+    $C$ & {\tt a} & $L^{*}$ {\tt 0} &  {\tt * 0} & Add RUP clause.  Inferred hint \\
+     & {\tt d} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause.  Explicit hint \\
+     & {\tt d} & $C$             & {\tt * 0} & Delete RUP clause.  Inferred hint \\
     \midrule
-        & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
-        & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
+    $C$    & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
+    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation.  Explicit hint \\ 
+    $C$    & {\tt s} & $V \; L \; L$    & {\tt * 0}  & Declare $\por$ operation.  Inferred hint \\ 
 %%        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
@@
 proofs.  Our syntax optionally allows input clauses to be listed with
-a rule of type {\tt i}.  Clauses can be added via a blocked-clause addition (command {\tt ab})
-or a RAT addition (command {\tt ar}).
-As described below, however, blocked clauses can only be added
-to define $\pand$ and $\por$
-%%, and $\pite$
-operations.  The hints portion
-of a blocked-clause addition lists all earlier clauses containing the
-negated version of the pivot literal, with the clause IDs negated.
-The hints portion of a RAT addition must contain a sequence of clause
-IDs such that the added clause is RAT with respect to these clauses.
-Similarly, a clause can be deleted if it blocked with respect to one
-of its literals (command {\tt db}), or because it is RAT (command {\tt
-  dr}) The hints portion of a RAT deletion operation must be an
-ordered list of clause IDs justifying the deletion.
+a rule of type {\tt i}.  Clauses can be added via RUP addition
+(command {\tt a}), with the sequence of antecedent clauses (the
+``hint'') either provided explicitly or to be inferred (indicated by the character `{\tt *}') by the proof
+checker.  Similarly for clause deletion (command {\tt d}).
 
-The second set of proof rules is unique to the CRAT format.  Each of
-these indicates the addition of an extension variable.  For each case,
-the rule must be followed by a sequence of blocked-clause additions
-providing the defining clauses for the extension variable.
+The declaration of a $\pand$ operation has the form:
+\begin{center}
+\begin{tabular}{ccccc}
+  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$l_1$} & \makebox[5mm]{$l_2$} \\
+\end{tabular}
+\end{center}
+where $i$ is a new clause ID, $v$ is a positive integer that does not
+correspond to any previous variable, and $l_1$ and $l_2$ are signed
+integers representing literals of existing variables.
+This declaration implicitly causes three clauses to be added to the proof:
+\begin{center}
+\begin{tabular}{cccc}
+\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
+  $i$ & $v$ & $-l_1$ & $-l_2$ \\
+  $i\!+\!1$ & $-v$ & $l_1$ & \\
+  $i\!+\!2$ & $-v$ & $l_2$ & \\
+\end{tabular}
+\end{center}
+The dependency sets for the arguments represented by literals $l_1$ and $l_2$ must be disjoint.
 
-A product rule of the form ${\tt p}\;v\;l_1\;l_2$ indicates that $v$
-will represent the product $l_1 \pand l_2$.  The blocked clause
-additions must encode the formula $v \leftrightarrow (l_1 \land l_2)$.
-Literals $l_1$ and $l_2$ must have disjoint dependency sets.
+The declaration of a $\por$ operation has the form:
+\begin{center}
+\begin{tabular}{ccccccc}
+  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$l_1$} & \makebox[5mm]{$l_2$} 
+\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
+\end{tabular}
+\end{center}
+where $i$ is a new clause ID, $v$ is a positive integer that does
+not correspond to any previous variable, and $l_1$ and $l_2$ are
+signed integers representing literals of existing variables.  Hint $H$
+can consist either of the single character \texttt{*}, or it can be a
+sequence of clause IDs.
 
-A sum rule of the form ${\tt s}\;v\;l_1\;l_2$ indicates that $v$ will
-represent the disjunction $l_1 \por l_2$.  The blocked clause
-additions must encode the formula $v \leftrightarrow (l_1 \lor l_2)$.
-The rule also contains a sequence of clause IDs such that the clause
-$\obar{l}_1 \lor \obar{l}_2$ is RAT with respect to the sequence.
-
-%% An if-then-else rule of the form ${\tt ite}\;v\;l_1\;l_2\;l_3$
-%% indicates that $v$ will represent $\pite(l_1, l_2, l_3)$.  The blocked
-%% clause additions must encode the formula
-%% $v \leftrightarrow \ite((l_1,l_2, l_3)$.
-%% The dependency set for $l_1$ must be disjoint from those
-%% of $l_2$ and $l_3$.
+This declaration implicitly causes three clauses to be added to the proof:
+\begin{center}
+\begin{tabular}{cccc}
+\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
+  $i$ & $-v$ & $l_1$ & $l_2$ \\
+  $i\!+\!1$ & $v$ & $-l_1$ \\
+  $i\!+\!2$ & $v$ & $-l_2$ \\
+\end{tabular}
+\end{center}
+The explicit or implied hints must provide a RUP proof of the clause $-l_1 \lor -l_2$.
 
@@
 
-A CRAT proof follows the same general form as a QRAT dual proof~\cite{bryant:cade:2021}---one
-that ensures that each clause addition and each clause deletion
-preserves equivalence.  Starting with the set of input clauses, it
-produces a sequence of steps that both add and delete clauses.  Each
-addition must be truth preserving and each deletion must be falsehood
-preserving.  At the end, all input clauses must have been deleted, and
-among the remaining clauses there must be only a single unit clause
-consisting of some variable or its complement.  Except for trivial
-cases, the final literal will be an extension variable or its
-complement.  That literal will indicate the root of the schema as the
-(possibly negated) output of a schema operation.  Working from that
-root backward, the schema can be extracted from the CRAT file.
+A CRAT proof follows the same general form as a QRAT dual
+proof~\cite{bryant:cade:2021}---one that ensures that each clause
+addition and each clause deletion preserves equivalence.  With CRAT,
+however, clauses are defined both explicitly and implicitly.  Starting
+with the set of input clauses, the proof consists of a sequence of steps that
+both add and delete clauses.  Each addition must be truth preserving
+and each deletion must be falsehood preserving.  At the end, all input
+clauses must have been deleted, and among the remaining clauses there
+must be only a single unit clause consisting of some variable or its
+complement.  Except for trivial cases, the final literal will be an
+extension variable or its complement.  That literal will indicate the
+root of the schema as the (possibly negated) output of a schema
+operation.  Working from that root backward, the schema can be
+extracted from the CRAT file.
 
@@
 \begin{figure}
-  \centering{
-  \begin{tabular}{llllll}
-    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
+  A).  CRAT File Contents
+  \begin{center}
+  \begin{tabular}{lcllll}
+    \multicolumn{4}{c}{File line} & & \multicolumn{1}{c}{Explanation} \\
 \midrule
-    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause}\\
-            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
-    {\tt 2} & {\tt ab}  & {\tt 4  2 -3 0}  & {\tt 0}    & & Defining clauses for $p_4$ \\ 
-    {\tt 3} & {\tt ab}  & {\tt -4 -2 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
-    {\tt 4} & {\tt ab}  & {\tt -4 3 0}    & {\tt -2 0} & & \\ %%Defining clause for $p_4$ \\ 
-            & {\tt s}   & {\tt 5 2 4}   & {\tt 3 0}  & & Declare $s_5 = x_2 \por p_4$ \\
-    {\tt 5} & {\tt ab}  & {\tt -5 2 4 0}  & {\tt 0}    & & Defining clauses for $s_5$ \\ 
-    {\tt 6} & {\tt ab}  & {\tt  5 -2 0}    & {\tt -5 0} &  \\  %%Defining clause for $s_5$ \\ 
-    {\tt 7} & {\tt ab}  & {\tt  5 -4 0}    & {\tt -5 0} & &  \\  %%Defining clause for $s_5$ \\ 
-            & {\tt p}   & {\tt 6 -1 5}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
-    {\tt 8} & {\tt ab}  & {\tt 6  1 -5 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
-    {\tt 9} & {\tt ab}  & {\tt -6 -1 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
-    {\tt 10} & {\tt ab}  & {\tt -6 5 0}    & {\tt -8 0} & &  \\  %%Defining clause for $p_6$ \\ 
-             & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
-    {\tt 11} & {\tt ab}  & {\tt -7 1 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
-    {\tt 12} & {\tt ab}  & {\tt  7 -1 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
-    {\tt 13} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -12 0} & &  \\  %%Defining clause for $s_7$ \\ 
-    {\tt 14} & {\tt ar}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
-             & {\tt dr}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
+    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause}\\
+    {\tt 2}        & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
+    {\tt 5}        & {\tt s}   & {\tt 5 2 4}   & {\tt 3 0}  & & Declare $s_5 = x_2 \por p_4$ \\
+    {\tt 8}        & {\tt p}   & {\tt 6 -1 5}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
+    {\tt 11}         & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
+    {\tt 14} & {\tt a}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
+             & {\tt d}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
   \end{tabular}
-  }
-  \caption{CRAT Proof \#1 for Formula $x_1 \lor x_2 \lor x_3$}
+  \end{center}
+B). Proof Clauses
+    \begin{center}
+  \begin{tabular}{lcllll}
+    \multicolumn{4}{c}{Clause} & & \multicolumn{1}{c}{Explanation} \\
+\midrule
+    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause}\\
+    {\tt 2} &     & {\tt 4  2 -3 0}  &     & & Defining clauses for $p_4$ \\ 
+    {\tt 3} &     & {\tt -4 -2 0}    & & & \\ %%Defining clause for $p_4$ \\ 
+    {\tt 4} &     & {\tt -4 3 0}    &  & & \\ %%Defining clause for $p_4$ \\ 
+            &     & {\tt -2 -4 0}   & {\tt 3 0}  & &  Mutual exclusion proof for $s_5$ \\
+    {\tt 5} &     & {\tt -5 2 4 0}  &     & & Defining clauses for $s_5$ \\ 
+    {\tt 6} &     & {\tt  5 -2 0}    &  &  \\  %%Defining clause for $s_5$ \\ 
+    {\tt 7} &     & {\tt  5 -4 0}    &  & &  \\  %%Defining clause for $s_5$ \\ 
+    {\tt 8} &     & {\tt 6  1 -5 0}  &    & & Defining clauses for $p_6$ \\ 
+    {\tt 9} &     & {\tt -6 -1 0}    &  & &  \\  %%Defining clause for $p_6$ \\ 
+    {\tt 10} &     & {\tt -6 5 0}    &  & &  \\  %%Defining clause for $p_6$ \\ 
+            &     & {\tt -1 -6 0}   & {\tt 9 0}  & &  Mutual exclusion proof for $s_7$ \\
+    {\tt 11} &     & {\tt -7 1 6 0}  &  & & Defining clauses for $s_7$ \\ 
+    {\tt 12} &     & {\tt  7 -1 0}     & &  \\  %%Defining clause for $s_7$ \\ 
+    {\tt 13} &     & {\tt  7 -6 0}     & &  \\  %%Defining clause for $s_7$ \\ 
+    {\tt 14} &     & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
+             & {\tt d}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
+  \end{tabular}
+  \end{center}
+  \caption{CRAT file \#1 for formula $x_1 \lor x_2 \lor x_3$, and the resulting set of proof clauses}
   \label{fig:c3:crat}
@@
 
-Figure~\ref{fig:c3:crat} shows an annotated version of the CRAT proof
-for this example.  Clause \#1
-is the input clause, and
-clauses \#2--\#13 are the defining clauses for the four operations.
-Each of the two sum operations lists one of the earlier defining
-clauses as a proof that its arguments are mutually exclusive.
-Proof clause \#14 adds the unit clause for the extension variable $s_7$.
+Figure~\ref{fig:c3:crat}A shows an annotated version of the CRAT file
+for this example, while \ref{fig:c3:crat}B shows the sequence of clauses that constitute the proof, including those defined implicitly, and those required for mutual exclusion checks.
+   Clause 1
+is the input clause.
+Clauses 2--13 are added explcitly as the defining clauses for the four operations.
+Also shown are the required mutual exclusion proofs for the two sum operations.
+Proof clause 14 adds the unit clause for the extension variable $s_7$.
 We refer to the literal representing formula $\phi_1$ as $t_1$,
@@
 \begin{figure}
-  \centering{
-  \begin{tabular}{llllll}
+  A).  CRAT File Contents
+  \begin{center}
+  \begin{tabular}{lcllll}
     \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
 \midrule
-    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause $C_1$}\\
 
+    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause $C_1$}\\
     {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
-            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
-    {\tt 3} & {\tt ab}  & {\tt 4 2 3 0} & {\tt 0}    & & Defining clauses for $p_4$ \\
-    {\tt 4} & {\tt ab}  & {\tt -4 -2 0} & {\tt -2 0}    & & \\
-    {\tt 5} & {\tt ab}  & {\tt -4 -3 0} & {\tt -2 0}    & & \\
-            & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
-    {\tt 6} & {\tt ab}  & {\tt 5 1 -4 0} & {\tt 0}    & & Defining clauses for $p_5$ \\
-    {\tt 7} & {\tt ab}  & {\tt -5 -1 0} & {\tt -6 0}    & & \\
-    {\tt 8} & {\tt ab}  & {\tt -5 4 0} & {\tt -6 0}    & & \\
-    {\tt 9} & {\tt ar}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
-            & {\tt dr}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
-            & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
-    {\tt 10} & {\tt ab}  & {\tt 6 -1 2 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
-    {\tt 11} & {\tt ab}  & {\tt -6 1 0}    & {\tt -10 0} & & \\
-    {\tt 12} & {\tt ab}  & {\tt -6 -2 0}    & {\tt -10 0} & & \\ 
-    {\tt 13} & {\tt ar} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
-            & {\tt dr}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
-            & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
-    {\tt 14} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
-    {\tt 15} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -14 0} &  \\  
-    {\tt 16} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -14 0} & &  \\
-    {\tt 17} & {\tt ar}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
-             & {\tt dr}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
-             & {\tt dr}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
+    {\tt 3} & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
+    {\tt 6} & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
+    {\tt 9} & {\tt a}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
+            & {\tt d}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
+    {\tt 10} & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
+    {\tt 13} & {\tt a} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
+            & {\tt d}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
+    {\tt 14} & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
+    {\tt 17} & {\tt a}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
+             & {\tt d}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
+             & {\tt d}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
   \end{tabular}
-}  
-  \caption{CRAT Proof \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
+  \end{center}  
+B). Proof Clauses
+  \begin{center}
+  \begin{tabular}{lcllll}
+    \multicolumn{4}{c}{Clause} & & \multicolumn{1}{c}{Explanation} \\
+\midrule
+    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause $C_1$}\\
+    {\tt 2} &    & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
+    {\tt 3} &   & {\tt 4 2 3 0} &     & & Defining clauses for $p_4$ \\
+    {\tt 4} &   & {\tt -4 -2 0} &     & & \\
+    {\tt 5} &   & {\tt -4 -3 0} &     & & \\
+
+    {\tt 6} &   & {\tt 5 1 -4 0} &    & & Defining clauses for $p_5$ \\
+    {\tt 7} &   & {\tt -5 -1 0} &     & & \\
+    {\tt 8} &   & {\tt -5 4 0} &      & & \\
+
+    {\tt 9} &   & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
+            & {\tt d}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
+
+    {\tt 10} &   & {\tt 6 -1 2 0}  &    & & Defining clauses for $p_6$ \\ 
+    {\tt 11} &   & {\tt -6 1 0}    &  & & \\
+    {\tt 12} &   & {\tt -6 -2 0}    &  & & \\ 
+    {\tt 13} &  & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
+            & {\tt d}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
+             &          & {\tt -5 -6 0} & {\tt 7 11 0}  & & Mutual exclusion proof for $s_7$ \\
+    {\tt 14} &   & {\tt -7 5 6 0}  &     & & Defining clauses for $s_7$ \\ 
+    {\tt 15} &   & {\tt  7 -5 0}    &  &  \\  
+    {\tt 16} &   & {\tt  7 -6 0}    &  & &  \\
+    {\tt 17} &   & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
+             & {\tt d}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
+             & {\tt d}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
+  \end{tabular}
+  \end{center}  
+  \caption{CRAT file \#2A for formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$, and the resulting set of proof clauses}
   \label{fig:p2-bdd:crat}
@@
 Figure~\ref{fig:p2-bdd:schema} shows a schema representing the formula,
-and Figure~\ref{fig:p2-bdd:crat} shows the associated CRAT proof.  This
+and Figure~\ref{fig:p2-bdd:crat} shows the associated CRAT file and proof clauses.  This
 proof was generated via a bottom-up strategy, such as would be created
@@
 \begin{figure}
-  \centering{
+  A).  CRAT File Contents
+  \begin{center}
   \begin{tabular}{llllll}
-    \multicolumn{4}{c}{Proof line} & & \multicolumn{1}{c}{Explanation} \\
+    \multicolumn{4}{c}{File line} & & \multicolumn{1}{c}{Explanation} \\
 \midrule
-    \makebox[5mm][l]{\tt 1} & \makebox[7mm][l]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[15mm]{}          & \makebox[5mm]{} & \makebox[20mm][l]{Input clause $C_1$}\\
-
+    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause $C_1$}\\
     {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
-            & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
-    {\tt 3} & {\tt ab}  & {\tt 4 2 3 0} & {\tt 0}    & & Defining clauses for $p_4$ \\
-    {\tt 4} & {\tt ab}  & {\tt -4 -2 0} & {\tt -2 0}    & & \\
-    {\tt 5} & {\tt ab}  & {\tt -4 -3 0} & {\tt -2 0}    & & \\
-            & {\tt p}   & {\tt 5 -1 -4}  &             & & Declare $p_5 = \obar{x}_1 \pand \obar{p_4}$ \\
-    {\tt 6} & {\tt ab}  & {\tt 5 1 4 0} & {\tt 0}    & & Defining clauses for $p_5$ \\
-    {\tt 7} & {\tt ab}  & {\tt -5 -1 0} & {\tt -6 0}    & & \\
-    {\tt 8} & {\tt ab}  & {\tt -5 -4 0} & {\tt -6 0}    & & \\
-            & {\tt p}   & {\tt 6 1 2} &              & & Declare $p_6 = x_1 \pand x_2$ \\    
-    {\tt 9} & {\tt ab}  & {\tt 6 -1 -2 0}  & {\tt 0}    & & Defining clauses for $p_6$ \\ 
-    {\tt 10} & {\tt ab}  & {\tt -6 1 0}    & {\tt -9 0} & & \\
-    {\tt 11} & {\tt ab}  & {\tt -6 2 0}    & {\tt -9 0} & & \\ 
-            & {\tt s}   & {\tt 7 5 6}   & {\tt 7 10 0}  & & Declare $s_7 = p_5 \por p_6$ \\
-    {\tt 12} & {\tt ab}  & {\tt -7 5 6 0}  & {\tt 0}    & & Defining clauses for $s_7$ \\ 
-    {\tt 13} & {\tt ab}  & {\tt  7 -5 0}    & {\tt -12 0} & & \\  
-    {\tt 14} & {\tt ab}  & {\tt  7 -6 0}    & {\tt -12 0} & & \\
-
-    {\tt 15} & {\tt ar}  & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p_4}$ \\
-    {\tt 16} & {\tt ar}  & {\tt 1 7 0}     & {\tt 13 6 15 0} & & Justify $\obar{x}_1 \rightarrow s_7$ \\
-    {\tt 17} & {\tt ar}  & {\tt -1 7 0}    & {\tt 14 9 2 0} & & Justify $x_1 \rightarrow s_7$ \\
-    {\tt 18} & {\tt ar}  & {\tt 7 0}       & {\tt 16 17 0}  & & Justify unit clause $t = s_7$ \\
-             & {\tt dr}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
-             & {\tt dr}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\
+    {\tt 3} & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
+    {\tt 6} & {\tt p}   & {\tt 5 -1 -4}  &             & & Declare $p_5 = \obar{x}_1 \pand \obar{p_4}$ \\
+    {\tt 9} & {\tt p}   & {\tt 6 1 2} &              & & Declare $p_6 = x_1 \pand x_2$ \\    
+    {\tt 12} & {\tt s}   & {\tt 7 5 6}   & {\tt 7 10 0}  & & Declare $s_7 = p_5 \por p_6$ \\
+    {\tt 15} & {\tt a}  & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p_4}$ \\
+    {\tt 16} & {\tt a}  & {\tt 1 7 0}     & {\tt 13 6 15 0} & & Justify $\obar{x}_1 \rightarrow s_7$ \\
+    {\tt 17} & {\tt a}  & {\tt -1 7 0}    & {\tt 14 9 2 0} & & Justify $x_1 \rightarrow s_7$ \\
+    {\tt 18} & {\tt a}  & {\tt 7 0}       & {\tt 16 17 0}  & & Justify unit clause $t = s_7$ \\
+             & {\tt d}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
+             & {\tt d}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\
   \end{tabular}
-}  
-  \caption{CRAT Proof \#2B for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$}
+  \end{center}  
+B). Proof Clauses
+  \begin{center}
+  \begin{tabular}{llllll}
+    \multicolumn{4}{c}{File line} & & \multicolumn{1}{c}{Explanation} \\
+\midrule
+    \makebox[5mm][l]{\tt 1} & \makebox[7mm]{}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause $C_1$}\\
+    {\tt 2} &    & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
+    {\tt 3} &   & {\tt 4 2 3 0} &     & & Defining clauses for $p_4$ \\
+    {\tt 4} &   & {\tt -4 -2 0} &     & & \\
+    {\tt 5} &   & {\tt -4 -3 0} &     & & \\
+    {\tt 6} &   & {\tt 5 1 4 0} &     & & Defining clauses for $p_5$ \\
+    {\tt 7} &   & {\tt -5 -1 0} &     & & \\
+    {\tt 8} &   & {\tt -5 -4 0} &     & & \\
+    {\tt 9} &   & {\tt 6 -1 -2 0}&     & & Defining clauses for $p_6$ \\ 
+    {\tt 10} &   & {\tt -6 1 0}    &  & & \\
+    {\tt 11} &   & {\tt -6 2 0}    &  & & \\ 
+     &           & {\tt -5 -6}   & {\tt 7 10 0}  & & Mutual exclusion proof for $s_7$ \\
+    {\tt 12} &   & {\tt -7 5 6 0}  &     & & Defining clauses for $s_7$ \\ 
+    {\tt 13} &   & {\tt  7 -5 0}    &  & & \\  
+    {\tt 14} &   & {\tt  7 -6 0}    &  & & \\
+    {\tt 15} &   & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p_4}$ \\
+    {\tt 16} &   & {\tt 1 7 0}     & {\tt 13 6 15 0} & & Justify $\obar{x}_1 \rightarrow s_7$ \\
+    {\tt 17} &   & {\tt -1 7 0}    & {\tt 14 9 2 0} & & Justify $x_1 \rightarrow s_7$ \\
+    {\tt 18} &   & {\tt 7 0}       & {\tt 16 17 0}  & & Justify unit clause $t = s_7$ \\
+             & {\tt d}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
+             & {\tt d}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\
+  \end{tabular}
+  \end{center}  
+  \caption{CRAT file \#2B for formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$, and the resulting set of proof clauses}
   \label{fig:p2-cdcl:crat}

### 2022-05-11 doc/mc-notes.tex
Fix formatting problems

@@
     {\tt 3} & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
-    {\tt 6} & {\tt p}   & {\tt 5 -1 -4}  &             & & Declare $p_5 = \obar{x}_1 \pand \obar{p_4}$ \\
+    {\tt 6} & {\tt p}   & {\tt 5 -1 -4}  &             & & Declare $p_5 = \obar{x}_1 \pand \obar{p}_4$ \\
     {\tt 9} & {\tt p}   & {\tt 6 1 2} &              & & Declare $p_6 = x_1 \pand x_2$ \\    
     {\tt 12} & {\tt s}   & {\tt 7 5 6}   & {\tt 7 10 0}  & & Declare $s_7 = p_5 \por p_6$ \\
-    {\tt 15} & {\tt a}  & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p_4}$ \\
+    {\tt 15} & {\tt a}  & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p}_4$ \\
     {\tt 16} & {\tt a}  & {\tt 1 7 0}     & {\tt 13 6 15 0} & & Justify $\obar{x}_1 \rightarrow s_7$ \\
@@
     {\tt 14} &   & {\tt  7 -6 0}    &  & & \\
-    {\tt 15} &   & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p_4}$ \\
+    {\tt 15} &   & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p}_4$ \\
     {\tt 16} &   & {\tt 1 7 0}     & {\tt 13 6 15 0} & & Justify $\obar{x}_1 \rightarrow s_7$ \\
@@
 in Figure~\ref{fig:p2-cdcl:schema}A\@.  The splitting on variable
-$x_1$ can be rejoined as $\ite(x_1, x_2, \obar{p_4})$, and this ITE
+$x_1$ can be rejoined as $\ite(x_1, x_2, \obar{p}_4)$, and this ITE
 operation can be expressed using the product operations $p_5$ and

### 2022-06-13 doc/mc-notes.tex
Expand example cost functions.  Add operator deletion

@@
 
-{\bf Example 1}: Let $\ring$ be the set of rational numbers of the form $a\cdot 2^b$ where $a$ and $b$ are integers.
+{\bf Example 1a}: Let $\ring$ be the set of rational numbers of the form $a\cdot 2^b$ where $a$ and $b$ are integers.
 Let $\cost(x_i) = 1/2$ for all variables $x_i$.  The cost of every
@@
 
-{\bf Example 2}: Let $\ring$ be the set of rational numbers.  Assign a
-{\em weight} $w_i$ to each variable $x_i$ such that $0 \leq w_i \leq
-1$ and let $\cost(x_i) = w_i$.  This implements {\em weighted model counting}
+{\bf Example 1b}: Let $\ring$ be the set of rational numbers.  Assign a
+{\em weight} $w(x_i)$ to each variable $x_i$ such that
+$0 \leq w(x_i) \leq 1$
+and let $\cost(x_i) = w(x_i)$.  This implements {\em weighted model counting}
 under the restrictions that: 
@@
 
-{\bf Example 3}: Let $\ring$ be a finite field with $z = |\ring| \geq 2n$,
+{\bf Example 1c}: Let $\ring$ be the set of rational numbers.  Assign
+{\em weights} to each literal.  In particular, for each variable
+$x_i$, define weights $w(x_i)$ and $w(\obar{x}_i)$, such that
+$\beta(x_i) \doteq w(x_i) + w(\obar{x}_i) \not = 0$.  We can convert
+this to the formulation of Example 1b by using weight $w'(x_i) =
+w(x_i)/\beta(x_i)$ and then scaling the final result by $\prod_{1 \leq
+  i \leq n} \beta(x_i)$.  Scaling eliminates the need to have the weight
+of each variable and its negation sum to 1.  This form is sufficiently general
+to handle the benchmarks used in weighted model counting track of the Model Counting Competition.
+Using $w(x_i) = w(\obar{x}_i) = 1$ for each variable $x_i$,
+gives the standard model counting of Example 1a.
+
+{\bf Example 2}: Let $\ring$ be a finite field with $z = |\ring| \geq 2n$,
 and let $\hashset$ be the set of functions
@@
 \begin{table}
-  \caption{Recursive Definition of Separable Schemas}
+  \caption{Recursive Definition of Separable Schemas and their Costs}
   \label{tab:schema}
@@
 \begin{center}
-\begin{tabular}{cccc}
+\begin{tabular}{ccccc}
 \toprule
-  \makebox[15mm]{$S$}           & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$}\\
+  \makebox[15mm]{$S$}  & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$} & \makebox[25mm]{$\cost(S)$}\\
 \midrule
-  $\nil$  & None  & $\emptyset$    & $\emptyset$ \\
-  $\tautology$        & None  & $\emptyset$    & $\uinterp$ \\
-$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ \\
-$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ \\
-$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$ \\
-$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$ \\
+  $\nil$  & None  & $\emptyset$    & $\emptyset$  & $0$\\
+  $\tautology$        & None  & $\emptyset$    & $\uinterp$ & $1$ \\
+$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ & $\cost(x_i)$ \\
+$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ & $\oneminus\cost(S)$ \\
+$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$
+& $\cost(S_1) \cdot \cost(S_2)$\\
+$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$
+& $\cost(S_1) + \cost(S_2)$\\
 \bottomrule
@@
 %% from basic operations.  The latter category consists of just the $\pite$ operation.
-
 A key property of a Boolean formula represented by separable schema
@@
 formula $\cost(S)$ can be computed with a linear number of ring
-operations.
+operations, as is shown in the final column of Table~\ref{tab:schema}.
 
@@
     $C$ & {\tt a} & $L^{*}$ {\tt 0} &  {\tt * 0} & Add RUP clause.  Inferred hint \\
-     & {\tt d} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause.  Explicit hint \\
-     & {\tt d} & $C$             & {\tt * 0} & Delete RUP clause.  Inferred hint \\
+     & {\tt dc} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause.  Explicit hint \\
+     & {\tt dc} & $C$             & {\tt * 0} & Delete RUP clause.  Inferred hint \\
     \midrule
@@
 %%        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
+     & {\tt do} & $V$             &  & Delete operation. \\
+
   \end{tabular}
@@
 ``hint'') either provided explicitly or to be inferred (indicated by the character `{\tt *}') by the proof
-checker.  Similarly for clause deletion (command {\tt d}).
+checker.  Similarly for clause deletion (command {\tt dc}).
 
@@
 sequence of clause IDs.
-
 This declaration implicitly causes three clauses to be added to the proof:
@@
 The explicit or implied hints must provide a RUP proof of the clause $-l_1 \lor -l_2$.
+
+Finally, operator deletion ({\tt do}) specifies which operator to
+delete via the extension variable associated with the operator.  Doing
+so for operator $v$ requires 1) there is no (undeleted) operator $u$
+having a literal of $v$ as one of its inputs, and 2) there is no
+(undeleted) RUP clause containing a literal of $v$.
 
@@
     {\tt 14} & {\tt a}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
-             & {\tt d}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
+             & {\tt dc}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
   \end{tabular}
@@
     {\tt 14} &     & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
-             & {\tt d}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
+             & {\tt dc}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
   \end{tabular}
@@
     {\tt 9} & {\tt a}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
-            & {\tt d}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
+            & {\tt dc}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
     {\tt 10} & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
     {\tt 13} & {\tt a} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
-            & {\tt d}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
+            & {\tt dc}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
     {\tt 14} & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
     {\tt 17} & {\tt a}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
-             & {\tt d}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
-             & {\tt d}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
+             & {\tt dc}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
+             & {\tt dc}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
   \end{tabular}
@@
     {\tt 9} &   & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
-            & {\tt d}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
+            & {\tt dc}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
 
@@
     {\tt 13} &  & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
-            & {\tt d}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
+            & {\tt dc}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
              &          & {\tt -5 -6 0} & {\tt 7 11 0}  & & Mutual exclusion proof for $s_7$ \\
@@
     {\tt 17} &   & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
-             & {\tt d}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
-             & {\tt d}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
+             & {\tt dc}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
+             & {\tt dc}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
   \end{tabular}
@@
     {\tt 18} & {\tt a}  & {\tt 7 0}       & {\tt 16 17 0}  & & Justify unit clause $t = s_7$ \\
-             & {\tt d}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
-             & {\tt d}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\
+             & {\tt dc}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
+             & {\tt dc}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\
   \end{tabular}
@@
     {\tt 18} &   & {\tt 7 0}       & {\tt 16 17 0}  & & Justify unit clause $t = s_7$ \\
-             & {\tt d}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
-             & {\tt d}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\
+             & {\tt dc}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
+             & {\tt dc}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\
   \end{tabular}

### 2022-06-15 doc/mc-notes.tex
Start discussion of Tseitin variables
Fix typo in proof

@@
     {\tt 11}         & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
-    {\tt 14} & {\tt a}  & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
+    {\tt 14} & {\tt a}  & {\tt 7 0} & {\tt 13 12 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
              & {\tt dc}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
@@
     {\tt 13} &     & {\tt  7 -6 0}     & &  \\  %%Defining clause for $s_7$ \\ 
-    {\tt 14} &     & {\tt 7 0} & {\tt 12 13 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
+    {\tt 14} &     & {\tt 7 0} & {\tt 13 12 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
              & {\tt dc}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
@@
     {\tt 6} & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
-    {\tt 9} & {\tt a}  & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
-            & {\tt dc}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
+    {\tt 9} & {\tt a}  & {\tt -5 0} & {\tt 7 8 5 4 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
+            & {\tt dc}  & {\tt 1 } & {\tt 3 6 9 0} & & Delete clause $C_1$\\
     {\tt 10} & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
-    {\tt 13} & {\tt a} & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
+    {\tt 13} & {\tt a} & {\tt -6 0} & {\tt 12 11 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
             & {\tt dc}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
@@
 
-    {\tt 9} &   & {\tt -5 0} & {\tt 7 8 4 5 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
-            & {\tt dc}  & {\tt 1 } & {\tt 3 6 7 0} & & Delete clause $C_1$\\
+    {\tt 9} &   & {\tt -5 0} & {\tt 7 8 5 4 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
+            & {\tt dc}  & {\tt 1 } & {\tt 3 6 9 0} & & Delete clause $C_1$\\
 
@@
     {\tt 12} &   & {\tt -6 -2 0}    &  & & \\ 
-    {\tt 13} &  & {\tt -6 0} & {\tt 11 12 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
+    {\tt 13} &  & {\tt -6 0} & {\tt 12 11 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
             & {\tt dc}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
@@
 
+%% \subsection{Tseitin Variable Optimization}
+
+%% Many problems cannot be directly encoded in CNF without an exponential
+%% blowup.  This can be avoided by introducing auxilliary variables.  A
+%% particular approach to these encodings makes use of {\em Tseitin}
+%% variables.  The general scheme with these is to let a variable $z$
+%% encode a subformula $F$ over input and prior Tseitin variables by
+%% adding clauses encoding $z \leftrightarrow F$.  A particular property
+%% of this encoding is that if we let ${\cal C}_z$ (respectively ${\cal
+%%   C}_{\obar{z}}$) denote the clauses in which literal $z$ (resp.,
+%% $\obar{z}$) occcurs
+
 \subsection{TO-DO List}

### 2022-06-15 benchmarks/generators/clause.py
Generate single clause CNF + CRAT

@@ -1,0 +1,86 @@
+#!/usr/bin/python3
+
+import sys
+import getopt
+import writer
+
+def usage(name):
+    print("Usage: %s [-h] [-v] [-p] -n N [-r ROOT]")
+    print(" -h      Print this message")
+    print(" -v      Add comments to files")
+    print(" -p      Generate CRAT proof")
+    print(" -n N    Set number of literals")
+    print(" -r ROOT Set file root name (default = clause-NNN)")
+
+# Generate CNF consisting of single clause with N literals
+literals = []
+clauseList = []
+
+def cnfGen(froot, n, verbose):
+    global literals, clauseList
+    cwriter = writer.LazyCnfWriter(froot)
+    if verbose:
+        cwriter.doComment("CNF file consisting of single clause with %d literals" % n)
+    vars = cwriter.newVariables(n)
+    literals = [v if v%2==0 else -v for v in vars]
+    cwriter.doClause(literals)
+    clauseList = cwriter.clauseList()
+    cwriter.finish()
+
+def proofGen(froot, n, verbose):
+    pwriter = writer.CratWriter(n, clauseList, froot, verbose)
+    # Create chain of AND operations, using negated inputs
+    pwriter.doComment("Create chain of %d AND operations with negated inputs" % (n-1))
+    args = [-lit for lit in literals]
+    upSteps = []
+    downSteps = []
+    while len(args) > 1:
+        step = pwriter.stepCount + 1
+        arg1 = args[-2]
+        arg2 = args[-1]
+        downSteps += [step]
+        upSteps += [step+1, step+2]
+        v = pwriter.doAnd(arg1, arg2)
+        args = args[:-2] + [v]
+    # Assert unit clause for root
+    root = -args[0]
+    pwriter.doComment("Assert unit clause for root")
+    upSteps.reverse()
+    pwriter.doClause([root], upSteps + [1])
+    step = pwriter.stepCount
+    # Delete input clause
+    pwriter.doComment("Delete input clause")
+    pwriter.doDeleteClause(1, downSteps + [step])
+    pwriter.finish()
+
+def run(name, args):
+    verbose = False
+    proof = False
+    sn = None
+    root = None
+    optlist, args = getopt.getopt(args, 'hvpn:r:')
+    for (opt, val) in optlist:
+        if opt == '-h':
+            usage(name)
+            return
+        elif opt == '-v':
+            verbose = True
+        elif opt == '-p':
+            proof = True
[…]

### 2022-06-15 benchmarks/generators/writer.py

@@ -1,4 +1,23 @@
-# Code for generating QCNF, orders, and clusters
+#####################################################################################
+# Copyright (c) 2021 Marijn Heule, Randal E. Bryant, Carnegie Mellon University
+# 
+# Permission is hereby granted, free of charge, to any person obtaining a copy of this software and
+# associated documentation files (the "Software"), to deal in the Software without restriction,
+# including without limitation the rights to use, copy, modify, merge, publish, distribute,
+# sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is
+# furnished to do so, subject to the following conditions:
+# 
+# The above copyright notice and this permission notice shall be included in all copies or
+# substantial portions of the Software.
+# 
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT
+# NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+# NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM,
+# DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT
+# OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+########################################################################################
+
+# Code for generating CNF, order, schedule, and crat proof files
 class WriterException(Exception):
 
     def __init__(self, value):
@@ -13,11 +32,17 @@ class Writer:
     outfile = None
     suffix = None
     verbose = False
-    variableCount = None
+    expectedVariableCount = None
+    isNull = False
+    froot = ""
 
-    def __init__(self, froot, suffix = None, verbose = False):
-        self.variableCount = 0
+    def __init__(self, count, froot, suffix = None, verbose = False, isNull = False):
+        self.expectedVariableCount = count
+        self.froot = froot
         self.verbose = verbose
+        self.isNull = isNull
+        if isNull:
+            return
         if suffix is not None:
             self.suffix = suffix 
             fname = froot if self.suffix is None else froot + "." + self.suffix
@@ -27,117 +52,276 @@ class Writer:
             print("Couldn't open file '%s'. Aborting" % fname)
[…]

### 2022-06-19 doc/mc-notes.tex
Clarification about operator deletion

@@
 delete via the extension variable associated with the operator.  Doing
-so for operator $v$ requires 1) there is no (undeleted) operator $u$
-having a literal of $v$ as one of its inputs, and 2) there is no
+so for operator $v$ requires that there is no
 (undeleted) RUP clause containing a literal of $v$.
+Implicitly, that means
+that there is no (undeleted) operator $u$
+having a literal of $v$ as one of its inputs.  The defining clauses for the operation are also deleted.
 

### 2022-06-19 crat/prototype/crat_checker.py
Added (weighted) counting

@@ -27,8 +27,10 @@ import getopt
 import datetime
 
 def usage(name):
-    print("Usage: %s [-v] -i FILE.cnf -p FILE.crat" % name)
-    print("   -v        Print more helpful diagnostic information if there is an error")
+    print("Usage: %s [-v] -i FILE.cnf -p FILE.crat [-w W1:W2:...:Wn]" % name)
+    print("   -v         Print more helpful diagnostic information if there is an error")
+    print("   -w WEIGHTS Provide colon-separated set of input weights.")
+    print("              Each should be between 0 and 100 (will be scaled by 1/100)")
 
 ######################################################################################
 # CRAT Syntax
@@ -340,8 +342,9 @@ class ClauseManager:
 
 class OperationManager:
     conjunction, disjunction = range(2)
-    typeName = ["p", "s"]
-
+    
+    # Number of input variables
+    inputVariableCount = 0
     # Operation indexed by output variable.  Each entry of form (op, arg1, arg2, id)
     operationDict = {}
     # For each variable, the variables on which it depends
@@ -352,6 +355,7 @@ class OperationManager:
     verbose = False
 
     def __init__(self, cmgr, varCount):
+        self.inputVariableCount = varCount
         self.cmgr = cmgr
         self.verbose = cmgr.verbose
         self.operationDict = {}
@@ -415,6 +419,32 @@ class OperationManager:
         del self.dependencySetDict[outVar]
         return (True, "")
 
+    # Optionally provide dictionary of weights.  Otherwise assume unweighted
+    def count(self, root, weights = None):
+        if weights is None:
+            weights = {}
+        beta = 1.0
+        for v in range(1, self.inputVariableCount+1):
+            if v not in weights:
+                weights[v] = 0.5
+                beta *= 2.0
+        for outVar in sorted(self.operationDict.keys()):
+            (op, arg1, arg2, cid) = self.operationDict[outVar]
+            var1 = abs(arg1)
+            val1 = weights[var1]
+            if arg1 < 0:
+                val1 = 1.0 - val1
+            var2 = abs(arg2)
+            val2 = weights[var2]
+            if arg2 < 0:
+                val2 = 1.0 - val2
+            result = val1 * val2 if op == self.conjunction else val1 + val2
+            weights[outVar] = result
+        rootVar = abs(root)
+        rval = weights[rootVar]
+        if root < 0:
+            rval = 1.0 - rval
+        return rval * beta
 
[…]

### 2022-06-19 doc/mc-notes.tex
Fixed proof bugs

@@
     {\tt 11}         & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
-    {\tt 14} & {\tt a}  & {\tt 7 0} & {\tt 13 12 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
+    {\tt 14} & {\tt a}  & {\tt 7 0} & {\tt 13 12 8 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
              & {\tt dc}  & {\tt 1}  & {\tt 4 5 10 11 14 0} & & Delete input clause \\
@@
     {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
-    {\tt 3} & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
+    {\tt 3} & {\tt p}   & {\tt 4 -2 -3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
     {\tt 6} & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
@@
     {\tt 14} & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
-    {\tt 17} & {\tt a}  & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
+    {\tt 17} & {\tt a}  & {\tt -7 0}       & {\tt 9 13 14 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
              & {\tt dc}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
@@
             {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
-    {\tt 3} & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
+            {\tt 3} & {\tt p}   & {\tt 4 -2 -3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
     {\tt 6} & {\tt p}   & {\tt 5 -1 -4}  &             & & Declare $p_5 = \obar{x}_1 \pand \obar{p}_4$ \\

### 2022-06-20 crat/prototype/crat_checker.py
Added precise arithmetic for unweighted count

@@ -61,6 +61,7 @@ def trim(s):
         s = s[:-1]
     return s
 
+
 # Clean up clause.
 # Remove duplicates
 # Sort in reverse order of variable number
@@ -105,6 +106,83 @@ def testClauseEquality(clause1, clause2):
         if l1 != l2:
             return False
     return True
+
+class P2Exception(Exception):
+    msg = ""
+
+    def __init__(self, msg):
+        self.msg = msg
+
+    def __str__(self):
+        return "P2 Number exception %s" % self.msg
+
+
+# Represent numbers of form a * 2^b, where a is odd integer and b is integer
+class P2Num:
+    a = 1
+    b = 0
+
+    def __init__(self, a, b=0):
+        if type(a) != type(1):
+            raise P2Exception("Nonintegral a (%s)" % str(a))
+        if type(b) != type(1):
+            raise P2Exception("Nonintegral b (%s)" % str(a))
+        if a == 0:
+            self.a = 0
+            self.b = 0
+            return
+        while a % 2 == 0:
+            a = a//2
+            b += 1
+        self.a = a
+        self.b = b
+
+    def __str__(self):
+        return "%d*2^(%d)" % (self.a, self.b)
+
+    def num(self):
+        p2 = 2**self.b
+        val = p2 * self.a
+        return val
+
+
+    def neg(self):
+        result = P2Num(-self.a, self.b)
+#        print("- %s --> %s" % (str(self), str(result)))
+        return result
+
+    def oneminus(self):
+        one = P2Num(1, 0)
+        result = one.add(self.neg())
+#        print("1- %s --> %s" % (str(self), str(result)))
+        return result
+
+    def mul(self, other):
+        na = self.a * other.a 
+        nb = self.b + other.b
+        result =  P2Num(na, nb)
+#        print("%s * %s --> %s" % (str(self), str(other), str(result)))
+        return result
+
+    def add(self, other):
+        a1 = self.a
+        b1 = self.b
+        a2 = other.a
+        b2 = other.b
+        if b1 > b2:
+            a1,a2 = a2,a1
+            b1,b2 = b2,b1
+        # Guarantee b2 >= b1
+        p2 = 2 ** (b2-b1)
+        a2 *= p2
+        na = a1 + a2
+        nb = b1
+        result = P2Num(na, nb)
+#        print("%s + %s --> %s" % (str(self), str(other), str(result)))
+        return result
+
+    def p2scale(self, p2):
+        return P2Num(self.a, p2+self.b)
 
 # Read CNF file.
[…]

### 2022-06-20 doc/mc-notes.tex
Simplified hint

@@
     {\tt 18} &   & {\tt 7 0}       & {\tt 16 17 0}  & & Justify unit clause $t = s_7$ \\
-             & {\tt dc}  & {\tt 1}         & {\tt 3 8 10 12 18 0} & & Delete $C_1$\\
+             & {\tt dc}  & {\tt 1}         & {\tt 3 15 0} & & Delete $C_1$\\
              & {\tt dc}  & {\tt 2}         & {\tt 11 7 12 18 0} & & Delete $C_2$\\

### 2022-06-20 crat/prototype/crat_checker.py
Fixes

@@ -590,7 +590,6 @@ class OperationManager:
     def count(self, root, weights = None):
         if weights is None:
             print("Precise count    = %s" % str(self.unweightedCount(root)))
-#            print("I think the count = %s" % str(self.unweightedCount(root)))
             weights = {}
         beta = 1.0
         for v in range(1, self.inputVariableCount+1):

### 2022-06-20 crat/prototype/crat_checker.py
Updated date

@@ -1,8 +1,8 @@
 #!/usr/bin/python3
 
 #####################################################################################
-# Copyright (c) 2021 Marijn Heule, Randal E. Bryant, Carnegie Mellon University
-# Last edit: Feb. 16, 2021
+# Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
+# Last edit: June 19, 2022
 # 
 # Permission is hereby granted, free of charge, to any person obtaining a copy of this software and
 # associated documentation files (the "Software"), to deal in the Software without restriction,

### 2022-07-16 doc/dd/p2-bdd-eval.tex
Expand example cost functions.  Add operator deletion
Start switch to k-argument And

@@
 \begin{tikzpicture}
+\definecolor{fillcolor}{RGB}{255,255,255}
+\definecolor{highcolor}{RGB}{0,0,0}
+\definecolor{lowcolor}{RGB}{0,0,0}
 \definecolor{neutralcolor}{RGB}{0,0,0}
 \definecolor{pathcolor}{RGB}{0,0,0}
-\definecolor{fillcolor}{RGB}{255,255,255}
 \definecolor{background}{RGB}{225,225,225}
-\definecolor{highcolor}{RGB}{0,0,0}
-\definecolor{lowcolor}{RGB}{0,0,0}
-\draw (1.76,7.41) [thin,neutralcolor] -- (1.76,5.65);
-\draw (1.76,5.65) [thin,neutralcolor] -- (0.44,3.88);
-\draw (1.76,5.65) [thin,neutralcolor] -- (3.09,3.88);
-\draw (0.44,3.88) [thin,neutralcolor] -- (1.76,3.00);
-\draw (0.44,3.88) [thin,neutralcolor] -- (0.44,2.12);
-\draw (3.09,3.88) [thin,neutralcolor] -- (1.76,3.00);
-\draw (3.09,3.88) [thin,neutralcolor] -- (3.09,2.12);
-\draw (1.76,3.00) [thin,neutralcolor] -- (1.76,2.12);
-\draw (0.44,2.12) [thin,neutralcolor] -- (0.44,0.35);
-\draw (0.44,2.12) [thin,neutralcolor] -- (1.76,1.24);
-\draw (3.09,2.12) [thin,neutralcolor] -- (1.76,1.24);
+\draw (1.76,6.53) [thin,neutralcolor] -- (1.76,4.77);
+\draw (1.76,4.77) [thin,neutralcolor] -- (0.44,3.00);
+\draw (1.76,4.77) [thin,neutralcolor] -- (3.09,3.00);
+\draw (0.44,3.00) [thin,neutralcolor] -- (3.09,1.24);
+\draw (0.44,3.00) [thin,neutralcolor] -- (0.44,0.35);
+\draw (0.44,3.00) [thin,neutralcolor] -- (1.76,1.24);
+\draw (3.09,3.00) [thin,neutralcolor] -- (3.09,1.24);
+\draw (3.09,3.00) [thin,neutralcolor] -- (1.76,1.24);
+\draw (3.09,1.24) [thin,neutralcolor] -- (3.09,0.35);
 \draw (1.76,1.24) [thin,neutralcolor] -- (1.76,0.35);
-\draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,3.00) circle [radius=0.00];
-\node at (1.76,3.00) {};
+\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,6.53) circle [radius=0.35];
+\node at (1.76,6.53) {$\sigma(\phi_2)$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,4.77) circle [radius=0.44];
+\node at (1.76,4.77) {$+$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.00) circle [radius=0.44];
+\node at (0.44,3.00) {$\times$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (3.09,3.00) circle [radius=0.44];
+\node at (3.09,3.00) {$\times$};
+\draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,1.24) circle [radius=0.00];
+\node at (3.09,1.24) {};
 \draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,1.24) circle [radius=0.00];
 \node at (1.76,1.24) {};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (3.09,3.88) circle [radius=0.44];
-\node at (3.09,3.88) {$\times$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,3.30) circle [radius=0.26];
-\node at (1.32,3.30) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,2.12) circle [radius=0.44];
-\node at (0.44,2.12) {$\times$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.88) circle [radius=0.44];
-\node at (0.44,3.88) {$\times$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,2.12) circle [radius=0.35];
-\node at (1.76,2.12) {$w_1$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (2.21,1.54) circle [radius=0.26];
-\node at (2.21,1.54) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.44];
-\node at (1.76,5.65) {$+$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,7.41) circle [radius=0.35];
-\node at (1.76,7.41) {$\sigma(\phi_2)$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,1.24) circle [radius=0.26];
-\node at (0.44,1.24) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,6.53) circle [radius=0.26];
-\node at (1.76,6.53) {$\sim$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (3.09,0.35) circle [radius=0.35];
+\node at (3.09,0.35) {$w_1$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,0.35) circle [radius=0.35];
@@
 \node at (0.44,0.35) {$w_3$};
-\draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,2.12) circle [radius=0.00];
-\node at (3.09,2.12) {};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,1.54) circle [radius=0.26];
-\node at (1.32,1.54) {$\sim$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.26];
+\node at (1.76,5.65) {$\sim$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,2.12) circle [radius=0.26];
+\node at (1.76,2.12) {$\sim$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.11,2.12) circle [radius=0.26];
+\node at (1.11,2.12) {$\sim$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (2.44,2.12) circle [radius=0.26];
+\node at (2.44,2.12) {$\sim$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,1.68) circle [radius=0.26];
+\node at (0.44,1.68) {$\sim$};
 \end{tikzpicture}%%

### 2022-07-16 doc/dd/p2-bdd.tex
Expand example cost functions.  Add operator deletion
Start switch to k-argument And

--- 2022-05-02 doc/dd/p2-bdd-eval.tex
@@
 \begin{tikzpicture}
+\definecolor{fillcolor}{RGB}{255,255,255}
+\definecolor{highcolor}{RGB}{0,0,0}
+\definecolor{lowcolor}{RGB}{0,0,0}
 \definecolor{neutralcolor}{RGB}{0,0,0}
 \definecolor{pathcolor}{RGB}{0,0,0}
-\definecolor{fillcolor}{RGB}{255,255,255}
 \definecolor{background}{RGB}{225,225,225}
-\definecolor{highcolor}{RGB}{0,0,0}
-\definecolor{lowcolor}{RGB}{0,0,0}
-\draw (1.76,7.41) [thin,neutralcolor] -- (1.76,5.65);
-\draw (1.76,5.65) [thin,neutralcolor] -- (0.44,3.88);
-\draw (1.76,5.65) [thin,neutralcolor] -- (3.09,3.88);
-\draw (0.44,3.88) [thin,neutralcolor] -- (1.76,3.00);
-\draw (0.44,3.88) [thin,neutralcolor] -- (0.44,2.12);
-\draw (3.09,3.88) [thin,neutralcolor] -- (1.76,3.00);
-\draw (3.09,3.88) [thin,neutralcolor] -- (3.09,2.12);
-\draw (1.76,3.00) [thin,neutralcolor] -- (1.76,2.12);
-\draw (0.44,2.12) [thin,neutralcolor] -- (0.44,0.35);
-\draw (0.44,2.12) [thin,neutralcolor] -- (1.76,1.24);
-\draw (3.09,2.12) [thin,neutralcolor] -- (1.76,1.24);
+\draw (0.44,6.53) [thin,neutralcolor] -- (0.44,3.00);
+\draw [fill=neutralcolor,draw=neutralcolor] (0.44,4.77) circle [radius=0.09];
+\draw (1.76,6.53) [thin,neutralcolor] -- (1.76,4.77);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.09];
+\draw (3.09,6.53) [thin,neutralcolor] -- (3.09,3.00);
+\draw [fill=neutralcolor,draw=neutralcolor] (3.09,4.77) circle [radius=0.09];
+\draw (1.76,4.77) [thin,neutralcolor] -- (0.44,3.00);
+\draw (1.76,4.77) [thin,neutralcolor] -- (3.09,3.00);
+\draw (0.44,3.00) [thin,neutralcolor] -- (3.09,1.24);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.76,2.12) circle [radius=0.09];
+\draw (0.44,3.00) [thin,neutralcolor] -- (1.76,1.24);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.10,2.12) circle [radius=0.09];
+\draw (0.44,3.00) [thin,neutralcolor] -- (0.44,0.35);
+\draw [fill=neutralcolor,draw=neutralcolor] (0.44,1.68) circle [radius=0.09];
+\draw (3.09,3.00) [thin,neutralcolor] -- (3.09,1.24);
+\draw (3.09,3.00) [thin,neutralcolor] -- (1.76,1.24);
+\draw [fill=neutralcolor,draw=neutralcolor] (2.43,2.12) circle [radius=0.09];
+\draw (3.09,1.24) [thin,neutralcolor] -- (3.09,0.35);
 \draw (1.76,1.24) [thin,neutralcolor] -- (1.76,0.35);
-\draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,3.00) circle [radius=0.00];
-\node at (1.76,3.00) {};
+\draw [thin,fill=fillcolor,draw=fillcolor] (0.44,6.53) circle [radius=0.35];
+\node at (0.44,6.53) {$t_1$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (3.09,6.53) circle [radius=0.35];
+\node at (3.09,6.53) {$t_2$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,6.53) circle [radius=0.35];
+\node at (1.76,6.53) {$t_3$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,4.77) circle [radius=0.44];
+\node at (1.76,4.77) {$\lor_{\textrm{a}}$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.00) circle [radius=0.44];
+\node at (0.44,3.00) {$\land_{\textrm{v}}$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (3.09,3.00) circle [radius=0.44];
+\node at (3.09,3.00) {$\land_{\textrm{v}}$};
+\draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,1.24) circle [radius=0.00];
+\node at (3.09,1.24) {};
 \draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,1.24) circle [radius=0.00];
 \node at (1.76,1.24) {};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (3.09,3.88) circle [radius=0.44];
-\node at (3.09,3.88) {$\times$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,3.30) circle [radius=0.26];
-\node at (1.32,3.30) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,2.12) circle [radius=0.44];
-\node at (0.44,2.12) {$\times$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.88) circle [radius=0.44];
-\node at (0.44,3.88) {$\times$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,2.12) circle [radius=0.35];
-\node at (1.76,2.12) {$w_1$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (2.21,1.54) circle [radius=0.26];
-\node at (2.21,1.54) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.44];
-\node at (1.76,5.65) {$+$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,7.41) circle [radius=0.35];
-\node at (1.76,7.41) {$\sigma(\phi_2)$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,1.24) circle [radius=0.26];
-\node at (0.44,1.24) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,6.53) circle [radius=0.26];
-\node at (1.76,6.53) {$\sim$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (3.09,0.35) circle [radius=0.35];
+\node at (3.09,0.35) {$x_1$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,0.35) circle [radius=0.35];
-\node at (1.76,0.35) {$w_2$};
+\node at (1.76,0.35) {$x_2$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (0.44,0.35) circle [radius=0.35];
-\node at (0.44,0.35) {$w_3$};
-\draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,2.12) circle [radius=0.00];
-\node at (3.09,2.12) {};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,1.54) circle [radius=0.26];
-\node at (1.32,1.54) {$\sim$};
+\node at (0.44,0.35) {$x_3$};
+\node [right] at (2.21,4.77) {$s_6$};
+\node [right] at (3.53,3.00) {$p_5$};
+\node [left] at (0.00,3.00) {$p_4$};
 \end{tikzpicture}%%

### 2022-07-16 doc/dd/p2-cdcl-eval.tex
Incorporate top-down proof example
Expand example cost functions.  Add operator deletion
Start switch to k-argument And

--- 2022-05-02 doc/dd/p2-bdd-eval.tex
@@
 \begin{tikzpicture}
+\definecolor{fillcolor}{RGB}{255,255,255}
+\definecolor{highcolor}{RGB}{0,0,0}
+\definecolor{lowcolor}{RGB}{0,0,0}
 \definecolor{neutralcolor}{RGB}{0,0,0}
 \definecolor{pathcolor}{RGB}{0,0,0}
-\definecolor{fillcolor}{RGB}{255,255,255}
 \definecolor{background}{RGB}{225,225,225}
-\definecolor{highcolor}{RGB}{0,0,0}
-\definecolor{lowcolor}{RGB}{0,0,0}
 \draw (1.76,7.41) [thin,neutralcolor] -- (1.76,5.65);
@@
 \draw (1.76,5.65) [thin,neutralcolor] -- (3.09,3.88);
-\draw (0.44,3.88) [thin,neutralcolor] -- (1.76,3.00);
+\draw (0.44,3.88) [thin,neutralcolor] -- (3.09,1.24);
 \draw (0.44,3.88) [thin,neutralcolor] -- (0.44,2.12);
-\draw (3.09,3.88) [thin,neutralcolor] -- (1.76,3.00);
-\draw (3.09,3.88) [thin,neutralcolor] -- (3.09,2.12);
-\draw (1.76,3.00) [thin,neutralcolor] -- (1.76,2.12);
+\draw (3.09,3.88) [thin,neutralcolor] -- (3.09,1.24);
+\draw (3.09,3.88) [thin,neutralcolor] -- (1.76,1.24);
+\draw (3.09,1.24) [thin,neutralcolor] -- (3.09,0.35);
 \draw (0.44,2.12) [thin,neutralcolor] -- (0.44,0.35);
 \draw (0.44,2.12) [thin,neutralcolor] -- (1.76,1.24);
-\draw (3.09,2.12) [thin,neutralcolor] -- (1.76,1.24);
 \draw (1.76,1.24) [thin,neutralcolor] -- (1.76,0.35);
-\draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,3.00) circle [radius=0.00];
-\node at (1.76,3.00) {};
+\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,7.41) circle [radius=0.35];
+\node at (1.76,7.41) {$\sigma(\phi_2)$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.44];
+\node at (1.76,5.65) {$+$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.88) circle [radius=0.44];
+\node at (0.44,3.88) {$\times$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (3.09,3.88) circle [radius=0.44];
+\node at (3.09,3.88) {$\times$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,2.12) circle [radius=0.44];
+\node at (0.44,2.12) {$\times$};
+\draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,1.24) circle [radius=0.00];
+\node at (3.09,1.24) {};
 \draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,1.24) circle [radius=0.00];
 \node at (1.76,1.24) {};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (3.09,3.88) circle [radius=0.44];
-\node at (3.09,3.88) {$\times$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,3.30) circle [radius=0.26];
-\node at (1.32,3.30) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,2.12) circle [radius=0.44];
-\node at (0.44,2.12) {$\times$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.88) circle [radius=0.44];
-\node at (0.44,3.88) {$\times$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,2.12) circle [radius=0.35];
-\node at (1.76,2.12) {$w_1$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (2.21,1.54) circle [radius=0.26];
-\node at (2.21,1.54) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.44];
-\node at (1.76,5.65) {$+$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,7.41) circle [radius=0.35];
-\node at (1.76,7.41) {$\sigma(\phi_2)$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,1.24) circle [radius=0.26];
-\node at (0.44,1.24) {$\sim$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,6.53) circle [radius=0.26];
-\node at (1.76,6.53) {$\sim$};
+\draw [thin,fill=fillcolor,draw=fillcolor] (3.09,0.35) circle [radius=0.35];
+\node at (3.09,0.35) {$w_1$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,0.35) circle [radius=0.35];
@@
 \node at (0.44,0.35) {$w_3$};
-\draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,2.12) circle [radius=0.00];
-\node at (3.09,2.12) {};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,2.56) circle [radius=0.26];
+\node at (1.76,2.56) {$\sim$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.00) circle [radius=0.26];
+\node at (0.44,3.00) {$\sim$};
 \draw [thin,fill=fillcolor,draw=neutralcolor] (1.32,1.54) circle [radius=0.26];
 \node at (1.32,1.54) {$\sim$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,1.24) circle [radius=0.26];
+\node at (0.44,1.24) {$\sim$};
 \end{tikzpicture}%%

### 2022-07-16 doc/dd/p2-cdcl.tex
Expand example cost functions.  Add operator deletion
Start switch to k-argument And

@@
 \draw (1.76,5.65) [thin,neutralcolor] -- (3.09,3.88);
-\draw (0.44,3.88) [thin,neutralcolor] -- (1.76,3.00);
-\draw [fill=neutralcolor,draw=neutralcolor] (1.10,3.44) circle [radius=0.09];
+\draw (0.44,3.88) [thin,neutralcolor] -- (3.09,1.24);
+\draw [fill=neutralcolor,draw=neutralcolor] (1.76,2.56) circle [radius=0.09];
 \draw (0.44,3.88) [thin,neutralcolor] -- (0.44,2.12);
 \draw [fill=neutralcolor,draw=neutralcolor] (0.44,3.00) circle [radius=0.09];
-\draw (3.09,3.88) [thin,neutralcolor] -- (1.76,3.00);
-\draw (3.09,3.88) [thin,neutralcolor] -- (3.09,2.12);
-\draw (1.76,3.00) [thin,neutralcolor] -- (1.76,2.12);
+\draw (3.09,3.88) [thin,neutralcolor] -- (3.09,1.24);
+\draw (3.09,3.88) [thin,neutralcolor] -- (1.76,1.24);
+\draw (3.09,1.24) [thin,neutralcolor] -- (3.09,0.35);
 \draw (0.44,2.12) [thin,neutralcolor] -- (0.44,0.35);
@@
 \draw [fill=neutralcolor,draw=neutralcolor] (1.10,1.68) circle [radius=0.09];
-\draw (3.09,2.12) [thin,neutralcolor] -- (1.76,1.24);
 \draw (1.76,1.24) [thin,neutralcolor] -- (1.76,0.35);
-\draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,3.00) circle [radius=0.00];
-\node at (1.76,3.00) {};
-\draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,1.24) circle [radius=0.00];
-\node at (1.76,1.24) {};
+\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,7.41) circle [radius=0.35];
+\node at (1.76,7.41) {$t$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.44];
+\node at (1.76,5.65) {$\lor_{\textrm{a}}$};
+\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.88) circle [radius=0.44];
+\node at (0.44,3.88) {$\land_{\textrm{v}}$};
 \draw [thin,fill=fillcolor,draw=neutralcolor] (3.09,3.88) circle [radius=0.44];
@@
 \node at (0.44,2.12) {$\land_{\textrm{v}}$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (0.44,3.88) circle [radius=0.44];
-\node at (0.44,3.88) {$\land_{\textrm{v}}$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,2.12) circle [radius=0.35];
-\node at (1.76,2.12) {$x_1$};
-\draw [thin,fill=fillcolor,draw=neutralcolor] (1.76,5.65) circle [radius=0.44];
-\node at (1.76,5.65) {$\lor_{\textrm{a}}$};
-\draw [thin,fill=fillcolor,draw=fillcolor] (1.76,7.41) circle [radius=0.35];
-\node at (1.76,7.41) {$t$};
+\draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,1.24) circle [radius=0.00];
+\node at (3.09,1.24) {};
+\draw [thin,fill=neutralcolor,draw=neutralcolor] (1.76,1.24) circle [radius=0.00];
+\node at (1.76,1.24) {};
+\draw [thin,fill=fillcolor,draw=fillcolor] (3.09,0.35) circle [radius=0.35];
+\node at (3.09,0.35) {$x_1$};
 \draw [thin,fill=fillcolor,draw=fillcolor] (1.76,0.35) circle [radius=0.35];
@@
 \node at (0.44,0.35) {$x_3$};
-\draw [thin,fill=neutralcolor,draw=neutralcolor] (3.09,2.12) circle [radius=0.00];
-\node at (3.09,2.12) {};
 \node [right] at (2.21,5.65) {$s_7$};

### 2022-07-16 doc/mc-notes.tex
Start switch to k-argument And

@@
     \midrule
-    $C$    & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
+    $C$    & {\tt p} & $V \; L^{2+}$ {\tt 0}    &                  & Declare $\pand$ operation \\
     $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation.  Explicit hint \\ 
@@
 ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
-such that clause $C$ can only reference clauses $C'$ such that $C' <
-C$.  Clause identifiers need not be consecutive.
+such that clause $C$ can only reference clauses $C'$ such that $C' < C$.
+Clause identifiers need not be consecutive.
 
@@
 \begin{center}
-\begin{tabular}{ccccc}
-  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$l_1$} & \makebox[5mm]{$l_2$} \\
+\begin{tabular}{ccccccccc}
+  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$l_1$} & \makebox[5mm]{$l_2$} &
+  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$l_k$} & \makebox[5mm]{\tt 0} \\
 \end{tabular}
@@
 where $i$ is a new clause ID, $v$ is a positive integer that does not
-correspond to any previous variable, and $l_1$ and $l_2$ are signed
+correspond to any previous variable, and $l_1, l_2, \ldots, l_k$ is a sequence of $k$  ($k \geq 2$)
 integers representing literals of existing variables.  
-This declaration implicitly causes three clauses to be added to the proof:
+This declaration implicitly causes $k+1$ clauses to be added to the proof:
 \begin{center}
-\begin{tabular}{cccc}
-\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
-  $i$ & $v$ & $-l_1$ & $-l_2$ \\
-  $i\!+\!1$ & $-v$ & $l_1$ & \\
-  $i\!+\!2$ & $-v$ & $l_2$ & \\
+\begin{tabular}{cccccc}
+\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
+  $i$ & $v$ & $-l_1$ & $-l_2$ & $\cdots$ & $-l_k$\\
+  $i\!+\!1$ & $-v$ & $l_1$  \\
+  $i\!+\!2$ & $-v$ & $l_2$  \\
+  & $\ldots$ \\
+  $i\!+\!k$ & $-v$ & $l_k$  \\
 \end{tabular}
 \end{center}
-The dependency sets for the arguments represented by literals $l_1$ and $l_2$ must be disjoint.
+The dependency sets for the arguments represented each pair of literals $l_{j_1}$ and $l_{j_2}$, for $1 \leq j_1 < j_2 \leq k$, must be disjoint.
 
@@
     \makebox[5mm][l]{\tt 1} & \makebox[7mm]{\tt i}   & \makebox[20mm][l]{\tt 1 2 3 0}   &  \makebox[30mm]{}          & \makebox[5mm]{} & \makebox[40mm][l]{Input clause}\\
-    {\tt 2}        & {\tt p}   & {\tt 4 -2 3}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
+    {\tt 2}        & {\tt p}   & {\tt 4 -2 3 0}  &            & & Declare $p_4 = \obar{x}_2 \pand x_3$ \\
     {\tt 5}        & {\tt s}   & {\tt 5 2 4}   & {\tt 3 0}  & & Declare $s_5 = x_2 \por p_4$ \\
-    {\tt 8}        & {\tt p}   & {\tt 6 -1 5}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
+    {\tt 8}        & {\tt p}   & {\tt 6 -1 5 0}  &            & & Declare $p_6 = \obar{x}_1 \pand s_5$ \\
     {\tt 11}         & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
@@
     {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
-    {\tt 3} & {\tt p}   & {\tt 4 -2 -3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
-    {\tt 6} & {\tt p}   & {\tt 5 -1 4}  &             & & Declare $p_5 = \obar{x}_1 \pand p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
-    {\tt 9} & {\tt a}  & {\tt -5 0} & {\tt 7 8 5 4 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
+    {\tt 3} & {\tt p}   & {\tt 4 -1 -2 -3 0}  &            & & Declare $p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
+    {\tt 7} & {\tt a}  & {\tt -4 0} & {\tt 7 8 5 4 1 0} & & Assert $t_1 = \obar{p}_4 = x_1 \lor x_2 \lor x_3$ \\
             & {\tt dc}  & {\tt 1 } & {\tt 3 6 9 0} & & Delete clause $C_1$\\
-    {\tt 10} & {\tt p}   & {\tt 6 1 -2} &              & & Declare $p_6 = x_1 \pand \obar{x}_2$ \\    
-    {\tt 13} & {\tt a} & {\tt -6 0} & {\tt 12 11 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
+    {\tt 8} & {\tt p}   & {\tt 5 1 -2} &              & & Declare $p_5 = x_1 \pand \obar{x}_2$ \\    
+    {\tt 11} & {\tt a} & {\tt -5 0} & {\tt 12 11 2 0}  & & Assert $t_2 = \obar{p}_5 = \obar{x}_1 \lor x_2$ \\
             & {\tt dc}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
-    {\tt 14} & {\tt s}   & {\tt 7 5 6}   & {\tt 7 11 0}  & & Declare $s_7 = \obar{t}_1 \por \obar{t}_2$ \\
-    {\tt 17} & {\tt a}  & {\tt -7 0}       & {\tt 9 13 14 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
-             & {\tt dc}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
-             & {\tt dc}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
+    {\tt 12} & {\tt s}   & {\tt 6 4 5}   & {\tt 7 11 0}  & & Declare $s_6 = \obar{t}_1 \por \obar{t}_2$ \\
+    {\tt 15} & {\tt a}  & {\tt -6 0}       & {\tt 9 13 14 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
+             & {\tt dc}  & {\tt 7}          & {\tt 15 17 0} & & Delete $t_1$\\
+             & {\tt dc}  & {\tt 11}         & {\tt 16 17 0} & & Delete $t_2$\\
   \end{tabular}
@@
     {\tt 2} &    & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
-    {\tt 3} &   & {\tt 4 2 3 0} &     & & Defining clauses for $p_4$ \\
-    {\tt 4} &   & {\tt -4 -2 0} &     & & \\
-    {\tt 5} &   & {\tt -4 -3 0} &     & & \\
-
-    {\tt 6} &   & {\tt 5 1 -4 0} &    & & Defining clauses for $p_5$ \\
-    {\tt 7} &   & {\tt -5 -1 0} &     & & \\
-    {\tt 8} &   & {\tt -5 4 0} &      & & \\
-
-    {\tt 9} &   & {\tt -5 0} & {\tt 7 8 5 4 1 0} & & Assert $t_1 = \obar{p}_5 = x_1 \lor x_2 \lor x_3$ \\
-            & {\tt dc}  & {\tt 1 } & {\tt 3 6 9 0} & & Delete clause $C_1$\\
+    {\tt 3} &   & {\tt 4 1 2 3 0} &     & & Defining clauses for $p_4$ \\
+    {\tt 4} &   & {\tt -4 -1 0} &     & & \\
+    {\tt 5} &   & {\tt -4 -2 0} &     & & \\
+    {\tt 6} &   & {\tt -4 -3 0} &     & & \\
+    {\tt 7} &   & {\tt -4 0} & {\tt 6 5 4 1 0} & & Assert $t_1 = \obar{p}_4 = x_1 \lor x_2 \lor x_3$ \\
+            & {\tt dc}  & {\tt 1 } & {\tt 3 7 0} & & Delete clause $C_1$\\
 

### 2022-07-17 doc/mc-notes.tex
Fixed example using three-input And

@@
     {\tt 3} & {\tt p}   & {\tt 4 -1 -2 -3 0}  &            & & Declare $p_4 = \obar{x}_1 \pand \obar{x}_2 \pand \obar{x}_3$ \\
-    {\tt 7} & {\tt a}  & {\tt -4 0} & {\tt 7 8 5 4 1 0} & & Assert $t_1 = \obar{p}_4 = x_1 \lor x_2 \lor x_3$ \\
-            & {\tt dc}  & {\tt 1 } & {\tt 3 6 9 0} & & Delete clause $C_1$\\
+    {\tt 7} & {\tt a}  & {\tt -4 0} & {\tt 6 5 4 1 0} & & Assert $t_1 = \obar{p}_4 = x_1 \lor x_2 \lor x_3$ \\
+            & {\tt dc}  & {\tt 1 } & {\tt 3 7 0} & & Delete clause $C_1$\\
     {\tt 8} & {\tt p}   & {\tt 5 1 -2} &              & & Declare $p_5 = x_1 \pand \obar{x}_2$ \\    
-    {\tt 11} & {\tt a} & {\tt -5 0} & {\tt 12 11 2 0}  & & Assert $t_2 = \obar{p}_5 = \obar{x}_1 \lor x_2$ \\
-            & {\tt dc}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
-    {\tt 12} & {\tt s}   & {\tt 6 4 5}   & {\tt 7 11 0}  & & Declare $s_6 = \obar{t}_1 \por \obar{t}_2$ \\
-    {\tt 15} & {\tt a}  & {\tt -6 0}       & {\tt 9 13 14 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
-             & {\tt dc}  & {\tt 7}          & {\tt 15 17 0} & & Delete $t_1$\\
-             & {\tt dc}  & {\tt 11}         & {\tt 16 17 0} & & Delete $t_2$\\
+    {\tt 11} & {\tt a} & {\tt -5 0} & {\tt 10 9 2 0}  & & Assert $t_2 = \obar{p}_5 = \obar{x}_1 \lor x_2$ \\
+            & {\tt dc}  & {\tt 2 } & {\tt 8 11 0} & & Delete clause $C_2$\\
+    {\tt 12} & {\tt s}   & {\tt 6 4 5}   & {\tt 4 9 0}  & & Declare $s_6 = \obar{t}_1 \por \obar{t}_2$ \\
+    {\tt 15} & {\tt a}  & {\tt -6 0}       & {\tt 7 11 12 0} & & Assert $t_3 = \obar{s}_6 = t_1 \land t_2$ \\
+             & {\tt dc}  & {\tt 7}          & {\tt 13 15 0} & & Delete $t_1$\\
+             & {\tt dc}  & {\tt 11}         & {\tt 14 15 0} & & Delete $t_2$\\
   \end{tabular}
@@
 
-    {\tt 10} &   & {\tt 6 -1 2 0}  &    & & Defining clauses for $p_6$ \\ 
-    {\tt 11} &   & {\tt -6 1 0}    &  & & \\
-    {\tt 12} &   & {\tt -6 -2 0}    &  & & \\ 
-    {\tt 13} &  & {\tt -6 0} & {\tt 12 11 2 0}  & & Assert $t_2 = \obar{p}_6 = \obar{x}_1 \lor x_2$ \\
-            & {\tt dc}  & {\tt 2 } & {\tt 10 13 0} & & Delete clause $C_2$\\
-             &          & {\tt -5 -6 0} & {\tt 7 11 0}  & & Mutual exclusion proof for $s_7$ \\
-    {\tt 14} &   & {\tt -7 5 6 0}  &     & & Defining clauses for $s_7$ \\ 
-    {\tt 15} &   & {\tt  7 -5 0}    &  &  \\  
-    {\tt 16} &   & {\tt  7 -6 0}    &  & &  \\
-    {\tt 17} &   & {\tt -7 0}       & {\tt 9 13 0} & & Assert $t_3 = \obar{s}_7 = t_1 \land t_2$ \\
-             & {\tt dc}  & {\tt 9}          & {\tt 15 17 0} & & Delete $t_1$\\
-             & {\tt dc}  & {\tt 13}         & {\tt 16 17 0} & & Delete $t_2$\\
+    {\tt 8} &   & {\tt 5 -1 2 0}  &    & & Defining clauses for $p_5$ \\ 
+    {\tt 9} &   & {\tt -5 1 0}    &  & & \\
+    {\tt 10} &   & {\tt -5 -2 0}    &  & & \\ 
+    {\tt 11} &  & {\tt -5 0} & {\tt 10 9 2 0}  & & Assert $t_2 = \obar{p}_5 = \obar{x}_1 \lor x_2$ \\
+             & {\tt dc}  & {\tt 2 } & {\tt 8 11 0} & & Delete clause $C_2$\\
+             &          & {\tt -4 -5 0} & {\tt 4 9 0}  & & Mutual exclusion proof for $s_6$ \\
+    {\tt 12} &   & {\tt -6 4 5 0}  &     & & Defining clauses for $s_6$ \\ 
+    {\tt 13} &   & {\tt  6 -4 0}    &  &  \\  
+    {\tt 14} &   & {\tt  6 -5 0}    &  & &  \\
+    {\tt 15} &   & {\tt -6 0}       & {\tt 7 11 12 0} & & Assert $t_3 = \obar{s}_6 = t_1 \land t_2$ \\
+             & {\tt dc}  & {\tt 7}          & {\tt 13 15 0} & & Delete $t_1$\\
+             & {\tt dc}  & {\tt 11}         & {\tt 14 15 0} & & Delete $t_2$\\
   \end{tabular}
@@
 $t_1 = \neg [\obar{x}_1\land \obar{x}_2\land \obar{x}_3]$, and this can be
-expressed using $\pand$ operations $s_4$ and $s_5$, shown in
+expressed using the $\pand$ operation $p_4$, shown in
 Figure~\ref{fig:p2-bdd:schema}A\@.  (Note that $t_1$ in this case is
@@
 Figure~\ref{fig:c3:schema}A, but the use of negation enables it to be
-represented with two operations rather than four.)  Similarly, $C_2$
+represented with one operation rather than four.)  Similarly, $C_2$
 can be written as $t_2 = \neg [x_1\land \obar{x}_2]$, and this can be
-represented by $\pand$ operation $p_6$.
-Terms $t_1$ and $t_2$ are asserted as unit clauses on proof lines 9 and 13,
+represented by $\pand$ operation $p_5$.
+Terms $t_1$ and $t_2$ are asserted as unit clauses on proof lines 7 and 11,
 allowing the input clauses to be deleted.
 The conjunction $t_1 \land t_2$
-can be written as $t_3 = \neg[\obar{t_1} \lor \obar{t_2}]$, represented by $\por$ operation $s_7$.
-Term $t_3$ is asserted as a unit clause on proof line 17.
+can be written as $t_3 = \neg[\obar{t_1} \lor \obar{t_2}]$, represented by $\por$ operation $s_6$.
+Term $t_3$ is asserted as a unit clause on proof line 15.
 Based on this, the unit clauses for terms $t_1$ and

### 2022-07-17 crat/prototype/crat_checker.py
Update to allow k-argument And

@@ -49,7 +49,7 @@ def usage(name):
 # Id  i [Lit*] 0             -- Input clause
 # Id  a [Lit*] 0    HINT 0   -- RUP clause addition
 #    dc Id          HINT 0   -- RUP clause deletion
-# Id  p Var Lit Lit          -- And operation
+# Id  p Var Lit Lit+     0   -- And operation
 # Id  a Var Lit Lit HINT 0   -- Or operation
 #    do Var                  -- Operation deletion
 
@@ -657,7 +657,7 @@ class OperationManager:
     
     # Number of input variables
     inputVariableCount = 0
-    # Operation indexed by output variable.  Each entry of form (op, arg1, arg2, id)
+    # Operation indexed by output variable.  Each entry of form (id, op, arg1, arg2, ...)
     operationDict = {}
     # For each variable, the variables on which it depends
     dependencySetDict = {}
@@ -673,42 +673,48 @@ class OperationManager:
         self.operationDict = {}
         self.dependencySetDict = { v : set([v]) for v in range(1, varCount+1) }
 
-    def addOperation(self, op, outVar, inLit1, inLit2, id):
+    def addOperation(self, op, outVar, inLits, id):
+        if op == self.disjunction:
+            if len(inLits) != 2:
+                return (False, "Cannot have %d arguments for disjunction" % len(inLits))
+        elif op == self.conjunction:
+            if len(inLits) < 2:
+                return (False, "Cannot have %d arguments for conjunction" % len(inLits))
         if outVar in self.dependencySetDict:
             return (False, "Operator output variable %d already in use" % outVar)
-        inVar1 = abs(inLit1)
-        inVar2 = abs(inLit2)
-        if inVar1 not in self.dependencySetDict:
-            return (False, "Operator input literal %d undefined" % inLit1)            
-        if inVar2 not in self.dependencySetDict:
-            return (False, "Operator input literal %d undefined" % inLit2)
-
-        dset1 = self.dependencySetDict[inVar1]
-        dset2 = self.dependencySetDict[inVar2]
-        if op == self.conjunction and not dset1.isdisjoint(dset2):
-            return (False, "Dependency sets of conjunction operation inputs %s and %s are not disjoint" % (inLit1, inLit2))
-        self.dependencySetDict[outVar] = dset1.union(dset2)
+        dset = set([])
+        for lit in inLits:
+            var = abs(lit)
+            if var not in self.dependencySetDict:
+                return (False, "Operator input literal %d undefined" % lit)
[…]

### 2022-07-17 crat/prototype/crat_checker.py
Fixed bugs related to k-argument And

@@ -361,21 +361,24 @@ class CratWriter(Writer):
     def doComment(self, line):
         self.show("c " + line)
         
-    def doAnd(self, i1, i2, xvar = None, id = None):
+    def doAnd(self, args, xvar = None, id = None):
         self.variableCount += 1
         self.stepCount += 1
         v = self.variableCount if xvar is None else xvar
         s = self.stepCount if id is None else id
-        self.doLine([s, 'p', v, i1, i2])
-        self.addClause(s, [v, -i1, -i2])
-        self.addClause(s+1, [-v, i1])        
-        self.addClause(s+2, [-v, i2])
+
+        self.doLine([s, 'p', v] + args + [0])
+        cpos = [v] + [-i for i in args]
+        self.addClause(s, cpos)
         if self.verbLevel >= 2:
             self.doComment("Implicit declarations:")
-            self.doComment("%d a %d %d %d 0" % (s, v, -i1, -i2))
-            self.doComment("%d a %d %d 0" % (s+1, -v, i1))
-            self.doComment("%d a %d %d 0" % (s+2, -v, i2))
-        self.stepCount += 2
+            slist = [str(lit) for lit in cpos]
+            self.doComment("%d a %s 0" % (s, " ".join(slist)))
+        for idx in range(len(args)):
+            self.addClause(s+idx, [-v, args[idx]])
+            if self.verbLevel >= 2:
+                self.doComment("%d a %d %d 0" % (s+1+idx, -v, args[idx]))
+        self.stepCount += len(args)
         return (v, s)
 
     def doOr(self, i1, i2, hints = None, xvar = None, id = None):
@@ -726,7 +729,7 @@ class OperationManager:
         entry = self.operationDict[outVar]
         id = entry[0]
         op = entry[1]
-        args in entry[2:]
+        args = entry[2:]
         for i in range(len(args)+1):
             (ok, msg) = self.cmgr.deleteClause(id+i)
             if not ok:

### 2022-07-17 doc/mc-notes.tex
Fixed typos

@@
             & {\tt dc}  & {\tt 1 } & {\tt 3 7 0} & & Delete clause $C_1$\\
-    {\tt 8} & {\tt p}   & {\tt 5 1 -2} &              & & Declare $p_5 = x_1 \pand \obar{x}_2$ \\    
+    {\tt 8} & {\tt p}   & {\tt 5 1 -2 0} &              & & Declare $p_5 = x_1 \pand \obar{x}_2$ \\    
     {\tt 11} & {\tt a} & {\tt -5 0} & {\tt 10 9 2 0}  & & Assert $t_2 = \obar{p}_5 = \obar{x}_1 \lor x_2$ \\
@@
             {\tt 2} & {\tt i}   & {\tt -1 2 0}  &            & & Input clause $C_2$ \\
-            {\tt 3} & {\tt p}   & {\tt 4 -2 -3}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
-    {\tt 6} & {\tt p}   & {\tt 5 -1 -4}  &             & & Declare $p_5 = \obar{x}_1 \pand \obar{p}_4$ \\
-    {\tt 9} & {\tt p}   & {\tt 6 1 2} &              & & Declare $p_6 = x_1 \pand x_2$ \\    
+            {\tt 3} & {\tt p}   & {\tt 4 -2 -3 0}  &            & & Declare $p_4 = \obar{x}_2 \pand \obar{x}_3$ \\
+    {\tt 6} & {\tt p}   & {\tt 5 -1 -4 0}  &             & & Declare $p_5 = \obar{x}_1 \pand \obar{p}_4$ \\
+    {\tt 9} & {\tt p}   & {\tt 6 1 2 0} &              & & Declare $p_6 = x_1 \pand x_2$ \\    
     {\tt 12} & {\tt s}   & {\tt 7 5 6}   & {\tt 7 10 0}  & & Declare $s_7 = p_5 \por p_6$ \\

### 2022-07-17 doc/mc-notes.tex
Fixed typo

@@
 \end{center}
-The dependency sets for the arguments represented each pair of literals $l_{j_1}$ and $l_{j_2}$, for $1 \leq j_1 < j_2 \leq k$, must be disjoint.
+The dependency sets for the arguments represented by each pair of literals $l_{j_1}$ and $l_{j_2}$, for $1 \leq j_1 < j_2 \leq k$, must be disjoint.
 

### 2022-08-05 doc/mc-notes.tex
Use POG as name of graph representation

@@
 
-For ring elemement $a$, we use the notation $\oneminus a$ to denote $1 - a$.
+For ring element $a$, we use the notation $\oneminus a$ to denote $1 - a$.
 
@@
 
-\section{Separable Schemas}
+\section{Partitioned-Operation Graphs (POGs)}
 
@@
 $\pand$, $\por$, and $\neg$, as is demonstrated by
-Propositions~\ref{prop:negation}--\ref{prop:disjunction}.  We define {\em separable schemas}
+Propositions~\ref{prop:negation}--\ref{prop:disjunction}.  We define a {\em partitioned-operation graph} (POG)
 as a direct-acyclic graph representation of such formulas.
@@
 
-\subsection{Schema Definition}
+\subsection{POG Definition}
 
 \begin{table}
-  \caption{Recursive Definition of Separable Schemas and their Costs}
-  \label{tab:schema}
+  \caption{Recursive Definition of POGs and their Costs}
+  \label{tab:pog}
 %%A) Basic Operations\\
@@
 \toprule
-  \makebox[15mm]{$S$}  & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$} & \makebox[25mm]{$\cost(S)$}\\
+  \makebox[15mm]{$P$}  & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(P)$} & \makebox[30mm]{$\interpset{P}$} & \makebox[25mm]{$\cost(P)$}\\
 \midrule
@@
 $x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ & $\cost(x_i)$ \\
-$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ & $\oneminus\cost(S)$ \\
-$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$
-& $\cost(S_1) \cdot \cost(S_2)$\\
-$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$
-& $\cost(S_1) + \cost(S_2)$\\
+$\neg P_1$      & None  & $\depend(P_1)$ & $\uinterp - \interpset{P_1}$ & $\oneminus\cost(P)$ \\
+$P_1 \pand P_2$ & $\depend(P_1) \cap \depend(P_2) = \emptyset$ & $\depend(P_1) \cup \depend(P_2)$ & $\interpset{P_1} \cap \interpset{P_2}$
+& $\cost(P_1) \cdot \cost(P_2)$\\
+$P_1 \por P_2$  & $\interpset{P_1} \cap \interpset{P_2} = \emptyset$ &  $\depend(P_1) \cup \depend(P_2)$ &  $\interpset{P_1} \cup \interpset{P_2}$
+& $\cost(P_1) + \cost(P_2)$\\
 \bottomrule
@@
 %% \toprule
-%%   $S$           & Definition \\
+%%   $P$           & Definition \\
 %% \midrule
-%% \makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
+%% \makebox[30mm]{$\pite(P_1, P_2, P_3)$} & \makebox[40mm]{$[P_1 \pand P_2] \por [\neg P_1 \pand P_2]$}\\
 %% \bottomrule
@@
 
-The set of schemas over a set of variables
+The set of POGs over a set of variables
 $\{x_1, x_2, \ldots, x_n\}$
 can be defined recursively, as is shown in Table
-\ref{tab:schema}.  Each schema $S$ has an associated dependency set
-$\depend(S)$ and an associated set of models $\interpset{S}$.
+\ref{tab:pog}.  Each POG $P$ has an associated dependency set
+$\depend(P)$ and an associated set of models $\interpset{P}$.
 %% We
@@
 %% from basic operations.  The latter category consists of just the $\pite$ operation.
-A key property of a Boolean formula represented by separable schema
-$S$ is that, for any separable cost function $\cost$, the cost of the
-formula $\cost(S)$ can be computed with a linear number of ring
-operations, as is shown in the final column of Table~\ref{tab:schema}.
+A key property of a Boolean formula represented by POG
+$P$ is that, for any separable cost function $\cost$, the cost of the
+formula $\cost(P)$ can be computed with a linear number of ring
+operations, as is shown in the final column of Table~\ref{tab:pog}.
 
-\subsection{Schema Normalization}
+\subsection{POG Normalization}
 
@@
     \makebox[20mm]{$\neg \tautology$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\nil$} \\
-    $\neg \neg S$ & $\rightarrow$ & $S$ \\
-    $S \pand \nil$ & $\rightarrow$ & $\nil$ && $\nil \pand S$ & $\rightarrow$ & $\nil$ \\
-    $S \pand \tautology$ & $\rightarrow$ & $S$ && $\tautology \pand S$ & $\rightarrow$ & $S$ \\
-    $S \por \nil$ & $\rightarrow$ & $S$ && $\nil \por S$ & $\rightarrow$ & $S$ \\
-    $S \por \tautology$ & $\rightarrow$ & $\tautology$ && $\tautology \por S$ & $\rightarrow$ & $\tautology$ \\
-%%     $\pite(\neg S_1, S_2, S_3)$ & $\rightarrow$ & $\pite(S_1, S_3, S_2)$ \\
-%%     $\pite(\tautology, S_2, S_3)$ & $\rightarrow$ & $S_2$ & &
-%%     $\pite(\nil, S_2, S_3)$ & $\rightarrow$ & $S_3$ \\
-%%     $\pite(S_1, \tautology, \nil)$ & $\rightarrow$ & $S_1$ &&
-%%     $\pite(S_1, \nil, \tautology)$ & $\rightarrow$ & $\neg S_1$ \\
-%%     $\pite(S_1, S_2, \nil)$ & $\rightarrow$ & $S_1 \pand S_2$ & &
-%%     $\pite(S_1, \nil, S_3)$ & $\rightarrow$ & $\neg S_1 \pand S_3$ \\
-%%     $\pite(S_1, S_2, \tautology)$ & $\rightarrow$ & $\neg [S_1 \pand \neg S_2]$ & &
-%%     $\pite(S_1, \tautology, S_3)$ & $\rightarrow$ & $\neg [\neg S_1 \pand \neg S_3]$ \\
+    $\neg \neg P$ & $\rightarrow$ & $P$ \\
+    $P \pand \nil$ & $\rightarrow$ & $\nil$ && $\nil \pand P$ & $\rightarrow$ & $\nil$ \\
+    $P \pand \tautology$ & $\rightarrow$ & $P$ && $\tautology \pand P$ & $\rightarrow$ & $P$ \\
+    $P \por \nil$ & $\rightarrow$ & $P$ && $\nil \por P$ & $\rightarrow$ & $P$ \\
+    $P \por \tautology$ & $\rightarrow$ & $\tautology$ && $\tautology \por P$ & $\rightarrow$ & $\tautology$ \\
+%%     $\pite(\neg P_1, P_2, P_3)$ & $\rightarrow$ & $\pite(P_1, P_3, P_2)$ \\
+%%     $\pite(\tautology, P_2, P_3)$ & $\rightarrow$ & $P_2$ & &
+%%     $\pite(\nil, P_2, P_3)$ & $\rightarrow$ & $P_3$ \\
+%%     $\pite(P_1, \tautology, \nil)$ & $\rightarrow$ & $P_1$ &&
+%%     $\pite(P_1, \nil, \tautology)$ & $\rightarrow$ & $\neg P_1$ \\
+%%     $\pite(P_1, P_2, \nil)$ & $\rightarrow$ & $P_1 \pand P_2$ & &
+%%     $\pite(P_1, \nil, P_3)$ & $\rightarrow$ & $\neg P_1 \pand P_3$ \\
+%%     $\pite(P_1, P_2, \tautology)$ & $\rightarrow$ & $\neg [P_1 \pand \neg P_2]$ & &
+%%     $\pite(P_1, \tautology, P_3)$ & $\rightarrow$ & $\neg [\neg P_1 \pand \neg P_3]$ \\
   \end{tabular}
@@
 Table~\ref{tab:normalize} shows a list of {\em normalizing}
-transformations to simplify a separable schema.  These 
+transformations to simplify a POG\@.  These 
 eliminate extra negations and remove constant
-terms, such that constants only occur in schemas when representing constant functions $\zero$ and $\one$.
+terms, such that constants only occur in POGs when representing constant functions $\zero$ and $\one$.
 %% The most interesting are the two bottom rules, which
@@
 %% transformation when using BDDs to convert a set of clauses into a
-%% separable schema.
+%%  POG\@.
 
@@
 
-The if-then-else (ITE) operation arises when converting the CNF
-representation of a formula into a separable schema, both for
+The if-then-else (ITE) operation arises when converting the CNF representation
+of a formula into a  POG, both for
 bottom-up approaches based on decision diagrams, and for top-down
@@
 $\nil$ for $f_1$.  Therefore, the only condition imposed on an
-expansion of $\ite$ into the allowed schema operations is that the
+expansion of $\ite$ into the allowed POG operations is that the
 dependency set for $f_1$ must be disjoint from those of $f_2$ and
@@
     \midrule
-     $\ite(S_1, S_2, S_3)$ & $$ & $(S_1 \pand S_2) \por (\neg S_1 \land S_3)$ \\
-     $\ite(\tautology, S_2, S_3)$ & $$ & $S_2$ \\
-     $\ite(\nil, S_2, S_3)$ & $$ & $S_3$ \\
-     $\ite(S_1, \tautology, \nil)$ & $$ & $S_1$ \\
-     $\ite(S_1, \nil, \tautology)$ & $$ & $\neg S_1$ \\
-     $\ite(S_1, S_2, \nil)$ & $$ & $S_1 \pand S_2$ \\
-     $\ite(S_1, \nil, S_3)$ & $$ & $\neg S_1 \pand S_3$ \\
-     $\ite(S_1, \tautology, S_3)$ & $$ & $\neg (\neg S_1 \pand \neg S_3)$ \\
-     $\ite(S_1, S_2, \tautology)$ & $$ & $\neg (S_1 \pand \neg S_2)$ \\
+     $\ite(P_1, P_2, P_3)$ & $$ & $(P_1 \pand P_2) \por (\neg P_1 \land P_3)$ \\
+     $\ite(\tautology, P_2, P_3)$ & $$ & $P_2$ \\
+     $\ite(\nil, P_2, P_3)$ & $$ & $P_3$ \\
+     $\ite(P_1, \tautology, \nil)$ & $$ & $P_1$ \\
+     $\ite(P_1, \nil, \tautology)$ & $$ & $\neg P_1$ \\
+     $\ite(P_1, P_2, \nil)$ & $$ & $P_1 \pand P_2$ \\
+     $\ite(P_1, \nil, P_3)$ & $$ & $\neg P_1 \pand P_3$ \\
+     $\ite(P_1, \tautology, P_3)$ & $$ & $\neg (\neg P_1 \pand \neg P_3)$ \\
+     $\ite(P_1, P_2, \tautology)$ & $$ & $\neg (P_1 \pand \neg P_2)$ \\
   \end{tabular}
@@
 Table~\ref{tab:ite} shows different ways to encode an ITE operation
-into schema operations.  All of these require the dependency set of
-the argument $S_1$ to be disjoint from those of arguments $S_2$ and
-$S_3$.  The first row shows the most general case, requiring one
+into POG operations.  All of these require the dependency set of
+the argument $P_1$ to be disjoint from those of arguments $P_2$ and
+$P_3$.  The first row shows the most general case, requiring one
 $\por$ and two $\pand$ operations.  The other rows show special cases,
@@
 checkable proof that a Boolean formula, given in conjunctive normal
-form, is logically equivalent to a separable schema.  Once this
-equivalence has been established, the schema can form the basis
+form, is logically equivalent to a POG\@.  Once this
+equivalence has been established, the POG can form the basis
 for computations enabled by the
@@
 
-Table~\ref{tab:crat:syntax} shows the set of proof rules for the CRAT
-format.
+Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
 As with other clausal proof formats, a variable is
@@
 extension variable or its complement.  That literal will indicate the
-root of the schema as the (possibly negated) output of a schema
-operation.  Working from that root backward, the schema can be
+root of the POG as the (possibly negated) output of a POG
+operation.  Working from that root backward, the POG can be
 extracted from the CRAT file.
@@
     \begin{minipage}{0.4\textwidth}
-      A) Schema   \\[2ex]
+      A) POG   \\[2ex]
     \centering{\input{dd/c3}}
@@
   }
-\caption{Schema \#1 for Formula $\phi_1 = x_1 \lor x_2 \lor x_3$ and its Cost Computation}
-\label{fig:c3:schema}
+\caption{POG \#1 for Formula $\phi_1 = x_1 \lor x_2 \lor x_3$ and its Cost Computation}
+\label{fig:c3:pog}
 \end{figure}
@@
 decompose this formula into a sequence of operations.
-Figure~\ref{fig:c3:schema}A shows one such decomposition.
+Figure~\ref{fig:c3:pog}A shows one such decomposition.
 The subscripts of the variables and the operator labels correspond to
@@
 $w_i$ for each variable $x_i$ for $1 \leq i \leq 3$.  This is
-illustrated in Figure~\ref{fig:c3:schema}B\@.  Note how the Boolean
+illustrated in Figure~\ref{fig:c3:pog}B\@.  Note how the Boolean
 negations become $\oneminus$ operations.  This formula is valid for
@@
     \begin{minipage}{0.4\textwidth}
-      A) Schema   \\[2ex]
+      A) POG   \\[2ex]
     \centering{\input{dd/p2-bdd}}
@@
   }
-  \caption{Schema \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
-\label{fig:p2-bdd:schema}
+  \caption{POG \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
+\label{fig:p2-bdd:pog}
 \end{figure}
@@
 
-Figure~\ref{fig:p2-bdd:schema} shows a schema representing the formula,
+Figure~\ref{fig:p2-bdd:pog} shows a POG representing the formula,
 and Figure~\ref{fig:p2-bdd:crat} shows the associated CRAT file and proof clauses.  This
 proof was generated via a bottom-up strategy, such as would be created
-using BDDs.  It creates schematic representations of the input
+using BDDs.  It creates POG representations of the input
 clauses and then forms their conjunction.  In the proof, unit
@@
 expressed using the $\pand$ operation $p_4$, shown in
-Figure~\ref{fig:p2-bdd:schema}A\@.  (Note that $t_1$ in this case is
-logically equivalent to root $t_1$ in the schema of
-Figure~\ref{fig:c3:schema}A, but the use of negation enables it to be
+Figure~\ref{fig:p2-bdd:pog}A\@.  (Note that $t_1$ in this case is
+logically equivalent to root $t_1$ in the POG of
+Figure~\ref{fig:c3:pog}A, but the use of negation enables it to be
 represented with one operation rather than four.)  Similarly, $C_2$
@@
 $t_2$ can be deleted.  Term $t_3$ then becomes the unique root of the cost
-computation shown in Figure~\ref{fig:p2-bdd:schema}B\@.
+computation shown in Figure~\ref{fig:p2-bdd:pog}B\@.
 
@@
     \begin{minipage}{0.4\textwidth}
-      A) Schema   \\[2ex]
+      A) POG   \\[2ex]
     \centering{\input{dd/p2-cdcl}}
@@
   }
-\caption{Schema \#2B for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
-\label{fig:p2-cdcl:schema}
+\caption{POG \#2B for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
+\label{fig:p2-cdcl:pog}
 \end{figure}
@@
 
-Figure~\ref{fig:p2-cdcl:schema} shows an alternate schema representing
+Figure~\ref{fig:p2-cdcl:pog} shows an alternate POG representing
 the same formula $\phi_2$ as in Example~2A, and
@@
 $x_1$ is assigned $\nil$, and clause $C_1$ becomes $C'_1 = x_2 \lor
-x_3$.  Clause $C'_1$ can be represented schematically as
+x_3$.  Clause $C'_1$ can be represented by POG operation
 $\neg[\obar{x}_2 \pand \obar{x}_3]$, with this operation labeled $p_4$
-in Figure~\ref{fig:p2-cdcl:schema}A\@.  The splitting on variable
+in Figure~\ref{fig:p2-cdcl:pog}A\@.  The splitting on variable
 $x_1$ can be rejoined as $\ite(x_1, x_2, \obar{p}_4)$, and this ITE
@@
 
-The two schemas shown in Figures~\ref{fig:p2-bdd:schema}A and
-\ref{fig:p2-cdcl:schema}A have similar structure.  They have very
+The two POGs shown in Figures~\ref{fig:p2-bdd:pog}A and
+\ref{fig:p2-cdcl:pog}A have similar structure.  They have very
 different negation patterns, but they are logically equivalent.  Their
-associated cost computations (Figures~\ref{fig:p2-bdd:schema}B and
-\ref{fig:p2-cdcl:schema}B) also yield the same results for arbitrary
+associated cost computations (Figures~\ref{fig:p2-bdd:pog}B and
+\ref{fig:p2-cdcl:pog}B) also yield the same results for arbitrary
 weights $w_1$, $w_2$, and $w_3$.
@@
 Given an arbitrary CNF formula, we can use BDD operations to generate
-a schematic representation.  The proof generation can follow the
+a POG representation.  The proof generation can follow the
 methods we have used for generating unsatisfiability proofs of Boolean
@@
 framework.  These choose a splitting variable $x$ and recursively
-construct schemas $S_1$ and $S_0$ for the two assignments to the
-variable.  These are combined as $\ite(x, S_1, S_0)$, using one of the
+construct POGs $P_1$ and $P_0$ for the two assignments to the
+variable.  These are combined as $\ite(x, P_1, P_0)$, using one of the
 encodings of $\ite$ shown in Table~\ref{tab:ite}.  Like CDCL, the
@@
 assignment to the variables, the input clauses decompose into two or
-more sets over disjoint variables.  Then the schema for each of these
+more sets over disjoint variables.  Then the POG for each of these
 partitions can be generated separately, and these are joined via the $\pand$

### 2022-08-05 crat/prototype/crat_checker.py

@@ -21,7 +21,7 @@
 ########################################################################################
 
 
-# Checker for CRAT schema.
+# Checker for CRAT POG.
 import sys
 import getopt
 import datetime

### 2022-08-05 crat/prototype/ddnnf.py

@@ -1,12 +1,12 @@
 #!/usr/bin/python3
 
-# Convert DNNF representation of Boolean formula into a counting schema
+# Convert DNNF representation of Boolean formula into a POG
 
 import sys
 import getopt
 import datetime
 import readwrite
-import schema
+import pog
 
 # Format documentation: http://www.cril.univ-artois.fr/kc/d-DNNF-reasoner.html
 # Input/output format:
@@ -75,7 +75,7 @@ class Node:
     ntype = None
     id = None
     children = []
-    # Corresponding schema node
+    # Corresponding POG node
     snode = None
 
     def __init__(self, ntype, id, children):
@@ -399,8 +399,8 @@ class Nnf:
                 self.nodes.append(node)
         self.topoSort(root)
 
-    def schematize(self, clauseList, fname):
-        sch = schema.Schema(self.inputCount, clauseList, fname, self.verbLevel)
+    def makePog(self, clauseList, fname):
+        sch = pog.Pog(self.inputCount, clauseList, fname, self.verbLevel)
         for node in self.nodes:
             schildren = [child.snode for child in node.children]
             if node.ntype == NodeType.constant:
@@ -421,7 +421,7 @@ class Nnf:
                 # Label for proof generation
                 node.snode.iteVar = node.splitVar
             if self.verbLevel >= 3:
-                print("NNF node %s --> Schema node %s" % (str(node), str(node.snode)))
+                print("NNF node %s --> POG node %s" % (str(node), str(node.snode)))
         sch.compress()
         return sch
                 
@@ -481,12 +481,12 @@ def run(name, args):
     if verbLevel >= 2:
         dag.show()
     if cratName is not None:
-        sch = dag.schematize(creader.clauses, cratName)
+        sch = dag.makePog(creader.clauses, cratName)
         if verbLevel == 1:
-            print("c Generated schema has %d nodes" % len(sch.nodes))
+            print("c Generated POG has %d nodes" % len(sch.nodes))
         if verbLevel >= 2:
             print("")
-            print("c Generated schema has %d nodes:" % len(sch.nodes))
+            print("c Generated POG has %d nodes:" % len(sch.nodes))
             sch.show()
         sch.doValidate()
         sch.finish()

### 2022-08-06 crat/prototype/crat_checker.py
Use arbitrary precision for weighted model counting

@@ -113,100 +113,150 @@ class PNumException(Exception):
         self.msg = msg
 
     def __str__(self):
-        return "P2 Number exception %s" % self.msg
+        return "P52 Number exception %s" % self.msg
 
 
-# Represent numbers of form a * base^b, where a is odd integer and b is integer
-# Default base = 2
-class PNum:
+
+# Represent numbers of form a * 2**m2 + 5**m5
+class P52:
     a = 1
-    b = 0
+    m2 = 0
+    m5 = 0
 
-    def __init__(self, a, b=0, base=2):
+    def __init__(self, a=0, m2=0, m5=0):
         if type(a) != type(1):
             raise PNumException("Nonintegral a (%s)" % str(a))
-        if type(b) != type(1):
-            raise PNumException("Nonintegral b (%s)" % str(b))
-        if type(base) != type(1):
-            raise PNumException("Nonintegral base (%s)" % str(base))
-        if base == 0:
-            raise PNumException("Zero base (%s)" % str(base))
-        self.base = base
+        if type(m2) != type(1):
+            raise PNumException("Nonintegral m2 (%s)" % str(m2))
+        if type(m5) != type(1):
+            raise PNumException("Nonintegral m5 (%s)" % str(m5))
         if a == 0:
-            self.a = 0
-            self.b = 0
+            self.a = a
+            self.m2 = 0
+            self.m5 = 0
             return
-        while a % base == 0:
-            a = a//base
-            b += 1
+        while a % 2 == 0:
+            a = a//2
+            m2 += 1
+        while a % 5 == 0:
+            a = a//5
+            m5 += 1
         self.a = a
-        self.b = b
+        self.m2 = m2
+        self.m5 = m5
 
     def __str__(self):
-        return "%d*%d^(%d)" % (self.a, self.base, self.b)
+        return "%d*2^(%d)5^(%d)" % (self.a, self.m2, self.m5)
 
     def num(self):
-        p = self.base**self.b
-        val = p * self.a
+        p2 = 2**self.m2
+        p5 = 5**self.m5
+        val = self.a * p2 * p5
         return val
 
 
     def neg(self):
-        result = PNum(-self.a, self.b, self.base)
+        result = P52(-self.a, self.m2, self.m5)
         return result
 
     def oneminus(self):
-        one = PNum(1, 0, self.base)
+        one = P52(1)
         result = one.add(self.neg())
         return result
 
     def mul(self, other):
-        if self.base != other.base:
[…]

### 2022-08-06 doc/mc-notes.tex
Update terminology

@@
 operations, as is shown in the final column of Table~\ref{tab:pog}.
+
+POGs are inspired by the DDNNF graphs devised by
+Darwiche for encoding Boolean functions, a process he refers to as
+``knowledge compilation.''
+POGs generalize DDNNF in two ways:
+\begin{itemize}
+\item It allows negation of arbitrary nodes in the graph, not just
+  variables.  It is therefore not a negation-normal form.  We will see
+  examples where this allows a more compact representation.
+\item It allows arbitrary arguments to a $\por$ operation, as long as
+  they have disjoint satisfying assignments.  By contrast, for each
+  disjunction node in a DDNNF, there must be a variable $x$ such that
+  the assignments for one argument have $x=1$ while those for
+  the other argument to have $x=0$.
+\end{itemize}
+  Both of these generalizations allow more flexibility in the POG
+  representation while maintaining the ability to easily compute their
+  cost for separable cost functions.
 

### 2022-08-27 doc/mc-notes.tex
Added additional ITE simplification
Split D4 from DSHARP
Further musings about number systems

@@
 
+{\bf Example 1d}: Restrict the rational numbers of Example~1c to be of
+the form $a\cdot 2^b\cdot 5^c$, where $a$, $b$, and $c$ are integers.
+This class of numbers can exactly represent any value written in
+base-10 decimal notation, and it is closed under addition and
+multiplication.  It can therefore be used to exactly
+encode the weights used in the weighted model counting track of the
+Model Counting Competition, as well as any results that are to be
+generated.  In terms of implementation, both $b$ and $c$ can be encoded
+as fixed-length integers, leaving only $a$ requiring an
+arbitrary-precision representation.  Manipulating numbers with this
+representation could be simpler and more efficient than using full rational
+arithmetic.
+
 {\bf Example 2}: Let $\ring$ be a finite field with $z = |\ring| \geq 2n$,
@@
 $\neg P_1$      & None  & $\depend(P_1)$ & $\uinterp - \interpset{P_1}$ & $\oneminus\cost(P)$ \\
-$P_1 \pand P_2$ & $\depend(P_1) \cap \depend(P_2) = \emptyset$ & $\depend(P_1) \cup \depend(P_2)$ & $\interpset{P_1} \cap \interpset{P_2}$
-& $\cost(P_1) \cdot \cost(P_2)$\\
+$\pand \{ P_1, \ldots, P_k \}$ & $\depend(P_i) \cap \depend(P_j) = \emptyset$, $1\leq i < j \leq k$ & $\bigcup_{1\leq i \leq k} \depend(P_i)$ & $\bigcap_{1\leq i \leq k} \interpset{P_i}$
+& $\prod_{1 \leq i \leq k} \cost(P_i)$\\
 $P_1 \por P_2$  & $\interpset{P_1} \cap \interpset{P_2} = \emptyset$ &  $\depend(P_1) \cup \depend(P_2)$ &  $\interpset{P_1} \cup \interpset{P_2}$
@@
     \midrule
-     $\ite(P_1, P_2, P_3)$ & $$ & $(P_1 \pand P_2) \por (\neg P_1 \land P_3)$ \\
-     $\ite(\tautology, P_2, P_3)$ & $$ & $P_2$ \\
-     $\ite(\nil, P_2, P_3)$ & $$ & $P_3$ \\
-     $\ite(P_1, \tautology, \nil)$ & $$ & $P_1$ \\
-     $\ite(P_1, \nil, \tautology)$ & $$ & $\neg P_1$ \\
-     $\ite(P_1, P_2, \nil)$ & $$ & $P_1 \pand P_2$ \\
-     $\ite(P_1, \nil, P_3)$ & $$ & $\neg P_1 \pand P_3$ \\
-     $\ite(P_1, \tautology, P_3)$ & $$ & $\neg (\neg P_1 \pand \neg P_3)$ \\
-     $\ite(P_1, P_2, \tautology)$ & $$ & $\neg (P_1 \pand \neg P_2)$ \\
+     $\ite(P_1, P_2, P_3)$ & $\;$ & $(P_1 \pand P_2) \por (\neg P_1 \land P_3)$ \\
+     $\ite(\tautology, P_2, P_3)$ & $\;$ & $P_2$ \\
+     $\ite(\nil, P_2, P_3)$ & $\;$ & $P_3$ \\
+     $\ite(P_1, P_2, P_2)$  & $\;$ & $P_2$ \\
+     $\ite(P_1, \tautology, \nil)$ & $\;$ & $P_1$ \\
+     $\ite(P_1, \nil, \tautology)$ & $\;$ & $\neg P_1$ \\
+     $\ite(P_1, P_2, \nil)$ & $\;$ & $P_1 \pand P_2$ \\
+     $\ite(P_1, \nil, P_3)$ & $\;$ & $\neg P_1 \pand P_3$ \\
+     $\ite(P_1, \tautology, P_3)$ & $\;$ & $\neg (\neg P_1 \pand \neg P_3)$ \\
+     $\ite(P_1, P_2, \tautology)$ & $\;$ & $\neg (P_1 \pand \neg P_2)$ \\
+
   \end{tabular}

### 2022-09-15 crat/prototype/crat_checker.py
Improved error message

@@ -693,7 +693,6 @@ class ClauseManager:
         # Should only be one unit clause
         self.root = None
 
-
         for id in sorted(self.clauseDict.keys()):
             if id in self.definingClauseSet:
                 continue
@@ -709,7 +708,7 @@ class ClauseManager:
                 notDeleted.append(id)
 
         if len(notDeleted) > 0:
-            return (False, "Input clauses %s not deleted" % str(notDeleted))
+            return (False, "Clauses %s not deleted" % str(notDeleted))
                 
 
                 

### 2022-09-15 crat/prototype/crat-lrat/convert.sh
hintgen.sh

@@ -14,12 +14,23 @@ cat $CNF  | awk '/ 0/ {print "d "$0;}' > d-lines
 cat $CNF  | awk '/ 0/ {print $0; print "d "$0;}' >> d-lines
 #cat $CNF  | awk '/ 0/ {print "d "$0; print $0; print "d "$0;}' > d-lines
 
-./expand p-lines > $DRAT
-cat a-lines d-lines >> $DRAT
+cp $CNF tmp-$$.cnf
+./expand p-lines >> tmp-$$.cnf
+NVAR=`tail -n 1 tmp-$$.cnf | awk '{if ($1 > 0) print $1; else print $1*-1}'`
+NCLS=`cat tmp-$$.cnf | grep -v "c" | wc | awk '{print $1}'`
 
-./crat-lrat $CNF $DRAT -f | grep -v -e "c " -e "s " > tmp.lrat
+cat tmp-$$.cnf | awk '/p cnf/ {print "p cnf '$NVAR' '$NCLS'"} / 0/ {print $0}' > base.cnf
+rm tmp-$$.cnf
+
+#echo $NVAR" "$NCLS
+
+#cat a-lines d-lines > $DRAT
+cat a-lines > $DRAT
+
+./crat-lrat base.cnf $DRAT -f | grep -v -e "c " -e "s " > tmp.lrat
 
 
-tail -n $BOTH tmp.lrat  | head -n $ASIZE | awk '{printf "%i a", $1; $1=""; print $0}'
+cat tmp.lrat | awk '{printf "%i a", $1; $1=""; print $0}'
+#tail -n $BOTH tmp.lrat  | head -n $ASIZE | awk '{printf "%i a", $1; $1=""; print $0}'
+#tail -n $DSIZE tmp.lrat | sed 's| 0 |:|' | cut -d: -f2- | awk 'BEGIN{i=1} {print "dc "i" "$0; i=i+1}'
 
-tail -n $DSIZE tmp.lrat | sed 's| 0 |:|' | cut -d: -f2- | awk 'BEGIN{i=1} {print "dc "i" "$0; i=i+1}'

### 2022-09-15 crat/prototype/crat-lrat/crat-lrat.c

@@ -1494,3 +1494,4 @@ int main (int argc, char** argv) {
   freeMemory (&S);
   return (sts != UNSAT); // 0 on success, 1 on any failure
 }
+

### 2022-09-15 crat/prototype/crat-lrat/hintgen.sh

@@ -1,0 +1,6 @@
+CNF=$1
+DRAT=$2
+
+cat $DRAT | awk '{print $0"\nd "$0}' > tmp-$$.drat
+./crat-lrat $CNF tmp-$$.drat -f | grep " 0" | grep -v "c "
+rm tmp-$$.drat

### 2022-09-15 crat/prototype/crat-lrat/hintgen.sh
make

@@ -1,6 +1,8 @@
 CNF=$1
 DRAT=$2
 
+make
+
 cat $DRAT | awk '{print $0"\nd "$0}' > tmp-$$.drat
 ./crat-lrat $CNF tmp-$$.drat -f | grep " 0" | grep -v "c "
 rm tmp-$$.drat

### 2022-09-16 crat/prototype/tdpog.py
Improve comments

@@ -1034,13 +1034,13 @@ class Pog:
     # Construct lemma
     def generateLemma(self, root):
         # Set up lemma for this node
-        self.addComment("Setting up Lemma %s" % str(root))
+        self.addComment("Prove Lemma %s" % str(root))
         ncontext = [lit for lit in root.lemma.shadowLiterals if lit != 0] 
         ncontext += list(root.lemma.assignedLiteralSet)
         root.lemma.isLemma = True
         # Prove the lemma
         hints = self.validateUp(root, ncontext, None)
-        self.addComment("Completed proof of lemma at node %s.  Assumed literals = %s" % (str(root), str(ncontext)))
+        self.addComment("Completed proof of Lemma %s.  Assumed literals = %s" % (str(root), str(ncontext)))
         root.lemma.lemmaHints = hints
         self.lemmaCount += 1
 
@@ -1250,7 +1250,7 @@ class Pog:
         hints = self.validateUp(root, [], parent = None)
         topUnitId = root.unitClauseId
         if self.verbLevel >= 2:
-            self.addComment("Delete all but final asserted clauses")
+            self.addComment("Delete all but final asserted clause")
         self.cwriter.doDeleteAssertedClauses()
         if self.verbLevel >= 2:
             self.addComment("Delete input clauses")
@@ -1371,7 +1371,7 @@ class Pog:
     def addLemmas(self):
         root = self.nodes[-1]
         root.lemma = Lemma(self.inputClauseList)
-        self.addComment("Defining (negated) conjunction operations to use as shadow clauses for lemmas")
+        self.addComment("Negated conjunction operations to use as lemma arguments")
         for node in reversed(self.nodes):
             if node.ntype not in [NodeType.conjunction, NodeType.disjunction]:
                 continue
@@ -1412,7 +1412,7 @@ class Pog:
                         print(str(ex))
                         print("Failed when adding lemma from node %s to child %s" % (str(node), str(child)))
                         sys.exit(1)
-        self.addComment("Completed declarations of shadow operations")
+        self.addComment("Operations for representing formula")
 
     def showNode(self, node):
         outs = str(node)

### 2022-09-16 doc/mc-notes.tex
Updated document

@@
      $\ite(\nil, P_2, P_3)$ & $\;$ & $P_3$ \\
-     $\ite(P_1, P_2, P_2)$  & $\;$ & $P_2$ \\
+%     $\ite(P_1, P_2, P_2)$  & $\;$ & $P_2$ \\
      $\ite(P_1, \tautology, \nil)$ & $\;$ & $P_1$ \\
@@
 
-\subsection{TO-DO List}
-
-\begin{itemize}
-\item Proof Framework
-  \begin{itemize}
-  \item Generalities and details of the format
-  \item Can some form of abstraction be incorporated?
-    \begin{itemize}
-    \item Want to abstract a subformula to consider only on its cost and dependency set
-    \item Could represent with fresh extension variable
-    \item But how to prove logical equivalence?
-    \end{itemize}
-  \end{itemize}
-
-\item Checker
-  \begin{itemize}
-  \item Working prototype
-  \item C/C++ (or Rust?)
-  \item Formally verified
-  \end{itemize}
-
-\item Counters
-\begin{itemize}
-\item BDD-based
-  \begin{itemize}
-  \item Prototype
-  \item C/C++
-  \end{itemize}
-
-  \item SDD-based
-  \begin{itemize}
-  \item Bottom-up
-  \item Top-down
-  \end{itemize}
-
-\item Others?
-\end{itemize}
-\end{itemize}
+\section{Exploiting Shared Subgraphs}
+
+\begin{figure}
+\begin{verbatim}
+p cnf 4 5
+-1 3 -4 0
+-1 -3 4 0
+1 3 -4 0
+1 -3 4 0
+-1 -2 0
+\end{verbatim}
+\caption{CNF representation of Formula \#4}
+\label{fig:eg4:cnf}
+\end{figure}
+
+\begin{figure}
+\begin{verbatim}
+c Node P5 = AND(var-3, var-4)
+12 p 5 3 4 0
+c Node P6 = AND(-var-3, -var-4)
+15 p 6 -3 -4 0
+c Node S7 = OR(P5, P6)
+18 s 7 5 6 13 16 0
+c Node P9 = AND(var-1, -var-2, S7)
+21 p 9 1 -2 7 0
+c Node P10 = AND(-var-1, S7)
+25 p 10 -1 7 0
+c Node S11 = OR(P9, P10)
+28 s 11 9 10 22 26 0
+\end{verbatim}
+\caption{Declarations of operations to represent Formula \#4}
+\label{fig:eg4:ops}
+\end{figure}
+
+\begin{figure}
+  \centering{
+    \centering{\input{dd/eg4}}
+  }
+\caption{POG for Formula \#4 yielding shared subgraph $s_7$}
+\label{fig:eg4:pog}
+\end{figure}
+
+\begin{figure}
+\begin{verbatim}
+c Shadow clause [3, -4]
+6 p 12 -3 4 0
+c Implicit declarations:
+c 6 a 12 3 -4 0
+c 7 a -12 -3 0
+c 8 a -12 4 0
+c Shadow clause [-3, 4]
+9 p 13 3 -4 0
+c Implicit declarations:
+c 9 a 13 -3 4 0
+c 10 a -13 3 0
+c 11 a -13 -4 0
+\end{verbatim}
+\caption{Declarations of operations to represent lemma arguments}
+\label{fig:eg4:args}
+\end{figure}
+
+\begin{figure}
+\begin{verbatim}
+c Proof of Lemma S7
+31 a 12 13 -3 4 0 * 0
+32 a 12 13 3 -4 0 * 0
+33 a -3 7 12 13 0 31 12 19 0
+34 a 7 12 13 0 33 32 15 20 0
+c Apply Lemma S7.  Context = [1, -2]
+35 a -1 2 -12 0 7 8 1 0
+36 a -1 2 -13 0 10 11 2 0
+37 a -1 2 7 0 35 36 34 0
+c Apply Lemma S7.  Context = [-1]
+38 a 1 -12 0 7 8 3 0
+39 a 1 -13 0 10 11 4 0
+40 a 1 7 0 38 39 34 0
+c Assert ITE at node S11 (root)
+41 a -1 11 0 5 37 21 29 0
+42 a 11 0 41 40 25 30 0
+\end{verbatim}
+\caption{Assertions leading to unit clause for $t=s_7$}
+\label{fig:eg4:assert}
+\end{figure}
+
+\begin{figure}
+\begin{verbatim}
+c Delete all but final asserted clause
+dc 41 5 37 21 29 0
+dc 40 38 39 34 0
+dc 39 10 11 4 0
+dc 38 7 8 3 0
+dc 37 35 36 34 0
+dc 36 10 11 2 0
+dc 35 7 8 1 0
+dc 34 33 32 15 20 0
+dc 33 31 12 19 0
+dc 32 * 0
+dc 31 * 0
+c Delete input clauses
+dc 1 13 17 18 24 26 28 42 0
+dc 2 14 16 18 24 26 28 42 0
+dc 3 22 13 17 18 27 28 42 0
+dc 4 22 14 16 18 27 28 42 0
+dc 5 23 26 28 42 0
+\end{verbatim}
+\caption{Clause deletions for Formula \#4}
+\label{fig:eg4:delete}
+\end{figure}
 

### 2022-09-16 doc/mc-notes.tex
Minor tweaks

@@
 \begin{verbatim}
-c Node P5 = AND(var-3, var-4)
+c P5 = AND(x-3, x-4)
 12 p 5 3 4 0
-c Node P6 = AND(-var-3, -var-4)
+c P6 = AND(-x-3, -x-4)
 15 p 6 -3 -4 0
-c Node S7 = OR(P5, P6)
+c S7 = OR(P5, P6)
 18 s 7 5 6 13 16 0
-c Node P9 = AND(var-1, -var-2, S7)
+c P9 = AND(x-1, -x-2, S7)
 21 p 9 1 -2 7 0
-c Node P10 = AND(-var-1, S7)
+c P10 = AND(-x-1, S7)
 25 p 10 -1 7 0
-c Node S11 = OR(P9, P10)
+c S11 = OR(P9, P10)
 28 s 11 9 10 22 26 0

### 2022-09-16 crat/prototype/hintify.py
Delete extra clauses at start of DRAT file; Better output info

@@ -44,8 +44,9 @@ prog = path + "/" + "crat-lrat"
 hintStart = 0
 hintList = []
 tmpList = []
-genLrat = True
-deleteTempFiles = False
+# Debugging options
+genLrat = False
+deleteTempFiles = True
 
 # Make up a name for intermediate files
 def getRoot(cnfName):
@@ -85,7 +86,6 @@ def splitFiles(cnfName, cratName, verbLevel):
         return
 
     lineNumber = 0
-    dratClauses = 0    
     getProducts = True
     for line in cratFile:
         line = readwrite.trim(line)
@@ -109,7 +109,7 @@ def splitFiles(cnfName, cratName, verbLevel):
                 ilist = [int(r) for r in rest[:-1]]
                 var = ilist[0]
                 args = ilist[1:]
-                cnfWriter.doProduct(var, args)
+                cnfWriter.doProduct(var, args, dwriter = dratWriter)
             except:
                 print("ERROR: File %s, line #%d.  Couldn't parse product line '%s' in CRAT file" % (cratName, lineNumber, line))                
                 return
@@ -119,7 +119,6 @@ def splitFiles(cnfName, cratName, verbLevel):
                 try:
                     lits = [int(r) for r in rest[:-3]]
                     dratWriter.doStep(lits)
-                    dratClauses += 1
                 except:
                     print("ERROR:  File %s, line #%d. Couldn't generate DRAT step from line %s" % (cratName, lineNumber, line))
                     return
@@ -128,13 +127,13 @@ def splitFiles(cnfName, cratName, verbLevel):
         else:
             break
     cratFile.close()
-    print("Input CNF file %s had %d variables and %d clauses" % (cnfName, cnfReader.nvar, len(cnfReader.clauses)))
+    print("HINTIFY: Input CNF file %s had %d variables and %d clauses" % (cnfName, cnfReader.nvar, len(cnfReader.clauses)))
     cnfWriter.finish()
     vcount = cnfWriter.variableCount()
     vmax = cnfWriter.expectedVariableCount
-    print("Augmented CNF file %s has %d variables (max=%d) and %d clauses" % (acnfName, vcount, vmax, cnfWriter.clauseCount))
+    print("HINTIFY: Augmented CNF file %s has %d variables (max=%d) and %d clauses" % (acnfName, vcount, vmax, cnfWriter.clauseCount))
[…]

### 2022-09-16 crat/prototype/tdpog.py
Enforce full context; better output info

@@ -441,7 +441,7 @@ class Lemma:
                     shadowClause = readwrite.cleanClause(readwrite.invertClause(node.ilist) + [node.xlit])
                     pog.reasoner.addPseudoInput(cid, shadowClause)
                 lit = -node.xlit
-                pog.addComment("Lemma %s, argument #%d: synthetic clause #%d" % (str(root), idx+1, node.definingClauseId))
+                pog.addComment("Lemma %s, argument #%d: shadow clause #%d" % (str(root), idx+1, node.definingClauseId))
             self.shadowLiterals.append(lit)
         self.assignedLiteralSet = externalLiteralSet
 
@@ -639,8 +639,6 @@ class Pog:
     inputClauseList = []
     reasoner = None
     # Options
-    # Should unit-propagated literals be added to context?
-    fullContext = True
     # Should the program attempt to fill out all hints?
     hintLevel = 2
     # What is minimum height for shared subgraph to generate lemma (None--> no lemmas)
@@ -954,7 +952,7 @@ class Pog:
                     root.lemma.assignLiteral(clit)
                 chints = self.validateUnit(clit, ncontext)
                 hints += chints
-                if clit not in ncontext and self.fullContext:
+                if clit not in ncontext:
                     ncontext.append(clit)
         if vcount > 1 or parent is None:
             # Assert extension literal
@@ -1055,7 +1053,7 @@ class Pog:
         for lit in root.lemma.assignedLiteralSet:
             if lit not in context:
                 self.addComment("Lemma %s.  Justify assigned literal %d in context %s" % (str(root), lit, str(context)))
-                chints, self.validateUnit(lit, context)
+                chints = self.validateUnit(lit, context)
                 lhints += chints
         for provenance,isOriginal,clause in root.lemma.argList:
             idx += 1
@@ -1269,29 +1267,29 @@ class Pog:
         if self.verbLevel >= 1:
             nnode = 0
             ndclause = self.definingClauseCounts
-            print("c Nodes by type")
+            print("GEN: Nodes by type")
             for t in range(NodeType.tcount):
                 if self.nodeCounts[t] == 0:
                     continue
-                print("c    %s: %d" % (NodeType.typeName[t], self.nodeCounts[t]))
+                print("GEN:    %s: %d" % (NodeType.typeName[t], self.nodeCounts[t]))
                 nnode += self.nodeCounts[t]
-            print("c    TOTAL: %d" % nnode)
-            print("c Total defining clauses: %d" % ndclause)
[…]

### 2022-09-16 crat/prototype/readwrite.py
Allow CNF operation generation to also delete extra clauses in DRAT file

@@ -323,19 +323,25 @@ class Writer:
 
 
 class DratWriter(Writer):
+    additions = 0
+    deletions = 0
     
     def __init__(self, fname):
         Writer.__init__(self, 0, fname)
+        self.additions = 0
+        self.deletions = 0
 
     def doStep(self, lits):
         slits = [str(lit) for lit in lits] + ['0']
         line = " ".join(slits)
         self.show(line)
+        self.additions += 1
 
     def doDelete(self, lits):
         slits = ['d'] + [str(lit) for lit in lits] + ['0']
         line = " ".join(slits)
         self.show(line)
+        self.deletions += 1
 
 # Creating CNF
 class CnfWriter(Writer):
@@ -385,14 +391,17 @@ class CnfWriter(Writer):
 # Version that allows adding clauses for product operators
 class AugmentedCnfWriter(CnfWriter):
     
-    def doProduct(self, var, args):
+    # Optionally have the DRAT file delete all but the first clause
+    def doProduct(self, var, args, dwriter = None):
         self.expectedVariableCount = max(self.expectedVariableCount, var)
         lits = [var] + [-arg for arg in args]
-        self.doClause(lits)
+        id = self.doClause(lits)
         for arg in args:
             lits = [-var, arg]
             self.doClause(lits)
-    
+            if dwriter is not None:
+                dwriter.doDelete(lits)
+        return id
 
 # Enable permuting of variables before emitting CNF
 class Permuter:
@@ -670,6 +679,7 @@ upperStepStart = 100 * 1000 * 1000
 
 class CratWriter(SplitWriter):
     variableCount = 0
+    usedVariableSet = set([])
     clauseDict = []
     lowerStepCount = 0
     upperStepCount = upperStepStart
@@ -680,6 +690,7 @@ class CratWriter(SplitWriter):
     def __init__(self, variableCount, clauseList, fname, verbLevel = 1):
         Writer.__init__(self, variableCount, fname, verbLevel=verbLevel, isNull=False)
         self.variableCount = variableCount
+        self.usedVariableSet = set([])
         self.lowerStepCount = len(clauseList)
         self.upperStepCount = upperStepStart
         self.clauseDict = {}
@@ -704,6 +715,8 @@ class CratWriter(SplitWriter):
         return cid+1
 
     def addClause(self, step, lits):
+        for lit in lits:
+            self.usedVariableSet.add(abs(lit))
         self.clauseDict[step] = lits
 
[…]

### 2022-09-16 crat/prototype/crat_checker.py
Better output information

@@ -276,7 +276,7 @@ class CnfReader():
             self.fail("Could not open file '%s'" % fname)
             return
         self.readCnf()
-        print("Read %d clauses from file %s" % (len(self.clauses), fname))
+        print("CHECKER: Read %d clauses from file %s" % (len(self.clauses), fname))
         self.file.close()
         
     def fail(self, msg):
@@ -357,7 +357,7 @@ class Writer:
         try:
             self.outfile = open(fname, 'w')
         except:
-            print("Couldn't open file '%s'. Aborting" % fname)
+            print("CHECKER: Couldn't open file '%s'. Aborting" % fname)
             sys.exit(1)
 
     def vcount(self):
@@ -470,7 +470,7 @@ class CratWriter(Writer):
         
         
     def finish(self):
-        print("c File '%s' has %d variables and %d steps" % (self.fname, self.variableCount, self.stepCount))
+        print("CHECKER: c File '%s' has %d variables and %d steps" % (self.fname, self.variableCount, self.stepCount))
         Writer.finish(self)
 
 
@@ -653,11 +653,11 @@ class ClauseManager:
                             usedIdSet.add(gen)
             hints.reverse()
             if self.verbose:
-                print("RUP finder: Target = %s.  Hints = %s" % (str(tclause), str(hints)))
+                print("CHECKER: RUP finder: Target = %s.  Hints = %s" % (str(tclause), str(hints)))
             return hints
         else:
             if self.verbose:
-                print("RUP finder failed: Target = %s.  Units = %s" % (str(tclause), str(list(unitSet))))
+                print("CHECKER: RUP finder failed: Target = %s.  Units = %s" % (str(tclause), str(list(unitSet))))
             return None
 
     # Check that clause is generated by set of antecedents
@@ -885,7 +885,7 @@ class Prover:
                 break
 
     def flagError(self, msg):
-        print("ERROR.  Line %d: %s" % (self.lineNumber, msg))
+        print("CHECKER: ERROR.  Line %d: %s" % (self.lineNumber, msg))
         self.failed = True
 
     # Find zero-terminated list of integers in fields (or single field consisting of '*').  Return (list, rest)
@@ -965,7 +965,7 @@ class Prover:
     def count(self, weights = None):
         root = self.cmgr.root
         if root is None:
-            print("Can't determine count.  Don't know root")
+            print("CHECKER: Can't determine count.  Don't know root")
[…]

### 2022-09-16 crat/prototype/ddnnf.py

@@ -812,32 +812,32 @@ def run(name, args):
     elif not dag.read(nfile):
         print("Read failed")
         return
-    elif verbLevel >= 1: 
-       print("c Input NNF DAG has %d inputs, %d nodes" % (dag.inputCount, dag.nodeCount()))
+    if verbLevel >= 1: 
+       print("GEN: Input NNF DAG has %d inputs, %d nodes" % (dag.inputCount, dag.nodeCount()))
     if verbLevel >= 3:
         dag.show()
     if verbLevel >= 2:
         print("")
-        print("c ITE extraction:")
+        print("GEN: ITE extraction:")
     dag.findIte()
     if verbLevel >= 1:
-        print("c NNF DAG with ITEs has %d nodes" % (dag.nodeCount()))
+        print("GEN: NNF DAG with ITEs has %d nodes" % (dag.nodeCount()))
     if verbLevel >= 3:
         dag.show()
     if cratName is not None:
         pg = dag.makePog(creader.clauses, cratName, hintLevel, lemmaHeight, splitProof)
         fcount = pg.nodeCounts[tdpog.NodeType.conjunction] + pg.nodeCounts[tdpog.NodeType.disjunction]
         if verbLevel == 1:
-            print("c Generated POG has %d And/Or nodes" % fcount)
+            print("GEN: Generated POG has %d And/Or nodes" % fcount)
         if verbLevel >= 3:
             print("")
-            print("c Generated POG has %d And/Or nodes:" % fcount)
+            print("GEN: Generated POG has %d And/Or nodes:" % fcount)
             pg.show()
         pg.doValidate()
         pg.finish()
     delta = datetime.datetime.now() - start
     seconds = delta.seconds + 1e-6 * delta.microseconds
-    print("Elapsed time for generation: %.2f seconds" % seconds)
+    print("GEN: Elapsed time for generation: %.2f seconds" % seconds)
 
 if __name__ == "__main__":
     run(sys.argv[0], sys.argv[1:])

### 2022-11-09 crat/checker/crat-check.c
Provide commandline options

@@ -50,6 +50,15 @@ bool early_rup = true;
 char *current_file = "";
 int line_count = 0;
 
+void usage(char *name) {
+    printf("Usage: %s [-h] [-v VERB] FILE.cnf [FILE.crat]\n", name);
+    printf(" -h VERB      Print this message\n");
+    printf(" -v           Set verbosity level\n");
+    printf("    FILE.cnf  Input CNF file\n");
+    printf("    FILE.crat Input CRAT file\n");
+    exit(0);
+}
+
 /*============================================
   Utility functions
 ============================================*/
@@ -1052,13 +1061,12 @@ void crat_add_clause(int cid) {
 	}
 
 	int lit = token_value;
+	clause_add_literal(lit);
 
 	if (lit == 0)
 	    break;
-	else {
-	    clause_add_literal(lit);
+	else
 	    lset_add_lit(-lit);
-	}
     }
     rup_run();
 
@@ -1360,17 +1368,13 @@ void crat_read(char *fname) {
     }
 }
 
-int main(int argc, char *argv[]) {
-    if (argc <= 1) {
-	fprintf(ERROUT, "Usage: %s FILE.cnf [FILE.crat]\n", argv[0]);
-	return 1;
-    }
+void run(char *cnf_name, char *crat_name) {
     double start = tod();
-    cnf_read(argv[1]);
+    cnf_read(cnf_name);
     if (verb_level >= 3)
 	cnf_show(stdout);
-    if (argc >= 3) {
-	crat_read(argv[2]);
+    if (crat_name != NULL) {
+	crat_read(crat_name);
 	if (verb_level >= 3) {
 	    crat_show(stdout);
 	    printf("All clauses:\n");
@@ -1380,10 +1384,44 @@ int main(int argc, char *argv[]) {
 	if (verb_level >= 1) {
 	    printf("CHECK: Final root literal %d\n", root);
 	}
+	printf("CHECK: SUCCESS.  CRAT representation verified\n");
     }
     if (verb_level >= 1) {
 	double secs = tod() - start;
 	printf("CHECK: Elapsed seconds: %.3f\n", secs);
     }
+}
+
+int main(int argc, char *argv[]) {
+    char *cnf_name = NULL;
+    char *crat_name = NULL;
+    verb_level = 1;
+    if (argc <= 1) 
+	usage(argv[0]);
+    int argi;
+    char *istring;
[…]

### 2022-11-09 crat/checker/crat-check.c
Better output info

@@ -938,7 +938,7 @@ void cnf_read(char *fname) {
     }
     token_finish();
     if (verb_level >= 1) {
-	printf("CHECK: Read CNF file with %d variables and %d clauses\n", input_variable_count, input_clause_count);
+	printf("FCHECK: Read CNF file with %d variables and %d clauses\n", input_variable_count, input_clause_count);
     }
 }
 
@@ -1361,10 +1361,10 @@ void crat_read(char *fname) {
     token_finish();
     if (verb_level >= 1) {
 	int all_clause_count = crat_assertion_count + crat_operation_clause_count;
-	printf("CHECK: Read CRAT file with %d operation,  %d+%d=%d clauses\n",
+	printf("FCHECK: Read CRAT file with %d operation,  %d+%d=%d clauses\n",
 	       crat_operation_count, crat_assertion_count,
 	       crat_operation_clause_count, all_clause_count);
-	printf("CHECK: Deleted %d input and asserted clauses\n", crat_assertion_deletion_count);
+	printf("FCHECK: Deleted %d input and asserted clauses\n", crat_assertion_deletion_count);
     }
 }
 
@@ -1382,13 +1382,13 @@ void run(char *cnf_name, char *crat_name) {
 	}
 	int root = crat_final_root();
 	if (verb_level >= 1) {
-	    printf("CHECK: Final root literal %d\n", root);
+	    printf("FCHECK: Final root literal %d\n", root);
 	}
-	printf("CHECK: SUCCESS.  CRAT representation verified\n");
+	printf("FCHECK: SUCCESS.  CRAT representation verified\n");
     }
     if (verb_level >= 1) {
 	double secs = tod() - start;
-	printf("CHECK: Elapsed seconds: %.3f\n", secs);
+	printf("FCHECK: Elapsed seconds: %.3f\n", secs);
     }
 }
 

### 2022-11-09 doc/mc-notes.tex
Clarify requirements for final status

@@
   Deleting clauses requires proving that the resulting formula is not weaker.
+  That is, it should not allow any additional satisfying assignment.
   \begin{itemize}
@@
     \midrule
-    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
-    $C$ & {\tt a} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RUP clause.  Explicit hint \\
+    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause.  Explicit hint} \\
     $C$ & {\tt a} & $L^{*}$ {\tt 0} &  {\tt * 0} & Add RUP clause.  Inferred hint \\
@@
 Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
+The checker is provided with the input formula as a separate file.
 As with other clausal proof formats, a variable is
@@
 The first set of proof rules are similar to those in other clausal
-proofs.  Our syntax optionally allows input clauses to be listed with
-a rule of type {\tt i}.  Clauses can be added via RUP addition
+proofs.
+Clauses can be added via RUP addition
 (command {\tt a}), with the sequence of antecedent clauses (the
@@
 both add and delete clauses.  Each addition must be truth preserving
-and each deletion must be falsehood preserving.  At the end, all input
-clauses must have been deleted, and among the remaining clauses there
-must be only a single unit clause consisting of some variable or its
+and each deletion must be falsehood preserving.  At the end, the following conditions must hold:
+\begin{itemize}
+\item All of the input clauses should have been deleted.
+\item There should be exactly one clause left that was added via RUP addition, and this should be a unit clause.
+\end{itemize}
+The remaining
+unit clause consists of some variable or its
 complement.  Except for trivial cases, the final literal will be an

### 2022-11-09 crat/checker/crat-check.c
Track blocks

@@ -594,6 +594,7 @@ typedef struct BELE {
 
 clause_block_t *clause_set = NULL;
 clause_block_t *clause_set_last = NULL;
+int clause_block_count = 0;
 
 void clause_error(char *msg) {
     fprintf(ERROUT, "ERROR: File %s, Line %d.  clause error in function %s\n", current_file, line_count+1, msg);
@@ -609,6 +610,7 @@ void clause_init() {
 	clause_error("clause_init");
     }
     clause_set = malloc(sizeof(clause_block_t));
+    clause_block_count ++;
     if (clause_set == NULL) {
 	fprintf(ERROUT, "Couldn't allocate space for clause block\n");
 	clause_error("clause_init");
@@ -666,6 +668,7 @@ void clause_new(int cid) {
     if (cid > clause_last_id + MAX_GAP) {
 	/* Need to start new block */
 	clause_block_t *nlast = malloc(sizeof(clause_block_t));
+	clause_block_count ++;
 	if (nlast == NULL) {
 	    fprintf(ERROUT, "Couldn't allocate space for clause block\n");
 	    clause_error("clause_new");
@@ -1364,6 +1367,7 @@ void crat_read(char *fname) {
 	printf("FCHECK: Read CRAT file with %d operation,  %d+%d=%d clauses\n",
 	       crat_operation_count, crat_assertion_count,
 	       crat_operation_clause_count, all_clause_count);
+	printf("FCHECK: Clauses divided into %d blocks\n", clause_block_count);
 	printf("FCHECK: Deleted %d input and asserted clauses\n", crat_assertion_deletion_count);
     }
 }

### 2023-02-19 doc/mc-data.tex
Data visualization

\documentclass{llncs}
\usepackage{latexsym}
\usepackage{times}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{cite}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{marvosym}
%% Colored hyperlink 
\newcommand{\cref}[2]{\href{#1}{\color{blue}#2}}
%% Colored hyperlink showing link in TT font
% \newcommand{\chref}[1]{\href{#1}{\small\tt \color{blue}#1}}
\newcommand{\hcref}[1]{\cref{#1}{\small\tt #1}}

\bibliographystyle{splncs04}
%\bibliographystyle{abbrv}

\definecolor{redorange}{rgb}{0.878431, 0.235294, 0.192157}
\definecolor{lightblue}{rgb}{0.552941, 0.72549, 0.792157}
\definecolor{clearyellow}{rgb}{0.964706, 0.745098, 0}
\definecolor{clearorange}{rgb}{0.917647, 0.462745, 0}
\definecolor{mildgray}{rgb}{0.54902, 0.509804, 0.47451}
\definecolor{softblue}{rgb}{0.643137, 0.858824, 0.909804}
\definecolor{bluegray}{rgb}{0.141176, 0.313725, 0.603922}
\definecolor{lightgreen}{rgb}{0.709804, 0.741176, 0}
\definecolor{darkgreen}{rgb}{0.152941, 0.776471, 0.172549}
\definecolor{redpurple}{rgb}{0.835294, 0, 0.196078}
\definecolor{midblue}{rgb}{0, 0.592157, 0.662745}
\definecolor{clearpurple}{rgb}{0.67451, 0.0784314, 0.352941}
\definecolor{browngreen}{rgb}{0.333333, 0.313725, 0.145098}
\definecolor{darkestpurple}{rgb}{0.396078, 0.113725, 0.196078}
\definecolor{greypurple}{rgb}{0.294118, 0.219608, 0.298039}
\definecolor{darkturquoise}{rgb}{0, 0.239216, 0.298039}
\definecolor{darkbrown}{rgb}{0.305882, 0.211765, 0.160784}
\definecolor{midgreen}{rgb}{0.560784, 0.6, 0.243137}
\definecolor{darkred}{rgb}{0.576471, 0.152941, 0.172549}
\definecolor{darkpurple}{rgb}{0.313725, 0.027451, 0.470588}
\definecolor{darkestblue}{rgb}{0, 0.156863, 0.333333}
\definecolor{lightpurple}{rgb}{0.776471, 0.690196, 0.737255}
\definecolor{softgreen}{rgb}{0.733333, 0.772549, 0.572549}
\definecolor{offwhite}{rgb}{0.839216, 0.823529, 0.768627}

\title{Validated Model Counting Data \\ Version of \today}

\author{Randal E. Bryant}

\authorrunning{R. E. Bryant}

\titlerunning{Validated Model Counting}

\institute{
Computer Science Department \\
Carnegie Mellon University, Pittsburgh, PA, United States
}

\begin{document}

\maketitle

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
                              x post scale=2.0, y post scale=2.0,
                              xmin=0,xmax=900,
                              xtick = {0,100,200,300,400,500,600,700,800,900},
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={D4 runtime}, ylabel={Defining Clauses},
%                              title={D4 Defining Clause Generation}
            ]
    \input{data-formatted/time+defining}
          \end{axis}
\end{tikzpicture}
} % centering
\caption{Number of defining clauses in generated POG as function of D4 runtime}
\label{fig:generate:d4}
\end{figure}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
                              x post scale=2.0, y post scale=2.0,
                              xmode=log, xmin=001,xmax=1000,
                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, 1000},
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={D4 runtime}, ylabel={Defining Clauses},
%                              title={D4 Defining Clause Generation}
            ]
    \input{data-formatted/time+defining}
          \end{axis}
\end{tikzpicture}
} % centering
\caption{Number of defining clauses in generated POG as function of D4 runtime}
\label{fig:generate:d4:log}
\end{figure}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
                              x post scale=2.0, y post scale=2.0,
                              xmode=log,xmin=100,xmax=1000000000, 
                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=darkestblue] coordinates{(100,100) (1e9,1e9)};
    \addplot[mark=none, color=midblue] coordinates{(100,1000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
    \input{data-formatted/defining+total}
    \node[left] at (axis cs: 900,1200) {$1{}.0\times$};
    \node[left] at (axis cs: 900,12000) {$10\times$};
    \node[left] at (axis cs: 900,120000) {$100\times$};

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Total number of clauses in CRAT file as function of number of defining clauses}
\label{fig:defining:total}

\end{figure}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 1.00]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=10000,
                              x post scale=2.0, y post scale=2.2,
                              xmode=log,xmin=0.01,xmax=1000,
                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, 1000},
                              ytick={0.01, 0.1,1.0,10,100,1000,10000}, yticklabels={0.01, 0.1, 1.0, 10, 100, 1000,10000},
                              xlabel={D4 runtime}, ylabel={CRAT generation runtime},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=darkestblue] coordinates{(0.01,0.01) (1000.0,1000.0)};
    \addplot[mark=none, color=midblue] coordinates{(0.01,0.1) (1000,10000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (100,10000)};
    \input{data-formatted/time-d4-gen}
    \node[left] at (axis cs: 0.09,0.12) {$1{}.0\times$};
    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
    \node[left] at (axis cs: 0.09,12.0) {$100\times$};

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Runtime for CRAT proof generation as function of D4 runtime}
\label{fig:d4:crat}

\end{figure}

\bibliography{references}

\end{document}

### 2023-02-19 tools/gen-data.py

@@ -1,0 +1,113 @@
+#!/usr/bin/python3
+import sys
+import getopt
+
+# Input is CSV file with entries x,y.
+# 05/19/2021 ... Allow entries of the form x,y1,y2,...,yN
+# Default is to use column N as value of y, but this can be changed as an option
+
+# Output is line like the following
+#          \addplot coordinates {(4, 900) (8, 8320) (9, 39628) (11, 252970) (12, 1324539) (13, 4095084) (14, 7131764) (15, 15225960)};
+
+# Threshold value.  Don't include data with Y values exceeding this value
+yThresh = 1000 * 1000 * 1000
+xThresh = 1000 * 1000 * 1000
+# Set any y value below yMin to yMin
+yMin = 0.0
+xColumn = -1
+yColumn = -1
+optionString = ""
+
+def usage(name):
+    print("Usage: %s [-h] [-x XTHRESH] [-y YTHRESH] < file.csv > file.tex")
+    print(" -h         Print this message")
+    print(" -X XCOL    Set column number to use as X value (counting from 1)")
+    print(" -x XTHRESH Set maximum X value included")
+    print(" -Y YCOL    Set column number to use as Y value (counting from 1)")
+    print(" -y YTHRESH Set maximum Y value included")
+    print(" -L YMIN    Set minimum Y value included")
+    print(" -O OSTRING Specify addplot options (usually quoted string)")
+
+def trim(s):
+    while len(s) > 0 and s[-1] in '\r\n':
+        s = s[:-1]
+    return s
+
+def gen(infile, outfile):
+    outfile.write("\\addplot %s coordinates {" % optionString)
+    for line in infile:
+        line = trim(line)
+        fields = line.split(",")
+        if len(fields) >= max(xColumn,yColumn):
+            try:
+                fx = float(fields[xColumn-1])
+                if fx > float(xThresh):
+                    sys.stderr.write("%f exceeds x threshold %d\n" % (fx, xThresh))
+                    continue
+            except:
+                pass
+            try:
+                fy = float(fields[yColumn-1])
+                if fy < yMin or fy > float(yThresh):
+                    continue
+            except:
+                pass
+            outfile.write(" (%s,%s)" % (fields[xColumn-1], fields[yColumn-1]))
+    outfile.write("};\n")
+        
+def run(name, args):
+    global xThresh, xColumn, yThresh, yColumn, yMin, optionString
+    optlist, args = getopt.getopt(args, "hx:X:y:Y:L:O:")
[…]

### 2023-02-19 tools/merge_csv.py

@@ -2,7 +2,10 @@
 
 ## Merge set of CSV file into single file with headings
 ## Each source file must have lines of form key,value
-## The files must all have the same keys in the same order
+## The files must all have the keys in the same order
+## The keys for the first entry must be a superset of the rest.
+
+## Optionally filter out keys for which one more fields is empty
 
 import sys
 import getopt
@@ -13,7 +16,8 @@ def eprint(s):
 
 
 def usage(name):
-    eprint("Usage: %s [-h] [-l L0,L1,L2,...,Ln] FILE1.csv FILE2.csv ... FILEn.csv" % name)
+    eprint("Usage: %s [-h] [-f] [-l L0,L1,L2,...,Ln] FILE1.csv FILE2.csv ... FILEn.csv" % name)
+    eprint("  -f            Filter out lines that have at least one field missing")
     eprint("  -h            Print this message")
     eprint("  -l LABELS     Provide comma-separated set of heading labels")
     eprint("  FILE1.csv ... Source files")
@@ -22,9 +26,11 @@ def usage(name):
 keys = []
 # Growing list of result lines
 lines = []
+# Set of keys with missing entries
+missingKeys = set([])
 
 def addData(fname, first = False):
-    global keys, lines
+    global keys, lines, missingKeys
     try:
         infile = open(fname)
         creader = csv.reader(infile)
@@ -46,40 +52,62 @@ def addData(fname, first = False):
             if row > len(keys):
                 eprint("File %s, row %d.  Too many entries" % (fname, row))
                 sys.exit(1)
-            if keys[row-1] != key:
-                eprint("File %s, row %d.  Expecting key '%s'.  Got '%s'." % (fname, row, keys[row-1], key))
-                sys.exit(1)
+            while row <= len(keys) and keys[row-1] != key:
+                missingKeys.add(keys[row-1])
+                lines[row-1] += ","
+                row += 1
+            if row > len(keys):
+                eprint("File %s, row %d.  Couldn't find key '%s'" % (fname, row, key))
+                sys.exit(1)                
             lines[row-1] += "," + val
-    while row < len(lines):
-        lines[row] += ","
+    while row <= len(lines):
+        lines[row-1] += ","
+        missingKeys.add(keys[row-1])
         row += 1
     infile.close()
         
-def build(lstring, flist):
+
+def filter():
+    global keys, lines
+    okeys = keys
+    olines = lines
+    keys = []
+    lines = []
+    for (key,line) in zip(okeys, olines):
[…]

### 2023-02-19 doc/mc-data.tex
Improve data gathering

@@
                               x post scale=2.0, y post scale=2.0,
-                              xmin=0,xmax=900,
-                              xtick = {0,100,200,300,400,500,600,700,800,900},
+                              xmin=0,xmax=1000,
+                              xtick = {0,100,200,300,400,500,600,700,800,900,1000},
                               ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              xlabel={D4 runtime}, ylabel={Defining Clauses},
+                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
 %                              title={D4 Defining Clause Generation}
             ]
-    \input{data-formatted/time+defining}
+    \input{data-formatted/time+defining-success}
+    \input{data-formatted/time+defining-failure}
           \end{axis}
@@
                               xmode=log, xmin=001,xmax=1000,
-                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, 1000},
+                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
                               ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              xlabel={D4 runtime}, ylabel={Defining Clauses},
+                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
 %                              title={D4 Defining Clause Generation}
             ]
-    \input{data-formatted/time+defining}
+    \input{data-formatted/time+defining-success}
+    \input{data-formatted/time+defining-failure}
           \end{axis}
@@
 \begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=10000,
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=20000,
                               x post scale=2.0, y post scale=2.2,
                               xmode=log,xmin=0.01,xmax=1000,
-                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, 1000},
-                              ytick={0.01, 0.1,1.0,10,100,1000,10000}, yticklabels={0.01, 0.1, 1.0, 10, 100, 1000,10000},
-                              xlabel={D4 runtime}, ylabel={CRAT generation runtime},
+                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
+                              ytick={0.01, 0.1,1.0,10,100,1000,10000,20000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{20,000}},
+                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
 %                              title={D4 Defining Clause Generation}
@@
     \addplot[mark=none, color=midblue] coordinates{(0.01,0.1) (1000,10000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (100,10000)};
-    \input{data-formatted/time-d4-gen}
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (200,20000)};
+    \input{data-formatted/time-d4-verify}
     \node[left] at (axis cs: 0.09,0.12) {$1{}.0\times$};
@@
 } % centering
-\caption{Runtime for CRAT proof generation as function of D4 runtime}
+\caption{Combined runtime for CRAT proof generation and checking  as function of D4 runtime}
 \label{fig:d4:crat}

### 2023-02-19 tools/merge_csv.py

@@ -16,8 +16,9 @@ def eprint(s):
 
 
 def usage(name):
-    eprint("Usage: %s [-h] [-f] [-l L0,L1,L2,...,Ln] FILE1.csv FILE2.csv ... FILEn.csv" % name)
+    eprint("Usage: %s [-h] [-f]  [-s] [-l L0,L1,L2,...] FILE1.csv FILE2.csv ... FILEn.csv" % name)
     eprint("  -f            Filter out lines that have at least one field missing")
+    eprint("  -s I1:I2:...:Ik Sum the values from specified files 1..n and add as new column")
     eprint("  -h            Print this message")
     eprint("  -l LABELS     Provide comma-separated set of heading labels")
     eprint("  FILE1.csv ... Source files")
@@ -60,6 +61,7 @@ def addData(fname, first = False):
                 eprint("File %s, row %d.  Couldn't find key '%s'" % (fname, row, key))
                 sys.exit(1)                
             lines[row-1] += "," + val
+    row += 1
     while row <= len(lines):
         lines[row-1] += ","
         missingKeys.add(keys[row-1])
@@ -78,7 +80,22 @@ def filter():
             keys.append(key)
             lines.append(line)
     
-def build(lstring, flist, doFilter):
+def sumLine(sumSet):
+    global lines
+    olines = lines
+    lines = []
+    for (key,line) in zip(keys, olines):
+        fields = line.split(",")
+        sfields = [fields[i] for i in range(len(fields)) if i in sumSet]
+        sval = ""
+        try:
+            nums = [float(field) if len(field) > 0 else 0.0 for field in sfields]
+            sval = str(sum(nums))
+        except:
+            eprint("key = %s.  Couldn't parse numbers in line '%s'" % (key, line))
+        lines.append(line + "," + sval)
+
+def build(lstring, flist, doFilter, sumSet):
     global lines
     first = True
     for fname in flist:
@@ -86,6 +103,8 @@ def build(lstring, flist, doFilter):
         first = False
     if doFilter:
         filter()
+    if sumSet is not None:
+        sumLine(sumSet)
     if len(lstring) > 0:
         lines = [lstring] + lines
     for line in lines:
@@ -93,21 +112,31 @@ def build(lstring, flist, doFilter):
 
 def run(name, args):
     doFilter = False
+    sumSet = None
     lstring = ""
-    optList, args = getopt.getopt(args, "hfl:")
+    optList, args = getopt.getopt(args, "hfs:l:")
     for (opt, val) in optList:
         if opt == '-h':
[…]

### 2023-02-20 doc/mc-data.tex
Document experiments

@@
 %% Colored hyperlink 
-\newcommand{\cref}[2]{\href{#1}{\color{blue}#2}}
-%% Colored hyperlink showing link in TT font
-% \newcommand{\chref}[1]{\href{#1}{\small\tt \color{blue}#1}}
-\newcommand{\hcref}[1]{\cref{#1}{\small\tt #1}}
+\newcommand{\hcref}[1]{\texttt{\small\url{#1}}}
 
@@
 
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
-                              x post scale=2.0, y post scale=2.0,
-                              xmin=0,xmax=1000,
-                              xtick = {0,100,200,300,400,500,600,700,800,900,1000},
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \input{data-formatted/time+defining-success}
-    \input{data-formatted/time+defining-failure}
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Number of defining clauses in generated POG as function of D4 runtime}
-\label{fig:generate:d4}
-\end{figure}
+\section*{Benchmark machine}
+\begin{itemize}
+\item 2021 MacBook Pro 18,4
+  \begin{itemize}
+  \item 3.2 Ghz Apple M1 processor
+  \item 64 GB Ram
+  \end{itemize}
+\item  Samsung T7 solid-state disk
+  \begin{itemize}
+  \item 2TB capacity
+  \item Up to 1 GB/s R/W
+  \item Critical for performance.  Wrote and read files of over 162 GB.
+  \end{itemize}
+\end{itemize}
+
+\section*{Methodology}
+\begin{enumerate}
+   \item Retrieved 200 public model counting benchmarks from 2022 Model Counting Competition website
+     \begin{center}
+       \hcref{https://mccompetition.org/2022/mc\_description.html}
+     \end{center}
+     \begin{itemize}
+     \item 100 public instances from Track 1 (unweighted model counting)
+     \item 100 public instances from Track 2 (weighted model counting)
+     \end{itemize}
+   \item Removed duplicates
+     \begin{itemize}
+    \item  20 instances were duplicated in the two tracks
+     \item Reduces set of benchmarks to 180
+     \end{itemize}
+   \item Ran D4 on 180 instances with time limit of 1000 seconds
+     \begin{itemize}
+     \item Completed for 114 instances
+     \end{itemize}
+   \item Generated POG representations from dDNNF graphs
+     \begin{itemize}
+     \item Simplifications via constant propagation
+       \begin{itemize}
+       \item Number of POG nodes around $0{.}90\times$ number of dDNNF nodes
+       \end{itemize}
+     \item Used heuristics to identify root node, and then put in topological ordering
+     \item Measure complexity by number of defining clauses $D$
+       \begin{itemize}
+       \item Product or sum node with degree $k$ requires $k+1$ clauses
+       \item $D$ also equals sum of number of edges and number of nodes
+       \item For 114 instances, Min = 304, Median = 674,179, Mean = 27,918,017, Max = 404,276,843
+       \end{itemize}
+     \end{itemize}
+   \item Ran program {\sc d2p} to generate CRAT file for maximum of 10,000 seconds
+     \begin{itemize}
+     \item Completed for 105 instances
+     \item Timed out for eight instances
+     \item Ran out of memory for one instance
+     \end{itemize}
+   \item For completed CRAT files, ran {\sc crat-check} CRAT checker
+     \begin{itemize}
+     \item Tests all conditions for correct CRAT
+     \item Most time spent performing RUP checks for clause addition and deletion
+     \item Completed successfully for all 105 instances
+     \item Time generally faster than {\sc d2p}.
+     \end{itemize}
+\end{enumerate}
 
@@
 \end{figure}
+
+\section*{Generated CRAT Files}
+
+\begin{itemize}
+\item Proof Clauses: Min 554, Average 44,335,021, Max 770,382,773
+\item Ratio to number of defining clauses: Min 1.54, Harmonic Mean 3.05, Max 6073
+\end{itemize}
 
@@
 
+\section*{Time to Generate and Validate CRAT}
+
+\begin{itemize}
+\item Min 0.01, Average 1190.0, Max 13,144.7
+\item Ratio to time for D4: Min 0.50, Harmonic Mean 6.16, Max 457.7
+\end{itemize}
+
 \begin{figure}
@@
 
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 1.00]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
+                              x post scale=2.0, y post scale=2.0,
+                              xmin=0,xmax=1000,
+                              xtick = {0,100,200,300,400,500,600,700,800,900,1000},
+                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
+                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
+%                              title={D4 Defining Clause Generation}
+            ]
+    \input{data-formatted/time+defining-success}
+    \input{data-formatted/time+defining-failure}
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Number of defining clauses in generated POG as function of D4 runtime (X linear scale)}
+\label{fig:generate:d4}
+\end{figure}
+
 \bibliography{references}

### 2023-02-21 doc/mc-data.tex
New information about weighted counts

@@
 
+\section*{Computing Counts and Weighted Counts}
+
+\begin{itemize}
+\item Implemented package to represent numbers of form $a \cdot 2^b \cdot 5^c$.
+  \begin{itemize}
+  \item For unweighted counts, first compute density and then scale by $2^n$
+  \item For unweighted counts, perform exact decimal arithmetic
+  \item The set of numbers of this form is not closed under a reciprocal operation, but the weights in the competion are either of the form
+    $w(x) + w(\overline{x}) = 1.0$ or $w(x) = w(\overline{x}) = 1$.  The latter case sums to 2, with reciprocal $1/2$.
+  \end{itemize}
+\item Can implement having $a$ be integer of arbitrary size and $b$ and $c$ being 32-bit integers.
+\item Weighted counts ranged up to 260,909 digits.
+\end{itemize}
+
 \bibliography{references}

### 2023-02-26 doc/mc-data.tex
Bring data up to date

@@
 \definecolor{offwhite}{rgb}{0.839216, 0.823529, 0.768627}
+\definecolor{medgreen}{rgb}{0.15, 0.6, 0.15}
 
@@
      \end{itemize}
-   \item Ran D4 on 180 instances with time limit of 1000 seconds
+   \item Ran D4 on 180 instances with time limit of 4000 seconds
      \begin{itemize}
-     \item Completed for 114 instances
+     \item Completed for 124 instances
      \end{itemize}
@@
        \item $D$ also equals sum of number of edges and number of nodes
-       \item For 114 instances, Min = 304, Median = 674,179, Mean = 27,918,017, Max = 404,276,843
+       \item For 124 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
        \end{itemize}
@@
      \begin{itemize}
-     \item Completed for 105 instances
-     \item Timed out for eight instances
-     \item Ran out of memory for one instance
+     \item Completed for 108 instances
+     \item Exceeded clause limit for 1 instance
+     \item Ran out of memory for 1 instance
+     \item Timed out for 14 instances
      \end{itemize}
@@
      \item Most time spent performing RUP checks for clause addition and deletion
-     \item Completed successfully for all 105 instances
-     \item Time generally faster than {\sc d2p}.
+     \item Completed successfully for all 108 instances
+     \item Time generally faster than {\sc d2p}. (min = 0.005, max = 1.4, harmonic mean = 0.10)
      \end{itemize}
@@
 \begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=10000000000,
+      legend style={at={(0.30,0.98)}},
+      legend cell align={left},
                               x post scale=2.0, y post scale=2.0,
-                              xmode=log, xmin=001,xmax=1000,
-                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
+                              xmode=log, xmin=001,xmax=10000,
+                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
+                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$,$10^{10}$},
                               xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
@@
             ]
+    \input{data-formatted/time+defining-failure}
     \input{data-formatted/time+defining-success}
-    \input{data-formatted/time+defining-failure}
+    \legend{
+      \scriptsize \textsf{Limit exceeded},
+      \scriptsize \textsf{Validated},
+    }
  \end{axis}
@@
 \begin{itemize}
-\item Proof Clauses: Min 554, Average 44,335,021, Max 770,382,773
-\item Ratio to number of defining clauses: Min 1.54, Harmonic Mean 3.05, Max 6073
+\item Proof Clauses: Min 554, Median 1,719,245, Average 45,001,945, Max 1,719,245
+\item Ratio to number of defining clauses: Min 1.54, Harmonic Mean 3.13, Max 6073
 \end{itemize}
@@
             ]
-    \addplot[mark=none, color=darkestblue] coordinates{(100,100) (1e9,1e9)};
-    \addplot[mark=none, color=midblue] coordinates{(100,1000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
     \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
@@
 \begin{itemize}
-\item Min 0.01, Average 1190.0, Max 13,144.7
-\item Ratio to time for D4: Min 0.50, Harmonic Mean 6.16, Max 457.7
+\item Min 0.01, Average 1318.3, Max 13,144.8, Median 71.64
+\item Ratio to time for D4: Min 0.50, Harmonic Mean 5.5, Max 457.7
 \end{itemize}
@@
 \begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=20000,
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=100000,
                               x post scale=2.0, y post scale=2.2,
-                              xmode=log,xmin=0.01,xmax=1000,
-                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
-                              ytick={0.01, 0.1,1.0,10,100,1000,10000,20000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{20,000}},
+                              xmode=log,xmin=0.01,xmax=10000,
+                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
+                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
                               xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
@@
             ]
-    \addplot[mark=none, color=darkestblue] coordinates{(0.01,0.01) (1000.0,1000.0)};
-    \addplot[mark=none, color=midblue] coordinates{(0.01,0.1) (1000,10000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (200,20000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
     \input{data-formatted/time-d4-verify}
@@
 \end{figure}
+
+\section*{Computing Counts and Weighted Counts}
+
+\begin{itemize}
+\item Implemented package to represent numbers of form $a \cdot 2^b \cdot 5^c$.
+  \begin{itemize}
+  \item For unweighted counts, first compute density and then scale by $2^n$
+  \item For unweighted counts, perform exact decimal arithmetic
+  \item The set of numbers of this form is not closed under a reciprocal operation, but the weights in the competion are either of the form
+    $w(x) + w(\overline{x}) = 1.0$ or $w(x) = w(\overline{x}) = 1$.  The latter case sums to 2, with reciprocal $1/2$.
+  \end{itemize}
+\item Can implement having $a$ be integer of arbitrary size and $b$ and $c$ being 32-bit integers.
+\item Weighted counts ranged up to 260,909 digits.
+\end{itemize}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 1.00]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1e9, 
+      legend style={at={(0.40,0.98)}},
+      legend cell align={left},
+                              x post scale=2.0, y post scale=2.3,
+                              xmode=log,xmin=100,xmax=1e7, 
+                              xtick={100,1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
+                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
+                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
+            ]
+    \input{data-formatted/clauses-lemma-nolemma}
+    \input{data-formatted/clauses-nosplit-split}
+    \legend{
+      \scriptsize \textsf{Without lemmas},
+      \scriptsize \textsf{Without literal grouping},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e8,1e8)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
+
+    \node[left] at (axis cs: 900,1200) {$1{}.0\times$};
+    \node[left] at (axis cs: 900,12000) {$10\times$};
+    \node[left] at (axis cs: 900,120000) {$100\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Proof Clauses when Disable Options}
+\label{fig:defining:unoptimized}
+
+\end{figure}
+
+\bibliography{references}
+
+\end{document}
 
@@
 \end{figure}
-
-\section*{Computing Counts and Weighted Counts}
-
-\begin{itemize}
-\item Implemented package to represent numbers of form $a \cdot 2^b \cdot 5^c$.
-  \begin{itemize}
-  \item For unweighted counts, first compute density and then scale by $2^n$
-  \item For unweighted counts, perform exact decimal arithmetic
-  \item The set of numbers of this form is not closed under a reciprocal operation, but the weights in the competion are either of the form
-    $w(x) + w(\overline{x}) = 1.0$ or $w(x) = w(\overline{x}) = 1$.  The latter case sums to 2, with reciprocal $1/2$.
-  \end{itemize}
-\item Can implement having $a$ be integer of arbitrary size and $b$ and $c$ being 32-bit integers.
-\item Weighted counts ranged up to 260,909 digits.
-\end{itemize}
-
-\bibliography{references}
-
-\end{document}

### 2023-02-26 tools/gen-data.py

@@ -10,7 +10,7 @@ import getopt
 #          \addplot coordinates {(4, 900) (8, 8320) (9, 39628) (11, 252970) (12, 1324539) (13, 4095084) (14, 7131764) (15, 15225960)};
 
 # Threshold value.  Don't include data with Y values exceeding this value
-yThresh = 1000 * 1000 * 1000
+yThresh = 10000 * 1000 * 1000
 xThresh = 1000 * 1000 * 1000
 # Set any y value below yMin to yMin
 yMin = 0.0

### 2023-02-27 doc/mc-data.tex
Data on optimization benefits

@@
     \input{data-formatted/defining+total}
-    \node[left] at (axis cs: 900,1200) {$1{}.0\times$};
+    \node[left] at (axis cs: 900,1200) {$1\times$};
     \node[left] at (axis cs: 900,12000) {$10\times$};
@@
     \input{data-formatted/time-d4-verify}
-    \node[left] at (axis cs: 0.09,0.12) {$1{}.0\times$};
+    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
     \node[left] at (axis cs: 0.09,1.2) {$10\times$};
@@
 
+\section*{Benefits of Optimizations}
+
+Two optimizations were implemented within the CRAT framework:
+\begin{description}
+\item[Lemmas:] The justification proof must show that the conjunction
+  of the input clauses implies the extension variable associated with
+  the root node.  It does so by showing that this property holds for
+  every node and for every path from the root that node, relative to
+  the partial assignment given by that path.  This could potentially
+  require enumerating exponential numbers of paths to some nodes,
+  effectively converting the graph into a tree.  Instead, the proof
+  can contain a {\em lemma} for each node having indegree greater than
+  one, showing that the root node is implied by a set of clauses.
+  Some of these are the original input clauses, while some are {\em
+    synthetic clauses} representing input clauses that have been
+  reduced by eliminating false literals.  Given such a proof, the
+  lemma can be instantiated for each parent node.  Proofs with lemmas
+  incur some overhead---the synthetic clauses are declared as POG
+  product nodes, and additional clauses must be added to show that,
+  under a given context, the synthetic clauses are implied by some
+  input or containing synthetic clause.  On the other hand, they
+  guarantee that no node in the graph is the root of more than one
+  justication proof.
+\item[Literal grouping:] The generated POG representation will
+  typically have a number of product nodes forming the conjunction of
+  tens, or even hundreds, of literals.  These literals represent ones
+  that we either implied by unit propagation or were added when
+  assigning the complement to the corresponding variable led to a
+  conflict.  Justifying this product node requires showing that each
+  of these literals must hold under the given context.  When this
+  cannot be done by unit propagation, the program must invoke a SAT
+  solver to prove that the complemented value yields a conflict.
+  Rather than doing so for each literal individually, the prover can
+  create a new product node consisting of the literals that have not
+  yet been validated and then prove that falsifying the extension
+  variable associated with the product node yields a conflict.  
+\end{description}
+
 \begin{figure}
@@
 \begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1e9, 
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
       legend style={at={(0.40,0.98)}},
@@
                               x post scale=2.0, y post scale=2.3,
-                              xmode=log,xmin=100,xmax=1e7, 
-                              xtick={100,1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
+                              xmode=log,xmin=1000,xmax=1e7, 
+                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
+                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
                               xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
@@
     \input{data-formatted/clauses-nosplit-split}
+    \input{data-formatted/clauses-optimized-unoptimized}    
     \legend{
@@
       \scriptsize \textsf{Without literal grouping},
+      \scriptsize \textsf{Without either optimization},
     }
-    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e8,1e8)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
 
-    \node[left] at (axis cs: 900,1200) {$1{}.0\times$};
-    \node[left] at (axis cs: 900,12000) {$10\times$};
-    \node[left] at (axis cs: 900,120000) {$100\times$};
+    \node[left] at (axis cs: 5000,7000) {$1\times$};
+    \node[left] at (axis cs: 5000,70000) {$10\times$};
+    \node[left] at (axis cs: 5000,700000) {$100\times$};
 
@@
 \end{figure}
+
+We conducted a set of experiments using the 80 benchmark instances for
+which the total number of proof clauses when both optimizations are
+performed was limited to $10^7$.  We then made three runs---with each
+of the two optimizations disabled individually and in combination.
+Figure \ref{fig:defining:unoptimized} illustrates the effect of the
+two optimizations.  Each set of points shows the number of clauses
+when both optimizations are performed for the $X$ value and the number
+with one or both of these optimizations disabled as the $Y$ value.
+
+The ratios between the number of clauses when either or both optimizations is disabled versus when both are enabled can be summarized as:
+\begin{description}
+\item[Without lemmas:] Min 1.00, Max 52.54, Harmonic mean 2.11
+\item[Without literal grouping:] Min 0.84, Max 39.60, Harmonic mean 1.18
+\item[Without either optimization:] Min 0.99, Max 52.54, Harmonic mean 2.95
+\end{description}
+Put another way: None of the 80 benchmarks require more than $10^7$
+proof clauses when both optimizations are used.  Without lemmas, 19
+rise above that threshold.  Without literal merging, 7 due.  Without
+either, 25 rise above that threshold and 2 even exceed $10^8$ proof
+clauses.
 

### 2023-02-27 doc/mc-data.tex
Data update

@@
        \end{itemize}
-     \item Used heuristics to identify root node, and then put in topological ordering
+     \item Used heuristics to identify root node, and then put in topological order with root node at end.
      \item Measure complexity by number of defining clauses $D$
@@
      \item Completed for 108 instances
-     \item Exceeded clause limit for 1 instance
+     \item Exceeded clause limit for 1 instance (Our programs represent a clause ID as a 32-bit signed value)
      \item Ran out of memory for 1 instance
@@
 \begin{itemize}
-\item Proof Clauses: Min 554, Median 1,719,245, Average 45,001,945, Max 1,719,245
-\item Ratio to number of defining clauses: Min 1.54, Harmonic Mean 3.13, Max 6073
+\item Proof Clauses: Min 554, Median 1,719,245, Average 45,001,945, Max 770,382,773
+\item Ratio to number of defining clauses: Min 1.54, Harmonic Mean 3.13, Max 6072.96
 \end{itemize}
@@
     \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
     \input{data-formatted/defining+total}
@@
     \node[left] at (axis cs: 900,120000) {$100\times$};
+    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
+    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};
 
@@
 \item Min 0.01, Average 1318.3, Max 13,144.8, Median 71.64
-\item Ratio to time for D4: Min 0.50, Harmonic Mean 5.5, Max 457.7
+\item Ratio to time for D4: Min 0.50, Harmonic Mean 5.50, Max 457.69
 \end{itemize}
@@
     \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
     \input{data-formatted/time-d4-verify}
@@
     \node[left] at (axis cs: 0.09,12.0) {$100\times$};
+    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
 
@@
 Two optimizations were implemented within the CRAT framework:
+\begin{description}
+\item[Lemmas:] Nodes with indegree greater than one were validated with separate lemmas, avoiding the possibly exponential expansion of the POG into a tree during the proof construction.
+
+\item[Literal grouping:] For a product node with multiple literal
+  arguments, those that cannot be validated by BCP are combined as
+  arguments to a newly defined product node.  The extension variable
+  associated with this node is validated, and the proof is then used
+  to validate the literals.  This reduces the number of
+  calls to the SAT solver to at most one per product node.
+\end{description}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 1.00]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
+      legend style={at={(0.40,0.98)}},
+      legend cell align={left},
+                              x post scale=2.0, y post scale=2.3,
+                              xmode=log,xmin=1000,xmax=1e7, 
+                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
+                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
+                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
+            ]
+    \input{data-formatted/clauses-lemma-nolemma}
+    \input{data-formatted/clauses-nosplit-split}
+    \legend{
+      \scriptsize \textsf{Without lemmas},
+      \scriptsize \textsf{Without literal grouping},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
+
+    \node[left] at (axis cs: 5000,7000) {$1\times$};
+    \node[left] at (axis cs: 5000,70000) {$10\times$};
+    \node[left] at (axis cs: 5000,700000) {$100\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Proof Clauses when Single Optimization Disabled}
+\label{fig:optimized:partially-optimized}
+\end{figure}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 1.00]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
+      legend style={at={(0.40,0.98)}},
+      legend cell align={left},
+                              x post scale=2.0, y post scale=2.3,
+                              xmode=log,xmin=1000,xmax=1e7, 
+                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
+                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
+                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
+            ]
+    \input{data-formatted/clauses-optimized-unoptimized}    
+    \legend{
+      \scriptsize \textsf{Without either optimization},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
+
+    \node[left] at (axis cs: 5000,7000) {$1\times$};
+    \node[left] at (axis cs: 5000,70000) {$10\times$};
+    \node[left] at (axis cs: 5000,700000) {$100\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Proof Clauses Both Optimizations are Disabled}
+\label{fig:optimized:unoptimized}
+\end{figure}
+
+We conducted a set of experiments using the 80 benchmark instances for
+which the total number of proof clauses when both optimizations are
+performed was limited to $10^7$.  We then made three additional runs
+of the CRAT generator and checker, with each of the two optimizations
+disabled individually and in combination.  Figures
+\ref{fig:optimized:partially-optimized}--\ref{fig:optimized:unoptimized} illustrates the effect of the two
+optimizations.  Each set of points shows the number of clauses when
+both optimizations are performed for the $X$ value and the number with
+one or both of these optimizations disabled as the $Y$ value.
+
+The ratios between the number of clauses when either or both optimizations is disabled versus when both are enabled can be summarized as:
+\begin{description}
+\item[Without lemmas:] Min 1.00, Max 52.54, Harmonic mean 2.11
+\item[Without literal grouping:] Min 0.84, Max 39.60, Harmonic mean 1.18
+\item[Without either optimization:] Min 0.99, Max 52.54, Harmonic mean 2.95
+\end{description}
+
+Here are other ways to summarize the impact on proof length:
+\begin{itemize}
+\item None of the 80 benchmarks require more than $10^7$
+proof clauses when both optimizations are used.  Without lemmas, 19
+rise above that threshold.  Without literal merging, 7 due.  Without
+either, 25 rise above the threshold.  Two of the benchmarks require
+more than $10^8$ proof clauses without lemmas.
+\item Of the 80 benchmark problems evaluated, and comparing with both
+  optimizations disabled, 55 had the number of proof clauses reduced
+  by at least 10\% by using lemmas, 19 by using literal merging, and
+  69 by enabling both optimizations.
+\end{itemize}
+
+The time performance is less dramatic, but still significant.
+Comparing the sum of CRAT generation and checking times for each of
+the four combinations of options, we get the following times relative
+to the one with both optimizations enabled:
+\begin{description}
+\item[Without lemmas:] Min 0.85, Max 37.98, Harmonic mean 1.80
+\item[Without literal grouping:] Min 0.83, Max 3.92, Harmonic mean 1.12
+\item[Without either optimization:] Min 0.91, Max 41.97, Harmonic mean 2.19
+\end{description}
+
+Here is another way to summarize the impact on the combined time to generate and validate the CRAT file:
+\begin{itemize}
+\item Of the 80 benchmark problems evaluated, and comparing with both
+  optimizations disabled, 50 had the time reduced by at least 10\% by
+  using lemmas, 27 by using literal merging, and 63 by enabling both
+  optimizations.
+\end{itemize}
+
+\bibliography{references}
+
+\end{document}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 1.00]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
+                              x post scale=2.0, y post scale=2.0,
+                              xmin=0,xmax=1000,
+                              xtick = {0,100,200,300,400,500,600,700,800,900,1000},
+                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
+                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
+%                              title={D4 Defining Clause Generation}
+            ]
+    \input{data-formatted/time+defining-success}
+    \input{data-formatted/time+defining-failure}
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Number of defining clauses in generated POG as function of D4 runtime (X linear scale)}
+\label{fig:generate:d4}
+\end{figure}
+
 \begin{description}
@@
 \end{description}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
-      legend style={at={(0.40,0.98)}},
-      legend cell align={left},
-                              x post scale=2.0, y post scale=2.3,
-                              xmode=log,xmin=1000,xmax=1e7, 
-                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
-                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
-                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
-            ]
-    \input{data-formatted/clauses-lemma-nolemma}
-    \input{data-formatted/clauses-nosplit-split}
-    \input{data-formatted/clauses-optimized-unoptimized}    
-    \legend{
-      \scriptsize \textsf{Without lemmas},
-      \scriptsize \textsf{Without literal grouping},
-      \scriptsize \textsf{Without either optimization},
-    }
-    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
-
-    \node[left] at (axis cs: 5000,7000) {$1\times$};
-    \node[left] at (axis cs: 5000,70000) {$10\times$};
-    \node[left] at (axis cs: 5000,700000) {$100\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Proof Clauses when Disable Options}
-\label{fig:defining:unoptimized}
-
-\end{figure}
-
-We conducted a set of experiments using the 80 benchmark instances for
-which the total number of proof clauses when both optimizations are
-performed was limited to $10^7$.  We then made three runs---with each
-of the two optimizations disabled individually and in combination.
-Figure \ref{fig:defining:unoptimized} illustrates the effect of the
-two optimizations.  Each set of points shows the number of clauses
-when both optimizations are performed for the $X$ value and the number
-with one or both of these optimizations disabled as the $Y$ value.
-
-The ratios between the number of clauses when either or both optimizations is disabled versus when both are enabled can be summarized as:
-\begin{description}
-\item[Without lemmas:] Min 1.00, Max 52.54, Harmonic mean 2.11
-\item[Without literal grouping:] Min 0.84, Max 39.60, Harmonic mean 1.18
-\item[Without either optimization:] Min 0.99, Max 52.54, Harmonic mean 2.95
-\end{description}
-Put another way: None of the 80 benchmarks require more than $10^7$
-proof clauses when both optimizations are used.  Without lemmas, 19
-rise above that threshold.  Without literal merging, 7 due.  Without
-either, 25 rise above that threshold and 2 even exceed $10^8$ proof
-clauses.
-
-\bibliography{references}
-
-\end{document}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
-                              x post scale=2.0, y post scale=2.0,
-                              xmin=0,xmax=1000,
-                              xtick = {0,100,200,300,400,500,600,700,800,900,1000},
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \input{data-formatted/time+defining-success}
-    \input{data-formatted/time+defining-failure}
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Number of defining clauses in generated POG as function of D4 runtime (X linear scale)}
-\label{fig:generate:d4}
-\end{figure}

### 2023-02-28 doc/mc-data.tex
Small updates

@@
 \begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=10000000000,
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=3000000000,
       legend style={at={(0.30,0.98)}},
@@
                               x post scale=2.0, y post scale=2.0,
-                              xmode=log, xmin=001,xmax=10000,
+                              xmode=log, xmin=001,xmax=4000,
                               xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
@@
 \begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=100000,
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=20000,
                               x post scale=2.0, y post scale=2.2,
-                              xmode=log,xmin=0.01,xmax=10000,
+                              xmode=log,xmin=0.01,xmax=4000,
                               xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
@@
             ]
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
     \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
@@
     \input{data-formatted/time-d4-verify}
+    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
     \node[left] at (axis cs: 0.09,0.12) {$1\times$};
@@
 more than $10^8$ proof clauses without lemmas.
-\item Of the 80 benchmark problems evaluated, and comparing with both
-  optimizations disabled, 55 had the number of proof clauses reduced
-  by at least 10\% by using lemmas, 19 by using literal merging, and
-  69 by enabling both optimizations.
+\item
+  Of the 80 benchmark problems evaluated, and comparing with both
+  optimizations disabled, 47 had the number of proof clauses reduced
+  by a factor of 2 by using lemmas, 14 by using literal merging, and
+  60 by enabling both optimizations.
 \end{itemize}
@@
 \item Of the 80 benchmark problems evaluated, and comparing with both
-  optimizations disabled, 50 had the time reduced by at least 10\% by
-  using lemmas, 27 by using literal merging, and 63 by enabling both
+  optimizations disabled, 40 had the time reduced by at least a factor of 2
+  using lemmas, 7 by using literal merging, and 49 by enabling both
   optimizations.

### 2023-03-01 doc/mc-data.tex
Start on conference paper

@@
                               x post scale=2.0, y post scale=2.0,
-                              xmode=log, xmin=001,xmax=4000,
+                              xmode=log, xmin=0.01,xmax=4000,
                               xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},

### 2023-03-01 doc/mc-paper.tex
Start on conference paper

\documentclass[letterpaper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\usepackage{amsmath}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{booktabs}

\newcommand{\pand}{\mathbin{\land^{\sf v}}}
\newcommand{\por}{\mathbin{\lor^{\sf a}}}
\DeclareMathOperator*{\Pand}{\bigwedge^{\sf v}}
\DeclareMathOperator*{\Por}{\bigvee^{\sf a}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

\newcommand{\varset}{X}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}

\newcommand{\assign}{\alpha}
\newcommand{\uassign}{{\cal U}}
\newcommand{\modelset}{{\cal M}}

\newcommand{\indegree}{\textrm{indegree}}
\newcommand{\outdegree}{\textrm{outdegree}}

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting}

\titlerunning{Certified Knowledge Compilation}

\author{Randal E. Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{rebryant@cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}
\author{Wojciech Nawrocki}{Department of Philosophy, Carnegie Mellon University}{wjnawrock@cmu.edu}{https://orcid.org/0000-0002-8839-0618}{}
\author{Jeremy Avigad}{Department of Philosophy, Carnegie Mellon University}{avigad@cmu.edu}{https://orcid.org/0000-0003-1275-315X}{}
\author{Marijn J. H. Heule}{Computer Science Department, Carnegie Mellon University}{marijn@cmu.edu}{https://orcid.org/0000-0002-5587-8801}{Supported by NSF grant CCF-2108521}

%\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

%\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

%\category{} %optional, e.g. invited paper

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\EventEditors{John Q. Open and Joan R. Access}
%\EventNoEds{2}
%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
%\EventShortTitle{CVIT 2016}
%\EventAcronym{CVIT}
%\EventYear{2016}
%\EventDate{December 24--27, 2016}
%\EventLocation{Little Whinging, United Kingdom}
%\EventLogo{}
%\SeriesVolume{42}
%\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\program}[1]{\textsc{#1}}
\newcommand{\lean}{\program{lean4}}

\begin{document}

\maketitle

\begin{abstract}
Knowledge compilers convert Boolean formulas into a form for which
useful tasks, including model counting and weighted model counting,
can easily be performed.  Existing knowledge compilers, however,
provide no guarantee that the generated representation accurately
describes the original formula.

We present the Partitioned Operation Graph (POG) representation that
generalizes existing knowledge compiler representations, and a proof
framework that can prove the equivalence of a POG representation with
a conjunctive normal form (CNF) Boolean formula.

We have developed a program that can generate POG representations and
equivalence proofs from the dDNNF graphs generated by a leading
knowledge compiler.  It scales to all but the largest formulas from a
recent model counting competition.  We also present two proof
checkers---a fast checker that can handle the large proofs generated,
and a formally verified checker written in the \lean{} proof system.
\end{abstract}

\section{Introduction}

\section{Logical Foundations}

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an {\em assignment} of truth values to the variables, where $0$
  denotes false and $1$ denotes true, i.e., $\assign \colon \varset
  \rightarrow \{0,1\}$.  The set of all possible assignments over
  $\varset$ is denoted $\uassign$.  For each variable $x \in \varset$,
  we define {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, where
  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item A variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em conjunction} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em disjunction} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{i=1,k} \modelset(\phi_i)$.
  \end{enumerate}

  We highlight the following special classes of Boolean formulas:
  \begin{itemize}
    \item In a {\em normal form} formula, negation is 
      applied only to variables
    \item A normal-form formula is in {\em conjunctive normal form} (CNF) if disjunction is applied only to literals.
    \item A {\em partitioned-operation formula} (or simply a ``partitioned formula'') satisfies the following restrictions on its conjunction and disjunction operations:
      \begin{itemize}
      \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
        $\phi = \bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a disjunction must have disjoint models.  That is, operation
        $\phi = \bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{itemize}
     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned formula.
  \end{itemize}

  \section{Ring Evaluation of a Boolean Formula}

We propose a general framework for describing a variety of tasks that
summarize some aspect of Boolean formulas.

\begin{definition}[Commutative Ring]
  A {\em commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$, 
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  {\em additive inverse} $-a$ such that $a + -a = \addident$.
\label{def:ring}
\end{definition}

\begin{definition}[Ring Evaluation Problem (REP)]
  A {\em ring labeling} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We define $w(\obar{x})$ to equal $1-w(x)$.

  For Boolean formula $\phi$, the {\em ring evaluation problem} (REP) is to compute:
  \begin{eqnarray}
    \rep(\phi) & = & \sum_{\alpha \in \modelset(\phi)} \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
\label{def:labeling}
\end{definition}

A wide variety of important tasks for Boolean functions can be
expressed as ring evaluation problems.  The 
{\em model counting} problem can be cast as an REP by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.  
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow independent
assignments of weights $W$ to the variables and their negations, such
that $W(x) + W(\obar{x}) \not = 1$.  We can convert this into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing algebraic evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$.

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, let $\ring$ be a
finite  field with $|\dset| \geq 2 |\varset|$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1) = \rep(\phi_2)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1) \not = \rep(\phi_2)$ will be at
least $\left(1-\frac{1}{z}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Therefore, ring evaluation can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

\section{Partitioned-Operation Graphs (POGs)}

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a partitioned formula:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following:
\begin{description}
\item[Negation] $\rep(\boolnot \phi) = \mulident - \rep(\phi)$
\item[Partitioned Conjunction] $\rep\left(\Pand_{i=1,k} \phi_i\right) = \prod_{i=1,k} \rep(\phi_i)$
\item[Partitioned Disjunction] $\rep\left(\Por_{i=1,k} \phi_i\right) = \sum_{i=1,k} \rep(\phi_i)$
\end{description}
\end{proposition}

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  When $(u,v) \in E$,
node $v$ is said to be a {\em child} of node $u$.
The in- and out-degrees of node $u$ are defined as $\indegree(u) = | E \cap (N \times \{u\}) |$, and
$\outdegree(u) = | E \cap (\{u\} \times N) |$.
Node $u$ is said to be {\em terminal} if $\outdegree(u) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $u$ is said to be {\em nonterminal} if $\outdegree(u) > 0$.  A nonterminal node is labeled by one of three Boolean operations:
$\boolnot$, $\pand$, or $\por$.  Node $u$ can be labeled with $\boolnot$ only if $\outdegree(u) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $r$ such that $\indegree(r) = 0$.

As can be seen, a POG is simply a way to represent a partitioned
formula with a sharing of common subformulas, and so we write
$\phi(P)$ as the formula denoted by $P$.

We define the {\em size} of POG $P$, written $|P|$ to be the sum of
the number of nonterminal nodes and the number of edges.  Ring
evaluation of $\phi(P)$ can be performed with at most $|P|$ ring
operations by traversing the graph from the terminal nodes up to
the root, computing a value $\rep(u)$ for each node $u$.

POGs are inspired by the d-DNNF graphs devised by Darwiche for
representing Boolean functions \cite{darwiche:jair:2002}.
POGs generalize d-DNNF in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.  As we have seen, ring evaluation can readily handle negation, and so we allow it in the interest of generality.
  Negation could be useful in some contexts, such as when
  converting binary decision diagrams (BDDs) having complemented
  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.

\item It allows arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  disjunction node in a d-DNNF can have at most two children, and there must be a variable $x$ such that
  the satisfying assignments for one child have $x=1$ while those for the
  other child have $x=0$.  Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

\section{The CRAT Proof System}

The CRAT clausal proof framework provides a means to express a
checkable proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to a POG\@, thus enabling certified
knowledge compilation.  It is based on extended
resolution~\cite{Tseitin:1983} with reverse unit propagation
(RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as its core method for
determining that a clause is implication-redundant with respect to other clauses.

Two programs are involved in producing a certified compilation of a formula:
\begin{itemize}
\item The {\em compiler} generates a CRAT file that both describes a POG and provides a proof that this POG is equivalent to the input formula.
  This compiler might start with the output of another knowledge compiler.
\item The {\em checker} ensures that all of the proof conditions are satisfied.  
\end{itemize}

The CRAT format draws its inspiration from the LRAT format for Boolean
formulas~\cite{lrat} and the QRAT format for quantified Boolean
formulas (QBF)~\cite{heule:JAR2014}.  The following are its key properties:
\begin{itemize}
  \item
  In addition to explicit clause additions and deletions, the proof contains declarations of $\pand$ and $\por$ operations.
\begin{itemize}
\item 
  Each such declaration implicitly adds an extension variable representing that operation and a set of {\em defining} clauses
  encoding the conjunction or disjunction operation.
  This is the only means for generating extension variables or adding blocked clauses to the proof.
\item The checker must track the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.  The resulting extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring
  $k\,(k-1)/2$ proofs of disjointness for larger values of $k$.
\item Boolean negation is provided implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\end{itemize}
\item
  Assertion clauses can can added when they satisfy the RUP
  property, with respect to a sequence of existing clauses.  This set must be provided explicitly as the {\em hints} for the assertion.
\item
  Deleting clauses requires proving that the resulting formula is not weaker.
  That is, it should not allow any additional satisfying assignment.
  Only the input clauses and those declared by an assertion can be deleted.
      The deletion must be accompanied by a sequence of remaining clauses
      providing a RUP proof of the clause.  This sequence must be provided as the {\em hints} for the deletion.
\end{itemize}

\subsection{Syntax}

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause} \\
     & {\tt dc} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause \\
    \midrule
    $C$    & {\tt p} & $V \; L^{*}$ {\tt 0}    &                  & Declare $\pand$ operation \\
    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
    \midrule
     & {\tt r} & $L$             &            & Declare root literal\\
  \end{tabular}
  }
\end{table}

Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $\lit$
is represented by a signed integer, with $-v$ being the complement of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that clause $C$ can only reference clauses $C'$ such that $C' < C$.

The first set of proof rules are similar to those in other clausal
proofs.
Clauses can be added via RUP addition
(command {\tt a}), with a sequence of antecedent clauses (the
``hint'').
Similarly for clause deletion (command {\tt dc}).

The declaration of a $\pand$ operation has the form:
\begin{center}
\begin{tabular}{ccccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\tt 0} \\
\end{tabular}
\end{center}
where $i$ is a new clause ID, $v$ is a positive integer that does not
correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$  
integers representing literals of existing variables.  
This declaration implicitly causes $k+1$ clauses to be added to the proof:
\begin{center}
\begin{tabular}{cccccc}
\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
  $i\!+\!1$ & $-v$ & $\lit_1$  \\
  $i\!+\!2$ & $-v$ & $\lit_2$  \\
  & $\ldots$ \\
  $i\!+\!k$ & $-v$ & $\lit_k$  \\
\end{tabular}
\end{center}
The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A $\pand$ operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then becomes a way
of representing constant $0$.

The declaration of a $\por$ operation has the form:
\begin{center}
\begin{tabular}{ccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} 
\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
\end{tabular}
\end{center}
where $i$ is a new clause ID, $v$ is a positive integer that does
not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
signed integers representing literals of existing variables.  Hint $H$
consists of a
sequence of clause IDs.
This declaration implicitly causes three clauses to be added to the proof:
\begin{center}
\begin{tabular}{cccc}
\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
  $i\!+\!1$ & $v$ & $-\lit_1$ \\
  $i\!+\!2$ & $v$ & $-\lit_2$ \\
\end{tabular}
\end{center}
The hints must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$.

Finally, the literal denoting the root of the POG is declared with the
{\tt r} command.  It can occur anywhere in the file.  Typically, it
will be the extension variable representing the root of a graph, but in
degenerate cases it can be the positive or negative literal of an
input variable.

\subsection{Semantics}

A CRAT proof follows the same general form as a QRAT dual
proof~\cite{bryant:cade:2021}---one that ensures that each clause
addition and each clause deletion preserves equivalence.  With CRAT,
however, clauses are defined both explicitly and implicitly.  Starting
with the set of input clauses, the proof consists of a sequence of
steps that both add and delete clauses.  Each addition must be truth
preserving, that is, any satisfying assignment for the set of clauses
before clause addition should still be a satisfying assignment after.
Each deletion must be falsehood preserving.  That is, there can be no new satisfying assignments when the clause is deleted.

At the completion of the proof, the following conditions must hold:
\begin{itemize}
\item All of the input clauses have been deleted.
\item There should be exactly one clause left that was added via RUP
  addition, and this should be a unit clause consisting of the root literal.
\end{itemize}

The sequence of operator declarations, asserted clauses, and deleted
clauses represents a systematic transformation of the input formula
into a POG\@.  Validating each of these steps serves to prove that the
POG is logically equivalent to the input formula.

\section{Generating CRAT from dDNNF}

\section{A Formally Verified CRAT Checker}

\section{Experimental Evaluation}

\section{Future Work}
\label{sect:future}

Each node $u$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
$s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(u)$ associated
with the node is defined recursively as $\phi(u) = \bigwedge_{i=1,k}
[\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
\cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
\cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
of such a node could have $k$ conjunction nodes: $t_i = p_i \pand s_i$
for $1 \leq i \leq k$, and then $k-1$ binary disjunction nodes to form
the disjunction of $t_1, \ldots, t_k$.

\bibliography{references}

\appendix

\section{Experimental Methodology}

\section{Detailed Experimental Results}

\end{document}

### 2023-03-02 doc/mc-paper.tex
Incremental progress on paper

@@
 \newcommand{\varset}{X}
+\newcommand{\exvarset}{Z}
 \newcommand{\dependencyset}{{\cal D}}
@@
 
+Observe that the defining clauses for the conjunction and disjunction
+operations provide a full Tseitin encoding of the operations.  Letting
+$\exvarset$ denote the set of added extension variables,  any
+assignment $\assign\colon \varset \rightarrow \{0,1\}$ has a unique extension,
+which we also call $\assign$, with $\assign\colon (\varset \cup \exvarset) \rightarrow \{0,1\}$.
+
 A CRAT proof follows the same general form as a QRAT dual
@@
 preserving, that is, any satisfying assignment for the set of clauses
-before clause addition should still be a satisfying assignment after.
-Each deletion must be falsehood preserving.  That is, there can be no new satisfying assignments when the clause is deleted.
+before clause addition should still be a satisfying assignment afterwards.
+Each deletion must be falsehood preserving.  That is, there can be no
+new satisfying assignments when the clause is deleted.
 
@@
 \item There should be exactly one clause left that was added via RUP
-  addition, and this should be a unit clause consisting of the root literal.
+  addition, and this should be a unit clause consisting of the root literal $r$.
 \end{itemize}
+The sequence of clause addition steps provide a proof that any
+assignment $\assign$ to the input variables that satisfies all of the
+clauses must, when extended, assign $\assign(r) = 1$.  Conversely,
+each proof step that deletes an input clause proves that any
+assignment $\alpha$ that falsifies the clause must, when extended,
+assign $\assign(r) = 0$.
 
@@
 
+\subsection{CRAT Example}
+
+\begin{figure}
+
+  \centering{
+\begin{minipage}{0.58\textwidth}
+(A)  Input Formula\\[1.2ex]
+\begin{tabular}{lll}
+\toprule
+\makebox[5mm]{ID} & \makebox[15mm]{Literals} & \\
+\midrule
+1 & \texttt{-1 3 -4} & \texttt{0} \\
+2 & \texttt{-1 -3 4} & \texttt{0} \\
+3 & \texttt{3 -4} & \texttt{0}\\
+4 & \texttt{1 -3 4} & \texttt{0} \\
+5 & \texttt{-1 -2} & \texttt{0} \\
+\bottomrule
+\end{tabular}
+\\[1.8ex]
+(B) CRAT Declaration of POG\\[1.2ex]
+\begin{tabular}{llll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Proof line} & Explanation \\
+\midrule
+6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
+9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
+12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
+15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\ 
+18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
+22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
+ & \texttt{r 10} && Root $r = s_{10}$\\
+\bottomrule
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.40\textwidth}
+(C) POG Visualization\\
+\input{dd/eg4}
+\end{minipage}
+}
+\caption{Example CNF Formula and POG Representation}
+\label{fig:eg4:pog}
+\end{figure}
+
+Figure \ref{fig:eg4:pog} illustrates an example formula and shows how
+the CRAT file declares its POG representation.  The input formula (A)
+consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
+$x_4$.  The generated POG has six nonterminal nodes representing four
+conjunctions and two disjunctions.  We name these by the node type (product or sum) and the extension variable.
+The first part of the CRAT file
+(B) declares these nodes using clause IDs that increment by three or
+four, depending on whether the node has two children or three.
+The last two nonzero values in the sum declarations are hints providing the required mutual exclusion proof.
+
+\begin{figure}
+\begin{minipage}{0.49\textwidth}
+(A) Implicit clauses\\[1.2ex]
+\begin{tabular}{llll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Proof line} & Explanation \\
+\midrule
+\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
+\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
+\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
+\midrule
+\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
+\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
+\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
+\midrule
+\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
+\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
+\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
+\midrule
+\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
+\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
+\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
+\midrule
+\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
+\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
+\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
+\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
+\midrule
+\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
+\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
+\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
+\bottomrule
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+(B) Assertions\\[1.2ex]
+\begin{tabular}{llllll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Proof line} & \multicolumn{2}{l}{Hints} & Explanation \\
+\midrule
+\texttt{25} & \texttt{5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \rightarrow p_5$ \\
+\texttt{26} & \texttt{6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \rightarrow p_6$ \\
+\texttt{27} & \texttt{3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \rightarrow s_7$  \\
+\texttt{28} & \texttt{7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \rightarrow s_7$  \\
+\texttt{29} & \texttt{8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \rightarrow p_8$  \\
+\texttt{30} & \texttt{5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \rightarrow p_5$ \\
+\texttt{31} & \texttt{6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \rightarrow p_6$ \\
+\texttt{32} & \texttt{3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \rightarrow s_7$  \\
+\texttt{33} & \texttt{7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \rightarrow s_7$  \\
+\texttt{34} & \texttt{9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \rightarrow p_9$  \\
+\texttt{35} & \texttt{1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \rightarrow s_{10}$  \\
+\texttt{36} & \texttt{10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
+\bottomrule
+\end{tabular}
+\\[1.2ex]
+(C) Input clause deletions\\[1.2ex]
+\begin{tabular}{lllll}
+  \toprule
+ \multicolumn{3}{l}{Proof line} & Explanation\\
+\midrule
+ \texttt{dc 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
+ \texttt{dc 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
+ \texttt{dc 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
+ \texttt{dc 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
+ \texttt{dc 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
+\end{tabular}
+\end{minipage}
+\caption{Defining Clauses (A), Assertions (B), and Input Clause Deletions (C) for Example Formula}
+\label{fig:eg4:proof}
+\end{figure}
+
+We step through a detailed example to provide a better understanding of the CRAT proof framework.
+Figure
+\ref{fig:eg4:proof}(A) shows the defining clauses that are implicitly
+defined by the POG operation declarations.  These do not appear in the
+CRAT file.  Referring back to the declarations of the sum nodes in
+Figure \ref{fig:eg4:pog}, we can see that the declaration of node
+$s_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
+\ref{fig:eg4:proof}(A) that these two clauses resolve to the clause
+$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $s_7$
+have disjoint models.  Similarly, node $s_{10}$ is declared as having
+clause IDs 16 and 19 as hints.  These resolve to the clause
+$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
+$s_{10}$ have disjoint models.
+
+Figure \ref{fig:eg4:proof}(B) provides the sequence of assertions
+leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $s_{10}$ is implied by the input clauses, i.e.,
+any assignment $\assign$
+satisfying the input clauses must extend to one with $\assign(s_{10}) = 1$.
+Working backward, we can see that
+clause 29 indicates that node $p_8$ will be implied by the input
+clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
+be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
+hints for clause 36.
+
+Figure \ref{fig:eg4:proof}(C) shows the RUP proof steps required to
+delete the input clauses.  Consider the first of these, deleting
+input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
+that there is no assignment $\assign$ that falsifies this clause, but assigns $\assign(s_{10}) = 1$.
+proof proceeds by first assuming that the clause is false, requiring
+$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hints then consist of unit
+clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
+clauses that cause unit propagation.  Hint clauses 8 and 10 force the
+assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 forces
+$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
+via clause 22 to $\assign(s_10) = 0$.  But this contradicts clause 22,
+completing the RUP proof.  The deletion hints for the other input
+clauses follow similar patterns---they work from the bottom nodes of
+the POG upward, showing that any assignment that falsifies the clause
+must assign $s_{10} = 0$.
+
+Deleting the asserted clauses so simple that we do not show it.  It
+involves simply deleting the clauses from clause number 35 down to
+clause number 25, with each deletion using the same hints as were used
+to add that clause.  In the end, therefore, only the defining clauses
+for the POG nodes and the unit clause asserting $s_{10}$ remain,
+completing a proof that the POG is logically equivalent to the input
+formula.
+
+Observe that this proof must ``visit'' nodes
+$p_5$ and $p_6$ twice, separately considering assignments where $\assign(x_1) = 0$
+(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).  In
+general, a naive proof must expand the POG into a tree.
+We address this shortcoming by introducing {\em lemmas}, as is described in Section~\ref{sect:lemma}
+
 \section{Generating CRAT from dDNNF}
+
+\label{sect:lemma}
 

### 2023-03-02 doc/mc-paper.tex
Tweaked figure

@@
 \newcommand{\mulident}{\textbf{1}}
+\newcommand{\imply}{\Rightarrow}
 
@@
 
-%\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
@@
 \begin{figure}
-
-  \centering{
-\begin{minipage}{0.58\textwidth}
+\begin{minipage}{0.62\textwidth}
 (A)  Input Formula\\[1.2ex]
@@
 \toprule
-\makebox[5mm]{ID} & \makebox[15mm]{Literals} & \\
+\makebox[5mm]{ID} & \makebox[15mm]{Clauses} & \\
 \midrule
@@
 \\[1.8ex]
-(B) CRAT Declaration of POG\\[1.2ex]
+(C) POG Declaration\\[1.2ex]
 \begin{tabular}{llll}
 \toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Proof line} & Explanation \\
+\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & Explanation \\
 \midrule
@@
 \end{minipage}
-\begin{minipage}{0.40\textwidth}
-(C) POG Visualization\\
+\begin{minipage}{0.35\textwidth}
+(B) POG Representation \\
 \input{dd/eg4}
 \end{minipage}
-}
-\caption{Example CNF Formula and POG Representation}
-\label{fig:eg4:pog}
-\end{figure}
-
-Figure \ref{fig:eg4:pog} illustrates an example formula and shows how
-the CRAT file declares its POG representation.  The input formula (A)
-consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
-$x_4$.  The generated POG has six nonterminal nodes representing four
-conjunctions and two disjunctions.  We name these by the node type (product or sum) and the extension variable.
-The first part of the CRAT file
-(B) declares these nodes using clause IDs that increment by three or
-four, depending on whether the node has two children or three.
-The last two nonzero values in the sum declarations are hints providing the required mutual exclusion proof.
-
-\begin{figure}
-\begin{minipage}{0.49\textwidth}
-(A) Implicit clauses\\[1.2ex]
+%% Break
+\\[2.5ex]
+\begin{minipage}{0.45\textwidth}
+(D) Implicit Clauses\\[1.2ex]
 \begin{tabular}{llll}
 \toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Proof line} & Explanation \\
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
 \midrule
@@
 \begin{minipage}{0.49\textwidth}
-(B) Assertions\\[1.2ex]
+(E) Assertions\\[1.2ex]
 \begin{tabular}{llllll}
 \toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Proof line} & \multicolumn{2}{l}{Hints} & Explanation \\
+\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & \multicolumn{2}{l}{Hints} & Explanation \\
 \midrule
-\texttt{25} & \texttt{5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \rightarrow p_5$ \\
-\texttt{26} & \texttt{6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \rightarrow p_6$ \\
-\texttt{27} & \texttt{3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \rightarrow s_7$  \\
-\texttt{28} & \texttt{7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \rightarrow s_7$  \\
-\texttt{29} & \texttt{8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \rightarrow p_8$  \\
-\texttt{30} & \texttt{5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \rightarrow p_5$ \\
-\texttt{31} & \texttt{6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \rightarrow p_6$ \\
-\texttt{32} & \texttt{3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \rightarrow s_7$  \\
-\texttt{33} & \texttt{7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \rightarrow s_7$  \\
-\texttt{34} & \texttt{9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \rightarrow p_9$  \\
-\texttt{35} & \texttt{1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \rightarrow s_{10}$  \\
-\texttt{36} & \texttt{10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
+\texttt{25} & \texttt{a 5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
+\texttt{26} & \texttt{a 6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
+\texttt{27} & \texttt{a 3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
+\texttt{28} & \texttt{a 7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
+\texttt{29} & \texttt{a 8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
+\texttt{30} & \texttt{a 5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
+\texttt{31} & \texttt{a 6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
+\texttt{32} & \texttt{a 3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
+\texttt{33} & \texttt{a 7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
+\texttt{34} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
+\texttt{35} & \texttt{a 1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
+\texttt{36} & \texttt{a 10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
 \bottomrule
 \end{tabular}
-\\[1.2ex]
-(C) Input clause deletions\\[1.2ex]
+\\[1.5ex]
+(F) Input Clause Deletions\\[1.2ex]
 \begin{tabular}{lllll}
   \toprule
- \multicolumn{3}{l}{Proof line} & Explanation\\
+ \multicolumn{3}{l}{CRAT line} & Explanation\\
 \midrule
@@
  \texttt{dc 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
+\bottomrule
 \end{tabular}
 \end{minipage}
-\caption{Defining Clauses (A), Assertions (B), and Input Clause Deletions (C) for Example Formula}
-\label{fig:eg4:proof}
+\caption{Example formula (A), its POG representation (B), and its CRAT proof (C)--(F)}
+\label{fig:eg4:all}
 \end{figure}
+
+Figure \ref{fig:eg4:all} illustrates an example formula and shows how
+the CRAT file declares its POG representation.  The input formula (A)
+consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
+$x_4$.  The generated POG (B) has six nonterminal nodes representing four
+conjunctions and two disjunctions.  We name these by the node type (product or sum) and the extension variable.
+The first part of the CRAT file
+(C) declares these nodes using clause IDs that increment by three or
+four, depending on whether the node has two children or three.
+The last two nonzero values in the sum declarations are hints providing the required mutual exclusion proof.
 
@@
 Figure
-\ref{fig:eg4:proof}(A) shows the defining clauses that are implicitly
+\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
 defined by the POG operation declarations.  These do not appear in the
 CRAT file.  Referring back to the declarations of the sum nodes in
-Figure \ref{fig:eg4:pog}, we can see that the declaration of node
+Figure \ref{fig:eg4:all}, we can see that the declaration of node
 $s_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
-\ref{fig:eg4:proof}(A) that these two clauses resolve to the clause
+\ref{fig:eg4:all}(A) that these two clauses resolve to the clause
 $\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $s_7$
@@
 
-Figure \ref{fig:eg4:proof}(B) provides the sequence of assertions
+Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
 leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $s_{10}$ is implied by the input clauses, i.e.,
@@
 
-Figure \ref{fig:eg4:proof}(C) shows the RUP proof steps required to
+Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
 delete the input clauses.  Consider the first of these, deleting

### 2023-03-03 doc/mc-paper.tex
Progress on paper

@@
 
-\newcommand{\pand}{\mathbin{\land^{\sf v}}}
-\newcommand{\por}{\mathbin{\lor^{\sf a}}}
+\newcommand{\pand}{\mathbin{\land^{\sf p}}}
+\newcommand{\por}{\mathbin{\lor^{\sf p}}}
 \DeclareMathOperator*{\Pand}{\bigwedge^{\sf v}}
@@
 \newcommand{\assign}{\alpha}
+\newcommand{\passign}{\rho}
 \newcommand{\uassign}{{\cal U}}
@@
 \newcommand{\outdegree}{\textrm{outdegree}}
+
+\newcommand{\simplify}[2]{#1|_{#2}}
+\newcommand{\expand}{\mathit{Expand}}
+
+\newcommand{\progname}[1]{\textsc{#1}}
+\newcommand{\dfour}{\progname{d4}}
+\newcommand{\Dfour}{\progname{d4}}
+\newcommand{\generator}{\progname{d2p}}
+\newcommand{\Generator}{\progname{d2p}}
+\newcommand{\checker}{\progname{crat-check}}
+\newcommand{\Checker}{\progname{Crat-check}}
 
@@
 We have developed a program that can generate POG representations and
-equivalence proofs from the dDNNF graphs generated by a leading
+equivalence proofs from the d-DNNF graphs generated by a leading
 knowledge compiler.  It scales to all but the largest formulas from a
@@
 
+\begin{figure}
+\centering{
+\begin{tikzpicture}[scale=0.18]
+  \draw (8,10) rectangle (20,18);
+  \node at (14,15.2) {Knowledge};
+  \node at (14,12.8) {Compiler};
+
+  \draw (28,10) rectangle (40,18);
+  \node at (34,15.2) {Proof};
+  \node at (34,12.8) {Generator};
+
+  \draw (48,10) rectangle (60,18);
+  \node at (54,15.2) {Proof};
+  \node at (54,12.8) {Checker};
+
+  \draw (48,0) rectangle (60,8);
+  \node at (54,5.2) {Ring};
+  \node at (54,2.8) {Evaluator};
+
+  %% CNF
+  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (48,16);
+  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
+  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
+  \node [left] at (-1,20) {\large {$\phi$}};
+  \node [above] at (2,20) {\texttt{.cnf}};
+
+  %% NNF
+  \draw[thick] (20,12) [-latex] -- (28,12);
+  \node [above] at (24,12) {\texttt{.nnf}};
+
+  %% CRAT
+  \draw[thick] (40,12) [-latex] -- (48,12);
+  \draw[thick] (44,12) -- (44,4) [-latex] -- (48,4);
+  \node [above] at (44,12) {\texttt{.crat}};
+
+  %% OK/Not
+  \draw[thick] (60,14) [-latex] -- (64,14) ;
+  \node [right] at (64,15.2) { OK /} ;
+  \node [right] at (64,12.8) { Not OK} ;
+  \draw[thick] (60,4) [-latex] -- (64,4) ;
+  \node [right] at (64,4) {\large {$\textbf{R}(\phi)$}};
+
+\end{tikzpicture}
+}
+\caption{Tool Chain. The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT) that can be independently checked and evaluated}
+\label{fig:chain}
+\end{figure}
+
 \section{Logical Foundations}
@@
   Let $\varset$ denote a set of Boolean variables, and let $\assign$
-  be an {\em assignment} of truth values to the variables, where $0$
-  denotes false and $1$ denotes true, i.e., $\assign \colon \varset
-  \rightarrow \{0,1\}$.  The set of all possible assignments over
-  $\varset$ is denoted $\uassign$.  For each variable $x \in \varset$,
+  be an {\em assignment} of truth values to some subset of the
+  variables, where $0$ denotes false and $1$ denotes true, i.e.,
+  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
+  \subseteq \varset$.  We say the assignment is {\em total} when it
+  assigns a value to every variable ($\varset' = \varset$), and that
+  it is {\em partial} otherwise.
+  The set of all possible total assignments over
+  $\varset$ is denoted $\uassign$.
+
+For each variable $x \in \varset$,
   we define {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
@@
 
+\begin{definition}[Boolean Formulas]
   The set of Boolean formulas is defined recursively.  Each
@@
   $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
-  consisting of assignments that satisfy the formula:
+  consisting of total assignments that satisfy the formula:
   \begin{enumerate}
@@
     with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
-  \item A variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
+  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
     and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
@@
   \end{enumerate}
+\label{def:boolean}
+\end{definition}
 
   We highlight the following special classes of Boolean formulas:
-  \begin{itemize}
-    \item In a {\em normal form} formula, negation is 
-      applied only to variables
-    \item A normal-form formula is in {\em conjunctive normal form} (CNF) if disjunction is applied only to literals.
-    \item A {\em partitioned-operation formula} (or simply a ``partitioned formula'') satisfies the following restrictions on its conjunction and disjunction operations:
+  \begin{description}
+    \item[Normal form:] A formula where negation is applied only to variables.
+    \item[Conjunctive Normal Form (CNF):] A normal form formula where  disjunction is applied only to literals.
+    \item[Partitioned-Operation:] A formula satisfying the following for all conjunction and disjunction operations: 
       \begin{itemize}
       \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
-        $\phi = \bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
+        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
       \item The arguments to a disjunction must have disjoint models.  That is, operation
-        $\phi = \bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
+        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
       \end{itemize}
-     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned formula.
-  \end{itemize}
+  \end{description}
+     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned-operation formula.
+     A formula in conjunctive normal form can be represented by a set
+     of {\em clauses}, each of which is a set of literals.  Each
+     clause then represents the disjunction of the literals, and the formula is then the conjunction of its clauses.
 
@@
 
-We propose a general framework for describing a variety of tasks that
-summarize some aspect of Boolean formulas.
+We propose a general framework for 
+summarizing properties of Boolean formulas.
 
@@
 \end{definition}
+We write $a - b$ as a shorthand for $a + -b$.
 
 \begin{definition}[Ring Evaluation Problem (REP)]
-  A {\em ring labeling} associates a value $w(x) \in \dset$ with
-  every variable $x \in \varset$.  We define $w(\obar{x})$ to equal $1-w(x)$.
+  For ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
+  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $1-w(x)$.
 
@@
   \begin{eqnarray}
-    \rep(\phi) & = & \sum_{\alpha \in \modelset(\phi)} \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
+    \rep(\phi) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
   \end{eqnarray}
@@
 expressed as ring evaluation problems.  The 
-{\em model counting} problem can be cast as an REP by having $\radd$ and
+{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$
+It can be cast as an REP by having $\radd$ and
 $\rmul$ be addition and multiplication over rational numbers and using
@@
 Ring evaluation of formula $\phi$ will then compute the {\em density} of
-the formula, i.e., the fraction of all possible assignments that are
+the formula, i.e., the fraction of all possible total assignments that are
 models.  For $n = |\varset|$, scaling the density by $2^n$ will then
@@
 The {\em weighted model counting} (WMC) problem is also defined over
-rational numbers.  Some formulations of WMC allow independent
-assignments of weights $W$ to the variables and their negations, such
-that $W(x) + W(\obar{x}) \not = 1$.  We can convert this into a
+rational numbers.  Some formulations of WMC allow 
+assigning weights $W$ to both the variables and their negations, such
+that there can be variables $x$ with $W(x) + W(\obar{x}) \not = 1$.  We can convert this more general form into a
 ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
-performing algebraic evaluation with label $w(x) = W(x)/r(x)$ for each
+performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
 variable $x$, and then computing the final result as $\rep(\phi)\;
-\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$.
+\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.
 
-The {\em function hashing problem} provides a test
-of inequivalence for Boolean formulas.  That is, let $\ring$ be a
-finite  field with $|\dset| \geq 2 |\varset|$.  Two formulas
+The {\em function hashing problem} provides a randomized test
+of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
+finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
 $\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1) = \rep(\phi_2)$
@@
 the probability that $\rep(\phi_1) \not = \rep(\phi_2)$ will be at
-least $\left(1-\frac{1}{z}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
+least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
 Therefore, ring evaluation can be used as part of a
@@
 
+Possible generalizations of this framework are discussed in Section~\ref{sect:future}.
+
 \section{Partitioned-Operation Graphs (POGs)}
 
-Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a partitioned formula:
+Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
 \begin{proposition}
@@
 
-As can be seen, a POG is simply a way to represent a partitioned
-formula with a sharing of common subformulas, and so we write
-$\phi(P)$ as the formula denoted by $P$.
+As can be seen, a POG is simply a way to represent a partitioned-operation
+formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
+$\phi(u)$ as the formula denoted by node $u$ and write $\phi(P) = \phi(r)$ for root $r$ of $P$.
 
@@
 
-\item It allows arbitrary arguments to a $\por$ operation, as long as
+\item They allow arbitrary arguments to a $\por$ operation, as long as
   they have disjoint models.  By contrast, each
-  disjunction node in a d-DNNF can have at most two children, and there must be a variable $x$ such that
-  the satisfying assignments for one child have $x=1$ while those for the
-  other child have $x=0$.  Our generalization is provided to allow an
+  disjunction node $v$ in a d-DNNF can have at most two children $u_1$ and $u_0$ for which there must be a {\em splitting variable} $x$ such that
+  any total assignment $\assign$ satisfying $\phi(u_1)$ has $\assign(x)=1$, while any total assignment satisfying $\phi(u_0)$ has
+  $\assign(x)=0$.  Our generalization is provided to allow an
   encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
@@
 
-\section{The CRAT Proof System}
+\section{The CRAT POG Representation and Proof System}
 
-The CRAT clausal proof framework provides a means to express a
-checkable proof that a Boolean formula, given in conjunctive normal
-form, is logically equivalent to a POG\@, thus enabling certified
-knowledge compilation.  It is based on extended
+A CRAT file provides both a declaration of a POG, as well as a checkable, clausal
+proof that a Boolean formula, given in conjunctive normal
+form, is logically equivalent to the POG, thus enabling certified
+knowledge compilation.  The proof system is based on extended
 resolution~\cite{Tseitin:1983} with reverse unit propagation
-(RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as its core method for
-determining that a clause is implication-redundant with respect to other clauses.
+(RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
+proving that a set of clauses logically implies another clause.
 
-Two programs are involved in producing a certified compilation of a formula:
+As is shown in Figure~\ref{fig:chain}
+two programs are involved in producing a certified compilation of a formula:
 \begin{itemize}
-\item The {\em compiler} generates a CRAT file that both describes a POG and provides a proof that this POG is equivalent to the input formula.
+\item The {\em proof generator} generates a CRAT file that both describes a POG and provides a proof that this POG is equivalent to the input formula.
   This compiler might start with the output of another knowledge compiler.
-\item The {\em checker} ensures that all of the proof conditions are satisfied.  
+\item The {\em proof checker} ensures that all of the proof conditions are satisfied.  
 \end{itemize}
@@
   \item
-  In addition to explicit clause additions and deletions, the proof contains declarations of $\pand$ and $\por$ operations.
-\begin{itemize}
+  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
+  Each such declaration implicitly adds an extension variable representing the POG node and a set of {\em defining} clauses
+  encoding the conjunction or disjunction operation.
+  This is the only means for adding extension variables to the proof.
+\item Boolean negation is supported implicitly by allowing the
+  arguments of the $\por$ and $\pand$ operations to be literals and not just
+  variables.
 \item
-  Each such declaration implicitly adds an extension variable representing that operation and a set of {\em defining} clauses
-  encoding the conjunction or disjunction operation.
-  This is the only means for generating extension variables or adding blocked clauses to the proof.
+  The file contains explicit clause addition steps.
+  A clause can only be added if it is logically implied by the existing clauses.
+  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
+\item
+  The file contains explicit clause deletion steps.
+  A clause can only be deleted if it is logically implied by the remaining clauses.
+  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
 \item The checker must track the dependency set for every input and
-  extension variable.  When an extension variable is introduced based
-  on the $\pand$ operation, the dependency sets of its arguments must
-  be disjoint.  The resulting extension variable has a dependency set equal to the union of those of its arguments.
+  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
+  The associated extension variable has a dependency set equal to the union of those of its arguments.
 \item Declaring a $\por$ operation requires a sequence of clauses
@@
   $k\,(k-1)/2$ proofs of disjointness for larger values of $k$.
-\item Boolean negation is provided implicitly by allowing the
-  arguments of the $\por$ and $\pand$ operations to be literals and not just
-  variables.
-\end{itemize}
-\item
-  Assertion clauses can can added when they satisfy the RUP
-  property, with respect to a sequence of existing clauses.  This set must be provided explicitly as the {\em hints} for the assertion.
-\item
-  Deleting clauses requires proving that the resulting formula is not weaker.
-  That is, it should not allow any additional satisfying assignment.
-  Only the input clauses and those declared by an assertion can be deleted.
-      The deletion must be accompanied by a sequence of remaining clauses
-      providing a RUP proof of the clause.  This sequence must be provided as the {\em hints} for the deletion.
 \end{itemize}
@@
 $\exvarset$ denote the set of added extension variables,  any
-assignment $\assign\colon \varset \rightarrow \{0,1\}$ has a unique extension,
+total assignment $\assign\colon \varset \rightarrow \{0,1\}$ has a unique extension,
 which we also call $\assign$, with $\assign\colon (\varset \cup \exvarset) \rightarrow \{0,1\}$.
@@
 steps that both add and delete clauses.  Each addition must be truth
-preserving, that is, any satisfying assignment for the set of clauses
+preserving, that is, any satisfying total assignment for the set of clauses
 before clause addition should still be a satisfying assignment afterwards.
@@
 The sequence of clause addition steps provide a proof that any
-assignment $\assign$ to the input variables that satisfies all of the
+total assignment $\assign$ to the input variables that satisfies all of the
 clauses must, when extended, assign $\assign(r) = 1$.  Conversely,
 each proof step that deletes an input clause proves that any
-assignment $\alpha$ that falsifies the clause must, when extended,
+total assignment $\alpha$ that falsifies the clause must, when extended,
 assign $\assign(r) = 0$.
@@
 \begin{minipage}{0.49\textwidth}
-(E) Assertions\\[1.2ex]
+(E) CRAT Assertions\\[1.2ex]
 \begin{tabular}{llllll}
 \toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & \multicolumn{2}{l}{Hints} & Explanation \\
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hints} & Explanation \\
 \midrule
@@
 leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $s_{10}$ is implied by the input clauses, i.e.,
-any assignment $\assign$
+any total assignment $\assign$
 satisfying the input clauses must extend to one with $\assign(s_{10}) = 1$.
@@
 input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
-that there is no assignment $\assign$ that falsifies this clause, but assigns $\assign(s_{10}) = 1$.
+that there is no total assignment $\assign$ that falsifies this clause, but assigns $\assign(s_{10}) = 1$.
 proof proceeds by first assuming that the clause is false, requiring
@@
 $\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
-via clause 22 to $\assign(s_10) = 0$.  But this contradicts clause 22,
+via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 22,
 completing the RUP proof.  The deletion hints for the other input
 clauses follow similar patterns---they work from the bottom nodes of
-the POG upward, showing that any assignment that falsifies the clause
+the POG upward, showing that any total assignment that falsifies the clause
 must assign $s_{10} = 0$.
@@
 Observe that this proof must ``visit'' nodes
-$p_5$ and $p_6$ twice, separately considering assignments where $\assign(x_1) = 0$
+$p_5$ and $p_6$ twice, separately considering total assignments where $\assign(x_1) = 0$
 (clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).  In
@@
 
-\section{Generating CRAT from dDNNF}
+\section{Generating CRAT from d-DNNF}
+
+Figure~\ref{fig:chain} illustrates one method of generating a CRAT
+file from an input formula in conjunctive normal form.  Our tool chain
+is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
+but it would also work for other tools that use a top-down search
+strategy to generate a d-DDNNF representation.  Converting a d-DNNF
+graph into a POG is straightforward, each node in a d-DNNF graph can
+be directly represented as a POG node.  We perform simplifications on
+the d-DNNF representation to eliminate any Boolean constants.  We can
+therefore assume that the POG contains only conjunction, disjunction,
+and literal nodes.
+
+\section{CNF Implies POG}
+
+The most challenging part of the proof is to show that any total
+assignment $\alpha$ that is a model of $\phi$ will, when extended,
+have $\assign(r) = 1$ for root literal
+$r$.  This part of the proof consists of a series of assertion steps
+leading to one adding $r$ as a unit clause.
+
+In the following, we describe proof generation as a process that
+recursively traverses the graph from the root downward and
+generates proof steps from the lower nodes upward.  This process
+effectively expands the POG into a tree and therefore can be very
+inefficient.  We eliminate this inefficiency with the addition of
+lemmas, as is described later.
+
+Assume that formula $\phi$ is 
+represented as a set of clauses.
+For partial assignment
+$\passign$, the expression  $\simplify{\phi}{\passign}$ denotes the
+reduced and simplified set of clauses obtained by 1) eliminating any
+clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
+2) for the remaining clauses eliminating those literals $\lit$ for
+which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
+In doing these simplifications, we also track the {\em provenance}
+of each simplified clause $C$, i.e., which of the original clauses simplified to become $C$.
+More formally, for $C \in \simplify{\phi}{\passign}$, we let $\expand(C)$ denote
+those clauses $C' \in \phi$, such that
+$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
+We then extend the definition of $\expand$ to any simplified formula
+$\psi$ as $\expand(\psi) = \bigcup_{C \in \psi} \expand(C)$.
+
+In following a path from the root node downward to a node $u$, the
+literals encountered in the conjunction nodes (and possibly a terminal
+literal) define a partial assignment $\passign$.  The subgraph with
+root $u$ should be a POG representation of the reduced formula
+$\simplify{\phi}{\passign}$.  Our recursive goal is to generate
+a sequence of proof steps finishing with the addition of the {\em target clause}
+$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating that any total
+assignment $\assign$ such that $\passign \subseteq \assign$ will, when extended, yield
+$\alpha(u) = 1$.
+
+The process for generating such a proof depends on the form of node $u$:
+\begin{enumerate}
+\item if $u$ is a literal $\lit$, then the formula
+  $simplify{\phi}{\passign}$ must consist of the single unit clause
+  $C = \lit$, such that $\expand{C}
+  this clause.  That means that every other literal $\lit' \in C$ must
+  have $\passign(\lit') = 0$, and therefore this input clause can
+  serve as the target clause.
+\item If $u$ is a disjunction node with children $u_1$ and $u_0$,
+  then, since the node originated from a d-DNNF graph, there must be
+  some variable $x$ such that either $u_1 = x$ or $u_1$ is a
+  conjunction node containing $x$ as a child.  In either case, we
+  recurse to node $u_1$ with the partial assignment $\passign \cup \{ x \}$.
+  This will yield a sequence of proof steps
+  $u \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{l}$.
+Similarly, either $u_0  = \obar{x}$ or $u_0$ is a conjunction node containing $\obar{x}$ as
+  a child.  In either case, we recurse to node $u_0$
+ with the partial assignment $\passign \cup \{ \obar{x} \}$.
+  This will yield a sequence of proof steps
+  $u \lor x \lor \bigvee_{\lit \in \passign} \obar{l}$.
+  These recursive results can be combined, via two proof steps, with second and third defining clauses for $u$
+  to generate the target clause for $u$.
+\item If $u$ is a conjunction node, then we can divide its children
+  into literal nodes and nonterminal nodes.
+  \begin{enumerate}
+    \item For each literal child
+  $\lit$, we must prove that any total assignment $\alpha$, such that
+  $\passign \subset \alpha$ must have $\alpha(\lit) = 1$.  In some
+  cases, this can be done by simple Boolean constraint propagation.
+  In other cases, we must prove that the formula
+  $\simplify{\phi}{\passign \cap \{\obar{l}\}} is unsatisfiable.  We
+  do so by writing the formula to a file, invoking a proof-generating
+  SAT solver, and then importing the generated unsatisfiability prove
+  as assertions in the CRAT file.
+\item if there are multiple nonterminal children $u_1, u_2, \ldots,
+  u_k$, it must be the case that the clauses in
+  $\simplify{\phi}{\passign}$ can be partitioned into $k$ subsets
+  $\psi_1, \psi_2, \ldots, \psi_k$ such that $\dependencyset(\psi_i)
+  \cap \dependencyset(\psi_j) = \emptyset$ for $1 \leq i < j \leq k$,
+  and we can match each node $u_i$ to subset $\psi_i$ based on its
+  literals.  Moreover, each clause $C$ in each subset $\psi_i$ must be
+  the simplified version of one or more input clauses $C'$ in the
+  original formula $\phi$.  We define $\phi_i$ to consist of all such
+  clauses $C'$ for each $C \in \psi_i$.
+
+We do so using a union-find algorithm and match the variables in each partition
+    $\phi_i$ to the POG node $u_i$ containing literals of the variables in
+    $\dependencyset(\phi_i)$.
+    We then recursively generate target clauses for nodes $u_1, u_2, \ldots, u_k$
+
+\end{itemize}
+
+\end{enumerate}
 

### 2023-03-04 doc/mc-paper.tex
Incremental progress

@@
 \newcommand{\outdegree}{\textrm{outdegree}}
+\newcommand{\validate}{\textsf{validate}}
 
 \newcommand{\simplify}[2]{#1|_{#2}}
-\newcommand{\expand}{\mathit{Expand}}
+\newcommand{\prov}{\mathit{Expand}}
 
@@
 In doing these simplifications, we also track the {\em provenance}
-of each simplified clause $C$, i.e., which of the original clauses simplified to become $C$.
-More formally, for $C \in \simplify{\phi}{\passign}$, we let $\expand(C)$ denote
+of each simplified clause $C$, i.e., which of the (possibly multiple) original clauses simplified to become $C$.
+More formally, for $C \in \simplify{\phi}{\passign}$, we let $\prov(C)$ denote
 those clauses $C' \in \phi$, such that
 $C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
-We then extend the definition of $\expand$ to any simplified formula
-$\psi$ as $\expand(\psi) = \bigcup_{C \in \psi} \expand(C)$.
+We then extend the definition of $\prov$ to any simplified formula
+$\psi$ as $\prov(\psi) = \bigcup_{C \in \psi} \prov(C)$.
+
+In the following we define a recursive routine $\validate(u, \passign,
+\psi)$ taking as arguments 1) POG node $u$, 2) partial assignment
+$\passign$, and 3) a set of clauses $\psi \subsetq \phi$.  The outcome of its execution should be to add a number of assertions culminating with
+the addition of the {\em target clause}
+$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating that any total
+assignment $\assign$ such that $\passign \subseteq \assign$
+will, when extended to extension variables, will assign $\assign(u) = 1$.
+The top-level call has $u = r$, $\passign = \emptyset$, and $\psi = \phi$.  The result will therefore be to add the unit clause $r$ to the proof.
 
@@
   $simplify{\phi}{\passign}$ must consist of the single unit clause
-  $C = \lit$, such that $\expand{C}
-  this clause.  That means that every other literal $\lit' \in C$ must
-  have $\passign(\lit') = 0$, and therefore this input clause can
+  $C = \lit$, such that any $C' \in \prov{C}$ must have $C' \subseteq \{ \lit \} \bigcup_{\lit \in \passign} \obar{\lit}$.
+  Any of these can
   serve as the target clause.
@@
   This will yield a sequence of proof steps
-  $u \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{l}$.
+  $u \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
 Similarly, either $u_0  = \obar{x}$ or $u_0$ is a conjunction node containing $\obar{x}$ as
@@
   This will yield a sequence of proof steps
-  $u \lor x \lor \bigvee_{\lit \in \passign} \obar{l}$.
+  $u \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
   These recursive results can be combined, via two proof steps, with second and third defining clauses for $u$
@@
 \item If $u$ is a conjunction node, then we can divide its children
-  into literal nodes and nonterminal nodes.
+  into a set of literal nodes $L$ and a set of nonterminal nodes $u_1, u_2, \ldots, u_k$.
   \begin{enumerate}
-    \item For each literal child
-  $\lit$, we must prove that any total assignment $\alpha$, such that
+    \item For each literal 
+  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
   $\passign \subset \alpha$ must have $\alpha(\lit) = 1$.  In some
-  cases, this can be done by simple Boolean constraint propagation.
+  cases, this can be done by simple Boolean constraint propagation (BCP).
   In other cases, we must prove that the formula
-  $\simplify{\phi}{\passign \cap \{\obar{l}\}} is unsatisfiable.  We
+  $\simplify{\phi}{\passign \cap \{\obar{\lit}\}}$ is unsatisfiable.  We
   do so by writing the formula to a file, invoking a proof-generating
@@
   as assertions in the CRAT file.
-\item if there are multiple nonterminal children $u_1, u_2, \ldots,
-  u_k$, it must be the case that the clauses in
+\item If there is a single nonterminal child ($k = 1$), we can recurse on node $u_1$
+  with partial assignment $\passign \cup \bigcup_{\lit \in L} \lit$.
+\item If there are multiple nonterminal children ($k > 1$)
+  it must be the case that the clauses in
   $\simplify{\phi}{\passign}$ can be partitioned into $k$ subsets
@@
   and we can match each node $u_i$ to subset $\psi_i$ based on its
-  literals.  Moreover, each clause $C$ in each subset $\psi_i$ must be
-  the simplified version of one or more input clauses $C'$ in the
-  original formula $\phi$.  We define $\phi_i$ to consist of all such
-  clauses $C'$ for each $C \in \psi_i$.
-
-We do so using a union-find algorithm and match the variables in each partition
-    $\phi_i$ to the POG node $u_i$ containing literals of the variables in
-    $\dependencyset(\phi_i)$.
-    We then recursively generate target clauses for nodes $u_1, u_2, \ldots, u_k$
-
-\end{itemize}
-
+  literals.
+  For each $i$ such that $1 \leq i \leq k$, we can then recurse on node $u_i$
+  with partial assignment $\passign \cup \bigcup_{\lit \in L} \lit$, and based on the input clauses in $\prov(\psi_i)$.
+\end{enumerate}
+  We then combine as hints the results from the BCP and SAT calls for
+  the literals, the recursively computed target clauses, and all but
+  the first defining clause for the conjunction operation to generate the target clause for node $u$.
 \end{enumerate}

### 2023-03-07 tools/merge_csv.py
Make more robust and able to merge multicolumn files

@@ -23,15 +23,16 @@ def usage(name):
     eprint("  -l LABELS     Provide comma-separated set of heading labels")
     eprint("  FILE1.csv ... Source files")
 
-# List of keys in order first encountered
-keys = []
-# Growing list of result lines
-lines = []
-# Set of keys with missing entries
-missingKeys = set([])
+# Growing set of result lines, indexed by key
+globalEntries = {}
+# Number of columns in entries
+globalCount = 0
 
-def addData(fname, first = False):
-    global keys, lines, missingKeys
+# Process a single file, building entries
+# Return dictionary of entries + count of number of entries per line
+def processFile(fname):
+    entries = {}
+    columnCount = 0
     try:
         infile = open(fname)
         creader = csv.reader(infile)
@@ -41,74 +42,77 @@ def addData(fname, first = False):
     row = 0
     for fields in creader:
         row += 1
-        if len(fields) != 2:
-            eprint("Error in file %s, row %d.  Can only handle CSV files with two items per row" % (fname, row))
-            sys.exit(1)
         key = fields[0]
-        val = fields[1]
-        if first:
-            keys.append(key)
-            lines.append(key + "," + val)
-        else:
-            if row > len(keys):
-                eprint("File %s, row %d.  Too many entries" % (fname, row))
-                sys.exit(1)
-            while row <= len(keys) and keys[row-1] != key:
-                missingKeys.add(keys[row-1])
-                lines[row-1] += ","
-                row += 1
-            if row > len(keys):
-                eprint("File %s, row %d.  Couldn't find key '%s'" % (fname, row, key))
-                sys.exit(1)                
-            lines[row-1] += "," + val
-    row += 1
-    while row <= len(lines):
-        lines[row-1] += ","
-        missingKeys.add(keys[row-1])
-        row += 1
+        entry = fields[1:]
+        dcount = len(entry)
+        if columnCount == 0:
+            columnCount = dcount
+        elif dcount != columnCount:
+            eprintf("File %s, row %d.  Expecting %d entries.  Found %d" % (fname, row, columnCount, dcount))
+            sys.exit(1)
+        entries[key] = entry
     infile.close()
-        
-
-def filter():
-    global keys, lines
-    okeys = keys
-    olines = lines
-    keys = []
-    lines = []
-    for (key,line) in zip(okeys, olines):
-        if key not in missingKeys:
-            keys.append(key)
-            lines.append(line)
[…]

### 2023-03-07 doc/mc-paper.tex
Updated

@@
 
+\newcommand{\makenode}[1]{\mathbf{#1}}
+\newcommand{\nodeu}{\makenode{u}}
+\newcommand{\nodev}{\makenode{v}}
+\newcommand{\nodes}{\makenode{s}}
+\newcommand{\nodep}{\makenode{p}}
+\newcommand{\noder}{\makenode{r}}
+
 \newcommand{\simplify}[2]{#1|_{#2}}
-\newcommand{\prov}{\mathit{Expand}}
+\newcommand{\prov}{\mathit{Prov}}
 
@@
 \begin{abstract}
-Knowledge compilers convert Boolean formulas into a form for which
-useful tasks, including model counting and weighted model counting,
-can easily be performed.  Existing knowledge compilers, however,
+  Knowledge compilers convert Boolean formulas into a form 
+  that enables efficient evaluation of properties of the formula,
+  including
+  weighted and unweighted model counting.
+ Existing knowledge compilers, however,
 provide no guarantee that the generated representation accurately
@@
 equivalence proofs from the d-DNNF graphs generated by a leading
-knowledge compiler.  It scales to all but the largest formulas from a
-recent model counting competition.  We also present two proof
+knowledge compiler.  We also present two proof
 checkers---a fast checker that can handle the large proofs generated,
 and a formally verified checker written in the \lean{} proof system.
+Our generation and checking tool chain scales to all but the largest formulas from a
+recent model counting competition.  
 \end{abstract}
@@
 \section{Introduction}
+
+\section{Logical Foundations}
+
+  Let $\varset$ denote a set of Boolean variables, and let $\assign$
+  be an {\em assignment} of truth values to some subset of the
+  variables, where $0$ denotes false and $1$ denotes true, i.e.,
+  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
+  \subseteq \varset$.  We say the assignment is {\em total} when it
+  assigns a value to every variable ($\varset' = \varset$), and that
+  it is {\em partial} otherwise.
+  The set of all possible total assignments over
+  $\varset$ is denoted $\uassign$.
+
+For each variable $x \in \varset$,
+  we define the {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
+  negation of $x$. An
+  assignment $\assign$ can be viewed as a set of literals, where
+  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
+  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
+$\obar{\lit} = x$ when $\lit = \obar{x}$.
+
+\begin{definition}[Boolean Formulas]
+  The set of Boolean formulas is defined recursively.  Each
+  formula $\phi$ has an associated {\em dependency set}
+  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
+  consisting of total assignments that satisfy the formula:
+  \begin{enumerate}
+  \item Boolean constants $0$ and $1$ are Boolean formulas,
+    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
+  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
+    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
+  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
+    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em conjunction} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
+      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
+      $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em disjunction} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
+      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
+      $\modelset(\phi) = \bigcup_{i=1,k} \modelset(\phi_i)$.
+  \end{enumerate}
+\label{def:boolean}
+\end{definition}
+
+  We highlight the following special classes of Boolean formulas:
+  \begin{description}
+    \item[Normal form:] A formula where negation is applied only to variables.
+    \item[Conjunctive Normal Form (CNF):] A normal form formula where  disjunction is applied only to literals.
+    \item[Partitioned-Operation:] A formula satisfying the following for all conjunction and disjunction operations: 
+      \begin{itemize}
+      \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
+        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
+      \item The arguments to a disjunction must have disjoint models.  That is, operation
+        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
+      \end{itemize}
+  \end{description}
+     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned-operation formula.
+     A formula in conjunctive normal form can be represented by a set
+     of {\em clauses}, each of which is a set of literals.  Each
+     clause represents the disjunction of the literals, and the formula is the conjunction of its clauses.
+
+  \section{Ring Evaluation of a Boolean Formula}
+
+We propose a general framework for 
+summarizing properties of Boolean formulas.
+
+\begin{definition}[Commutative Ring]
+  A {\em commutative ring} $\ring$ is an algebraic structure
+  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$, 
+  with elements in the set $\dset$ and with commutative and
+  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
+  such that multiplication distributes
+  over addition.  $\addident$ is the additive identity and $\mulident$ is
+  the multiplicative identity.  Every element $a \in \dset$ has an
+  {\em additive inverse} $-a$ such that $a + -a = \addident$.
+\label{def:ring}
+\end{definition}
+We write $a - b$ as a shorthand for $a + -b$.
+
+\begin{definition}[Ring Evaluation Problem (REP)]
+  For ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
+  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $1-w(x)$.
+
+  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} (REP) is to compute:
+  \begin{eqnarray}
+    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
+  \end{eqnarray}
+  In this equation, the summation $\sum$ is computed using operation $\radd$, and the product $\prod$ is computed using operation $\rmul$.
+\label{def:labeling}
+\end{definition}
+
+A wide variety of important tasks for Boolean functions can be
+expressed as ring evaluation problems.  The 
+{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
+It can be cast as a REP by having $\radd$ and
+$\rmul$ be addition and multiplication over rational numbers and using
+label $w(x) = 1/2$ for every variable $x$.  
+Ring evaluation of formula $\phi$ will then compute the {\em density} of
+the formula, i.e., the fraction of all possible total assignments that are
+models.  For $n = |\varset|$, scaling the density by $2^n$ will then
+yield the number of models.
+
+The {\em weighted model counting} (WMC) problem is also defined over
+rational numbers.  Some formulations of WMC allow 
+indepdently assigning weights $W(x)$ and $W(\obar{x})$ to each variable $x$, potentially having
+$W(x) + W(\obar{x}) \not = 1$.
+We can convert this more general form into a
+ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
+performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
+variable $x$, and then computing the final result as $\rep(\phi, w)\;
+\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.
+
+The {\em function hashing problem} provides a randomized test
+of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
+finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
+$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
+if they are logically equivalent.  If they are not equivalent, then
+the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
+least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
+Therefore, ring evaluation can be used as part of a
+randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
+
+Generalizations of this framework are discussed in Section~\ref{sect:future}.
+
+\section{Partitioned-Operation Graphs (POGs)}
+
+Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
+\begin{proposition}
+Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
+\begin{description}
+\item[Negation] $\rep(\boolnot \phi, w) = \mulident - \rep(\phi, w)$
+\item[Partitioned Conjunction] $\rep\left(\Pand_{i=1,k} \phi_i, w \right) = \prod_{i=1,k} \rep(\phi_i, w)$
+\item[Partitioned Disjunction] $\rep\left(\Por_{i=1,k} \phi_i, w\right) = \sum_{i=1,k} \rep(\phi_i, w)$
+\end{description}
+\end{proposition}
+
+A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
+with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
+When $(\nodeu,\nodev) \in E$,
+node $\nodev$ is said to be a {\em child} of node $\nodeu$.
+The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
+$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
+Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
+Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.  A nonterminal node is labeled by one of three Boolean operations:
+$\boolnot$, $\pand$, or $\por$.  Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
+It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
+There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.
+
+As can be seen, a POG is simply a way to represent a partitioned-operation
+formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
+$\phi(\nodeu)$ as the formula denoted by node $\nodeu$.
+Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi(\nodeu))$.
+
+We define the {\em size} of POG $P$, written $|P|$ to be the sum of
+the number of nonterminal nodes and the number of edges.  Ring
+evaluation of $P$ can be performed with at most $|P|$ ring
+operations by traversing the graph from the terminal nodes up to
+the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
+The final result is then $\rep(\noder, w)$.
+
+POGs are inspired by the d-DNNF graphs devised by Darwiche for
+representing Boolean functions \cite{darwiche:jair:2002}.
+POGs generalize d-DNNF in two ways:
+\begin{itemize}
+\item They allow negation of arbitrary nodes in the graph, not just
+  variables.  As we have seen, ring evaluation can readily handle negation, and so we allow it in the interest of generality.
+  Negation could be useful in some contexts, such as when
+  converting binary decision diagrams (BDDs) having complemented
+  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
+
+\item They allow arbitrary arguments to a $\por$ operation, as long as
+  they have disjoint models.  By contrast, each
+  disjunction node $\nodev$ in a d-DNNF can have at most two children $\nodeu_1$ and $\nodeu_0$ for which there must be a {\em decision variable} $x$ such that
+  any total assignment $\assign \in  \modelset(u_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
+Our generalization is provided to allow an
+  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
+  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
+\end{itemize}
+  Both of these generalizations allow more flexibility in the POG
+  representation while maintaining the ability to efficiently perform ring evaluation.
+
+\section{The CRAT POG Representation and Proof System}
 
@@
 
-\section{Logical Foundations}
-
-  Let $\varset$ denote a set of Boolean variables, and let $\assign$
-  be an {\em assignment} of truth values to some subset of the
-  variables, where $0$ denotes false and $1$ denotes true, i.e.,
-  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
-  \subseteq \varset$.  We say the assignment is {\em total} when it
-  assigns a value to every variable ($\varset' = \varset$), and that
-  it is {\em partial} otherwise.
-  The set of all possible total assignments over
-  $\varset$ is denoted $\uassign$.
-
-For each variable $x \in \varset$,
-  we define {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
-  negation of $x$. An
-  assignment $\assign$ can be viewed as a set of literals, where
-  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
-  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
-$\obar{\lit} = x$ when $\lit = \obar{x}$.
-
-\begin{definition}[Boolean Formulas]
-  The set of Boolean formulas is defined recursively.  Each
-  formula $\phi$ has an associated {\em dependency set}
-  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
-  consisting of total assignments that satisfy the formula:
-  \begin{enumerate}
-  \item Boolean constants $0$ and $1$ are Boolean formulas,
-    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
-  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
-    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
-  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
-    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em conjunction} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em disjunction} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcup_{i=1,k} \modelset(\phi_i)$.
-  \end{enumerate}
-\label{def:boolean}
-\end{definition}
-
-  We highlight the following special classes of Boolean formulas:
-  \begin{description}
-    \item[Normal form:] A formula where negation is applied only to variables.
-    \item[Conjunctive Normal Form (CNF):] A normal form formula where  disjunction is applied only to literals.
-    \item[Partitioned-Operation:] A formula satisfying the following for all conjunction and disjunction operations: 
-      \begin{itemize}
-      \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
-        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \item The arguments to a disjunction must have disjoint models.  That is, operation
-        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \end{itemize}
-  \end{description}
-     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned-operation formula.
-     A formula in conjunctive normal form can be represented by a set
-     of {\em clauses}, each of which is a set of literals.  Each
-     clause then represents the disjunction of the literals, and the formula is then the conjunction of its clauses.
-
-  \section{Ring Evaluation of a Boolean Formula}
-
-We propose a general framework for 
-summarizing properties of Boolean formulas.
-
-\begin{definition}[Commutative Ring]
-  A {\em commutative ring} $\ring$ is an algebraic structure
-  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$, 
-  with elements in the set $\dset$ and with commutative and
-  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
-  such that multiplication distributes
-  over addition.  $\addident$ is the additive identity and $\mulident$ is
-  the multiplicative identity.  Every element $a \in \dset$ has an
-  {\em additive inverse} $-a$ such that $a + -a = \addident$.
-\label{def:ring}
-\end{definition}
-We write $a - b$ as a shorthand for $a + -b$.
-
-\begin{definition}[Ring Evaluation Problem (REP)]
-  For ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
-  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $1-w(x)$.
-
-  For Boolean formula $\phi$, the {\em ring evaluation problem} (REP) is to compute:
-  \begin{eqnarray}
-    \rep(\phi) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
-  \end{eqnarray}
-\label{def:labeling}
-\end{definition}
-
-A wide variety of important tasks for Boolean functions can be
-expressed as ring evaluation problems.  The 
-{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$
-It can be cast as an REP by having $\radd$ and
-$\rmul$ be addition and multiplication over rational numbers and using
-label $w(x) = 1/2$ for every variable $x$.  
-Ring evaluation of formula $\phi$ will then compute the {\em density} of
-the formula, i.e., the fraction of all possible total assignments that are
-models.  For $n = |\varset|$, scaling the density by $2^n$ will then
-yield the number of models.
-
-The {\em weighted model counting} (WMC) problem is also defined over
-rational numbers.  Some formulations of WMC allow 
-assigning weights $W$ to both the variables and their negations, such
-that there can be variables $x$ with $W(x) + W(\obar{x}) \not = 1$.  We can convert this more general form into a
-ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
-performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
-variable $x$, and then computing the final result as $\rep(\phi)\;
-\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.
-
-The {\em function hashing problem} provides a randomized test
-of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
-finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
-$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1) = \rep(\phi_2)$
-if they are logically equivalent.  If they are not equivalent, then
-the probability that $\rep(\phi_1) \not = \rep(\phi_2)$ will be at
-least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
-Therefore, ring evaluation can be used as part of a
-randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
-
-Possible generalizations of this framework are discussed in Section~\ref{sect:future}.
-
-\section{Partitioned-Operation Graphs (POGs)}
-
-Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
-\begin{proposition}
-Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following:
-\begin{description}
-\item[Negation] $\rep(\boolnot \phi) = \mulident - \rep(\phi)$
-\item[Partitioned Conjunction] $\rep\left(\Pand_{i=1,k} \phi_i\right) = \prod_{i=1,k} \rep(\phi_i)$
-\item[Partitioned Disjunction] $\rep\left(\Por_{i=1,k} \phi_i\right) = \sum_{i=1,k} \rep(\phi_i)$
-\end{description}
-\end{proposition}
-
-A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
-with nodes $N$ and edges $E \subseteq N \times N$.  When $(u,v) \in E$,
-node $v$ is said to be a {\em child} of node $u$.
-The in- and out-degrees of node $u$ are defined as $\indegree(u) = | E \cap (N \times \{u\}) |$, and
-$\outdegree(u) = | E \cap (\{u\} \times N) |$.
-Node $u$ is said to be {\em terminal} if $\outdegree(u) = 0$.  A terminal node is labeled by a Boolean constant or variable.
-Node $u$ is said to be {\em nonterminal} if $\outdegree(u) > 0$.  A nonterminal node is labeled by one of three Boolean operations:
-$\boolnot$, $\pand$, or $\por$.  Node $u$ can be labeled with $\boolnot$ only if $\outdegree(u) = 1$.
-It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
-There is a unique {\em root node} $r$ such that $\indegree(r) = 0$.
-
-As can be seen, a POG is simply a way to represent a partitioned-operation
-formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
-$\phi(u)$ as the formula denoted by node $u$ and write $\phi(P) = \phi(r)$ for root $r$ of $P$.
-
-We define the {\em size} of POG $P$, written $|P|$ to be the sum of
-the number of nonterminal nodes and the number of edges.  Ring
-evaluation of $\phi(P)$ can be performed with at most $|P|$ ring
-operations by traversing the graph from the terminal nodes up to
-the root, computing a value $\rep(u)$ for each node $u$.
-
-POGs are inspired by the d-DNNF graphs devised by Darwiche for
-representing Boolean functions \cite{darwiche:jair:2002}.
-POGs generalize d-DNNF in two ways:
-\begin{itemize}
-\item They allow negation of arbitrary nodes in the graph, not just
-  variables.  As we have seen, ring evaluation can readily handle negation, and so we allow it in the interest of generality.
-  Negation could be useful in some contexts, such as when
-  converting binary decision diagrams (BDDs) having complemented
-  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
-
-\item They allow arbitrary arguments to a $\por$ operation, as long as
-  they have disjoint models.  By contrast, each
-  disjunction node $v$ in a d-DNNF can have at most two children $u_1$ and $u_0$ for which there must be a {\em splitting variable} $x$ such that
-  any total assignment $\assign$ satisfying $\phi(u_1)$ has $\assign(x)=1$, while any total assignment satisfying $\phi(u_0)$ has
-  $\assign(x)=0$.  Our generalization is provided to allow an
-  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
-  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
-\end{itemize}
-  Both of these generalizations allow more flexibility in the POG
-  representation while maintaining the ability to efficiently perform ring evaluation.
-
-\section{The CRAT POG Representation and Proof System}
-
 A CRAT file provides both a declaration of a POG, as well as a checkable, clausal
@@
 \begin{itemize}
-\item The {\em proof generator} generates a CRAT file that both describes a POG and provides a proof that this POG is equivalent to the input formula.
-  This compiler might start with the output of another knowledge compiler.
+\item The {\em proof generator} produces a CRAT file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
+As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
 \item The {\em proof checker} ensures that all of the proof conditions are satisfied.  
@@
   The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
-  Each such declaration implicitly adds an extension variable representing the POG node and a set of {\em defining} clauses
+  Each such declaration implicitly adds an extension variable $u$ representing each  POG node $\nodeu$ and a set of {\em defining} clauses
   encoding the conjunction or disjunction operation.
@@
 variables and successive ones being extension variables.  Literal $\lit$
-is represented by a signed integer, with $-v$ being the complement of
+is represented by a signed integer, with $-v$ being the logical negation of
 variable $v$.  Each clause is indicated by a positive integer
@@
 ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
-such that clause $C$ can only reference clauses $C'$ such that $C' < C$.
+such that any clause identifier $C'$ given in the hints when adding clause $C$ must have $C' < C$.
 
@@
 (command {\tt a}), with a sequence of antecedent clauses (the
-``hint'').
+``hints'').
 Similarly for clause deletion (command {\tt dc}).
@@
 \end{center}
-where $i$ is a new clause ID, $v$ is a positive integer that does not
+Symbol `{\tt p}' refers to ``product.''
+Integer $i$ is a new clause ID, $v$ is a positive integer that does not
 correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$  
 integers representing literals of existing variables.  
-This declaration implicitly causes $k+1$ clauses to be added to the proof:
+This declaration implicitly causes $k+1$ clauses to be added to the proof, defining extension variable $v$ to be the conjunction of its arguments:
 \begin{center}
@@
 \end{center}
-where $i$ is a new clause ID, $v$ is a positive integer that does
+Symbol `{\tt s}' refers to ``sum.''
+Integer $i$ is a new clause ID, $v$ is a positive integer that does
 not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
@@
 sequence of clause IDs.
-This declaration implicitly causes three clauses to be added to the proof:
+This declaration implicitly causes three clauses to be added to the proof, defining extension variable $v$ to be the disjunction of its arguments:
 \begin{center}
@@
 
+The sequence of operator declarations, asserted clause, and
+clause deletions represents a systematic transformation of the input formula
+into a POG\@.  Validating each of these steps serves to prove that the
+POG is logically equivalent to the input formula.
+
 Observe that the defining clauses for the conjunction and disjunction
-operations provide a full Tseitin encoding of the operations.  Letting
-$\exvarset$ denote the set of added extension variables,  any
-total assignment $\assign\colon \varset \rightarrow \{0,1\}$ has a unique extension,
-which we also call $\assign$, with $\assign\colon (\varset \cup \exvarset) \rightarrow \{0,1\}$.
+operations provide a full Tseitin encoding of the operations.  For any node $\nodeu$ and associated extension variable $u$,
+we can therefore assume that any total assignment $\assign$ to the input variables will also assign
+a value to the extension variables such that $\assign(u) = 1$ if and
+only if $\assign \in \modelset(\nodeu)$.
 
@@
 \end{itemize}
-The sequence of clause addition steps provide a proof that any
-total assignment $\assign$ to the input variables that satisfies all of the
-clauses must, when extended, assign $\assign(r) = 1$.  Conversely,
+The sequence of clause addition steps provide a proof that
+$\modelset(\phi) \subseteq \modelset(\noder)$.  That is, any total
+assignment $\assign$ satisfying the input formula must also satisfy
+the formula represented by the POG\@.
+Conversely,
 each proof step that deletes an input clause proves that any
-total assignment $\alpha$ that falsifies the clause must, when extended,
-assign $\assign(r) = 0$.
-
-The sequence of operator declarations, asserted clauses, and deleted
-clauses represents a systematic transformation of the input formula
-into a POG\@.  Validating each of these steps serves to prove that the
-POG is logically equivalent to the input formula.
+total assignment $\alpha$ that falsifies the clause must
+assign $\assign(r) = 0$, and therefore 
+$\modelset(\noder) \subseteq \modelset(\phi)$.
 
@@
 consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
-$x_4$.  The generated POG (B) has six nonterminal nodes representing four
-conjunctions and two disjunctions.  We name these by the node type (product or sum) and the extension variable.
-The first part of the CRAT file
-(C) declares these nodes using clause IDs that increment by three or
-four, depending on whether the node has two children or three.
-The last two nonzero values in the sum declarations are hints providing the required mutual exclusion proof.
+$x_4$.  The generated POG (B) has six nonterminal nodes representing
+four conjunctions and two disjunctions.  We name these by the node
+type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
+extension variable.  We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
+  The first part of the CRAT file (C) declares
+these nodes using clause IDs that increment by three or four,
+depending on whether the node has two children or three.  The last two
+nonzero values in the sum declarations are hints providing the
+required mutual exclusion proof.
 
@@
 Figure \ref{fig:eg4:all}, we can see that the declaration of node
-$s_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
+$\nodes_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
 \ref{fig:eg4:all}(A) that these two clauses resolve to the clause
-$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $s_7$
-have disjoint models.  Similarly, node $s_{10}$ is declared as having
+$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
+have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
 clause IDs 16 and 19 as hints.  These resolve to the clause
 $\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
-$s_{10}$ have disjoint models.
+$\nodes_{10}$ have disjoint models.
 
 Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
-leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $s_{10}$ is implied by the input clauses, i.e.,
+leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $\nodes_{10}$ is implied by the input clauses, i.e.,
 any total assignment $\assign$
-satisfying the input clauses must extend to one with $\assign(s_{10}) = 1$.
+satisfying the input clauses must have $\assign(s_{10}) = 1$.
 Working backward, we can see that
-clause 29 indicates that node $p_8$ will be implied by the input
+clause 29 indicates that variable $p_8$ will be implied by the input
 clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
@@
 input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
-that there is no total assignment $\assign$ that falsifies this clause, but assigns $\assign(s_{10}) = 1$.
-proof proceeds by first assuming that the clause is false, requiring
+that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
+The proof proceeds by first assuming that the clause is false, requiring
 $\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hints then consist of unit
@@
 $\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
-via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 22,
+via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
 completing the RUP proof.  The deletion hints for the other input
@@
 the POG upward, showing that any total assignment that falsifies the clause
-must assign $s_{10} = 0$.
+must assign $\assign(s_{10}) = 0$.
 
-Deleting the asserted clauses so simple that we do not show it.  It
+Deleting the asserted clauses is so simple that we do not show it.  It
 involves simply deleting the clauses from clause number 35 down to
@@
 Observe that this proof must ``visit'' nodes
-$p_5$ and $p_6$ twice, separately considering total assignments where $\assign(x_1) = 0$
+$\nodep_5$ and $\nodep_6$ twice, separately considering total assignments where $\assign(x_1) = 1$
 (clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).  In
@@
 strategy to generate a d-DDNNF representation.  Converting a d-DNNF
-graph into a POG is straightforward, each node in a d-DNNF graph can
-be directly represented as a POG node.  We perform simplifications on
-the d-DNNF representation to eliminate any Boolean constants.  We can
-therefore assume that the POG contains only conjunction, disjunction,
+graph into a POG is straightforward---each node in a d-DNNF graph can
+be directly represented as a POG node.  Rather than representing every d-DNNF node explicitly,
+our progrma perform simplifications on
+to eliminate any Boolean constants.  Except for the degenerate cases where the formula is unsatisfiable or a tautology, we can
+ssume that the POG contains only conjunction, disjunction,
 and literal nodes.
 
-\section{CNF Implies POG}
+\section{Forward implication}
 
-The most challenging part of the proof is to show that any total
-assignment $\alpha$ that is a model of $\phi$ will, when extended,
-have $\assign(r) = 1$ for root literal
-$r$.  This part of the proof consists of a series of assertion steps
-leading to one adding $r$ as a unit clause.
+For input formula $\phi$ and its translation into a POG with root
+$\noder$, the most challenging part of the proof is to show that
+$\modelset(\phi) \subseteq \modelset(\noder)$, i.e., that any total
+assignment $\alpha$ that is a model of $\phi$ will yield $\assign(r) =
+1$, where $r$ is the extension variable associated with root node
+$\noder$.  This part of the proof consists of a series of assertion
+steps leading to one adding $r$ as a unit clause.
 
-In the following, we describe proof generation as a process that
-recursively traverses the graph from the root downward and
-generates proof steps from the lower nodes upward.  This process
-effectively expands the POG into a tree and therefore can be very
-inefficient.  We eliminate this inefficiency with the addition of
-lemmas, as is described later.
+We describe this part of the proof generation as a recursive procedure
+$\validate(u, \passign, \psi)$ taking as arguments 1) POG node $\nodeu$, 2) partial assignment
+$\passign$, and 3) a set of clauses $\psi \subseteq \phi$.
+The procedure adds a number of clauses to the proof, culminating with 
+the addition of the {\em target clause}
+$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating that any total
+assignment $\assign$ such that $\passign \subseteq \assign$
+will assign $\assign(u) = 1$.
+The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \phi$.
+The result will therefore be to add the unit clause $r$ to the proof.
+Here we present a correct, but somewhat inefficient formulation of
+$\validate$.  We then later refine it with some optimizations.
 
-Assume that formula $\phi$ is 
-represented as a set of clauses.
+Assume that formulas $\psi$ and $\phi$ are
+represented as sets of clauses, with $\psi \subseteq \phi$.
 For partial assignment
-$\passign$, the expression  $\simplify{\phi}{\passign}$ denotes the
+$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the
 reduced and simplified set of clauses obtained by 1) eliminating any
@@
 of each simplified clause $C$, i.e., which of the (possibly multiple) original clauses simplified to become $C$.
-More formally, for $C \in \simplify{\phi}{\passign}$, we let $\prov(C)$ denote
-those clauses $C' \in \phi$, such that
+More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov(C, \psi)$ denote
+those clauses $C' \in \psi$, such that
 $C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
 We then extend the definition of $\prov$ to any simplified formula
-$\psi$ as $\prov(\psi) = \bigcup_{C \in \psi} \prov(C)$.
+$\psi'$ as $\prov(\psi', \psi) = \bigcup_{C \in \psi'} \prov(C, \psi)$.
 
-In the following we define a recursive routine $\validate(u, \passign,
-\psi)$ taking as arguments 1) POG node $u$, 2) partial assignment
-$\passign$, and 3) a set of clauses $\psi \subsetq \phi$.  The outcome of its execution should be to add a number of assertions culminating with
-the addition of the {\em target clause}
-$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating that any total
-assignment $\assign$ such that $\passign \subseteq \assign$
-will, when extended to extension variables, will assign $\assign(u) = 1$.
-The top-level call has $u = r$, $\passign = \emptyset$, and $\psi = \phi$.  The result will therefore be to add the unit clause $r$ to the proof.
+The recursive call $\validate(u, \passign, \psi)$ assumes that we have
+traversed a path from the root node down to node $\nodeu$, with the
+literals encountered in the conjunction nodes forming the partial
+assignment $\passign$.  The set of clauses $\psi$ can be a proper
+subset of the input clauses $\phi$ when a conjunction node has caused
+a splitting into clauses containing disjoint variables.
+The subgraph with root node $\nodeu$ should be a POG representation of the formula
+$\simplify{\psi}{\passign}$.  Having the
+routine add target clause indicates that any total assignment 
+$\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.
 
-In following a path from the root node downward to a node $u$, the
-literals encountered in the conjunction nodes (and possibly a terminal
-literal) define a partial assignment $\passign$.  The subgraph with
-root $u$ should be a POG representation of the reduced formula
-$\simplify{\phi}{\passign}$.  Our recursive goal is to generate
-a sequence of proof steps finishing with the addition of the {\em target clause}
-$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating that any total
-assignment $\assign$ such that $\passign \subseteq \assign$ will, when extended, yield
-$\alpha(u) = 1$.
-
-The process for generating such a proof depends on the form of node $u$:
+The process for generating such a proof depends on the form of node $\nodeu$:
 \begin{enumerate}
-\item if $u$ is a literal $\lit$, then the formula
-  $simplify{\phi}{\passign}$ must consist of the single unit clause
-  $C = \lit$, such that any $C' \in \prov{C}$ must have $C' \subseteq \{ \lit \} \bigcup_{\lit \in \passign} \obar{\lit}$.
+\item If $\nodeu$ is a literal $\lit$, then the formula
+  $\simplify{\psi}{\passign}$ must consist of the single unit clause
+  $C = \lit$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit \} \bigcup_{\lit \in \passign} \obar{\lit}$.
   Any of these can
   serve as the target clause.
-\item If $u$ is a disjunction node with children $u_1$ and $u_0$,
+\item If $\nodeu$ is a disjunction node with children $\nodeu_1$ and $\nodeu_0$,
   then, since the node originated from a d-DNNF graph, there must be
-  some variable $x$ such that either $u_1 = x$ or $u_1$ is a
+  some variable $x$ such that either $\nodeu_1$ is a terminal node with literal $x$ or $\nodeu_1$ is a
   conjunction node containing $x$ as a child.  In either case, we
-  recurse to node $u_1$ with the partial assignment $\passign \cup \{ x \}$.
-  This will yield a sequence of proof steps
-  $u \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-Similarly, either $u_0  = \obar{x}$ or $u_0$ is a conjunction node containing $\obar{x}$ as
-  a child.  In either case, we recurse to node $u_0$
- with the partial assignment $\passign \cup \{ \obar{x} \}$.
-  This will yield a sequence of proof steps
-  $u \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-  These recursive results can be combined, via two proof steps, with second and third defining clauses for $u$
-  to generate the target clause for $u$.
-\item If $u$ is a conjunction node, then we can divide its children
-  into a set of literal nodes $L$ and a set of nonterminal nodes $u_1, u_2, \ldots, u_k$.
+  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
+  This will cause the addition of the target clause
+  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
+Similarly, either $\nodeu_0$ is a terminal node with literal $\obar{x}$ or $\nodeu_0$ is a conjunction node containing $\obar{x}$ as
+  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
+  causing the addition of the target clause
+  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
+  These recursive results can be combined, via two proof steps, with the second and third defining clauses for $\nodeu$
+  to generate the target clause for $\nodeu$.
+\item If $\nodeu$ is a conjunction node, then we can divide its children
+  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
   \begin{enumerate}
@@
   In other cases, we must prove that the formula
-  $\simplify{\phi}{\passign \cap \{\obar{\lit}\}}$ is unsatisfiable.  We
+  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
   do so by writing the formula to a file, invoking a proof-generating
@@
   as assertions in the CRAT file.
-\item If there is a single nonterminal child ($k = 1$), we can recurse on node $u_1$
-  with partial assignment $\passign \cup \bigcup_{\lit \in L} \lit$.
+\item If there is a single nonterminal child ($k = 1$), we recursively call
+  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$, 
 \item If there are multiple nonterminal children ($k > 1$)
   it must be the case that the clauses in
-  $\simplify{\phi}{\passign}$ can be partitioned into $k$ subsets
+  $\simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
   $\psi_1, \psi_2, \ldots, \psi_k$ such that $\dependencyset(\psi_i)
   \cap \dependencyset(\psi_j) = \emptyset$ for $1 \leq i < j \leq k$,
-  and we can match each node $u_i$ to subset $\psi_i$ based on its
+  and we can match each node $\nodeu_i$ to subset $\psi_i$ based on its
   literals.
-  For each $i$ such that $1 \leq i \leq k$, we can then recurse on node $u_i$
-  with partial assignment $\passign \cup \bigcup_{\lit \in L} \lit$, and based on the input clauses in $\prov(\psi_i)$.
+  For each $i$ such that $1 \leq i \leq k$, we then recursively call
+  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \prov(\psi_i, \psi)\right)$
 \end{enumerate}
@@
   the literals, the recursively computed target clauses, and all but
-  the first defining clause for the conjunction operation to generate the target clause for node $u$.
+  the first defining clause for the conjunction operation to generate the target clause for node $\nodeu$.
 \end{enumerate}
+Observe that most of these steps involve a polynomial number of
+operations per recursive call.  The only potential source of
+exponential growth is when a SAT solver must be invoked to validate a
+literal.
 
+\section{Reverse Implication}
+
+\section{Optimizations}
+
+\subsection{Literal Grouping}
+
+\subsection{Lemmas}
 \label{sect:lemma}
@@
 
-Each node $u$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
-$s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(u)$ associated
-with the node is defined recursively as $\phi(u) = \bigwedge_{i=1,k}
+Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
+$s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
+with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{i=1,k}
 [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$

### 2023-03-09 doc/mc-paper.tex
Progress on paper

@@
 \newcommand{\por}{\mathbin{\lor^{\sf p}}}
-\DeclareMathOperator*{\Pand}{\bigwedge^{\sf v}}
-\DeclareMathOperator*{\Por}{\bigvee^{\sf a}}
+\DeclareMathOperator*{\Pand}{\bigwedge^{\sf p}}
+\DeclareMathOperator*{\Por}{\bigvee^{\sf p}}
 \newcommand{\boolnot}{\neg}
@@
 \newcommand{\passign}{\rho}
+\newcommand{\lassign}{\beta}
 \newcommand{\uassign}{{\cal U}}
@@
 \newcommand{\validate}{\textsf{validate}}
+\newcommand{\prov}{\textrm{Prov}}
 
@@
 \newcommand{\simplify}[2]{#1|_{#2}}
-\newcommand{\prov}{\mathit{Prov}}
 
@@
     with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em conjunction} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
       $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
       $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em disjunction} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
       $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
@@
 
-  We highlight the following special classes of Boolean formulas:
-  \begin{description}
-    \item[Normal form:] A formula where negation is applied only to variables.
-    \item[Conjunctive Normal Form (CNF):] A normal form formula where  disjunction is applied only to literals.
-    \item[Partitioned-Operation:] A formula satisfying the following for all conjunction and disjunction operations: 
+  We highlight two special classes of Boolean formulas.  A formula is
+  in {\em normal form} when negation is applied only to variables.  A
+  formula is in {\em conjunctive normal form} (CNF) when 1) it is in
+  normal form, and 2) sum is applied only to literals.  A CNF
+  formula can be represented as a set of {\em clauses}, each of which is a
+  set of literals.  Each clause represents the sum of the
+  literals, and the formula is the product of its clauses.  We use
+  set notation to reference the clauses in a formula and the
+  literals in a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause.
+This literal must be assigned value $1$ by any satisfying assignment of the formula.
+
+  A {\em partitioned-operation formula} 
+ satisfies the following for all product and sum operations: 
       \begin{itemize}
-      \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
+      \item The arguments to a product must have disjoint dependency sets.  That is, operation
         $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \item The arguments to a disjunction must have disjoint models.  That is, operation
+      \item The arguments to a sum must have disjoint models.  That is, operation
         $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
       \end{itemize}
-  \end{description}
-     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned-operation formula.
-     A formula in conjunctive normal form can be represented by a set
-     of {\em clauses}, each of which is a set of literals.  Each
-     clause represents the disjunction of the literals, and the formula is the conjunction of its clauses.
+     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.
 
@@
 
-We propose a general framework for 
-summarizing properties of Boolean formulas.
+We propose a general framework for summarizing properties of Boolean
+formulas inspired by the formulation of {\em algebraic model counting}
+by Kimmig, et al.~\cite{kimmig:jal:2017}.
 
@@
   \end{eqnarray}
-  In this equation, the summation $\sum$ is computed using operation $\radd$, and the product $\prod$ is computed using operation $\rmul$.
+  In this equation, the summation $\sum$ is computed using addition operation $\radd$, and the product $\prod$ is computed using multiplication operation $\rmul$.
 \label{def:labeling}
@@
 
-A wide variety of important tasks for Boolean functions can be
+Many important queries about Boolean formulas can be
 expressed as ring evaluation problems.  The 
@@
 \item[Negation] $\rep(\boolnot \phi, w) = \mulident - \rep(\phi, w)$
-\item[Partitioned Conjunction] $\rep\left(\Pand_{i=1,k} \phi_i, w \right) = \prod_{i=1,k} \rep(\phi_i, w)$
-\item[Partitioned Disjunction] $\rep\left(\Por_{i=1,k} \phi_i, w\right) = \sum_{i=1,k} \rep(\phi_i, w)$
+\item[Partitioned Product] $\rep\left(\Pand_{i=1,k} \phi_i, w \right) = \prod_{i=1,k} \rep(\phi_i, w)$
+\item[Partitioned Sum] $\rep\left(\Por_{i=1,k} \phi_i, w\right) = \sum_{i=1,k} \rep(\phi_i, w)$
 \end{description}
@@
   variables.  As we have seen, ring evaluation can readily handle negation, and so we allow it in the interest of generality.
-  Negation could be useful in some contexts, such as when
-  converting binary decision diagrams (BDDs) having complemented
-  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
+%%  Negation could be useful in some contexts, such as when
+%%  converting binary decision diagrams (BDDs) having complemented
+%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
 
@@
   they have disjoint models.  By contrast, each
-  disjunction node $\nodev$ in a d-DNNF can have at most two children $\nodeu_1$ and $\nodeu_0$ for which there must be a {\em decision variable} $x$ such that
-  any total assignment $\assign \in  \modelset(u_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
+  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
+  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
 Our generalization is provided to allow an
@@
   \draw[thick] (20,12) [-latex] -- (28,12);
-  \node [above] at (24,12) {\texttt{.nnf}};
+  \node [above] at (24,12) {\texttt{.ddnnf}};
 
@@
 
-A CRAT file provides both a declaration of a POG, as well as a checkable, clausal
+A CRAT file provides both a declaration of a POG, as well as an easily checkable
 proof that a Boolean formula, given in conjunctive normal
@@
   The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
-  Each such declaration implicitly adds an extension variable $u$ representing each  POG node $\nodeu$ and a set of {\em defining} clauses
-  encoding the conjunction or disjunction operation.
+  Declaring a node $\nodeu$ 
+implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses
+  encoding the product or sum operation.
   This is the only means for adding extension variables to the proof.
@@
   providing a RUP proof that the arguments are mutually exclusive.
-  Only binary $\por$ operations are allowed to avoid requiring
-  $k\,(k-1)/2$ proofs of disjointness for larger values of $k$.
+  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
+  $k\,(k-1)/2$ proofs of disjointness.
 \end{itemize}
@@
   \begin{tabular}{lllll}
+    \toprule
     \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
@@
     \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause} \\
-     & {\tt dc} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause \\
+     & {\tt d} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause \\
     \midrule
@@
      & {\tt r} & $L$             &            & Declare root literal\\
+    \bottomrule
   \end{tabular}
@@
 ``hints'').
-Similarly for clause deletion (command {\tt dc}).
+Similarly for clause deletion (command {\tt d}).
 
-The declaration of a $\pand$ operation has the form:
+\begin{table}
+\caption{Defining Clauses for Product (A) and Sum (B) Operations}
+\begin{minipage}{0.54\textwidth} 
+\begin{center}
+\begin{tabular}{cccccc}
+\multicolumn{6}{c}{(A).  Product Operation $\pand$}\\
+\toprule
+\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
+\midrule
+  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
+  $i\!+\!1$ & $-v$ & $\lit_1$  \\
+  $i\!+\!2$ & $-v$ & $\lit_2$  \\
+  & $\ldots$ \\
+  $i\!+\!k$ & $-v$ & $\lit_k$  \\
+\bottomrule
+\end{tabular}
+\end{center}
+\end{minipage}
+\begin{minipage}{0.44\textwidth}
+\begin{center}
+\begin{tabular}{cccc}
+\multicolumn{4}{c}{(B).  Sum Operation $\por$}\\
+\toprule
+\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
+\midrule
+  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
+  $i\!+\!1$ & $v$ & $-\lit_1$ \\
+  $i\!+\!2$ & $v$ & $-\lit_2$ \\
+\bottomrule
+$\;$ \\
+$\;$ \\
+\end{tabular}
+\end{center}
+\end{minipage}
+\label{tab:defining}
+\end{table}
+
+The declaration of a {\em product} operation, creating a node with operation $\pand$,
+ has the form:
 \begin{center}
@@
 \end{center}
-Symbol `{\tt p}' refers to ``product.''
 Integer $i$ is a new clause ID, $v$ is a positive integer that does not
@@
 integers representing literals of existing variables.  
-This declaration implicitly causes $k+1$ clauses to be added to the proof, defining extension variable $v$ to be the conjunction of its arguments:
-\begin{center}
-\begin{tabular}{cccccc}
-\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
-  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
-  $i\!+\!1$ & $-v$ & $\lit_1$  \\
-  $i\!+\!2$ & $-v$ & $\lit_2$  \\
-  & $\ldots$ \\
-  $i\!+\!k$ & $-v$ & $\lit_k$  \\
-\end{tabular}
-\end{center}
+As Table~\ref{tab:defining}(A) shows,
+this declaration implicitly causes $k+1$ clauses to be added to the proof, defining extension variable $v$ to be the product of its arguments.
+
 The dependency sets for the arguments represented by each pair of
@@
 and $\lit_{j}$ must
-be disjoint, for $1 \leq i < j \leq k$.  A $\pand$ operation may have no arguments,
+be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
 representing Boolean constant $1$.  The only clause added to the proof will be
@@
 
-The declaration of a $\por$ operation has the form:
+The declaration of a {\em sum} operation, creating a node with operation $\por$, has the form:
 \begin{center}
@@
 \end{center}
-Symbol `{\tt s}' refers to ``sum.''
 Integer $i$ is a new clause ID, $v$ is a positive integer that does
@@
 consists of a
-sequence of clause IDs.
-This declaration implicitly causes three clauses to be added to the proof, defining extension variable $v$ to be the disjunction of its arguments:
-\begin{center}
-\begin{tabular}{cccc}
-\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
-  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
-  $i\!+\!1$ & $v$ & $-\lit_1$ \\
-  $i\!+\!2$ & $v$ & $-\lit_2$ \\
-\end{tabular}
-\end{center}
-The hints must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$.
+sequence of clause IDs, all of which must be defining clauses for other POG operations.
+As Table~\ref{tab:defining}(B) shows,
+this declaration implicitly causes three clauses to be added to the proof, defining extension variable $v$ to be the sum of its arguments.
+The hints must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.
 
@@
 
-The sequence of operator declarations, asserted clause, and
+The defining clauses for a product or sum
+operation uniquely define the value of its extension variable for any assignment of values to the argument variables.
+For the
+extension variable $u$ associated with any POG node $\nodeu$, we can therefore
+assume that any total assignment $\assign$ to the input variables will
+also assign a value to $u$ such that $\assign(u) =
+1$ if and only if $\assign \in \modelset(\nodeu)$.  For a POG with
+root literal $r$, we can define its set of models $\modelset(P)$ as
+the set of all total assignments $\assign$ such that $\assign(r) = 1$.
+
+%% A CRAT proof follows the same general form as a QRAT dual
+%% proof~\cite{bryant:cade:2021}---one that ensures that each clause
+%% addition and each clause deletion preserves equivalence.  With CRAT,
+%% however, clauses are defined both explicitly and implicitly.  Starting
+%% with the set of input clauses, the proof consists of a sequence of
+%% steps that both add and delete clauses.  Each addition must be truth
+%% preserving, that is, any satisfying total assignment for the set of clauses
+%% before clause addition should still be a satisfying assignment afterwards.
+%% Each deletion must be falsehood preserving.  That is, there can be no
+%% new satisfying assignments when the clause is deleted.
+
+The sequence of operator declarations, asserted clauses, and
 clause deletions represents a systematic transformation of the input formula
-into a POG\@.  Validating each of these steps serves to prove that the
+into a POG\@.  Validating all of these steps serves to prove that the
 POG is logically equivalent to the input formula.
-
-Observe that the defining clauses for the conjunction and disjunction
-operations provide a full Tseitin encoding of the operations.  For any node $\nodeu$ and associated extension variable $u$,
-we can therefore assume that any total assignment $\assign$ to the input variables will also assign
-a value to the extension variables such that $\assign(u) = 1$ if and
-only if $\assign \in \modelset(\nodeu)$.
-
-A CRAT proof follows the same general form as a QRAT dual
-proof~\cite{bryant:cade:2021}---one that ensures that each clause
-addition and each clause deletion preserves equivalence.  With CRAT,
-however, clauses are defined both explicitly and implicitly.  Starting
-with the set of input clauses, the proof consists of a sequence of
-steps that both add and delete clauses.  Each addition must be truth
-preserving, that is, any satisfying total assignment for the set of clauses
-before clause addition should still be a satisfying assignment afterwards.
-Each deletion must be falsehood preserving.  That is, there can be no
-new satisfying assignments when the clause is deleted.
-
 At the completion of the proof, the following conditions must hold:
 \begin{itemize}
+\item There should exactly one remaining clause that was added via RUP
+  addition, and this should be a unit clause consisting of the root literal $r$.
 \item All of the input clauses have been deleted.
-\item There should be exactly one clause left that was added via RUP
-  addition, and this should be a unit clause consisting of the root literal $r$.
 \end{itemize}
-The sequence of clause addition steps provide a proof that
-$\modelset(\phi) \subseteq \modelset(\noder)$.  That is, any total
+The sequence of clause addition steps provide a {\em forward implication} proof that
+$\modelset(\phi) \subseteq \modelset(P)$.  That is, any total
 assignment $\assign$ satisfying the input formula must also satisfy
@@
 total assignment $\alpha$ that falsifies the clause must
-assign $\assign(r) = 0$, and therefore 
-$\modelset(\noder) \subseteq \modelset(\phi)$.
+assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
+that
+$\modelset(P) \subseteq \modelset(\phi)$.
 
@@
 \midrule
- \texttt{dc 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
- \texttt{dc 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
- \texttt{dc 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
- \texttt{dc 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
- \texttt{dc 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
+ \texttt{d 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
+ \texttt{d 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
+ \texttt{d 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
+ \texttt{d 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
+ \texttt{d 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
 \bottomrule
@@
 $x_4$.  The generated POG (B) has six nonterminal nodes representing
-four conjunctions and two disjunctions.  We name these by the node
+four products and two sums.  We name these by the node
 type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
-extension variable.  We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
+extension variable.
+% We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
   The first part of the CRAT file (C) declares
@@
 CRAT file.  Referring back to the declarations of the sum nodes in
-Figure \ref{fig:eg4:all}, we can see that the declaration of node
+Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
 $\nodes_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
@@
 clauses that cause unit propagation.  Hint clauses 8 and 10 force the
-assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 forces
+assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
 $\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
@@
 Observe that this proof must ``visit'' nodes
-$\nodep_5$ and $\nodep_6$ twice, separately considering total assignments where $\assign(x_1) = 1$
+$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
 (clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).  In
@@
 but it would also work for other tools that use a top-down search
-strategy to generate a d-DDNNF representation.  Converting a d-DNNF
+strategy to generate a d-DNNF representation.  Converting a d-DNNF
 graph into a POG is straightforward---each node in a d-DNNF graph can
 be directly represented as a POG node.  Rather than representing every d-DNNF node explicitly,
-our progrma perform simplifications on
-to eliminate any Boolean constants.  Except for the degenerate cases where the formula is unsatisfiable or a tautology, we can
-ssume that the POG contains only conjunction, disjunction,
+our program performs simplifications
+to eliminate Boolean constants.  Except for the degenerate cases where the formula is unsatisfiable or a tautology, we can
+therefore assume that the POG contains only product, sum,
 and literal nodes.
 
-\section{Forward implication}
+\subsection{Forward implication proof}
 
-For input formula $\phi$ and its translation into a POG with root
-$\noder$, the most challenging part of the proof is to show that
-$\modelset(\phi) \subseteq \modelset(\noder)$, i.e., that any total
+For input formula $\phi$ and its translation into a POG with root literal
+$r$, the most challenging part of the proof is to show that
+$\modelset(\phi) \subseteq \modelset(P)$, i.e., that any total
 assignment $\alpha$ that is a model of $\phi$ will yield $\assign(r) =
-1$, where $r$ is the extension variable associated with root node
-$\noder$.  This part of the proof consists of a series of assertion
-steps leading to one adding $r$ as a unit clause.
+1$, for root literal $r$.
+This part of the proof consists of a series of clause assertions
+leading to one adding $r$ as a unit clause.
 
 We describe this part of the proof generation as a recursive procedure
-$\validate(u, \passign, \psi)$ taking as arguments 1) POG node $\nodeu$, 2) partial assignment
-$\passign$, and 3) a set of clauses $\psi \subseteq \phi$.
+$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
+node $\nodeu$, partial assignment
+$\passign$, and a set of clauses $\psi \subseteq \phi$.
 The procedure adds a number of clauses to the proof, culminating with 
@@
 Here we present a correct, but somewhat inefficient formulation of
-$\validate$.  We then later refine it with some optimizations.
+$\validate$.  We then refine it with some optimizations.
 
@@
 For partial assignment
-$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the
-reduced and simplified set of clauses obtained by 1) eliminating any
+$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
+obtained from $\psi$ by: 1) eliminating any
 clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
@@
 In doing these simplifications, we also track the {\em provenance}
-of each simplified clause $C$, i.e., which of the (possibly multiple) original clauses simplified to become $C$.
+of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
 More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov(C, \psi)$ denote
@@
 We then extend the definition of $\prov$ to any simplified formula
-$\psi'$ as $\prov(\psi', \psi) = \bigcup_{C \in \psi'} \prov(C, \psi)$.
+$\gamma$ as $\prov(\gamma, \psi) = \bigcup_{C \in \gamma} \prov(C, \psi)$.
 
@@
 traversed a path from the root node down to node $\nodeu$, with the
-literals encountered in the conjunction nodes forming the partial
+literals encountered in the product nodes forming the partial
 assignment $\passign$.  The set of clauses $\psi$ can be a proper
-subset of the input clauses $\phi$ when a conjunction node has caused
+subset of the input clauses $\phi$ when a product node has caused
 a splitting into clauses containing disjoint variables.
@@
 $\simplify{\psi}{\passign}$.  Having the
-routine add target clause indicates that any total assignment 
+routine add its target clause indicates that any total assignment 
 $\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.
@@
 \begin{enumerate}
-\item If $\nodeu$ is a literal $\lit$, then the formula
+\item If $\nodeu$ is a literal $\lit'$, then the formula
   $\simplify{\psi}{\passign}$ must consist of the single unit clause
-  $C = \lit$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit \} \bigcup_{\lit \in \passign} \obar{\lit}$.
+  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \bigcup_{\lit \in \passign} \obar{\lit}$.
   Any of these can
   serve as the target clause.
-\item If $\nodeu$ is a disjunction node with children $\nodeu_1$ and $\nodeu_0$,
+\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
   then, since the node originated from a d-DNNF graph, there must be
   some variable $x$ such that either $\nodeu_1$ is a terminal node with literal $x$ or $\nodeu_1$ is a
-  conjunction node containing $x$ as a child.  In either case, we
+  product node containing $x$ as a child.  In either case, we
   recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
@@
   $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-Similarly, either $\nodeu_0$ is a terminal node with literal $\obar{x}$ or $\nodeu_0$ is a conjunction node containing $\obar{x}$ as
+Similarly, either $\nodeu_0$ is a terminal node with literal $\obar{x}$ or $\nodeu_0$ is a product node containing $\obar{x}$ as
   a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
@@
   These recursive results can be combined, via two proof steps, with the second and third defining clauses for $\nodeu$
+(see Table~\ref{tab:defining}(B))
   to generate the target clause for $\nodeu$.
-\item If $\nodeu$ is a conjunction node, then we can divide its children
+\item If $\nodeu$ is a product node, then we can divide its children
   into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
@@
   $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$, 
-\item If there are multiple nonterminal children ($k > 1$)
+\item If there are multiple nonterminal children ($k > 1$),
   it must be the case that the clauses in
-  $\simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
-  $\psi_1, \psi_2, \ldots, \psi_k$ such that $\dependencyset(\psi_i)
-  \cap \dependencyset(\psi_j) = \emptyset$ for $1 \leq i < j \leq k$,
-  and we can match each node $\nodeu_i$ to subset $\psi_i$ based on its
+  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
+  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
+  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
+  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
   literals.
-  For each $i$ such that $1 \leq i \leq k$, we then recursively call
-  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \prov(\psi_i, \psi)\right)$
+  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
+  We recursively call
+  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \psi_i\right)$.
 \end{enumerate}
@@
   the literals, the recursively computed target clauses, and all but
-  the first defining clause for the conjunction operation to generate the target clause for node $\nodeu$.
+  the first defining clause for node $\nodeu$
+(see Table~\ref{tab:defining}(A))
+to generate the target clause for node $\nodeu$.
 \end{enumerate}
@@
 
-\section{Reverse Implication}
+\subsection{Reverse implication proof}
+
+Completing the equivalence proof of input formula $\phi$ and its POG
+representation with root literal $r$ requires showing that
+$\modelset(P) \subseteq \modelset(\phi)$.  This is done in the
+CRAT framework by first deleting all asserted clauses, except for the
+final unit clause for root literal $r$ and then deleting all of the
+input clauses.  Suppose at some point in this process that we want to
+delete clause $C$ such that the remaining set of clauses will form the
+CNF formula $\psi$.  Deleting $C$ requires proving that
+$\modelset(\psi) = \modelset(\psi \lor C)$, i.e., that deleting $C$
+does not make the formula easier to satisfy.
+
+The asserted clauses can be deleted in reverse order, using the same
+hints that were used in their original assertions.  By reversing the
+order, those clauses that were used as hints when a clause was
+added will still remain when it is deleted.
+
+Each input clause deletion can be done as a single RUP step.  The
+proof generator builds a sequence of hints from among the defining
+clauses of the POG nodes via a single, bottom-up pass through the
+graph.  A RUP proof for input clause $C$ effectively proves that any
+total assignment $\assign$ that does not satisfy $C$ will yield
+$\assign(r) = 0$.  It starts with the set of literals
+$\{ \obar{\lit} | \lit \in C\}$, describing the required condition for
+assignment $\assign$ to falsify clause $C$.
+It then
+adds literals via unit propagation until a
+conflict arises.    Unit literal $r$ gets
+added right away, setting up a potential conflict.
+
+Working upward through the graph, each node $\nodeu$ is {\em marked} when
+the collected set of literals force $u$ to evaluate to $0$.  The
+program adds $\obar{u}$ to the RUP literals and adds the appropriate
+defining clause to the list of hints.  A terminal node with literal
+$\lit$ will be marked if $\lit \in C$, with no hint required.  If
+product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
+$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
+added to the hints.  Marking sum node $\nodeu$ requires that both of
+is children $\nodeu_1$ and $\nodeu_2$ be marked.  The first defining
+clause for this node (see Table~\ref{tab:defining}(B)) will then be added.  At the very end, the program
+(assuming the reverse implication holds) will attempt to mark root
+node $\noder$, which would require $\assign(r) = 0$, yielding a
+conflict.
+
+It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
+each clause deletion requires a single RUP step, where 
+the number of hints will be less than the number of
+nodes in the POG\@.
 
@@
 
+The performance of forward implication, both in its execution time and
+the size of the proof generated by two optimizations described here.
+A key feature of both is that they do not require any changes to the
+proof framework---they build on the power of extended resolution to
+create new logical structures within the proof.
+
 \subsection{Literal Grouping}
+
+A single recursive step of $\validate$ can encounter product nodes
+having many---tens or even hundreds---of literal as children.  As described
+earlier, the standard approach for each child literal $\lit \in \L$ is to first attempt BCP, but calling a SAT solver
+on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$.
+
+An alternative approach is to collect those literals $\lit_1, \lit_2,
+\ldots, \lit_m$ that cannot be validated by BCP and then define a
+product node $\nodev$ having these literals as children.  The goal
+then becomes to prove that this product must yield 1 for extension
+variable $v$.  This can be done by contradiction using a SAT solver.
+In particular, looking at the defining clauses for $\nodev$ (Table
+\ref{tab:defining}(A)), we can see that assigning 0 to $v$ will cause
+  the first defining clause to simplify to $C_p = \obar{lit}_1 \lor
+  \obar{\lit}_2 \lor \cdots \lor \obar{\lit}_m$, and for the remaining
+  clauses to become tautologies.  The formula
+  $\simplify{\psi}{\passign} \cup \{ C_p \}$ should therefore be
+  unsatisfiable.  The proof steps and hints generated by the SAT
+  solver can mapped back to the input clauses, plus the defining
+  clauses for $\nodev$.
 
@@
 \label{sect:lemma}
+
+As we have noted, the recursive calling of $\validate$ starting at
+root $\noder$ effectively expands the POG into a tree, and this can
+lead to an exponential number of calls.  We can exploit the property
+that a shared subgraph in the graph generated by a top-down d-DNNF
+compiler arises when the program detects that the simplified set of
+clauses arising from a partial assignment to the literals matches that
+of a previous traversal.  When this d-DNNF node is translated into POG
+node $\nodeu$, the proof generator can assume (and later check), that
+there is a simplified set of clauses $\gamma_{\nodeu}$
+for which the subgraph with root $\nodeu$ is its POG representation.
+
+The forward implication proof can exploit this property of shared
+subgraphs by constructing and proving a {\em lemma} for each node
+$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
+total assignment $\assign$ that satisfies formula $\gamma$ must yield
+$\assign(u) = 1$.  This lemma is then invoked for every node having
+$\nodeu$ as a child.  The net effect is to allow $\validate$ to
+maintain the graph structure of the POG, invoking $\validate$ on each
+node at most once.
+
+The challenge for implementing this strategy is to find a way to
+represent the clauses in a simplified formula $\gamma$.  Some may be
+unalterered input clauses, and these can be used directly.  Others,
+however will be clauses that do not appear in the input formula.  We
+implement these by adding POG product nodes to the CRAT file to create
+the appropriate clauses.  Imagine that there is an {\em argument} clause
+$C \in \gamma$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  If we
+define product node $\nodev$ with arguments
+$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, then by
+Table~\ref{tab:defining}(A), we will introduce a defining clause
+$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
+  synthetic} clause having $\obar{v}$ as the {\em guard literal}.
+That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
+  activate} the clause, causing it to represent argument clause $C$.  On the other
+hand, a partial assignment with $\passign(v) = 1$ will
+cause the clause to become a tautology and therefore have no effect.
+
+Suppose for every clause $C_j \in \gamma$ that does not correspond to
+an input clause, we generate a synthetic clause $C'_j$ with guard literal
+$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
+and
+let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
+Invoking $\validate(\nodeu, \lassign, \gamma)$
+ will then prove a lemma, given by the target clause
+ $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
+ showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u)$.
+
+Later, when node $\nodeu$ is encountered by a call to $\validate(\nodeu, \passign, \psi)$, we invoke the lemma
+by showing that each synthetic clause
+$C_j$ matches some simplified clause in $\simplify{\psi}{\passign}$.  More precisely,
+we assert the clause
+$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$
+to show that synthetic clause $C_j$ will be activated for $1 \leq j \leq m$.
+
+Observe that the lemma structure can be hierarchical, since a shared
+subgraph may contain nodes that are themselves roots of shared
+subgraphs.  Nonetheless, the principles described allow the
+definition, proof, and applications of a lemma for each shared node in
+the graph.
 
@@
 \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
-of such a node could have $k$ conjunction nodes: $t_i = p_i \pand s_i$
-for $1 \leq i \leq k$, and then $k-1$ binary disjunction nodes to form
-the disjunction of $t_1, \ldots, t_k$.
+of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
+for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
+the sum of $t_1, \ldots, t_k$.
 

### 2023-03-09 crat/prototype/pigeon-d4-split2/run.sh
merge

@@ -4,6 +4,7 @@ cat $BASE.drat | grep -v "d " > $BASE.rat
 ./negate $BASE.rat > $BASE.cubes
 cat $BASE.acnf $BASE.cubes | awk '/p cnf/ {print "p inccnf"} / 0/ {print $0}'> $BASE.icnf
 ~/cadical/build/cadical $BASE.icnf $BASE.drup --no-binary
+#~/GitHub/Mario/cadical/build/cadical $BASE.icnf $BASE.lrat --no-binary --lrat=true
 cat $BASE.drup | grep -v "d " > $BASE.rup
 ../crat-lrat/crat-lrat $BASE.acnf $BASE.rup -f | grep -v "c " | grep " 0" > result.txt
 cat result.txt

### 2023-03-09 doc/mc-data.tex

@@
-\documentclass{llncs}
-\usepackage{latexsym}
-\usepackage{times}
+\documentclass[letterpaper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+%This is a template for producing LIPIcs articles. 
+%See lipics-v2021-authors-guidelines.pdf for further information.
+%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
+%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
+%for section-numbered lemmas etc., use "numberwithinsect"
+%for enabling cleveref support, use "cleveref"
+%for enabling autoref support, use "autoref"
+%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
+%for enabling thm-restate support, use "thm-restate"
+%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
+%for producing a PDF according the PDF/A standard, add "pdfa"
+
+%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
+%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+
+%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+
 \usepackage{amsmath}
-\usepackage{graphicx}
 \usepackage{tikz}
 \usepackage{pgfplots}
-\usepackage{cite}
 \usepackage{booktabs}
-\usepackage{adjustbox}
-\usepackage{marvosym}
+
+\newcommand{\pand}{\mathbin{\land^{\sf p}}}
+\newcommand{\por}{\mathbin{\lor^{\sf p}}}
+\DeclareMathOperator*{\Pand}{\bigwedge^{\sf v}}
+\DeclareMathOperator*{\Por}{\bigvee^{\sf a}}
+\newcommand{\boolnot}{\neg}
+\newcommand{\tautology}{\top}
+\newcommand{\nil}{\bot}
+\newcommand{\obar}[1]{\overline{#1}}
+\newcommand{\oneminus}{{\sim}}
+\newcommand{\lit}{\ell}
+
+\newcommand{\varset}{X}
+\newcommand{\exvarset}{Z}
+\newcommand{\dependencyset}{{\cal D}}
+\newcommand{\litset}{{\cal L}}
+\newcommand{\ring}{{\cal R}}
+\newcommand{\dset}{{\cal A}}
+\newcommand{\rep}{\textbf{R}}
+\newcommand{\radd}{+}
+\newcommand{\rmul}{\times}
+\newcommand{\addident}{\textbf{0}}
+\newcommand{\mulident}{\textbf{1}}
+\newcommand{\imply}{\Rightarrow}
+
+\newcommand{\assign}{\alpha}
+\newcommand{\passign}{\rho}
+\newcommand{\uassign}{{\cal U}}
+\newcommand{\modelset}{{\cal M}}
+
+\newcommand{\indegree}{\textrm{indegree}}
+\newcommand{\outdegree}{\textrm{outdegree}}
+\newcommand{\validate}{\textsf{validate}}
+
+\newcommand{\makenode}[1]{\mathbf{#1}}
+\newcommand{\nodeu}{\makenode{u}}
+\newcommand{\nodev}{\makenode{v}}
+\newcommand{\nodes}{\makenode{s}}
+\newcommand{\nodep}{\makenode{p}}
+\newcommand{\noder}{\makenode{r}}
+
+\newcommand{\simplify}[2]{#1|_{#2}}
+\newcommand{\prov}{\mathit{Prov}}
+
+\newcommand{\progname}[1]{\textsc{#1}}
+\newcommand{\dfour}{\progname{d4}}
+\newcommand{\Dfour}{\progname{d4}}
+\newcommand{\generator}{\progname{d2p}}
+\newcommand{\Generator}{\progname{d2p}}
+\newcommand{\checker}{\progname{crat-check}}
+\newcommand{\Checker}{\progname{Crat-check}}
+
 %% Colored hyperlink 
-\newcommand{\hcref}[1]{\texttt{\small\url{#1}}}
+%\newcommand{\cref}[2]{\href{#1}{\color{blue}#2}}
+%% Colored hyperlink showing link in TT font
+% \newcommand{\chref}[1]{\href{#1}{\small\tt \color{blue}#1}}
+\newcommand{\hcref}[1]{\cref{#1}{\small\tt #1}}
 
-\bibliographystyle{splncs04}
-%\bibliographystyle{abbrv}
+\bibliographystyle{plainurl}% the mandatory bibstyle
+
+\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting}
+
+\titlerunning{Certified Knowledge Compilation}
+
+\author{Randal E.
+Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{rebryant@cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}
+\author{Wojciech Nawrocki}{Department of Philosophy, Carnegie Mellon University}{wjnawrock@cmu.edu}{https://orcid.org/0000-0002-8839-0618}{}
+\author{Jeremy Avigad}{Department of Philosophy, Carnegie Mellon University}{avigad@cmu.edu}{https://orcid.org/0000-0003-1275-315X}{}
+\author{Marijn J. H. Heule}{Computer Science Department, Carnegie Mellon University}{marijn@cmu.edu}{https://orcid.org/0000-0002-5587-8801}{Supported by NSF grant CCF-2108521}
+
+\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+
+%\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+
+%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+
+%\category{} %optional, e.g. invited paper
+
+%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
+Please enter author specific funding statements as fifth argument of the \author macro.
+
+%\nolinenumbers %uncomment to disable line numbering
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\EventEditors{John Q. Open and Joan R. Access}
+%\EventNoEds{2}
+%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+%\EventShortTitle{CVIT 2016}
+%\EventAcronym{CVIT}
+%\EventYear{2016}
+%\EventDate{December 24--27, 2016}
+%\EventLocation{Little Whinging, United Kingdom}
+%\EventLogo{}
+%\SeriesVolume{42}
+%\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\program}[1]{\textsc{#1}}
+\newcommand{\lean}{\program{lean4}}
+
+\begin{document}
+
+\maketitle
 
@@
 
-\title{Validated Model Counting Data \\ Version of \today}
-
-\author{Randal E. Bryant}
-
-\authorrunning{R. E. Bryant}
-
-\titlerunning{Validated Model Counting}
-
-\institute{
-Computer Science Department \\
-Carnegie Mellon University, Pittsburgh, PA, United States
-}
-
-\begin{document}
-
-\maketitle
-
 \section*{Benchmark machine}
@@
      \begin{center}
-       \hcref{https://mccompetition.org/2022/mc\_description.html}
+       \hcref{https://mccompetition.org/2022/mc-description.html}
      \end{center}
@@
 
+\section{Comparison with MICE}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 1.00]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=1000,
+                              x post scale=2.0, y post scale=2.2,
+                              xmode=log,xmin=0.01,xmax=100,
+                              xtick={0.01, 0.1,1.0,10,100}, xticklabels={0.01, 0.1, 1.0, 10, 100},
+                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
+                              xlabel={CRAT total time (seconds)}, ylabel={MICE total time (seconds)},
+            ]
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
+    \input{data-formatted/time-mice-compare}
+    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
+    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
+    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
+    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
+    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Running Time for MICE versus CRAT proof chains}
+\label{fig:mice}
+\end{figure}
+
 \bibliography{references}

### 2023-03-09 doc/mc-notes.tex

@@
     \midrule
-    $C$    & {\tt p} & $V \; L^{2+}$ {\tt 0}    &                  & Declare $\pand$ operation \\
+    $C$    & {\tt p} & $V \; L^{*}$ {\tt 0}    &                  & Declare $\pand$ operation \\
     $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation.  Explicit hint \\ 
@@
      & {\tt do} & $V$             &  & Delete operation. \\
-
+    \midrule
+     & {\tt r} & $L$             &            & Declare root literal\\
   \end{tabular}
@@
 \end{center}
-The dependency sets for the arguments represented by each pair of literals $l_{j_1}$ and $l_{j_2}$, for $1 \leq j_1 < j_2 \leq k$, must be disjoint.
+The dependency sets for the arguments represented by each pair of
+literals $l_{j_1}$ and $l_{j_2}$, for $1 \leq j_1 < j_2 \leq k$, must
+be disjoint.  Note that a $\pand$ operation can have no arguments,
+representing $\tautology$.  The only clause added to the proof will be
+the unit literal $v$.  A reference to literal $-v$ then becomes a way
+of representing $\nil$.
 
@@
 
-Finally, operator deletion ({\tt do}) specifies which operator to
+Operator deletion ({\tt do}) specifies which operator to
 delete via the extension variable associated with the operator.  Doing
@@
 having a literal of $v$ as one of its inputs.  The defining clauses for the operation are also deleted.
+
+Finally, the literal denoting the root of the POG is declared with the
+{\tt r} command.  It can occur anywhere in the file.  Typically, it
+will be the extension node representing the root of a graph, but in
+degenerate cases it can be the positive or negative literal of one of
+the input variables.
 
@@
 complement.  Except for trivial cases, the final literal will be an
-extension variable or its complement.  That literal will indicate the
-root of the POG as the (possibly negated) output of a POG
-operation.  Working from that root backward, the POG can be
-extracted from the CRAT file.
+extension variable or its complement.  That literal must match the one declared as the root literal.
 
@@
     {\tt 11}         & {\tt s}   & {\tt 7 1 6}   & {\tt 9 0}  & & Declare $s_7 = x_1 \por p_6$ \\
+             & {\tt r}  & {\tt 7} & && Declare root literal \\
     {\tt 14} & {\tt a}  & {\tt 7 0} & {\tt 13 12 8 7 6 2 1 0} & & Assert unit clause $[s_7]$ \\
@@
     {\tt 12} & {\tt s}   & {\tt 6 4 5}   & {\tt 4 9 0}  & & Declare $s_6 = \obar{t}_1 \por \obar{t}_2$ \\
+             & {\tt r}  & {\tt -6} & && Declare root literal \\
     {\tt 15} & {\tt a}  & {\tt -6 0}       & {\tt 7 11 12 0} & & Assert $t_3 = \obar{s}_6 = t_1 \land t_2$ \\
@@
     {\tt 12} & {\tt s}   & {\tt 7 5 6}   & {\tt 7 10 0}  & & Declare $s_7 = p_5 \por p_6$ \\
+             & {\tt r}  & {\tt 7} & && Declare root literal \\
     {\tt 15} & {\tt a}  & {\tt 1 -4 0}    & {\tt 4 5 1 0} & & Justify $\obar{x}_1 \rightarrow \obar{p}_4$ \\

### 2023-03-10 doc/mc-paper.tex

@@
 
-\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{R. E. Bryant et al.} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
@@
 \begin{abstract}
-  Knowledge compilers convert Boolean formulas into a form 
-  that enables efficient evaluation of properties of the formula,
-  including
-  weighted and unweighted model counting.
- Existing knowledge compilers, however,
-provide no guarantee that the generated representation accurately
-describes the original formula.
-
-We present the Partitioned Operation Graph (POG) representation that
-generalizes existing knowledge compiler representations, and a proof
-framework that can prove the equivalence of a POG representation with
-a conjunctive normal form (CNF) Boolean formula.
-
-We have developed a program that can generate POG representations and
-equivalence proofs from the d-DNNF graphs generated by a leading
-knowledge compiler.  We also present two proof
-checkers---a fast checker that can handle the large proofs generated,
-and a formally verified checker written in the \lean{} proof system.
-Our generation and checking tool chain scales to all but the largest formulas from a
-recent model counting competition.  
+\emph{Knowledge compilation} is the process of converting a Boolean
+formula into a form that makes it possible to calculate associated
+values efficiently, such as the weighted or unweighted model
+count. Existing knowledge compilers, however,
+provide no guarantee that the representation they generate
+is logically equivalent to the original formula, and therefore they cannot validate the generated counts.
+
+We present \emph{Partitioned Operation Graphs} (POGs), a generalization
+of representations commonly used in knowledge compilation.
+We have designed a proof framework that can prove the equivalence of a
+POG representation with a Boolean formula in conjunctive normal form (CNF),
+the standard input format for SAT solvers and model counting tools.
+
+We have developed a program that can generate POG representations
+from the d-DNNF graphs generated by the state-of-the-art knowledge
+compiler D4, as well as independently checkable proofs that the
+output POGs are equivalent to the input CNF formulas.
+We have implemented an efficient checker
+that can handle very large proofs. Our tool
+chain for generating and verifying POGs scales to all but the largest d-DNNF graphs
+generated by D4 for formulas
+from a recent model counting competition. Moreover, we
+present a checker that has been formally verified
+in the Lean~4 proof assistant, thereby creating the
+first formally verified toolchain for weighted and unweighted model
+counting.
+
+%  Knowledge compilers convert Boolean formulas into a form 
+%  that enables efficient evaluation of properties of the formula,
+%  including
+%  weighted and unweighted model counting.
+% Existing knowledge compilers, however,
+%provide no guarantee that the generated representation accurately
+%describes the original formula.
+%
+%We present the Partitioned Operation Graph (POG) representation that
+%generalizes existing knowledge compiler representations, and a proof
+%framework that can prove the equivalence of a POG representation with
+%a conjunctive normal form (CNF) Boolean formula.
+%
+%We have developed a program that can generate POG representations and
+%equivalence proofs from the d-DNNF graphs generated by a leading
+%knowledge compiler.  We also present two proof
+%checkers---a fast checker that can handle the large proofs generated,
+%and a formally verified checker written in the \lean{} proof system.
+%Our generation and checking tool chain scales to all but the largest formulas from a
+%recent model counting competition.  
 \end{abstract}

### 2023-03-10 doc/mc-data.tex
Incorporated MICE data; added graphs to paper

@@
 \definecolor{lightgreen}{rgb}{0.709804, 0.741176, 0}
-\definecolor{darkgreen}{rgb}{0.152941, 0.776471, 0.172549}
+\definecolor{darkgreen}{rgb}{0.152941, 0.576471, 0.172549}
 \definecolor{redpurple}{rgb}{0.835294, 0, 0.196078}
@@
 \centering{%
-\begin{tikzpicture}[scale = 1.00]
+\begin{tikzpicture}[scale = 0.90]
   \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=3000000000,
@@
       legend cell align={left},
-                              x post scale=2.0, y post scale=2.0,
+                              x post scale=1.8, y post scale=1.8,
                               xmode=log, xmin=0.01,xmax=4000,
@@
             ]
+    \input{data-formatted/time+defining-success}
+    \input{data-formatted/time+defining-onesided-only}
     \input{data-formatted/time+defining-failure}
-    \input{data-formatted/time+defining-success}
     \legend{
-      \scriptsize \textsf{Limit exceeded},
-      \scriptsize \textsf{Validated},
+      \scriptsize \textsf{Full validation},
+      \scriptsize \textsf{One-sided validation},
+      \scriptsize \textsf{Resource limit exceeded},
     }
@@
 \centering{%
-\begin{tikzpicture}[scale = 1.00]
+\begin{tikzpicture}[scale = 0.90]
   \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
-                              x post scale=2.0, y post scale=2.0,
+                              x post scale=1.8, y post scale=2.15,
                               xmode=log,xmin=100,xmax=1000000000, 
@@
 \label{fig:defining:total}
-
 \end{figure}
@@
 \centering{%
-\begin{tikzpicture}[scale = 1.00]
+\begin{tikzpicture}[scale = 0.90]
   \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=20000,
-                              x post scale=2.0, y post scale=2.2,
+      legend style={at={(0.30,0.98)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.4,
                               xmode=log,xmin=0.01,xmax=4000,
@@
             ]
+    \input{data-formatted/time-d4-verify}
+    \input{data-formatted/time-d4-verify-onesided-only}
+    \legend{
+      \scriptsize \textsf{Full validation},
+      \scriptsize \textsf{One-sided validation},
+    }
     \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
@@
     \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
-    \input{data-formatted/time-d4-verify}
     \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
@@
 \label{fig:d4:crat}
-
 \end{figure}
@@
 \begin{figure}
+\begin{minipage}{0.49\textwidth}
 \centering{%
-\begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
+\begin{tikzpicture}[scale = 0.80]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
+      legend style={at={(0.50,0.98)}},
+      legend cell align={left},
+                              x post scale=1.1, y post scale=1.8,
+                              xmode=log,xmin=1000,xmax=1e7, 
+                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
+                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
+%                              xlabel={Optimized Proof Clauses},
+%                              ylabel={Proof Clauses},
+            ]
+    \input{data-formatted/clauses-lemma-nolemma}
+    \input{data-formatted/clauses-nosplit-split}
+    \legend{
+      \scriptsize \textsf{No lemmas},
+      \scriptsize \textsf{No literal grouping},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
+
+    \node[left] at (axis cs: 5000,7000) {$1\times$};
+    \node[left] at (axis cs: 5000,70000) {$10\times$};
+    \node[left] at (axis cs: 5000,700000) {$100\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\centering{%
+\begin{tikzpicture}[scale = 0.80]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
+      legend style={at={(0.50,0.98)}},
+      legend cell align={left},
+                              x post scale=1.1, y post scale=1.8,
+                              xmode=log,xmin=1000,xmax=1e7, 
+                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
+                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
+%                              xlabel={Optimized Proof Clauses},y
+%                              ylabel={Proof Clauses},
+            ]
+    \input{data-formatted/clauses-optimized-unoptimized}    
+    \legend{
+      \scriptsize \textsf{Neither optimization},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
+
+    \node[left] at (axis cs: 5000,7000) {$1\times$};
+    \node[left] at (axis cs: 5000,70000) {$10\times$};
+    \node[left] at (axis cs: 5000,700000) {$100\times$};
+  \end{axis}
+\end{tikzpicture}
+} % centering
+\end{minipage}
+\caption{Proof clauses when one (left) or both (right) optimizations are disabled, versus when both optimizations are enabled}
+\label{fig:optimized:lessoptimized}
+\end{figure}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.90]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
       legend style={at={(0.40,0.98)}},
       legend cell align={left},
-                              x post scale=2.0, y post scale=2.3,
+                              x post scale=1.4, y post scale=2.3,
                               xmode=log,xmin=1000,xmax=1e7, 
@@
 \centering{%
-\begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=1e9, 
+\begin{tikzpicture}[scale = 0.90]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
       legend style={at={(0.40,0.98)}},
       legend cell align={left},
-                              x post scale=2.0, y post scale=2.3,
+                              x post scale=1.4, y post scale=2.3,
                               xmode=log,xmin=1000,xmax=1e7, 
@@
     \input{data-formatted/clauses-optimized-unoptimized}    
-    \legend{
-      \scriptsize \textsf{Without either optimization},
-    }
     \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
@@
 proof clauses when both optimizations are used.  Without lemmas, 19
-rise above that threshold.  Without literal merging, 7 due.  Without
+rise above that threshold.  Without literal merging, 7 do.  Without
 either, 25 rise above the threshold.  Two of the benchmarks require
@@
 \centering{%
-\begin{tikzpicture}[scale = 1.00]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=1000,
-                              x post scale=2.0, y post scale=2.2,
-                              xmode=log,xmin=0.01,xmax=100,
-                              xtick={0.01, 0.1,1.0,10,100}, xticklabels={0.01, 0.1, 1.0, 10, 100},
+\begin{tikzpicture}[scale = 0.90]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=1500,
+      legend style={at={(0.22,0.98)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.2,
+                              xmode=log,xmin=0.01,xmax=1500,
+                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100,{1,000}},
                               ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
@@
             ]
+    \input{data-formatted/time-mice-compare}
+    \input{data-formatted/time-mice-error}
+    \legend{
+      \scriptsize \textsf{Correct or timeout},
+      \scriptsize \textsf{MICE error},
+    }
+    \input{data-formatted/time-mice-timeout}
+%    \addplot[mark=none, color=black] coordinates{(0.01,1500) (1500,1500)};
+%    \addplot[mark=none, color=black] coordinates{(1500,0.01) (1500,1500)};
+    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (10000.0,10.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (10000.0,100.0)};
     \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
@@
     \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
-    \input{data-formatted/time-mice-compare}
-    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
-    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
-    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
-    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
-    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
+    \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
+    \node[left] at (axis cs: 5.0,0.06) {$0{.}01\times$};
+    \node[left] at (axis cs: 0.5,0.06) {$0{.}1\times$};
+    \node[left] at (axis cs: 0.05,0.06) {$1\times$};
+    \node[left] at (axis cs: 0.05,0.6) {$10\times$};
+    \node[left] at (axis cs: 0.05,6.0) {$100\times$};
+    \node[left] at (axis cs: 0.05,60.0) {$1000\times$};
 
@@
 
+For comparison, we also ran the MICE tool chain for certifying model
+counters~\cite{fichte:sat:2022}.  Figure~\ref{fig:mice} compares the
+runtimes of our toolchain versus theirs.  The X-axis shows the sum of
+the elapsed times for our generator, checker, and counter, while the
+Y-axis shows the sum of the elapsed times for \textsc{nnf2trace},
+generating a file in their trace format from a d-DNNF graph generated
+by D4, and for \textsc{sharptrace}, a checker for the trace format.
+The plot shows data points for those benchmarks for at least one of
+the two tools completed in less than 1000 seconds.
+
+We encountered a significant number of errors with both
+\textsc{nnf2trace} and \textsc{sharptrace}.  Based on the error
+messages, these appear to be mostly due to features of some of the
+d-DNNF graphs they had not anticipated.  We provided a bug report to
+the authors, but they were unable to fix them on short notice.
+
+In the figure, red dots indicate cases where neither program had an
+error, but it possibly timed out.  The time-out cases are shown along
+the upper edge (MICE) or the righthand edge (CRAT).
+
+As can be seen, there is not a strong correlation between the program
+runtimes.  This is, in part, because the two tools perform different
+roles: one validates the algorithmic steps used by a top-down model
+counting problem, while the other verifies the logical equivalence of
+a Boolean formula and a representation for which model counting can be
+done efficiently.
+
+Overall, we can summarize the results as:
+\begin{itemize}
+\item 85 total data points
+\item 64 were completed by both programs in under 1000 seconds
+\item Of these, the MICE toolchain was faster for 20, and the CRAT toolchain was faster for 45
+\item 5 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
+\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
+\item 15 completed with CRAT but generated errors for MICE.
+\end{itemize}
+
 \bibliography{references}
@@
 \centering{%
-\begin{tikzpicture}[scale = 1.00]
+\begin{tikzpicture}[scale = 0.90]
   \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
-                              x post scale=2.0, y post scale=2.0,
+                              x post scale=1.8, y post scale=1.8,
                               xmin=0,xmax=1000,

### 2023-03-10 doc/mc-paper.tex
Incorporated MICE data; added graphs to paper

@@
 \newcommand{\imply}{\Rightarrow}
+\newcommand{\drational}{\mathbb{Q}_{2,5}}
 
@@
 \newcommand{\progname}[1]{\textsc{#1}}
-\newcommand{\dfour}{\progname{d4}}
-\newcommand{\Dfour}{\progname{d4}}
+\newcommand{\dfour}{\progname{D4}}
+\newcommand{\Dfour}{\progname{D4}}
 \newcommand{\generator}{\progname{d2p}}
-\newcommand{\Generator}{\progname{d2p}}
+\newcommand{\Generator}{\progname{D2p}}
 \newcommand{\checker}{\progname{crat-check}}
 \newcommand{\Checker}{\progname{Crat-check}}
+\newcommand{\cadical}{\progname{CaDiCal}}
+\newcommand{\dtrim}{\progname{drat-trim}}
+
+\definecolor{redorange}{rgb}{0.878431, 0.235294, 0.192157}
+\definecolor{lightblue}{rgb}{0.552941, 0.72549, 0.792157}
+\definecolor{clearyellow}{rgb}{0.964706, 0.745098, 0}
+\definecolor{clearorange}{rgb}{0.917647, 0.462745, 0}
+\definecolor{mildgray}{rgb}{0.54902, 0.509804, 0.47451}
+\definecolor{softblue}{rgb}{0.643137, 0.858824, 0.909804}
+\definecolor{bluegray}{rgb}{0.141176, 0.313725, 0.603922}
+\definecolor{lightgreen}{rgb}{0.709804, 0.741176, 0}
+\definecolor{darkgreen}{rgb}{0.152941, 0.576471, 0.172549}
+\definecolor{redpurple}{rgb}{0.835294, 0, 0.196078}
+\definecolor{midblue}{rgb}{0, 0.592157, 0.662745}
+\definecolor{clearpurple}{rgb}{0.67451, 0.0784314, 0.352941}
+\definecolor{browngreen}{rgb}{0.333333, 0.313725, 0.145098}
+\definecolor{darkestpurple}{rgb}{0.396078, 0.113725, 0.196078}
+\definecolor{greypurple}{rgb}{0.294118, 0.219608, 0.298039}
+\definecolor{darkturquoise}{rgb}{0, 0.239216, 0.298039}
+\definecolor{darkbrown}{rgb}{0.305882, 0.211765, 0.160784}
+\definecolor{midgreen}{rgb}{0.560784, 0.6, 0.243137}
+\definecolor{darkred}{rgb}{0.576471, 0.152941, 0.172549}
+\definecolor{darkpurple}{rgb}{0.313725, 0.027451, 0.470588}
+\definecolor{darkestblue}{rgb}{0, 0.156863, 0.333333}
+\definecolor{lightpurple}{rgb}{0.776471, 0.690196, 0.737255}
+\definecolor{softgreen}{rgb}{0.733333, 0.772549, 0.572549}
+\definecolor{offwhite}{rgb}{0.839216, 0.823529, 0.768627}
+\definecolor{medgreen}{rgb}{0.15, 0.6, 0.15}
 
@@
 
-\authorrunning{R. E. Bryant et al.} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
@@
 \newcommand{\program}[1]{\textsc{#1}}
-\newcommand{\lean}{\program{lean4}}
+\newcommand{\lean}{Lean~4}
 
@@
 \begin{abstract}
+
 \emph{Knowledge compilation} is the process of converting a Boolean 
@@
 from the d-DNNF graphs generated by the state-of-the-art knowledge 
-compiler D4, as well as independently checkable proofs that the
+compiler \dfour, as well as independently checkable proofs that the 
 output POGs are equivalent to the input CNF formulas.
@@
 chain for generating and verifying POGs scales to all but the largest d-DNNF graphs
-generated by D4 for formulas
+generated by \dfour for formulas
 from a recent model counting competition. Moreover, we 
 present a checker that has been formally verified
-in the Lean~4 proof assistant, thereby creating the
-first formally verified toolchain for weighted and unweighted model
-counting.
-
-%  Knowledge compilers convert Boolean formulas into a form 
-%  that enables efficient evaluation of properties of the formula,
-%  including
-%  weighted and unweighted model counting.
-% Existing knowledge compilers, however,
-%provide no guarantee that the generated representation accurately
-%describes the original formula.
-%
-%We present the Partitioned Operation Graph (POG) representation that
-%generalizes existing knowledge compiler representations, and a proof
-%framework that can prove the equivalence of a POG representation with
-%a conjunctive normal form (CNF) Boolean formula.
-%
-%We have developed a program that can generate POG representations and
-%equivalence proofs from the d-DNNF graphs generated by a leading
-%knowledge compiler.  We also present two proof
-%checkers---a fast checker that can handle the large proofs generated,
-%and a formally verified checker written in the \lean{} proof system.
-%Our generation and checking tool chain scales to all but the largest formulas from a
-%recent model counting competition.  
+in the \lean{} proof assistant, thereby creating the
+first formally verified toolchain for
+% weighted and unweighted
+model counting.
 \end{abstract}
@@
 \section{Introduction}
+
+Modern Boolean satisfiability (SAT) solvers provide the ability to
+find a satisfying assignment to a Boolean formula or to prove that no
+such assignment exists.  The have applications across a variety of
+domains, including formal verification of hardware, software, and
+security protocols, computational mathematics, and combinatorial
+optimization.  Some problems of interest, however, require going
+beyond standard Boolean satisfiability.  For example, the {\em model
+  counting problem} requires computing the number of satisfying
+assignments, often in cases where there are far to many satisfying
+assignments to enumerate individually.  Model counting, in turn, has
+applications in artificial intelligence, computer security, and
+statistical sampling.  A variant on standard model counting is {\em
+  weighted model counting}, where a weighting function is defined for
+each assignment and the goal becomes to compute the sum of the weights
+of the satisfying assignments.
+
+Model counting is a challenging problem---more challenging that the
+already NP-hard problem of Boolean (un)satisfiability.  Several
+tractable variants of Boolean satisfiability, including 2-SAT, become
+intractable when the goal becomes model counting and not just
+satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
+model counters have developed that scale to very large formulas, as
+witnessed by their progress in the annual model counting competitions.
+
+One approach to model counting, known as {\em knowledge compilation}
+is to transform the formula into a structured form for which model
+counting becomes more tractable.  For example, the {\em deterministic,
+  decomposable, negation normal form} (d-DNNF) introduced by
+Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a Boolean formula as a
+directed acyclic graph, with terminal nodes labeled by Boolean
+variables and their complements, and with each nonterminal node
+labeled by Boolean operation And or Or.  Restrictions are place on the
+structure of the graph such that a count of the models can be computed
+by a single bottom-up traversal of the graph.
+
+Once a Boolean formula has been transformed into a d-DNNF
+representation, a number of useful properties of the formula can
+readily be computed.  Kimmig, et al., present a very general
+framework, known as {\em algebraic model
+counting}~\cite{kimmig:jal:2017}, that go well beyond simple model counting.
+
+One shortcoming of existing knowledge compilers is that they have no
+way to validate that their generated results are corrected, i.e., that
+the compiled representation is logically equivalent to the original
+formula.  By contrast, all modern SAT solvers are able to produce
+checkable proofs when they encounter unsatisfiable formulas.  The
+guarantee provided by a checkable certificate of correctness enables
+users of SAT solvers to fully trust their results.  Experience has
+shown that adding this capability has also greatly improved the quality of
+SAT solvers and the ability of developers to quickly detect and
+identify bugs in their solvers.
+
+In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
+a generalization of the representations commonly used in knowledge
+compilation, including d-DNNF\@.  We present the CRAT file format that
+both captures the structure of a POG and can provide a checkable proof
+that the POG is logically equivalent to a Boolean formula given in
+conjunctive normal form (CNF).  A proof is expressed as a sequence of
+clause addition and deletion steps, based on the extended resolution
+proof system~\cite{Tseitin:1983}.
+
+We have developed a program that can generate CRAT descriptions from
+the d-DNNF graphs generated by the state-of-the-art knowledge compiler
+\dfour, as well as a fast CRAT proof checker.  We evaluate the
+capabilities of our tool chain using benchmark formulas from the 2022
+standard and weighted model competitions.  Our tools are able to
+handle all but the largest d-DNNF graphs generated by \dfour.
+Moreover, we 
+present a checker that has been formally verified
+in the \lean{} proof assistant, thereby creating the
+first formally verified toolchain for
+% weighted and unweighted
+model counting.
+
+\section{Related Work}
+
+Having a SAT solver generate a proof of unsatisfiability has a long
+tradition~\cite{ZhangMalik} and has become widely accepted due to the
+formulation of clausal proof systems for which proofs can readily be
+generated and efficiently checked
+\cite{cruz-cade-2017,RAT,wetzler14_drattrim}.  Some progress has also
+been made on extending these proof systems to quantified Boolean
+formulas~\cite{bryant:cade:2021,heule:JAR2014}.
+
+Fichte, Hecher, and Roland~\cite{fichte:sat:2021} devised the MICE
+proof framework for model counting programs.  Their proof rules are
+based on the algorithmic steps commonly used by model counting
+programs, and they present a series of lemmas and a theorem stating
+that their checking framework is sound and complete.  They also
+present experimental results for cases where they modify existing
+model counters to also generate proof traces, and where they convert
+d-DNNF graphs into a checkable form.  We compare the performance of
+our tool chain to theirs as part of our experimental evluation in
+Section~\ref{sect:experimental}.  Our experience with developing a
+formally verified proof system has shown us that, even for the well
+understood framework of extended resolution, ensuring that the system
+is sound can be challenging.
+
+We know of know other work addressing the more general question of
+certifying the output of a knowledge compiler.
 
@@
   For ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
-  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $1-w(x)$.
+  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $\mulident-w(x)$.
 
@@
 rational numbers.  Some formulations of WMC allow 
-indepdently assigning weights $W(x)$ and $W(\obar{x})$ to each variable $x$, potentially having
+independently assigning weights $W(x)$ and $W(\obar{x})$ for variable $x$ and its complement, with the possibility that
 $W(x) + W(\obar{x}) \not = 1$.
@@
 Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
-\begin{description}
-\item[Negation] $\rep(\boolnot \phi, w) = \mulident - \rep(\phi, w)$
-\item[Partitioned Product] $\rep\left(\Pand_{i=1,k} \phi_i, w \right) = \prod_{i=1,k} \rep(\phi_i, w)$
-\item[Partitioned Sum] $\rep\left(\Por_{i=1,k} \phi_i, w\right) = \sum_{i=1,k} \rep(\phi_i, w)$
-\end{description}
+\begin{eqnarray}
+\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
+\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum} 
+\end{eqnarray}
 \end{proposition}
@@
 \end{minipage}
-\caption{Example formula (A), its POG representation (B), and its CRAT proof (C)--(F)}
+\caption{Example formula (A), its POG representation (B), and its CRAT proof (C), (E), and (F)}
 \label{fig:eg4:all}
@@
   do so by writing the formula to a file, invoking a proof-generating
-  SAT solver, and then importing the generated unsatisfiability prove
-  as assertions in the CRAT file.
+  SAT solver, and then converting the generated unsatisfiability proof
+  into a sequence of clause additions in the CRAT file.
 \item If there is a single nonterminal child ($k = 1$), we recursively call
@@
 Observe that most of these steps involve a polynomial number of
-operations per recursive call.  The only potential source of
-exponential growth is when a SAT solver must be invoked to validate a
+operations per recursive call.  The only step that could have
+exponential complexity is when a SAT solver must be invoked to validate a
 literal.
@@
 CRAT framework by first deleting all asserted clauses, except for the
-final unit clause for root literal $r$ and then deleting all of the
+final unit clause for root literal $r$, and then deleting all of the
 input clauses.  Suppose at some point in this process that we want to
@@
 clauses of the POG nodes via a single, bottom-up pass through the
-graph.  A RUP proof for input clause $C$ effectively proves that any
+graph.  The RUP deletion proof for input clause $C$ effectively proves that any
 total assignment $\assign$ that does not satisfy $C$ will yield
@@
 
-Working upward through the graph, each node $\nodeu$ is {\em marked} when
-the collected set of literals force $u$ to evaluate to $0$.  The
+Working upward through the graph, node $\nodeu$ is {\em marked} when
+the collected set of literals force $u$ to evaluate to $0$.  When marking $\nodeu$, the
 program adds $\obar{u}$ to the RUP literals and adds the appropriate
@@
 $\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
-added to the hints.  Marking sum node $\nodeu$ requires that both of
-is children $\nodeu_1$ and $\nodeu_2$ be marked.  The first defining
-clause for this node (see Table~\ref{tab:defining}(B)) will then be added.  At the very end, the program
+added to the hints.  Marking sum node $\nodeu$ requires that its two children are marked.
+The first defining
+clause for this node (see Table~\ref{tab:defining}(B)) will then be added as a hint.  At the very end, the program
 (assuming the reverse implication holds) will attempt to mark root
@@
 It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
-each clause deletion requires a single RUP step, where 
-the number of hints will be less than the number of
-nodes in the POG\@.
+each clause deletion requires a single RUP step having a set of hints that is
+bounded by the number of POG nodes.
 
@@
 the size of the proof generated by two optimizations described here.
-A key feature of both is that they do not require any changes to the
+A key feature of these optimizations is that they do not require any changes to the
 proof framework---they build on the power of extended resolution to
@@
 A single recursive step of $\validate$ can encounter product nodes
-having many---tens or even hundreds---of literal as children.  As described
-earlier, the standard approach for each child literal $\lit \in \L$ is to first attempt BCP, but calling a SAT solver
-on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$.
-
-An alternative approach is to collect those literals $\lit_1, \lit_2,
-\ldots, \lit_m$ that cannot be validated by BCP and then define a
+having many---tens or even hundreds---of literals as children.  The earlier formulation of $\validate$
+considered each literal $\lit \in L$ separately,
+calling a SAT solver
+on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ when $\lit$ could not be validated with BCP\@.
+
+Literal grouping allows all literals to be validated with a single call to the SAT solver.
+It collects those literals
+$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
 product node $\nodev$ having these literals as children.  The goal
-then becomes to prove that this product must yield 1 for extension
-variable $v$.  This can be done by contradiction using a SAT solver.
+then becomes to prove that any total assignment must yield 1 for extension
+variable $v$.  This can be done by contradiction using the SAT solver.
 In particular, looking at the defining clauses for $\nodev$ (Table
 \ref{tab:defining}(A)), we can see that assigning 0 to $v$ will cause
-  the first defining clause to simplify to $C_p = \obar{lit}_1 \lor
-  \obar{\lit}_2 \lor \cdots \lor \obar{\lit}_m$, and for the remaining
-  clauses to become tautologies.  The formula
-  $\simplify{\psi}{\passign} \cup \{ C_p \}$ should therefore be
-  unsatisfiable.  The proof steps and hints generated by the SAT
-  solver can mapped back to the input clauses, plus the defining
-  clauses for $\nodev$.
+the first defining clause to simplify to
+$C = \obar{\lit}_1 \lor  \obar{\lit}_2 \lor \cdots \lor \obar{\lit}_m$, and for the remaining
+  clauses to become tautologies.  The SAT solver can be invoked on formula
+  $\simplify{\psi}{\passign} \cup \{ C \}$ to show that is unsatisfiable.
+  The proof steps can be mapped back into clause addition steps in the CRAT file, using as hints the
+  input clauses and the defining clauses for $\nodev$.
 
@@
 compiler arises when the program detects that the simplified set of
-clauses arising from a partial assignment to the literals matches that
-of a previous traversal.  When this d-DNNF node is translated into POG
-node $\nodeu$, the proof generator can assume (and later check), that
+clauses arising from one partial assignment to the literals matches that
+of a previous partial assignment.  When this d-DNNF node is translated into POG
+node $\nodeu$, the proof generator can assume (and also check), that
 there is a simplified set of clauses $\gamma_{\nodeu}$
@@
 $\nodeu$ as a child.  The net effect is to allow $\validate$ to
-maintain the graph structure of the POG, invoking $\validate$ on each
+exploit the graph structure of the POG, invoking $\validate$ on each
 node at most once.
@@
 The challenge for implementing this strategy is to find a way to
-represent the clauses in a simplified formula $\gamma$.  Some may be
+represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CRAT file.  Some may be
 unalterered input clauses, and these can be used directly.  Others,
@@
 implement these by adding POG product nodes to the CRAT file to create
-the appropriate clauses.  Imagine that there is an {\em argument} clause
-$C \in \gamma$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  If we
+the appropriate clauses.  Consider an {\em argument} clause
+$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
 define product node $\nodev$ with arguments
@@
 
-Suppose for every clause $C_j \in \gamma$ that does not correspond to
+Suppose for every clause $C_j \in \gamma_{\nodeu}$ that does not correspond to
 an input clause, we generate a synthetic clause $C'_j$ with guard literal
-$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
-and
-let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
-Invoking $\validate(\nodeu, \lassign, \gamma)$
+$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma'_{\nodeu}$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
+while input clauses in $\gamma_{\nodeu}$ are left unchanged.
+Let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
+Invoking $\validate(\nodeu, \lassign, \gamma'_{\nodeu})$
  will then prove a lemma, given by the target clause
  $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
- showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u)$.
+ showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u) = 1$.
 
@@
 $C_j$ matches some simplified clause in $\simplify{\psi}{\passign}$.  More precisely,
-we assert the clause
-$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$
-to show that synthetic clause $C_j$ will be activated for $1 \leq j \leq m$.
+for $1 \leq j \leq m$,
+we use clause addition to assert the clause
+$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$,
+showing that synthetic clause $C_j$ will be activated.
+Combining the lemma with these activations provides a derivation of the target clause for the call to $\validate$.
 
@@
 definition, proof, and applications of a lemma for each shared node in
-the graph.
+the graph.  For any node $\nodeu$, the first call to
+$\validate(\nodeu, \passign, \psi)$ may require further recursion,
+but any subsequent call can simply reuse the lemma proved by the first call.
 
@@
 \section{Experimental Evaluation}
+\label{sect:experimental}
+
+We implemented the three programs illustrated in
+Figure~\ref{fig:chain} to supplement the knowledge compiler \dfour{}.
+The proof generator and proof checker were written in C/C++, while the
+ring evaluator was written in Python.\footnote{The programs are
+available at \texttt{https://github.com/rebryant/crat}.}  The proof
+generator uses \cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert proof steps
+back into hinted CRAT clause additions, the generator can use either
+its own RUP proof generator, or it can invoke \dtrim{}~\cite{RAT}.
+The latter yields shorter proofs and scales well to large proofs, but
+each invocation has a high startup cost.  We therefore set a threshold
+of 1000 clauses in the SAT problem for using \dtrim{}.
+
+The proof generator can optionally be instructed to generate a {\em
+  one-sided} proof, providing only reverse-implication portion via
+input clause deletion.  This can provide useful information---any
+assignment that is a model for the compiled representation
+must also be a model for the input formula---even when generation of a
+full validation is impractical.
+
+Our ring evaluator can perform both standard and weighted model
+counting.  It performs arithmetic over a subset of the rational
+numbers we term $\drational$, consisting of numbers of the form $a
+\cdot 2^{b} \cdot 5^{c}$, where $a$, $b$, and $c$ are 
+integers, and with $a$ implemented to have arbitary range.
+Allowing scaling by powers of 2 enables the density computation
+and rescaling required for standard model counting.  Allowing scaling
+by powers of both 2 and 5 enables exact decimal arithmetic, handling
+the weights used in the weighted model counting competitions.
 
@@
 
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.80]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=3000000000,
+      legend style={at={(0.30,0.98)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=1.8,
+                              xmode=log, xmin=0.01,xmax=4000,
+                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
+                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$,$10^{10}$},
+                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
+%                              title={D4 Defining Clause Generation}
+            ]
+    \input{data-formatted/time+defining-success}
+    \input{data-formatted/time+defining-onesided-only}
+    \input{data-formatted/time+defining-failure}
+    \legend{
+      \scriptsize \textsf{Full validation},
+      \scriptsize \textsf{One-sided validation},
+      \scriptsize \textsf{Resource limit exceeded},
+    }
+ \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Number of defining clauses in generated POG as function of D4 runtime}
+\label{fig:generate:d4:log}
+\end{figure}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.80]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=20000,
+      legend style={at={(0.30,0.98)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.4,
+                              xmode=log,xmin=0.01,xmax=4000,
+                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
+                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
+                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
+%                              title={D4 Defining Clause Generation}
+            ]
+    \input{data-formatted/time-d4-verify}
+    \input{data-formatted/time-d4-verify-onesided-only}
+    \legend{
+      \scriptsize \textsf{Full validation},
+      \scriptsize \textsf{One-sided validation},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
+    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
+    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
+    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
+    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
+    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Combined runtime for CRAT proof generation and checking  as function of D4 runtime}
+\label{fig:d4:crat}
+\end{figure}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.80]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
+                              x post scale=1.8, y post scale=2.15,
+                              xmode=log,xmin=100,xmax=1000000000, 
+                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
+                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
+                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
+%                              title={D4 Defining Clause Generation}
+            ]
+    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
+    \input{data-formatted/defining+total}
+    \node[left] at (axis cs: 900,1200) {$1\times$};
+    \node[left] at (axis cs: 900,12000) {$10\times$};
+    \node[left] at (axis cs: 900,120000) {$100\times$};
+    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
+    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Total number of clauses in CRAT file as function of number of defining clauses}
+\label{fig:defining:total}
+\end{figure}
+
+\begin{figure}
+\begin{minipage}{0.49\textwidth}
+\centering{%
+\begin{tikzpicture}[scale = 0.80]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
+      legend style={at={(0.50,0.98)}},
+      legend cell align={left},
+                              x post scale=1.1, y post scale=1.8,
+                              xmode=log,xmin=1000,xmax=1e7, 
+                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
+                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
+%                              xlabel={Optimized Proof Clauses},
+%                              ylabel={Proof Clauses},
+            ]
+    \input{data-formatted/clauses-lemma-nolemma}
+    \input{data-formatted/clauses-nosplit-split}
+    \legend{
+      \scriptsize \textsf{No lemmas},
+      \scriptsize \textsf{No literal grouping},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
+
+    \node[left] at (axis cs: 5000,7000) {$1\times$};
+    \node[left] at (axis cs: 5000,70000) {$10\times$};
+    \node[left] at (axis cs: 5000,700000) {$100\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\centering{%
+\begin{tikzpicture}[scale = 0.80]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
+      legend style={at={(0.50,0.98)}},
+      legend cell align={left},
+                              x post scale=1.1, y post scale=1.8,
+                              xmode=log,xmin=1000,xmax=1e7, 
+                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
+                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
+%                              xlabel={Optimized Proof Clauses},y
+%                              ylabel={Proof Clauses},
+            ]
+    \input{data-formatted/clauses-optimized-unoptimized}    
+    \legend{
+      \scriptsize \textsf{Neither optimization},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
+
+    \node[left] at (axis cs: 5000,7000) {$1\times$};
+    \node[left] at (axis cs: 5000,70000) {$10\times$};
+    \node[left] at (axis cs: 5000,700000) {$100\times$};
+  \end{axis}
+\end{tikzpicture}
+} % centering
+\end{minipage}
+\caption{Proof clauses when one (left) or both (right) optimizations are disabled, versus when both optimizations are enabled}
+\label{fig:optimized:lessoptimized}
+\end{figure}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.80]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=1500,
+      legend style={at={(0.22,0.98)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.2,
+                              xmode=log,xmin=0.01,xmax=1500,
+                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100,{1,000}},
+                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
+                              xlabel={CRAT total time (seconds)}, ylabel={MICE total time (seconds)},
+            ]
+    \input{data-formatted/time-mice-compare}
+    \input{data-formatted/time-mice-error}
+    \legend{
+      \scriptsize \textsf{Correct or timeout},
+      \scriptsize \textsf{MICE error},
+    }
+    \input{data-formatted/time-mice-timeout}
+%    \addplot[mark=none, color=black] coordinates{(0.01,1500) (1500,1500)};
+%    \addplot[mark=none, color=black] coordinates{(1500,0.01) (1500,1500)};
+    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (10000.0,10.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (10000.0,100.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
+    \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
+    \node[left] at (axis cs: 5.0,0.06) {$0{.}01\times$};
+    \node[left] at (axis cs: 0.5,0.06) {$0{.}1\times$};
+    \node[left] at (axis cs: 0.05,0.06) {$1\times$};
+    \node[left] at (axis cs: 0.05,0.6) {$10\times$};
+    \node[left] at (axis cs: 0.05,6.0) {$100\times$};
+    \node[left] at (axis cs: 0.05,60.0) {$1000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Running Time for MICE versus CRAT proof chains}
+\label{fig:mice}
+\end{figure}
+
 \end{document}

### 2023-03-10 tools/merge_csv.py

@@ -22,6 +22,7 @@ def usage(name):
     eprint("  -h            Print this message")
     eprint("  -l LABELS     Provide comma-separated set of heading labels")
     eprint("  FILE1.csv ... Source files")
+    eprint("  (If give numeric value, then it gets used for all entries")
 
 # Growing set of result lines, indexed by key
 globalEntries = {}
@@ -37,7 +38,7 @@ def processFile(fname):
         infile = open(fname)
         creader = csv.reader(infile)
     except:
-        eprint("Coudn't open CSV file '%s'" % fname)
+        eprint("Couldn't open CSV file '%s'" % fname)
         sys.exit(1)
     row = 0
     for fields in creader:
@@ -75,8 +76,25 @@ def merge(entries1, count1, entries2, count2, subset = True):
             entries[k] = entry1 + entry2
     return entries
         
+def mergeConstant(entries1, value):
+    entries = {}
+    for k in entries1.keys():
+        entry1 = entries1[k]
+        entries[k] = entry1 + [str(value)]
+    return entries
+
 def nextFile(fname, first, subset):
     global globalEntries, globalCount
+    value = None
+    try:
+        value = int(fname)
+    except:
+        pass
+    if value is not None:
+        globalEntries = mergeConstant(globalEntries, value)
+        globalCount += 1
+        return
+
     entries, ccount = processFile(fname)
     if first:
         globalEntries = entries
@@ -84,8 +102,6 @@ def nextFile(fname, first, subset):
     else:
         globalEntries = merge(globalEntries, globalCount, entries, ccount, subset)
         globalCount += ccount
-#    print("globalCount = %d" % globalCount)
-#    print("globalEntries = %s" % str(globalEntries))
 
 def sumEntries(sumSet):
     global globalEntries, globalCount

### 2023-03-10 doc/lean.tex
Separate section

\section{A Formally Verified CRAT Checker and POG Model Counter}

### 2023-03-10 doc/mc-paper.tex
Separate section for Lean

@@
 Moreover, we 
-present a checker that has been formally verified
+present a CRAT checker, and a POG model counter that have been formally verified
 in the \lean{} proof assistant, thereby creating the
@@
 
-\section{A Formally Verified CRAT Checker}
+\input{lean}
 
@@
 counting.  It performs arithmetic over a subset of the rational
-numbers we term $\drational$, consisting of numbers of the form $a
-\cdot 2^{b} \cdot 5^{c}$, where $a$, $b$, and $c$ are 
-integers, and with $a$ implemented to have arbitary range.
+numbers we term $\drational$, consisting of numbers of the form
+$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
+with $a$ implemented to have arbitary range.
 Allowing scaling by powers of 2 enables the density computation
@@
 
-Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
-$s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
-with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{i=1,k}
-[\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
-such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
-\cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
-such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
-\cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
-of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
-for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
-the sum of $t_1, \ldots, t_k$.
+%% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
+%% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
+%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{i=1,k}
+%% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
+%% such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
+%% \cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
+%% such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
+%% \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
+%% of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
+%% for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
+%% the sum of $t_1, \ldots, t_k$.
 

### 2023-03-12 doc/mc-paper.tex
typos

@@
 variables and their complements, and with each nonterminal node
-labeled by Boolean operation And or Or.  Restrictions are place on the
+labeled by Boolean operation And or Or.  Restrictions are placed on the
 structure of the graph such that a count of the models can be computed
@@
 d-DNNF graphs into a checkable form.  We compare the performance of
-our tool chain to theirs as part of our experimental evluation in
+our tool chain to theirs as part of our experimental evaluation in
 Section~\ref{sect:experimental}.  Our experience with developing a
@@
 represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CRAT file.  Some may be
-unalterered input clauses, and these can be used directly.  Others,
+unaltered input clauses, and these can be used directly.  Others,
 however will be clauses that do not appear in the input formula.  We
@@
 $a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
-with $a$ implemented to have arbitary range.
+with $a$ implemented to have arbitrary range.
 Allowing scaling by powers of 2 enables the density computation

### 2023-03-13 doc/lean.tex
start paper section on the verified checker

@@
 \section{A Formally Verified CRAT Checker and POG Model Counter}
+
+[NOTE (JA): I am not worrying about length for the moment. This entire section can become an appendix, and then we can give an executive summary using whatever space
+we have available]
+
+We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
+proof checker and both model and weighted model counters, in the Lean 4 programming language
+and proof assistant [cite]. Lean 4 implements a logical foundation in which expressions
+have a computational interpretation, and just as with other proof assistants like Isabelle and Coq,
+a function defined within the formal system can be compiled to efficient code.
+At the same time, we can state and prove claims about the function within the system, thereby
+verifying that the functions compute the intended results.
+
+Verifying our checker and the model counters in this way serves at least two purposes.
+When we run them on particular data sets, they provide high confidence
+that the particular results we obtain are correct.
+More broadly, however, our success also validates our general approach,
+showing, in particular, that the CRAT proof format is sound.
+Indeed, our formalization helped us discover a bug in our implementation
+of the informal proof checker: in the ``\emph{s}'' step described at the end of Subsection~\ref{subsection:syntax}, it is sound (and necessary) to allow defining
+clauses for the new operations in the proof of $\obar{\lit}_1 \lor \obar{\lit}_2$,
+but it is generally not sound to allow the input clauses or other added clauses. [NOTE: let's add this specification to the end of Section 6.1.]
+Our verified implementation of the checker tells us that the algorithm is sound and provides a reference for other implementations.
+
+In this section, we describe the implementation in Lean and the specifications we have proved.
+Lean 4's logical foundation describes a functional programming language with
+inductive data types,
+but our procedures also make use of calculations with natural numbers, integers, and rationals,
+and data structures for arrays and hashmaps.
+Lean's code extraction uses efficient but unverified versions of these,
+and so in Subsection~\ref{subsection:verification:specifics} we clarify
+what exactly we have verified and what we currently trust.
+
+\subsection{Logical objects and operations}
+
+Our specifications are built on a generic library for the syntax and semantics of
+propositional logic. The data type of propositional formulas with variables indexed by
+a datatype \lstinline{ν} is defined inductively as follows:
+\begin{lstlisting}
+inductive PropForm (ν : Type)
+  | var (x : ν)
+  | tr
+  | fls
+  | neg (φ : PropForm ν)
+  | conj (φ₁ φ₂ : PropForm ν)
+  | disj (φ₁ φ₂ : PropForm ν)
+  | impl (φ₁ φ₂ : PropForm ν)
+  | biImpl (φ₁ φ₂ : PropForm ν)
+\end{lstlisting}
+The last two constructors denote implication and bi-implication, which are not used in this
+project. The usual syntactic and semantic notions are defined in the expected ways.
+It is convenient to take a truth assignment \lstinline{PropAssignment ν} to a function that
+assigns a boolean value to every assignment, and later restrict attention to subsets of the
+variables. In Lean, we introduce the notation \lstinline{τ ⊨ φ} for the statement that
+\lstinline{τ} satisfies \lstinline{φ}. We write \lstinline{entails φ₁ φ₂} to say that
+\lstinline{φ₁} and \lstinline{φ₂} and \lstinline{equivalent φ₁ φ₂}.
+For the purpose of model counting, if \lstinline{s} is any finite set of variables,
+we define \lstinline{models φ s} to be the finite set of truth assignments \lstinline{τ}
+satisfying \lstinline{φ} that assign arbitrary values on \lstinline{s} and
+\lstinline{false} outside \lstinline{s}.
+
+In logic, the set of propositional formulas modulo equivalence is sometimes called the \emph{Lindenbaum--Tarski algebra}. We have found it convenient to define this quotient,
+which we denote \lstinline{PropTerm ν}, and lift the boolean operations and the entailment
+relation to this new type. The advantage is that equivalent formulas give
+rise to equal elements in the quotient.
+This makes easier to prove equivalences as we go, since we can calculate with the associated
+elements of the quotient and substitute one for another when the corresponding formulas are equivalent.
+
+In our application, we instantiate \lstinline{ν} to a type \lstinline{Var} consisting
+of positive natural numbers, corresponding to the indexing of variables in the DIMACS
+format for SAT. It may be helpful to think of the elements of \lstinline{PropForm Var},
+\lstinline{PropTerm Var}, and \lstinline{PropAssignment Var} as mathematical objects
+rather than computational data. Our proof checker and model counters do not compute
+with them; rather, we use them to specify what the checker and model counters are supposed to do.
+The starting point for the proof checker is, instead, a CNF formula, which \emph{is}
+presented as data.
+Specifically, a literal is a nonzero integer, with the understanding that $-\ell$ represents
+the negation of $\ell$. A clause is an array of literals, and a CNF formula is an array
+of clauses:
+\begin{lstlisting}
+def ILit := { i : Int // i ≠ 0 }
+abbrev IClause := Array ILit
+abbrev ICnf := Array IClause
+\end{lstlisting}
+We define operations that translate these objects to the elements of \lstinline{PropForm Var}
+that they represent. Lean provides us with helpful ``anonymous projection'' syntax that allows
+us to write, for example, \lstinline{C.toPropForm} for the expression
+\lstinline{Clause.toPropForm C}.
+Similarly, we have a representation \lstinline{C.toPropTerm}
+of \lstinline{C} as an equivalence class of propositional formulas, and we can speak of the
+value of \lstinline{C} under any truth assignment.
+Our code represents partial truth assignments as hashmaps:
+\begin{lstlisting}
+abbrev PartPropAssignment := HashMap Var Bool
+\end{lstlisting}
+We have written a procedure that reduces a clause \lstinline{C} with respect to a partial
+assignment, and another procedure that uses a hashset to test whether a clause \lstinline{C} is a
+tautology.
+We have also verified that these low-level procedures do what they are supposed to do.
+
+\subsection{Proof Checker}
+
+Our proof checker interprets the input clauses listed in a CRAT file as a CNF formula $\varphi$.
+It then processes and checks each rule, throwing an exception if a precondition is not met.
+If it terminates without raising an exception, it outputs a POG, $P$, with distinguished root
+$\bf r$.
+We have formally proved the claim that in this case the formula corresponding to node $\bf r$ is a partitioned formula and is
+logically equivalent to $\varphi$.
+The POG $P$ can be saved to a file [TODO: implement this?] or passed on to one of the model counting
+procedures described in the next section.
+
+In our code, each element of a POG  is either a variable, a binary disjunction, or an arbitrary conjunction:
+\begin{lstlisting}
+inductive PogElt where
+  | var  : Var → PogElt
+  | disj : Var → ILit → ILit → PogElt
+  | conj : Var → Array ILit → PogElt
+\end{lstlisting}
+In the first case, the argument \lstinline{x} in the expression
+\lstinline{var x} is the index
+of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
+it is the definition number in the CRAT file.
+The CRAT generator described in Section~\ref{section:generating:crat}
+declares new variables sequentially, so we currently make the simplifying assumption
+that the $i$th element of the array has the node with index $i + 1$,
+but it will not be hard to generalize this.
+Notice that the arguments to \lstinline{disj} and \lstinline{conj} are literals,
+corresponding to positive or negative instances of variables declared earlier in the POG.
+So our POG data structure is as follows:
+\begin{lstlisting}
+structure Pog where
+  elts : Array PogElt
+  wf : ∀ i : Fin elts.size, elts[i].args_decreasing
+  inv : ∀ i : Fin elts.size, i = elts[i].varNum.natPred
+\end{lstlisting}
+The invariant \lstinline{wf} says that the graph is well founded, in the
+sense that if \lstinline{elts[i]} is a disjunction
+or a conjunction, then the variables occurring in the arguments have smaller indices.
+Note that \lstinline{Fin elts.size} is the data type consisting of natural numbers
+less than the size of the array \lstinline{elts}; our code enforces statically
+that the array accesses are within bounds.
+
+Note that we have not yet said that the corresponding formula is partitioned.
+For each POG \lstinline{P} and literal \lstinline{l}, we define
+\lstinline{P.toPropForm l} to be the
+propositional formula that arises from interpreting \lstinline{l} as a propositional
+formula, unfolding all the defined conjunctions and disjunctions. We define
+what it means for such a formula to be partitioned:
+\begin{lstlisting}
+def partitioned : PropForm ν → Prop
+  | tr         => True
+  | fls        => True
+  | var _      => True
+  | neg φ      => φ.partitioned
+  | disj φ ψ   => φ.partitioned ∧ ψ.partitioned ∧
+                    ∀ τ, ¬ (φ.eval τ ∧ ψ.eval τ)
+  | conj φ ψ   => φ.partitioned ∧ ψ.partitioned ∧ (φ.vars ∩ ψ.vars = ∅)
+  | impl _ _   => False
+  | biImpl _ _ => False
+\end{lstlisting}
+In other words, we rule out formulas with implication and bi-implication, and otherwise we follow the straightforward recursive definition.
+
+With these definitions in place, we can explain the specification for our checker.
+Our parser reads a CRAT file and returns an input formula \lstinline{cnf : ICnf} and a CRAT proof {pf : Array CratStep}, that is, an array of single proof steps:
+\begin{lstlisting}
+inductive CratStep (α ν β : Type)
+  | /-- Add asymmetric tautology. -/
+    addAt (idx : α) (C : Array β) (upHints : Array α)
+  | /-- Delete asymmetric tautology. -/
+    delAt (idx : α) (upHints : Array α)
+  | /-- Declare product operation. -/
+    prod (idx : α) (x : ν) (ls : Array β)
+  | /-- Declare sum operation. -/
+    sum (idx : α) (x : ν) (l₁ l₂ : β) (upHints : Array α)
+  | /-- Delete operation. -/
+    delOp (x : ν)
+\end{lstlisting}
+The core checker takes those as input, as well as a specification of the final node \lstinline{r}, and either throws an exception, indicating that the proof is not well formed, or returns a POG \lstinline{P}.
+The specification asserts that, in the latter case, \lstinline{P.toPropForm r} is partitioned
+and is equivalent to \lstinline{cnf.toPropForm}.
+
+[TODO(JA): I'll leave stop here now because I am running out of steam today,
+and also because Wojciech is still working on it.
+But here is where we can sketch what the checker does.
+For each line, it does some checks, it adds clauses to a clause database,
+and it add a definition to the POG.
+There is a list of invariants having to do with the original formula,
+the clause database, and the POG, and we prove that if the checks are satisfied,
+then at each step the invariants hold.
+We should list the invariants and say something about how the checks are
+performed.]
+
+\subsection{The Model Counter}
+
+The number of models that a propositional formula has depends on what we take the
+variables to be. For example, there is only one assignment to the variables
+$\{ x, y \}$ that makes $x \wedge y$ true, but there are two assignments to the
+variables $\{ x, y, z \}$ that make $x \wedge y$ true.
+It is not hard to show that for partitioned formulas \lstinline{PropForm ν} whose
+variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
+we can count the number of models of the formula on the variables in \lstinline{s}
+recursively as follows:
+\begin{lstlisting}
+def countModels (nVars : Nat) : PropForm ν → Nat
+  | tr         => 2^nVars
+  | fls        => 0
+  | var _      => 2^(nVars - 1)
+  | neg φ      => 2^nVars - φ.countModels nVars
+  | disj φ ψ   => φ.countModels nVars + ψ.countModels nVars
+  | conj φ ψ   => φ.countModels nVars * ψ.countModels nVars / 2^nVars
+  | impl _ _   => 0
+  | biImpl _ _ => 0
+\end{lstlisting}
+Note that we can carry out this calculation for any formula at all, though the
+calculation is meaningless if the formula is not partitioned.
+Also recall that if the formula is partitioned, it does not contain implication or bi-implication.
+
+We have implemented an efficient counting function for POGs and proved that
+it computes that same quantity on the corresponding formulas:
+\begin{lstlisting}
+def count (pog : Pog) (nVars : Nat) (x : Var) : Nat := ...
+
+theorem count_eq_countModels (pog : Pog) (nVars : Nat) (x : Var) :
+  pog.count nVars x = (pog.toPropForm (.mkPos x)).countModels nVars
+\end{lstlisting}
+Here \lstinline{.mkPos x} denotes the positive literal for the variable \lstinline{x}. We have also proved that the calculation above really does
+return the number of models:
+\begin{lstlisting}
+theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
+    (hvars : φ.vars ⊆ s) (hpar : φ.partitioned) :
+  φ.countModels s.card = card (φ.models s)
+\end{lstlisting}
+In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
+we have that the number of models on its variables is \lstinline{φ.countModels φ.vars.card}.
+
+Given a weighting function \lstinline{weight : ν → Rat} that assigns a rational
+number in $[0, 1]$ to each variable, we can calculate the weighted model count on
+partitioned formulas as follows:
+\begin{lstlisting}
+def countWeight (weight : ν → Rat) : PropForm ν → Rat
+  | tr         => 1
+  | fls        => 0
+  | var x      => weight x
+  | neg φ      => 1 - φ.countWeight weight
+  | disj φ ψ   => φ.countWeight weight + ψ.countWeight weight
+  | conj φ ψ   => φ.countWeight weight * ψ.countWeight weight
+  | impl _ _   => 0
+  | biImpl _ _ => 0
+\end{lstlisting}
+We have implemented the corresponding count function on POGs and proved that
+it calculates this quantity on the associated formulas,
+though we have not yet proved the combinatorial lemma
+that relates this calculation to the sum of the weights of the corresponding models.
+
+\subsection{Verification specifics}
+\label{subsection:verification:specifics}
+
+In this subsection, we clarify what has been verified and what has been trusted.
+Recall that our first step is to parse a CRAT file to read in the initial CNF formula
+and the CRAT proof. We do not verify this step. There is no strong need to verify the
+parsing of the proof, because if the checker claims success for \emph{any} proof, it does not really matter
+whether the proof was the one the user intended.
+On the other hand, it would be nice to have reassurance that the CNF formula that serves
+as input to our verified checker matches the one described in the file.
+An easy and fairly standard way of addressing this is to print out the CNF formula
+right after parsing it and then do a diff with the original input.
+This involves trusting only the correctness of the print procedure and diff.
+
+Lean's code extraction replaces calculations with natural numbers and integers with
+efficient but unverified arbitrary precision versions.
+Lean also uses an efficient implementation of arrays; within the
+formal system, these are defined in terms of lists, but code extraction replaces them
+with dynamic arrays and uses reference counting to allow destructive updates when it is safe
+to do so [cite counting immutable beans paper].
+Code extraction also provides efficient implementations of tail recursive functions and loops
+[cite ``do unchained'' paper].
+Finally, we have implemented hashmaps in terms of arrays, but we have not yet proved
+the correctness of all the primitive operations on these hashmaps.
+Thus our proofs depend on the unverified assumption that those operations have the
+expected properties.
+
+In sum, in addition to trusting Lean's foundation and trusted kernel,
+we also have to trust that code extraction respects that foundation,
+that the implementation of arrays and hashmaps satisfy their internal descriptions,
+and that, after parsing, the computation has the correct input formula.
+All of our specifications have been completely proved and verified relative to that.
+
+\subsection{Evaluation}
+
+[TODO: Say something to establish that the procedure run in reasonable time and produce useful results; or maybe this does in the next section?]

### 2023-03-14 doc/mc-paper.tex
start paper section on the verified checker
Fixed reference
merge

@@
 $\assign(r) = 0$.  It starts with the set of literals
-$\{ \obar{\lit} | \lit \in C\}$, describing the required condition for
+$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
 assignment $\assign$ to falsify clause $C$.

### 2023-03-14 doc/lean.tex

@@
 
-\subsection{Proof Checker}
+\subsection{The Proof Checker}
 
@@
 
-\subsection{The Model Counter}
+\subsection{The Model Counters}
 
@@
 
-In sum, in addition to trusting Lean's foundation and trusted kernel,
+In sum, in addition to trusting Lean's foundation and kernel checker,
 we also have to trust that code extraction respects that foundation,

### 2023-03-14 doc/mc-data.tex
Add Monolithic results

@@
 
+\section{Comparing Monolithic and Structured Validation}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.90]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=1500,
+      legend style={at={(0.22,0.98)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.2,
+                              xmode=log,xmin=0.01,xmax=1500,
+                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100,{1,000}},
+                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
+                              xlabel={Structured generation + checking (seconds)}, ylabel={Monolithic generation + checking (seconds)},
+            ]
+    \input{data-formatted/time-structured+monolithic}
+    \input{data-formatted/time-structured+monolithic_timeout}
+%    \addplot[mark=none, color=black] coordinates{(0.01,1500) (1500,1500)};
+%    \addplot[mark=none, color=black] coordinates{(1500,0.01) (1500,1500)};
+%    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (10000.0,10.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (10000.0,100.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
+%    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
+%    \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
+    \node[left] at (axis cs: 5.0,0.06) {$0{.}01\times$};
+    \node[left] at (axis cs: 0.5,0.06) {$0{.}1\times$};
+    \node[left] at (axis cs: 0.05,0.06) {$1\times$};
+    \node[left] at (axis cs: 0.05,0.6) {$10\times$};
+    \node[left] at (axis cs: 0.05,6.0) {$100\times$};
+%    \node[left] at (axis cs: 0.05,60.0) {$1000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Monolithic versus Structured Validation: Time}
+\label{fig:monolithic:time}
+\end{figure}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.90]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10},
+      legend style={at={(0.22,0.98)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.2,
+                              xmode=log,xmin=1e3,xmax=1e7,
+                              ymode=log, ymin=1e3, ymax=1e7,
+                              xlabel={Structured (total clauses)}, ylabel={Monolithic (total clauses)},
+            ]
+    \input{data-formatted/clauses-total-structured+monolithic}
+%    \addplot[mark=none, color=black] coordinates{(0.01,1500) (1500,1500)};
+%    \addplot[mark=none, color=black] coordinates{(1500,0.01) (1500,1500)};
+%    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (10000.0,10.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(1e3,1e4) (1e7,1e8)};
+    \addplot[mark=none, color=lightblue] coordinates{(1e3,1e3) (1e7,1e7)};
+    \addplot[mark=none, color=lightblue] coordinates{(1e3,1e2) (1e7,1e6)};
+%    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
+%    \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
+    \node[left] at (axis cs: 5000,60000) {$10\times$};
+    \node[left] at (axis cs: 5000,6000) {$1\times$};
+    \node[left] at (axis cs: 50000,6000) {$0.1\times$};
+%    \node[left] at (axis cs: 0.05,60.0) {$1000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Monolithic versus Structured Validation: Clauses}
+\label{fig:monolithic:clauses}
+\end{figure}
+
 \section{Comparison with MICE}

### 2023-03-14 doc/mc-data.tex
Remove duplicate graphs

@@
 
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
-      legend style={at={(0.40,0.98)}},
-      legend cell align={left},
-                              x post scale=1.4, y post scale=2.3,
-                              xmode=log,xmin=1000,xmax=1e7, 
-                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
-                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
-                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
-            ]
-    \input{data-formatted/clauses-lemma-nolemma}
-    \input{data-formatted/clauses-nosplit-split}
-    \legend{
-      \scriptsize \textsf{Without lemmas},
-      \scriptsize \textsf{Without literal grouping},
-    }
-    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
-
-    \node[left] at (axis cs: 5000,7000) {$1\times$};
-    \node[left] at (axis cs: 5000,70000) {$10\times$};
-    \node[left] at (axis cs: 5000,700000) {$100\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Proof Clauses when Single Optimization Disabled}
-\label{fig:optimized:partially-optimized}
-\end{figure}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
-      legend style={at={(0.40,0.98)}},
-      legend cell align={left},
-                              x post scale=1.4, y post scale=2.3,
-                              xmode=log,xmin=1000,xmax=1e7, 
-                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
-                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
-                              xlabel={Proof Clauses using Lemmas and Literal Grouping}, ylabel={Proof Clauses},
-            ]
-    \input{data-formatted/clauses-optimized-unoptimized}    
-    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
-
-    \node[left] at (axis cs: 5000,7000) {$1\times$};
-    \node[left] at (axis cs: 5000,70000) {$10\times$};
-    \node[left] at (axis cs: 5000,700000) {$100\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Proof Clauses Both Optimizations are Disabled}
-\label{fig:optimized:unoptimized}
-\end{figure}
-
 We conducted a set of experiments using the 80 benchmark instances for

### 2023-03-14 doc/mc-paper.tex
Ring must be commutative

@@
 \newcommand{\prov}{\textrm{Prov}}
+\newcommand{\inputformula}{\phi_I}
+\newcommand{\pogformula}{\phi_P}
 
@@
 
+% Lean code:
+\usepackage{listings}
+\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
+\definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
+\definecolor{commentcolor}{rgb}{0.4, 0.4, 0.4}   % grey
+\definecolor{symbolcolor}{rgb}{0.0, 0.1, 0.6}    % blue
+\definecolor{sortcolor}{rgb}{0.1, 0.5, 0.1}      % green
+\definecolor{attributecolor}{rgb}{0.7, 0.1, 0.1} % red
+\def\lstlanguagefiles{lstlean.tex}
+% set default language
+\lstset{language=lean, xleftmargin=1em}
+\lstset{backgroundcolor=\color{white}}
+
 \bibliographystyle{plainurl}% the mandatory bibstyle
@@
 
-Fichte, Hecher, and Roland~\cite{fichte:sat:2021} devised the MICE
+Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
 proof framework for model counting programs.  Their proof rules are
@@
 
-\begin{definition}[Ring Evaluation Problem (REP)]
-  For ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
+\begin{definition}[Ring Evaluation Problem]
+  For commutative ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
   every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $\mulident-w(x)$.
@@
 {\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
-It can be cast as a REP by having $\radd$ and
+It can be cast as a ring evaluation problem by having $\radd$ and
 $\rmul$ be addition and multiplication over rational numbers and using
@@
 \subsection{Syntax}
+\label{subsection:syntax}
 
@@
 \section{Generating CRAT from d-DNNF}
+\label{section:generating:crat}
 

### 2023-03-14 doc/mc-data.tex
Add URL to Github

@@
 \usepackage{booktabs}
+\usepackage{hyperref}
 
@@
 \newcommand{\Checker}{\progname{Crat-check}}
-
-%% Colored hyperlink 
-%\newcommand{\cref}[2]{\href{#1}{\color{blue}#2}}
-%% Colored hyperlink showing link in TT font
-% \newcommand{\chref}[1]{\href{#1}{\small\tt \color{blue}#1}}
-\newcommand{\hcref}[1]{\cref{#1}{\small\tt #1}}
 
@@
      \begin{center}
-       \hcref{https://mccompetition.org/2022/mc-description.html}
+       \url{https://mccompetition.org/2022/mc_description.html}
      \end{center}
@@
 \begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=1500,
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
       legend style={at={(0.22,0.98)}},
@@
                               xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100,{1,000}},
+                              ymin=0.01, ymax=1900,
                               ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
@@
     \input{data-formatted/time-structured+monolithic_timeout}
-%    \addplot[mark=none, color=black] coordinates{(0.01,1500) (1500,1500)};
-%    \addplot[mark=none, color=black] coordinates{(1500,0.01) (1500,1500)};
-%    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (10000.0,10.0)};
     \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (10000.0,100.0)};
@@
     \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
-%    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
-%    \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
+    \addplot[color=black,dashed] coordinates{(0.01,1500) (1500,1500)};
     \node[left] at (axis cs: 5.0,0.06) {$0{.}01\times$};
@@
     \node[left] at (axis cs: 0.05,6.0) {$100\times$};
-%    \node[left] at (axis cs: 0.05,60.0) {$1000\times$};
 
@@
 %    \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
-    \node[left] at (axis cs: 5000,60000) {$10\times$};
-    \node[left] at (axis cs: 5000,6000) {$1\times$};
-    \node[left] at (axis cs: 50000,6000) {$0.1\times$};
+    \node[left] at (axis cs: 3500,40000) {$10\times$};
+    \node[left] at (axis cs: 3500,4000) {$1\times$};
+    \node[left] at (axis cs: 35000,4000) {$0.1\times$};
 %    \node[left] at (axis cs: 0.05,60.0) {$1000\times$};
@@
 \begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=1500,
-      legend style={at={(0.22,0.98)}},
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10},
+      legend style={at={(0.22,0.97)}},
       legend cell align={left},
                               x post scale=1.8, y post scale=2.2,
-                              xmode=log,xmin=0.01,xmax=1500,
+                              xmode=log,xmin=0.01,xmax=1900,
                               xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100,{1,000}},
+                              ymode=log, ymin=0.01, ymax=1900,
                               ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
@@
     \input{data-formatted/time-mice-timeout}
+
+    \addplot[color=black,dashed] coordinates{(0.01,1500) (1500,1500)};
+    \addplot[color=black,dashed] coordinates{(1500,0.01) (1500,1500)};
+
 %    \addplot[mark=none, color=black] coordinates{(0.01,1500) (1500,1500)};

### 2023-03-15 doc/mc-data.tex
Did not meet criteria

@@
      \item Timed out for 14 instances
+     \item Was able to do one-sided verification for 9 of these.
      \end{itemize}
@@
                               x post scale=1.8, y post scale=2.2,
-                              xmode=log,xmin=0.01,xmax=1500,
+                              xmode=log,xmin=0.01,xmax=1900,
                               xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100,{1,000}},
@@
     \addplot[color=black,dashed] coordinates{(0.01,1500) (1500,1500)};
+    \addplot[color=black,dashed] coordinates{(1500,0.01) (1500,1500)};
     \node[left] at (axis cs: 5.0,0.06) {$0{.}01\times$};
@@
 \end{figure}
+
+Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 83.  All completed in less than 1000 seconds with structured.
+Compare both runtimes and total number of clauses for structured vs. monolithic:
+\begin{itemize}
+\item Structured complete 83.
+\item Monolithic completed  69, timed out on 14.
+\item Of the 69 completed, 23 were faster with monolithic, 36 were slower (min ratio = 0.015, max ratio = 9.66, hmean = 0.40)
+\item Of the 69 completed, 59 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
+\end{itemize}
 
@@
 runtimes of our toolchain versus theirs.  The X-axis shows the sum of
-the elapsed times for our generator, checker, and counter, while the
+the elapsed times for our generator, checker, and counter (since the MICE checker also provides a count), while the
 Y-axis shows the sum of the elapsed times for \textsc{nnf2trace},
@@
 \begin{itemize}
-\item 85 total data points
-\item 64 were completed by both programs in under 1000 seconds
-\item Of these, the MICE toolchain was faster for 20, and the CRAT toolchain was faster for 45
-\item 5 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
+\item 84 total data points
+\item 63 were completed by both programs in under 1000 seconds
+\item Of those 20 were faster with MICE, 43 with CRAT
+\item 6 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
 \item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
-\item 15 completed with CRAT but generated errors for MICE.
+\item 14 completed with CRAT but generated errors for MICE.
 \end{itemize}

### 2023-03-15 doc/mc-paper.tex

@@
 ring evaluator was written in Python.\footnote{The programs are
-available at \texttt{https://github.com/rebryant/crat}.}  The proof
+available at \url{https://github.com/rebryant/crat}.}  The proof
 generator uses \cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert proof steps

### 2023-03-15 doc/mc-paper.tex
orcid
RUP

@@
 \rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
-\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
-\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
+\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
+%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
 \end{eqnarray}
@@
 proving that a set of clauses logically implies another clause.
+
+As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$ 
+and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
+A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
+is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the 
+involved clauses (antecedents) are shown in the order that they became unit / falsified.
+
+\begin{center}
+  \begin{tabular}{lcccc}
+         & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
+  Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
+  \midrule
+  Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
+  \end{tabular}
+\end{center}
+
+RUP is an alternative formulation of resolution.  For target clause
+$C$, it can be seen that applying resolution operations to the
+antecedent clauses from right to left will derive a clause $C'$ such
+that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
+we then have $C' \rightarrow C$.  Compared to listing each resolution
+operation as a separate step, using RUP as the basic proof step makes
+the proofs more compact.
 

### 2023-03-15 doc/mc-data.tex
Lots of updates

@@
 \newcommand{\prov}{\mathit{Prov}}
-
-\newcommand{\progname}[1]{\textsc{#1}}
-\newcommand{\dfour}{\progname{d4}}
-\newcommand{\Dfour}{\progname{d4}}
-\newcommand{\generator}{\progname{d2p}}
-\newcommand{\Generator}{\progname{d2p}}
-\newcommand{\checker}{\progname{crat-check}}
-\newcommand{\Checker}{\progname{Crat-check}}
 
@@
 
+\section*{Time to Generate and Validate CRAT}
+
+\begin{itemize}
+\item Min 0.01, Average 1318.3, Max 13,144.8, Median 71.64
+\item Ratio to time for D4: Min 0.50, Harmonic Mean 5.50, Max 457.69
+\end{itemize}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.90]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
+      legend style={at={(0.30,0.96)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.4,
+                              xmode=log,xmin=0.01,xmax=4000,
+                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
+                              ymin=0.01, ymax=29000,
+                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
+                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
+%                              title={D4 Defining Clause Generation}
+            ]
+    \input{data-formatted/time-d4-verify}
+    \input{data-formatted/time-d4-verify-onesided-only}
+    \input{data-formatted/time-d4-failure}
+    \legend{
+      \scriptsize \textsf{Full validation},
+      \scriptsize \textsf{One-sided validation},
+      \scriptsize \textsf{Resource limit exceeded},
+    }
+    \addplot[mark=none, dashed] coordinates{(0.01,25000) (4000, 25000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (2000,20000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (200, 20000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (20, 20000)};
+    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
+    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
+    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
+    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
+    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Combined runtime for CRAT proof generation and checking  as function of D4 runtime}
+\label{fig:d4:crat}
+\end{figure}
+
 \begin{figure}
@@
 \label{fig:defining:total}
-\end{figure}
-
-\section*{Time to Generate and Validate CRAT}
-
-\begin{itemize}
-\item Min 0.01, Average 1318.3, Max 13,144.8, Median 71.64
-\item Ratio to time for D4: Min 0.50, Harmonic Mean 5.50, Max 457.69
-\end{itemize}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=20000,
-      legend style={at={(0.30,0.98)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=2.4,
-                              xmode=log,xmin=0.01,xmax=4000,
-                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
-                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
-                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \input{data-formatted/time-d4-verify}
-    \input{data-formatted/time-d4-verify-onesided-only}
-    \legend{
-      \scriptsize \textsf{Full validation},
-      \scriptsize \textsf{One-sided validation},
-    }
-    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
-    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
-    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
-    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
-    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
-    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Combined runtime for CRAT proof generation and checking  as function of D4 runtime}
-\label{fig:d4:crat}
 \end{figure}
@@
     \input{data-formatted/time-structured+monolithic_timeout}
-    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (10000.0,100.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
+    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (1000.0,10.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (1000.0,100.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (1000.0,1000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (100, 1000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (10, 1000)};
     \addplot[color=black,dashed] coordinates{(0.01,1500) (1500,1500)};
@@
 
-Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 83.  All completed in less than 1000 seconds with structured.
+Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 82.  All completed in less than 1000 seconds with structured.
 Compare both runtimes and total number of clauses for structured vs. monolithic:
 \begin{itemize}
-\item Structured complete 83.
-\item Monolithic completed  69, timed out on 14.
+\item Structured complete 82.
+\item Monolithic completed  69, timed out on 13.
 \item Of the 69 completed, 23 were faster with monolithic, 36 were slower (min ratio = 0.015, max ratio = 9.66, hmean = 0.40)
@@
 %    \addplot[mark=none, color=black] coordinates{(1500,0.01) (1500,1500)};
-    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (10000.0,10.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (10000.0,100.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
+    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (1000.0,1.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (1000.0,10.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (1000.0,100.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (1000.0,1000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (100,1000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (10,1000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1,1000)};
     \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
@@
 messages, these appear to be mostly due to features of some of the
-d-DNNF graphs they had not anticipated.  We provided a bug report to
+CNF files or d-DNNF graphs they had not anticipated.  We provided a bug report to
 the authors, but they were unable to fix them on short notice.
@@
 \item 84 total data points
-\item 63 were completed by both programs in under 1000 seconds
-\item Of those 20 were faster with MICE, 43 with CRAT
-\item 6 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
+\item 67 were completed by both programs in under 1000 seconds
+\item Of those 21 were faster with MICE, 46 with CRAT
+\item 7 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
 \item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
-\item 14 completed with CRAT but generated errors for MICE.
+\item 9 completed with CRAT but generated errors for MICE.
 \end{itemize}

### 2023-03-15 doc/mc-paper.tex
Lots of updates

--- 2023-03-15 doc/mc-paper.tex
@@
 \newcommand{\imply}{\Rightarrow}
+\newcommand{\ifandonlyif}{\Leftrightarrow}
 \newcommand{\drational}{\mathbb{Q}_{2,5}}
@@
 \newcommand{\Dfour}{\progname{D4}}
-\newcommand{\generator}{\progname{d2p}}
-\newcommand{\Generator}{\progname{D2p}}
-\newcommand{\checker}{\progname{crat-check}}
-\newcommand{\Checker}{\progname{Crat-check}}
 \newcommand{\cadical}{\progname{CaDiCal}}
@@
 
-We have developed a program that can generate POG representations
-from the d-DNNF graphs generated by the state-of-the-art knowledge
-compiler \dfour, as well as independently checkable proofs that the
-output POGs are equivalent to the input CNF formulas.
-We have implemented an efficient checker
-that can handle very large proofs. Our tool
-chain for generating and verifying POGs scales to all but the largest d-DNNF graphs
-generated by \dfour for formulas
-from a recent model counting competition. Moreover, we
-present a checker that has been formally verified
-in the \lean{} proof assistant, thereby creating the
-first formally verified toolchain for
-% weighted and unweighted
+We have developed a program that can generate POG representations from
+the d-DNNF graphs generated by the state-of-the-art knowledge compiler
+\dfour, as well as independently checkable proofs that the output POGs
+are equivalent to the input CNF formulas.  We have implemented an
+efficient checker that can handle very large proofs. Our tool chain
+for generating and verifying POGs scales to all but the largest d-DNNF
+graphs generated by \dfour{} for formulas from a recent model counting
+competition. Moreover, we present a checker and a model counter for POGs that have
+been formally verified in the \lean{} proof assistant, thereby
+creating the first formally verified toolchain for
+weighted and unweighted
 model counting.
@@
 find a satisfying assignment to a Boolean formula or to prove that no
-such assignment exists.  The have applications across a variety of
+such assignment exists.  They have applications across a variety of
 domains, including formal verification of hardware, software, and
@@
   counting problem} requires computing the number of satisfying
-assignments, often in cases where there are far to many satisfying
-assignments to enumerate individually.  Model counting, in turn, has
+assignments of a formula, including in cases where there are far too many
+to enumerate individually.  Model counting, in turn, has
 applications in artificial intelligence, computer security, and
-statistical sampling.  A variant on standard model counting is {\em
-  weighted model counting}, where a weighting function is defined for
-each assignment and the goal becomes to compute the sum of the weights
+statistical sampling.  There are also many useful extensions of standard model counting,
+including {\em
+  weighted model counting}, where a weight is defined for
+each possible assignment, and the goal becomes to compute the sum of the weights
 of the satisfying assignments.
@@
 
-One approach to model counting, known as {\em knowledge compilation}
-is to transform the formula into a structured form for which model
-counting becomes more tractable.  For example, the {\em deterministic,
+One approach to model counting, known as {\em knowledge compilation},
+transforms the formula into a structured form for which model counting
+becomes straightforward.  For example, the {\em deterministic,
   decomposable, negation normal form} (d-DNNF) introduced by
-Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a Boolean formula as a
-directed acyclic graph, with terminal nodes labeled by Boolean
-variables and their complements, and with each nonterminal node
-labeled by Boolean operation And or Or.  Restrictions are placed on the
-structure of the graph such that a count of the models can be computed
-by a single bottom-up traversal of the graph.
-
-Once a Boolean formula has been transformed into a d-DNNF
-representation, a number of useful properties of the formula can
-readily be computed.  Kimmig, et al., present a very general
-framework, known as {\em algebraic model
-counting}~\cite{kimmig:jal:2017}, that go well beyond simple model counting.
+Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
+Boolean formula as a directed acyclic graph, with terminal nodes
+labeled by Boolean variables and their complements, and with each
+nonterminal node labeled by Boolean operation And or Or.  Restrictions
+are placed on the structure of the graph such that a count of the
+models can be computed by a single bottom-up traversal of the graph.
+Kimmig, et al., present a very general framework, known as {\em
+  algebraic model counting}~\cite{kimmig:jal:2017}, describing
+properties of Boolean functions that can be efficiently computing from
+its d-DNNF representation.  These include standard and weighted model
+counting, and much more.
 
 One shortcoming of existing knowledge compilers is that they have no
-way to validate that their generated results are corrected, i.e., that
+way to validate that their generated results are correct, i.e., that
 the compiled representation is logically equivalent to the original
-formula.  By contrast, all modern SAT solvers are able to produce
+formula.  By contrast, all modern SAT solvers can generate
 checkable proofs when they encounter unsatisfiable formulas.  The
 guarantee provided by a checkable certificate of correctness enables
-users of SAT solvers to fully trust their results.  Experience has
-shown that adding this capability has also greatly improved the quality of
-SAT solvers and the ability of developers to quickly detect and
-identify bugs in their solvers.
+users of SAT solvers to fully trust their results.  Experience has also
+shown that having this capability gives developers the ability quickly
+detect and identify bugs in their solvers, and this, in turn, has led
+to more reliable SAT solvers.
 
@@
 a generalization of the representations commonly used in knowledge
-compilation, including d-DNNF\@.  We present the CRAT file format that
-both captures the structure of a POG and can provide a checkable proof
+compilation, including d-DNNF\@.  We present the CRAT file format.
+It captures both the structure of a POG and a checkable proof
 that the POG is logically equivalent to a Boolean formula given in
-conjunctive normal form (CNF).  A proof is expressed as a sequence of
+conjunctive normal form (CNF).  A proof consists of a sequence of
 clause addition and deletion steps, based on the extended resolution
-proof system~\cite{Tseitin:1983}.
-
-We have developed a program that can generate CRAT descriptions from
-the d-DNNF graphs generated by the state-of-the-art knowledge compiler
-\dfour, as well as a fast CRAT proof checker.  We evaluate the
-capabilities of our tool chain using benchmark formulas from the 2022
-standard and weighted model competitions.  Our tools are able to
-handle all but the largest d-DNNF graphs generated by \dfour.
-Moreover, we
-present a CRAT checker, and a POG model counter that have been formally verified
-in the \lean{} proof assistant, thereby creating the
-first formally verified toolchain for
-% weighted and unweighted
-model counting.
+proof system~\cite{Tseitin:1983}.  We present a set of conditions that
+a CRAT file must satisfy to provide an equivalence proof.
+
+\begin{figure}
+\centering{
+\begin{tikzpicture}[scale=0.17]
+  \draw (8,10) rectangle (20,18);
+  \node at (14,15.2) {Knowledge};
+  \node at (14,12.8) {Compiler};
+
+  \draw (28,10) rectangle (40,18);
+  \node at (34,15.2) {Proof};
+  \node at (34,12.8) {Generator};
+
+  \draw (50,10) rectangle (62,18);
+  \node at (56,15.2) {Proof};
+  \node at (56,12.8) {Checker};
+
+  \draw (50,0) rectangle (62,8);
+  \node at (56,5.2) {Ring};
+  \node at (56,2.8) {Evaluator};
+
+  \draw[dashed] (48,-2) rectangle (64,22);
+  \node at (56,20) {Trusted Code};
+
+  %% CNF
+  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (50,16);
+  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
+  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
+  \node [left] at (-1,20) {\large {$\phi$}};
+  \node [above] at (2,20) {\texttt{.cnf}};
+
+  %% NNF
+  \draw[thick] (20,12) [-latex] -- (28,12);
+  \node [above] at (24,12) {\texttt{.ddnnf}};
+
+  %% CRAT
+  \draw[thick] (40,12) [-latex] -- (50,12);
+  \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
+  \node [above] at (44,12) {\texttt{.crat}};
+
+  %% OK/Not
+  \draw[thick] (62,14) [-latex] -- (66,14) ;
+  \node [right] at (66,15.2) { OK /} ;
+  \node [right] at (66,12.8) { Not OK} ;
+  \draw[thick] (62,4) [-latex] -- (66,4) ;
+  \node [right] at (66,4) {\large {$\textbf{R}(\phi)$}};
+
+\end{tikzpicture}
+}
+\caption{Certifying Tool Chain.
+  The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT) that can be independently checked and evaluated}
+\label{fig:chain}
+\end{figure}
+
+Figure~\ref{fig:chain} illustrates the certifying knowledge compilation
+and (weighted) model counting tool chain that we have developed.  A
+{\em proof generator} generates a CRAT description from a d-DNNF
+graph generated by \dfour{}, a state-of-the-art knowledge compiler.
+A {\em proof checker} then verifies that the combination of CNF and CRAT files
+satisfy our rules, and a {\em ring evaluator} can compute
+either a standard or weighted model count from the POG\@ representation.
+We evaluate the capabilities of this tool chain using
+benchmark formulas from the 2022 standard and weighted model
+competitions.  Our tools are able to handle all but the largest d-DNNF
+graphs generated by \dfour{}. 
+As the dashed box in Figure~\ref{fig:chain} indicates, this tool chain
+shifts the need for trustworthy code away from the complex and highly
+optimized knowledge compiler to a relatively simple checker and
+evaluator.  Importantly, the proof generator need not be
+trustworthy---its errors will also be caught by the proof checker.
+
+To further enhance the trustworthiness of our toolchain, we have
+developed formally verified versions of the proof checker and ring
+evaluator in the \lean{} proof assistant.  Running these two programs
+has the effect of formally verifying both the logical equivalence of the POG to the input formula and the correctness of the count.
+In addition, the
+existence of a formally verified checker serves the important role of
+establishing the soundness of our proof framework.
 
@@
 formulation of clausal proof systems for which proofs can readily be
-generated and efficiently checked
-\cite{cruz-cade-2017,RAT,wetzler14_drattrim}.  Some progress has also
+generated and efficiently (and, in some cases, formally) checked
+\cite{cruz-cade-2017,RAT,wetzler14_drattrim,Tan:2021}.  Some progress has also
 been made on extending these proof systems to quantified Boolean
@@
 model counters to also generate proof traces, and where they convert
-d-DNNF graphs into a checkable form.  We compare the performance of
-our tool chain to theirs as part of our experimental evaluation in
-Section~\ref{sect:experimental}.  Our experience with developing a
-formally verified proof system has shown us that, even for the well
-understood framework of extended resolution, ensuring that the system
-is sound can be challenging.
+d-DNNF graphs into a checkable form.  Our tool chain is not directly
+comparable to theirs---they certify the output of a standard modeling
+program, while we certify the output of a knowledge compiler and then
+produce a certified standard or weighted model count.  There tool
+chain can be applied to some programs that are pure model counters,
+while ours can be applied to pure knowledge compilers.  Nonetheless,
+we compare the performance of our tool chain to theirs as part of our
+experimental evaluation in Section~\ref{sect:experimental}.  Our
+experience with developing a formally verified CRAT proof checker has shown
+us that, even for the well understood framework of extended
+resolution, it can be challenging to formulate the full set of requirements to guarantee soundness.
 
@@
   We highlight two special classes of Boolean formulas.  A formula is
-  in {\em normal form} when negation is applied only to variables.  A
+  in {\em negation normal form} when negation is applied only to variables.  A
   formula is in {\em conjunctive normal form} (CNF) when 1) it is in
-  normal form, and 2) sum is applied only to literals.  A CNF
+  negation normal form, and 2) sum is applied only to literals.  A CNF
   formula can be represented as a set of {\em clauses}, each of which is a
@@
 
-  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} (REP) is to compute:
+  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} is to compute:
   \begin{eqnarray}
@@
   \end{eqnarray}
-  In this equation, the summation $\sum$ is computed using addition operation $\radd$, and the product $\prod$ is computed using multiplication operation $\rmul$.
+  In this equation, sum $\sum$ is computed using addition operation $\radd$, and product $\prod$ is computed using multiplication operation $\rmul$.
 \label{def:labeling}
@@
 
-Many important queries about Boolean formulas can be
+Many important properties of Boolean formulas can be
 expressed as ring evaluation problems.  The
@@
 rational numbers.  Some formulations of WMC allow
-independently assigning weights $W(x)$ and $W(\obar{x})$ for variable $x$ and its complement, with the possibility that
+independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
 $W(x) + W(\obar{x}) \not = 1$.
@@
 
-The {\em function hashing problem} provides a randomized test
+The {\em function hashing problem} provides a test
 of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
@@
 
-Generalizations of this framework are discussed in Section~\ref{sect:future}.
+We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.
 
@@
 \begin{proposition}
-Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
+For any labeling $w$, ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies:
 \begin{eqnarray}
-\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
-\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
-\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
+\textstyle
+\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
+\textstyle
+\rep\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+\textstyle
+\rep\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
 \end{eqnarray}
@@
 Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
-Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.  A nonterminal node is labeled by one of three Boolean operations:
-$\boolnot$, $\pand$, or $\por$.  Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
+Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
+A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
+Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
 It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
 There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.
-
 As can be seen, a POG is simply a way to represent a partitioned-operation
 formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
-$\phi(\nodeu)$ as the formula denoted by node $\nodeu$.
-Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi(\nodeu))$.
-
-We define the {\em size} of POG $P$, written $|P|$ to be the sum of
-the number of nonterminal nodes and the number of edges.  Ring
+$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
+Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.
+
+We define the {\em size} of POG $P$, written $|P|$ to be the 
+the number of nodes labeled $\pand$ or $\por$ plus the number of edges from these nodes to their children.  Ring
 evaluation of $P$ can be performed with at most $|P|$ ring
@@
 Our generalization is provided to allow an
-  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
+  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
   POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
@@
 
+\section{Clausal Proof Frameworks}
+
+A proof in our framework takes the form of a sequence of clause addition
+and deletion steps, with each step maintaining logical equivalence to
+the original formula.  More precisely, the status of the proof at any step will be represented by
+a set of {\em active} clauses $\phi$, i.e., those that
+have been added but not yet deleted.  Our framework is based
+on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
+steps can introduce new {\em extension variables}.  Let $Z$
+denote the set of extension variables occuring in formula $\phi$.
+Starting with input formula $\inputformula$, 
+we want to maintain the invariant that
+$\inputformula \ifandonlyif \exists Z\,\phi$.
+
+Clauses can be added in two different ways.  One is when they serve as
+the {\em defining clauses} for an extension variable.  We will use
+this form only when defining $\pand$ and $\por$ operations, as is
+described in Section~\ref{sect:crat}.  The other is by {\em
+  implication redundancy}.  That is, we can add clause $C$ to the set
+of clauses forming formula $\phi$ when $\phi \imply C$.  Clauses can
+also be deleted through implication redundancy.  That is, given a set
+of clauses $\{C\} \cup \phi$, we can remove $C$ if $\phi \imply C$.
+
+We use {\em reverse unit propagation} (RUP) to certify
+implication redundancy when adding or deleting
+clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
+RUP
+is the core rule supported by standard
+proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic.
+It is a way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
+Let $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_p$ be a clause to be
+proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
+{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
+A RUP step
+proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
+that the combination of the antecedents plus the negation of $C$ leads
+to a contradiction.  The negation of $C$ is the formula
+$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
+\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
+clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
+check processes the clauses of the antecedent in sequence, inferring
+additional unit clauses.  In processing clause $D_i$, if all but one
+of the literals in the clause is the negation of one of the
+accumulated unit clauses, then we can add this literal to the
+accumulated set.  That is, all but this literal have been falsified,
+and so it must be set to true for the clause to be satisfied.  The
+final step with clause $D_k$ must cause a contradiction, i.e., all of
+its literals are falsified by the accumulated unit clauses.
+
+Compared to the proofs of unsatisfiability generated by SAT solvers,
+it can be seen that ours has important differences.  Most
+significantly, we want our proofs to maintain logical equivalence at
+each step, whereas an unsatisfiability proof need only guarantee that
+no proof step makes a satisfiable set of clauses become
+unsatisfiable.  Therefore, our proofs must justify clause deletions as well as clause additions.
+
 \section{The CRAT POG Representation and Proof System}
-
-\begin{figure}
-\centering{
-\begin{tikzpicture}[scale=0.18]
-  \draw (8,10) rectangle (20,18);
-  \node at (14,15.2) {Knowledge};
-  \node at (14,12.8) {Compiler};
-
-  \draw (28,10) rectangle (40,18);
-  \node at (34,15.2) {Proof};
-  \node at (34,12.8) {Generator};
-
-  \draw (48,10) rectangle (60,18);
-  \node at (54,15.2) {Proof};
-  \node at (54,12.8) {Checker};
-
-  \draw (48,0) rectangle (60,8);
-  \node at (54,5.2) {Ring};
-  \node at (54,2.8) {Evaluator};
-
-  %% CNF
-  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (48,16);
-  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
-  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
-  \node [left] at (-1,20) {\large {$\phi$}};
-  \node [above] at (2,20) {\texttt{.cnf}};
-
-  %% NNF
-  \draw[thick] (20,12) [-latex] -- (28,12);
-  \node [above] at (24,12) {\texttt{.ddnnf}};
-
-  %% CRAT
-  \draw[thick] (40,12) [-latex] -- (48,12);
-  \draw[thick] (44,12) -- (44,4) [-latex] -- (48,4);
-  \node [above] at (44,12) {\texttt{.crat}};
-
-  %% OK/Not
-  \draw[thick] (60,14) [-latex] -- (64,14) ;
-  \node [right] at (64,15.2) { OK /} ;
-  \node [right] at (64,12.8) { Not OK} ;
-  \draw[thick] (60,4) [-latex] -- (64,4) ;
-  \node [right] at (64,4) {\large {$\textbf{R}(\phi)$}};
-
-\end{tikzpicture}
-}
-\caption{Tool Chain. The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT) that can be independently checked and evaluated}
-\label{fig:chain}
-\end{figure}
+\label{sect:crat}
 
@@
 form, is logically equivalent to the POG, thus enabling certified
-knowledge compilation.  The proof system is based on extended
-resolution~\cite{Tseitin:1983} with reverse unit propagation
-(RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
-proving that a set of clauses logically implies another clause.
-
-As is shown in Figure~\ref{fig:chain}
-two programs are involved in producing a certified compilation of a formula:
-\begin{itemize}
-\item The {\em proof generator} produces a CRAT file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
-As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
-\item The {\em proof checker} ensures that all of the proof conditions are satisfied.
-\end{itemize}
+knowledge compilation.
+%% The proof system is based on extended
+%% resolution~\cite{Tseitin:1983} with reverse unit propagation
+%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
+%% proving that a set of clauses logically implies another clause.
 
@@
 graph into a POG is straightforward---each node in a d-DNNF graph can
-be directly represented as a POG node.  Rather than representing every d-DNNF node explicitly,
-our program performs simplifications
-to eliminate Boolean constants.  Except for the degenerate cases where the formula is unsatisfiable or a tautology, we can
-therefore assume that the POG contains only product, sum,
-and literal nodes.
+be directly represented as a POG node.  Rather than representing every
+d-DNNF node explicitly, our program performs simplifications to
+eliminate Boolean constants.
+Except for the degenerate cases
+where the formula is unsatisfiable or a tautology, we can also assume
+that the POG does not contain any constant nodes.
+In addition, negation operators will
+only be applied to variables, and so we can view the POG as consisting
+of {\em literal} nodes corresponding to input variables and their negations, along with
+{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.
 
@@
 
-For input formula $\phi$ and its translation into a POG with root literal
+For input formula $\inputformula$ and its translation into a POG with root literal
 $r$, the most challenging part of the proof is to show that
-$\modelset(\phi) \subseteq \modelset(P)$, i.e., that any total
-assignment $\alpha$ that is a model of $\phi$ will yield $\assign(r) =
+$\modelset(\inputformula) \subseteq \modelset(P)$, i.e., that any total
+assignment $\alpha$ that is a model of $\inputformula$ will yield $\assign(r) =
 1$, for root literal $r$.
@@
 node $\nodeu$, partial assignment
-$\passign$, and a set of clauses $\psi \subseteq \phi$.
+$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
 The procedure adds a number of clauses to the proof, culminating with
 the addition of the {\em target clause}
-$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating that any total
+$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating 
+that $\left(\bigwedge_{\lit \in \passign} \lit\right) \imply u$, i.e.,
+that any total
 assignment $\assign$ such that $\passign \subseteq \assign$
 will assign $\assign(u) = 1$.
-The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \phi$.
+The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
 The result will therefore be to add the unit clause $r$ to the proof.
@@
 
-Assume that formulas $\psi$ and $\phi$ are
-represented as sets of clauses, with $\psi \subseteq \phi$.
+Let $\psi$ be a subset of the clauses in $\inputformula$.
 For partial assignment
@@
 assignment $\passign$.  The set of clauses $\psi$ can be a proper
-subset of the input clauses $\phi$ when a product node has caused
+subset of the input clauses $\inputformula$ when a product node has caused
 a splitting into clauses containing disjoint variables.
@@
   $\simplify{\psi}{\passign}$ must consist of the single unit clause
-  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \bigcup_{\lit \in \passign} \obar{\lit}$.
+  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
   Any of these can
@@
   then, since the node originated from a d-DNNF graph, there must be
-  some variable $x$ such that either $\nodeu_1$ is a terminal node with literal $x$ or $\nodeu_1$ is a
-  product node containing $x$ as a child.  In either case, we
+  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
+  product node containing a literal node for $x$ as a child.  In either case, we
   recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
@@
   $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-Similarly, either $\nodeu_0$ is a terminal node with literal $\obar{x}$ or $\nodeu_0$ is a product node containing $\obar{x}$ as
+Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
   a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
@@
   $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-  These recursive results can be combined, via two proof steps, with the second and third defining clauses for $\nodeu$
+  These recursive results can be combined with the second and third defining clauses for $\nodeu$
 (see Table~\ref{tab:defining}(B))
@@
   into a sequence of clause additions in the CRAT file.
-\item If there is a single nonterminal child ($k = 1$), we recursively call
-  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$,
-\item If there are multiple nonterminal children ($k > 1$),
+\item For a single nonterminal child ($k = 1$), we recursively call
+  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$.
+\item For multiple nonterminal children ($k > 1$),
   it must be the case that the clauses in
@@
 
-Completing the equivalence proof of input formula $\phi$ and its POG
+Completing the equivalence proof of input formula $\inputformula$ and its POG
 representation with root literal $r$ requires showing that
-$\modelset(P) \subseteq \modelset(\phi)$.  This is done in the
+$\modelset(P) \subseteq \modelset(\inputformula)$.  This is done in the
 CRAT framework by first deleting all asserted clauses, except for the
@@
 program adds $\obar{u}$ to the RUP literals and adds the appropriate
-defining clause to the list of hints.  A terminal node with literal
+defining clause to the list of hints.  A literal node for
 $\lit$ will be marked if $\lit \in C$, with no hint required.  If

### 2023-03-16 doc/lean.tex
describe model counting in Lean

@@
 
-[NOTE (JA): I am not worrying about length for the moment. This entire section can become an appendix, and then we can give an executive summary using whatever space
+[NOTE (JA): I am not worrying about length for the moment. This entire section can become an appendix,
+and then we can give an executive summary using whatever space
 we have available]
@@
 Indeed, our formalization helped us discover a bug in our implementation
-of the informal proof checker: in the ``\emph{s}'' step described at the end of Subsection~\ref{subsection:syntax}, it is sound (and necessary) to allow defining
+of the informal proof checker: in the ``\emph{s}'' step described at the end of
+Subsection~\ref{subsection:syntax}, it is sound (and necessary) to allow defining
 clauses for the new operations in the proof of $\obar{\lit}_1 \lor \obar{\lit}_2$,
-but it is generally not sound to allow the input clauses or other added clauses. [NOTE: let's add this specification to the end of Section 6.1.]
-Our verified implementation of the checker tells us that the algorithm is sound and provides a reference for other implementations.
+but it is generally not sound to allow the input clauses or other added clauses.
+[NOTE: let's add this specification to the end of Section 6.1.]
+Our verified implementation of the checker tells us that the algorithm is sound and
+provides a reference for other implementations.
 
@@
 
-In logic, the set of propositional formulas modulo equivalence is sometimes called the \emph{Lindenbaum--Tarski algebra}. We have found it convenient to define this quotient,
+In logic, the set of propositional formulas modulo equivalence is sometimes called the
+\emph{Lindenbaum--Tarski algebra}. We have found it convenient to define this quotient,
 which we denote \lstinline{PropTerm ν}, and lift the boolean operations and the entailment
@@
 \subsection{The Proof Checker}
+\label{subsection:proof:checker}
 
@@
 $\bf r$.
-We have formally proved the claim that in this case the formula corresponding to node $\bf r$ is a partitioned formula and is
+We have formally proved the claim that in this case the formula corresponding to node
+$\bf r$ is a partitioned formula and is
 logically equivalent to $\varphi$.
@@
 
-In our code, each element of a POG  is either a variable, a binary disjunction, or an arbitrary conjunction:
+In our code, each element of a POG  is either a variable, a binary disjunction,
+or an arbitrary conjunction:
 \begin{lstlisting}
@@
 \end{lstlisting}
-In other words, we rule out formulas with implication and bi-implication, and otherwise we follow the straightforward recursive definition.
+In other words, we rule out formulas with implication and bi-implication, and otherwise
+we follow the straightforward recursive definition.
 
 With these definitions in place, we can explain the specification for our checker.
-Our parser reads a CRAT file and returns an input formula \lstinline{cnf : ICnf} and a CRAT proof {pf : Array CratStep}, that is, an array of single proof steps:
+Our parser reads a CRAT file and returns an input formula \lstinline{cnf : ICnf} and a
+CRAT proof {pf : Array CratStep}, that is, an array of single proof steps:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The core checker takes those as input, as well as a specification of the final node \lstinline{r}, and either throws an exception, indicating that the proof is not well formed, or returns a POG \lstinline{P}.
+The core checker takes those as input, as well as a specification of the final node \lstinline{r},
+and either throws an exception, indicating that the proof is not well formed,
+or returns a POG \lstinline{P}.
 The specification asserts that, in the latter case, \lstinline{P.toPropForm r} is partitioned
@@
 
-[TODO(JA): I'll leave stop here now because I am running out of steam today,
-and also because Wojciech is still working on it.
-But here is where we can sketch what the checker does.
-For each line, it does some checks, it adds clauses to a clause database,
-and it add a definition to the POG.
-There is a list of invariants having to do with the original formula,
-the clause database, and the POG, and we prove that if the checks are satisfied,
-then at each step the invariants hold.
-We should list the invariants and say something about how the checks are
-performed.]
-
 \subsection{The Model Counters}
 
+We have formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem,
+Definition~\ref{def:ring_evaluation}, as follows:
+\begin{lstlisting}
+def weightCount {R : Type} [CommRing R]
+    (weight : ν → R) (φ : PropForm ν) (s : Finset ν) : R :=
+  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
+\end{lstlisting}
+Here \lstinline{R} is assumed to be a commutative ring. The curly brackets around \lstinline{R} indicate that
+this argument is meant ot be left implicit, because it can inferred from the return type of
+the weight function. Similarly, the square brackets indicate that the commutative ring structure
+is a type class argument, generally inferred from the context.
+
+More precisely, the quantity \lstinline{weightCount weight φ s} adds up the weights of the models
+of \lstinline{φ} over the set of variables \lstinline{s}. We prove that as long as \lstinline{s}
+contains the variables of \lstinline{φ}, adding extra variables does not change the sum, that is, \lstinline{vars φ ⊆ s} implies
+\begin{lstlisting}
+  weightCount weight φ s = weightCount weight φ (vars φ)
+\end{lstlisting}
+The counting scheme of Proposition~\ref{prop:ring:eval} for partitioned formulas is expressed as follows:
+\begin{lstlisting}
+ def countWeight (weight : ν → Rat) : PropForm ν → Rat
+  | tr       => 1
+  | fls      => 0
+  | var x    => weight x
+  | neg φ    => 1 - countWeight weight φ
+  | disj φ ψ => countWeight weight φ + countWeight weight ψ
+  | conj φ ψ => countWeight weight φ * countWeight weight ψ
+\end{lstlisting}
+We arbitrarily assign 0 to implications and bi-implications;
+recall that, by definition, these do not appear in partitioned formulas.
+Proposition~\ref{prop:ring:eval} is then formalized as follows:
+\begin{lstlisting}
+theorem countWeight_eq_weightCount (weight : ν → R) {φ : PropForm ν}
+    (hdec : partitioned φ) :
+  countWeight weight φ = weightCount weight φ (vars φ)
+\end{lstlisting}
+
+The functions just described are not meant to be computed, but,
+rather, serve as our mathematical reference.
+We have implemented an efficient function to calculate the weighted model count
+for an arbitrary node of a Pog, and we have proved that it computes the function above on the
+associated formula:
+\begin{lstlisting}
+theorem countWeight_eq_countWeight (pog : Pog) (weight : Var → Rat)
+    (x : Var) :
+  pog.countWeight weight x = (pog.toPropForm (.mkPos x)).countWeight weight
+\end{lstlisting}
+Here \lstinline{.mkPos x} denotes the positive literal for the variable \lstinline{x}.
+Applying this to the output of our verified CRAT proof checker,
+we obtain a proof that
+our toolchain computes the correct weighted model count of the input CNF.
+
+Lean 4 does not yet have an efficient implementation of rational arithmetic, and
+we have not yet implemented a verified version of $\drational$,
+defined in Section~\ref{sect:experimental}.
+We have therefore implemented a separate procedure that carries out an integer
+calculation of the number of models.
 The number of models that a propositional formula has depends on what we take the
@@
   | var _    => 2^(nVars - 1)
-  | neg φ      => 2^nVars - φ.countModels nVars
-  | disj φ ψ   => φ.countModels nVars + ψ.countModels nVars
-  | conj φ ψ   => φ.countModels nVars * ψ.countModels nVars / 2^nVars
-  | impl _ _   => 0
-  | biImpl _ _ => 0
+  | neg φ    => 2^nVars - countModels nVars φ
+  | disj φ ψ => countModels nVars φ + countModels nVars ψ
+  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
 \end{lstlisting}
-Note that we can carry out this calculation for any formula at all, though the
-calculation is meaningless if the formula is not partitioned.
-Also recall that if the formula is partitioned, it does not contain implication or bi-implication.
-
-We have implemented an efficient counting function for POGs and proved that
-it computes that same quantity on the corresponding formulas:
-\begin{lstlisting}
-def count (pog : Pog) (nVars : Nat) (x : Var) : Nat := ...
-
-theorem count_eq_countModels (pog : Pog) (nVars : Nat) (x : Var) :
-  pog.count nVars x = (pog.toPropForm (.mkPos x)).countModels nVars
-\end{lstlisting}
-Here \lstinline{.mkPos x} denotes the positive literal for the variable \lstinline{x}. We have also proved that the calculation above really does
+We have also implemented an efficient counting function for POGs and proved that
+it computes that same quantity on the corresponding formulas.
+Finally,
+We have proved that the calculation above really does
 return the number of models:
@@
 theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
-    (hvars : φ.vars ⊆ s) (hpar : φ.partitioned) :
-  φ.countModels s.card = card (φ.models s)
+    (hvars : vars φ ⊆ s) (hpar :  partitioned φ) :
+  countModels (card s) φ = card (models φ s)
 \end{lstlisting}
 In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
-we have that the number of models on its variables is \lstinline{φ.countModels φ.vars.card}.
+we have that the number of models on its variables is \lstinline{countModels φ (card (vars φ))}.
 
-Given a weighting function \lstinline{weight : ν → Rat} that assigns a rational
-number in $[0, 1]$ to each variable, we can calculate the weighted model count on
-partitioned formulas as follows:
-\begin{lstlisting}
-def countWeight (weight : ν → Rat) : PropForm ν → Rat
-  | tr         => 1
-  | fls        => 0
-  | var x      => weight x
-  | neg φ      => 1 - φ.countWeight weight
-  | disj φ ψ   => φ.countWeight weight + ψ.countWeight weight
-  | conj φ ψ   => φ.countWeight weight * ψ.countWeight weight
-  | impl _ _   => 0
-  | biImpl _ _ => 0
-\end{lstlisting}
-We have implemented the corresponding count function on POGs and proved that
-it calculates this quantity on the associated formulas,
-though we have not yet proved the combinatorial lemma
-that relates this calculation to the sum of the weights of the corresponding models.
+\subsection{Implementation of the Proof Checker}
+
+We have specified the behavior of the proof checker in Section~\ref{subsection:proof:checker};
+in this section we sketch the implementation.
+Establishing its correctness was, by far, the most challenging part of the formalization
+process.
+For clarity, we describe the algorithm with ordinary mathematical notation rather than
+Lean snippets.
+
+We begin by reading the input CNF formula $\varphi$ and adding its clauses to a clause
+database $C$. We initialize that POG graph $G$ with the variables of $\varphi$,
+and then iteratively process each line of the CRAT proof as follows:
+\begin{itemize}
+  \item (Summarize what each step does.)
+\end{itemize}
+We establish that all the following invariants hold at each step:
+\begin{itemize}
+  \item (Wojciech has a long list of invariants)
+\end{itemize}
+At the end \ldots [Explain how the invariants at the end imply the desired conclusion.]
 
@@
 Recall that our first step is to parse a CRAT file to read in the initial CNF formula
-and the CRAT proof. We do not verify this step. There is no strong need to verify the
-parsing of the proof, because if the checker claims success for \emph{any} proof, it does not really matter
-whether the proof was the one the user intended.
-On the other hand, it would be nice to have reassurance that the CNF formula that serves
+and the CRAT proof. We do not verify this step.
+If one is only interested in obtaining a model count or a verified model count,
+there is no strong need to verify the parsing of the proof,
+because if the checker succeeds in constructing an equivalent POG,
+it does not really matter whether the proof was the one the user intended.
+But it would be nice to have reassurance that the CNF formula that serves
 as input to our verified checker matches the one described in the file.
@@
 This involves trusting only the correctness of the print procedure and diff.
+Similarly, if one wants to establish the correctness of the POG contained in the CRAT file,
+one can print out the POG that is constructed by the checker and compare.
 
@@
 [cite ``do unchained'' paper].
-Finally, we have implemented hashmaps in terms of arrays, but we have not yet proved
-the correctness of all the primitive operations on these hashmaps.
-Thus our proofs depend on the unverified assumption that those operations have the
+Finally, Lean's standard library implements hashmaps in terms of array.
+Many of the basic properties of hashmaps have been formally verified, but not all.
+In particular, we make use of a fold operation whose verification is not yet complete.
+Thus our proofs depend on the assumption that the fold operation has the
 expected properties.
@@
 
-[TODO: Say something to establish that the procedure run in reasonable time and produce useful results; or maybe this does in the next section?]
+[TODO: Say something to establish that the procedure run in reasonable time and produce
+useful results; or maybe this goes in the next section?]

### 2023-03-16 doc/mc-paper.tex
fix merge error

@@
 \usepackage{listings}
-\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
+%\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
+\definecolor{keywordcolor}{rgb}{0.0, 0.1, 0.6}   % blue
 \definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
@@
 \begin{definition}[Ring Evaluation Problem]
+\label{def:ring_evaluation}
   For commutative ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
@@
 \begin{proposition}
-For any labeling $w$, ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies:
+\label{prop:ring:eval}
+Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
 \begin{eqnarray}
+%% V1
 \textstyle
@@
 \rep\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
+%% V2
+%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
+%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
+%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
 \end{eqnarray}
@@
 %% proving that a set of clauses logically implies another clause.
+%% Alternate version
+%% knowledge compilation.  The proof system is based on extended
+%% resolution~\cite{Tseitin:1983} with reverse unit propagation
+%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
+%% proving that a set of clauses logically implies another clause.
+%%
+%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
+%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
+%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
+%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
+%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
+%%
+%% \begin{center}
+%%   \begin{tabular}{lcccc}
+%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
+%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
+%%   \midrule
+%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
+%%   \end{tabular}
+%% \end{center}
+%%
+%% RUP is an alternative formulation of resolution.  For target clause
+%% $C$, it can be seen that applying resolution operations to the
+%% antecedent clauses from right to left will derive a clause $C'$ such
+%% that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
+%% we then have $C' \rightarrow C$.  Compared to listing each resolution
+%% operation as a separate step, using RUP as the basic proof step makes
+%% the proofs more compact.
+%%
+%% As is shown in Figure~\ref{fig:chain}
+%% two programs are involved in producing a certified compilation of a formula:
+%% \begin{itemize}
+%% \item The {\em proof generator} produces a CRAT file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
+%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
+%% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
+%% \end{itemize}
+%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b
 

### 2023-03-16 doc/mc-paper.tex
Most of the technical stuff is done

@@
 \usepackage{listings}
-%\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
-\definecolor{keywordcolor}{rgb}{0.0, 0.1, 0.6}   % blue
+\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
 \definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
@@
 \begin{definition}[Ring Evaluation Problem]
-\label{def:ring_evaluation}
   For commutative ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
@@
 \begin{proposition}
-\label{prop:ring:eval}
-Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
+For any labeling $w$, ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies:
 \begin{eqnarray}
@@
 and deletion steps, with each step maintaining logical equivalence to
-the original formula.  More precisely, the status of the proof at any step will be represented by
+the original formula.  More precisely, the status of the proof at any step is represented as
 a set of {\em active} clauses $\phi$, i.e., those that
@@
 It is a way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
-Let $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_p$ be a clause to be
+Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
 proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
@@
 Except for the degenerate cases
-where the formula is unsatisfiable or a tautology, we can also assume
+where the formula is unsatisfiable or a tautology, we can therefore assume
 that the POG does not contain any constant nodes.
@@
 
-For input formula $\inputformula$ and its translation into a POG with root literal
-$r$, the most challenging part of the proof is to show that
-$\modelset(\inputformula) \subseteq \modelset(P)$, i.e., that any total
-assignment $\alpha$ that is a model of $\inputformula$ will yield $\assign(r) =
-1$, for root literal $r$.
-This part of the proof consists of a series of clause assertions
-leading to one adding $r$ as a unit clause.
-
-We describe this part of the proof generation as a recursive procedure
+For input formula $\inputformula$ and its translation into a POG $P$
+with root literal $r$, the most challenging part of the proof is to
+show that $\modelset(\inputformula) \subseteq \modelset(P)$, i.e.,
+that any total assignment $\alpha$ that is a model of $\inputformula$
+will yield $\assign(r) = 1$, for root literal $r$.  This part of the
+proof consists of a series of clause assertions leading to one adding
+$r$ as a unit clause.  We have devised two methods of generating this
+proof.  The {\em monolithic} approach makes just one call to a
+proof-generating SAT solver and has it determine the relationship
+between the two representations.  The {\em structural} approach
+recursively traverses the POG, generating proof obligations at each
+node encountered.  It too may require calls to a proof-generating SAT
+solver.
+
+As notation,
+let $\psi$ be a subset of the clauses in $\inputformula$.
+For partial assignment
+$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
+obtained from $\psi$ by: 1) eliminating any
+clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
+2) for the remaining clauses eliminating those literals $\lit$ for
+which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
+In doing these simplifications, we also track the {\em provenance}
+of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
+More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov(C, \psi)$ denote
+those clauses $C' \in \psi$, such that
+$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
+We then extend the definition of $\prov$ to any simplified formula
+$\gamma$ as $\prov(\gamma, \psi) = \bigcup_{C \in \gamma} \prov(C, \psi)$.
+
+The monolithic approach
+takes advantage of the clausal representations of
+the input formula $\inputformula$ and the POG, given by its set of defining clauses $\pogformula$.
+Furthermore, it can
+negate the POG formula by simply negating its root literal.  Forward implication
+will hold
+when $\inputformula \imply (r \land \pogformula)$.  Equivalently,
+the formula $\inputformula \land \simplify{\pogformula}{\{\obar{r}\}}$
+should be unsatisfiable, where the
+conjunction can be expressed as the union
+of the two sets of clauses.  The proof generator writes the clauses to a file, invokes a proof-generating solver, and
+imports the proof steps as assertion steps in the CRAT file.
+Our experimental results show
+that this approach can be very effective and generates short proofs
+for smaller problems, but it does not scale well enough for routine
+use.
+
+We describe the structural approach to proof generation as a recursive procedure
 $\validate(\nodeu, \passign, \psi)$ taking as arguments POG
@@
 $\validate$.  We then refine it with some optimizations.
-
-Let $\psi$ be a subset of the clauses in $\inputformula$.
-For partial assignment
-$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
-obtained from $\psi$ by: 1) eliminating any
-clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
-2) for the remaining clauses eliminating those literals $\lit$ for
-which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
-In doing these simplifications, we also track the {\em provenance}
-of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
-More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov(C, \psi)$ denote
-those clauses $C' \in \psi$, such that
-$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
-We then extend the definition of $\prov$ to any simplified formula
-$\gamma$ as $\prov(\gamma, \psi) = \bigcup_{C \in \gamma} \prov(C, \psi)$.
 
@@
 Observe that most of these steps involve a polynomial number of
-operations per recursive call.  The only step that could have
+operations per recursive call.  The only one that could have
 exponential complexity is when a SAT solver must be invoked to validate a
@@
 delete clause $C$ such that the remaining set of clauses will form the
-CNF formula $\psi$.  Deleting $C$ requires proving that
-$\modelset(\psi) = \modelset(\psi \lor C)$, i.e., that deleting $C$
-does not make the formula easier to satisfy.
+CNF formula $\phi$.  Deleting $C$ requires proving that
+$\phi \imply C$.
 
@@
 The performance of forward implication, both in its execution time and
-the size of the proof generated by two optimizations described here.
-A key feature of these optimizations is that they do not require any changes to the
+the size of the proof generated can be improved by two optimizations described here.
+A key feature is that they do not require any changes to the
 proof framework---they build on the power of extended resolution to
@@
 having many---tens or even hundreds---of literals as children.  The earlier formulation of $\validate$
-considered each literal $\lit \in L$ separately,
+considers each literal $\lit \in L$ separately,
 calling a SAT solver
-on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ when $\lit$ could not be validated with BCP\@.
+on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ when $\lit$ cannot be validated with BCP\@.
 
@@
 then becomes to prove that any total assignment must yield 1 for extension
-variable $v$.  This can be done by contradiction using the SAT solver.
-In particular, looking at the defining clauses for $\nodev$ (Table
+variable $v$.  A single call to the solver can generate this proof.
+Looking at the defining clauses for $\nodev$ (Table
 \ref{tab:defining}(A)), we can see that assigning 0 to $v$ will cause
@@
 root $\noder$ effectively expands the POG into a tree, and this can
-lead to an exponential number of calls.  We can exploit the property
-that a shared subgraph in the graph generated by a top-down d-DNNF
-compiler arises when the program detects that the simplified set of
+lead to an exponential number of calls.  
+These shared subgraphs arise when the knowledge compiler employs {\em clause caching}
+detecting that the simplified set of
 clauses arising from one partial assignment to the literals matches that
-of a previous partial assignment.  When this d-DNNF node is translated into POG
+of a previous partial assignment~\cite{darwiche:aaai:2002}.
+When this d-DNNF node is translated into POG
 node $\nodeu$, the proof generator can assume (and also check), that
@@
 
-The forward implication proof can exploit this property of shared
-subgraphs by constructing and proving a {\em lemma} for each node
+The proof generator can exploit the sharing of subgraphs
+by constructing and proving a {\em lemma} for each node
 $\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
@@
 $\assign(u) = 1$.  This lemma is then invoked for every node having
-$\nodeu$ as a child.  The net effect is to allow $\validate$ to
-exploit the graph structure of the POG, invoking $\validate$ on each
-node at most once.
+$\nodeu$ as a child.  The net effect is that generator will recurse during a call to $\validate$ only once for each node in the POG\@.
 

### 2023-03-16 doc/mc-paper.tex

@@
 \begin{proposition}
-For any labeling $w$, ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies:
+\label{prop:ring:eval}
+Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
 \begin{eqnarray}

### 2023-03-16 doc/mc-paper.tex
Polished technical presentation

@@
 \newcommand{\inputformula}{\phi_I}
-\newcommand{\pogformula}{\phi_P}
+\newcommand{\pogformula}{\theta_P}
 
@@
 %\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10003752.10003790.10003794</concept_id>
+       <concept_desc>Theory of computation~Automated reasoning</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Automated reasoning}
 
 %\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+\keywords{Propositional model counting, Proof checking}
 
@@
 Model counting is a challenging problem---more challenging that the
-already NP-hard problem of Boolean (un)satisfiability.  Several
+already NP-hard problem of Boolean satisfiability.  Several
 tractable variants of Boolean satisfiability, including 2-SAT, become
@@
 evaluator in the \lean{} proof assistant.  Running these two programs
-has the effect of formally verifying both the logical equivalence of the POG to the input formula and the correctness of the count.
-In addition, the
-existence of a formally verified checker serves the important role of
-establishing the soundness of our proof framework.
+has the effect of formally verifying both the logical equivalence of
+the POG to the input formula and the correctness of the count.  In
+addition, the existence of a formally verified checker serves the
+important role of establishing the soundness of our proof framework.
+Indeed, in the course of developing a formally verified checker, we
+identified subtle aspects of the proof obligations required to assure
+soundness.
 
@@
 generated and efficiently (and, in some cases, formally) checked
-\cite{cruz-cade-2017,RAT,wetzler14_drattrim,Tan:2021}.  Some progress has also
+\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress has also
 been made on extending these proof systems to quantified Boolean
@@
 program, while we certify the output of a knowledge compiler and then
-produce a certified standard or weighted model count.  There tool
+produce a certified standard or weighted model count.  Their tool
 chain can be applied to some programs that are pure model counters,
@@
 experimental evaluation in Section~\ref{sect:experimental}.  Our
-experience with developing a formally verified CRAT proof checker has shown
+experience with developing a formally verified proof checker has shown
 us that, even for the well understood framework of extended
@@
   literals, and the formula is the product of its clauses.  We use
-  set notation to reference the clauses in a formula and the
-  literals in a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause.
+  set notation to reference the clauses within a formula and the
+  literals within a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause and the literal as a {\em unit} literal.
 This literal must be assigned value $1$ by any satisfying assignment of the formula.
@@
  satisfies the following for all product and sum operations:
-      \begin{itemize}
+      \begin{enumerate}
       \item The arguments to a product must have disjoint dependency sets.  That is, operation
@@
         $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \end{itemize}
+      \end{enumerate}
      We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.
@@
 We propose a general framework for summarizing properties of Boolean
-formulas inspired by the formulation of {\em algebraic model counting}
-by Kimmig, et al.~\cite{kimmig:jal:2017}.
+formulas inspired by the formulation of {\em algebraic model counting}.~\cite{kimmig:jal:2017}.
 
@@
 label $w(x) = 1/2$ for every variable $x$.
-Ring evaluation of formula $\phi$ will then compute the {\em density} of
+Ring evaluation of formula $\phi$ gives the {\em density} of
 the formula, i.e., the fraction of all possible total assignments that are
-models.  For $n = |\varset|$, scaling the density by $2^n$ will then
-yield the number of models.
+models.  For $n = |\varset|$, scaling the density by $2^n$ 
+yields the number of models.
 
@@
 $W(x) + W(\obar{x}) \not = 1$.
-We can convert this more general form into a
+We can cast this as a
 ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
 performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
-variable $x$, and then computing the final result as $\rep(\phi, w)\;
-\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.
+variable $x$, and computing the weighted count
+as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
+Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.
 
@@
 least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
-Therefore, ring evaluation can be used as part of a
+Function hashing can therefore be used as part of a
 randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
 
-We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.
+%We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.
 
@@
 \textstyle
-\rep\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+{\Large \rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
 \textstyle
-\rep\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
+{\Large \rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
 %% V2
@@
 There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.
-As can be seen, a POG is simply a way to represent a partitioned-operation
-formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
+As can be seen, a POG is represents a partitioned-operation
+formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
 $\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
@@
 \item They allow negation of arbitrary nodes in the graph, not just
-  variables.  As we have seen, ring evaluation can readily handle negation, and so we allow it in the interest of generality.
+  variables.  As we have seen, ring evaluation can readily handle negation.
+  %, and so we allow it in the interest of generality.
 %%  Negation could be useful in some contexts, such as when
@@
   any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
-Our generalization is provided to allow an
-  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
-  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
+%%Our generalization is provided to allow an
+%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
+%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
 \end{itemize}
-  Both of these generalizations allow more flexibility in the POG
+  Although not required for representing d-DNNF graphs,
+  both of these generalizations allow more flexibility in the POG
   representation while maintaining the ability to efficiently perform ring evaluation.
@@
 on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
-steps can introduce new {\em extension variables}.  Let $Z$
+steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
+Let $Z$
 denote the set of extension variables occuring in formula $\phi$.
@@
 Clauses can be added in two different ways.  One is when they serve as
-the {\em defining clauses} for an extension variable.  We will use
-this form only when defining $\pand$ and $\por$ operations, as is
+the {\em defining clauses} for an extension variable.  This form
+occurs only when defining $\pand$ and $\por$ operations, as is
 described in Section~\ref{sect:crat}.  The other is by {\em
@@
 
-The CRAT format draws its inspiration from the LRAT format for Boolean
-formulas~\cite{lrat} and the QRAT format for quantified Boolean
-formulas (QBF)~\cite{heule:JAR2014}.  The following are its key properties:
+The CRAT format draws its inspiration from the LRAT~\cite{lrat} and
+QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
+The following are its key properties:
 \begin{itemize}
@@
   Declaring a node $\nodeu$
-implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses
+implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{\nodeu}$
   encoding the product or sum operation.
@@
 Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
-The checker is provided with the input formula as a separate file.
+%% The checker is provided with the input formula as a separate file.
 As with other clausal proof formats, a variable is
@@
 Finally, the literal denoting the root of the POG is declared with the
-{\tt r} command.  It can occur anywhere in the file.  Typically, it
-will be the extension variable representing the root of a graph, but in
-degenerate cases it can be the positive or negative literal of an
-input variable.
+{\tt r} command.  It can occur anywhere in the file.  Except in degenerate cases, it
+will be the extension variable representing the root of a graph
+%, but in
+%degenerate cases it can be the positive or negative literal of an
+%input variable.
 
@@
 clause deletions represents a systematic transformation of the input formula
-into a POG\@.  Validating all of these steps serves to prove that the
+into a POG formula.  Validating all of these steps serves to prove that the
 POG is logically equivalent to the input formula.
 At the completion of the proof, the following conditions must hold:
-\begin{itemize}
-\item There should exactly one remaining clause that was added via RUP
-  addition, and this should be a unit clause consisting of the root literal $r$.
+\begin{enumerate}
+\item There is exactly one remaining clause that was added via RUP
+  addition, and this is a unit clause consisting of root literal $r$.
 \item All of the input clauses have been deleted.
-\end{itemize}
+\end{enumerate}
+Let $\inputformula$ denote the input formula and $\pogformula$ represent the formula encoded by unit clause $r$ and the defining clauses for the POG nodes:
+$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{\nodeu}$.  We write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
+any assignment to the input variables implicitly defines the assignments to the extension variables.
 The sequence of clause addition steps provide a {\em forward implication} proof that
-$\modelset(\phi) \subseteq \modelset(P)$.  That is, any total
+$\modelset(\inputformula) \subseteq \modelset(P)$.  That is, any total
 assignment $\assign$ satisfying the input formula must also satisfy
@@
 that
-$\modelset(P) \subseteq \modelset(\phi)$.
+$\modelset(P) \subseteq \modelset(\inputformula)$.
 
@@
 
-Figure~\ref{fig:chain} illustrates one method of generating a CRAT
-file from an input formula in conjunctive normal form.  Our tool chain
-is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
-but it would also work for other tools that use a top-down search
-strategy to generate a d-DNNF representation.  Converting a d-DNNF
+%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
+%%file from an input formula in conjunctive normal form.  Our tool chain
+%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
+%%but it would also work for other tools that use a top-down search
+%%strategy to generate a d-DNNF representation.
+Converting a d-DNNF
 graph into a POG is straightforward---each node in a d-DNNF graph can
@@
 eliminate Boolean constants.
-Except for the degenerate cases
-where the formula is unsatisfiable or a tautology, we can therefore assume
+Except for the degenerate cases,
+%where the formula is unsatisfiable or a tautology,
+we can therefore assume
 that the POG does not contain any constant nodes.
-In addition, negation operators will
-only be applied to variables, and so we can view the POG as consisting
+In addition, negation is only
+applied to variables, and so we can merge the negation operations into the terminal nodes,
+viewing the POG as consisting
 of {\em literal} nodes corresponding to input variables and their negations, along with
@@
 takes advantage of the clausal representations of
-the input formula $\inputformula$ and the POG, given by its set of defining clauses $\pogformula$.
+the input formula $\inputformula$ and the POG formula $\pogformula$, consisting of the defining clauses of its nodes plus a unit clause for root literal $r$.
 Furthermore, it can
-negate the POG formula by simply negating its root literal.  Forward implication
+negate the POG formula by simply negating its root literal.  More precisely $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{\nodeu}}{\{\obar{r}\}}$
+Forward implication
 will hold
-when $\inputformula \imply (r \land \pogformula)$.  Equivalently,
-the formula $\inputformula \land \simplify{\pogformula}{\{\obar{r}\}}$
+when $\inputformula \imply \pogformula$.  Equivalently,
+the formula $\inputformula \land \neg \pogformula$
 should be unsatisfiable, where the
 conjunction can be expressed as the union
-of the two sets of clauses.  The proof generator writes the clauses to a file, invokes a proof-generating solver, and
-imports the proof steps as assertion steps in the CRAT file.
+of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
+For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
 Our experimental results show
@@
 The procedure adds a number of clauses to the proof, culminating with
-the addition of the {\em target clause}
+the addition of the {\em target} clause:
 $u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
@@
 The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
-The result will therefore be to add the unit clause $r$ to the proof.
+The result will therefore be to add unit clause $r$ to the proof.
 Here we present a correct, but somewhat inefficient formulation of
@@
   $\lit \in L$, we must prove that any total assignment $\alpha$, such that
-  $\passign \subset \alpha$ must have $\alpha(\lit) = 1$.  In some
+  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
   cases, this can be done by simple Boolean constraint propagation (BCP).
@@
 The performance of forward implication, both in its execution time and
-the size of the proof generated can be improved by two optimizations described here.
-A key feature is that they do not require any changes to the
-proof framework---they build on the power of extended resolution to
-create new logical structures within the proof.
+the size of the proof generated can be improved by two optimizations
+described here.  A key feature is that they do not require any changes
+to the proof framework---they build on the power of extended
+resolution to create new logical structures within the proof.  They
+involve declaring new product nodes to encode both conjunctions and
+disjunctions (via DeMorgan's Law) of literals.  These nodes are not
+part of the POG representation of the formula; they serve only to enable the forward implication proof.
 
@@
 then becomes to prove that any total assignment must yield 1 for extension
-variable $v$.  A single call to the solver can generate this proof.
-Looking at the defining clauses for $\nodev$ (Table
-\ref{tab:defining}(A)), we can see that assigning 0 to $v$ will cause
-the first defining clause to simplify to
-$C = \obar{\lit}_1 \lor  \obar{\lit}_2 \lor \cdots \lor \obar{\lit}_m$, and for the remaining
-  clauses to become tautologies.  The SAT solver can be invoked on formula
-  $\simplify{\psi}{\passign} \cup \{ C \}$ to show that is unsatisfiable.
+variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
+  $\simplify{\psi}{\passign} \cup \simplify{\theta_{\nodev}}{\{ \obar{v} \}}$ to show that is unsatisfiable.
   The proof steps can be mapped back into clause addition steps in the CRAT file, using as hints the
@@
 These shared subgraphs arise when the knowledge compiler employs {\em clause caching}
-detecting that the simplified set of
+to detect that the simplified set of
 clauses arising from one partial assignment to the literals matches that
@@
 $\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
-total assignment $\assign$ that satisfies formula $\gamma$ must yield
+total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ must yield
 $\assign(u) = 1$.  This lemma is then invoked for every node having
-$\nodeu$ as a child.  The net effect is that generator will recurse during a call to $\validate$ only once for each node in the POG\@.
+$\nodeu$ as a child.  As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.
 

### 2023-03-16 doc/mc-data.tex
Put figures in separate files ; move parts of presentation into appendices

@@
 
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
-      legend style={at={(0.30,0.96)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=2.4,
-                              xmode=log,xmin=0.01,xmax=4000,
-                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
-                              ymin=0.01, ymax=29000,
-                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
-                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \input{data-formatted/time-d4-verify}
-    \input{data-formatted/time-d4-verify-onesided-only}
-    \input{data-formatted/time-d4-failure}
-    \legend{
-      \scriptsize \textsf{Full validation},
-      \scriptsize \textsf{One-sided validation},
-      \scriptsize \textsf{Resource limit exceeded},
-    }
-    \addplot[mark=none, dashed] coordinates{(0.01,25000) (4000, 25000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (2000,20000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (200, 20000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (20, 20000)};
-    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
-    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
-    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
-    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
-    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Combined runtime for CRAT proof generation and checking  as function of D4 runtime}
-\label{fig:d4:crat}
-\end{figure}
+\input{figure-runtime-d4-verify}
 
@@
 
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
-                              x post scale=1.8, y post scale=2.15,
-                              xmode=log,xmin=100,xmax=1000000000, 
-                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
-    \input{data-formatted/defining+total}
-    \node[left] at (axis cs: 900,1200) {$1\times$};
-    \node[left] at (axis cs: 900,12000) {$10\times$};
-    \node[left] at (axis cs: 900,120000) {$100\times$};
-    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
-    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Total number of clauses in CRAT file as function of number of defining clauses}
-\label{fig:defining:total}
-\end{figure}
+\input{figure-clauses-d4-verify}
 
@@
 
-\begin{figure}
-\begin{minipage}{0.49\textwidth}
-\centering{%
-\begin{tikzpicture}[scale = 0.80]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
-      legend style={at={(0.50,0.98)}},
-      legend cell align={left},
-                              x post scale=1.1, y post scale=1.8,
-                              xmode=log,xmin=1000,xmax=1e7, 
-                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
-                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
-%                              xlabel={Optimized Proof Clauses},
-%                              ylabel={Proof Clauses},
-            ]
-    \input{data-formatted/clauses-lemma-nolemma}
-    \input{data-formatted/clauses-nosplit-split}
-    \legend{
-      \scriptsize \textsf{No lemmas},
-      \scriptsize \textsf{No literal grouping},
-    }
-    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
-
-    \node[left] at (axis cs: 5000,7000) {$1\times$};
-    \node[left] at (axis cs: 5000,70000) {$10\times$};
-    \node[left] at (axis cs: 5000,700000) {$100\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-\centering{%
-\begin{tikzpicture}[scale = 0.80]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8, 
-      legend style={at={(0.50,0.98)}},
-      legend cell align={left},
-                              x post scale=1.1, y post scale=1.8,
-                              xmode=log,xmin=1000,xmax=1e7, 
-                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
-                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
-%                              xlabel={Optimized Proof Clauses},y
-%                              ylabel={Proof Clauses},
-            ]
-    \input{data-formatted/clauses-optimized-unoptimized}    
-    \legend{
-      \scriptsize \textsf{Neither optimization},
-    }
-    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
-
-    \node[left] at (axis cs: 5000,7000) {$1\times$};
-    \node[left] at (axis cs: 5000,70000) {$10\times$};
-    \node[left] at (axis cs: 5000,700000) {$100\times$};
-  \end{axis}
-\end{tikzpicture}
-} % centering
-\end{minipage}
-\caption{Proof clauses when one (left) or both (right) optimizations are disabled, versus when both optimizations are enabled}
-\label{fig:optimized:lessoptimized}
-\end{figure}
+\input{figure-clauses-optimizations}
 
@@
 
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
-      legend style={at={(0.22,0.98)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=2.2,
-                              xmode=log,xmin=0.01,xmax=1900,
-                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100,{1,000}},
-                              ymin=0.01, ymax=1900,
-                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
-                              xlabel={Structured generation + checking (seconds)}, ylabel={Monolithic generation + checking (seconds)},
-            ]
-    \input{data-formatted/time-structured+monolithic}
-    \input{data-formatted/time-structured+monolithic_timeout}
-    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (1000.0,10.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (1000.0,100.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (1000.0,1000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (100, 1000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (10, 1000)};
-    \addplot[color=black,dashed] coordinates{(0.01,1500) (1500,1500)};
-    \addplot[color=black,dashed] coordinates{(1500,0.01) (1500,1500)};
-    \node[left] at (axis cs: 5.0,0.06) {$0{.}01\times$};
-    \node[left] at (axis cs: 0.5,0.06) {$0{.}1\times$};
-    \node[left] at (axis cs: 0.05,0.06) {$1\times$};
-    \node[left] at (axis cs: 0.05,0.6) {$10\times$};
-    \node[left] at (axis cs: 0.05,6.0) {$100\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Monolithic versus Structured Validation: Time}
-\label{fig:monolithic:time}
-\end{figure}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10},
-      legend style={at={(0.22,0.98)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=2.2,
-                              xmode=log,xmin=1e3,xmax=1e7,
-                              ymode=log, ymin=1e3, ymax=1e7,
-                              xlabel={Structured (total clauses)}, ylabel={Monolithic (total clauses)},
-            ]
-    \input{data-formatted/clauses-total-structured+monolithic}
-%    \addplot[mark=none, color=black] coordinates{(0.01,1500) (1500,1500)};
-%    \addplot[mark=none, color=black] coordinates{(1500,0.01) (1500,1500)};
-%    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (10000.0,10.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(1e3,1e4) (1e7,1e8)};
-    \addplot[mark=none, color=lightblue] coordinates{(1e3,1e3) (1e7,1e7)};
-    \addplot[mark=none, color=lightblue] coordinates{(1e3,1e2) (1e7,1e6)};
-%    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
-%    \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
-    \node[left] at (axis cs: 3500,40000) {$10\times$};
-    \node[left] at (axis cs: 3500,4000) {$1\times$};
-    \node[left] at (axis cs: 35000,4000) {$0.1\times$};
-%    \node[left] at (axis cs: 0.05,60.0) {$1000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Monolithic versus Structured Validation: Clauses}
-\label{fig:monolithic:clauses}
-\end{figure}
+\input{figure-runtime-monolithic}
 
@@
 
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10},
-      legend style={at={(0.22,0.97)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=2.2,
-                              xmode=log,xmin=0.01,xmax=1900,
-                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100,{1,000}},
-                              ymode=log, ymin=0.01, ymax=1900,
-                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
-                              xlabel={CRAT total time (seconds)}, ylabel={MICE total time (seconds)},
-            ]
-    \input{data-formatted/time-mice-compare}
-    \input{data-formatted/time-mice-error}
-    \legend{
-      \scriptsize \textsf{Correct or timeout},
-      \scriptsize \textsf{MICE error},
-    }
-    \input{data-formatted/time-mice-timeout}
-
-    \addplot[color=black,dashed] coordinates{(0.01,1500) (1500,1500)};
-    \addplot[color=black,dashed] coordinates{(1500,0.01) (1500,1500)};
-
-%    \addplot[mark=none, color=black] coordinates{(0.01,1500) (1500,1500)};
-%    \addplot[mark=none, color=black] coordinates{(1500,0.01) (1500,1500)};
-    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (1000.0,1.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (1000.0,10.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (1000.0,100.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (1000.0,1000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (100,1000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (10,1000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1,1000)};
-    \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
-    \node[left] at (axis cs: 5.0,0.06) {$0{.}01\times$};
-    \node[left] at (axis cs: 0.5,0.06) {$0{.}1\times$};
-    \node[left] at (axis cs: 0.05,0.06) {$1\times$};
-    \node[left] at (axis cs: 0.05,0.6) {$10\times$};
-    \node[left] at (axis cs: 0.05,6.0) {$100\times$};
-    \node[left] at (axis cs: 0.05,60.0) {$1000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Running Time for MICE versus CRAT proof chains}
-\label{fig:mice}
-\end{figure}
+\input{figure-mice-crat}
 

### 2023-03-16 doc/mc-paper.tex
Put figures in separate files ; move parts of presentation into appendices

@@
 
-\subsection{CRAT Example}
+Appendix~\ref{app:crat:example} shows the CRAT description for
+an input formula with five clauses, yielding a POG with five
+nonterminal nodes.  It explains how the clause addition and deletion
+steps yield a proof of equivalence between the input formula and its POG
+representation.
+
+\section{Generating CRAT from d-DNNF}
+\label{section:generating:crat}
+
+%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
+%%file from an input formula in conjunctive normal form.  Our tool chain
+%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
+%%but it would also work for other tools that use a top-down search
+%%strategy to generate a d-DNNF representation.
+Converting a d-DNNF
+graph into a POG is straightforward---each node in a d-DNNF graph can
+be directly represented as a POG node.  Rather than representing every
+d-DNNF node explicitly, our program performs simplifications to
+eliminate Boolean constants.
+Except for the degenerate cases,
+%where the formula is unsatisfiable or a tautology,
+we can therefore assume
+that the POG does not contain any constant nodes.
+In addition, negation is only
+applied to variables, and so we can merge the negation operations into the terminal nodes,
+viewing the POG as consisting
+of {\em literal} nodes corresponding to input variables and their negations, along with
+{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.
+
+\subsection{Forward implication proof}
+
+For input formula $\inputformula$ and its translation into a POG $P$
+with root literal $r$, the most challenging part of the proof is to
+show that $\modelset(\inputformula) \subseteq \modelset(P)$, i.e.,
+that any total assignment $\alpha$ that is a model of $\inputformula$
+will yield $\assign(r) = 1$, for root literal $r$.  This part of the
+proof consists of a series of clause assertions leading to one adding
+$r$ as a unit clause.  We have devised two methods of generating this
+proof.  The {\em monolithic} approach makes just one call to a
+proof-generating SAT solver and has it determine the relationship
+between the two representations.  The {\em structural} approach
+recursively traverses the POG, generating proof obligations at each
+node encountered.  It too may require calls to a proof-generating SAT
+solver.
+
+As notation,
+let $\psi$ be a subset of the clauses in $\inputformula$.
+For partial assignment
+$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
+obtained from $\psi$ by: 1) eliminating any
+clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
+2) for the remaining clauses eliminating those literals $\lit$ for
+which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
+In doing these simplifications, we also track the {\em provenance}
+of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
+More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov(C, \psi)$ denote
+those clauses $C' \in \psi$, such that
+$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
+We then extend the definition of $\prov$ to any simplified formula
+$\gamma$ as $\prov(\gamma, \psi) = \bigcup_{C \in \gamma} \prov(C, \psi)$.
+
+The monolithic approach
+takes advantage of the clausal representations of
+the input formula $\inputformula$ and the POG formula $\pogformula$, consisting of the defining clauses of its nodes plus a unit clause for root literal $r$.
+Furthermore, it can
+negate the POG formula by simply negating its root literal.  More precisely $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{\nodeu}}{\{\obar{r}\}}$
+Forward implication
+will hold
+when $\inputformula \imply \pogformula$.  Equivalently,
+the formula $\inputformula \land \neg \pogformula$
+should be unsatisfiable, where the
+conjunction can be expressed as the union
+of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
+For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
+Our experimental results show
+that this approach can be very effective and generates short proofs
+for smaller problems, but it does not scale well enough for routine
+use.
+
+We describe the structural approach to proof generation as a recursive procedure
+$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
+node $\nodeu$, partial assignment
+$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
+The procedure adds a number of clauses to the proof, culminating with
+the addition of the {\em target} clause:
+$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
+that $\left(\bigwedge_{\lit \in \passign} \lit\right) \imply u$, i.e.,
+that any total
+assignment $\assign$ such that $\passign \subseteq \assign$
+will assign $\assign(u) = 1$.
+The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
+The result will therefore be to add unit clause $r$ to the proof.
+Here we present a correct, but somewhat inefficient formulation of
+$\validate$.  We then refine it with some optimizations.
+
+The recursive call $\validate(u, \passign, \psi)$ assumes that we have
+traversed a path from the root node down to node $\nodeu$, with the
+literals encountered in the product nodes forming the partial
+assignment $\passign$.  The set of clauses $\psi$ can be a proper
+subset of the input clauses $\inputformula$ when a product node has caused
+a splitting into clauses containing disjoint variables.
+The subgraph with root node $\nodeu$ should be a POG representation of the formula
+$\simplify{\psi}{\passign}$.  Having the
+routine add its target clause indicates that any total assignment
+$\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.
+
+The process for generating such a proof depends on the form of node $\nodeu$:
+\begin{enumerate}
+\item If $\nodeu$ is a literal $\lit'$, then the formula
+  $\simplify{\psi}{\passign}$ must consist of the single unit clause
+  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
+  Any of these can
+  serve as the target clause.
+\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
+  then, since the node originated from a d-DNNF graph, there must be
+  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
+  product node containing a literal node for $x$ as a child.  In either case, we
+  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
+  This will cause the addition of the target clause
+  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
+Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
+  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
+  causing the addition of the target clause
+  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
+  These recursive results can be combined with the second and third defining clauses for $\nodeu$
+(see Table~\ref{tab:defining}(B))
+  to generate the target clause for $\nodeu$.
+\item If $\nodeu$ is a product node, then we can divide its children
+  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
+  \begin{enumerate}
+    \item For each literal
+  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
+  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
+  cases, this can be done by simple Boolean constraint propagation (BCP).
+  In other cases, we must prove that the formula
+  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
+  do so by writing the formula to a file, invoking a proof-generating
+  SAT solver, and then converting the generated unsatisfiability proof
+  into a sequence of clause additions in the CRAT file.
+\item For a single nonterminal child ($k = 1$), we recursively call
+  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$.
+\item For multiple nonterminal children ($k > 1$),
+  it must be the case that the clauses in
+  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
+  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
+  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
+  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
+  literals.
+  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
+  We recursively call
+  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \psi_i\right)$.
+\end{enumerate}
+  We then combine as hints the results from the BCP and SAT calls for
+  the literals, the recursively computed target clauses, and all but
+  the first defining clause for node $\nodeu$
+(see Table~\ref{tab:defining}(A))
+to generate the target clause for node $\nodeu$.
+\end{enumerate}
+Observe that most of these steps involve a polynomial number of
+operations per recursive call.  The only one that could have
+exponential complexity is when a SAT solver must be invoked to validate a
+literal.
+
+\subsection{Reverse implication proof}
+
+Completing the equivalence proof of input formula $\inputformula$ and its POG
+representation with root literal $r$ requires showing that
+$\modelset(P) \subseteq \modelset(\inputformula)$.  This is done in the
+CRAT framework by first deleting all asserted clauses, except for the
+final unit clause for root literal $r$, and then deleting all of the
+input clauses.  Suppose at some point in this process that we want to
+delete clause $C$ such that the remaining set of clauses will form the
+CNF formula $\phi$.  Deleting $C$ requires proving that
+$\phi \imply C$.
+
+The asserted clauses can be deleted in reverse order, using the same
+hints that were used in their original assertions.  By reversing the
+order, those clauses that were used as hints when a clause was
+added will still remain when it is deleted.
+
+Each input clause deletion can be done as a single RUP step.  The
+proof generator builds a sequence of hints from among the defining
+clauses of the POG nodes via a single, bottom-up pass through the
+graph.  The RUP deletion proof for input clause $C$ effectively proves that any
+total assignment $\assign$ that does not satisfy $C$ will yield
+$\assign(r) = 0$.  It starts with the set of literals
+$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
+assignment $\assign$ to falsify clause $C$.
+It then
+adds literals via unit propagation until a
+conflict arises.    Unit literal $r$ gets
+added right away, setting up a potential conflict.
+
+Working upward through the graph, node $\nodeu$ is {\em marked} when
+the collected set of literals force $u$ to evaluate to $0$.  When marking $\nodeu$, the
+program adds $\obar{u}$ to the RUP literals and adds the appropriate
+defining clause to the list of hints.  A literal node for
+$\lit$ will be marked if $\lit \in C$, with no hint required.  If
+product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
+$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
+added to the hints.  Marking sum node $\nodeu$ requires that its two children are marked.
+The first defining
+clause for this node (see Table~\ref{tab:defining}(B)) will then be added as a hint.  At the very end, the program
+(assuming the reverse implication holds) will attempt to mark root
+node $\noder$, which would require $\assign(r) = 0$, yielding a
+conflict.
+
+It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
+each clause deletion requires a single RUP step having a set of hints that is
+bounded by the number of POG nodes.
+
+\section{Optimizations}
+
+The performance of the proof generator for forward implication, both in its execution time and
+the size of the proof generated, can be improved by two optimizations
+described here.  A key feature is that they do not require any changes
+to the proof framework---they build on the power of extended
+resolution to create new logical structures within the proof.  They
+involve declaring new product nodes to encode both products of literals.
+These nodes are not part of the POG
+representation of the formula; they serve only to enable the forward
+implication proof.  Here we only summarize the two optimizations.
+More details are presented in Appendix~\ref{app:optimizations}.
+
+{\em Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
+having many---tens or even hundreds---of literals as children.  The
+earlier formulation of $\validate$ considers each literal $\lit \in L$
+separately, calling a SAT solver for every literal that cannot be
+validated with BCP\@.  The literal grouping optimization handles
+all of these literals together.  It defines product node $\nodev$
+having the literals as children.  The goal then becomes to prove that
+any total assignment must yield 1 for extension variable $v$.  Calling
+a solver with $v$ set to 0 yields an unsatisfiability proof that can
+be mapped back to a sequence of clause additions in the CRAT file that
+validates all of the literals.
+
+{\em Lemmas:} Our formulation of $\validate$ requires each call at a
+node $\nodeu$ to recursively validate all of its children.  This
+effectively expands the graph into a tree, potentially requiring an
+exponential number of recursive steps.  Instead, for each node
+$\nodeu$ having $\indegree(u) > 1$, the program can define and
+generate the proof of a lemma for $\nodeu$ when it is first reached by
+a call to $\validate$ and then apply this lemma for this and
+subsequent calls.  The lemma states that the POG with root node
+$\nodeu$ satisfies forward implication for a formula
+$\gamma_{\nodeu}$, where some of the clauses in this formula can be
+input clauses from $\inputformula$, but others can be simplified
+versions of input clauses.
+The key idea is to introduce product nodes to encode (via DeMorgan's
+Laws) the simplified clauses and have these serve as lemma arguments.
+
+The combination of these two optimization guarantees that 1) each call
+to $\validate$ for a product node will cause at most one invocation of
+the SAT solver, and 2) each call to $\validate$ for any node $\nodeu$
+will cause further recursive calls only once.  Our experimental
+results (Appendix~\ref{app:experiments}) shows that these
+optimizations yield substantial benefits.
+
+\input{lean}
+
+\section{Experimental Evaluation}
+\label{sect:experimental}
+We have implemented versions of the three programs that, in addition to the \dfour{} knowledge compiler,  form the
+toolchain illustrated in Figure~\ref{fig:chain}.  One version,
+including a proof generator, proof checker, and ring evaluator for
+standard and weighted model counting, is designed to be fully
+scalable.  The other version, consisting of a proof checker and
+ring evaluator has been formally verified within the \lean{} proof
+system.
+
+Our long-term goal is to have a toolchain suitable for use in a
+knowledge compilation competition, along the lines of the annual
+competitions for other automated reasoning tools.  Our experiments
+consider the benchmark problems and computing capabilities that would
+arise in such a competition and evaluate whether our implementations
+could handle these problems.  As with the original proof-generating
+SAT competitions, we assume that the proofs at a competition could be
+checked with the unverified checker.  Having a verified checker
+that works on some subset of the benchmarks would provide an
+additional degree of certainty.
+
+Our proof generator is written in C/C++ and uses
+\cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
+proof steps back into hinted CRAT clause additions, the generator can
+use either its own RUP proof generator, or it can invoke
+\dtrim{}~\cite{RAT}.  The latter yields shorter proofs and scales well
+to large proofs, but each invocation has a high startup cost.  We
+therefore only use it when solving larger problems (currently ones
+with over 1000 clauses).
+
+The proof generator can optionally be instructed to generate a {\em
+one-sided} proof, providing only the reverse-implication portion of the proof via
+input clause deletion.  This can provide useful information---any
+assignment that is a model for the compiled representation
+must also be a model for the input formula---even when generating a
+full validation is impractical.
+
+Our ring evaluator can perform both standard and weighted model
+counting.  It performs arithmetic over a subset of the rational
+numbers we term $\drational$, consisting of numbers of the form
+$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
+with $a$ implemented to have arbitrary range.
+Allowing scaling by powers of 2 enables the density computation
+and rescaling required for standard model counting.  Allowing scaling
+by powers of both 2 and 5 enables exact decimal arithmetic, handling
+the weights used in the weighted model counting competitions.
+
+We provide only a summary of our experiments here.  A more complete
+description is provided in Appendix~\ref{app:experiments}\@. For our
+evaluation, we used the public benchmark problems from the 2022
+standard and weighted model competitions.  We found that there were
+180 unique CNF files among these, ranging in size from 250 to
+2,753,207 clauses.  We ran our programs on a processor with 64~GB of
+memory and having an attached high-speed, solid-state
+disk---capabilities that should be included in the cluster nodes for
+future competitions.
+
+With a runtime limit of 4000 seconds (the model counting competition
+has a limit of 3600 seconds), \dfour{} completed for 124 of the
+benchmark problems.  Our proof generator was able to convert all of
+these into POGs, with their declarations ranging from 304 to
+2,761,457,765 (median 848,784) defining clauses.
+%(The
+%maximum count overflowed the overflowed the 32-bit signed integers we
+%used to represent clause identifiers in both the proof generator and
+%checker.)
+We then ran our proof generator with a time limit of 10,000 seconds.
+It was able to generate full proofs for 108 of the problems and
+one-sided proofs for an additional 9 of them, leaving just 7 with no
+verification.  The checker successfully verified all of the generated
+proofs.  The longest runtime for the combination of proof generator
+and checker for a full proof was 13,145 seconds.
+It is worth noting that, to date, we have not found
+any errors in the d-DNNF graph generated by \dfour{}.
+
+Appendix~\ref{app:experiments} also presents experimental results
+comparing the monolithic and structural methods for generating forward
+implication proofs, the effects of our two optimizations of structural
+proofs, and a comparison with the MICE proof framework.
+
+\section{Conclusions}
+\label{sect:future}
+
+Our experiments demonstrate that our toolchain could handle a
+substantial portion of the solutions generated in a knowledge compiler
+competition.  Further engineering and optimization of our proof
+generator and checker could improve their performance and capacity,
+and we could increase the time limit for these programs.
+
+%% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
+%% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
+%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{i=1,k}
+%% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
+%% such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
+%% \cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
+%% such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
+%% \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
+%% of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
+%% for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
+%% the sum of $t_1, \ldots, t_k$.
+
+\bibliography{references}
+
+\newpage
+
+\appendix
+
+\section{CRAT Example}
+\label{app:crat:example}
 
@@
 
+\subsection{Node Declarations}
+
 We step through a detailed example to provide a better understanding of the CRAT proof framework.
@@
 
+\subsection{Forward Implication Proof}
+
 Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
@@
 hints for clause 36.
+
+\subsection{Reverse Implication Proof}
 
@@
 
-Observe that this proof must ``visit'' nodes
-$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
-(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).  In
-general, a naive proof must expand the POG into a tree.
-We address this shortcoming by introducing {\em lemmas}, as is described in Section~\ref{sect:lemma}
-
-\section{Generating CRAT from d-DNNF}
-\label{section:generating:crat}
-
-%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
-%%file from an input formula in conjunctive normal form.  Our tool chain
-%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
-%%but it would also work for other tools that use a top-down search
-%%strategy to generate a d-DNNF representation.
-Converting a d-DNNF
-graph into a POG is straightforward---each node in a d-DNNF graph can
-be directly represented as a POG node.  Rather than representing every
-d-DNNF node explicitly, our program performs simplifications to
-eliminate Boolean constants.
-Except for the degenerate cases,
-%where the formula is unsatisfiable or a tautology,
-we can therefore assume
-that the POG does not contain any constant nodes.
-In addition, negation is only
-applied to variables, and so we can merge the negation operations into the terminal nodes,
-viewing the POG as consisting
-of {\em literal} nodes corresponding to input variables and their negations, along with
-{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.
-
-\subsection{Forward implication proof}
-
-For input formula $\inputformula$ and its translation into a POG $P$
-with root literal $r$, the most challenging part of the proof is to
-show that $\modelset(\inputformula) \subseteq \modelset(P)$, i.e.,
-that any total assignment $\alpha$ that is a model of $\inputformula$
-will yield $\assign(r) = 1$, for root literal $r$.  This part of the
-proof consists of a series of clause assertions leading to one adding
-$r$ as a unit clause.  We have devised two methods of generating this
-proof.  The {\em monolithic} approach makes just one call to a
-proof-generating SAT solver and has it determine the relationship
-between the two representations.  The {\em structural} approach
-recursively traverses the POG, generating proof obligations at each
-node encountered.  It too may require calls to a proof-generating SAT
-solver.
-
-As notation,
-let $\psi$ be a subset of the clauses in $\inputformula$.
-For partial assignment
-$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
-obtained from $\psi$ by: 1) eliminating any
-clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
-2) for the remaining clauses eliminating those literals $\lit$ for
-which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
-In doing these simplifications, we also track the {\em provenance}
-of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
-More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov(C, \psi)$ denote
-those clauses $C' \in \psi$, such that
-$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
-We then extend the definition of $\prov$ to any simplified formula
-$\gamma$ as $\prov(\gamma, \psi) = \bigcup_{C \in \gamma} \prov(C, \psi)$.
-
-The monolithic approach
-takes advantage of the clausal representations of
-the input formula $\inputformula$ and the POG formula $\pogformula$, consisting of the defining clauses of its nodes plus a unit clause for root literal $r$.
-Furthermore, it can
-negate the POG formula by simply negating its root literal.  More precisely $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{\nodeu}}{\{\obar{r}\}}$
-Forward implication
-will hold
-when $\inputformula \imply \pogformula$.  Equivalently,
-the formula $\inputformula \land \neg \pogformula$
-should be unsatisfiable, where the
-conjunction can be expressed as the union
-of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
-For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
-Our experimental results show
-that this approach can be very effective and generates short proofs
-for smaller problems, but it does not scale well enough for routine
-use.
-
-We describe the structural approach to proof generation as a recursive procedure
-$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
-node $\nodeu$, partial assignment
-$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
-The procedure adds a number of clauses to the proof, culminating with
-the addition of the {\em target} clause:
-$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
-that $\left(\bigwedge_{\lit \in \passign} \lit\right) \imply u$, i.e.,
-that any total
-assignment $\assign$ such that $\passign \subseteq \assign$
-will assign $\assign(u) = 1$.
-The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
-The result will therefore be to add unit clause $r$ to the proof.
-Here we present a correct, but somewhat inefficient formulation of
-$\validate$.  We then refine it with some optimizations.
-
-The recursive call $\validate(u, \passign, \psi)$ assumes that we have
-traversed a path from the root node down to node $\nodeu$, with the
-literals encountered in the product nodes forming the partial
-assignment $\passign$.  The set of clauses $\psi$ can be a proper
-subset of the input clauses $\inputformula$ when a product node has caused
-a splitting into clauses containing disjoint variables.
-The subgraph with root node $\nodeu$ should be a POG representation of the formula
-$\simplify{\psi}{\passign}$.  Having the
-routine add its target clause indicates that any total assignment
-$\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.
-
-The process for generating such a proof depends on the form of node $\nodeu$:
-\begin{enumerate}
-\item If $\nodeu$ is a literal $\lit'$, then the formula
-  $\simplify{\psi}{\passign}$ must consist of the single unit clause
-  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
-  Any of these can
-  serve as the target clause.
-\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
-  then, since the node originated from a d-DNNF graph, there must be
-  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
-  product node containing a literal node for $x$ as a child.  In either case, we
-  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
-  This will cause the addition of the target clause
-  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
-  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
-  causing the addition of the target clause
-  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-  These recursive results can be combined with the second and third defining clauses for $\nodeu$
-(see Table~\ref{tab:defining}(B))
-  to generate the target clause for $\nodeu$.
-\item If $\nodeu$ is a product node, then we can divide its children
-  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
-  \begin{enumerate}
-    \item For each literal
-  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
-  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
-  cases, this can be done by simple Boolean constraint propagation (BCP).
-  In other cases, we must prove that the formula
-  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
-  do so by writing the formula to a file, invoking a proof-generating
-  SAT solver, and then converting the generated unsatisfiability proof
-  into a sequence of clause additions in the CRAT file.
-\item For a single nonterminal child ($k = 1$), we recursively call
-  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$.
-\item For multiple nonterminal children ($k > 1$),
-  it must be the case that the clauses in
-  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
-  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
-  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
-  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
-  literals.
-  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
-  We recursively call
-  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \psi_i\right)$.
-\end{enumerate}
-  We then combine as hints the results from the BCP and SAT calls for
-  the literals, the recursively computed target clauses, and all but
-  the first defining clause for node $\nodeu$
-(see Table~\ref{tab:defining}(A))
-to generate the target clause for node $\nodeu$.
-\end{enumerate}
-Observe that most of these steps involve a polynomial number of
-operations per recursive call.  The only one that could have
-exponential complexity is when a SAT solver must be invoked to validate a
-literal.
-
-\subsection{Reverse implication proof}
-
-Completing the equivalence proof of input formula $\inputformula$ and its POG
-representation with root literal $r$ requires showing that
-$\modelset(P) \subseteq \modelset(\inputformula)$.  This is done in the
-CRAT framework by first deleting all asserted clauses, except for the
-final unit clause for root literal $r$, and then deleting all of the
-input clauses.  Suppose at some point in this process that we want to
-delete clause $C$ such that the remaining set of clauses will form the
-CNF formula $\phi$.  Deleting $C$ requires proving that
-$\phi \imply C$.
-
-The asserted clauses can be deleted in reverse order, using the same
-hints that were used in their original assertions.  By reversing the
-order, those clauses that were used as hints when a clause was
-added will still remain when it is deleted.
-
-Each input clause deletion can be done as a single RUP step.  The
-proof generator builds a sequence of hints from among the defining
-clauses of the POG nodes via a single, bottom-up pass through the
-graph.  The RUP deletion proof for input clause $C$ effectively proves that any
-total assignment $\assign$ that does not satisfy $C$ will yield
-$\assign(r) = 0$.  It starts with the set of literals
-$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
-assignment $\assign$ to falsify clause $C$.
-It then
-adds literals via unit propagation until a
-conflict arises.    Unit literal $r$ gets
-added right away, setting up a potential conflict.
-
-Working upward through the graph, node $\nodeu$ is {\em marked} when
-the collected set of literals force $u$ to evaluate to $0$.  When marking $\nodeu$, the
-program adds $\obar{u}$ to the RUP literals and adds the appropriate
-defining clause to the list of hints.  A literal node for
-$\lit$ will be marked if $\lit \in C$, with no hint required.  If
-product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
-$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
-added to the hints.  Marking sum node $\nodeu$ requires that its two children are marked.
-The first defining
-clause for this node (see Table~\ref{tab:defining}(B)) will then be added as a hint.  At the very end, the program
-(assuming the reverse implication holds) will attempt to mark root
-node $\noder$, which would require $\assign(r) = 0$, yielding a
-conflict.
-
-It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
-each clause deletion requires a single RUP step having a set of hints that is
-bounded by the number of POG nodes.
-
-\section{Optimizations}
-
-The performance of forward implication, both in its execution time and
-the size of the proof generated can be improved by two optimizations
-described here.  A key feature is that they do not require any changes
-to the proof framework---they build on the power of extended
-resolution to create new logical structures within the proof.  They
-involve declaring new product nodes to encode both conjunctions and
-disjunctions (via DeMorgan's Law) of literals.  These nodes are not
-part of the POG representation of the formula; they serve only to enable the forward implication proof.
+\section{Optimizations for Forward Implication Proofs}
+\label{app:optimizations}
 
@@
 \subsection{Lemmas}
-\label{sect:lemma}
+\label{app:lemma}
 
@@
 
-\input{lean}
-
-\section{Experimental Evaluation}
-\label{sect:experimental}
-
-We implemented the three programs illustrated in
-Figure~\ref{fig:chain} to supplement the knowledge compiler \dfour{}.
-The proof generator and proof checker were written in C/C++, while the
-ring evaluator was written in Python.\footnote{The programs are
-available at \url{https://github.com/rebryant/crat}.}  The proof
-generator uses \cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert proof steps
-back into hinted CRAT clause additions, the generator can use either
-its own RUP proof generator, or it can invoke \dtrim{}~\cite{RAT}.
-The latter yields shorter proofs and scales well to large proofs, but
-each invocation has a high startup cost.  We therefore set a threshold
-of 1000 clauses in the SAT problem for using \dtrim{}.
-
-The proof generator can optionally be instructed to generate a {\em
-  one-sided} proof, providing only reverse-implication portion via
-input clause deletion.  This can provide useful information---any
-assignment that is a model for the compiled representation
-must also be a model for the input formula---even when generation of a
-full validation is impractical.
-
-Our ring evaluator can perform both standard and weighted model
-counting.  It performs arithmetic over a subset of the rational
-numbers we term $\drational$, consisting of numbers of the form
-$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
-with $a$ implemented to have arbitrary range.
-Allowing scaling by powers of 2 enables the density computation
-and rescaling required for standard model counting.  Allowing scaling
-by powers of both 2 and 5 enables exact decimal arithmetic, handling
-the weights used in the weighted model counting competitions.
-
-\section{Future Work}
-\label{sect:future}
-
-%% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
-%% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
-%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{i=1,k}
-%% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
-%% such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
-%% \cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
-%% such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
-%% \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
-%% of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
-%% for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
-%% the sum of $t_1, \ldots, t_k$.
-
-\bibliography{references}
-
-\appendix
-
-\section{Experimental Methodology}
+\subsection{Lemma Example}
+Observe that the forward implication proof shown in Figure~\ref{fig:eg4:all}  must ``visit'' nodes
+$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
+(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).
 
 \section{Detailed Experimental Results}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.80]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=3000000000,
-      legend style={at={(0.30,0.98)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=1.8,
-                              xmode=log, xmin=0.01,xmax=4000,
-                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$,$10^{10}$},
-                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \input{data-formatted/time+defining-success}
-    \input{data-formatted/time+defining-onesided-only}
-    \input{data-formatted/time+defining-failure}
-    \legend{
-      \scriptsize \textsf{Full validation},
-      \scriptsize \textsf{One-sided validation},
-      \scriptsize \textsf{Resource limit exceeded},
-    }
- \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Number of defining clauses in generated POG as function of D4 runtime}
-\label{fig:generate:d4:log}
-\end{figure}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.80]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=20000,
-      legend style={at={(0.30,0.98)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=2.4,
-                              xmode=log,xmin=0.01,xmax=4000,
-                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
-                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
-                              xlabel={D4 runtime (seconds)}, ylabel={CRAT generation and checking runtime (seconds)},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \input{data-formatted/time-d4-verify}
-    \input{data-formatted/time-d4-verify-onesided-only}
-    \legend{
-      \scriptsize \textsf{Full validation},
-      \scriptsize \textsf{One-sided validation},
-    }
-    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
-    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
-    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
-    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
-    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
-    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Combined runtime for CRAT proof generation and checking  as function of D4 runtime}
-\label{fig:d4:crat}
-\end{figure}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.80]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000,
-                              x post scale=1.8, y post scale=2.15,
-                              xmode=log,xmin=100,xmax=1000000000,
-                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
-    \input{data-formatted/defining+total}
-    \node[left] at (axis cs: 900,1200) {$1\times$};
-    \node[left] at (axis cs: 900,12000) {$10\times$};
-    \node[left] at (axis cs: 900,120000) {$100\times$};
-    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
-    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Total number of clauses in CRAT file as function of number of defining clauses}
-\label{fig:defining:total}
-\end{figure}
-
-\begin{figure}
-\begin{minipage}{0.49\textwidth}
-\centering{%
-\begin{tikzpicture}[scale = 0.80]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8,
-      legend style={at={(0.50,0.98)}},
-      legend cell align={left},
-                              x post scale=1.1, y post scale=1.8,
-                              xmode=log,xmin=1000,xmax=1e7,
-                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
-                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
-%                              xlabel={Optimized Proof Clauses},
-%                              ylabel={Proof Clauses},
-            ]
-    \input{data-formatted/clauses-lemma-nolemma}
-    \input{data-formatted/clauses-nosplit-split}
-    \legend{
-      \scriptsize \textsf{No lemmas},
-      \scriptsize \textsf{No literal grouping},
-    }
-    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
-
-    \node[left] at (axis cs: 5000,7000) {$1\times$};
-    \node[left] at (axis cs: 5000,70000) {$10\times$};
-    \node[left] at (axis cs: 5000,700000) {$100\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-\centering{%
-\begin{tikzpicture}[scale = 0.80]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=1000, ymax=3e8,
-      legend style={at={(0.50,0.98)}},
-      legend cell align={left},
-                              x post scale=1.1, y post scale=1.8,
-                              xmode=log,xmin=1000,xmax=1e7,
-                              xtick={1000, 10000, 100000, 1000000, 10000000}, xticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$},
-                              ytick={1000, 10000, 100000, 1000000, 10000000, 100000000, 1e9}, yticklabels={$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$, $10^9$},
-%                              xlabel={Optimized Proof Clauses},y
-%                              ylabel={Proof Clauses},
-            ]
-    \input{data-formatted/clauses-optimized-unoptimized}
-    \legend{
-      \scriptsize \textsf{Neither optimization},
-    }
-    \addplot[mark=none, color=lightblue] coordinates{(1000,1000) (1e8,1e8)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,10000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(1000,100000) (1e7,1e9)};
-
-    \node[left] at (axis cs: 5000,7000) {$1\times$};
-    \node[left] at (axis cs: 5000,70000) {$10\times$};
-    \node[left] at (axis cs: 5000,700000) {$100\times$};
-  \end{axis}
-\end{tikzpicture}
-} % centering
-\end{minipage}
-\caption{Proof clauses when one (left) or both (right) optimizations are disabled, versus when both optimizations are enabled}
-\label{fig:optimized:lessoptimized}
-\end{figure}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.80]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=0.01, ymax=1500,
-      legend style={at={(0.22,0.98)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=2.2,
-                              xmode=log,xmin=0.01,xmax=1500,
-                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100,{1,000}},
-                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
-                              xlabel={CRAT total time (seconds)}, ylabel={MICE total time (seconds)},
-            ]
-    \input{data-formatted/time-mice-compare}
-    \input{data-formatted/time-mice-error}
-    \legend{
-      \scriptsize \textsf{Correct or timeout},
-      \scriptsize \textsf{MICE error},
-    }
-    \input{data-formatted/time-mice-timeout}
-%    \addplot[mark=none, color=black] coordinates{(0.01,1500) (1500,1500)};
-%    \addplot[mark=none, color=black] coordinates{(1500,0.01) (1500,1500)};
-    \addplot[mark=none, color=lightblue] coordinates{(10.0,0.01) (10000.0,10.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(1.0,0.01) (10000.0,100.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (10000,100000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (1000,100000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (1000,1000000)};
-    \node[left] at (axis cs: 50,0.06) {$0{.}001\times$};
-    \node[left] at (axis cs: 5.0,0.06) {$0{.}01\times$};
-    \node[left] at (axis cs: 0.5,0.06) {$0{.}1\times$};
-    \node[left] at (axis cs: 0.05,0.06) {$1\times$};
-    \node[left] at (axis cs: 0.05,0.6) {$10\times$};
-    \node[left] at (axis cs: 0.05,6.0) {$100\times$};
-    \node[left] at (axis cs: 0.05,60.0) {$1000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Running Time for MICE versus CRAT proof chains}
-\label{fig:mice}
-\end{figure}
+\label{app:experiments}
+
+\input{figure-runtime-d4-verify}
+
+\input{figure-clauses-d4-verify}
+
+\input{figure-runtime-monolithic}
+
+\input{figure-clauses-optimizations}
+
+\input{figure-mice-crat}
 

### 2023-03-16 doc/mc-data.tex
Filled out experimental results

@@
 
+\input{figure-clauses-monolithic}
+
 Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 82.  All completed in less than 1000 seconds with structured.
@@
 \begin{itemize}
-\item Structured complete 82.
-\item Monolithic completed  69, timed out on 13.
-\item Of the 69 completed, 23 were faster with monolithic, 36 were slower (min ratio = 0.015, max ratio = 9.66, hmean = 0.40)
-\item Of the 69 completed, 59 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
+\item Structured complete 83.
+\item Monolithic completed  70,timed out on 13.
+\item Of the 70 completed, 23 were faster with monolithic, 37 were slower (min ratio = 0.015, max ratio = 9.66, hmean = 0.40)
+\item Of the 70 completed, 60 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
 \end{itemize}

### 2023-03-16 doc/mc-paper.tex
Filled out experimental results

@@
 It is worth noting that, to date, we have not found
-any errors in the d-DNNF graph generated by \dfour{}.
+any errors in the d-DNNF graphs generated by \dfour{}.
 
@@
 \section{Optimizations for Forward Implication Proofs}
+
+The two optimizations we have implemented exploit the power of our
+general resolution framework to define new structures within the
+proof. In particular, they create new product nodes that are not part
+of the POG representation.
+
 \label{app:optimizations}
@@
 A single recursive step of $\validate$ can encounter product nodes
-having many---tens or even hundreds---of literals as children.  The earlier formulation of $\validate$
-considers each literal $\lit \in L$ separately,
-calling a SAT solver
-on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ when $\lit$ cannot be validated with BCP\@.
-
-Literal grouping allows all literals to be validated with a single call to the SAT solver.
+having many literals as children.  The naive formulation of $\validate$
+considers each literal $\lit \in L$ separately.
+Literal grouping allows all literals to be validated with a single call to a SAT solver.
 It collects those literals
@@
 
+The primary purpose of our experiments is to assess the ability of our
+implementation of the toolchain illustrated in Figure~\ref{fig:chain}
+to check the results from a possible knowledge compiler competition,
+along the lines of the annual competitions in SAT solvers, model
+counters, and other automated reasoning tools.  As secondary goals, we
+seek insights into how possible algorithmic choices affect the
+performance of the toolchain and how our toolchain compares to the
+MICE framework for verifying model counters~\cite{fichte:sat:2022}.
+
+\subsection{Experimental Setup}
+
+As a test machine, we used a 2021~MacBook~Pro laptop having a 3.2~Ghz
+Apple M1 processor and with 64~GB of physical memory.  We also used an
+external 3~TB solid-state disk drive with an advertised read/write
+speed of one gigabit per second.  Despite being a laptop, this
+configuration is somewhat more powerful than the nodes typically
+found in server clusters.  The fast access to storage is especially
+important due to the very large files generated and processed by the
+tools.  Our tools generated individual files with over 160 gigabytes
+of data.
+
+For benchmarks, we downloaded 100 files each from the 2022 standard and weighted model counting competitions:
+\begin{center}
+       \url{https://mccompetition.org/2022/mc_description.html}
+\end{center}
+We found that 20 of these were duplicates across the two tracks,
+yielding a total of 180 unique benchmark problems, ranging in size
+from 250 to 2,753,207 clauses.
+
+Our default configuration for the proof generator used the structural
+method for the forward implication proof, with the two optimizations
+of literal grouping and lemmas.
+Running with a time limit of 4000 seconds\footnote{All measured times
+in this document are actual elapsed times, not CPU times.}, \dfour{}
+was able to generated d-DNNF representations for 124 of these, and it timed out on the
+rest.  Our proof generator was able to convert all of the generated
+d-DNNF graphs into POGs and determine how many defining clauses they
+would generate.  A POG operation with $k$ arguments requires $k+1$
+defining clauses, and so the total number of defining clauses is equal
+to the sum of the number of nonterminal nodes plus the sum of number
+of children for each of these nodes.  The number of defining clauses
+ranged from 304 to 2,761,457,765 with a median of 848,784.
+
+\subsection{Performance of the Proof Generator and Checker}
+
 \input{figure-runtime-d4-verify}
 
+For each of the 124 d-DNNF graphs generated by \dfour{}, we attempted
+to use our proof generator and unverified checker, limiting proof
+generation to 10,000 seconds.  We also attempted to generate and check
+one-sided proofs for 124 graphs, again with a time limit of 10,000
+seconds for proof generation.
+
+Figure~\ref{fig:d4:crat} summarizes our results in terms of the time
+required by \dfour{} (X axis) versus the sum of the times for the
+proof generator and checker (Y axis).
+We were able to complete a full validation for 108 of the 124
+benchmarks, with times ranging from fractions of a second to 13,144.
+seconds, with a median of 71.6 seconds.
+
+Relative to the runtime for \dfour{}, two problems ran faster
+with the generator and checker.  These were for problems having small numbers of
+models (relative to the number of variables), and so most of the time
+spent by both \dfour{} and our proof generator was in running a CDCL
+solver to search the very sparse solution space.  Our solver,
+\cadical{} generally outperforms the miniSAT-based CDCL solver used by
+\dfour{}.  At the other extreme, one problem that required only 19
+seconds for \dfour{} required 8657 seconds to generate a proof and 45
+seconds to check it.  It is unclear why the proof generation required
+so much time.  Overall the ratios between the combined generation plus
+checking times versus the time for \dfour{} had a harmonic mean of
+5.5.
+
+Of the 16 benchmarks that could not be fully validated, one had so
+many defining clauses that it overflowed the 32-bit signed integers
+our programs used for clause identifiers.  Another exited due to the
+virtual memory limit imposed by the operating system, and the
+other 14 timed out.  Of the 15 that did not overflow the clause
+counter, we were able to complete a one-sided verification for 9, but
+the other 6 timed out.
+
 \input{figure-clauses-d4-verify}
 
+Figure~\ref{fig:defining:total} compares the number of defining
+clauses (X axis) with the total number of clauses (Y axis) for the 108
+problems that were fully verified.  These totals include the defining
+clauses for the POG, the additional defining clauses introduced to
+support literal grouping and lemmas, and the clauses added by RUP
+steps in the forward implication proof.  These totals ranged from 554
+to 770,382,773, with a median of 1,719,245.  The ratios of the total
+clauses versus definining clauses ranged from 1.54 to 6073, with a
+harmonic mean of 3.13.  The high numbers were for problems having very
+few models, and so the proof clauses must encode large proofs of
+unsatisfiability.
+
+\subsection{Monolithic Forward Implication Proofs}
+
 \input{figure-runtime-monolithic}
 
+\input{figure-clauses-monolithic}
+
+To assess the relative merits of the two approaches to forward
+implication proof generation, we ran the proof generator in monolithic
+mode on the 92 problems for which the combination of structural proof generation
+and checking required at most 3000 seconds and then kept only those
+measurements for which at least one of the two approaches had combined
+times (generation plus checking) of at most 1000 seconds.  There were
+a total of 83 such problems.  Figures~\ref{fig:monolithic:time} and
+\ref{fig:monolithic:clauses} show comparisons between the two
+approaches in terms of time and total clauses.
+
+Examining Figure~\ref{fig:monolithic:time}, we can see that the
+monolithic approach ran faster than the structural approach for 23 of
+the problems, including a case where the generator and checker
+required only one second, versus 71 seconds for the structural
+approach.  On the other hand, the monolithic approach timed out for 13
+of the cases and was slower for the other 37.  In general, we can see
+the monolithic approach faring worse on larger problems.
+
+Figure~\ref{fig:monolithic:clauses} shows the total numbers of clauses
+for the two approaches for the 70 cases where both approaches completed within 1000 seconds.
+As can be seen, the monolithic approach consistently produces
+shorter proofs, perhaps due to the benefits of the proof trimming by
+\dtrim{}.
+
+These experiments suggest that a hybrid of the two approaches might
+yield the best results in terms of consistency, runtime, and proof
+size.  It would begin the recursive descent of $\validate$ according
+to a structural approach, but monolithically generate forward
+implication proofs once it encounters sufficiently small subgraphs.
+
+\subsection{Impact of Optimizations}
+
 \input{figure-clauses-optimizations}
 
-\input{figure-mice-crat}
+To assess the impact of the two optimizations: literal grouping and
+lemmas, we ran the proof generator on the 80 problems for which the
+combination of the two optimizations yielded proofs totaling at most
+100~million clauses.  Compared to the problems used to assess monolithic
+proof generation, this set included one that had a short proof but
+long runtime and excluded four that had short runtimes but long
+proofs.  We then ran the proof generator on all three combinations of
+the optimizations being partially or totally disabled.
+Figure~\ref{fig:optimized:lessoptimized} shows how disabling these optimizations affected the
+total number of clauses in the proofs, where the X axis indicates the
+total number with both optimizations enabled, while the Y axis
+indicates the number with one (left) or both (right) optimizations
+enabled.
+
+As the figures illustrated, each optimization on its own can shorten
+the proofs, sometimes substantially, and the two combine to have even
+greater effect.  Of the 80 benchmark problems evaluated, and comparing
+with both optimizations disabled, 47 had the number of proof clauses
+reduced by a factor of 2 by using lemmas, 14 by using literal merging,
+and 60 by enabling both optimizations.  In the extreme cases, a lack
+of lemmas caused one proof to grow by a factor of 52.5, while a lack
+of literal grouping causes another proof to grow by a factor of 39.6.
+
+The time performance of these optimizations is less dramatic, but
+still significant.  Lemmas sometimes made proof generation slower, but
+in one case ran $38\times$ faster.  Similarly, literal grouping could
+take longer, but in one case ran $3.9\times$ faster.
+
+Overall, it is clear that these two optimizations are worthwhile, and
+even critical to success for some benchmarks.
+
+\subsection{Comparison with MICE Framework}
+
+To compare the performance of our toolchain versus that of
+MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
+generation, proof checking, and ring evaluation in the CRAT toolchain compare to the time to
+run their proof generator \progname{nnf2trace} and checker
+\progname{sharptrace}.  We add the time for ring evaluation on our
+side, since their checker provides a count, although only an
+unweighted one.  We ran their tools on the 92 benchmark problems for
+which our toolchain requires at most 3000 seconds and retained
+the 84 data points for which at least one of the toolchains completes in under 1000 seconds.
+
+We encountered a significant number of errors running their tools.
+These seem mostly to be easily fixable, and we are currently
+supporting their debugging efforts.  At present, we can summarize the results as:
+\begin{itemize}
+\item 84 total data points
+\item 67 were completed by both programs in under 1000 seconds
+\item Of those 21 were faster with MICE, 46 with CRAT
+\item 7 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
+\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
+\item 9 completed with CRAT but generated errors for MICE.
+\end{itemize}
+
+\input{figure-mice-crat} The detailed results are shown in
+Figure~\ref{fig:mice}.  As can be seen, the comparative runtimes are
+highly variable, reflecting the fact that the two toolchains differ
+fundamentally in their objectives and their approaches.
+
+One shortcoming of \progname{nnf2trace} can be seen in
+the near-vertical series of points in the upper-lefthand corner.
+These correlate with a similar set of points along the left in
+Figure~\ref{fig:optimized:lessoptimized}.  Like the naive
+implementation of $\validate$, their program recursively traverses the
+graph representation of a formula, effectively expanding it into a
+tree.  They lack an optimization analogous to lemmas.
+
+We have not tried the MICE toolchain on larger benchmarks, but we
+suspect it would encounter significant scaling limitations.
 

### 2023-03-17 doc/mc-paper.tex
Added lemma example and polished

@@
 model counters have developed that scale to very large formulas, as
-witnessed by their progress in the annual model counting competitions.
+witnessed by the progress in recent model counting competitions.
 
@@
 are placed on the structure of the graph such that a count of the
-models can be computed by a single bottom-up traversal of the graph.
+models can be computed by a single bottom-up traversal.
 Kimmig, et al., present a very general framework, known as {\em
@@
 properties of Boolean functions that can be efficiently computing from
-its d-DNNF representation.  These include standard and weighted model
+a d-DNNF representation.  These include standard and weighted model
 counting, and much more.
@@
 conjunctive normal form (CNF).  A proof consists of a sequence of
-clause addition and deletion steps, based on the extended resolution
+clause addition and deletion steps, based on an extended resolution
 proof system~\cite{Tseitin:1983}.  We present a set of conditions that
@@
 evaluator.  Importantly, the proof generator need not be
-trustworthy---its errors will also be caught by the proof checker.
+trusted---its errors will also be caught by the proof checker.
 
@@
 \cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress has also
-been made on extending these proof systems to quantified Boolean
+been made on extending these proof systems to other domains, including quantified Boolean
 formulas~\cite{bryant:cade:2021,heule:JAR2014}.
@@
 programs, and they present a series of lemmas and a theorem stating
-that their checking framework is sound and complete.  They also
+that their checking framework is sound.  They also
 present experimental results for cases where they modify existing
@@
 we compare the performance of our tool chain to theirs as part of our
-experimental evaluation in Section~\ref{sect:experimental}.  Our
+experimental evaluation in Appendix~\ref{app:experiments}.  Our
 experience with developing a formally verified proof checker has shown
@@
 
-We know of know other work addressing the more general question of
+We know of no other work addressing the question of
 certifying the output of a knowledge compiler.
@@
 We propose a general framework for summarizing properties of Boolean
-formulas inspired by the formulation of {\em algebraic model counting}.~\cite{kimmig:jal:2017}.
+formulas inspired by the formulation of {\em algebraic model counting}~\cite{kimmig:jal:2017}.
 
@@
   \end{eqnarray}
-  In this equation, sum $\sum$ is computed using addition operation $\radd$, and product $\prod$ is computed using multiplication operation $\rmul$.
+  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
 \label{def:labeling}
@@
 \textstyle
-{\Large \rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+\scalebox{1.2}{\rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
 \textstyle
-{\Large \rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
+\scalebox{1.2}{\rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
 %% V2
@@
 QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
-The following are its key properties:
+Key properties include:
 \begin{itemize}
@@
 eliminate Boolean constants.
-Except for the degenerate cases,
+Except in degenerate cases,
 %where the formula is unsatisfiable or a tautology,
@@
 
-The performance of the proof generator for forward implication, both in its execution time and
+The performance of the structural proof generator for forward implication, both in its execution time and
 the size of the proof generated, can be improved by two optimizations
@@
 resolution to create new logical structures within the proof.  They
-involve declaring new product nodes to encode both products of literals.
+involve declaring new product nodes to encode products of literals.
 These nodes are not part of the POG
@@
 $\nodeu$ having $\indegree(u) > 1$, the program can define and
-generate the proof of a lemma for $\nodeu$ when it is first reached by
+generate the proof of a {\em lemma} for $\nodeu$ when it is first reached by
 a call to $\validate$ and then apply this lemma for this and
@@
 \label{sect:experimental}
-We have implemented versions of the three programs that, in addition to the \dfour{} knowledge compiler,  form the
+We have implemented versions of the three programs that, along with the \dfour{} knowledge compiler,  form the
 toolchain illustrated in Figure~\ref{fig:chain}.  One version,
@@
 system.
-
 Our long-term goal is to have a toolchain suitable for use in a
-knowledge compilation competition, along the lines of the annual
+knowledge compiler competition, along the lines of the annual
 competitions for other automated reasoning tools.  Our experiments
@@
 assignment that is a model for the compiled representation
-must also be a model for the input formula---even when generating a
+must also be a model for the input formula---even when 
 full validation is impractical.
@@
 Our ring evaluator can perform both standard and weighted model
-counting.  It performs arithmetic over a subset of the rational
-numbers we term $\drational$, consisting of numbers of the form
+counting.  It performs arithmetic over a subset of the rationals
+consisting of numbers of the form
 $a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
@@
 
+Observe that the forward implication proof shown in Figure~\ref{fig:eg4:all}  must ``visit'' nodes
+$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
+(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).  
+Section~\ref{app:lemma:eg} shows how to use a lemma such that these nodes are each only visited once.
+
 \section{Optimizations for Forward Implication Proofs}
@@
 \subsection{Lemma Example}
-Observe that the forward implication proof shown in Figure~\ref{fig:eg4:all}  must ``visit'' nodes
-$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
-(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).
+\label{app:lemma:eg}
+
+\begin{figure}
+(A) Additional nodes\\[1.0em]
+\begin{tabular}{llll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & Explanation \\
+\midrule
+25 & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
+28 & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
+\bottomrule
+\end{tabular}
+\\[1.0em]
+(B) Implicit Clauses\\[1.2em]
+\begin{tabular}{llll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
+\midrule
+\texttt{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
+\texttt{26} & \texttt{-11 -3} & \texttt{0} & \\
+\texttt{27} & \texttt{-11 4} & \texttt{0} & \\
+\midrule
+\texttt{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
+\texttt{29} & \texttt{-12 3} & \texttt{0} & \\
+\texttt{30} & \texttt{-12 -4} & \texttt{0} & \\
+\bottomrule
+\end{tabular}
+\\[1.0em]
+(C) CRAT Assertions\\[1.0em]
+\begin{tabular}{llllll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hints} & Explanation \\
+\midrule
+\multicolumn{6}{l}{Lemma Proof} \\
+\texttt{31} & \texttt{a 5 11 12 3} & \texttt{0} & \texttt{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
+\texttt{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \texttt{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
+\texttt{33} & \texttt{a 3 7 11 12} & \texttt{0} & \texttt{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
+\texttt{34} & \texttt{a 7 11 12} & \texttt{0} & \texttt{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
+\midrule
+\multicolumn{6}{l}{Lemma Application \#1} \\
+\texttt{35} & \texttt{a -11 1} & \texttt{0} & \texttt{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
+\texttt{36} & \texttt{a -12 1} & \texttt{0} & \texttt{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
+\texttt{37} & \texttt{a 7 1} & \texttt{0} & \texttt{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
+\midrule
+\texttt{38} & \texttt{a 8 1} & \texttt{0} & \texttt{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
+\midrule
+\multicolumn{6}{l}{Lemma Application \#2} \\
+\texttt{39} & \texttt{a -11 -1} & \texttt{0} & \texttt{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
+\texttt{40} & \texttt{a -12 -1} & \texttt{0} & \texttt{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
+\texttt{41} & \texttt{a 7 -1} & \texttt{0} & \texttt{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
+\midrule
+\texttt{42} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
+\texttt{43} & \texttt{a 1 10} & \texttt{0} & \texttt{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
+\texttt{44} & \texttt{a 10} & \texttt{0} & \texttt{43 24 42} & \texttt{0} & $s_{10}$ \\
+\bottomrule
+\end{tabular}
+\caption{Example of lemma definition, proof, and application}
+\label{fig:eg4:lemmas}
+\end{figure}
+
+Figure~\ref{fig:eg4:lemmas} shows an alternate forward implication
+proof for the example of Figure~\ref{fig:eg4:all} using a lemma to
+represent the shared node $\nodes_7$.  We can see that the POG with
+this node as root encodes the Boolean formula $x_3 \leftrightarrow x_4$, having a CNF representation consisting of the clauses
+$\{x_3 ,\, \obar{x}_4\}$ and $\{\obar{x}_3 ,\, {x}_4\}$.  The product node
+declarations shown in Figure~\ref{fig:eg4:lemmas}(A) create synthetic
+clauses 25 and 28 to encode these arguments with activating literals
+$\obar{v}_{11}$ and $\obar{v}_{12}$, respectively.  Clauses 31--34
+then provide a proof of the lemma, stating that any assignment
+$\assign$ that activates these clauses will  assign $1$ to $s_7$.
+Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
+will cause the first synthetic clause to activate due to input clause
+3, and it will cause the second synthetic clause to activate due to
+input clause 4.  From this, clause 37 can use the lemma to state that
+assigning $0$ to $x_1$ will cause $s_7$ to evaluate to $1$.  Similarly,
+clauses 39 and 40 serve to activate the synthetic clauses when
+$\assign(x_1) = 1$, due to input clauses 1 and 2, and clause 41 then
+uses the lemma to state that assigning $1$ to $x_1$ will cause $s_7$ to
+evaluate to $1$.
+
+In this example, adding the lemma increases the proof length, but that
+is only because it is such a simple formula.
 
@@
 in this document are actual elapsed times, not CPU times.}, \dfour{}
-was able to generated d-DNNF representations for 124 of these, and it timed out on the
+was able to generate d-DNNF representations for 124 of these, and it timed out on the
 rest.  Our proof generator was able to convert all of the generated
@@
 would generate.  A POG operation with $k$ arguments requires $k+1$
-defining clauses, and so the total number of defining clauses is equal
-to the sum of the number of nonterminal nodes plus the sum of number
-of children for each of these nodes.  The number of defining clauses
-ranged from 304 to 2,761,457,765 with a median of 848,784.
+defining clauses, and so the total number of defining clauses in pog $P$ equals
+the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$.
+The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.
 
@@
 generation to 10,000 seconds.  We also attempted to generate and check
-one-sided proofs for 124 graphs, again with a time limit of 10,000
+one-sided proofs for the graphs, again with a time limit of 10,000
 seconds for proof generation.
@@
 spent by both \dfour{} and our proof generator was in running a CDCL
-solver to search the very sparse solution space.  Our solver,
-\cadical{} generally outperforms the miniSAT-based CDCL solver used by
+solver to search the very sparse solution space.  
+\cadical{} generally outperforms the miniSAT-based solver used by
 \dfour{}.  At the other extreme, one problem that required only 19
@@
 many defining clauses that it overflowed the 32-bit signed integers
-our programs used for clause identifiers.  Another exited due to the
+our programs use for clause identifiers.  Another exited due to the
 virtual memory limit imposed by the operating system, and the
@@
 
-Figure~\ref{fig:monolithic:clauses} shows the total numbers of clauses
+Figure~\ref{fig:monolithic:clauses} shows the total number of clauses
 for the two approaches for the 70 cases where both approaches completed within 1000 seconds.
@@
 indicates the number with one (left) or both (right) optimizations
-enabled.
+disabled.
 
@@
 Overall, it is clear that these two optimizations are worthwhile, and
-even critical to success for some benchmarks.
+sometimes critical, to success for some benchmarks.
 
@@
 MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
-generation, proof checking, and ring evaluation in the CRAT toolchain compare to the time to
+generation, proof checking, and ring evaluation in the CRAT toolchain compares to the time to
 run their proof generator \progname{nnf2trace} and checker

### 2023-03-17 doc/lean.tex
create Lean appendix and shorten section

@@
 \section{A Formally Verified CRAT Checker and POG Model Counter}
-
-[NOTE (JA): I am not worrying about length for the moment. This entire section can become an appendix,
-and then we can give an executive summary using whatever space
-we have available]
 
@@
 proof checker and both model and weighted model counters, in the Lean 4 programming language
-and proof assistant [cite]. Lean 4 implements a logical foundation in which expressions
-have a computational interpretation, and just as with other proof assistants like Isabelle and Coq,
-a function defined within the formal system can be compiled to efficient code.
-At the same time, we can state and prove claims about the function within the system, thereby
-verifying that the functions compute the intended results.
-
-Verifying our checker and the model counters in this way serves at least two purposes.
-When we run them on particular data sets, they provide high confidence
-that the particular results we obtain are correct.
-More broadly, however, our success also validates our general approach,
-showing, in particular, that the CRAT proof format is sound.
-Indeed, our formalization helped us discover a bug in our implementation
-of the informal proof checker: in the ``\emph{s}'' step described at the end of
-Subsection~\ref{subsection:syntax}, it is sound (and necessary) to allow defining
-clauses for the new operations in the proof of $\obar{\lit}_1 \lor \obar{\lit}_2$,
-but it is generally not sound to allow the input clauses or other added clauses.
-[NOTE: let's add this specification to the end of Section 6.1.]
-Our verified implementation of the checker tells us that the algorithm is sound and
-provides a reference for other implementations.
-
-In this section, we describe the implementation in Lean and the specifications we have proved.
-Lean 4's logical foundation describes a functional programming language with
-inductive data types,
-but our procedures also make use of calculations with natural numbers, integers, and rationals,
-and data structures for arrays and hashmaps.
-Lean's code extraction uses efficient but unverified versions of these,
-and so in Subsection~\ref{subsection:verification:specifics} we clarify
-what exactly we have verified and what we currently trust.
+and proof assistant [cite].
+In this section, we briefly describe the functions we have implemented in Lean and
+the specifications we have proved.
+More information is provided in the Appendix~\ref{appendix:lean}.
 
@@
 assigns a boolean value to every assignment, and later restrict attention to subsets of the
-variables. In Lean, we introduce the notation \lstinline{τ ⊨ φ} for the statement that
-\lstinline{τ} satisfies \lstinline{φ}. We write \lstinline{entails φ₁ φ₂} to say that
-\lstinline{φ₁} and \lstinline{φ₂} and \lstinline{equivalent φ₁ φ₂}.
-For the purpose of model counting, if \lstinline{s} is any finite set of variables,
-we define \lstinline{models φ s} to be the finite set of truth assignments \lstinline{τ}
-satisfying \lstinline{φ} that assign arbitrary values on \lstinline{s} and
-\lstinline{false} outside \lstinline{s}.
+variables. In our application, we instantiate \lstinline{ν} to a type \lstinline{Var} consisting
+of positive natural numbers, corresponding to the indexing of variables in the DIMACS
+format for SAT.
 
-In logic, the set of propositional formulas modulo equivalence is sometimes called the
-\emph{Lindenbaum--Tarski algebra}. We have found it convenient to define this quotient,
-which we denote \lstinline{PropTerm ν}, and lift the boolean operations and the entailment
-relation to this new type. The advantage is that equivalent formulas give
-rise to equal elements in the quotient.
-This makes easier to prove equivalences as we go, since we can calculate with the associated
-elements of the quotient and substitute one for another when the corresponding formulas are equivalent.
-
-In our application, we instantiate \lstinline{ν} to a type \lstinline{Var} consisting
-of positive natural numbers, corresponding to the indexing of variables in the DIMACS
-format for SAT. It may be helpful to think of the elements of \lstinline{PropForm Var},
-\lstinline{PropTerm Var}, and \lstinline{PropAssignment Var} as mathematical objects
-rather than computational data. Our proof checker and model counters do not compute
-with them; rather, we use them to specify what the checker and model counters are supposed to do.
-The starting point for the proof checker is, instead, a CNF formula, which \emph{is}
-presented as data.
-Specifically, a literal is a nonzero integer, with the understanding that $-\ell$ represents
-the negation of $\ell$. A clause is an array of literals, and a CNF formula is an array
-of clauses:
+We define the following representations for CNF formulas.
+A literal is represented as a nonzero integer, a clause is an array of literals,
+and a CNF formula is an array of clauses.
 \begin{lstlisting}
@@
 \end{lstlisting}
-We define operations that translate these objects to the elements of \lstinline{PropForm Var}
-that they represent. Lean provides us with helpful ``anonymous projection'' syntax that allows
-us to write, for example, \lstinline{C.toPropForm} for the expression
-\lstinline{Clause.toPropForm C}.
-Similarly, we have a representation \lstinline{C.toPropTerm}
-of \lstinline{C} as an equivalence class of propositional formulas, and we can speak of the
-value of \lstinline{C} under any truth assignment.
-Our code represents partial truth assignments as hashmaps:
-\begin{lstlisting}
-abbrev PartPropAssignment := HashMap Var Bool
-\end{lstlisting}
-We have written a procedure that reduces a clause \lstinline{C} with respect to a partial
-assignment, and another procedure that uses a hashset to test whether a clause \lstinline{C} is a
-tautology.
-We have also verified that these low-level procedures do what they are supposed to do.
+We also define functions \lstinline{ILit.toPropForm}, \lstinline{IClause.toPropForm},
+and \lstinline{ICnf.toPropForm} to relate these to propositional formulas and their
+semantics.
 
@@
 
-Our proof checker interprets the input clauses listed in a CRAT file as a CNF formula $\varphi$.
-It then processes and checks each rule, throwing an exception if a precondition is not met.
-If it terminates without raising an exception, it outputs a POG, $P$, with distinguished root
-$\bf r$.
-We have formally proved the claim that in this case the formula corresponding to node
-$\bf r$ is a partitioned formula and is
-logically equivalent to $\varphi$.
-The POG $P$ can be saved to a file [TODO: implement this?] or passed on to one of the model counting
-procedures described in the next section.
+Our proof checker starts by parsing the CNF formula and storing it internally.
+It then processes and checks each rule of a CRAT proof, throwing an exception if
+a precondition is not met. We have proved that if the checker terminates successfully,
+it succeeds in constructing a POG that is equivalent to that CNF.
 
-In our code, each element of a POG  is either a variable, a binary disjunction,
+In greater detail, each element of a POG  is either a variable, a binary disjunction,
 or an arbitrary conjunction:
@@
 
-Note that we have not yet said that the corresponding formula is partitioned.
 For each POG \lstinline{P} and literal \lstinline{l}, we define
@@
 
-With these definitions in place, we can explain the specification for our checker.
-Our parser reads a CRAT file and returns an input formula \lstinline{cnf : ICnf} and a
-CRAT proof {pf : Array CratStep}, that is, an array of single proof steps:
-\begin{lstlisting}
-inductive CratStep (α ν β : Type)
-  | /-- Add asymmetric tautology. -/
-    addAt (idx : α) (C : Array β) (upHints : Array α)
-  | /-- Delete asymmetric tautology. -/
-    delAt (idx : α) (upHints : Array α)
-  | /-- Declare product operation. -/
-    prod (idx : α) (x : ν) (ls : Array β)
-  | /-- Declare sum operation. -/
-    sum (idx : α) (x : ν) (l₁ l₂ : β) (upHints : Array α)
-  | /-- Delete operation. -/
-    delOp (x : ν)
-\end{lstlisting}
-The core checker takes those as input, as well as a specification of the final node \lstinline{r},
-and either throws an exception, indicating that the proof is not well formed,
+The core checker reads in a CRAT proof, which includes a specification of the final node \lstinline{r}.
+It either throws an exception, indicating that the proof is not well formed,
 or returns a POG \lstinline{P}.
-The specification asserts that, in the latter case, \lstinline{P.toPropForm r} is partitioned
+We have formally proved the claim that in the latter case the propositional formula \lstinline{P.toPropForm r} corresponding to node \lstinline{r} is partitioned
 and is equivalent to \lstinline{cnf.toPropForm}.
@@
 \end{lstlisting}
-Here \lstinline{R} is assumed to be a commutative ring. The curly brackets around \lstinline{R} indicate that
-this argument is meant ot be left implicit, because it can inferred from the return type of
-the weight function. Similarly, the square brackets indicate that the commutative ring structure
-is a type class argument, generally inferred from the context.
-
-More precisely, the quantity \lstinline{weightCount weight φ s} adds up the weights of the models
-of \lstinline{φ} over the set of variables \lstinline{s}. We prove that as long as \lstinline{s}
-contains the variables of \lstinline{φ}, adding extra variables does not change the sum, that is, \lstinline{vars φ ⊆ s} implies
-\begin{lstlisting}
-  weightCount weight φ s = weightCount weight φ (vars φ)
-\end{lstlisting}
+Here \lstinline{R} is assumed to be a commutative ring.
 The counting scheme of Proposition~\ref{prop:ring:eval} for partitioned formulas is expressed as follows:
@@
 
-The functions just described are not meant to be computed, but,
-rather, serve as our mathematical reference.
-We have implemented an efficient function to calculate the weighted model count
+We have implemented an efficient function \lstinline{Pog.countWeight} to calculate the weighted model count
 for an arbitrary node of a Pog, and we have proved that it computes the function above on the
@@
 
-Lean 4 does not yet have an efficient implementation of rational arithmetic, and
-we have not yet implemented a verified version of $\drational$,
-defined in Section~\ref{sect:experimental}.
-We have therefore implemented a separate procedure that carries out an integer
+We have also implemented a separate procedure that directly carries out an integer
 calculation of the number of models.
-The number of models that a propositional formula has depends on what we take the
-variables to be. For example, there is only one assignment to the variables
-$\{ x, y \}$ that makes $x \wedge y$ true, but there are two assignments to the
-variables $\{ x, y, z \}$ that make $x \wedge y$ true.
-It is not hard to show that for partitioned formulas \lstinline{PropForm ν} whose
+For partitioned formulas \lstinline{PropForm ν} whose
 variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
@@
 we have that the number of models on its variables is \lstinline{countModels φ (card (vars φ))}.
-
-\subsection{Implementation of the Proof Checker}
-
-We have specified the behavior of the proof checker in Section~\ref{subsection:proof:checker};
-in this section we sketch the implementation.
-Establishing its correctness was, by far, the most challenging part of the formalization
-process.
-For clarity, we describe the algorithm with ordinary mathematical notation rather than
-Lean snippets.
-
-We begin by reading the input CNF formula $\varphi$ and adding its clauses to a clause
-database $C$. We initialize that POG graph $G$ with the variables of $\varphi$,
-and then iteratively process each line of the CRAT proof as follows:
-\begin{itemize}
-  \item (Summarize what each step does.)
-\end{itemize}
-We establish that all the following invariants hold at each step:
-\begin{itemize}
-  \item (Wojciech has a long list of invariants)
-\end{itemize}
-At the end \ldots [Explain how the invariants at the end imply the desired conclusion.]
-
-\subsection{Verification specifics}
-\label{subsection:verification:specifics}
-
-In this subsection, we clarify what has been verified and what has been trusted.
-Recall that our first step is to parse a CRAT file to read in the initial CNF formula
-and the CRAT proof. We do not verify this step.
-If one is only interested in obtaining a model count or a verified model count,
-there is no strong need to verify the parsing of the proof,
-because if the checker succeeds in constructing an equivalent POG,
-it does not really matter whether the proof was the one the user intended.
-But it would be nice to have reassurance that the CNF formula that serves
-as input to our verified checker matches the one described in the file.
-An easy and fairly standard way of addressing this is to print out the CNF formula
-right after parsing it and then do a diff with the original input.
-This involves trusting only the correctness of the print procedure and diff.
-Similarly, if one wants to establish the correctness of the POG contained in the CRAT file,
-one can print out the POG that is constructed by the checker and compare.
-
-Lean's code extraction replaces calculations with natural numbers and integers with
-efficient but unverified arbitrary precision versions.
-Lean also uses an efficient implementation of arrays; within the
-formal system, these are defined in terms of lists, but code extraction replaces them
-with dynamic arrays and uses reference counting to allow destructive updates when it is safe
-to do so [cite counting immutable beans paper].
-Code extraction also provides efficient implementations of tail recursive functions and loops
-[cite ``do unchained'' paper].
-Finally, Lean's standard library implements hashmaps in terms of array.
-Many of the basic properties of hashmaps have been formally verified, but not all.
-In particular, we make use of a fold operation whose verification is not yet complete.
-Thus our proofs depend on the assumption that the fold operation has the
-expected properties.
-
-In sum, in addition to trusting Lean's foundation and kernel checker,
-we also have to trust that code extraction respects that foundation,
-that the implementation of arrays and hashmaps satisfy their internal descriptions,
-and that, after parsing, the computation has the correct input formula.
-All of our specifications have been completely proved and verified relative to that.
-
-\subsection{Evaluation}
-
-[TODO: Say something to establish that the procedure run in reasonable time and produce
-useful results; or maybe this goes in the next section?]

### 2023-03-17 doc/leanappendix.tex
create Lean appendix and shorten section

\section{More information about the formally verified checker}
\label{appendix:lean}

We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
proof checker and both model and weighted model counters, in the Lean 4 programming language
and proof assistant [cite]. Lean 4 implements a logical foundation in which expressions
have a computational interpretation, and just as with other proof assistants like Isabelle and Coq,
a function defined within the formal system can be compiled to efficient code.
At the same time, we can state and prove claims about the function within the system, thereby
verifying that the functions compute the intended results.

Verifying our checker and the model counters in this way serves at least two purposes.
When we run them on particular data sets, they provide high confidence
that the particular results we obtain are correct.
More broadly, however, our success also validates our general approach,
showing, in particular, that the CRAT proof format is sound.
Indeed, our formalization helped us discover a bug in our implementation
of the informal proof checker: in the ``\emph{s}'' step described at the end of
Subsection~\ref{subsection:syntax}, it is sound (and necessary) to allow defining
clauses for the new operations in the proof of $\obar{\lit}_1 \lor \obar{\lit}_2$,
but it is generally not sound to allow the input clauses or other added clauses.
[NOTE: let's add this specification to the end of Section 6.1.]
Our verified implementation of the checker tells us that the algorithm is sound and
provides a reference for other implementations.

In this section, we describe the implementation in Lean and the specifications we have proved.
Lean 4's logical foundation describes a functional programming language with
inductive data types,
but our procedures also make use of calculations with natural numbers, integers, and rationals,
and data structures for arrays and hashmaps.
Lean's code extraction uses efficient but unverified versions of these,
and so in Subsection~\ref{subsection:verification:specifics} we clarify
what exactly we have verified and what we currently trust.

\subsection{Logical objects and operations}

Our specifications are built on a generic library for the syntax and semantics of
propositional logic. The data type of propositional formulas with variables indexed by
a datatype \lstinline{ν} is defined inductively as follows:
\begin{lstlisting}
inductive PropForm (ν : Type)
  | var (x : ν)
  | tr
  | fls
  | neg (φ : PropForm ν)
  | conj (φ₁ φ₂ : PropForm ν)
  | disj (φ₁ φ₂ : PropForm ν)
  | impl (φ₁ φ₂ : PropForm ν)
  | biImpl (φ₁ φ₂ : PropForm ν)
\end{lstlisting}
The last two constructors denote implication and bi-implication, which are not used in this
project. The usual syntactic and semantic notions are defined in the expected ways.
It is convenient to take a truth assignment \lstinline{PropAssignment ν} to a function that
assigns a boolean value to every assignment, and later restrict attention to subsets of the
variables. In Lean, we introduce the notation \lstinline{τ ⊨ φ} for the statement that
\lstinline{τ} satisfies \lstinline{φ}. We write \lstinline{entails φ₁ φ₂} to say that
\lstinline{φ₁} and \lstinline{φ₂} and \lstinline{equivalent φ₁ φ₂}.
For the purpose of model counting, if \lstinline{s} is any finite set of variables,
we define \lstinline{models φ s} to be the finite set of truth assignments \lstinline{τ}
satisfying \lstinline{φ} that assign arbitrary values on \lstinline{s} and
\lstinline{false} outside \lstinline{s}.

In logic, the set of propositional formulas modulo equivalence is sometimes called the
\emph{Lindenbaum--Tarski algebra}. We have found it convenient to define this quotient,
which we denote \lstinline{PropTerm ν}, and lift the boolean operations and the entailment
relation to this new type. The advantage is that equivalent formulas give
rise to equal elements in the quotient.
This makes easier to prove equivalences as we go, since we can calculate with the associated
elements of the quotient and substitute one for another when the corresponding formulas are equivalent.

In our application, we instantiate \lstinline{ν} to a type \lstinline{Var} consisting
of positive natural numbers, corresponding to the indexing of variables in the DIMACS
format for SAT. It may be helpful to think of the elements of \lstinline{PropForm Var},
\lstinline{PropTerm Var}, and \lstinline{PropAssignment Var} as mathematical objects
rather than computational data. Our proof checker and model counters do not compute
with them; rather, we use them to specify what the checker and model counters are supposed to do.
The starting point for the proof checker is, instead, a CNF formula, which \emph{is}
presented as data.
Specifically, a literal is a nonzero integer, with the understanding that $-\ell$ represents
the negation of $\ell$. A clause is an array of literals, and a CNF formula is an array
of clauses:
\begin{lstlisting}
def ILit := { i : Int // i ≠ 0 }
abbrev IClause := Array ILit
abbrev ICnf := Array IClause
\end{lstlisting}
We define operations that translate these objects to the elements of \lstinline{PropForm Var}
that they represent. Lean provides us with helpful ``anonymous projection'' syntax that allows
us to write, for example, \lstinline{C.toPropForm} for the expression
\lstinline{Clause.toPropForm C}.
Similarly, we have a representation \lstinline{C.toPropTerm}
of \lstinline{C} as an equivalence class of propositional formulas, and we can speak of the
value of \lstinline{C} under any truth assignment.
Our code represents partial truth assignments as hashmaps:
\begin{lstlisting}
abbrev PartPropAssignment := HashMap Var Bool
\end{lstlisting}
We have written a procedure that reduces a clause \lstinline{C} with respect to a partial
assignment, and another procedure that uses a hashset to test whether a clause \lstinline{C} is a
tautology.
We have also verified that these low-level procedures do what they are supposed to do.

\subsection{The Proof Checker}
\label{subsection:proof:checker}

Our proof checker interprets the input clauses listed in a CRAT file as a CNF formula $\varphi$.
It then processes and checks each rule, throwing an exception if a precondition is not met.
If it terminates without raising an exception, it outputs a POG, $P$, with distinguished root
$\bf r$.
We have formally proved the claim that in this case the formula corresponding to node
$\bf r$ is a partitioned formula and is
logically equivalent to $\varphi$.
The POG $P$ can be saved to a file [TODO: implement this?] or passed on to one of the model counting
procedures described in the next section.

In our code, each element of a POG  is either a variable, a binary disjunction,
or an arbitrary conjunction:
\begin{lstlisting}
inductive PogElt where
  | var  : Var → PogElt
  | disj : Var → ILit → ILit → PogElt
  | conj : Var → Array ILit → PogElt
\end{lstlisting}
In the first case, the argument \lstinline{x} in the expression
\lstinline{var x} is the index
of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
it is the definition number in the CRAT file.
The CRAT generator described in Section~\ref{section:generating:crat}
declares new variables sequentially, so we currently make the simplifying assumption
that the $i$th element of the array has the node with index $i + 1$,
but it will not be hard to generalize this.
Notice that the arguments to \lstinline{disj} and \lstinline{conj} are literals,
corresponding to positive or negative instances of variables declared earlier in the POG.
So our POG data structure is as follows:
\begin{lstlisting}
structure Pog where
  elts : Array PogElt
  wf : ∀ i : Fin elts.size, elts[i].args_decreasing
  inv : ∀ i : Fin elts.size, i = elts[i].varNum.natPred
\end{lstlisting}
The invariant \lstinline{wf} says that the graph is well founded, in the
sense that if \lstinline{elts[i]} is a disjunction
or a conjunction, then the variables occurring in the arguments have smaller indices.
Note that \lstinline{Fin elts.size} is the data type consisting of natural numbers
less than the size of the array \lstinline{elts}; our code enforces statically
that the array accesses are within bounds.

Note that we have not yet said that the corresponding formula is partitioned.
For each POG \lstinline{P} and literal \lstinline{l}, we define
\lstinline{P.toPropForm l} to be the
propositional formula that arises from interpreting \lstinline{l} as a propositional
formula, unfolding all the defined conjunctions and disjunctions. We define
what it means for such a formula to be partitioned:
\begin{lstlisting}
def partitioned : PropForm ν → Prop
  | tr         => True
  | fls        => True
  | var _      => True
  | neg φ      => φ.partitioned
  | disj φ ψ   => φ.partitioned ∧ ψ.partitioned ∧
                    ∀ τ, ¬ (φ.eval τ ∧ ψ.eval τ)
  | conj φ ψ   => φ.partitioned ∧ ψ.partitioned ∧ (φ.vars ∩ ψ.vars = ∅)
  | impl _ _   => False
  | biImpl _ _ => False
\end{lstlisting}
In other words, we rule out formulas with implication and bi-implication, and otherwise
we follow the straightforward recursive definition.

With these definitions in place, we can explain the specification for our checker.
Our parser reads a CRAT file and returns an input formula \lstinline{cnf : ICnf} and a
CRAT proof {pf : Array CratStep}, that is, an array of single proof steps:
\begin{lstlisting}
inductive CratStep (α ν β : Type)
  | /-- Add asymmetric tautology. -/
    addAt (idx : α) (C : Array β) (upHints : Array α)
  | /-- Delete asymmetric tautology. -/
    delAt (idx : α) (upHints : Array α)
  | /-- Declare product operation. -/
    prod (idx : α) (x : ν) (ls : Array β)
  | /-- Declare sum operation. -/
    sum (idx : α) (x : ν) (l₁ l₂ : β) (upHints : Array α)
  | /-- Delete operation. -/
    delOp (x : ν)
\end{lstlisting}
The core checker takes those as input, as well as a specification of the final node \lstinline{r},
and either throws an exception, indicating that the proof is not well formed,
or returns a POG \lstinline{P}.
The specification asserts that, in the latter case, \lstinline{P.toPropForm r} is partitioned
and is equivalent to \lstinline{cnf.toPropForm}.

\subsection{The Model Counters}

We have formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem,
Definition~\ref{def:ring_evaluation}, as follows:
\begin{lstlisting}
def weightCount {R : Type} [CommRing R]
    (weight : ν → R) (φ : PropForm ν) (s : Finset ν) : R :=
  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
\end{lstlisting}
Here \lstinline{R} is assumed to be a commutative ring. The curly brackets around \lstinline{R} indicate that
this argument is meant ot be left implicit, because it can inferred from the return type of
the weight function. Similarly, the square brackets indicate that the commutative ring structure
is a type class argument, generally inferred from the context.

More precisely, the quantity \lstinline{weightCount weight φ s} adds up the weights of the models
of \lstinline{φ} over the set of variables \lstinline{s}. We prove that as long as \lstinline{s}
contains the variables of \lstinline{φ}, adding extra variables does not change the sum, that is, \lstinline{vars φ ⊆ s} implies
\begin{lstlisting}
  weightCount weight φ s = weightCount weight φ (vars φ)
\end{lstlisting}
The counting scheme of Proposition~\ref{prop:ring:eval} for partitioned formulas is expressed as follows:
\begin{lstlisting}
 def countWeight (weight : ν → Rat) : PropForm ν → Rat
  | tr       => 1
  | fls      => 0
  | var x    => weight x
  | neg φ    => 1 - countWeight weight φ
  | disj φ ψ => countWeight weight φ + countWeight weight ψ
  | conj φ ψ => countWeight weight φ * countWeight weight ψ
\end{lstlisting}
We arbitrarily assign 0 to implications and bi-implications;
recall that, by definition, these do not appear in partitioned formulas.
Proposition~\ref{prop:ring:eval} is then formalized as follows:
\begin{lstlisting}
theorem countWeight_eq_weightCount (weight : ν → R) {φ : PropForm ν}
    (hdec : partitioned φ) :
  countWeight weight φ = weightCount weight φ (vars φ)
\end{lstlisting}

The functions just described are not meant to be computed, but,
rather, serve as our mathematical reference.
We have implemented an efficient function to calculate the weighted model count
for an arbitrary node of a Pog, and we have proved that it computes the function above on the
associated formula:
\begin{lstlisting}
theorem countWeight_eq_countWeight (pog : Pog) (weight : Var → Rat)
    (x : Var) :
  pog.countWeight weight x = (pog.toPropForm (.mkPos x)).countWeight weight
\end{lstlisting}
Here \lstinline{.mkPos x} denotes the positive literal for the variable \lstinline{x}.
Applying this to the output of our verified CRAT proof checker,
we obtain a proof that
our toolchain computes the correct weighted model count of the input CNF.

Lean 4 does not yet have an efficient implementation of rational arithmetic, and
we have not yet implemented a verified version of $\drational$,
defined in Section~\ref{sect:experimental}.
We have therefore implemented a separate procedure that carries out an integer
calculation of the number of models.
The number of models that a propositional formula has depends on what we take the
variables to be. For example, there is only one assignment to the variables
$\{ x, y \}$ that makes $x \wedge y$ true, but there are two assignments to the
variables $\{ x, y, z \}$ that make $x \wedge y$ true.
It is not hard to show that for partitioned formulas \lstinline{PropForm ν} whose
variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
we can count the number of models of the formula on the variables in \lstinline{s}
recursively as follows:
\begin{lstlisting}
def countModels (nVars : Nat) : PropForm ν → Nat
  | tr       => 2^nVars
  | fls      => 0
  | var _    => 2^(nVars - 1)
  | neg φ    => 2^nVars - countModels nVars φ
  | disj φ ψ => countModels nVars φ + countModels nVars ψ
  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
\end{lstlisting}
We have also implemented an efficient counting function for POGs and proved that
it computes that same quantity on the corresponding formulas.
Finally,
We have proved that the calculation above really does
return the number of models:
\begin{lstlisting}
theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
    (hvars : vars φ ⊆ s) (hpar :  partitioned φ) :
  countModels (card s) φ = card (models φ s)
\end{lstlisting}
In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
we have that the number of models on its variables is \lstinline{countModels φ (card (vars φ))}.

\subsection{Implementation of the Proof Checker}

We have specified the behavior of the proof checker in Section~\ref{subsection:proof:checker};
in this section we sketch the implementation.
Establishing its correctness was, by far, the most challenging part of the formalization
process.
For clarity, we describe the algorithm with ordinary mathematical notation rather than
Lean snippets.

We begin by reading the input CNF formula $\varphi$ and adding its clauses to a clause
database $C$. We initialize that POG graph $G$ with the variables of $\varphi$,
and then iteratively process each line of the CRAT proof as follows:
\begin{itemize}
  \item (Summarize what each step does.)
\end{itemize}
We establish that all the following invariants hold at each step:
\begin{itemize}
  \item (Wojciech has a long list of invariants)
\end{itemize}
At the end \ldots [Explain how the invariants at the end imply the desired conclusion.]

\subsection{Verification specifics}
\label{subsection:verification:specifics}

In this subsection, we clarify what has been verified and what has been trusted.
Recall that our first step is to parse a CRAT file to read in the initial CNF formula
and the CRAT proof. We do not verify this step.
If one is only interested in obtaining a model count or a verified model count,
there is no strong need to verify the parsing of the proof,
because if the checker succeeds in constructing an equivalent POG,
it does not really matter whether the proof was the one the user intended.
But it would be nice to have reassurance that the CNF formula that serves
as input to our verified checker matches the one described in the file.
An easy and fairly standard way of addressing this is to print out the CNF formula
right after parsing it and then do a diff with the original input.
This involves trusting only the correctness of the print procedure and diff.
Similarly, if one wants to establish the correctness of the POG contained in the CRAT file,
one can print out the POG that is constructed by the checker and compare.

Lean's code extraction replaces calculations with natural numbers and integers with
efficient but unverified arbitrary precision versions.
Lean also uses an efficient implementation of arrays; within the
formal system, these are defined in terms of lists, but code extraction replaces them
with dynamic arrays and uses reference counting to allow destructive updates when it is safe
to do so [cite counting immutable beans paper].
Code extraction also provides efficient implementations of tail recursive functions and loops
[cite ``do unchained'' paper].
Finally, Lean's standard library implements hashmaps in terms of array.
Many of the basic properties of hashmaps have been formally verified, but not all.
In particular, we make use of a fold operation whose verification is not yet complete.
Thus our proofs depend on the assumption that the fold operation has the
expected properties.

In sum, in addition to trusting Lean's foundation and kernel checker,
we also have to trust that code extraction respects that foundation,
that the implementation of arrays and hashmaps satisfy their internal descriptions,
and that, after parsing, the computation has the correct input formula.
All of our specifications have been completely proved and verified relative to that.

\subsection{Evaluation}

[TODO: Say something to establish that the procedure run in reasonable time and produce
useful results; or maybe this goes in the next section?]

### 2023-03-17 doc/lean.tex
tinkering

@@
 \end{lstlisting}
-We have also implemented an efficient counting function for POGs and proved that
+We have implemented an efficient counting function for POGs and proved that
 it computes that same quantity on the corresponding formulas.
 Finally,
-We have proved that the calculation above really does
+we have proved that the calculation above really does
 return the number of models:

### 2023-03-17 doc/mc-paper.tex
create Lean appendix and shorten section
Bug fixes in MICE
et al.

@@
 models can be computed by a single bottom-up traversal.
-Kimmig, et al., present a very general framework, known as {\em
+Kimmig et al.\ present a very general framework, known as {\em
   algebraic model counting}~\cite{kimmig:jal:2017}, describing
@@
 \item 84 total data points
-\item 67 were completed by both programs in under 1000 seconds
-\item Of those 21 were faster with MICE, 46 with CRAT
+\item 76 were completed by both programs in under 1000 seconds
+\item Of those 21 were faster with MICE, 55 with CRAT
 \item 7 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
 \item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
-\item 9 completed with CRAT but generated errors for MICE.
 \end{itemize}
@@
 
+\input{leanappendix}
+
 \end{document}

### 2023-03-17 doc/lean.tex
clean up and shorten Lean section

@@
 
-\subsection{Logical objects and operations}
+\subsection{The Proof Checker}
+\label{subsection:proof:checker}
 
 Our specifications are built on a generic library for the syntax and semantics of
-propositional logic. The data type of propositional formulas with variables indexed by
-a datatype \lstinline{ν} is defined inductively as follows:
-\begin{lstlisting}
-inductive PropForm (ν : Type)
-  | var (x : ν)
-  | tr
-  | fls
-  | neg (φ : PropForm ν)
-  | conj (φ₁ φ₂ : PropForm ν)
-  | disj (φ₁ φ₂ : PropForm ν)
-  | impl (φ₁ φ₂ : PropForm ν)
-  | biImpl (φ₁ φ₂ : PropForm ν)
-\end{lstlisting}
-The last two constructors denote implication and bi-implication, which are not used in this
-project. The usual syntactic and semantic notions are defined in the expected ways.
-It is convenient to take a truth assignment \lstinline{PropAssignment ν} to a function that
+propositional logic. We define the data type \lstinline{PropForm Var} of propositional
+formulas with variables \lstinline{Var}, which are positive natural numbers,
+corresponding to the indexing of variables in the DIMACS format for SAT.
+It is convenient to take a truth assignment \lstinline{PropAssignment Var} to a function that
 assigns a boolean value to every assignment, and later restrict attention to subsets of the
-variables. In our application, we instantiate \lstinline{ν} to a type \lstinline{Var} consisting
-of positive natural numbers, corresponding to the indexing of variables in the DIMACS
-format for SAT.
+variables.
 
@@
 
-\subsection{The Proof Checker}
-\label{subsection:proof:checker}
-
-Our proof checker starts by parsing the CNF formula and storing it internally.
-It then processes and checks each rule of a CRAT proof, throwing an exception if
-a precondition is not met. We have proved that if the checker terminates successfully,
-it succeeds in constructing a POG that is equivalent to that CNF.
-
-In greater detail, each element of a POG  is either a variable, a binary disjunction,
+The goal of the proof checker is to construct a POG that is equivalent to the input CNF.
+Each element of a POG  is either a variable, a binary disjunction,
 or an arbitrary conjunction:
@@
 of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
-it is the definition number in the CRAT file.
-The CRAT generator described in Section~\ref{section:generating:crat}
-declares new variables sequentially, so we currently make the simplifying assumption
-that the $i$th element of the array has the node with index $i + 1$,
-but it will not be hard to generalize this.
-Notice that the arguments to \lstinline{disj} and \lstinline{conj} are literals,
-corresponding to positive or negative instances of variables declared earlier in the POG.
-So our POG data structure is as follows:
-\begin{lstlisting}
-structure Pog where
-  elts : Array PogElt
-  wf : ∀ i : Fin elts.size, elts[i].args_decreasing
-  inv : ∀ i : Fin elts.size, i = elts[i].varNum.natPred
-\end{lstlisting}
-The invariant \lstinline{wf} says that the graph is well founded, in the
-sense that if \lstinline{elts[i]} is a disjunction
-or a conjunction, then the variables occurring in the arguments have smaller indices.
-Note that \lstinline{Fin elts.size} is the data type consisting of natural numbers
-less than the size of the array \lstinline{elts}; our code enforces statically
-that the array accesses are within bounds.
-
+it is the definition number in the CRAT file. Note that representing edges as literals allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
+A \lstinline{Pog} is then simply an array of \lstinline{PogElt}.
 For each POG \lstinline{P} and literal \lstinline{l}, we define
@@
 propositional formula that arises from interpreting \lstinline{l} as a propositional
-formula, unfolding all the defined conjunctions and disjunctions. We define
-what it means for such a formula to be partitioned:
+formula, unfolding all the defined conjunctions and disjunctions.
+Our formulation of what it means for such a formula to be partitioned, corresponding
+to the definition at the end of \cref{section:logical:foundations}, is as follows:
 \begin{lstlisting}
-def partitioned : PropForm ν → Prop
-  | tr         => True
-  | fls        => True
-  | var _      => True
+def partitioned : PropForm Var → Prop
+  | tr | fls | var _ => True
   | neg φ    => φ.partitioned
@@
   | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ (φ.vars ∩ ψ.vars = ∅)
-  | impl _ _   => False
-  | biImpl _ _ => False
 \end{lstlisting}
-In other words, we rule out formulas with implication and bi-implication, and otherwise
-we follow the straightforward recursive definition.
 
-The core checker reads in a CRAT proof, which includes a specification of the final node \lstinline{r}.
-It either throws an exception, indicating that the proof is not well formed,
-or returns a POG \lstinline{P}.
-We have formally proved the claim that in the latter case the propositional formula \lstinline{P.toPropForm r} corresponding to node \lstinline{r} is partitioned
-and is equivalent to \lstinline{cnf.toPropForm}.
+Our proof checker starts by parsing the CNF formula and storing it internally.
+It then processes and checks each rule of a CRAT proof, throwing an exception if
+the proof is not well-formed. We have proved that if the checker terminates successfully,
+it succeeds in constructing a POG that is partitioned and equivalent to that CNF. This is formalized as \lstinline{(pog.toPropForm r).partitioned ∧ inputCnf.toPropForm ≡ pog.toPropForm r}.
 
@@
 We have formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem,
-Definition~\ref{def:ring_evaluation}, as follows:
+\cref{def:ring_evaluation}, as follows:
 \begin{lstlisting}
 def weightCount {R : Type} [CommRing R]
-    (weight : ν → R) (φ : PropForm ν) (s : Finset ν) : R :=
+    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
   ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
@@
 \begin{lstlisting}
- def countWeight (weight : ν → Rat) : PropForm ν → Rat
+ def ringEval (weight : Var → R) : PropForm Var → R
   | tr       => 1
@@
   | var x    => weight x
-  | neg φ    => 1 - countWeight weight φ
-  | disj φ ψ => countWeight weight φ + countWeight weight ψ
-  | conj φ ψ => countWeight weight φ * countWeight weight ψ
+  | neg φ    => 1 - ringEval weight φ
+  | disj φ ψ => ringEval weight φ + ringEval weight ψ
+  | conj φ ψ => ringEval weight φ * ringEval weight ψ
 \end{lstlisting}
@@
 \begin{lstlisting}
-theorem countWeight_eq_weightCount (weight : ν → R) {φ : PropForm ν}
-    (hdec : partitioned φ) :
-  countWeight weight φ = weightCount weight φ (vars φ)
+theorem ringEval_eq_weightCount (weight : Var → R) {φ : PropForm Var} :
+    partitioned φ → ringEval weight φ = weightCount weight φ (vars φ)
 \end{lstlisting}
 
-We have implemented an efficient function \lstinline{Pog.countWeight} to calculate the weighted model count
+We have implemented an efficient function \lstinline{Pog.ringEval} to calculate the weighted model count
 for an arbitrary node of a Pog, and we have proved that it computes the function above on the
@@
 \begin{lstlisting}
-theorem countWeight_eq_countWeight (pog : Pog) (weight : Var → Rat)
-    (x : Var) :
-  pog.countWeight weight x = (pog.toPropForm (.mkPos x)).countWeight weight
+theorem ringEval_eq_ringEval (pog : Pog) (weight : Var → R) (x : Var) :
+  pog.ringEval weight x = (pog.toPropForm (.mkPos x)).ringEval weight
 \end{lstlisting}
@@
 calculation of the number of models.
-For partitioned formulas \lstinline{PropForm ν} whose
+For partitioned formulas \lstinline{PropForm Var} whose
 variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
@@
 \begin{lstlisting}
-def countModels (nVars : Nat) : PropForm ν → Nat
+def countModels (nVars : Nat) : PropForm Var → Nat
   | tr       => 2^nVars
@@
 \begin{lstlisting}
-theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
-    (hvars : vars φ ⊆ s) (hpar :  partitioned φ) :
-  countModels (card s) φ = card (models φ s)
+theorem countModels_eq_card_models {φ : PropForm Var} {s : Finset Var} :
+  vars φ ⊆ s → partitioned φ → countModels (card s) φ = card (models φ s)
 \end{lstlisting}

### 2023-03-17 doc/mc-paper.tex
clean up and shorten Lean section
ring labeling -> ring weight function

@@
 \section{Logical Foundations}
+\label{section:logical:foundations}
 
@@
 \label{def:ring_evaluation}
-  For commutative ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
+  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
   every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $\mulident-w(x)$.
 
-  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} is to compute:
+  For Boolean formula $\phi$, and ring weight function $w$, the {\em ring evaluation problem} is to compute:
   \begin{eqnarray}
@@
   In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
-\label{def:labeling}
 \end{definition}

### 2023-03-17 doc/lean.tex
tinkering

@@
  | neg φ    => φ.partitioned
-  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧
-                    ∀ τ, ¬ (φ.eval τ ∧ ψ.eval τ)
+ | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(φ.eval τ ∧ ψ.eval τ)
  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ (φ.vars ∩ ψ.vars = ∅)
@@
 \end{lstlisting}
-We arbitrarily assign 0 to implications and bi-implications;
-recall that, by definition, these do not appear in partitioned formulas.
 Proposition~\ref{prop:ring:eval} is then formalized as follows:

### 2023-03-17 doc/lean.tex
General updates

--- 2023-03-17 doc/lean.tex
@@
 We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
-proof checker and both model and weighted model counters, in the Lean 4 programming language
-and proof assistant [cite].
+proof checker and both model and weighted model counters, in the \lean{} programming language
+and proof assistant~\cite{demoura:cade:2021}.
 In this section, we briefly describe the functions we have implemented in Lean and
@@
 formulas with variables \lstinline{Var}, which are positive natural numbers,
-corresponding to the indexing of variables in the DIMACS format for SAT.
+corresponding to the indexing of variables in the CNF file.
 It is convenient to take a truth assignment \lstinline{PropAssignment Var} to a function that
-assigns a boolean value to every assignment, and later restrict attention to subsets of the
+assigns a Boolean value to every assignment, and later restrict attention to subsets of the
 variables.
@@
 formula, unfolding all the defined conjunctions and disjunctions.
-Our formulation of what it means for such a formula to be partitioned, corresponding
-to the definition at the end of \cref{section:logical:foundations}, is as follows:
+A partitioned-operation formula, as defined in \cref{section:logical:foundations}, is expressed as:
 \begin{lstlisting}
@@
 \subsection{The Model Counters}
+\label{subsection:counting}
 
@@
 \end{lstlisting}
-We arbitrarily assign 0 to implications and bi-implications;
-recall that, by definition, these do not appear in partitioned formulas.
+%%We arbitrarily assign 0 to implications and bi-implications;
+%%recall that, by definition, these do not appear in partitioned formulas.
 Proposition~\ref{prop:ring:eval} is then formalized as follows:

### 2023-03-17 doc/leanappendix.tex
General updates

@@
 We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
-proof checker and both model and weighted model counters, in the Lean 4 programming language
-and proof assistant [cite]. Lean 4 implements a logical foundation in which expressions
+proof checker and both model and weighted model counters, in the \lean{} programming language
+and proof assistant [cite]. \lean{} implements a logical foundation in which expressions
 have a computational interpretation, and just as with other proof assistants like Isabelle and Coq,
@@
 In this section, we describe the implementation in Lean and the specifications we have proved.
-Lean 4's logical foundation describes a functional programming language with
+\lean{}'s logical foundation describes a functional programming language with
 inductive data types,
@@
 \subsection{The Proof Checker}
-\label{subsection:proof:checker}
+\label{appendix:proof:checker}
 
@@
 
-Lean 4 does not yet have an efficient implementation of rational arithmetic, and
+\lean{} does not yet have an efficient implementation of rational arithmetic, and
 we have not yet implemented a verified version of $\drational$,
@@
 
-We have specified the behavior of the proof checker in Section~\ref{subsection:proof:checker};
+We have specified the behavior of the proof checker in Section~\ref{appendix:proof:checker};
 in this section we sketch the implementation.
@@
 
-\subsection{Evaluation}
+%% \subsection{Evaluation}
 
-[TODO: Say something to establish that the procedure run in reasonable time and produce
-useful results; or maybe this goes in the next section?]
+%% [TODO: Say something to establish that the procedure run in reasonable time and produce
+%% useful results; or maybe this goes in the next section?]

### 2023-03-17 doc/mc-paper.tex
General updates

@@
 \newcommand{\ifandonlyif}{\Leftrightarrow}
-\newcommand{\drational}{\mathbb{Q}_{2,5}}
+%\newcommand{\drational}{\mathbb{Q}_{2,5}}
+\newcommand{\drational}{\textbf{Q}_{2,5}}
 
@@
 values efficiently, such as the weighted or unweighted model
-count. Existing knowledge compilers, however,
+count. Existing knowledge compilers
 provide no guarantee that the representation they generate
-is logically equivalent to the original formula, and therefore they cannot validate the generated counts.
+correctly represents the original formula, and therefore they cannot validate the generated counts.
 
 We present \emph{Partitioned Operation Graphs} (POGs), a generalization
-of representations commonly used in knowledge compilation.
+of the representations used by existing knowledge compilers.
 We have designed a proof framework that can prove the equivalence of a
@@
 
-We have developed a program that can generate POG representations from
+We have developed a program that generates POG representations from
 the d-DNNF graphs generated by the state-of-the-art knowledge compiler
-\dfour, as well as independently checkable proofs that the output POGs
+\dfour, as well as checkable proofs that the output POGs
 are equivalent to the input CNF formulas.  We have implemented an
-efficient checker that can handle very large proofs. Our tool chain
+efficient checker that can handle large proofs. Our tool chain
 for generating and verifying POGs scales to all but the largest d-DNNF
 graphs generated by \dfour{} for formulas from a recent model counting
-competition. Moreover, we present a checker and a model counter for POGs that have
-been formally verified in the \lean{} proof assistant, thereby
-creating the first formally verified toolchain for
-weighted and unweighted
-model counting.
+competition. We also present a checker and a model counter for POGs that have
+been formally verified in the \lean{} proof assistant.  These programs both prove the soundness of our proof framework and
+create the first formally verified toolchain for weighted and unweighted model counting.
 \end{abstract}
@@
 developed formally verified versions of the proof checker and ring
-evaluator in the \lean{} proof assistant.  Running these two programs
+evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.  Running these two programs
 has the effect of formally verifying both the logical equivalence of
@@
 generated and efficiently (and, in some cases, formally) checked
-\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress has also
-been made on extending these proof systems to other domains, including quantified Boolean
+\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress
+has also been made on extending these proof systems to other domains,
+including quantified Boolean
 formulas~\cite{bryant:cade:2021,heule:JAR2014}.
@@
 
-We know of no other work addressing the question of
-certifying the output of a knowledge compiler.
+We know of no other
+work addressing the question of certifying the output of a knowledge
+compiler.
 
@@
   In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
+\label{def:weight}
 \end{definition}
@@
 
-The {\em weighted model counting} (WMC) problem is also defined over
-rational numbers.  Some formulations of WMC allow
+The {\em weighted model counting}  problem is also defined over
+rational numbers.  Some formulations  allow
 independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
@@
 $\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
-if they are logically equivalent.  If they are not equivalent, then
+if they are logically equivalent.
+%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
+If they are not equivalent, then
 the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
@@
 \section{Partitioned-Operation Graphs (POGs)}
+\label{sect:pog}
 
@@
 \label{prop:ring:eval}
-Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
+Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
 \begin{eqnarray}
@@
 \end{proposition}
+As is described in Section~\ref{subsection:counting}, we have proved these three equations using \lean{}.
 
@@
 \end{itemize}
-  Although not required for representing d-DNNF graphs,
-  both of these generalizations allow more flexibility in the POG
+  These generalizations allow more flexibility in the POG
   representation while maintaining the ability to efficiently perform ring evaluation.
@@
 A proof in our framework takes the form of a sequence of clause addition
-and deletion steps, with each step maintaining logical equivalence to
-the original formula.  More precisely, the status of the proof at any step is represented as
+and deletion steps, with each step preserving the solutions to
+the original formula.  The status of the proof at any step is represented as
 a set of {\em active} clauses $\phi$, i.e., those that
@@
 Starting with input formula $\inputformula$,
-we want to maintain the invariant that
+the proof must maintain the invariant that
 $\inputformula \ifandonlyif \exists Z\,\phi$.
@@
 occurs only when defining $\pand$ and $\por$ operations, as is
-described in Section~\ref{sect:crat}.  The other is by {\em
-  implication redundancy}.  That is, we can add clause $C$ to the set
-of clauses forming formula $\phi$ when $\phi \imply C$.  Clauses can
-also be deleted through implication redundancy.  That is, given a set
-of clauses $\{C\} \cup \phi$, we can remove $C$ if $\phi \imply C$.
+described in Section~\ref{sect:crat}.  Clauses can also be added or deleted based on
+{\em  implication redundancy}.
+That is, when clause $C$ satisfies $\phi \imply C$ for some formula $\phi$, then it can either be added to form the set $\phi \cup \{C\}$ or it can be deleted from that set to form $\phi$.
 
@@
 is the core rule supported by standard
-proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic.
-It is a way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
+proof checkers~\cite{wetzler14_drattrim,RAT} for propositional logic. It provides a simple and efficient
+way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
 Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
@@
 it can be seen that ours has important differences.  Most
-significantly, we want our proofs to maintain logical equivalence at
+significantly, we want our proofs to maintain preserve the set of solutions at
 each step, whereas an unsatisfiability proof need only guarantee that
-no proof step makes a satisfiable set of clauses become
-unsatisfiable.  Therefore, our proofs must justify clause deletions as well as clause additions.
+a proof step does not makes a satisfiable set of clauses become
+unsatisfiable.  In particular, our proofs must justify clause deletions as well as clause additions.
 
@@
   Declaring a node $\nodeu$
-implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{\nodeu}$
+implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
   encoding the product or sum operation.
@@
 Let $\inputformula$ denote the input formula and $\pogformula$ represent the formula encoded by unit clause $r$ and the defining clauses for the POG nodes:
-$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{\nodeu}$.  We write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
+$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.  We write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
 any assignment to the input variables implicitly defines the assignments to the extension variables.
@@
 graph into a POG is straightforward---each node in a d-DNNF graph can
-be directly represented as a POG node.  Rather than representing every
-d-DNNF node explicitly, our program performs simplifications to
+be directly represented as a POG node, although
+our program performs simplifications to
 eliminate Boolean constants.
@@
 of {\em literal} nodes corresponding to input variables and their negations, along with
-{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.
+{\em nonterminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.
 
@@
 takes advantage of the clausal representations of
-the input formula $\inputformula$ and the POG formula $\pogformula$, consisting of the defining clauses of its nodes plus a unit clause for root literal $r$.
-Furthermore, it can
-negate the POG formula by simply negating its root literal.  More precisely $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{\nodeu}}{\{\obar{r}\}}$
+the input formula $\inputformula$ and the POG formula $\pogformula$.
+It can
+express the negation of $\pogformula$ as $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
 Forward implication
 will hold
-when $\inputformula \imply \pogformula$.  Equivalently,
-the formula $\inputformula \land \neg \pogformula$
-should be unsatisfiable, where the
+when $\inputformula \imply \pogformula$, or  equivalently,
+that the formula $\inputformula \land \neg \pogformula$
+is unsatisfiable, where the
 conjunction can be expressed as the union
@@
 $u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
-that $\left(\bigwedge_{\lit \in \passign} \lit\right) \imply u$, i.e.,
+that $(\bigwedge_{\lit \in \passign} \lit) \imply u$, i.e.,
 that any total
@@
 \end{enumerate}
-Observe that most of these steps involve a polynomial number of
+Most of these steps involve a polynomial number of
 operations per recursive call.  The only one that could have
@@
 knowledge compiler competition, along the lines of the annual
-competitions for other automated reasoning tools.  Our experiments
+competitions for other automated reasoning tools.  (The results could be evaluated for correctness
+either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
+Our experiments
 consider the benchmark problems and computing capabilities that would
-arise in such a competition and evaluate whether our implementations
+arise in such a competition and evaluate whether our implementationed toolchain
 could handle these problems.  As with the original proof-generating
@@
 counting.  It performs arithmetic over a subset of the rationals
+we call $\drational$,
 consisting of numbers of the form
@@
 
-Our experiments demonstrate that our toolchain could handle a
-substantial portion of the solutions generated in a knowledge compiler
-competition.  Further engineering and optimization of our proof
-generator and checker could improve their performance and capacity,
-and we could increase the time limit for these programs.
+We are hopeful that having checkable proofs for knowledge compilers
+will allow them to be used in applications where high levels of trust
+are required, and that it will provide a useful tool for developers of
+knowledge compilers.  Although our current implementation only handles
+the outputs of one particular program, we believe it could be adapted
+to work with others, as well as with some model counters.
+
+Our experiments demonstrate that our toolchain can already
+ handle problems nearly at the limits of current knowledge compilers.
+Further engineering and optimization of our proof
+generator and checker could improve their performance and capacity substantially.
 
@@
 defining clauses, and so the total number of defining clauses in pog $P$ equals
-the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$.
+the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$ in Section~\ref{sect:pog}.
 The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.

### 2023-03-17 doc/lean.tex

@@
 \end{lstlisting}
-%%We arbitrarily assign 0 to implications and bi-implications;
-%%recall that, by definition, these do not appear in partitioned formulas.
 Proposition~\ref{prop:ring:eval} is then formalized as follows:

### 2023-03-18 doc/mc-paper.tex
Minor changes

@@
 models.  For $n = |\varset|$, scaling the density by $2^n$
-yields the number of models.
+yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
+in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.
 
@@
 \item They allow negation of arbitrary nodes in the graph, not just
-  variables.  As we have seen, ring evaluation can readily handle negation.
+  variables.
   %, and so we allow it in the interest of generality.
@@
 A proof in our framework takes the form of a sequence of clause addition
-and deletion steps, with each step preserving the solutions to
-the original formula.  The status of the proof at any step is represented as
+and deletion steps, with each step preserving the set of solutions to
+the original formula.
+The status of the proof at any step is represented as
 a set of {\em active} clauses $\phi$, i.e., those that
-have been added but not yet deleted.  Our framework is based
-on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
+have been added but not yet deleted.
+Our framework is based
+on {\em extended} resolution~\cite{Tseitin:1983}, where proof
 steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
@@
 Compared to the proofs of unsatisfiability generated by SAT solvers,
-it can be seen that ours has important differences.  Most
-significantly, we want our proofs to maintain preserve the set of solutions at
-each step, whereas an unsatisfiability proof need only guarantee that
-a proof step does not makes a satisfiable set of clauses become
-unsatisfiable.  In particular, our proofs must justify clause deletions as well as clause additions.
+ours have important differences.  Most
+significantly, each proof step must preserve the set of solutions with respect to the input variables.
+By contrast, an unsatisfiability proof need only guarantee that
+no proof step cause a satisfiable set of clauses to become
+unsatisfiable.  As a consequence, our proofs must justify clause deletions as well as clause additions.
 
@@
 
-A CRAT file provides both a declaration of a POG, as well as an easily checkable
-proof that a Boolean formula, given in conjunctive normal
-form, is logically equivalent to the POG, thus enabling certified
-knowledge compilation.
 %% The proof system is based on extended
@@
 
-The CRAT format draws its inspiration from the LRAT~\cite{lrat} and
+A CRAT file provides both a declaration of a POG, as well as a checkable
+proof that a Boolean formula, given in conjunctive normal
+form, is logically equivalent to the POG\@.
+The proof format draws its inspiration from the LRAT~\cite{lrat} and
 QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.

### 2023-03-18 doc/mc-data.tex
Bug fixes in MICE
Added Lean checker data

@@
 \item 84 total data points
-\item 67 were completed by both programs in under 1000 seconds
-\item Of those 21 were faster with MICE, 46 with CRAT
+\item 76 were completed by both programs in under 1000 seconds
+\item Of those 21 were faster with MICE, 55 with CRAT
 \item 7 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
 \item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
-\item 9 completed with CRAT but generated errors for MICE.
 \end{itemize}
+
+\section{Benchmarking the \lean{} Proof Checker}
+
+Ran \lean{} checker on the benchmarks yielding at most $10^7$ clauses, with time limit of 1000 seconds.
+\begin{itemize}
+\item 80 total
+\item Had 2 that failed RUP check (need to investigate)
+\item One of them needs to be rerun (Very long generation time)
+\item Two timed out for \lean{}
+\end{itemize}
+
+\input{figure-check}
 

### 2023-03-18 doc/mc-paper.tex
Added Lean checker data

@@
 \subsection{Monolithic Forward Implication Proofs}
+\label{app:experiment:monolithic}
 
@@
 \subsection{Impact of Optimizations}
+\label{app:experiment:optimize}
 
@@
 combination of the two optimizations yielded proofs totaling at most
-100~million clauses.  Compared to the problems used to assess monolithic
-proof generation, this set included one that had a short proof but
+10~million clauses.  Compared to the problems used to assess monolithic
+proof generation (Section~\ref{app:experiment:monolithic}), this set included one that had a short proof but
 long runtime and excluded four that had short runtimes but long

### 2023-03-18 doc/mc-paper.tex
minor changes

--- 2023-03-17 doc/mc-paper.tex
@@
 {\tt r} command.  It can occur anywhere in the file.  Except in degenerate cases, it
-will be the extension variable representing the root of a graph
+will be the extension variable representing the root of a graph.
 %, but in
@@
 $\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
-obtained from $\psi$ by: 1) eliminating any
+obtained from $\psi$ by: i) eliminating any
 clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
-2) for the remaining clauses eliminating those literals $\lit$ for
-which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
+ii) for the remaining clauses eliminating those literals $\lit$ for
+which $\passign(\lit) = 0$, and iii) eliminating any duplicate clauses.
 In doing these simplifications, we also track the {\em provenance}

### 2023-03-18 doc/leanappendix.tex

@@
 
-%% \subsection{Evaluation}
+\subsection{Evaluation}
 
-%% [TODO: Say something to establish that the procedure run in reasonable time and produce
-%% useful results; or maybe this goes in the next section?]
+\input{figure-check}
+
+We ran the \lean{} checker on the 80 benchmark problems for which the
+generated CRAT file had at most 10~million total clauses.  These are the
+same benchmarks used in evaluating the optimizations (Section
+\ref{app:experiment:optimize}).  Figure~\ref{fig:lean} shows the
+comparative performance of the \lean{} checker versus our unverified
+checker on the Y axis, with the total number of clauses on the X axis.
+
+As would be expected, the \lean{} checker consistently runs slower
+than the unverified checker.  Two of the benchmarks exceeded the time
+limit of 1000 seconds with the \lean{} checker.  Of those that
+completed, the ratio of the runtime for \lean{} versus the runtime for
+the unverified checker had a harmonic mean of 5.9.
+
+Importantly, however, it can be seen that both checkers show the same
+overall asymptotic performance.  The ratio of the two runtimes was
+less than $8.0$ for all but two very small benchmarks.  This seems like a
+reasonable price to pay for a rigorous guarantee of correctness.

### 2023-03-18 doc/lean.tex
Polished main part of paper

@@
 formulas with variables \lstinline{Var}, which are positive natural numbers,
-corresponding to the indexing of variables in the CNF file.
+corresponding to the indexing of variables in the CNF and CRAT files.
 It is convenient to take a truth assignment \lstinline{PropAssignment Var} to a function that
@@
 The goal of the proof checker is to construct a POG that is equivalent to the input CNF.
-Each element of a POG  is either a variable, a binary disjunction,
-or an arbitrary conjunction:
+Each element of a POG  is either a variable, a binary disjunction (sum),
+or an arbitrary conjunction (product):
 \begin{lstlisting}

### 2023-03-18 doc/mc-paper.tex
Polished main part of paper

@@
 count. Existing knowledge compilers
-provide no guarantee that the representation they generate
+provide no guarantee that the representation they produce
 correctly represents the original formula, and therefore they cannot validate the generated counts.
@@
 We have developed a program that generates POG representations from
-the d-DNNF graphs generated by the state-of-the-art knowledge compiler
+the d-DNNF graphs produced by the state-of-the-art knowledge compiler
 \dfour, as well as checkable proofs that the output POGs
-are equivalent to the input CNF formulas.  We have implemented an
-efficient checker that can handle large proofs. Our tool chain
+are equivalent to the input CNF formulas.  We have also implemented an
+efficient checker for these proofs. Our tool chain
 for generating and verifying POGs scales to all but the largest d-DNNF
-graphs generated by \dfour{} for formulas from a recent model counting
+graphs produced by \dfour{} for formulas from a recent model counting
 competition. We also present a checker and a model counter for POGs that have
@@
 
-Modern Boolean satisfiability (SAT) solvers provide the ability to
-find a satisfying assignment to a Boolean formula or to prove that no
+Modern Boolean satisfiability (SAT) solvers can
+find a satisfying assignment to a Boolean formula or generate a proof that no
 such assignment exists.  They have applications across a variety of
-domains, including formal verification of hardware, software, and
-security protocols, computational mathematics, and combinatorial
-optimization.  Some problems of interest, however, require going
+domains, including computational mathematics, combinatorial
+optimization, and the formal verification of hardware, software, and
+security protocols.  Some applications, however, require going
 beyond standard Boolean satisfiability.  For example, the {\em model
@@
 users of SAT solvers to fully trust their results.  Experience has also
-shown that having this capability gives developers the ability quickly
-detect and identify bugs in their solvers, and this, in turn, has led
+shown that having this capability gives SAT solver developers an ability to quickly
+detect and diagnose bugs in their programs, and this, in turn, has led
 to more reliable SAT solvers.
@@
 In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
-a generalization of the representations commonly used in knowledge
-compilation, including d-DNNF\@.  We present the CRAT file format.
+a generalization of the representations produce by current knowledge
+compilers, including d-DNNF\@.  We present the CRAT file format.
 It captures both the structure of a POG and a checkable proof
@@
 {\em proof generator} generates a CRAT description from a d-DNNF
-graph generated by \dfour{}, a state-of-the-art knowledge compiler.
+graph produced by \dfour{}, a state-of-the-art knowledge compiler.
 A {\em proof checker} then verifies that the combination of CNF and CRAT files
@@
 important role of establishing the soundness of our proof framework.
-Indeed, in the course of developing a formally verified checker, we
-identified subtle aspects of the proof obligations required to assure
-soundness.
+Our
+experience with developing a formally verified proof checker has shown
+us that, even within the well-understood framework of extended
+resolution, it can be challenging to formulate a full set of requirements to guarantee soundness.
+Our efforts to formalize the framework
+identified subtle details for some of the proof obligations.
 
@@
 programs, and they present a series of lemmas and a theorem stating
-that their checking framework is sound.  They also
-present experimental results for cases where they modify existing
-model counters to also generate proof traces, and where they convert
-d-DNNF graphs into a checkable form.  Our tool chain is not directly
+that their checking framework is sound.  They also present
+experimental results for cases where they modify existing model
+counters to generate proof traces, and where they generate proof
+traces from d-DNNF graphs.
+%% They have not formally verified their
+%%proof framework or created a formally verified checker.
+
+Our tool chain is not directly
 comparable to theirs---they certify the output of a standard modeling
 program, while we certify the output of a knowledge compiler and then
-produce a certified standard or weighted model count.  Their tool
-chain can be applied to some programs that are pure model counters,
-while ours can be applied to pure knowledge compilers.  Nonetheless,
+produce a certified standard or weighted model count.
+%% Their tool
+%% chain can be applied to some programs that are pure model counters,
+%%while ours can be applied to pure knowledge compilers.
+Nonetheless,
 we compare the performance of our tool chain to theirs as part of our
-experimental evaluation in Appendix~\ref{app:experiments}.  Our
-experience with developing a formally verified proof checker has shown
-us that, even for the well understood framework of extended
-resolution, it can be challenging to formulate the full set of requirements to guarantee soundness.
+experimental evaluation in Appendix~\ref{app:experiments}.
 
@@
 work addressing the question of certifying the output of a knowledge
-compiler.
+compiler or model counter.
 
@@
     with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcup_{i=1,k} \modelset(\phi_i)$.
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{i=1 \leq i \leq k} \phi_i$ is a Boolean formula, with
+      $\dependencyset(\phi) = \bigcup_{i=1 \leq i \leq k} \dependencyset(\phi_i)$ and
+      $\modelset(\phi) = \bigcap_{i=1 \leq i \leq k} \modelset(\phi_i)$.
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{i=1 \leq i \leq k} \phi_i$ is a Boolean formula, with
+      $\dependencyset(\phi) = \bigcup_{i=1 \leq i \leq k} \dependencyset(\phi_i)$ and
+      $\modelset(\phi) = \bigcup_{i=1 \leq i \leq k} \modelset(\phi_i)$.
   \end{enumerate}
@@
       \begin{enumerate}
-      \item The arguments to a product must have disjoint dependency sets.  That is, operation
-        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \item The arguments to a sum must have disjoint models.  That is, operation
-        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
+      \item The arguments to each product must have disjoint dependency sets.  That is, operation
+        $\bigwedge_{i=1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
+      \item The arguments to each sum must have disjoint models.  That is, operation
+        $\bigvee_{i=1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
       \end{enumerate}
@@
 We propose a general framework for summarizing properties of Boolean
-formulas inspired by the formulation of {\em algebraic model counting}~\cite{kimmig:jal:2017}.
+formulas along the lines of algebraic model counting~\cite{kimmig:jal:2017}.
 
@@
 $\rmul$ be addition and multiplication over rational numbers and using
-label $w(x) = 1/2$ for every variable $x$.
+weight function $w(x) = 1/2$ for every variable $x$.
 Ring evaluation of formula $\phi$ gives the {\em density} of
@@
 ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
-performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
+performing ring evaluation with weight function $w(x) = W(x)/r(x)$ for each
 variable $x$, and computing the weighted count
@@
 \textstyle
-\scalebox{1.2}{\rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+\scalebox{1.2}{\rep}\left(\Pand_{i=1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
 \textstyle
-\scalebox{1.2}{\rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
+\scalebox{1.2}{\rep}\left(\Por_{i=1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
 %% V2
@@
 %\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
-%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
-%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
+%\rep\left(\Pand_{i=1 \leq i \leq k} \phi_i, w \right) &=& \prod_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+%\rep\left(\Por_{i=1 \leq i \leq k} \phi_i, w\right) &=& \sum_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
 \end{eqnarray}
@@
 There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.
+
 As can be seen, a POG is represents a partitioned-operation
@@
 
-POGs are inspired by the d-DNNF graphs devised by Darwiche for
-representing Boolean functions \cite{darwiche:jair:2002}.
-POGs generalize d-DNNF in two ways:
+POGs generalize the d-DNNF graphs in two ways:
 \begin{itemize}
@@
 %%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
-
 \item They allow arbitrary arguments to a $\por$ operation, as long as
   they have disjoint models.  By contrast, each
-  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
+  sum node $\nodeu$ in a d-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
   any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
@@
 occurs only when defining $\pand$ and $\por$ operations, as is
-described in Section~\ref{sect:crat}.  Clauses can also be added or deleted based on
-{\em  implication redundancy}.
-That is, when clause $C$ satisfies $\phi \imply C$ for some formula $\phi$, then it can either be added to form the set $\phi \cup \{C\}$ or it can be deleted from that set to form $\phi$.
+described in Section~\ref{sect:crat}.  Clauses can also be added or
+deleted based on {\em implication redundancy}.  That is, when clause
+$C$ satisfies $\phi \imply C$ for formula $\phi$, then it can either
+be added $\phi$ to create the formula $\phi \cup \{C\}$ or it can be deleted
+from $\phi \cup \{C\}$ to create $\phi$.
 
@@
 is the core rule supported by standard
-proof checkers~\cite{wetzler14_drattrim,RAT} for propositional logic. It provides a simple and efficient
+proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
 way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
@@
   encoding the product or sum operation.
-  This is the only means for adding extension variables to the proof.
+%  This is the only means for adding extension variables to the proof.
 \item Boolean negation is supported implicitly by allowing the
@@
   A clause can only be added if it is logically implied by the existing clauses.
-  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
+  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
 \item
@@
   A clause can only be deleted if it is logically implied by the remaining clauses.
-  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
+  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
 \item The checker must track the dependency set for every input and
@@
   providing a RUP proof that the arguments are mutually exclusive.
-  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
-  $k\,(k-1)/2$ proofs of disjointness.
+  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness
+%  Generalizing to $k$ arguments would require
+%  $k\,(k-1)/2$ proofs of disjointness.
 \end{itemize}
@@
 ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
-such that any clause identifier $C'$ given in the hints when adding clause $C$ must have $C' < C$.
+such that any clause identifier $C'$ given in the hint when adding clause $C$ must have $C' < C$.
 
@@
 (command {\tt a}), with a sequence of antecedent clauses (the
-``hints'').
+``hint'').
 Similarly for clause deletion (command {\tt d}).
@@
 representing Boolean constant $1$.  The only clause added to the proof will be
-the unit literal $v$.  A reference to literal $-v$ then becomes a way
-of representing constant $0$.
+the unit literal $v$.  A reference to literal $-v$ then provides a way
+to representing constant $0$.
 
@@
 this declaration implicitly causes three clauses to be added to the proof, defining extension variable $v$ to be the sum of its arguments.
-The hints must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.
+The hint must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.
 
@@
 Let $\inputformula$ denote the input formula and $\pogformula$ represent the formula encoded by unit clause $r$ and the defining clauses for the POG nodes:
-$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.  We write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
+$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.  We can write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
 any assignment to the input variables implicitly defines the assignments to the extension variables.
@@
 of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
-More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov(C, \psi)$ denote
+More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov_{\passign}(C, \psi)$ denote
 those clauses $C' \in \psi$, such that
@@
 We then extend the definition of $\prov$ to any simplified formula
-$\gamma$ as $\prov(\gamma, \psi) = \bigcup_{C \in \gamma} \prov(C, \psi)$.
+$\gamma$ as $\prov_{\passign}(\gamma, \psi) = \bigcup_{C \in \gamma} \prov_{\passign}(C, \psi)$.
 
@@
   $\simplify{\psi}{\passign}$ must consist of the single unit clause
-  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
+  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
   Any of these can
@@
   literals.
-  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
+  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
   We recursively call
@@
 \end{enumerate}
-  We then combine as hints the results from the BCP and SAT calls for
+  We then generate the target clause for node $\nodeu$,
+creating the hint by combining the results from the BCP and SAT calls for
   the literals, the recursively computed target clauses, and all but
   the first defining clause for node $\nodeu$
-(see Table~\ref{tab:defining}(A))
-to generate the target clause for node $\nodeu$.
+(see Table~\ref{tab:defining}(A)).
 \end{enumerate}
 Most of these steps involve a polynomial number of
-operations per recursive call.  The only one that could have
-exponential complexity is when a SAT solver must be invoked to validate a
-literal.
+operations per recursive call, with the exception of those that call
+a SAT solver to validate a literal.
 
@@
 hints that were used in their original assertions.  By reversing the
-order, those clauses that were used as hints when a clause was
+order, those clauses that were used in the hint when a clause was
 added will still remain when it is deleted.
@@
 Each input clause deletion can be done as a single RUP step.  The
-proof generator builds a sequence of hints from among the defining
+proof generator constructs the hint sequence from the defining
 clauses of the POG nodes via a single, bottom-up pass through the
@@
 program adds $\obar{u}$ to the RUP literals and adds the appropriate
-defining clause to the list of hints.  A literal node for
+defining clause to the hint.  A literal node for
 $\lit$ will be marked if $\lit \in C$, with no hint required.  If
@@
 $\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
-added to the hints.  Marking sum node $\nodeu$ requires that its two children are marked.
+added to the hint.  Marking sum node $\nodeu$ requires that its two children are marked.
 The first defining
@@
 It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
-each clause deletion requires a single RUP step having a set of hints that is
+each clause deletion requires a single RUP step having a hint with length
 bounded by the number of POG nodes.
@@
 to the proof framework---they build on the power of extended
-resolution to create new logical structures within the proof.  They
+resolution to create new logical structures.  They
 involve declaring new product nodes to encode products of literals.
@@
 representation of the formula; they serve only to enable the forward
-implication proof.  Here we only summarize the two optimizations.
-More details are presented in Appendix~\ref{app:optimizations}.
+implication proof.  Here we summarize the two optimizations.
+Appendix~\ref{app:optimizations} provides more details.
 
@@
 separately, calling a SAT solver for every literal that cannot be
-validated with BCP\@.  The literal grouping optimization handles
+validated with BCP\@.  Literal grouping handles
 all of these literals together.  It defines product node $\nodev$
@@
 a solver with $v$ set to 0 yields an unsatisfiability proof that can
-be mapped back to a sequence of clause additions in the CRAT file that
-validates all of the literals.
+be mapped back to a sequence of clause additions in the CRAT file validating
+all of the literals.
 
@@
 $\nodeu$ satisfies forward implication for a formula
-$\gamma_{\nodeu}$, where some of the clauses in this formula can be
-input clauses from $\inputformula$, but others can be simplified
+$\gamma_{\nodeu}$, where some of the clauses in this formula are
+input clauses from $\inputformula$, but others are simplified
 versions of input clauses.
@@
 
-\section{Experimental Evaluation}
-\label{sect:experimental}
+\section{Implementations}
 We have implemented versions of the three programs that, along with the \dfour{} knowledge compiler,  form the
@@
 system.
-Our long-term goal is to have a toolchain suitable for use in a
-knowledge compiler competition, along the lines of the annual
-competitions for other automated reasoning tools.  (The results could be evaluated for correctness
-either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
-Our experiments
-consider the benchmark problems and computing capabilities that would
-arise in such a competition and evaluate whether our implementationed toolchain
-could handle these problems.  As with the original proof-generating
-SAT competitions, we assume that the proofs at a competition could be
-checked with the unverified checker.  Having a verified checker
-that works on some subset of the benchmarks would provide an
-additional degree of certainty.
+%% Our long-term goal is to have a toolchain suitable for use in a
+%% knowledge compiler competition, along the lines of the annual
+%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
+%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
+%% Our experiments
+%% consider the benchmark problems and computing capabilities that would
+%% arise in such a competition and evaluate whether our implementationed toolchain
+%% could handle these problems.  As with the original proof-generating
+%% SAT competitions, we assume that the proofs at a competition could be
+%% checked with the unverified checker.  Having a verified checker
+%% that works on some subset of the benchmarks would provide an
+%% additional degree of certainty.
 
@@
 
+\section{Experimental Evaluation}
+\label{sect:experimental}
+
+%% Our experiments are designed to evaluate the following:
+%% \begin{itemize}
+%% \item
+%%   whether or toolchain can handle challenging benchmark problems,
+%% \item  
+%% the effectiveness of some design choices and optimizations,
+%% \item
+%% the tradeoffs between the unverified and verified checker, and
+%% \item
+%% how the performance of our toolchain compares to that of the MICE model counter verifier.
+%% \end{itemize}
+
 We provide only a summary of our experiments here.  A more complete
@@
 180 unique CNF files among these, ranging in size from 250 to
-2,753,207 clauses.  We ran our programs on a processor with 64~GB of
-memory and having an attached high-speed, solid-state
-disk---capabilities that should be included in the cluster nodes for
-future competitions.
-
-With a runtime limit of 4000 seconds (the model counting competition
-has a limit of 3600 seconds), \dfour{} completed for 124 of the
+2,753,207 clauses.
+%%We ran our programs on a processor with 64~GB of
+%%memory and having an attached high-speed, solid-state
+%%disk.
+With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
 benchmark problems.  Our proof generator was able to convert all of
@@
 
-Appendix~\ref{app:experiments} also presents experimental results
-comparing the monolithic and structural methods for generating forward
-implication proofs, the effects of our two optimizations of structural
-proofs, and a comparison with the MICE proof framework.
+We found that the monolithic approach for generating the forward
+implication proof works well for smaller POGs, but it becomes
+unreliable and inefficient for larger ones.  These experiments suggest
+a possible hybrid approach, stopping the recursion of the structural
+approach and shifting to monolithic mode once a subgraph size is below some
+threshold.  We found that our two optimizations: literal grouping and
+lemmas can provide substantial improvements in proof size and runtime.
+
+Running the verified proof checker in \lean{} showed the same scaling
+with respect to the proof size as did the unverified checker, but it
+required, on average, around 6 times longer.
+
+Finally, the runtimes for the MICE toolchain versus our CRAT toolchain
+showed little correlation, reflecting the fact that the two solve
+slightly different problems and use different approaches.  In general,
+our CRAT toolchain showed better scaling, in part due to its ability
+to control the recursion through lemmas.
 
@@
 are required, and that it will provide a useful tool for developers of
-knowledge compilers.  Although our current implementation only handles
-the outputs of one particular program, we believe it could be adapted
-to work with others, as well as with some model counters.
-
+knowledge compilers.
+%%Although our current implementation only handles
+%%the outputs of one particular program, it could be adapted
 Our experiments demonstrate that our toolchain can already
@@
 %% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
-%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{i=1,k}
+%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{i=1 \leq i \leq k}
 %% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
@@
 \toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hints} & Explanation \\
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
 \midrule
@@
 depending on whether the node has two children or three.  The last two
-nonzero values in the sum declarations are hints providing the
+nonzero values in the sum declarations are the hint providing the
 required mutual exclusion proof.
@@
 Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
-$\nodes_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
+$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
 \ref{fig:eg4:all}(A) that these two clauses resolve to the clause
@@
 have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
-clause IDs 16 and 19 as hints.  These resolve to the clause
+clause IDs 16 and 19 as the hint.  These resolve to the clause
 $\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
@@
 be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
-hints for clause 36.
+hint for clause 36.
 
@@
 The proof proceeds by first assuming that the clause is false, requiring
-$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hints then consist of unit
+$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
 clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
@@
 involves simply deleting the clauses from clause number 35 down to
-clause number 25, with each deletion using the same hints as were used
+clause number 25, with each deletion using the same hint as were used
 to add that clause.  In the end, therefore, only the defining clauses

### 2023-03-18 doc/mc-paper.tex
Minor edits of appendix sections

@@
 
-%% Our experiments are designed to evaluate the following:
-%% \begin{itemize}
-%% \item
-%%   whether or toolchain can handle challenging benchmark problems,
-%% \item  
-%% the effectiveness of some design choices and optimizations,
-%% \item
-%% the tradeoffs between the unverified and verified checker, and
-%% \item
-%% how the performance of our toolchain compares to that of the MICE model counter verifier.
-%% \end{itemize}
-
 We provide only a summary of our experiments here.  A more complete
@@
 
-We step through a detailed example to provide a better understanding of the CRAT proof framework.
+We step through portions of the file to provide a better understanding of the CRAT proof framework.
 Figure
@@
 $\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
-\ref{fig:eg4:all}(A) that these two clauses resolve to the clause
+\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
 $\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
 have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
-clause IDs 16 and 19 as the hint.  These resolve to the clause
+clause IDs 16 and 19 as the hint.  These form a RUP proof for the clause
 $\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
@@
 general resolution framework to define new structures within the
-proof. In particular, they create new product nodes that are not part
+proof.
+They create new product nodes that are not part
 of the POG representation.
@@
 $\assign(u) = 1$.  This lemma is then invoked for every node having
-$\nodeu$ as a child.  As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.
+$\nodeu$ as a child.
+As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.
 
@@
 
-The primary purpose of our experiments is to assess the ability of our
-implementation of the toolchain illustrated in Figure~\ref{fig:chain}
-to check the results from a possible knowledge compiler competition,
-along the lines of the annual competitions in SAT solvers, model
-counters, and other automated reasoning tools.  As secondary goals, we
-seek insights into how possible algorithmic choices affect the
-performance of the toolchain and how our toolchain compares to the
-MICE framework for verifying model counters~\cite{fichte:sat:2022}.
+Our experiments are designed to evaluate the following:
+\begin{itemize}
+\item
+  whether or toolchain can handle challenging benchmark problems,
+\item  
+the effectiveness of some design choices and optimizations,
+\item
+the tradeoffs between the unverified and verified checker, and
+\item
+how the performance of our toolchain compares to that of the MICE model counter verifier.
+\end{itemize}
 
@@
 would generate.  A POG operation with $k$ arguments requires $k+1$
-defining clauses, and so the total number of defining clauses in pog $P$ equals
+defining clauses, and so the total number of defining clauses in POG $P$ equals
 the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$ in Section~\ref{sect:pog}.
@@
 
-For each of the 124 d-DNNF graphs generated by \dfour{}, we attempted
-to use our proof generator and unverified checker, limiting proof
-generation to 10,000 seconds.  We also attempted to generate and check
+For each of the 124 d-DNNF graphs generated by \dfour{}, we ran
+our proof generator and unverified checker, limiting proof
+generation to 10,000 seconds.  We ran the programs to generate and check
 one-sided proofs for the graphs, again with a time limit of 10,000
@@
 We were able to complete a full validation for 108 of the 124
-benchmarks, with times ranging from fractions of a second to 13,144.
+benchmarks, with times ranging from fractions of a second to 13,144
 seconds, with a median of 71.6 seconds.
 
-Relative to the runtime for \dfour{}, two problems ran faster
-with the generator and checker.  These were for problems having small numbers of
-models (relative to the number of variables), and so most of the time
-spent by both \dfour{} and our proof generator was in running a CDCL
-solver to search the very sparse solution space.
-\cadical{} generally outperforms the miniSAT-based solver used by
-\dfour{}.  At the other extreme, one problem that required only 19
-seconds for \dfour{} required 8657 seconds to generate a proof and 45
-seconds to check it.  It is unclear why the proof generation required
-so much time.  Overall the ratios between the combined generation plus
-checking times versus the time for \dfour{} had a harmonic mean of
-5.5.
+Relative to the runtime for \dfour{}, two problems ran faster with the
+generator and checker.  These were for problems having small numbers
+of models (relative to the number of variables), and so most of the
+time spent by both \dfour{} and our proof generator was in running a
+CDCL solver to search the very sparse solution space.  \cadical{}
+generally outperforms the miniSAT-based solver used by \dfour{}.  At
+the other extreme, one problem that required only 19 seconds for
+\dfour{} required 8657 seconds to generate a proof and 45 seconds to
+check it.  This one particular benchmark appears to expose weak performance
+limitations by our implementation of BCP\@.  Overall the ratios
+between the combined generation plus checking times versus the time
+for \dfour{} had a harmonic mean of 5.5.
 
@@
 our programs use for clause identifiers.  Another exited due to the
-virtual memory limit imposed by the operating system, and the
-other 14 timed out.  Of the 15 that did not overflow the clause
-counter, we were able to complete a one-sided verification for 9, but
-the other 6 timed out.
+virtual memory limit imposed by the operating system, and the other 14
+timed out.  Of the 15 that did not overflow the clause counter, we
+were able to complete a one-sided verification for 9, but the other 6
+timed out.  Overall, we were able to provide some level of
+verification for 94\% of the benchmark problems.
 
@@
 size.  It would begin the recursive descent of $\validate$ according
-to a structural approach, but monolithically generate forward
-implication proofs once it encounters sufficiently small subgraphs.
+to a structural approach, but monolithically generate
+the proof once it encounters a sufficiently small subgraph.
 

### 2023-03-19 doc/mc-paper.tex
adjust phrasing in introduction

@@
 \emph{Knowledge compilation} is the process of converting a Boolean
-formula into a form that makes it possible to calculate associated
-values efficiently, such as the weighted or unweighted model
+formula into a form that makes it possible to efficiently calculate
+associated values, such as the weighted or unweighted model
 count. Existing knowledge compilers
@@
 are equivalent to the input CNF formulas.  We have also implemented an
-efficient checker for these proofs. Our tool chain
+efficient checker for these proofs. Our toolchain
 for generating and verifying POGs scales to all but the largest d-DNNF
@@
 
-Modern Boolean satisfiability (SAT) solvers can
-find a satisfying assignment to a Boolean formula or generate a proof that no
+Given a Boolean formula $\phi$, modern Boolean satisfiability (SAT) solvers can
+find an assignment satisfying $\phi$ or generate a proof that no
 such assignment exists.  They have applications across a variety of
-domains, including computational mathematics, combinatorial
+domains including computational mathematics, combinatorial
 optimization, and the formal verification of hardware, software, and
 security protocols.  Some applications, however, require going
-beyond standard Boolean satisfiability.  For example, the {\em model
+beyond Boolean satisfiability.  For example, the {\em model
   counting problem} requires computing the number of satisfying
 assignments of a formula, including in cases where there are far too many
-to enumerate individually.  Model counting, in turn, has
+to enumerate individually.  Model counting has
 applications in artificial intelligence, computer security, and
@@
 
-Model counting is a challenging problem---more challenging that the
-already NP-hard problem of Boolean satisfiability.  Several
+Model counting is a challenging problem---more challenging than the
+already NP-hard Boolean satisfiability.  Several
 tractable variants of Boolean satisfiability, including 2-SAT, become
-intractable when the goal becomes model counting and not just
+intractable when the goal is to count models and not just determine
 satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
-model counters have developed that scale to very large formulas, as
+model counters that scale to very large formulas have been developed, as
 witnessed by the progress in recent model counting competitions.
@@
 transforms the formula into a structured form for which model counting
-becomes straightforward.  For example, the {\em deterministic,
+is straightforward.  For example, the {\em deterministic,
   decomposable, negation normal form} (d-DNNF) introduced by
@@
 labeled by Boolean variables and their complements, and with each
-nonterminal node labeled by Boolean operation And or Or.  Restrictions
+nonterminal node labeled by a Boolean And/Or operation.  Restrictions
 are placed on the structure of the graph such that a count of the
 models can be computed by a single bottom-up traversal.
-Kimmig et al.\ present a very general framework, known as {\em
-  algebraic model counting}~\cite{kimmig:jal:2017}, describing
-properties of Boolean functions that can be efficiently computing from
+Kimmig et al.\ present a very general {\em
+  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
+properties of Boolean functions that can be efficiently computed from
 a d-DNNF representation.  These include standard and weighted model
@@
 users of SAT solvers to fully trust their results.  Experience has also
-shown that having this capability gives SAT solver developers an ability to quickly
-detect and diagnose bugs in their programs, and this, in turn, has led
+shown that proof production capabilites allow SAT solver developers to quickly
+detect and diagnose bugs in their programs. This, in turn, has led
 to more reliable SAT solvers.
 
-In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
-a generalization of the representations produce by current knowledge
-compilers, including d-DNNF\@.  We present the CRAT file format.
-It captures both the structure of a POG and a checkable proof
-that the POG is logically equivalent to a Boolean formula given in
-conjunctive normal form (CNF).  A proof consists of a sequence of
-clause addition and deletion steps, based on an extended resolution
-proof system~\cite{Tseitin:1983}.  We present a set of conditions that
-a CRAT file must satisfy to provide an equivalence proof.
+In this paper, we introduce \emph{Partitioned Operation Graphs} (POGs),
+a generalization of the representations produced by current knowledge
+compilers, including d-DNNF\@. To capture both the structure of a POG
+and a checkable proof of its logical equivalence to a Boolean formula in
+conjunctive normal form (CNF), we propose the CRAT file format. A CRAT
+proof consists of a sequence of clause addition and deletion steps,
+based on an extended resolution proof system~\cite{Tseitin:1983}.
+We establish a set of conditions that a CRAT file must satisfy in order to
+encode a well-formed POG and provide a valid equivalence proof.
 
@@
 }
-\caption{Certifying Tool Chain.
-  The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT) that can be independently checked and evaluated}
+\caption{Certifying toolchain.
+  The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT)
+  which can be independently checked and evaluated}
 \label{fig:chain}
@@
 
-Figure~\ref{fig:chain} illustrates the certifying knowledge compilation
-and (weighted) model counting tool chain that we have developed.  A
+Figure~\ref{fig:chain} illustrates our certifying knowledge compilation
+and (weighted) model counting toolchain.  A
 {\em proof generator} generates a CRAT description from a d-DNNF
@@
 either a standard or weighted model count from the POG\@ representation.
-We evaluate the capabilities of this tool chain using
+We evaluate the capabilities of this toolchain using
 benchmark formulas from the 2022 standard and weighted model
@@
 graphs generated by \dfour{}.
-As the dashed box in Figure~\ref{fig:chain} indicates, this tool chain
-shifts the need for trustworthy code away from the complex and highly
+As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
+moves the root of trust away from the complex and highly
 optimized knowledge compiler to a relatively simple checker and
 evaluator.  Importantly, the proof generator need not be
-trusted---its errors will also be caught by the proof checker.
-
-To further enhance the trustworthiness of our toolchain, we have
-developed formally verified versions of the proof checker and ring
-evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.  Running these two programs
-has the effect of formally verifying both the logical equivalence of
-the POG to the input formula and the correctness of the count.  In
-addition, the existence of a formally verified checker serves the
-important role of establishing the soundness of our proof framework.
-Our
-experience with developing a formally verified proof checker has shown
-us that, even within the well-understood framework of extended
-resolution, it can be challenging to formulate a full set of requirements to guarantee soundness.
-Our efforts to formalize the framework
-identified subtle details for some of the proof obligations.
+trusted---its errors will be caught by the proof checker.
+
+To ensure soundness of the abstract CRAT proof system, as well as correctness of its
+concrete implementation, we formally verified the proof system as well as versions
+of the proof checker and ring evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
+Running these two programs on a particular CRAT file gives assurance that the proof
+and the model count are correct. Our experience with developing a formally verified
+proof checker has shown that, even within the well-understood framework of extended
+resolution, it can be challenging to formulate a full set of requirements that guarantee soundness.
+In fact, subtle details in the partitioned sum rule were identified during this process.
 
@@
 
-Having a SAT solver generate a proof of unsatisfiability has a long
+Generating proofs of unsatisfiability in SAT solvers has a long
 tradition~\cite{ZhangMalik} and has become widely accepted due to the
@@
 
-Our tool chain is not directly
-comparable to theirs---they certify the output of a standard modeling
+Our toolchain is not directly
+comparable to theirs---they directly certify the output of a model counting
 program, while we certify the output of a knowledge compiler and then
-produce a certified standard or weighted model count.
+produce a certified model count.
 %% Their tool
@@
 Nonetheless,
-we compare the performance of our tool chain to theirs as part of our
+we compare the performance of our toolchain to theirs as part of our
 experimental evaluation in Appendix~\ref{app:experiments}.
@@
 %%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
-%%file from an input formula in conjunctive normal form.  Our tool chain
+%%file from an input formula in conjunctive normal form.  Our toolchain
 %%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},

### 2023-03-19 doc/mc-paper.tex
more style changes

@@
 
-As can be seen, a POG is represents a partitioned-operation
+In essence, a POG represents a partitioned-operation
 formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
@@
 
-We define the {\em size} of POG $P$, written $|P|$ to be the
+We define the {\em size} of POG $P$, written $|P|$, to be the
 the number of nodes labeled $\pand$ or $\por$ plus the number of edges from these nodes to their children.  Ring
@@
 \item They allow arbitrary arguments to a $\por$ operation, as long as
-  they have disjoint models.  By contrast, each
+  these have disjoint models.  By contrast, each
   sum node $\nodeu$ in a d-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
@@
 $C$ satisfies $\phi \imply C$ for formula $\phi$, then it can either
-be added $\phi$ to create the formula $\phi \cup \{C\}$ or it can be deleted
+be added to $\phi$ to create the formula $\phi \cup \{C\}$ or it can be deleted
 from $\phi \cup \{C\}$ to create $\phi$.
@@
 $\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
-\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
+\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
 clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
@@
 the unit literal $v$.  A reference to literal $-v$ then provides a way
-to representing constant $0$.
+of representing constant $0$.
 
@@
 \end{enumerate}
-Let $\inputformula$ denote the input formula and $\pogformula$ represent the formula encoded by unit clause $r$ and the defining clauses for the POG nodes:
-$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.  We can write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
+In other words, at the end of the proof it must hold that the active clauses be exactly those
+in $\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
+of unit clause $r$ and the defining clauses of the POG nodes. We can write
+$\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
 any assignment to the input variables implicitly defines the assignments to the extension variables.
-The sequence of clause addition steps provide a {\em forward implication} proof that
+Let $\inputformula$ denote the input formula.
+The sequence of clause addition steps provides a {\em forward implication} proof that
 $\modelset(\inputformula) \subseteq \modelset(P)$.  That is, any total
@@
 scalable.  The other version, consisting of a proof checker and
-ring evaluator has been formally verified within the \lean{} proof
-system.
+ring evaluator has been formally verified within the \lean{} theorem prover.
 %% Our long-term goal is to have a toolchain suitable for use in a

### 2023-03-19 doc/leanappendix.tex
One more Lean data point
Final data
Change explanation
Final Lean data point

@@
 As would be expected, the \lean{} checker consistently runs slower
-than the unverified checker.  Two of the benchmarks exceeded the time
-limit of 1000 seconds with the \lean{} checker.  Of those that
-completed, the ratio of the runtime for \lean{} versus the runtime for
-the unverified checker had a harmonic mean of 5.9.
+than the unverified checker.  All of the checks completed and confirmed the outcome of the unverified checker.
+The ratio of the runtime for \lean{} versus the runtime for
+the unverified checker had a harmonic mean of $5.94$.
 
 Importantly, however, it can be seen that both checkers show the same
-overall asymptotic performance.  The ratio of the two runtimes was
-less than $8.0$ for all but two very small benchmarks.  This seems like a
-reasonable price to pay for a rigorous guarantee of correctness.
+overall performance trends.  The ratio of the two runtimes was less
+than $8.0$ for all but three small benchmarks.  This seems like a
+reasonable price to pay for a rigorous guarantee of correctness, and
+we are confident that we can reduce this gap with more effort in
+optimizing the checker.

### 2023-03-19 doc/mc-paper.tex
adjust phi/theta notation

@@
 extension variable $u$ associated with any POG node $\nodeu$, we can therefore
-assume that any total assignment $\assign$ to the input variables will
-also assign a value to $u$ such that $\assign(u) =
-1$ if and only if $\assign \in \modelset(\nodeu)$.  For a POG with
-root literal $r$, we can define its set of models $\modelset(P)$ as
-the set of all total assignments $\assign$ such that $\assign(r) = 1$.
+prove that any total assignment $\assign$ to the input variables which
+also satisfies the POG definitions will
+assign a value to $u$ such that $\assign(u) =
+1$ if and only if $\assign \in \modelset(\nodeu)$.
 
@@
 in $\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
-of unit clause $r$ and the defining clauses of the POG nodes. We can write
-$\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
-any assignment to the input variables implicitly defines the assignments to the extension variables.
+of unit clause $r$ and the defining clauses of the POG nodes. Recognizing that
+any assignment to the input variables implicitly defines the assignments to the extension variables,
+we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
 Let $\inputformula$ denote the input formula.
 The sequence of clause addition steps provides a {\em forward implication} proof that
-$\modelset(\inputformula) \subseteq \modelset(P)$.  That is, any total
+$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
 assignment $\assign$ satisfying the input formula must also satisfy
@@
 that
-$\modelset(P) \subseteq \modelset(\inputformula)$.
+$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.
 
@@
 For input formula $\inputformula$ and its translation into a POG $P$
-with root literal $r$, the most challenging part of the proof is to
-show that $\modelset(\inputformula) \subseteq \modelset(P)$, i.e.,
-that any total assignment $\alpha$ that is a model of $\inputformula$
+with root node $\noder$, the most challenging part of the proof is to
+show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
+that any total assignment $\assign$ that is a model of $\inputformula$
+and the POG definition clauses
 will yield $\assign(r) = 1$, for root literal $r$.  This part of the
@@
 takes advantage of the clausal representations of
-the input formula $\inputformula$ and the POG formula $\pogformula$.
+the input formula $\inputformula$ and the POG formula $\phi_\noder$.
 It can
-express the negation of $\pogformula$ as $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
+express the negation of $\phi_\noder$ as $\neg \phi_\noder = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
 Forward implication
 will hold
-when $\inputformula \imply \pogformula$, or  equivalently,
-that the formula $\inputformula \land \neg \pogformula$
+when $\inputformula \imply \phi_\noder$, or  equivalently,
+that the formula $\inputformula \land \neg \phi_\noder$
 is unsatisfiable, where the
@@
 Completing the equivalence proof of input formula $\inputformula$ and its POG
-representation with root literal $r$ requires showing that
-$\modelset(P) \subseteq \modelset(\inputformula)$.  This is done in the
+representation with root node $\noder$ requires showing that
+$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
 CRAT framework by first deleting all asserted clauses, except for the
@@
 graph.  The RUP deletion proof for input clause $C$ effectively proves that any
-total assignment $\assign$ that does not satisfy $C$ will yield
+total assignment $\assign$ that satisfies the POG definition clauses
+but does not satisfy $C$ will yield
 $\assign(r) = 0$.  It starts with the set of literals

### 2023-03-19 doc/leanappendix.tex
Refer to checkers as prototype and verified

@@
 \subsection{Evaluation}
+\label{app:lean:evaluation}
 
@@
 \ref{app:experiment:optimize}).  Figure~\ref{fig:lean} shows the
-comparative performance of the \lean{} checker versus our unverified
+comparative performance of the \lean{} checker versus our unverified prototype
 checker on the Y axis, with the total number of clauses on the X axis.
@@
 As would be expected, the \lean{} checker consistently runs slower
-than the unverified checker.  All of the checks completed and confirmed the outcome of the unverified checker.
+than the prototype checker.  All of the checks completed and confirmed the outcome of the prototype checker.
 The ratio of the runtime for \lean{} versus the runtime for
-the unverified checker had a harmonic mean of $5.94$.
+the prototype checker had a harmonic mean of $5.94$.
 

### 2023-03-19 doc/mc-paper.tex
Refer to checkers as prototype and verified

--- 2023-03-19 doc/mc-paper.tex
@@
 
-Our toolchain is not directly
-comparable to theirs---they directly certify the output of a model counting
-program, while we certify the output of a knowledge compiler and then
-produce a certified model count.
+Our toolchain is not directly comparable to theirs---they directly
+certify the numeric value produced by a model counter, while we 
+certify that the graph generated by a knowledge compiler is logically
+equivalent to the input formula.  We can also validate the unweighted or
+weighted count generated from that representation.
 %% Their tool
@@
 work addressing the question of certifying the output of a knowledge
-compiler or model counter.
+compiler.
 
@@
 
+\begin{definition}[Partitioned-Operation Formula]
   A {\em partitioned-operation formula}
@@
       \end{enumerate}
+\end{definition}
      We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.
@@
 \section{Implementations}
-We have implemented versions of the three programs that, along with the \dfour{} knowledge compiler,  form the
-toolchain illustrated in Figure~\ref{fig:chain}.  One version,
-including a proof generator, proof checker, and ring evaluator for
-standard and weighted model counting, is designed to be fully
-scalable.  The other version, consisting of a proof checker and
-ring evaluator has been formally verified within the \lean{} theorem prover.
+We have implemented versions of the three programs that, along with
+the \dfour{} knowledge compiler, form the toolchain illustrated in
+Figure~\ref{fig:chain}.  The proof generator is the same in both
+cases, since it need not be trusted.
+Our {\em verified}
+versions of the proof checker and ring evaluator have been formally
+verified within the \lean{} theorem prover.  Our long term goal is to
+rely on these.  Our {\em prototype} versions are written in C
+(checker) and Python (ring evaluator).  These are currently faster and
+more scalable, but we anticipate their need will diminish as the
+verified version is further optimized.
 %% Our long-term goal is to have a toolchain suitable for use in a
@@
 
-Our ring evaluator can perform both standard and weighted model
-counting.  It performs arithmetic over a subset of the rationals
-we call $\drational$,
-consisting of numbers of the form
-$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
-with $a$ implemented to have arbitrary range.
-Allowing scaling by powers of 2 enables the density computation
-and rescaling required for standard model counting.  Allowing scaling
-by powers of both 2 and 5 enables exact decimal arithmetic, handling
-the weights used in the weighted model counting competitions.
+Our prototype ring evaluator can perform both standard and weighted
+model counting.  It performs arithmetic over a subset of the rationals
+we call $\drational$, consisting of numbers of the form $a \cdot 2^{b}
+\cdot 5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented
+to have arbitrary range.  Allowing scaling by powers of 2 enables the
+density computation and rescaling required for standard model
+counting.  Allowing scaling by powers of both 2 and 5 enables exact
+decimal arithmetic, handling the weights used in the weighted model
+counting competitions.  For one of the weighted benchmarks, it
+generated a result with 260,909 decimal digits.
 
@@
 one-sided proofs for an additional 9 of them, leaving just 7 with no
-verification.  The checker successfully verified all of the generated
+verification.  The prototype checker successfully verified all of the generated
 proofs.  The longest runtime for the combination of proof generator
@@
 Running the verified proof checker in \lean{} showed the same scaling
-with respect to the proof size as did the unverified checker, but it
-required, on average, around 6 times longer.
+with respect to the proof size as did the prototype checker, but it
+required, on average, around 5.9 times longer.
 
@@
 \item
-the tradeoffs between the unverified and verified checker, and
+the tradeoffs between the prototype and verified checker, and
 \item
@@
 For each of the 124 d-DNNF graphs generated by \dfour{}, we ran
-our proof generator and unverified checker, limiting proof
+our proof generator and prototype checker, limiting proof
 generation to 10,000 seconds.  We ran the programs to generate and check
@@
 side, since their checker provides a count, although only an
-unweighted one.  We ran their tools on the 92 benchmark problems for
+unweighted one.  We used the prototype versions of the checker and ring evaluator.
+We ran their tools on the 92 benchmark problems for
 which our toolchain requires at most 3000 seconds and retained

### 2023-03-19 doc/mc-paper.tex
fix negation statement

--- 2023-03-19 doc/mc-paper.tex
@@
 the input formula $\inputformula$ and the POG formula $\phi_\noder$.
-It can
-express the negation of $\phi_\noder$ as $\neg \phi_\noder = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
-Forward implication
-will hold
-when $\inputformula \imply \phi_\noder$, or  equivalently,
-that the formula $\inputformula \land \neg \phi_\noder$
+We can express the negation of $\phi_\noder$ in clausal form as
+$\theta_{\obar{\noder}} = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
+Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
+when the formula $\inputformula \land \theta_{\obar{\noder}}$
 is unsatisfiable, where the

### 2023-03-20 doc/lean.tex
Overall tuning

@@
 We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
-proof checker and both model and weighted model counters, in the \lean{} programming language
+proof checker and both standard and weighted model counters, in the \lean{} programming language
 and proof assistant~\cite{demoura:cade:2021}.

### 2023-03-20 doc/mc-paper.tex
phi_I
Minor wording changes
Overall tuning

@@
   \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
-  \node [left] at (-1,20) {\large {$\phi$}};
+  \node [left] at (-1,20) {\large {$\inputformula$}};
   \node [above] at (2,20) {\texttt{.cnf}};
@@
   \draw[thick] (62,4) [-latex] -- (66,4) ;
-  \node [right] at (66,4) {\large {$\textbf{R}(\phi)$}};
+  \node [right] at (66,4) {\large {$\textbf{R}(\inputformula)$}};
 
@@
 certify that the graph generated by a knowledge compiler is logically
-equivalent to the input formula.  We can also validate the unweighted or
+equivalent to the input formula.  We can also validate the standard or
 weighted count generated from that representation.
@@
 expressed as ring evaluation problems.  The
-{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
+(standard) {\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
 It can be cast as a ring evaluation problem by having $\radd$ and
@@
 It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
-There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.
+Every POG must have a designated {\em root node} $\noder$.
 
@@
 In other words, at the end of the proof it must hold that the active clauses be exactly those
-in $\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
-of unit clause $r$ and the defining clauses of the POG nodes. Recognizing that
-any assignment to the input variables implicitly defines the assignments to the extension variables,
+in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
+of unit clause $\{r\}$ and the defining clauses of the POG nodes. Recognizing that
+any total assignment to the input variables implicitly defines the assignments to the extension variables,
 we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
@@
 %%strategy to generate a d-DNNF representation.
-Converting a d-DNNF
-graph into a POG is straightforward---each node in a d-DNNF graph can
-be directly represented as a POG node, although
-our program performs simplifications to
+A d-DNNF
+graph can be directly translated  into a POG,
+although in doing this conversion,
+our program also performs simplifications to
 eliminate Boolean constants.
@@
 of {\em literal} nodes corresponding to input variables and their negations, along with
-{\em nonterminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.
+{\em nonterminal} nodes, which can be further classified as {\em product} and {\em sum} nodes.
 
@@
 proof consists of a series of clause assertions leading to one adding
-$r$ as a unit clause.  We have devised two methods of generating this
+$\{r\}$ as a unit clause.  We have devised two methods of generating this
 proof.  The {\em monolithic} approach makes just one call to a
@@
 We can express the negation of $\phi_\noder$ in clausal form as
-$\theta_{\obar{\noder}} = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
+$\theta_{\obar{\noder}} \doteq \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
 Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
@@
 of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
-For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
+For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
 Our experimental results show
@@
 The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
-The result will therefore be to add unit clause $r$ to the proof.
+The result will therefore be to add unit clause $\{r\}$ to the proof.
 Here we present a correct, but somewhat inefficient formulation of
@@
 The subgraph with root node $\nodeu$ should be a POG representation of the formula
-$\simplify{\psi}{\passign}$.  Having the
-routine add its target clause indicates that any total assignment
-$\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.
+$\simplify{\psi}{\passign}$.
+%% Having the
+%% routine add its target clause indicates that any total assignment
+%% $\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.
 
@@
 final unit clause for root literal $r$, and then deleting all of the
-input clauses.  Suppose at some point in this process that we want to
-delete clause $C$ such that the remaining set of clauses will form the
-CNF formula $\phi$.  Deleting $C$ requires proving that
-$\phi \imply C$.
+input clauses.
+%%  Suppose at some point in this process that we want to
+%% delete clause $C$ such that the remaining set of clauses will form the
+%% CNF formula $\phi$.  Deleting $C$ requires proving that
+%% $\phi \imply C$.
 
@@
 The first defining
-clause for this node (see Table~\ref{tab:defining}(B)) will then be added as a hint.  At the very end, the program
+clause for this node (see Table~\ref{tab:defining}(B)) will then be added to the hint.  At the very end, the program
 (assuming the reverse implication holds) will attempt to mark root
@@
 exponential number of recursive steps.  Instead, for each node
-$\nodeu$ having $\indegree(u) > 1$, the program can define and
+$\nodeu$ having $\indegree(\nodeu) > 1$, the program can define and
 generate the proof of a {\em lemma} for $\nodeu$ when it is first reached by
@@
 a possible hybrid approach, stopping the recursion of the structural
-approach and shifting to monolithic mode once a subgraph size is below some
+approach and shifting to monolithic mode once the subgraph size is below some
 threshold.  We found that our two optimizations: literal grouping and
@@
 showed little correlation, reflecting the fact that the two solve
-slightly different problems and use different approaches.  In general,
+different problems and use different approaches.  In general,
 our CRAT toolchain showed better scaling, in part due to its ability

### 2023-03-20 doc/mc-paper.tex
Minor changes to appendix

@@
 variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
-  $\simplify{\psi}{\passign} \cup \simplify{\theta_{\nodev}}{\{ \obar{v} \}}$ to show that is unsatisfiable.
-  The proof steps can be mapped back into clause addition steps in the CRAT file, using as hints the
-  input clauses and the defining clauses for $\nodev$.
+  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
+  The proof steps can be mapped back into clause addition steps in the CRAT file, incorporating the 
+  input clauses and the defining clauses for $\nodev$ into the hints.
 
@@
 $\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
-total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ must yield
+total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ and the defining clauses for the POG must yield
 $\assign(u) = 1$.  This lemma is then invoked for every node having

### 2023-03-20 doc/leanappendix.tex
Fix up parts of appendix

@@
 same benchmarks used in evaluating the optimizations (Section
-\ref{app:experiment:optimize}).  Figure~\ref{fig:lean} shows the
-comparative performance of the \lean{} checker versus our unverified prototype
-checker on the Y axis, with the total number of clauses on the X axis.
+\ref{app:experiment:optimize}).
+All of the checks completed and confirmed the outcome of the prototype checker.
 
+Figure~\ref{fig:lean} shows the
+performance of both the \lean{} checker and the prototype
+checker on the Y axis, as functions of the total number of clauses on the X axis.
 As would be expected, the \lean{} checker consistently runs slower
-than the prototype checker.  All of the checks completed and confirmed the outcome of the prototype checker.
+than the prototype checker.  
 The ratio of the runtime for \lean{} versus the runtime for
@@
 we are confident that we can reduce this gap with more effort in
-optimizing the checker.
+optimizing the verified checker.

### 2023-03-20 doc/mc-paper.tex
Fix up parts of appendix

@@
 \item
-the tradeoffs between the prototype and verified checker, and
+how the performance of our toolchain compares to that of the MICE model counter verifier,
 \item
-how the performance of our toolchain compares to that of the MICE model counter verifier.
+the tradeoffs between the prototype and verified checker.  These experiments are described in Section~\ref{app:lean:evaluation}.
 \end{itemize}
@@
 \dfour{} required 8657 seconds to generate a proof and 45 seconds to
-check it.  This one particular benchmark appears to expose weak performance
-limitations by our implementation of BCP\@.  Overall the ratios
+check it.  This benchmark appears to stem from weak performance
+by our implementation of BCP\@.  Overall the ratios
 between the combined generation plus checking times versus the time
@@
 
-We encountered a significant number of errors running their tools.
-These seem mostly to be easily fixable, and we are currently
-supporting their debugging efforts.  At present, we can summarize the results as:
+We can summarize the results as:
 \begin{itemize}

### 2023-03-20 doc/mc-paper.tex
add labels

@@
 
-\begin{definition}[Partitioned-Operation Formula]
+\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
   A {\em partitioned-operation formula}
@@
 \subsection{Semantics}
+\label{subsection:semantics}
 
@@
 POG is logically equivalent to the input formula.
-At the completion of the proof, the following conditions must hold:
+At the completion of the proof, the following \textsc{final conditions} must hold:
 \begin{enumerate}

### 2023-03-20 doc/lean.tex
new Lean section

--- 2023-03-17 doc/lean.tex
@@
-\section{A Formally Verified CRAT Checker and POG Model Counter}
+\section{A Formally Verified Toolchain}
+\label{section:formally-verified-toolchain}
 
-We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
-proof checker and both model and weighted model counters, in the \lean{} programming language
-and proof assistant~\cite{demoura:cade:2021}.
-In this section, we briefly describe the functions we have implemented in Lean and
-the specifications we have proved.
-More information is provided in the Appendix~\ref{appendix:lean}.
+We set out to formally verify the system with two goals in mind: first, to ensure that the CRAT framework is mathematically
+sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \cref{fig:chain}).
+These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}.
+Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic.
+In this section, we briefly describe the functionality we implemented and what we proved about it.
+More information is provided in \cref{appendix:lean}.
 
-\subsection{The Proof Checker}
-\label{subsection:proof:checker}
+\vspace{1em}\noindent
+\textbf{Proof checking.} The goal of a CRAT proof is to construct a POG that is equivalent to the input CNF $\inputformula$.
+The checker begins by parsing the input formula, initializing the set of active clauses to $\theta \leftarrow \inputformula$, and the POG $P$ to an empty one.
+It then processes every step of the CRAT proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
+Afterwards, it carries out the \textsc{final conditions} check of \cref{subsection:semantics}.
+Throughout the process, we maintain invariants which ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root.
 
-Our specifications are built on a generic library for the syntax and semantics of
-propositional logic. We define the data type \lstinline{PropForm Var} of propositional
-formulas with variables \lstinline{Var}, which are positive natural numbers,
-corresponding to the indexing of variables in the CNF file.
-It is convenient to take a truth assignment \lstinline{PropAssignment Var} to a function that
-assigns a Boolean value to every assignment, and later restrict attention to subsets of the
-variables.
+The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \cref{section:logical:foundations}.
+Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm Var} of propositional formulas.
+Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}.
+Enforcing totality is no limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$.
+We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.
 
-We define the following representations for CNF formulas.
-A literal is represented as a nonzero integer, a clause is an array of literals,
-and a CNF formula is an array of clauses.
+The invariants refer to the checker state \lstinline{st} containing fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far.
+For any $\nodeu\in P$, \lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.
+The first two invariants state that assignments to original variables extend uniquely to extension variables defining the POG nodes.
+In the formalization, we split this into two parts---extension and uniqueness:
 \begin{lstlisting}
-def ILit := { i : Int // i ≠ 0 }
-abbrev IClause := Array ILit
-abbrev ICnf := Array IClause
+/-- Any assignment satisfying φ₁ extends to φ₂ while preserving values on X. -/
+def extendsOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
+  ∀ (σ₁ : PropAssignment Var), σ₁ ⊨ φ₁ → ∃ σ₂, σ₁.agreeOn X σ₂ ∧ σ₂ ⊨ φ₂
+
+invariants.extends_pogDefsForm : extendsOver st.inputCnf.vars ⊤ st.pogDefsForm
+
+/-- Assignments satisfying φ are determined on Y by their values on X. -/
+def uniqueExt (X Y : Set Var) (φ : PropForm Var) :=
+  ∀ (σ₁ σ₂ : PropAssignment Var), σ₁ ⊨ φ → σ₂ ⊨ φ → σ₁.agreeOn X σ₂ →
+    σ₁.agreeOn Y σ₂
+
+invariants.uep_pogDefsForm : uniqueExt st.inputCnf.vars st.allVars st.pogDefsForm
 \end{lstlisting}
-We also define functions \lstinline{ILit.toPropForm}, \lstinline{IClause.toPropForm},
-and \lstinline{ICnf.toPropForm} to relate these to propositional formulas and their
-semantics.
+The next one guarantees that the set of solutions over the original variables is preserved:
+\begin{lstlisting}
+def equivalentOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
+  extendsOver X φ₁ φ₂ ∧ extendsOver X φ₂ φ₁
 
-The goal of the proof checker is to construct a POG that is equivalent to the input CNF.
-Each element of a POG  is either a variable, a binary disjunction,
-or an arbitrary conjunction:
-\begin{lstlisting}
-inductive PogElt where
-  | var  : Var → PogElt
-  | disj : Var → ILit → ILit → PogElt
-  | conj : Var → Array ILit → PogElt
+invariants.equivalent_clauseDb : equivalentOver st.inputCnf.vars
+  st.inputCnf st.clauseDb
 \end{lstlisting}
-In the first case, the argument \lstinline{x} in the expression
-\lstinline{var x} is the index
-of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
-it is the definition number in the CRAT file. Note that representing edges as literals allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
-A \lstinline{Pog} is then simply an array of \lstinline{PogElt}.
-For each POG \lstinline{P} and literal \lstinline{l}, we define
-\lstinline{P.toPropForm l} to be the
-propositional formula that arises from interpreting \lstinline{l} as a propositional
-formula, unfolding all the defined conjunctions and disjunctions.
-A partitioned-operation formula, as defined in \cref{section:logical:foundations}, is expressed as:
+Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\cref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
 \begin{lstlisting}
@@
   | neg φ    => φ.partitioned
- | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(φ.eval τ ∧ ψ.eval τ)
- | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ (φ.vars ∩ ψ.vars = ∅)
+  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(τ ⊨ φ ∧ τ ⊨ ψ)
+  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ φ.vars ∩ ψ.vars = ∅
+
+invariants.partitioned : ∀ u : ILit, (st.pog.toPropForm u).partitioned
+
+invariants.equivalent_lits : ∀ u : ILit, equivalentOver st.inputCnf.vars
+    (u ∧ st.pogDefsForm) (st.pog.toPropForm x)
 \end{lstlisting}
+These invariants are maintained by valid CRAT proofs. Together with a few additional, omitted invariants
+which ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
+node $\noder$: the \textsc{final conditions} ensure that $\theta$ is logically equivalent to
+$\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$
+(\lstinline{equivalent st.clauseDb (r ∧ st.pogDefsForm)}), and from the invariants we can prove that
+$\inputformula \iff \phi_\noder$ (\lstinline{equivalent st.inputCnf (st.pog.toPropForm r)}). 
+After certifying a proof, the checker can pass its in-memory POG representation to the ring evaluator.
 
-Our proof checker starts by parsing the CNF formula and storing it internally.
-It then processes and checks each rule of a CRAT proof, throwing an exception if
-the proof is not well-formed. We have proved that if the checker terminates successfully,
-it succeeds in constructing a POG that is partitioned and equivalent to that CNF. This is formalized as \lstinline{(pog.toPropForm r).partitioned ∧ inputCnf.toPropForm ≡ pog.toPropForm r}.
-
-\subsection{The Model Counters}
-\label{subsection:counting}
-
-We have formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem,
-\cref{def:ring_evaluation}, as follows:
+\vspace{1em}\noindent
+\textbf{Ring evaluation}. We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
+(\cref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Here \lstinline{R} is assumed to be a commutative ring.
-The counting scheme of Proposition~\ref{prop:ring:eval} for partitioned formulas is expressed as follows:
+The rules for efficient ring evaluation of partitioned formulas are expressed as:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Proposition~\ref{prop:ring:eval} is then formalized as follows:
+\Cref{prop:ring:eval} is then formalized as follows:
 \begin{lstlisting}
@@
 
-We have implemented an efficient function \lstinline{Pog.ringEval} to calculate the weighted model count
-for an arbitrary node of a Pog, and we have proved that it computes the function above on the
-associated formula:
+To efficiently compute the ring evaluation of a formula represented by a POG node, we implemented
+\lstinline{Pog.ringEval}. We then proved that it matches the specification above:
 \begin{lstlisting}
 theorem ringEval_eq_ringEval (pog : Pog) (weight : Var → R) (x : Var) :
-  pog.ringEval weight x = (pog.toPropForm (.mkPos x)).ringEval weight
+  pog.ringEval weight x = (pog.toPropForm x).ringEval weight
 \end{lstlisting}
-Here \lstinline{.mkPos x} denotes the positive literal for the variable \lstinline{x}.
-Applying this to the output of our verified CRAT proof checker,
-we obtain a proof that
-our toolchain computes the correct weighted model count of the input CNF.
 
-We have also implemented a separate procedure that directly carries out an integer
-calculation of the number of models.
-For partitioned formulas \lstinline{PropForm Var} whose
-variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
-we can count the number of models of the formula on the variables in \lstinline{s}
-recursively as follows:
+Applying this to the output of our verified CRAT proof checker which we know to be partitioned
+and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes
+the correct ring evaluation of $\inputformula$.
+
+\vspace{1em}\noindent
+\textbf{Model counting.} Finally, we established that ring evaluation with the appropriate weights
+corresponds to the standard model count. To this end we specified \lstinline{countModels}
+which directly carries out an integer calculation of the number of models over a set of variables
+of cardinality \lstinline{nVars}:
 \begin{lstlisting}
@@
 \end{lstlisting}
-We have implemented an efficient counting function for POGs and proved that
-it computes that same quantity on the corresponding formulas.
-Finally,
-we have proved that the calculation above really does
-return the number of models:
+We then formally proved that for a partitioned formula whose variables are among a finite set
+\lstinline{s}, this computation really does count the number of models over \lstinline{s}:
 \begin{lstlisting}

### 2023-03-20 doc/leanappendix.tex
new Lean section

@@
 \subsection{Logical objects and operations}
+
+% This didn't fit in the main section:
+\iffalse
+We define the following representations for CNF formulas.
+A literal is represented as a nonzero integer, a clause is an array of literals,
+and a CNF formula is an array of clauses.
+\begin{lstlisting}
+def ILit := { i : Int // i ≠ 0 }
+abbrev IClause := Array ILit
+abbrev ICnf := Array IClause
+\end{lstlisting}
+We also define functions \lstinline{ILit.toPropForm}, \lstinline{IClause.toPropForm},
+and \lstinline{ICnf.toPropForm} to relate these to propositional formulas and their
+semantics.
+\fi
+
+% This didn't fit in the main section:
+\iffalse
+Each element of a POG  is either a variable, a binary disjunction (sum),
+or an arbitrary conjunction (product):
+\begin{lstlisting}
+inductive PogElt where
+  | var  : Var → PogElt
+  | disj : Var → ILit → ILit → PogElt
+  | conj : Var → Array ILit → PogElt
+\end{lstlisting}
+In the first case, the argument \lstinline{x} in the expression
+\lstinline{var x} is the index
+of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
+it is the definition number in the CRAT file. Note that representing edges as literals allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
+A \lstinline{Pog} is then simply an array of \lstinline{PogElt}.
+\fi
 

### 2023-03-20 crat/checker/crat-check.c
Check for defining clause in mutual exclusion proof hint

@@ -632,6 +632,14 @@ clause_block_t *clause_blocks = NULL;
 int clause_block_alloc = 0;
 int clause_block_count = 0;
 
+/*
+  For mutual exclusion proof of sum operation, must ensure that only
+  defining clauses are referenced in hints.  Keep track of ranges of
+  clause IDs that form hints as pair of ilist's
+ */
+ilist clause_defining_start_id = NULL;
+ilist clause_defining_end_id = NULL;
+
 /* Operations */
 void clause_init() {
     /* Allocate a single starting chunk */
@@ -650,6 +658,8 @@ void clause_init() {
     clause_blocks[clause_block_count-1].chunk = ilist_new(MIN_SIZE);
     clause_blocks[clause_block_count-1].offset = ilist_new(MIN_SIZE);
     clause_xvar_reference = ilist_new(MIN_SIZE);
+    clause_defining_start_id = ilist_new(MIN_SIZE);
+    clause_defining_end_id = ilist_new(MIN_SIZE);
 }
 
 /* Return -1, 0, or +1 depending on whether clause is below, within, or beyond block */
@@ -689,6 +699,51 @@ int *clause_locate(int cid) {
 	    return clause_locate_within(bid, cid);
     }
     return NULL;
+}
+
+/* Record range of defining clause IDs */
+void clause_add_defining(int start_cid, int end_cid) {
+    int len = ilist_length(clause_defining_start_id);
+    if (len == 0 || clause_defining_end_id[len-1] < start_cid - 1) {
+	clause_defining_start_id = ilist_push(clause_defining_start_id, start_cid);
+	clause_defining_end_id = ilist_push(clause_defining_end_id, end_cid);
+    } else {
+	clause_defining_end_id[len-1] = end_cid;
+    }
+}
+
+/* Return -1, 0, or +1 depending on whether clause ID is below,
+   within, or beyond indicated range */
+int clause_probe_range(int idx, int cid) {
+    int start_id = clause_defining_start_id[idx];
+    int end_id = clause_defining_end_id[idx];
+    if (cid < start_id)
+	return -1;
+    if (cid > end_id)
+	return +1;
+    return 0;
+}
+
+/* Is this the ID of a defining clause? */
+bool is_defining(int cid) {
+    int len = ilist_length(clause_defining_start_id);
+    if (len <= 0)
+	return false;
[…]

### 2023-03-20 doc/leanappendix.tex
Rerrange appendix

@@
 All of our specifications have been completely proved and verified relative to that.
-
-\subsection{Evaluation}
-\label{app:lean:evaluation}
-
-\input{figure-check}
-
-We ran the \lean{} checker on the 80 benchmark problems for which the
-generated CRAT file had at most 10~million total clauses.  These are the
-same benchmarks used in evaluating the optimizations (Section
-\ref{app:experiment:optimize}).
-All of the checks completed and confirmed the outcome of the prototype checker.
-
-Figure~\ref{fig:lean} shows the
-performance of both the \lean{} checker and the prototype
-checker on the Y axis, as functions of the total number of clauses on the X axis.
-As would be expected, the \lean{} checker consistently runs slower
-than the prototype checker.  
-The ratio of the runtime for \lean{} versus the runtime for
-the prototype checker had a harmonic mean of $5.94$.
-
-Importantly, however, it can be seen that both checkers show the same
-overall performance trends.  The ratio of the two runtimes was less
-than $8.0$ for all but three small benchmarks.  This seems like a
-reasonable price to pay for a rigorous guarantee of correctness, and
-we are confident that we can reduce this gap with more effort in
-optimizing the verified checker.

### 2023-03-20 doc/mc-paper.tex
new Lean section
Tweaked box placements
Rerrange appendix

@@
 \end{proposition}
-As is described in Section~\ref{subsection:counting}, we have proved these three equations using \lean{}.
+As is described in \cref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.
 
@@
 \begin{figure}
-\begin{minipage}{0.62\textwidth}
+\begin{minipage}{0.58\textwidth}
 (A)  Input Formula\\[1.2ex]
@@
 \\[2.5ex]
-\begin{minipage}{0.45\textwidth}
+\begin{minipage}{0.42\textwidth}
 (D) Implicit Clauses\\[1.2ex]
@@
 
+\input{leanappendix}
+
 \section{Detailed Experimental Results}
@@
 \item
-how the performance of our toolchain compares to that of the MICE model counter verifier,
+the comparative performance of the prototype and verified checkers, and
 \item
-the tradeoffs between the prototype and verified checker.  These experiments are described in Section~\ref{app:lean:evaluation}.
+how the performance of our toolchain compares to that of the MICE model counter verifier.
 \end{itemize}
@@
 
+\subsection{Evaluation of the Verified Checker}
+\label{app:lean:evaluation}
+
+\input{figure-check}
+
+We ran the \lean{} checker on the 80 benchmark problems for which the
+generated CRAT file had at most 10~million total clauses.  These are the
+same benchmarks used in evaluating the optimizations (Section
+\ref{app:experiment:optimize}).
+All of the checks completed and confirmed the outcome of the prototype checker.
+
+Figure~\ref{fig:lean} shows the
+performance of both the \lean{} checker and the prototype
+checker on the Y axis, as functions of the total number of clauses on the X axis.
+As would be expected, the \lean{} checker consistently runs slower
+than the prototype checker.  
+The ratio of the runtime for \lean{} versus the runtime for
+the prototype checker had a harmonic mean of $5.94$.
+
+Importantly, however, it can be seen that both checkers show the same
+overall performance trends.  The ratio of the two runtimes was less
+than $8.0$ for all but three small benchmarks.  This seems like a
+reasonable price to pay for a rigorous guarantee of correctness, and
+we are confident that we can reduce this gap with more effort in
+optimizing the verified checker.
+
 \subsection{Comparison with MICE Framework}
@@
 
-\input{leanappendix}
-
 \end{document}

### 2023-03-20 doc/mc-paper.tex
Minor tweaks

@@
 labeled by Boolean variables and their complements, and with each
-nonterminal node labeled by a Boolean And/Or operation.  Restrictions
+nonterminal node labeled by a Boolean And or Or operation.  Restrictions
 are placed on the structure of the graph such that a count of the
@@
 
-In this paper, we introduce \emph{Partitioned Operation Graphs} (POGs),
+This paper introduces \emph{Partitioned Operation Graphs} (POGs),
 a generalization of the representations produced by current knowledge
-compilers, including d-DNNF\@. To capture both the structure of a POG
+compilers. To capture both the structure of a POG
 and a checkable proof of its logical equivalence to a Boolean formula in
@@
 
-Figure~\ref{fig:chain} illustrates our certifying knowledge compilation
-and (weighted) model counting toolchain.  A
-{\em proof generator} generates a CRAT description from a d-DNNF
-graph produced by \dfour{}, a state-of-the-art knowledge compiler.
-A {\em proof checker} then verifies that the combination of CNF and CRAT files
-satisfy our rules, and a {\em ring evaluator} can compute
-either a standard or weighted model count from the POG\@ representation.
-We evaluate the capabilities of this toolchain using
-benchmark formulas from the 2022 standard and weighted model
-competitions.  Our tools are able to handle all but the largest d-DNNF
-graphs generated by \dfour{}.
+Figure~\ref{fig:chain} illustrates our certifying knowledge
+compilation and model counting toolchain.  Starting with
+input formula $\inputformula$, the \dfour{} knowledge compiler
+generates a d-DNNF representation, and the {\em proof
+  generator} uses this to generate a CRAT file.
+The {\em proof checker} verifies the equivalence of the CNF and CRAT representations.
+The {\em ring evaluator} computes
+a standard or weighted model count from the POG\@ representation.
 As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
@@
 In fact, subtle details in the partitioned sum rule were identified during this process.
+
+We evaluate the capabilities of this toolchain using benchmark
+formulas from the 2022 standard and weighted model competitions.  Our
+tools are able to handle all but the largest d-DNNF graphs generated
+by \dfour{}.  We also measure the benefits of several optimizations as
+well as the performance of the verified checker relative to one
+written in C.
 

### 2023-03-20 doc/mc-paper.tex
Fix capitalization

@@
 
-\section{Clausal Proof Frameworks}
+\section{Clausal Proof Framework}
 
@@
 
-\subsection{Forward implication proof}
+\subsection{Forward Implication Proof}
 
@@
 
-\subsection{Reverse implication proof}
+\subsection{Reverse Implication Proof}
 
@@
 implication proof works well for smaller POGs, but it becomes
-unreliable and inefficient for larger ones.  These experiments suggest
-a possible hybrid approach, stopping the recursion of the structural
-approach and shifting to monolithic mode once the subgraph size is below some
-threshold.  We found that our two optimizations: literal grouping and
+unreliable and inefficient for larger ones.
+%% These experiments suggest
+%% a possible hybrid approach, stopping the recursion of the structural
+%% approach and shifting to monolithic mode once the subgraph size is below some
+%% threshold.
+We also found that our two optimizations: literal grouping and
 lemmas can provide substantial improvements in proof size and runtime.
-
 Running the verified proof checker in \lean{} showed the same scaling
@@
 \input{figure-runtime-d4-verify}
+\input{figure-clauses-d4-verify}
 
@@
 
-\input{figure-clauses-d4-verify}
-
 Figure~\ref{fig:defining:total} compares the number of defining
@@
 
-\input{figure-runtime-monolithic}
-
-\input{figure-clauses-monolithic}
-
 To assess the relative merits of the two approaches to forward
@@
 \dtrim{}.
+
+\input{figure-runtime-monolithic}
+\input{figure-clauses-monolithic}
 
@@
 
-\subsection{Comparison with MICE Framework}
+\subsection{Comparison with the MICE Framework}
 

### 2023-03-20 doc/leanappendix.tex
Fix capitalization
Clarified the unsound condition we discovered

@@
-\section{More information about the formally verified checker}
+\section{Detailed Information about the Formally Verified Checker}
 \label{appendix:lean}
@@
 proof checker and both model and weighted model counters, in the \lean{} programming language
-and proof assistant [cite]. \lean{} implements a logical foundation in which expressions
+and proof assistant~\cite{demoura:cade:2021}. \lean{} implements a logical foundation in which expressions
 have a computational interpretation, and just as with other proof assistants like Isabelle and Coq,
@@
 
-Verifying our checker and the model counters in this way serves at least two purposes.
-When we run them on particular data sets, they provide high confidence
-that the particular results we obtain are correct.
-More broadly, however, our success also validates our general approach,
-showing, in particular, that the CRAT proof format is sound.
-Indeed, our formalization helped us discover a bug in our implementation
-of the informal proof checker: in the ``\emph{s}'' step described at the end of
-Subsection~\ref{subsection:syntax}, it is sound (and necessary) to allow defining
-clauses for the new operations in the proof of $\obar{\lit}_1 \lor \obar{\lit}_2$,
-but it is generally not sound to allow the input clauses or other added clauses.
-[NOTE: let's add this specification to the end of Section 6.1.]
-Our verified implementation of the checker tells us that the algorithm is sound and
-provides a reference for other implementations.
+Verifying our checker and the model counters in this way serves at
+least two purposes.  When we run them on particular data sets, they
+provide high confidence that the particular results we obtain are
+correct.  More broadly, however, our success also validates our
+general approach, showing, in particular, that the CRAT proof format
+is sound.  Indeed, our formalization helped us discover a bug in our
+prototype proof checker: the declaration a sum operation via the
+``\texttt{s}'' step (Section~\ref{subsection:syntax}) requires a hint
+consisting of a sequence of clause Ids that provide a RUP proof of the
+mutual exclusion condition $\obar{\lit}_1 \lor \obar{\lit}_2$.  It is
+sound (and necessary) to allow these Ids to be those of defining
+clauses for previous POG nodes, but, as we discovered, it is generally
+not sound to allow them to be for input clauses or other added
+clauses.  We therefore added the condition that only defining clauses
+could be included in the hint, and we implemented a test of this rule
+in our prototype checker.  Our verified implementation of the checker
+tells us that the algorithm is sound and provides a reference for
+other implementations.
 
@@
 
-\subsection{Logical objects and operations}
+\subsection{Logical Objects and Operations}
 
@@
 
-\subsection{Verification specifics}
+\subsection{Verification Specifics}
 \label{subsection:verification:specifics}

### 2023-03-20 doc/mc-paper.tex
add other checkers to related work

@@
 experimental evaluation in Appendix~\ref{app:experiments}.
+
+Verified checkers for unsatisfiability include \cite{Cruz:Filipe:et:al:16,Heule:et:al:17,Lammich:20,Tan:2021}.
+The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
+In contrast, our proofs construct a new propositional formula, and we have to verify that it
+is partitioned and is equivalent to the input.
+This requires verifying additional proof steps and subtle invariants, as described in
+Sections~\ref{sect:crat} and \ref{section:formally-verified-toolchain}.
 

### 2023-03-20 doc/leanappendix.tex
revise the Lean appendix

--- 2023-03-20 doc/leanappendix.tex
@@
-\section{More information about the formally verified checker}
+\section{More Information about the Formally Verified Checker}
 \label{appendix:lean}
 
-We have implemented the rightmost two components of Figure~\ref{fig:chain}, namely the
-proof checker and both model and weighted model counters, in the \lean{} programming language
-and proof assistant [cite]. \lean{} implements a logical foundation in which expressions
-have a computational interpretation, and just as with other proof assistants like Isabelle and Coq,
+Our verified toolchain has been implemented in \lean{}, which is based on
+a formal logical foundation in which expressions
+have a computational interpretation.
+As with other proof assistants like Isabelle and Coq,
 a function defined within the formal system can be compiled to efficient code.
-At the same time, we can state and prove claims about the function within the system, thereby
-verifying that the functions compute the intended results.
+At the same time, we can state and prove claims about the function within the system, thereby verifying that the functions compute the intended results.
 
-Verifying our checker and the model counters in this way serves at least two purposes.
-When we run them on particular data sets, they provide high confidence
-that the particular results we obtain are correct.
-More broadly, however, our success also validates our general approach,
-showing, in particular, that the CRAT proof format is sound.
-Indeed, our formalization helped us discover a bug in our implementation
-of the informal proof checker: in the ``\emph{s}'' step described at the end of
-Subsection~\ref{subsection:syntax}, it is sound (and necessary) to allow defining
-clauses for the new operations in the proof of $\obar{\lit}_1 \lor \obar{\lit}_2$,
-but it is generally not sound to allow the input clauses or other added clauses.
-[NOTE: let's add this specification to the end of Section 6.1.]
-Our verified implementation of the checker tells us that the algorithm is sound and
-provides a reference for other implementations.
+Our code is contained in the directory {\tt ProofChecker}.
+The main checker loop is implemented in {\tt Checker/CheckerCore.lean}.
+That file defines a structure {\tt PreState} that contains all the
+relevant data structures, which include
+the input CNF formula,
+the clause database,
+the POG under construction,
+and the root literal.
+We define a {\tt State} of the checker to be a {\tt PreState}
+that satisfies all the invariants that we need to establish
+the final result,
+and the bulk of our work involved showing that
+these are indeed preserved by each step of the proof, which
+modifies the clause database and the POG.
+The ring evaluator and model counter are contained in the
+{\tt Count} directory.
+Here we provide a few additional notes on the implementation
+and clarify the sense in which the toolchain has been verified.
 
-In this section, we describe the implementation in Lean and the specifications we have proved.
-\lean{}'s logical foundation describes a functional programming language with
-inductive data types,
-but our procedures also make use of calculations with natural numbers, integers, and rationals,
-and data structures for arrays and hashmaps.
-Lean's code extraction uses efficient but unverified versions of these,
-and so in Subsection~\ref{subsection:verification:specifics} we clarify
-what exactly we have verified and what we currently trust.
+\subsection{Implementation notes}
 
-\subsection{Logical objects and operations}
-
-% This didn't fit in the main section:
-\iffalse
-We define the following representations for CNF formulas.
-A literal is represented as a nonzero integer, a clause is an array of literals,
+When thinking about the formal verification, it is helpful to distinguish between
+the data structures that play a role in the code that is
+executed and the definitions that serve as a mathematical reference,
+allowing us to state our specifications and prove that they are met.
+Our representation of CNF formulas is among the former:
+a literal is represented as a nonzero integer, a clause is an array of literals,
 and a CNF formula is an array of clauses.
@@
 \end{lstlisting}
-We also define functions \lstinline{ILit.toPropForm}, \lstinline{IClause.toPropForm},
-and \lstinline{ICnf.toPropForm} to relate these to propositional formulas and their
-semantics.
-\fi
-
-% This didn't fit in the main section:
-\iffalse
+We define the clause database {\tt ClauseDb} as a hashmap that
+stores each clause together with a flag indicating whether it has
+been deleted.
 Each element of a POG is either a variable, a binary disjunction (sum),
@@
 it is the definition number in the CRAT file. Note that representing edges as literals allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
-A \lstinline{Pog} is then simply an array of \lstinline{PogElt}.
-\fi
+A \lstinline{Pog} is then an array of \lstinline{PogElt}
+that is well founded in the sense that each element depends only on prior elements
+in the array.
 
-Our specifications are built on a generic library for the syntax and semantics of
-propositional logic. The data type of propositional formulas with variables indexed by
-a datatype \lstinline{ν} is defined inductively as follows:
-\begin{lstlisting}
-inductive PropForm (ν : Type)
-  | var (x : ν)
-  | tr
-  | fls
-  | neg (φ : PropForm ν)
-  | conj (φ₁ φ₂ : PropForm ν)
-  | disj (φ₁ φ₂ : PropForm ν)
-  | impl (φ₁ φ₂ : PropForm ν)
-  | biImpl (φ₁ φ₂ : PropForm ν)
-\end{lstlisting}
-The last two constructors denote implication and bi-implication, which are not used in this
-project. The usual syntactic and semantic notions are defined in the expected ways.
-It is convenient to take a truth assignment \lstinline{PropAssignment ν} to a function that
-assigns a boolean value to every assignment, and later restrict attention to subsets of the
-variables. In Lean, we introduce the notation \lstinline{τ ⊨ φ} for the statement that
-\lstinline{τ} satisfies \lstinline{φ}. We write \lstinline{entails φ₁ φ₂} to say that
-\lstinline{φ₁} and \lstinline{φ₂} and \lstinline{equivalent φ₁ φ₂}.
-For the purpose of model counting, if \lstinline{s} is any finite set of variables,
-we define \lstinline{models φ s} to be the finite set of truth assignments \lstinline{τ}
-satisfying \lstinline{φ} that assign arbitrary values on \lstinline{s} and
-\lstinline{false} outside \lstinline{s}.
+On the mathematical side, we define propositional formulas and
+their semantics in the usual way.
+Functions \lstinline{ILit.toPropForm}, \lstinline{IClause.toPropForm},
+\lstinline{ICnf.toPropForm}, and \lstinline{Pog.toPropForm}
+relate our data structures to propositional formulas and their semantics.
+For example, given a literal \lstinline{l}, \lstinline{Pog.toPropForm P l}
+denotes the (unnegated or negated) interpretation of the node of the POG
+\lstinline{P} specified by \lstinline{l} as a propositional formula over the
+original variables.
+Lean provides us with convenient ``anonymous projector'' notation that allows
+us to write \lstinline{P.toPropForm l} instead of \lstinline{Pog.toPropForm P l}
+when \lstinline{P} has type \lstinline{Pog},
+\lstinline{C.toPropForm} instead of \lstinline{IClause.toPropForm C} when \lstinline{C}
+has type \lstinline{IClause}, etc.
 
-In logic, the set of propositional formulas modulo equivalence is sometimes called the
-\emph{Lindenbaum--Tarski algebra}. We have found it convenient to define this quotient,
-which we denote \lstinline{PropTerm ν}, and lift the boolean operations and the entailment
+In order to reason about the behavior of the checker, we found it important to work
+with propositional formulas modulo logical equivalence, a structure known in logic
+as the \emph{Lindenbaum--Tarski algebra}.
+We defined the quotient and lifted the boolean operations and the entailment
 relation to this new type. The advantage is that equivalent formulas give
-rise to equal elements in the quotient.
-This makes easier to prove equivalences as we go, since we can calculate with the associated
-elements of the quotient and substitute one for another when the corresponding formulas are equivalent.
+rise to equal elements in the quotient, which
+made is much easier to prove equivalences involving complicated expressions,
+since it enabled us to calculate with the associated
+elements of the quotient and substitute one for another when the corresponding formulas are equivalent. This gave rise to auxiliary challenges, however.
+For example, it is no longer straightforward to say that an element of the quotient ``depends'' on a variable,
+since equivalent formulas can have difference variables.
+Instead, we made use of a semantic notion of dependence, in which an element of the
+quotient depends on a variable if and only if for some truth assignment its value
+changes when the assignment to that variable is flipped.
 
-In our application, we instantiate \lstinline{ν} to a type \lstinline{Var} consisting
-of positive natural numbers, corresponding to the indexing of variables in the DIMACS
-format for SAT. It may be helpful to think of the elements of \lstinline{PropForm Var},
-\lstinline{PropTerm Var}, and \lstinline{PropAssignment Var} as mathematical objects
-rather than computational data. Our proof checker and model counters do not compute
-with them; rather, we use them to specify what the checker and model counters are supposed to do.
-The starting point for the proof checker is, instead, a CNF formula, which \emph{is}
-presented as data.
-Specifically, a literal is a nonzero integer, with the understanding that $-\ell$ represents
-the negation of $\ell$. A clause is an array of literals, and a CNF formula is an array
-of clauses:
-\begin{lstlisting}
-def ILit := { i : Int // i ≠ 0 }
-abbrev IClause := Array ILit
-abbrev ICnf := Array IClause
-\end{lstlisting}
-We define operations that translate these objects to the elements of \lstinline{PropForm Var}
-that they represent. Lean provides us with helpful ``anonymous projection'' syntax that allows
-us to write, for example, \lstinline{C.toPropForm} for the expression
-\lstinline{Clause.toPropForm C}.
-Similarly, we have a representation \lstinline{C.toPropTerm}
-of \lstinline{C} as an equivalence class of propositional formulas, and we can speak of the
-value of \lstinline{C} under any truth assignment.
-Our code represents partial truth assignments as hashmaps:
-\begin{lstlisting}
-abbrev PartPropAssignment := HashMap Var Bool
-\end{lstlisting}
-We have written a procedure that reduces a clause \lstinline{C} with respect to a partial
-assignment, and another procedure that uses a hashset to test whether a clause \lstinline{C} is a
-tautology.
-We have also verified that these low-level procedures do what they are supposed to do.
-
-\subsection{The Proof Checker}
-\label{appendix:proof:checker}
-
-Our proof checker interprets the input clauses listed in a CRAT file as a CNF formula $\varphi$.
-It then processes and checks each rule, throwing an exception if a precondition is not met.
-If it terminates without raising an exception, it outputs a POG, $P$, with distinguished root
-$\bf r$.
-We have formally proved the claim that in this case the formula corresponding to node
-$\bf r$ is a partitioned formula and is
-logically equivalent to $\varphi$.
-The POG $P$ can be saved to a file [TODO: implement this?] or passed on to one of the model counting
-procedures described in the next section.
-
-In our code, each element of a POG  is either a variable, a binary disjunction,
-or an arbitrary conjunction:
-\begin{lstlisting}
-inductive PogElt where
-  | var  : Var → PogElt
-  | disj : Var → ILit → ILit → PogElt
-  | conj : Var → Array ILit → PogElt
-\end{lstlisting}
-In the first case, the argument \lstinline{x} in the expression
-\lstinline{var x} is the index
-of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
-it is the definition number in the CRAT file.
-The CRAT generator described in Section~\ref{section:generating:crat}
-declares new variables sequentially, so we currently make the simplifying assumption
-that the $i$th element of the array has the node with index $i + 1$,
-but it will not be hard to generalize this.
-Notice that the arguments to \lstinline{disj} and \lstinline{conj} are literals,
-corresponding to positive or negative instances of variables declared earlier in the POG.
-So our POG data structure is as follows:
-\begin{lstlisting}
-structure Pog where
-  elts : Array PogElt
-  wf : ∀ i : Fin elts.size, elts[i].args_decreasing
-  inv : ∀ i : Fin elts.size, i = elts[i].varNum.natPred
-\end{lstlisting}
-The invariant \lstinline{wf} says that the graph is well founded, in the
-sense that if \lstinline{elts[i]} is a disjunction
-or a conjunction, then the variables occurring in the arguments have smaller indices.
-Note that \lstinline{Fin elts.size} is the data type consisting of natural numbers
-less than the size of the array \lstinline{elts}; our code enforces statically
-that the array accesses are within bounds.
-
-Note that we have not yet said that the corresponding formula is partitioned.
-For each POG \lstinline{P} and literal \lstinline{l}, we define
-\lstinline{P.toPropForm l} to be the
-propositional formula that arises from interpreting \lstinline{l} as a propositional
-formula, unfolding all the defined conjunctions and disjunctions. We define
-what it means for such a formula to be partitioned:
-\begin{lstlisting}
-def partitioned : PropForm ν → Prop
-  | tr         => True
-  | fls        => True
-  | var _      => True
-  | neg φ      => φ.partitioned
-  | disj φ ψ   => φ.partitioned ∧ ψ.partitioned ∧
-                    ∀ τ, ¬ (φ.eval τ ∧ ψ.eval τ)
-  | conj φ ψ   => φ.partitioned ∧ ψ.partitioned ∧ (φ.vars ∩ ψ.vars = ∅)
-  | impl _ _   => False
-  | biImpl _ _ => False
-\end{lstlisting}
-In other words, we rule out formulas with implication and bi-implication, and otherwise
-we follow the straightforward recursive definition.
-
-With these definitions in place, we can explain the specification for our checker.
-Our parser reads a CRAT file and returns an input formula \lstinline{cnf : ICnf} and a
-CRAT proof {pf : Array CratStep}, that is, an array of single proof steps:
-\begin{lstlisting}
-inductive CratStep (α ν β : Type)
-  | /-- Add asymmetric tautology. -/
-    addAt (idx : α) (C : Array β) (upHints : Array α)
-  | /-- Delete asymmetric tautology. -/
-    delAt (idx : α) (upHints : Array α)
-  | /-- Declare product operation. -/
-    prod (idx : α) (x : ν) (ls : Array β)
-  | /-- Declare sum operation. -/
-    sum (idx : α) (x : ν) (l₁ l₂ : β) (upHints : Array α)
-  | /-- Delete operation. -/
-    delOp (x : ν)
-\end{lstlisting}
-The core checker takes those as input, as well as a specification of the final node \lstinline{r},
-and either throws an exception, indicating that the proof is not well formed,
-or returns a POG \lstinline{P}.
-The specification asserts that, in the latter case, \lstinline{P.toPropForm r} is partitioned
-and is equivalent to \lstinline{cnf.toPropForm}.
-
-\subsection{The Model Counters}
-
-We have formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem,
-Definition~\ref{def:ring_evaluation}, as follows:
-\begin{lstlisting}
-def weightCount {R : Type} [CommRing R]
-    (weight : ν → R) (φ : PropForm ν) (s : Finset ν) : R :=
-  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
-\end{lstlisting}
-Here \lstinline{R} is assumed to be a commutative ring. The curly brackets around \lstinline{R} indicate that
-this argument is meant ot be left implicit, because it can inferred from the return type of
-the weight function. Similarly, the square brackets indicate that the commutative ring structure
-is a type class argument, generally inferred from the context.
-
-More precisely, the quantity \lstinline{weightCount weight φ s} adds up the weights of the models
-of \lstinline{φ} over the set of variables \lstinline{s}. We prove that as long as \lstinline{s}
-contains the variables of \lstinline{φ}, adding extra variables does not change the sum, that is, \lstinline{vars φ ⊆ s} implies
-\begin{lstlisting}
-  weightCount weight φ s = weightCount weight φ (vars φ)
-\end{lstlisting}
-The counting scheme of Proposition~\ref{prop:ring:eval} for partitioned formulas is expressed as follows:
-\begin{lstlisting}
- def countWeight (weight : ν → Rat) : PropForm ν → Rat
-  | tr       => 1
-  | fls      => 0
-  | var x    => weight x
-  | neg φ    => 1 - countWeight weight φ
-  | disj φ ψ => countWeight weight φ + countWeight weight ψ
-  | conj φ ψ => countWeight weight φ * countWeight weight ψ
-\end{lstlisting}
-We arbitrarily assign 0 to implications and bi-implications;
-recall that, by definition, these do not appear in partitioned formulas.
-Proposition~\ref{prop:ring:eval} is then formalized as follows:
-\begin{lstlisting}
-theorem countWeight_eq_weightCount (weight : ν → R) {φ : PropForm ν}
-    (hdec : partitioned φ) :
-  countWeight weight φ = weightCount weight φ (vars φ)
-\end{lstlisting}
-
-The functions just described are not meant to be computed, but,
-rather, serve as our mathematical reference.
-We have implemented an efficient function to calculate the weighted model count
-for an arbitrary node of a Pog, and we have proved that it computes the function above on the
-associated formula:
-\begin{lstlisting}
-theorem countWeight_eq_countWeight (pog : Pog) (weight : Var → Rat)
-    (x : Var) :
-  pog.countWeight weight x = (pog.toPropForm (.mkPos x)).countWeight weight
-\end{lstlisting}
-Here \lstinline{.mkPos x} denotes the positive literal for the variable \lstinline{x}.
-Applying this to the output of our verified CRAT proof checker,
-we obtain a proof that
-our toolchain computes the correct weighted model count of the input CNF.
-
-\lean{} does not yet have an efficient implementation of rational arithmetic, and
-we have not yet implemented a verified version of $\drational$,
-defined in Section~\ref{sect:experimental}.
-We have therefore implemented a separate procedure that carries out an integer
-calculation of the number of models.
-The number of models that a propositional formula has depends on what we take the
-variables to be. For example, there is only one assignment to the variables
-$\{ x, y \}$ that makes $x \wedge y$ true, but there are two assignments to the
-variables $\{ x, y, z \}$ that make $x \wedge y$ true.
-It is not hard to show that for partitioned formulas \lstinline{PropForm ν} whose
-variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
-we can count the number of models of the formula on the variables in \lstinline{s}
-recursively as follows:
-\begin{lstlisting}
-def countModels (nVars : Nat) : PropForm ν → Nat
-  | tr       => 2^nVars
-  | fls      => 0
-  | var _    => 2^(nVars - 1)
-  | neg φ    => 2^nVars - countModels nVars φ
-  | disj φ ψ => countModels nVars φ + countModels nVars ψ
-  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
-\end{lstlisting}
-We have also implemented an efficient counting function for POGs and proved that
-it computes that same quantity on the corresponding formulas.
-Finally,
-We have proved that the calculation above really does
-return the number of models:
-\begin{lstlisting}
-theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
-    (hvars : vars φ ⊆ s) (hpar :  partitioned φ) :
-  countModels (card s) φ = card (models φ s)
-\end{lstlisting}
-In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
-we have that the number of models on its variables is \lstinline{countModels φ (card (vars φ))}.
-
-\subsection{Implementation of the Proof Checker}
-
-We have specified the behavior of the proof checker in Section~\ref{appendix:proof:checker};
-in this section we sketch the implementation.
-Establishing its correctness was, by far, the most challenging part of the formalization
-process.
-For clarity, we describe the algorithm with ordinary mathematical notation rather than
-Lean snippets.
-
-We begin by reading the input CNF formula $\varphi$ and adding its clauses to a clause
-database $C$. We initialize that POG graph $G$ with the variables of $\varphi$,
-and then iteratively process each line of the CRAT proof as follows:
-\begin{itemize}
-  \item (Summarize what each step does.)
-\end{itemize}
-We establish that all the following invariants hold at each step:
-\begin{itemize}
-  \item (Wojciech has a long list of invariants)
-\end{itemize}
-At the end \ldots [Explain how the invariants at the end imply the desired conclusion.]
-
-\subsection{Verification specifics}
-\label{subsection:verification:specifics}
+\subsection{Trust}
 
@@
 and the CRAT proof. We do not verify this step.
-If one is only interested in obtaining a model count or a verified model count,
+If one is only interested in obtaining a model count or a weighted model count,
 there is no strong need to verify the parsing of the proof,
@@
 as input to our verified checker matches the one described in the file.
-An easy and fairly standard way of addressing this is to print out the CNF formula
-right after parsing it and then do a diff with the original input.
-This involves trusting only the correctness of the print procedure and diff.
+An easy and standard way of addressing this is to print out the CNF formula
+after parsing it and then do a {\tt diff} with the original input.
+This involves trusting only the correctness of the print procedure and {\tt diff}.
 Similarly, if one wants to establish the correctness of the POG contained in the CRAT file,
@@
 with dynamic arrays and uses reference counting to allow destructive updates when it is safe
-to do so [cite counting immutable beans paper].
-Code extraction also provides efficient implementations of tail recursive functions and loops
-[cite ``do unchained'' paper].
-Finally, Lean's standard library implements hashmaps in terms of array.
+to do so \cite{Ullrich:de:Moura:19}.
+Finally, Lean's standard library implements hashmaps in terms of arrays.
 Many of the basic properties of hashmaps have been formally verified, but not all.
@@
 we also have to trust that code extraction respects that foundation,
-that the implementation of arrays and hashmaps satisfy their internal descriptions,
+that the implementation of arrays and the fold operation for hashmaps
+satisfy their internal descriptions,
 and that, after parsing, the computation has the correct input formula.

### 2023-03-20 doc/lean.tex
revise Lean section

@@
 \textbf{Proof checking.} The goal of a CRAT proof is to construct a POG that is equivalent to the input CNF $\inputformula$.
-The checker begins by parsing the input formula, initializing the set of active clauses to $\theta \leftarrow \inputformula$, and the POG $P$ to an empty one.
+The checker begins by parsing the input formula, initializing the set of active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one.
 It then processes every step of the CRAT proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
@@
 Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}.
-Enforcing totality is no limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$.
+Requiring totality is not a limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$.
 We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.
 
-The invariants refer to the checker state \lstinline{st} containing fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far.
+The invariants refer to the checker state \lstinline{st} with fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far.
 For any $\nodeu\in P$, \lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.
 The first two invariants state that assignments to original variables extend uniquely to extension variables defining the POG nodes.
-In the formalization, we split this into two parts---extension and uniqueness:
+In the formalization, we split this into two parts, namely, extension and uniqueness:
 \begin{lstlisting}
@@
 \end{lstlisting}
-These invariants are maintained by valid CRAT proofs. Together with a few additional, omitted invariants
-which ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
+These are maintained by valid CRAT proofs. Together with a few additional invariants
+that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
 node $\noder$: the \textsc{final conditions} ensure that $\theta$ is logically equivalent to
 $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$
-(\lstinline{equivalent st.clauseDb (r ∧ st.pogDefsForm)}), and from the invariants we can prove that
+(\lstinline{equivalent st.clauseDb (r ∧ st.pogDefsForm)}), and from the invariants we can prove
 $\inputformula \iff \phi_\noder$ (\lstinline{equivalent st.inputCnf (st.pog.toPropForm r)}).
 After certifying a proof, the checker can pass its in-memory POG representation to the ring evaluator.
+% JA: I cut a few words here to save a line.
 
@@
 \begin{lstlisting}
-def weightCount {R : Type} [CommRing R]
+def weightSum {R : Type} [CommRing R]
     (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
@@
 \begin{lstlisting}
-theorem ringEval_eq_weightCount (weight : Var → R) {φ : PropForm Var} :
-    partitioned φ → ringEval weight φ = weightCount weight φ (vars φ)
+theorem ringEval_eq_weightSum (weight : Var → R) {φ : PropForm Var} :
+    partitioned φ → ringEval weight φ = weightSum weight φ (vars φ)
 \end{lstlisting}
@@
 
-Applying this to the output of our verified CRAT proof checker which we know to be partitioned
+Applying this to the output of our verified CRAT proof checker, which we know to be partitioned
 and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes
@@
 \textbf{Model counting.} Finally, we established that ring evaluation with the appropriate weights
-corresponds to the standard model count. To this end we specified \lstinline{countModels}
-which directly carries out an integer calculation of the number of models over a set of variables
+corresponds to the standard model count. To do so, we defined a function that
+% JA: deleted the word "directly" to save a line
+carries out an integer calculation of the number of models over a set of variables
 of cardinality \lstinline{nVars}:

### 2023-03-20 doc/mc-paper.tex
mention clause deletions in related work

@@
 is partitioned and is equivalent to the input.
-This requires verifying additional proof steps and subtle invariants, as described in
+This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
 Sections~\ref{sect:crat} and \ref{section:formally-verified-toolchain}.

### 2023-03-20 doc/lean.tex
clean up code; weightCount -> weightSum, etc.

@@
 \end{lstlisting}
-
 To efficiently compute the ring evaluation of a formula represented by a POG node, we implemented
-\lstinline{Pog.ringEval}. We then proved that it matches the specification above:
+\lstinline{Pog.ringEval} and then proved that it matches the specification above:
 \begin{lstlisting}
@@
 \end{lstlisting}
-
 Applying this to the output of our verified CRAT proof checker, which we know to be partitioned

### 2023-03-20 doc/mc-paper.tex
Tweaked intro

@@
 compilation and model counting toolchain.  Starting with
-input formula $\inputformula$, the \dfour{} knowledge compiler
+input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
 generates a d-DNNF representation, and the {\em proof
@@
 formulation of clausal proof systems for which proofs can readily be
-generated and efficiently (and, in some cases, formally) checked
-\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress
-has also been made on extending these proof systems to other domains,
-including quantified Boolean
-formulas~\cite{bryant:cade:2021,heule:JAR2014}.
+generated and efficiently checked
+\cite{RAT,wetzler14_drattrim}.
+A number formally verified checkers for this task have been developed within different frameworks~\cite{cruz-cade-2017,Heule:et:al:17,Lammich:20,Tan:2021}.
+The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
+By contrast, our proofs construct a new propositional formula, and we must verify both that it
+is partitioned and that it is equivalent to the input formula.
+This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
+Sections~\ref{sect:crat} and \ref{section:formally-verified-toolchain}.
 
@@
 proof framework for model counting programs.  Their proof rules are
-based on the algorithmic steps commonly used by model counting
-programs, and they present a series of lemmas and a theorem stating
+based on the algorithms commonly used by model counters.
+They present a series of lemmas and a theorem stating
 that their checking framework is sound.  They also present
@@
 %%proof framework or created a formally verified checker.
-
 Our toolchain is not directly comparable to theirs---they directly
@@
 certify that the graph generated by a knowledge compiler is logically
-equivalent to the input formula.  We can also validate the standard or
-weighted count generated from that representation.
+equivalent to the input formula.  We can then validate a standard or
+weighted count from that representation.
 %% Their tool
@@
 experimental evaluation in Appendix~\ref{app:experiments}.
-
-Verified checkers for unsatisfiability include \cite{Cruz:Filipe:et:al:16,Heule:et:al:17,Lammich:20,Tan:2021}.
-The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
-In contrast, our proofs construct a new propositional formula, and we have to verify that it
-is partitioned and is equivalent to the input.
-This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
-Sections~\ref{sect:crat} and \ref{section:formally-verified-toolchain}.
 

### 2023-03-20 doc/mc-paper.tex
Edits

@@
 \cite{RAT,wetzler14_drattrim}.
-A number formally verified checkers for this task have been developed within different frameworks~\cite{cruz-cade-2017,Heule:et:al:17,Lammich:20,Tan:2021}.
+A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
 The associated proofs add clauses while preserving satisfiability until the empty clause is derived.

### 2023-03-21 doc/mc-paper.tex
Polishing

@@
 We know of no other
-work addressing the question of certifying the output of a knowledge
-compiler.
+work on certifying the output of a knowledge compiler.
 
@@
     with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{i=1 \leq i \leq k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{i=1 \leq i \leq k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcap_{i=1 \leq i \leq k} \modelset(\phi_i)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{i=1 \leq i \leq k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{i=1 \leq i \leq k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcup_{i=1 \leq i \leq k} \modelset(\phi_i)$.
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
+      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
+      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
+      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
+      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
   \end{enumerate}
@@
   in {\em negation normal form} when negation is applied only to variables.  A
-  formula is in {\em conjunctive normal form} (CNF) when 1) it is in
-  negation normal form, and 2) sum is applied only to literals.  A CNF
+  formula is in {\em conjunctive normal form} (CNF) when i) it is in
+  negation normal form, and ii) sum is applied only to literals.  A CNF
   formula can be represented as a set of {\em clauses}, each of which is a
@@
       \item The arguments to each product must have disjoint dependency sets.  That is, operation
-        $\bigwedge_{i=1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
+        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
       \item The arguments to each sum must have disjoint models.  That is, operation
-        $\bigvee_{i=1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
+        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
       \end{enumerate}
@@
   For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
-  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $\mulident-w(x)$.
-
-  For Boolean formula $\phi$, and ring weight function $w$, the {\em ring evaluation problem} is to compute:
+  every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.
+
+  For Boolean formula $\phi$ and ring weight function $w$, the {\em ring evaluation problem} computes
   \begin{eqnarray}
@@
 randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
+For example, it can compare different runs on a single formula,
+either from different compilers or from a single compiler with different configuration parameters.
 
@@
 \textstyle
-\scalebox{1.2}{\rep}\left(\Pand_{i=1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
+\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
 \textstyle
-\scalebox{1.2}{\rep}\left(\Por_{i=1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
-%% V2
-%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
-%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
-%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
-%\rep\left(\Pand_{i=1 \leq i \leq k} \phi_i, w \right) &=& \prod_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
-%\rep\left(\Por_{i=1 \leq i \leq k} \phi_i, w\right) &=& \sum_{i=1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
+\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
 \end{eqnarray}
@@
 It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
-Every POG must have a designated {\em root node} $\noder$.
+Every POG has a designated {\em root node} $\noder$.
 
@@
 
-The combination of these two optimization guarantees that 1) each call
+The combination of these two optimization guarantees that i) each call
 to $\validate$ for a product node will cause at most one invocation of
-the SAT solver, and 2) each call to $\validate$ for any node $\nodeu$
+the SAT solver, and ii) each call to $\validate$ for any node $\nodeu$
 will cause further recursive calls only once.  Our experimental
@@
 %% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
-%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{i=1 \leq i \leq k}
+%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{1 \leq i \leq k}
 %% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$

### 2023-03-21 doc/mc-paper.tex
Minor edits

@@
 the \dfour{} knowledge compiler, form the toolchain illustrated in
-Figure~\ref{fig:chain}.  The proof generator is the same in both
+Figure~\ref{fig:chain}\footnote{All tools, along with a set of 
+benchmark problems are available at \url{https://github.com/rebryant/crat}}.  The proof generator is the same in both
 cases, since it need not be trusted.
@@
 decimal arithmetic, handling the weights used in the weighted model
-counting competitions.  For one of the weighted benchmarks, it
-generated a result with 260,909 decimal digits.
+counting competitions.  To give a sense of scale, the counter
+generated a result with 260,909 decimal digits
+for one of the weighted benchmarks.
 
@@
 
-We provide only a summary of our experiments here.  A more complete
-description is provided in Appendix~\ref{app:experiments}\@. For our
+We summarize the results of our experiments here.  
+Appendix~\ref{app:experiments} provides a more complete description. For our
 evaluation, we used the public benchmark problems from the 2022
@@
 %checker.)
-We then ran our proof generator with a time limit of 10,000 seconds.
+We ran our proof generator with a time limit of 10,000 seconds.
 It was able to generate full proofs for 108 of the problems and
@@
 generator and checker could improve their performance and capacity substantially.
+
+\newpage
 

### 2023-03-21 doc/mc-paper.tex
chore: emails and funding

@@
 \author{Randal E.
-Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{rebryant@cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}
-\author{Wojciech Nawrocki}{Department of Philosophy, Carnegie Mellon University}{wjnawrock@cmu.edu}{https://orcid.org/0000-0002-8839-0618}{}
-\author{Jeremy Avigad}{Department of Philosophy, Carnegie Mellon University}{avigad@cmu.edu}{https://orcid.org/0000-0003-1275-315X}{}
+Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{Randy.Bryant@cs.cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}
+\author{Wojciech Nawrocki}{Department of Philosophy, Carnegie Mellon University}{wjnawrocki@cmu.edu}{https://orcid.org/0000-0002-8839-0618}{Hoskinson Center for Formal Mathematics}
+\author{Jeremy Avigad}{Department of Philosophy, Carnegie Mellon University}{avigad@cmu.edu}{https://orcid.org/0000-0003-1275-315X}{Hoskinson Center for Formal Mathematics}
 \author{Marijn J. H. Heule}{Computer Science Department, Carnegie Mellon University}{marijn@cmu.edu}{https://orcid.org/0000-0002-5587-8801}{Supported by NSF grant CCF-2108521}

### 2023-03-21 doc/lean.tex
save some lines in Lean

@@
 The first two invariants state that assignments to original variables extend uniquely to extension variables defining the POG nodes.
-In the formalization, we split this into two parts, namely, extension and uniqueness:
+In the formalization, we split this into extension and uniqueness:
 \begin{lstlisting}
@@
   ∀ (σ₁ : PropAssignment Var), σ₁ ⊨ φ₁ → ∃ σ₂, σ₁.agreeOn X σ₂ ∧ σ₂ ⊨ φ₂
-
-invariants.extends_pogDefsForm : extendsOver st.inputCnf.vars ⊤ st.pogDefsForm
 
@@
 
+invariants.extends_pogDefsForm : extendsOver st.inputCnf.vars ⊤ st.pogDefsForm
 invariants.uep_pogDefsForm : uniqueExt st.inputCnf.vars st.allVars st.pogDefsForm
@@
 
-invariants.equivalent_clauseDb : equivalentOver st.inputCnf.vars
-  st.inputCnf st.clauseDb
+invariants.equivInput : equivalentOver st.inputCnf.vars st.inputCnf st.clauseDb
 \end{lstlisting}
@@
 invariants.partitioned : ∀ u : ILit, (st.pog.toPropForm u).partitioned
-
 invariants.equivalent_lits : ∀ u : ILit, equivalentOver st.inputCnf.vars
@@
 In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
-we have that the number of models on its variables is \lstinline{countModels φ (card (vars φ))}.
+we have that the number of models of \lstinline{φ} on its variables is
+\lstinline{countModels φ (card (vars φ))}.

### 2023-03-21 doc/mc-paper.tex
more theta/phi corrections

@@
 
-This paper introduces \emph{Partitioned Operation Graphs} (POGs),
+This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
 a generalization of the representations produced by current knowledge
@@
 The status of the proof at any step is represented as
-a set of {\em active} clauses $\phi$, i.e., those that
+a set of {\em active} clauses $\theta$, i.e., those that
 have been added but not yet deleted.
@@
 Let $Z$
-denote the set of extension variables occuring in formula $\phi$.
+denote the set of extension variables occuring in formula $\theta$.
 Starting with input formula $\inputformula$,
 the proof must maintain the invariant that
-$\inputformula \ifandonlyif \exists Z\,\phi$.
+$\inputformula \ifandonlyif \exists Z\,\theta$. We write $\theta_\text{foo}$ for
+clausal encodings possibly using extension variables, and $\phi_\text{bar}$ for formulas that do not.
 
@@
 deleted based on {\em implication redundancy}.  That is, when clause
-$C$ satisfies $\phi \imply C$ for formula $\phi$, then it can either
-be added to $\phi$ to create the formula $\phi \cup \{C\}$ or it can be deleted
-from $\phi \cup \{C\}$ to create $\phi$.
+$C$ satisfies $\theta \imply C$ for formula $\theta$, then it can either
+be added to $\theta$ to create the formula $\theta \cup \{C\}$ or it can be deleted
+from $\theta \cup \{C\}$ to create $\theta$.
 
@@
 Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
-proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
-{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
+proved redundant with respect to formula $\theta$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
+{\em antecedent} clauses, such that each $D_i$ is in $\theta$.
 A RUP step
@@
 total assignment $\alpha$ that falsifies the clause must
-assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
+falsify $\phi_\noder$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
 that

### 2023-03-21 doc/lean.tex
changes to abstract and foo/bar

@@
   ∀ (σ₁ : PropAssignment Var), σ₁ ⊨ φ₁ → ∃ σ₂, σ₁.agreeOn X σ₂ ∧ σ₂ ⊨ φ₂
-
 /-- Assignments satisfying φ are determined on Y by their values on X. -/

### 2023-03-21 doc/mc-paper.tex
changes to abstract and foo/bar

@@
 
-\emph{Knowledge compilation} is the process of converting a Boolean
-formula into a form that makes it possible to efficiently calculate
-associated values, such as the weighted or unweighted model
-count. Existing knowledge compilers
-provide no guarantee that the representation they produce
-correctly represents the original formula, and therefore they cannot validate the generated counts.
-
-We present \emph{Partitioned Operation Graphs} (POGs), a generalization
+Computing certain properties of Boolean formulas, such as their weighted or unweighted model count,
+is intractable on general representations. It can become tractable when formulas are expressed in a
+special form, such as d-DNNF. \emph{Knowledge compilation} is the process of converting a formula
+into such a form. Existing knowledge compilers provide no guarantee that their output correctly
+represents the original formula, and therefore they cannot validate the resulting model counts.
+
+We present \emph{Partitioned-Operation Graphs} (POGs), a generalization
 of the representations used by existing knowledge compilers.
-We have designed a proof framework that can prove the equivalence of a
+We have designed CRAT, a proof framework that can prove the equivalence of a
 POG representation with a Boolean formula in conjunctive normal form (CNF),
@@
 the d-DNNF graphs produced by the state-of-the-art knowledge compiler
-\dfour, as well as checkable proofs that the output POGs
-are equivalent to the input CNF formulas.  We have also implemented an
+\dfour{}, as well as checkable CRAT proofs certifying that the output POGs
+are equivalent to the input CNF formulas. We have implemented an
 efficient checker for these proofs. Our toolchain
@@
 graphs produced by \dfour{} for formulas from a recent model counting
-competition. We also present a checker and a model counter for POGs that have
-been formally verified in the \lean{} proof assistant.  These programs both prove the soundness of our proof framework and
-create the first formally verified toolchain for weighted and unweighted model counting.
+competition. Additionally, we have developed a formally verified CRAT
+checker and model counter for POGs in the \lean{} proof assistant.
+In doing so, we proved the soundness of our proof framework. These programs
+comprise the first formally verified toolchain for weighted and unweighted
+model counting.
 \end{abstract}
@@
 the proof must maintain the invariant that
-$\inputformula \ifandonlyif \exists Z\,\theta$. We write $\theta_\text{foo}$ for
-clausal encodings possibly using extension variables, and $\phi_\text{bar}$ for formulas that do not.
+$\inputformula \ifandonlyif \exists Z\,\theta$. We write (subscripted) $\theta$ for
+clausal encoding formulas possibly using extension variables, and (subscripted) $\phi$ for formulas
+that do not.
 

### 2023-03-21 doc/leanappendix.tex
clarify Lean appendix

@@
 allowing us to state our specifications and prove that they are met.
-Our representation of CNF formulas is among the former:
+Among the former is our representation of CNF formulas:
 a literal is represented as a nonzero integer, a clause is an array of literals,
@@
 \lstinline{var x} is the index
-of the variable; in \lstinline{disj x left right} and \lstinline{conj x args}
-it is the definition number in the CRAT file. Note that representing edges as literals allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
+of an original variable; in \lstinline{disj x left right} and \lstinline{conj x args}
+it is an extension variable appearing in the CRAT file. Note that representing edges as literals
+allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
 A \lstinline{Pog} is then an array of \lstinline{PogElt}
-that is well founded in the sense that each element depends only on prior elements
+that is well-founded in the sense that each element depends only on prior elements
 in the array.
@@
 relate our data structures to propositional formulas and their semantics.
-For example, given a literal \lstinline{l}, \lstinline{Pog.toPropForm P l}
-denotes the (unnegated or negated) interpretation of the node of the POG
-\lstinline{P} specified by \lstinline{l} as a propositional formula over the
+For example, given a literal \lstinline{u}, \lstinline{Pog.toPropForm P u}
+denotes the (unnegated or negated) interpretation of the node corresponding to \lstinline{u} in the POG
+\lstinline{P} as a propositional formula $\phi_\nodeu/\neg\phi_\nodeu$ over the
 original variables.
 Lean provides us with convenient ``anonymous projector'' notation that allows
-us to write \lstinline{P.toPropForm l} instead of \lstinline{Pog.toPropForm P l}
+us to write \lstinline{P.toPropForm u} instead of \lstinline{Pog.toPropForm P u}
 when \lstinline{P} has type \lstinline{Pog},
@@
 as the \emph{Lindenbaum--Tarski algebra}.
-We defined the quotient and lifted the boolean operations and the entailment
+We defined the quotient and lifted the Boolean operations and the entailment
 relation to this new type. The advantage is that equivalent formulas give
-rise to equal elements in the quotient, which
-made is much easier to prove equivalences involving complicated expressions,
-since it enabled us to calculate with the associated
-elements of the quotient and substitute one for another when the corresponding formulas are equivalent. This gave rise to auxiliary challenges, however.
+rise to equal elements in the quotient.
+This makes it much easier to prove equivalences involving complicated expressions,
+since it enables us to substitute elements of the quotient for others even when the corresponding
+formulas are merely equivalent but not equal. This gave rise to auxiliary challenges, however.
 For example, it is no longer straightforward to say that an element of the quotient ``depends'' on a variable,
-since equivalent formulas can have difference variables.
-Instead, we made use of a semantic notion of dependence, in which an element of the
-quotient depends on a variable if and only if for some truth assignment its value
+since equivalent formulas can have different variables---consider $x \vee \neg x$ and $\top$.
+Instead, we made use of a semantic notion of dependence, in which an element $\phi$ of the
+quotient depends on a variable if and only if for some truth assignment the value of $\phi$
 changes when the assignment to that variable is flipped.
@@
 
-In this subsection, we clarify what has been verified and what has been trusted.
-Recall that our first step is to parse a CRAT file to read in the initial CNF formula
-and the CRAT proof. We do not verify this step.
-If one is only interested in obtaining a model count or a weighted model count,
-there is no strong need to verify the parsing of the proof,
-because if the checker succeeds in constructing an equivalent POG,
-it does not really matter whether the proof was the one the user intended.
-But it would be nice to have reassurance that the CNF formula that serves
-as input to our verified checker matches the one described in the file.
-An easy and standard way of addressing this is to print out the CNF formula
-after parsing it and then do a {\tt diff} with the original input.
-This involves trusting only the correctness of the print procedure and {\tt diff}.
-Similarly, if one wants to establish the correctness of the POG contained in the CRAT file,
-one can print out the POG that is constructed by the checker and compare.
+In this subsection, we clarify what has been verified and what has to be trusted.
+Recall that our first step is to parse CNF and CRAT files in order to read in the initial
+formula and the proof. We do not verify this step. Instead, the verified checker exposes
+flags \verb+--print-cnf+ and \verb+--print-crat+ which reprint the consumed formula or proof,
+respectively. Comparing this to the actual files using {\tt diff} provides an easy way of ensuring
+that what was parsed matches their contents. This involves trusting only the correctness of the
+print procedure and {\tt diff}. Similarly, if one wants to establish the correctness of the POG
+contained in the CRAT file, one can print out the POG that is constructed by the checker and compare.
 
-Lean's code extraction replaces calculations with natural numbers and integers with
+Lean's code extraction replaces calculations on natural numbers and integers with
 efficient but unverified arbitrary precision versions.
@@
 
-In sum, in addition to trusting Lean's foundation and kernel checker,
+In summary, in addition to trusting Lean's foundation and kernel checker,
 we also have to trust that code extraction respects that foundation,
-that the implementation of arrays and the fold operation for hashmaps
-satisfy their internal descriptions,
-and that, after parsing, the computation has the correct input formula.
-All of our specifications have been completely proved and verified relative to that.
+that the implementation of the fold operation for hashmaps satisfies its description,
+and that, after parsing, the computation uses the correct input formula.
+All of our specifications have been completely proven and verified relative to these assumptions.

### 2023-03-22 doc/lean.tex
CRAT --> CPOG

@@
 
-We set out to formally verify the system with two goals in mind: first, to ensure that the CRAT framework is mathematically
+We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
 sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \cref{fig:chain}).
@@
 \vspace{1em}\noindent
-\textbf{Proof checking.} The goal of a CRAT proof is to construct a POG that is equivalent to the input CNF $\inputformula$.
+\textbf{Proof checking.} The goal of a CPOG proof is to construct a POG that is equivalent to the input CNF $\inputformula$.
 The checker begins by parsing the input formula, initializing the set of active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one.
-It then processes every step of the CRAT proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
+It then processes every step of the CPOG proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
 Afterwards, it carries out the \textsc{final conditions} check of \cref{subsection:semantics}.
@@
 \end{lstlisting}
-These are maintained by valid CRAT proofs. Together with a few additional invariants
+These are maintained by valid CPOG proofs. Together with a few additional invariants
 that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
@@
 \end{lstlisting}
-Applying this to the output of our verified CRAT proof checker, which we know to be partitioned
+Applying this to the output of our verified CPOG proof checker, which we know to be partitioned
 and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes

### 2023-03-22 doc/leanappendix.tex
CRAT --> CPOG

@@
 of an original variable; in \lstinline{disj x left right} and \lstinline{conj x args}
-it is an extension variable appearing in the CRAT file. Note that representing edges as literals
+it is an extension variable appearing in the CPOG file. Note that representing edges as literals
 allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
@@
 In this subsection, we clarify what has been verified and what has to be trusted.
-Recall that our first step is to parse CNF and CRAT files in order to read in the initial
+Recall that our first step is to parse CNF and CPOG files in order to read in the initial
 formula and the proof. We do not verify this step. Instead, the verified checker exposes
-flags \verb+--print-cnf+ and \verb+--print-crat+ which reprint the consumed formula or proof,
+flags \verb+--print-cnf+ and \verb+--print-cpog+ which reprint the consumed formula or proof,
 respectively. Comparing this to the actual files using {\tt diff} provides an easy way of ensuring
@@
 print procedure and {\tt diff}. Similarly, if one wants to establish the correctness of the POG
-contained in the CRAT file, one can print out the POG that is constructed by the checker and compare.
+contained in the CPOG file, one can print out the POG that is constructed by the checker and compare.
 

### 2023-03-22 doc/mc-data.tex
Final Lean data point
CRAT --> CPOG

@@
      \end{itemize}
-   \item Ran program {\sc d2p} to generate CRAT file for maximum of 10,000 seconds
+   \item Ran program {\sc d2p} to generate CPOG file for maximum of 10,000 seconds
      \begin{itemize}
@@
      \end{itemize}
-   \item For completed CRAT files, ran {\sc crat-check} CRAT checker
+   \item For completed CPOG files, ran {\sc cpog-check} CPOG checker
      \begin{itemize}
-     \item Tests all conditions for correct CRAT
+     \item Tests all conditions for correct CPOG
      \item Most time spent performing RUP checks for clause addition and deletion
@@
 
-\section*{Time to Generate and Validate CRAT}
+\section*{Time to Generate and Validate CPOG}
 
@@
 
-\section*{Generated CRAT Files}
+\section*{Generated CPOG Files}
 
@@
 
-Two optimizations were implemented within the CRAT framework:
+Two optimizations were implemented within the CPOG framework:
 \begin{description}
@@
 performed was limited to $10^7$.  We then made three additional runs
-of the CRAT generator and checker, with each of the two optimizations
+of the CPOG generator and checker, with each of the two optimizations
 disabled individually and in combination.  Figures
@@
 The time performance is less dramatic, but still significant.
-Comparing the sum of CRAT generation and checking times for each of
+Comparing the sum of CPOG generation and checking times for each of
 the four combinations of options, we get the following times relative
@@
 
-Here is another way to summarize the impact on the combined time to generate and validate the CRAT file:
+Here is another way to summarize the impact on the combined time to generate and validate the CPOG file:
 \begin{itemize}
@@
 
-\input{figure-mice-crat}
+\input{figure-mice-cpog}
 
@@
 error, but it possibly timed out.  The time-out cases are shown along
-the upper edge (MICE) or the righthand edge (CRAT).
+the upper edge (MICE) or the righthand edge (CPOG).
 
@@
 \item 76 were completed by both programs in under 1000 seconds
-\item Of those 21 were faster with MICE, 55 with CRAT
-\item 7 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
-\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
+\item Of those 21 were faster with MICE, 55 with CPOG
+\item 7 completed with CPOG under 1000 seconds but exceeded that threshold for MICE
+\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CPOG
 \end{itemize}
@@
 \item 80 total
-\item Had 2 that failed RUP check (need to investigate)
-\item One of them needs to be rerun (Very long generation time)
-\item Two timed out for \lean{}
+\item All ran to completion
+\item All but three ran within factor of $8\times$.  Outliers were small ones
+\item Harmonic mean = $5.94$.
 \end{itemize}

### 2023-03-22 doc/mc-paper.tex
CRAT --> CPOG

@@
 
-Computing certain properties of Boolean formulas, such as their weighted or unweighted model count,
+Computing many useful properties of Boolean formulas, such as their weighted or unweighted model count,
 is intractable on general representations. It can become tractable when formulas are expressed in a
-special form, such as d-DNNF. \emph{Knowledge compilation} is the process of converting a formula
+special form, such as a deterministic,
+  decomposable, negation normal form (d-DNNF) graph.
+\emph{Knowledge compilation} is the process of converting a formula
 into such a form. Existing knowledge compilers provide no guarantee that their output correctly
-represents the original formula, and therefore they cannot validate the resulting model counts.
+represents the original formula, and therefore they cannot validate the model counts.
 
@@
 of the representations used by existing knowledge compilers.
-We have designed CRAT, a proof framework that can prove the equivalence of a
-POG representation with a Boolean formula in conjunctive normal form (CNF),
-the standard input format for SAT solvers and model counting tools.
+We have designed  CPOG, a framework that can express proofs of equivalence between a
+POG  and a Boolean formula in conjunctive normal form (CNF).
 
@@
 the d-DNNF graphs produced by the state-of-the-art knowledge compiler
-\dfour{}, as well as checkable CRAT proofs certifying that the output POGs
-are equivalent to the input CNF formulas. We have implemented an
-efficient checker for these proofs. Our toolchain
+\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
+are equivalent to the input CNF formulas.  Our toolchain
 for generating and verifying POGs scales to all but the largest d-DNNF
 graphs produced by \dfour{} for formulas from a recent model counting
-competition. Additionally, we have developed a formally verified CRAT
+competition. Additionally, we have developed a formally verified CPOG
 checker and model counter for POGs in the \lean{} proof assistant.
@@
 and a checkable proof of its logical equivalence to a Boolean formula in
-conjunctive normal form (CNF), we propose the CRAT file format. A CRAT
+conjunctive normal form (CNF), we propose the CPOG (for ``checkable'' POG) file format. A CPOG
 proof consists of a sequence of clause addition and deletion steps,
 based on an extended resolution proof system~\cite{Tseitin:1983}.
-We establish a set of conditions that a CRAT file must satisfy in order to
+We establish a set of conditions that a CPOG file must satisfy in order to
 encode a well-formed POG and provide a valid equivalence proof.
@@
 
-  %% CRAT
+  %% CPOG
   \draw[thick] (40,12) [-latex] -- (50,12);
   \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
-  \node [above] at (44,12) {\texttt{.crat}};
+  \node [above] at (44,12) {\texttt{.cpog}};
 
@@
 \caption{Certifying toolchain.
-  The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT)
+  The output of a standard knowledge compiler is converted into a combined graph/proof (CPOG)
   which can be independently checked and evaluated}
@@
 generates a d-DNNF representation, and the {\em proof
-  generator} uses this to generate a CRAT file.
-The {\em proof checker} verifies the equivalence of the CNF and CRAT representations.
+  generator} uses this to generate a CPOG file.
+The {\em proof checker} verifies the equivalence of the CNF and CPOG representations.
 The {\em ring evaluator} computes
@@
 
-To ensure soundness of the abstract CRAT proof system, as well as correctness of its
+To ensure soundness of the abstract CPOG proof system, as well as correctness of its
 concrete implementation, we formally verified the proof system as well as versions
 of the proof checker and ring evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
-Running these two programs on a particular CRAT file gives assurance that the proof
+Running these two programs on a particular CPOG file gives assurance that the proof
 and the model count are correct. Our experience with developing a formally verified
@@
 This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
-Sections~\ref{sect:crat} and \ref{section:formally-verified-toolchain}.
+Sections~\ref{sect:cpog} and \ref{section:formally-verified-toolchain}.
 
@@
 
+We write (possibly subscripted) $\theta$ for
+formulas encoded as clauses, possibly with extension variables.
+We write (possibly subscripted)
+$\phi$ for formulas
+that use no extension variables.
+
 A proof in our framework takes the form of a sequence of clause addition
@@
 the proof must maintain the invariant that
-$\inputformula \ifandonlyif \exists Z\,\theta$. We write (subscripted) $\theta$ for
-clausal encoding formulas possibly using extension variables, and (subscripted) $\phi$ for formulas
-that do not.
+$\inputformula \ifandonlyif \exists Z\,\theta$.
 
@@
 occurs only when defining $\pand$ and $\por$ operations, as is
-described in Section~\ref{sect:crat}.  Clauses can also be added or
+described in Section~\ref{sect:cpog}.  Clauses can also be added or
 deleted based on {\em implication redundancy}.  That is, when clause
@@
 By contrast, an unsatisfiability proof need only guarantee that
-no proof step cause a satisfiable set of clauses to become
+no proof step causes a satisfiable set of clauses to become
 unsatisfiable.  As a consequence, our proofs must justify clause deletions as well as clause additions.
 
-\section{The CRAT POG Representation and Proof System}
-\label{sect:crat}
+\section{The CPOG POG Representation and Proof System}
+\label{sect:cpog}
 
@@
 %% \begin{itemize}
-%% \item The {\em proof generator} produces a CRAT file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
+%% \item The {\em proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
 %% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
@@
 
-A CRAT file provides both a declaration of a POG, as well as a checkable
+A CPOG file provides both a declaration of a POG, as well as a checkable
 proof that a Boolean formula, given in conjunctive normal
@@
 \begin{table}
-  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
-  \label{tab:crat:syntax}
+  \caption{CPOG Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
+  \label{tab:cpog:syntax}
 \centering{
@@
 
-Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
+Table~\ref{tab:cpog:syntax} shows the declarations that can occur in a CPOG file.
 %% The checker is provided with the input formula as a separate file.
@@
 
-%% A CRAT proof follows the same general form as a QRAT dual
+%% A CPOG proof follows the same general form as a QRAT dual
 %% proof~\cite{bryant:cade:2021}---one that ensures that each clause
-%% addition and each clause deletion preserves equivalence.  With CRAT,
+%% addition and each clause deletion preserves equivalence.  With CPOG,
 %% however, clauses are defined both explicitly and implicitly.  Starting
@@
 
-Appendix~\ref{app:crat:example} shows the CRAT description for
+Appendix~\ref{app:cpog:example} shows the CPOG description for
 an input formula with five clauses, yielding a POG with five
@@
 
-\section{Generating CRAT from d-DNNF}
-\label{section:generating:crat}
-
-%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
+\section{Generating CPOG from d-DNNF}
+\label{section:generating:cpog}
+
+%%Figure~\ref{fig:chain} illustrates one method of generating a CPOG
 %%file from an input formula in conjunctive normal form.  Our toolchain
@@
 of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
-For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
+For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CPOG proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
 Our experimental results show
@@
   SAT solver, and then converting the generated unsatisfiability proof
-  into a sequence of clause additions in the CRAT file.
+  into a sequence of clause additions in the CPOG file.
 \item For a single nonterminal child ($k = 1$), we recursively call
@@
 $\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
-CRAT framework by first deleting all asserted clauses, except for the
+CPOG framework by first deleting all asserted clauses, except for the
 final unit clause for root literal $r$, and then deleting all of the
@@
 a solver with $v$ set to 0 yields an unsatisfiability proof that can
-be mapped back to a sequence of clause additions in the CRAT file validating
+be mapped back to a sequence of clause additions in the CPOG file validating
 all of the literals.
@@
 Figure~\ref{fig:chain}\footnote{All tools, along with a set of 
-benchmark problems are available at \url{https://github.com/rebryant/crat}}.  The proof generator is the same in both
+benchmark problems are available at \url{https://github.com/rebryant/cpog}}.  The proof generator is the same in both
 cases, since it need not be trusted.
@@
 \cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
-proof steps back into hinted CRAT clause additions, the generator can
+proof steps back into hinted CPOG clause additions, the generator can
 use either its own RUP proof generator, or it can invoke
@@
 
-Finally, the runtimes for the MICE toolchain versus our CRAT toolchain
+Finally, the runtimes for the MICE toolchain versus our CPOG toolchain
 showed little correlation, reflecting the fact that the two solve
 different problems and use different approaches.  In general,
-our CRAT toolchain showed better scaling, in part due to its ability
+our CPOG toolchain showed better scaling, in part due to its ability
 to control the recursion through lemmas.
@@
 
-\section{CRAT Example}
-\label{app:crat:example}
+\section{CPOG Example}
+\label{app:cpog:example}
 
@@
 \toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & Explanation \\
+\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
 \midrule
@@
 \begin{minipage}{0.49\textwidth}
-(E) CRAT Assertions\\[1.2ex]
+(E) CPOG Assertions\\[1.2ex]
 \begin{tabular}{llllll}
@@
   \toprule
- \multicolumn{3}{l}{CRAT line} & Explanation\\
+ \multicolumn{3}{l}{CPOG line} & Explanation\\
 \midrule
@@
 \end{minipage}
-\caption{Example formula (A), its POG representation (B), and its CRAT proof (C), (E), and (F)}
+\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F)}
 \label{fig:eg4:all}
@@
 Figure \ref{fig:eg4:all} illustrates an example formula and shows how
-the CRAT file declares its POG representation.  The input formula (A)
+the CPOG file declares its POG representation.  The input formula (A)
 consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
@@
 % We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
-  The first part of the CRAT file (C) declares
+  The first part of the CPOG file (C) declares
 these nodes using clause IDs that increment by three or four,
@@
 
-We step through portions of the file to provide a better understanding of the CRAT proof framework.
+We step through portions of the file to provide a better understanding of the CPOG proof framework.
 Figure
@@
 defined by the POG operation declarations.  These do not appear in the
-CRAT file.  Referring back to the declarations of the sum nodes in
+CPOG file.  Referring back to the declarations of the sum nodes in
 Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
@@
   $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
-  The proof steps can be mapped back into clause addition steps in the CRAT file, incorporating the
+  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
   input clauses and the defining clauses for $\nodev$ into the hints.
@@
 The challenge for implementing this strategy is to find a way to
-represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CRAT file.  Some may be
+represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CPOG file.  Some may be
 unaltered input clauses, and these can be used directly.  Others,
 however will be clauses that do not appear in the input formula.  We
-implement these by adding POG product nodes to the CRAT file to create
+implement these by adding POG product nodes to the CPOG file to create
 the appropriate clauses.  Consider an {\em argument} clause
@@
 \\[1.0em]
-(C) CRAT Assertions\\[1.0em]
+(C) CPOG Assertions\\[1.0em]
 \begin{tabular}{llllll}
@@
 \item
-  whether or toolchain can handle challenging benchmark problems,
+  whether our toolchain can handle challenging benchmark problems,
 \item
@@
 
-Figure~\ref{fig:d4:crat} summarizes our results in terms of the time
+Figure~\ref{fig:d4:cpog} summarizes our results in terms of the time
 required by \dfour{} (X axis) versus the sum of the times for the
@@
 We ran the \lean{} checker on the 80 benchmark problems for which the
-generated CRAT file had at most 10~million total clauses.  These are the
+generated CPOG file had at most 10~million total clauses.  These are the
 same benchmarks used in evaluating the optimizations (Section
@@
 MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
-generation, proof checking, and ring evaluation in the CRAT toolchain compares to the time to
+generation, proof checking, and ring evaluation in the CPOG toolchain compares to the time to
 run their proof generator \progname{nnf2trace} and checker
@@
 \item 76 were completed by both programs in under 1000 seconds
-\item Of those 21 were faster with MICE, 55 with CRAT
-\item 7 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
-\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
+\item Of those 21 were faster with MICE, 55 with CPOG
+\item 7 completed with CPOG under 1000 seconds but exceeded that threshold for MICE
+\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CPOG
 \end{itemize}
 
-\input{figure-mice-crat} The detailed results are shown in
+\input{figure-mice-cpog} The detailed results are shown in
 Figure~\ref{fig:mice}.  As can be seen, the comparative runtimes are

### 2023-03-22 doc/mc-paper.tex
Describe soundness, not completeness

@@
 is intractable on general representations. It can become tractable when formulas are expressed in a
-special form, such as a deterministic,
-  decomposable, negation normal form (d-DNNF) graph.
+special form, such as the deterministic,
+decomposable, negation normal form (d-DNNF)\@.
 \emph{Knowledge compilation} is the process of converting a formula
@@
 a generalization of the representations produced by current knowledge
-compilers. To capture both the structure of a POG
+compilers. The CPOG (for ``checkable'' POG) file format then
+captures both the structure of a POG
 and a checkable proof of its logical equivalence to a Boolean formula in
-conjunctive normal form (CNF), we propose the CPOG (for ``checkable'' POG) file format. A CPOG
+conjunctive normal form (CNF).  A CPOG
 proof consists of a sequence of clause addition and deletion steps,
 based on an extended resolution proof system~\cite{Tseitin:1983}.
-We establish a set of conditions that a CPOG file must satisfy in order to
-encode a well-formed POG and provide a valid equivalence proof.
+We establish a set of conditions that, when satisified by a CPOG file, guarantee that it
+encodes a well-formed POG and provides a valid equivalence proof.
 

### 2023-03-22 doc/mc-paper.tex
checkable POG --> certified POG

@@
 a generalization of the representations produced by current knowledge
-compilers. The CPOG (for ``checkable'' POG) file format then
+compilers. The CPOG (for ``certified'' POG) file format then
 captures both the structure of a POG

### 2023-03-22 doc/mc-paper.tex
Incremental changes

@@
 \emph{Knowledge compilation} is the process of converting a formula
-into such a form. Existing knowledge compilers provide no guarantee that their output correctly
-represents the original formula, and therefore they cannot validate the model counts.
+into such a form. Unfortunately existing knowledge compilers provide no guarantee that their output correctly
+represents the original formula, and therefore they cannot validate a model count, or any other computed value.
 
@@
 
-We evaluate the capabilities of this toolchain using benchmark
+We evaluate our toolchain using benchmark
 formulas from the 2022 standard and weighted model competitions.  Our
-tools are able to handle all but the largest d-DNNF graphs generated
+tools handle all but the largest d-DNNF graphs generated
 by \dfour{}.  We also measure the benefits of several optimizations as
-well as the performance of the verified checker relative to one
-written in C.
+well as the relative performance of the verified checker with one
+designed for high performance and capacity.
 
@@
 The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
-By contrast, our proofs construct a new propositional formula, and we must verify both that it
-is partitioned and that it is equivalent to the input formula.
+By contrast, our proofs construct a new propositional formula, and we must verify
+%both
+that it
+%satisfies a set of rules and that it
+is equivalent to the input formula.
 This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
@@
 
-A proof in our framework takes the form of a sequence of clause addition
+A proof in our framework consists of a sequence of clause addition
 and deletion steps, with each step preserving the set of solutions to
@@
 denote the set of extension variables occuring in formula $\theta$.
-Starting with input formula $\inputformula$,
+Starting with  $\theta$ equal to input formula $\inputformula$,
 the proof must maintain the invariant that
@@
 ours have important differences.  Most
-significantly, each proof step must preserve the set of solutions with respect to the input variables.
+significantly, each proof step must preserve the set of solutions with respect to the input variables;
+our proofs must therefore justify both clause deletions and additions.
 By contrast, an unsatisfiability proof need only guarantee that
 no proof step causes a satisfiable set of clauses to become
-unsatisfiable.  As a consequence, our proofs must justify clause deletions as well as clause additions.
+unsatisfiable, and therefore it need only justify clause additions.
 
@@
 recursively traverses the POG, generating proof obligations at each
-node encountered.  It too may require calls to a proof-generating SAT
+node encountered.  It may require multiple calls to a proof-generating SAT
 solver.
@@
 to the proof framework---they build on the power of extended
-resolution to create new logical structures.  They
+resolution to enable new logical structures.  They
 involve declaring new product nodes to encode products of literals.
@@
 will cause further recursive calls only once.  Our experimental
-results (Appendix~\ref{app:experiments}) shows that these
+results (Appendix~\ref{app:experiments}) show that these
 optimizations yield substantial benefits.

### 2023-03-22 crat/buddy/config.h
Housecleaning

@@ -1,20 +1,0 @@
-/* config.h.  Reduced from version generated for BuDDy 2.4 */
-
-/* BuDDy's major version. */
-#define MAJOR_VERSION 2
-
-/* BuDDy's minor version. */
-#define MINOR_VERSION 5
-
-/* Define to the full name of this package. */
-#define PACKAGE_NAME "buddy"
-
-/* Define to the full name and version of this package. */
-#define PACKAGE_STRING "buddy 2.5"
-
-/* Define to the one symbol short name of this package. */
-#define PACKAGE_TARNAME "buddy"
-
-/* Define to the version of this package. */
-#define PACKAGE_VERSION "2.5"
-

### 2023-03-22 crat/buddy/src/bdd.h

@@ -1,922 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  As noted below, this code is a modified version of code authored and
-  copywrited by Jorn Lind-Nielsen.  Permisssion to use the original
-  code is subject to the terms noted below.
-
-  Regarding the modifications, and subject to any constraints on the
-  use of the original code, permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
[…]

### 2023-03-22 crat/buddy/src/bddio.c

@@ -1,593 +1,0 @@
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
-    for any purpose, provided that (1) the above copyright notice and
-    the following two paragraphs appear in all copies of the source code
-    and (2) redistributions, including without limitation binaries,
-    reproduce these notices in the supporting documentation. Substantial
-    modifications to this software may be copyrighted by their authors
-    and need not follow the licensing terms described here, provided
-    that the new terms are clearly indicated in all files where they apply.
-
-    IN NO EVENT SHALL JORN LIND-NIELSEN, OR DISTRIBUTORS OF THIS
-    SOFTWARE BE LIABLE TO ANY PARTY FOR DIRECT, INDIRECT, SPECIAL,
-    INCIDENTAL, OR CONSEQUENTIAL DAMAGES ARISING OUT OF THE USE OF THIS
-    SOFTWARE AND ITS DOCUMENTATION, EVEN IF THE AUTHORS OR ANY OF THE
-    ABOVE PARTIES HAVE BEEN ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-
-    JORN LIND-NIELSEN SPECIFICALLY DISCLAIM ANY WARRANTIES, INCLUDING,
-    BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND
-    FITNESS FOR A PARTICULAR PURPOSE. THE SOFTWARE PROVIDED HEREUNDER IS
-    ON AN "AS IS" BASIS, AND THE AUTHORS AND DISTRIBUTORS HAVE NO
-    OBLIGATION TO PROVIDE MAINTENANCE, SUPPORT, UPDATES, ENHANCEMENTS, OR
-    MODIFICATIONS.
-========================================================================*/
-
-/*************************************************************************
[…]

### 2023-03-22 crat/buddy/src/bddop.c

@@ -1,3485 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  As noted below, this code is a modified version of code authored and
-  copywrited by Jorn Lind-Nielsen.  Permisssion to use the original
-  code is subject to the terms noted below.
-
-  Regarding the modifications, and subject to any constraints on the
-  use of the original code, permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
-    for any purpose, provided that (1) the above copyright notice and
-    the following two paragraphs appear in all copies of the source code
[…]

### 2023-03-22 crat/buddy/src/bvec.c

@@ -1,1352 +1,0 @@
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
-    for any purpose, provided that (1) the above copyright notice and
-    the following two paragraphs appear in all copies of the source code
-    and (2) redistributions, including without limitation binaries,
-    reproduce these notices in the supporting documentation. Substantial
-    modifications to this software may be copyrighted by their authors
-    and need not follow the licensing terms described here, provided
-    that the new terms are clearly indicated in all files where they apply.
-
-    IN NO EVENT SHALL JORN LIND-NIELSEN, OR DISTRIBUTORS OF THIS
-    SOFTWARE BE LIABLE TO ANY PARTY FOR DIRECT, INDIRECT, SPECIAL,
-    INCIDENTAL, OR CONSEQUENTIAL DAMAGES ARISING OUT OF THE USE OF THIS
-    SOFTWARE AND ITS DOCUMENTATION, EVEN IF THE AUTHORS OR ANY OF THE
-    ABOVE PARTIES HAVE BEEN ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-
-    JORN LIND-NIELSEN SPECIFICALLY DISCLAIM ANY WARRANTIES, INCLUDING,
-    BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND
-    FITNESS FOR A PARTICULAR PURPOSE. THE SOFTWARE PROVIDED HEREUNDER IS
-    ON AN "AS IS" BASIS, AND THE AUTHORS AND DISTRIBUTORS HAVE NO
-    OBLIGATION TO PROVIDE MAINTENANCE, SUPPORT, UPDATES, ENHANCEMENTS, OR
-    MODIFICATIONS.
-========================================================================*/
-
-/*************************************************************************
[…]

### 2023-03-22 crat/buddy/src/cache.c

@@ -1,156 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  As noted below, this code is a modified version of code authored and
-  copywrited by Jorn Lind-Nielsen.  Permisssion to use the original
-  code is subject to the terms noted below.
-
-  Regarding the modifications, and subject to any constraints on the
-  use of the original code, permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
-    for any purpose, provided that (1) the above copyright notice and
[…]

### 2023-03-22 crat/buddy/src/cache.h

@@ -1,123 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  As noted below, this code is a modified version of code authored and
-  copywrited by Jorn Lind-Nielsen.  Permisssion to use the original
-  code is subject to the terms noted below.
-
-  Regarding the modifications, and subject to any constraints on the
-  use of the original code, permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
-    for any purpose, provided that (1) the above copyright notice and
-    the following two paragraphs appear in all copies of the source code
[…]

### 2023-03-22 crat/buddy/src/fdd.c

@@ -1,1077 +1,0 @@
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
-    for any purpose, provided that (1) the above copyright notice and
-    the following two paragraphs appear in all copies of the source code
-    and (2) redistributions, including without limitation binaries,
-    reproduce these notices in the supporting documentation. Substantial
-    modifications to this software may be copyrighted by their authors
-    and need not follow the licensing terms described here, provided
-    that the new terms are clearly indicated in all files where they apply.
-
-    IN NO EVENT SHALL JORN LIND-NIELSEN, OR DISTRIBUTORS OF THIS
-    SOFTWARE BE LIABLE TO ANY PARTY FOR DIRECT, INDIRECT, SPECIAL,
-    INCIDENTAL, OR CONSEQUENTIAL DAMAGES ARISING OUT OF THE USE OF THIS
-    SOFTWARE AND ITS DOCUMENTATION, EVEN IF THE AUTHORS OR ANY OF THE
-    ABOVE PARTIES HAVE BEEN ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-
-    JORN LIND-NIELSEN SPECIFICALLY DISCLAIM ANY WARRANTIES, INCLUDING,
-    BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND
-    FITNESS FOR A PARTICULAR PURPOSE. THE SOFTWARE PROVIDED HEREUNDER IS
-    ON AN "AS IS" BASIS, AND THE AUTHORS AND DISTRIBUTORS HAVE NO
-    OBLIGATION TO PROVIDE MAINTENANCE, SUPPORT, UPDATES, ENHANCEMENTS, OR
-    MODIFICATIONS.
-========================================================================*/
-
-/*************************************************************************
-  $Header: /cvsroot/buddy/buddy/src/fdd.c,v 1.1.1.1 2004/06/25 13:22:40 haimcohen Exp $
-  FILE:  fdd.c
-  DESCR: Finite domain extensions to BDD package
-  AUTH:  Jorn Lind
-  DATE:  (C) june 1997
-
[…]

### 2023-03-22 crat/buddy/src/ilist.c

@@ -1,329 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  This code was not included in the original BuDDy distribution and is
-  therefore not subject to any of its licensing terms.
-
-  Permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-#include <stdlib.h>
-#include <stdio.h>
-#include <string.h>
-#include "ilist.h"
-
-#ifdef ENABLE_TBDD
-int bdd_error(int);
-#define ILIST_ALLOC (-23)
-#endif
-
-#define ERROUT stdout
-
-/*============================================
-   Integer lists
-============================================*/
-
-/* Macros */
-#define ILIST_BASE(ils) ((ils)-ILIST_OVHD)
-#define ILIST_LENGTH(ils) ((ils)[-1])
-#define IABS(x) ((x)<0?(-x):(x))
[…]

### 2023-03-22 crat/buddy/src/ilist.h

@@ -1,190 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  This code was not included in the original BuDDy distribution and is
-  therefore not subject to any of its licensing terms.
-
-  Permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-#ifndef _ILIST_H
-#define _ILIST_H
-
-#include <stdio.h>
-#include <stdbool.h>
-#include <limits.h>
-
-
-/*============================================
-   Integer lists
-============================================*/
-
-/* Allow this headerfile to define C++ constructs if requested */
-#ifdef __cplusplus
-#define CPLUSPLUS
-#endif
-
-#ifdef CPLUSPLUS
-extern "C" {
-#endif
-
-/*
-  Data type ilist is used to represent clauses and clause id lists.
-  These are simply lists of integers, where the value at position -1
-  indicates the list length and the value at position -2 indicates the
-  maximum list length.  The value at position -2 is positive for
[…]

### 2023-03-22 crat/buddy/src/kernel.c

@@ -1,1851 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  As noted below, this code is a modified version of code authored and
-  copywrited by Jorn Lind-Nielsen.  Permisssion to use the original
-  code is subject to the terms noted below.
-
-  Regarding the modifications, and subject to any constraints on the
-  use of the original code, permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
-    for any purpose, provided that (1) the above copyright notice and
[…]

### 2023-03-22 crat/buddy/src/kernel.h

@@ -1,331 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  As noted below, this code is a modified version of code authored and
-  copywrited by Jorn Lind-Nielsen.  Permisssion to use the original
-  code is subject to the terms noted below.
-
-  Regarding the modifications, and subject to any constraints on the
-  use of the original code, permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
[…]

### 2023-03-22 crat/buddy/src/pairs.c

@@ -1,335 +1,0 @@
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
-    for any purpose, provided that (1) the above copyright notice and
-    the following two paragraphs appear in all copies of the source code
-    and (2) redistributions, including without limitation binaries,
-    reproduce these notices in the supporting documentation. Substantial
-    modifications to this software may be copyrighted by their authors
-    and need not follow the licensing terms described here, provided
-    that the new terms are clearly indicated in all files where they apply.
-
-    IN NO EVENT SHALL JORN LIND-NIELSEN, OR DISTRIBUTORS OF THIS
-    SOFTWARE BE LIABLE TO ANY PARTY FOR DIRECT, INDIRECT, SPECIAL,
-    INCIDENTAL, OR CONSEQUENTIAL DAMAGES ARISING OUT OF THE USE OF THIS
-    SOFTWARE AND ITS DOCUMENTATION, EVEN IF THE AUTHORS OR ANY OF THE
-    ABOVE PARTIES HAVE BEEN ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-
-    JORN LIND-NIELSEN SPECIFICALLY DISCLAIM ANY WARRANTIES, INCLUDING,
-    BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND
-    FITNESS FOR A PARTICULAR PURPOSE. THE SOFTWARE PROVIDED HEREUNDER IS
-    ON AN "AS IS" BASIS, AND THE AUTHORS AND DISTRIBUTORS HAVE NO
-    OBLIGATION TO PROVIDE MAINTENANCE, SUPPORT, UPDATES, ENHANCEMENTS, OR
-    MODIFICATIONS.
-========================================================================*/
-
-/*************************************************************************
[…]

### 2023-03-22 crat/buddy/src/prover.c

@@ -1,1018 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  This code was not included in the original BuDDy distribution and is
-  therefore not subject to any of its licensing terms.
-
-  Permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-#include <stdio.h>
-#include <stdlib.h>
-#include "tbdd.h"
-#include "prover.h"
-#include "kernel.h"
-
-
-/* Global variables exported by prover */
-proof_type_t proof_type = PROOF_FRAT;
-int verbosity_level = 1;
-int *variable_counter = NULL;
-int *clause_id_counter = NULL;
-int clause_limit = CLAUSE_LIMIT_LRAT;
-int total_clause_count = 0;
-int input_clause_count = 0;
-int input_variable_count = 0;
-int max_live_clause_count = 0;
[…]

### 2023-03-22 crat/buddy/src/prover.h

@@ -1,125 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  This code was not included in the original BuDDy distribution and is
-  therefore not subject to any of its licensing terms.
-
-  Permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-/* Routines to manage proof generation */
-
-#ifndef PROVER_H
-#define PROVER_H
-
-#include <stdarg.h>
-
-/* Size limits */
-/* Maximum number of clauses for each proof type */
-#define CLAUSE_LIMIT_LRAT (1<<30)
-#define CLAUSE_LIMIT_FRAT (1<<27)
-#define CLAUSE_LIMIT_DRAT (1<<25)
-
-/* Allow this headerfile to define C++ constructs if requested */
-#ifdef __cplusplus
-#define CPLUSPLUS
-#endif
-
-#ifdef CPLUSPLUS
-extern "C" {
-#endif
-
-/* Global variables exported by prover */
-extern proof_type_t proof_type;
-extern int verbosity_level;
-extern int clause_limit;
-extern int *variable_counter;
-extern int *clause_id_counter;
-extern int total_clause_count;
[…]

### 2023-03-22 crat/buddy/src/pseudoboolean.h

@@ -1,197 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  This code was not included in the original BuDDy distribution and is
-  therefore not subject to any of its licensing terms.
-
-  Permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-/* Interface for proof-generating operations on Pseudo-Boolean functions */
-/* Initial version on supports XOR constraints */
-
-#ifndef _PSEUDOBOOLEAN_H
-#define _PSEUDOBOOLEAN_H
-
-#include <vector>
-#include <unordered_set>
-#include <cstdint>
-#include "tbdd.h"
-
-namespace trustbdd{
-
-// A handy class for generating random numbers with own seed
-// Pseudo-random number generator based on the Lehmer MINSTD RNG
-// Use to both randomize selection and to provide a unique value
-// for each cost.
-
-class Sequencer {
-
-public:
-    Sequencer(int s) { seed = s; }
-
-    Sequencer(void) { seed = default_seed; }
-
-    Sequencer(Sequencer &s) { seed = s.seed; }
-
[…]

### 2023-03-22 crat/buddy/src/reorder.c

@@ -1,2328 +1,0 @@
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
-    for any purpose, provided that (1) the above copyright notice and
-    the following two paragraphs appear in all copies of the source code
-    and (2) redistributions, including without limitation binaries,
-    reproduce these notices in the supporting documentation. Substantial
-    modifications to this software may be copyrighted by their authors
-    and need not follow the licensing terms described here, provided
-    that the new terms are clearly indicated in all files where they apply.
-
-    IN NO EVENT SHALL JORN LIND-NIELSEN, OR DISTRIBUTORS OF THIS
-    SOFTWARE BE LIABLE TO ANY PARTY FOR DIRECT, INDIRECT, SPECIAL,
-    INCIDENTAL, OR CONSEQUENTIAL DAMAGES ARISING OUT OF THE USE OF THIS
-    SOFTWARE AND ITS DOCUMENTATION, EVEN IF THE AUTHORS OR ANY OF THE
-    ABOVE PARTIES HAVE BEEN ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-
-    JORN LIND-NIELSEN SPECIFICALLY DISCLAIM ANY WARRANTIES, INCLUDING,
-    BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND
-    FITNESS FOR A PARTICULAR PURPOSE. THE SOFTWARE PROVIDED HEREUNDER IS
-    ON AN "AS IS" BASIS, AND THE AUTHORS AND DISTRIBUTORS HAVE NO
-    OBLIGATION TO PROVIDE MAINTENANCE, SUPPORT, UPDATES, ENHANCEMENTS, OR
-    MODIFICATIONS.
-========================================================================*/
-
-/*************************************************************************
-  $Header: /cvsroot/buddy/buddy/src/reorder.c,v 1.1.1.1 2004/06/25 13:22:59 haimcohen Exp $
-  FILE:  reorder.c
-  DESCR: BDD reordering functions
-  AUTH:  Jorn Lind
-  DATE:  (C) january 1998
[…]

### 2023-03-22 crat/buddy/src/schema.c

@@ -1,939 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  This code is derived from the file 'kernel.c', part of the BuDDy BDD
-  package authored and copywrited by Jorn Lind-Nielsen.  Permisssion
-  to use the original code is subject to the terms noted below.
-
-  Regarding the modifications, and subject to any constraints on the
-  use of the original code, permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
-    license or royalty fees, to use, reproduce, prepare derivative
-    works, distribute, and display this software and its documentation
[…]

### 2023-03-22 crat/buddy/src/schema.h

@@ -1,330 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  This code is derived from the files 'kernel.h' and 'bdd.h', parts of
-  the BuDDy BDD package authored and copywrited by Jorn Lind-Nielsen.
-  Permisssion to use the original code is subject to the terms noted
-  below.
-
-  Regarding the modifications, and subject to any constraints on the
-  use of the original code, permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-/*========================================================================
-               Copyright (C) 1996-2002 by Jorn Lind-Nielsen
-                            All rights reserved
-
-    Permission is hereby granted, without written agreement and without
[…]

### 2023-03-22 crat/buddy/src/tbdd.c

@@ -1,902 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  This code was not included in the original BuDDy distribution and is
-  therefore not subject to any of its licensing terms.
-
-  Permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-#include <stdlib.h>
-#include <stdio.h>
-#include <string.h>
-#include "tbdd.h"
-#include "kernel.h"
-
-
-/*============================================
-  Local data
-============================================*/
-
-#define BUFLEN 2048
-// For formatting information
-static char ibuf[BUFLEN];
-
-#define FUN_MAX 10
-static tbdd_info_fun ifuns[FUN_MAX];
-static int ifun_count = 0;
-
-static tbdd_done_fun dfuns[FUN_MAX];
-static int dfun_count = 0;
-
-static int last_variable = 0;
[…]

### 2023-03-22 crat/buddy/src/tbdd.h

@@ -1,378 +1,0 @@
-/*========================================================================
-  Copyright (c) 2022 Randal E. Bryant, Carnegie Mellon University
-  
-  This code was not included in the original BuDDy distribution and is
-  therefore not subject to any of its licensing terms.
-
-  Permission is hereby granted, free of
-  charge, to any person obtaining a copy of this software and
-  associated documentation files (the "Software"), to deal in the
-  Software without restriction, including without limitation the
-  rights to use, copy, modify, merge, publish, distribute, sublicense,
-  and/or sell copies of the Software, and to permit persons to whom
-  the Software is furnished to do so, subject to the following
-  conditions:
-  
-  The above copyright notice and this permission notice shall be
-  included in all copies or substantial portions of the Software.
-  
-  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
-  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
-  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
-  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
-  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-  SOFTWARE.
-========================================================================*/
-
-
-#ifndef _TBDD_H
-#define _TBDD_H
-
-/* Turn on TBDD-specific information in files */
-#ifndef ENABLE_TBDD
-#define ENABLE_TBDD 1
-#endif
-
-/* Allow this headerfile to define C++ constructs if requested */
-#ifdef __cplusplus
-#define CPLUSPLUS
-#endif
-
-/*============================================
- API for trusted extension to BuDDy BDD package
-============================================*/
-
-
-#include <limits.h>
-#include "ilist.h"
-#include "bdd.h"
-
-/* Value representing logical truth */
-#define TAUTOLOGY INT_MAX 
-
-/* 
[…]

### 2023-03-22 crat/prototype/pigeon-d4-hintgen/prun.sh

@@ -1,12 +1,0 @@
-#!/bin/bash
-for N in {3..9}
-do
-    make psd-all N=$N
-    make pst-all N=$N
-done
-
-for N in {10..11}
-do
-    make pst-all N=$N
-done
-

### 2023-03-22 crat/prototype/pigeon-d4-split1/prun.sh

@@ -1,7 +1,0 @@
-#!/bin/bash
-for N in {3..10}
-do
-    make pst-all N=$N
-    make psd-all N=$N
-done
-

### 2023-03-22 crat/prototype/pigeon-d4-split2-cadical/prun.sh

@@ -1,12 +1,0 @@
-#!/bin/bash
-for N in {3..10}
-do
-    make pst-all N=$N
-    make psd-all N=$N
-done
-
-for N in {11..13}
-do
-    make pst-all N=$N
-done
-

### 2023-03-22 crat/prototype/pigeon-d4-split2-h3/prun.sh

@@ -1,12 +1,0 @@
-#!/bin/bash
-for N in {3..10}
-do
-    make pst-all N=$N
-    make psd-all N=$N
-done
-
-for N in {11..12}
-do
-    make pst-all N=$N
-done
-

### 2023-03-22 crat/prototype/pigeon-d4-split2/negate.c

@@ -1,20 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-
-int main (int argc, char** argv) {
-
-  FILE *input = fopen (argv[1], "r");
-
-  int size = 0;
-  while (1) {
-    int lit;
-    int tmp = fscanf (input, " %i ", &lit);
-    if (tmp == EOF || tmp == 0) break;
-    if (size == 0) printf ("a");
-    size++;
-    printf (" %i", -lit);
-    if (lit == 0) { size = 0; printf ("\n"); }
-  }
-
-  fclose (input);
-}

### 2023-03-22 crat/prototype/pigeon-d4-split2/prun.sh

@@ -1,12 +1,0 @@
-#!/bin/bash
-for N in {5..10}
-do
-    make pst-all N=$N
-    make psd-all N=$N
-done
-
-for N in {11..12}
-do
-    make pst-all N=$N
-done
-

### 2023-03-22 crat/prototype/pigeon-d4-split2/run.sh

@@ -1,11 +1,0 @@
-BASE=$1
-
-cat $BASE.drat | grep -v "d " > $BASE.rat
-./negate $BASE.rat > $BASE.cubes
-cat $BASE.acnf $BASE.cubes | awk '/p cnf/ {print "p inccnf"} / 0/ {print $0}'> $BASE.icnf
-~/cadical/build/cadical $BASE.icnf $BASE.drup --no-binary
-#~/GitHub/Mario/cadical/build/cadical $BASE.icnf $BASE.lrat --no-binary --lrat=true
-cat $BASE.drup | grep -v "d " > $BASE.rup
-../crat-lrat/crat-lrat $BASE.acnf $BASE.rup -f | grep -v "c " | grep " 0" > result.txt
-cat result.txt
-

### 2023-03-22 crat/prototype/pigeon-dsharp-hintgen/prun.sh

@@ -1,7 +1,0 @@
-#!/bin/bash
-for N in {3..8}
-do
-    make psd-all N=$N
-    make pst-all N=$N
-done
-

### 2023-03-27 doc/pmc-notes.tex
Initial thoughts on projected model counting

--- 2023-03-22 doc/mc-paper.tex
@@
 
-\newcommand{\varset}{X}
+\newcommand{\dvarset}{X}
+\newcommand{\avarset}{Y}
 \newcommand{\exvarset}{Z}
@@
 \newcommand{\rep}{\textbf{R}}
+\newcommand{\drep}{\textbf{R}_D}
 \newcommand{\radd}{+}
@@
 
-\newcommand{\assign}{\alpha}
+\newcommand{\dassign}{\alpha}
+\newcommand{\aassign}{\beta}
+\newcommand{\acombine}{\!\cdot\!}
 \newcommand{\passign}{\rho}
 \newcommand{\lassign}{\beta}
+\newcommand{\skolem}{\sigma}
 \newcommand{\uassign}{{\cal U}}
+\newcommand{\udassign}{\uassign_X}
+\newcommand{\uaassign}{\uassign_Y}
 \newcommand{\modelset}{{\cal M}}
+\newcommand{\dmodelset}{{\cal M_{D}}}
 
@@
 \newcommand{\dtrim}{\progname{drat-trim}}
+\newcommand{\lean}{Lean~4}
 
@@
 
-\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting}
+\title{Notes on \\ Certified Projected Knowledge Compilation \\ \today }
 
-\titlerunning{Certified Knowledge Compilation}
+\titlerunning{Projected Knowledge Compilation}
 
@@
 
-\newcommand{\program}[1]{\textsc{#1}}
-\newcommand{\lean}{Lean~4}
-
 \begin{document}
@@
 
-\begin{abstract}
+\section{Notation}
 
-Computing many useful properties of Boolean formulas, such as their weighted or unweighted model count,
-is intractable on general representations. It can become tractable when formulas are expressed in a
-special form, such as the deterministic,
-decomposable, negation normal form (d-DNNF)\@.
-\emph{Knowledge compilation} is the process of converting a formula
-into such a form. Existing knowledge compilers provide no guarantee that their output correctly
-represents the original formula, and therefore they cannot validate the model counts.
+\subsection{Data Model}
 
-We present \emph{Partitioned-Operation Graphs} (POGs), a generalization
-of the representations used by existing knowledge compilers.
-We have designed  CPOG, a framework that can express proofs of equivalence between a
-POG  and a Boolean formula in conjunctive normal form (CNF).
+The set of Boolean variables is divided into two sets: {\em data} variables
+$\dvarset$ and {\em auxiliary} variables $\avarset$.  We use separate
+assignments for the two sets of variables:
+$\dassign \colon \dvarset \rightarrow \{0,1\}$
+ and
+ $\aassign \colon \avarset \rightarrow \{0,1\}$, and the symbol $\acombine$ to combine the two assignments:
+ $\dassign\acombine\aassign \colon \dvarset\cup\avarset \rightarrow \{0,1\}$.
+We let $\udassign$, $\uaassign$, and $\uassign$ denote the set of all
+possible assignments to the data variables, auxiliary variables, and
+all variables, respectively.  From this, we can see that
+$\uassign = \udassign \times \uaassign$.
 
-We have developed a program that generates POG representations from
-the d-DNNF graphs produced by the state-of-the-art knowledge compiler
-\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
-are equivalent to the input CNF formulas.  Our toolchain
-for generating and verifying POGs scales to all but the largest d-DNNF
-graphs produced by \dfour{} for formulas from a recent model counting
-competition. Additionally, we have developed a formally verified CPOG
-checker and model counter for POGs in the \lean{} proof assistant.
-In doing so, we proved the soundness of our proof framework. These programs
-comprise the first formally verified toolchain for weighted and unweighted
-model counting.
-\end{abstract}
+For Boolean formula $\phi$ over the variables $\dvarset$ and
+$\avarset$, the set of {\em models}, written $\modelset(\phi)$
+consists of those assignments $\dassign \acombine \aassign$ that
+satisfy the formula.  The set of {\em data models}, written $\dmodelset(\phi)$ is defined as:
+\begin{eqnarray}
+\dmodelset(\phi) & \doteq & \{ \dassign \in \udassign | \exists \aassign \in \uaassign \; \dassign\acombine\aassign \in \modelset(\phi) \} \label{eqn:dmodelset}
+\end{eqnarray}
 
-\section{Introduction}
+\subsection{Data Ring Evaluation}
 
-Given a Boolean formula $\phi$, modern Boolean satisfiability (SAT) solvers can
-find an assignment satisfying $\phi$ or generate a proof that no
-such assignment exists.  They have applications across a variety of
-domains including computational mathematics, combinatorial
-optimization, and the formal verification of hardware, software, and
-security protocols.  Some applications, however, require going
-beyond Boolean satisfiability.  For example, the {\em model
-  counting problem} requires computing the number of satisfying
-assignments of a formula, including in cases where there are far too many
-to enumerate individually.  Model counting has
-applications in artificial intelligence, computer security, and
-statistical sampling.  There are also many useful extensions of standard model counting,
-including {\em
-  weighted model counting}, where a weight is defined for
-each possible assignment, and the goal becomes to compute the sum of the weights
-of the satisfying assignments.
+  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
+  every data variable $x \in \dvarset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.
 
-Model counting is a challenging problem---more challenging than the
-already NP-hard Boolean satisfiability.  Several
-tractable variants of Boolean satisfiability, including 2-SAT, become
-intractable when the goal is to count models and not just determine
-satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
-model counters that scale to very large formulas have been developed, as
-witnessed by the progress in recent model counting competitions.
+  For Boolean formula $\phi$ over data variables $\dvarset$ and auxiliary variables $\avarset$, and ring weight function $w$, the {\em data ring evaluation problem} computes
+  \begin{eqnarray}
+    \drep(\phi, w) & = & \sum_{\alpha \in \dmodelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
+  \end{eqnarray}
 
-One approach to model counting, known as {\em knowledge compilation},
-transforms the formula into a structured form for which model counting
-is straightforward.  For example, the {\em deterministic,
-  decomposable, negation normal form} (d-DNNF) introduced by
-Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
-Boolean formula as a directed acyclic graph, with terminal nodes
-labeled by Boolean variables and their complements, and with each
-nonterminal node labeled by a Boolean And or Or operation.  Restrictions
-are placed on the structure of the graph such that a count of the
-models can be computed by a single bottom-up traversal.
-Kimmig et al.\ present a very general {\em
-  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
-properties of Boolean functions that can be efficiently computed from
-a d-DNNF representation.  These include standard and weighted model
-counting, and much more.
+The goal of {\em projected knowledge compilation} is to generate a
+representation of an input formula $\inputformula$ such that data ring
+evaluation can be performed in polynomial time with respect to the
+size of the representation.
 
-One shortcoming of existing knowledge compilers is that they have no
-way to validate that their generated results are correct, i.e., that
-the compiled representation is logically equivalent to the original
-formula.  By contrast, all modern SAT solvers can generate
-checkable proofs when they encounter unsatisfiable formulas.  The
-guarantee provided by a checkable certificate of correctness enables
-users of SAT solvers to fully trust their results.  Experience has also
-shown that proof production capabilites allow SAT solver developers to quickly
-detect and diagnose bugs in their programs. This, in turn, has led
-to more reliable SAT solvers.
+\subsection{Skolem Formulas}
 
-This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
-a generalization of the representations produced by current knowledge
-compilers. The CPOG (for ``certified'' POG) file format then
-captures both the structure of a POG
-and a checkable proof of its logical equivalence to a Boolean formula in
-conjunctive normal form (CNF).  A CPOG
-proof consists of a sequence of clause addition and deletion steps,
-based on an extended resolution proof system~\cite{Tseitin:1983}.
-We establish a set of conditions that, when satisified by a CPOG file, guarantee that it
-encodes a well-formed POG and provides a valid equivalence proof.
+One method of instantiating the existential quantifier
+(\ref{eqn:dmodelset}) is to provide formulas that compute an
+assignment $\aassign$ for each possible $\dassign$.  For $y \in
+\avarset$, define a {\em Skolem formula} $\skolem_y$ as a Boolean
+formula over $\dvarset$.  It implicitly defines a function
+$\skolem_y \colon \udassign \rightarrow \{0,1\}$.  Combining these functions for all $y \in
+\avarset$ then defines a function $\skolem \colon \udassign
+\rightarrow \uaassign$, where $\skolem(\dassign)$ gives the assignment
+$\aassign$ such that $\aassign(y) = \skolem_y(\dassign)$.
 
-\begin{figure}
-\centering{
-\begin{tikzpicture}[scale=0.17]
-  \draw (8,10) rectangle (20,18);
-  \node at (14,15.2) {Knowledge};
-  \node at (14,12.8) {Compiler};
+A {\em valid Skolemization} of formula $\phi$ is then a set of Skolem formulas $\skolem_y$
+such that every data assignment $\dassign\in \dmodelset(\phi)$ is extended by $\skolem$ to become a model of $\phi$, i.e.,
+$\dassign \acombine \skolem(\dassign) \in \modelset(\phi)$.
 
-  \draw (28,10) rectangle (40,18);
-  \node at (34,15.2) {Proof};
-  \node at (34,12.8) {Generator};
+\section{Example Formula}
 
-  \draw (50,10) rectangle (62,18);
-  \node at (56,15.2) {Proof};
-  \node at (56,12.8) {Checker};
+Consider an encoding of the formula $(x_1 \land x_2) \lor x_3$ using
+encoding variable $y$.  A full {\em Tseitin} encoding of $y
+\leftrightarrow (x_1 \land x_2)$ requires three clauses, but we can
+use a Plaisted-Greenbaum encoding in this context, encoding just $y
+\rightarrow (x_1 \land x_2)$, represented by the clauses
+$x_1 \lor \obar{y}$ and $x_2 \lor \obar{y}$.  We then encode the disujunction
+with the clause $x_3 \lor y$.  The DIMACS representation is then as follows (using variable 4 to represent $y$):
 
-  \draw (50,0) rectangle (62,8);
-  \node at (56,5.2) {Ring};
-  \node at (56,2.8) {Evaluator};
-
-  \draw[dashed] (48,-2) rectangle (64,22);
-  \node at (56,20) {Trusted Code};
-
-  %% CNF
-  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (50,16);
-  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
-  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
-  \node [left] at (-1,20) {\large {$\inputformula$}};
-  \node [above] at (2,20) {\texttt{.cnf}};
-
-  %% NNF
-  \draw[thick] (20,12) [-latex] -- (28,12);
-  \node [above] at (24,12) {\texttt{.ddnnf}};
-
-  %% CPOG
-  \draw[thick] (40,12) [-latex] -- (50,12);
-  \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
-  \node [above] at (44,12) {\texttt{.cpog}};
-
-  %% OK/Not
-  \draw[thick] (62,14) [-latex] -- (66,14) ;
-  \node [right] at (66,15.2) { OK /} ;
-  \node [right] at (66,12.8) { Not OK} ;
-  \draw[thick] (62,4) [-latex] -- (66,4) ;
-  \node [right] at (66,4) {\large {$\textbf{R}(\inputformula)$}};
-
-\end{tikzpicture}
-}
-\caption{Certifying toolchain.
-  The output of a standard knowledge compiler is converted into a combined graph/proof (CPOG)
-  which can be independently checked and evaluated}
-\label{fig:chain}
-\end{figure}
-
-Figure~\ref{fig:chain} illustrates our certifying knowledge
-compilation and model counting toolchain.  Starting with
-input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
-generates a d-DNNF representation, and the {\em proof
-  generator} uses this to generate a CPOG file.
-The {\em proof checker} verifies the equivalence of the CNF and CPOG representations.
-The {\em ring evaluator} computes
-a standard or weighted model count from the POG\@ representation.
-As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
-moves the root of trust away from the complex and highly
-optimized knowledge compiler to a relatively simple checker and
-evaluator.  Importantly, the proof generator need not be
-trusted---its errors will be caught by the proof checker.
-
-To ensure soundness of the abstract CPOG proof system, as well as correctness of its
-concrete implementation, we formally verified the proof system as well as versions
-of the proof checker and ring evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
-Running these two programs on a particular CPOG file gives assurance that the proof
-and the model count are correct. Our experience with developing a formally verified
-proof checker has shown that, even within the well-understood framework of extended
-resolution, it can be challenging to formulate a full set of requirements that guarantee soundness.
-In fact, subtle details in the partitioned sum rule were identified during this process.
-
-We evaluate the capabilities of this toolchain using benchmark
-formulas from the 2022 standard and weighted model competitions.  Our
-tools are able to handle all but the largest d-DNNF graphs generated
-by \dfour{}.  We also measure the benefits of several optimizations as
-well as the performance of the verified checker relative to one
-written in C.
-
-\section{Related Work}
-
-Generating proofs of unsatisfiability in SAT solvers has a long
-tradition~\cite{ZhangMalik} and has become widely accepted due to the
-formulation of clausal proof systems for which proofs can readily be
-generated and efficiently checked
-\cite{RAT,wetzler14_drattrim}.
-A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
-The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
-By contrast, our proofs construct a new propositional formula, and we must verify both that it
-is partitioned and that it is equivalent to the input formula.
-This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
-Sections~\ref{sect:cpog} and \ref{section:formally-verified-toolchain}.
-
-Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
-proof framework for model counting programs.  Their proof rules are
-based on the algorithms commonly used by model counters.
-They present a series of lemmas and a theorem stating
-that their checking framework is sound.  They also present
-experimental results for cases where they modify existing model
-counters to generate proof traces, and where they generate proof
-traces from d-DNNF graphs.
-%% They have not formally verified their
-%%proof framework or created a formally verified checker.
-Our toolchain is not directly comparable to theirs---they directly
-certify the numeric value produced by a model counter, while we
-certify that the graph generated by a knowledge compiler is logically
-equivalent to the input formula.  We can then validate a standard or
-weighted count from that representation.
-%% Their tool
-%% chain can be applied to some programs that are pure model counters,
-%%while ours can be applied to pure knowledge compilers.
-Nonetheless,
-we compare the performance of our toolchain to theirs as part of our
-experimental evaluation in Appendix~\ref{app:experiments}.
-
-We know of no other
-work on certifying the output of a knowledge compiler.
-
-\section{Logical Foundations}
-\label{section:logical:foundations}
-
-  Let $\varset$ denote a set of Boolean variables, and let $\assign$
-  be an {\em assignment} of truth values to some subset of the
-  variables, where $0$ denotes false and $1$ denotes true, i.e.,
-  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
-  \subseteq \varset$.  We say the assignment is {\em total} when it
-  assigns a value to every variable ($\varset' = \varset$), and that
-  it is {\em partial} otherwise.
-  The set of all possible total assignments over
-  $\varset$ is denoted $\uassign$.
-
-For each variable $x \in \varset$,
-  we define the {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
-  negation of $x$. An
-  assignment $\assign$ can be viewed as a set of literals, where
-  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
-  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
-$\obar{\lit} = x$ when $\lit = \obar{x}$.
-
-\begin{definition}[Boolean Formulas]
-  The set of Boolean formulas is defined recursively.  Each
-  formula $\phi$ has an associated {\em dependency set}
-  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
-  consisting of total assignments that satisfy the formula:
-  \begin{enumerate}
-  \item Boolean constants $0$ and $1$ are Boolean formulas,
-    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
-  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
-    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
-  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
-    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
-  \end{enumerate}
-\label{def:boolean}
-\end{definition}
-
-  We highlight two special classes of Boolean formulas.  A formula is
-  in {\em negation normal form} when negation is applied only to variables.  A
-  formula is in {\em conjunctive normal form} (CNF) when i) it is in
-  negation normal form, and ii) sum is applied only to literals.  A CNF
-  formula can be represented as a set of {\em clauses}, each of which is a
-  set of literals.  Each clause represents the sum of the
-  literals, and the formula is the product of its clauses.  We use
-  set notation to reference the clauses within a formula and the
-  literals within a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause and the literal as a {\em unit} literal.
-This literal must be assigned value $1$ by any satisfying assignment of the formula.
-
-\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
-  A {\em partitioned-operation formula}
- satisfies the following for all product and sum operations:
-      \begin{enumerate}
-      \item The arguments to each product must have disjoint dependency sets.  That is, operation
-        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \item The arguments to each sum must have disjoint models.  That is, operation
-        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \end{enumerate}
-\end{definition}
-     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.
-
-  \section{Ring Evaluation of a Boolean Formula}
-
-We propose a general framework for summarizing properties of Boolean
-formulas along the lines of algebraic model counting~\cite{kimmig:jal:2017}.
-
-\begin{definition}[Commutative Ring]
-  A {\em commutative ring} $\ring$ is an algebraic structure
-  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
-  with elements in the set $\dset$ and with commutative and
-  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
-  such that multiplication distributes
-  over addition.  $\addident$ is the additive identity and $\mulident$ is
-  the multiplicative identity.  Every element $a \in \dset$ has an
-  {\em additive inverse} $-a$ such that $a + -a = \addident$.
-\label{def:ring}
-\end{definition}
-We write $a - b$ as a shorthand for $a + -b$.
-
-\begin{definition}[Ring Evaluation Problem]
-\label{def:ring_evaluation}
-  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
-  every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.
-
-  For Boolean formula $\phi$ and ring weight function $w$, the {\em ring evaluation problem} computes
-  \begin{eqnarray}
-    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
-  \end{eqnarray}
-  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
-\label{def:weight}
-\end{definition}
-
-Many important properties of Boolean formulas can be
-expressed as ring evaluation problems.  The
-(standard) {\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
-It can be cast as a ring evaluation problem by having $\radd$ and
-$\rmul$ be addition and multiplication over rational numbers and using
-weight function $w(x) = 1/2$ for every variable $x$.
-Ring evaluation of formula $\phi$ gives the {\em density} of
-the formula, i.e., the fraction of all possible total assignments that are
-models.  For $n = |\varset|$, scaling the density by $2^n$
-yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
-in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.
-
-The {\em weighted model counting}  problem is also defined over
-rational numbers.  Some formulations  allow
-independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
-$W(x) + W(\obar{x}) \not = 1$.
-We can cast this as a
-ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
-performing ring evaluation with weight function $w(x) = W(x)/r(x)$ for each
-variable $x$, and computing the weighted count
-as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
-Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.
-
-The {\em function hashing problem} provides a test
-of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
-finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
-$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
-if they are logically equivalent.
-%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
-If they are not equivalent, then
-the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
-least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
-Function hashing can therefore be used as part of a
-randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
-For example, it can compare different runs on a single formula,
-either from different compilers or from a single compiler with different configuration parameters.
-
-%We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.
-
-\section{Partitioned-Operation Graphs (POGs)}
-\label{sect:pog}
-
-Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
-\begin{proposition}
-\label{prop:ring:eval}
-Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
-\begin{eqnarray}
-%% V1
-\textstyle
-\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
-\textstyle
-\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
-\textstyle
-\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
-\end{eqnarray}
-\end{proposition}
-As is described in \cref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.
-
-A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
-with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
-When $(\nodeu,\nodev) \in E$,
-node $\nodev$ is said to be a {\em child} of node $\nodeu$.
-The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
-$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
-Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
-Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
-A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
-Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
-It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
-Every POG has a designated {\em root node} $\noder$.
-
-In essence, a POG represents a partitioned-operation
-formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
-$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
-Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.
-
-We define the {\em size} of POG $P$, written $|P|$, to be the
-the number of nodes labeled $\pand$ or $\por$ plus the number of edges from these nodes to their children.  Ring
-evaluation of $P$ can be performed with at most $|P|$ ring
-operations by traversing the graph from the terminal nodes up to
-the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
-The final result is then $\rep(\noder, w)$.
-
-POGs generalize the d-DNNF graphs in two ways:
-\begin{itemize}
-\item They allow negation of arbitrary nodes in the graph, not just
-  variables.
-  %, and so we allow it in the interest of generality.
-%%  Negation could be useful in some contexts, such as when
-%%  converting binary decision diagrams (BDDs) having complemented
-%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
-\item They allow arbitrary arguments to a $\por$ operation, as long as
-  these have disjoint models.  By contrast, each
-  sum node $\nodeu$ in a d-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
-  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
-%%Our generalization is provided to allow an
-%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
-%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
-\end{itemize}
-  These generalizations allow more flexibility in the POG
-  representation while maintaining the ability to efficiently perform ring evaluation.
-
-\section{Clausal Proof Framework}
-
-We write (possibly subscripted) $\theta$ for
-formulas encoded as clauses, possibly with extension variables.
-We write (possibly subscripted)
-$\phi$ for formulas
-that use no extension variables.
-
-A proof in our framework takes the form of a sequence of clause addition
-and deletion steps, with each step preserving the set of solutions to
-the original formula.
-The status of the proof at any step is represented as
-a set of {\em active} clauses $\theta$, i.e., those that
-have been added but not yet deleted.
-Our framework is based
-on {\em extended} resolution~\cite{Tseitin:1983}, where proof
-steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
-Let $Z$
-denote the set of extension variables occuring in formula $\theta$.
-Starting with input formula $\inputformula$,
-the proof must maintain the invariant that
-$\inputformula \ifandonlyif \exists Z\,\theta$.
-
-Clauses can be added in two different ways.  One is when they serve as
-the {\em defining clauses} for an extension variable.  This form
-occurs only when defining $\pand$ and $\por$ operations, as is
-described in Section~\ref{sect:cpog}.  Clauses can also be added or
-deleted based on {\em implication redundancy}.  That is, when clause
-$C$ satisfies $\theta \imply C$ for formula $\theta$, then it can either
-be added to $\theta$ to create the formula $\theta \cup \{C\}$ or it can be deleted
-from $\theta \cup \{C\}$ to create $\theta$.
-
-We use {\em reverse unit propagation} (RUP) to certify
-implication redundancy when adding or deleting
-clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
-RUP
-is the core rule supported by standard
-proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
-way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
-Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
-proved redundant with respect to formula $\theta$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
-{\em antecedent} clauses, such that each $D_i$ is in $\theta$.
-A RUP step
-proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
-that the combination of the antecedents plus the negation of $C$ leads
-to a contradiction.  The negation of $C$ is the formula
-$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
-\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
-clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
-check processes the clauses of the antecedent in sequence, inferring
-additional unit clauses.  In processing clause $D_i$, if all but one
-of the literals in the clause is the negation of one of the
-accumulated unit clauses, then we can add this literal to the
-accumulated set.  That is, all but this literal have been falsified,
-and so it must be set to true for the clause to be satisfied.  The
-final step with clause $D_k$ must cause a contradiction, i.e., all of
-its literals are falsified by the accumulated unit clauses.
-
-Compared to the proofs of unsatisfiability generated by SAT solvers,
-ours have important differences.  Most
-significantly, each proof step must preserve the set of solutions with respect to the input variables.
-By contrast, an unsatisfiability proof need only guarantee that
-no proof step causes a satisfiable set of clauses to become
-unsatisfiable.  As a consequence, our proofs must justify clause deletions as well as clause additions.
-
-\section{The CPOG POG Representation and Proof System}
-\label{sect:cpog}
-
-%% The proof system is based on extended
-%% resolution~\cite{Tseitin:1983} with reverse unit propagation
-%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
-%% proving that a set of clauses logically implies another clause.
-%% Alternate version
-%% knowledge compilation.  The proof system is based on extended
-%% resolution~\cite{Tseitin:1983} with reverse unit propagation
-%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
-%% proving that a set of clauses logically implies another clause.
-%%
-%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
-%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
-%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
-%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
-%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
-%%
-%% \begin{center}
-%%   \begin{tabular}{lcccc}
-%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
-%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
-%%   \midrule
-%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
-%%   \end{tabular}
-%% \end{center}
-%%
-%% RUP is an alternative formulation of resolution.  For target clause
-%% $C$, it can be seen that applying resolution operations to the
-%% antecedent clauses from right to left will derive a clause $C'$ such
-%% that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
-%% we then have $C' \rightarrow C$.  Compared to listing each resolution
-%% operation as a separate step, using RUP as the basic proof step makes
-%% the proofs more compact.
-%%
-%% As is shown in Figure~\ref{fig:chain}
-%% two programs are involved in producing a certified compilation of a formula:
-%% \begin{itemize}
-%% \item The {\em proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
-%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
-%% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
-%% \end{itemize}
-%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b
-
-A CPOG file provides both a declaration of a POG, as well as a checkable
-proof that a Boolean formula, given in conjunctive normal
-form, is logically equivalent to the POG\@.
-The proof format draws its inspiration from the LRAT~\cite{lrat} and
-QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
-Key properties include:
-\begin{itemize}
-  \item
-  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
-  Declaring a node $\nodeu$
-implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
-  encoding the product or sum operation.
-%  This is the only means for adding extension variables to the proof.
-\item Boolean negation is supported implicitly by allowing the
-  arguments of the $\por$ and $\pand$ operations to be literals and not just
-  variables.
-\item
-  The file contains explicit clause addition steps.
-  A clause can only be added if it is logically implied by the existing clauses.
-  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
-\item
-  The file contains explicit clause deletion steps.
-  A clause can only be deleted if it is logically implied by the remaining clauses.
-  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
-\item The checker must track the dependency set for every input and
-  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
-  The associated extension variable has a dependency set equal to the union of those of its arguments.
-\item Declaring a $\por$ operation requires a sequence of clauses
-  providing a RUP proof that the arguments are mutually exclusive.
-  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness
-%  Generalizing to $k$ arguments would require
-%  $k\,(k-1)/2$ proofs of disjointness.
-\end{itemize}
-
-\subsection{Syntax}
-\label{subsection:syntax}
-
-\begin{table}
-  \caption{CPOG Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
-  \label{tab:cpog:syntax}
-\centering{
-  \begin{tabular}{lllll}
-    \toprule
-    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
-    \midrule
-    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause} \\
-     & {\tt d} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause \\
-    \midrule
-    $C$    & {\tt p} & $V \; L^{*}$ {\tt 0}    &                  & Declare $\pand$ operation \\
-    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\
-    \midrule
-     & {\tt r} & $L$             &            & Declare root literal\\
-    \bottomrule
-  \end{tabular}
-  }
-\end{table}
-
-Table~\ref{tab:cpog:syntax} shows the declarations that can occur in a CPOG file.
-%% The checker is provided with the input formula as a separate file.
-As with other clausal proof formats, a variable is
-represented by a positive integer $v$, with the first ones being input
-variables and successive ones being extension variables.  Literal $\lit$
-is represented by a signed integer, with $-v$ being the logical negation of
-variable $v$.  Each clause is indicated by a positive integer
-identifier $C$, with the first ones being the IDs of the input clauses and successive
-ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
-such that any clause identifier $C'$ given in the hint when adding clause $C$ must have $C' < C$.
-
-The first set of proof rules are similar to those in other clausal
-proofs.
-Clauses can be added via RUP addition
-(command {\tt a}), with a sequence of antecedent clauses (the
-``hint'').
-Similarly for clause deletion (command {\tt d}).
-
-\begin{table}
-\caption{Defining Clauses for Product (A) and Sum (B) Operations}
-\begin{minipage}{0.54\textwidth}
 \begin{center}
-\begin{tabular}{cccccc}
-\multicolumn{6}{c}{(A).  Product Operation $\pand$}\\
-\toprule
-\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
-\midrule
-  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
-  $i\!+\!1$ & $-v$ & $\lit_1$  \\
-  $i\!+\!2$ & $-v$ & $\lit_2$  \\
-  & $\ldots$ \\
-  $i\!+\!k$ & $-v$ & $\lit_k$  \\
-\bottomrule
-\end{tabular}
-\end{center}
-\end{minipage}
-\begin{minipage}{0.44\textwidth}
-\begin{center}
-\begin{tabular}{cccc}
-\multicolumn{4}{c}{(B).  Sum Operation $\por$}\\
-\toprule
-\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
-\midrule
-  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
-  $i\!+\!1$ & $v$ & $-\lit_1$ \\
-  $i\!+\!2$ & $v$ & $-\lit_2$ \\
-\bottomrule
-$\;$ \\
-$\;$ \\
-\end{tabular}
-\end{center}
-\end{minipage}
-\label{tab:defining}
-\end{table}
-
-The declaration of a {\em product} operation, creating a node with operation $\pand$,
- has the form:
-\begin{center}
-\begin{tabular}{ccccccccc}
-  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
-  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\tt 0} \\
-\end{tabular}
-\end{center}
-Integer $i$ is a new clause ID, $v$ is a positive integer that does not
-correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$
-integers representing literals of existing variables.
-As Table~\ref{tab:defining}(A) shows,
-this declaration implicitly causes $k+1$ clauses to be added to the proof, defining extension variable $v$ to be the product of its arguments.
-
-The dependency sets for the arguments represented by each pair of
-literals $\lit_i$
-and $\lit_{j}$ must
-be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
-representing Boolean constant $1$.  The only clause added to the proof will be
-the unit literal $v$.  A reference to literal $-v$ then provides a way
-of representing constant $0$.
-
-The declaration of a {\em sum} operation, creating a node with operation $\por$, has the form:
-\begin{center}
-\begin{tabular}{ccccccc}
-  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$}
-\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
-\end{tabular}
-\end{center}
-Integer $i$ is a new clause ID, $v$ is a positive integer that does
-not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
-signed integers representing literals of existing variables.  Hint $H$
-consists of a
-sequence of clause IDs, all of which must be defining clauses for other POG operations.
-As Table~\ref{tab:defining}(B) shows,
-this declaration implicitly causes three clauses to be added to the proof, defining extension variable $v$ to be the sum of its arguments.
-The hint must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.
-
-Finally, the literal denoting the root of the POG is declared with the
-{\tt r} command.  It can occur anywhere in the file.  Except in degenerate cases, it
-will be the extension variable representing the root of a graph.
-%, but in
-%degenerate cases it can be the positive or negative literal of an
-%input variable.
-
-\subsection{Semantics}
-\label{subsection:semantics}
-
-The defining clauses for a product or sum
-operation uniquely define the value of its extension variable for any assignment of values to the argument variables.
-For the
-extension variable $u$ associated with any POG node $\nodeu$, we can therefore
-prove that any total assignment $\assign$ to the input variables which
-also satisfies the POG definitions will
-assign a value to $u$ such that $\assign(u) =
-1$ if and only if $\assign \in \modelset(\nodeu)$.
-
-%% A CPOG proof follows the same general form as a QRAT dual
-%% proof~\cite{bryant:cade:2021}---one that ensures that each clause
-%% addition and each clause deletion preserves equivalence.  With CPOG,
-%% however, clauses are defined both explicitly and implicitly.  Starting
-%% with the set of input clauses, the proof consists of a sequence of
-%% steps that both add and delete clauses.  Each addition must be truth
-%% preserving, that is, any satisfying total assignment for the set of clauses
-%% before clause addition should still be a satisfying assignment afterwards.
-%% Each deletion must be falsehood preserving.  That is, there can be no
-%% new satisfying assignments when the clause is deleted.
-
-The sequence of operator declarations, asserted clauses, and
-clause deletions represents a systematic transformation of the input formula
-into a POG formula.  Validating all of these steps serves to prove that the
-POG is logically equivalent to the input formula.
-At the completion of the proof, the following \textsc{final conditions} must hold:
-\begin{enumerate}
-\item There is exactly one remaining clause that was added via RUP
-  addition, and this is a unit clause consisting of root literal $r$.
-\item All of the input clauses have been deleted.
-\end{enumerate}
-In other words, at the end of the proof it must hold that the active clauses be exactly those
-in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
-of unit clause $\{r\}$ and the defining clauses of the POG nodes. Recognizing that
-any total assignment to the input variables implicitly defines the assignments to the extension variables,
-we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
-Let $\inputformula$ denote the input formula.
-The sequence of clause addition steps provides a {\em forward implication} proof that
-$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
-assignment $\assign$ satisfying the input formula must also satisfy
-the formula represented by the POG\@.
-Conversely,
-each proof step that deletes an input clause proves that any
-total assignment $\alpha$ that falsifies the clause must
-falsify $\phi_\noder$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
-that
-$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.
-
-Appendix~\ref{app:cpog:example} shows the CPOG description for
-an input formula with five clauses, yielding a POG with five
-nonterminal nodes.  It explains how the clause addition and deletion
-steps yield a proof of equivalence between the input formula and its POG
-representation.
-
-\section{Generating CPOG from d-DNNF}
-\label{section:generating:cpog}
-
-%%Figure~\ref{fig:chain} illustrates one method of generating a CPOG
-%%file from an input formula in conjunctive normal form.  Our toolchain
-%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
-%%but it would also work for other tools that use a top-down search
-%%strategy to generate a d-DNNF representation.
-A d-DNNF
-graph can be directly translated  into a POG,
-although in doing this conversion,
-our program also performs simplifications to
-eliminate Boolean constants.
-Except in degenerate cases,
-%where the formula is unsatisfiable or a tautology,
-we can therefore assume
-that the POG does not contain any constant nodes.
-In addition, negation is only
-applied to variables, and so we can merge the negation operations into the terminal nodes,
-viewing the POG as consisting
-of {\em literal} nodes corresponding to input variables and their negations, along with
-{\em nonterminal} nodes, which can be further classified as {\em product} and {\em sum} nodes.
-
-\subsection{Forward Implication Proof}
-
-For input formula $\inputformula$ and its translation into a POG $P$
-with root node $\noder$, the most challenging part of the proof is to
-show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
-that any total assignment $\assign$ that is a model of $\inputformula$
-and the POG definition clauses
-will yield $\assign(r) = 1$, for root literal $r$.  This part of the
-proof consists of a series of clause assertions leading to one adding
-$\{r\}$ as a unit clause.  We have devised two methods of generating this
-proof.  The {\em monolithic} approach makes just one call to a
-proof-generating SAT solver and has it determine the relationship
-between the two representations.  The {\em structural} approach
-recursively traverses the POG, generating proof obligations at each
-node encountered.  It too may require calls to a proof-generating SAT
-solver.
-
-As notation,
-let $\psi$ be a subset of the clauses in $\inputformula$.
-For partial assignment
-$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
-obtained from $\psi$ by: i) eliminating any
-clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
-ii) for the remaining clauses eliminating those literals $\lit$ for
-which $\passign(\lit) = 0$, and iii) eliminating any duplicate clauses.
-In doing these simplifications, we also track the {\em provenance}
-of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
-More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov_{\passign}(C, \psi)$ denote
-those clauses $C' \in \psi$, such that
-$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
-We then extend the definition of $\prov$ to any simplified formula
-$\gamma$ as $\prov_{\passign}(\gamma, \psi) = \bigcup_{C \in \gamma} \prov_{\passign}(C, \psi)$.
-
-The monolithic approach
-takes advantage of the clausal representations of
-the input formula $\inputformula$ and the POG formula $\phi_\noder$.
-We can express the negation of $\phi_\noder$ in clausal form as
-$\theta_{\obar{\noder}} \doteq \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
-Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
-when the formula $\inputformula \land \theta_{\obar{\noder}}$
-is unsatisfiable, where the
-conjunction can be expressed as the union
-of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
-For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CPOG proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
-Our experimental results show
-that this approach can be very effective and generates short proofs
-for smaller problems, but it does not scale well enough for routine
-use.
-
-We describe the structural approach to proof generation as a recursive procedure
-$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
-node $\nodeu$, partial assignment
-$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
-The procedure adds a number of clauses to the proof, culminating with
-the addition of the {\em target} clause:
-$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
-that $(\bigwedge_{\lit \in \passign} \lit) \imply u$, i.e.,
-that any total
-assignment $\assign$ such that $\passign \subseteq \assign$
-will assign $\assign(u) = 1$.
-The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
-The result will therefore be to add unit clause $\{r\}$ to the proof.
-Here we present a correct, but somewhat inefficient formulation of
-$\validate$.  We then refine it with some optimizations.
-
-The recursive call $\validate(u, \passign, \psi)$ assumes that we have
-traversed a path from the root node down to node $\nodeu$, with the
-literals encountered in the product nodes forming the partial
-assignment $\passign$.  The set of clauses $\psi$ can be a proper
-subset of the input clauses $\inputformula$ when a product node has caused
-a splitting into clauses containing disjoint variables.
-The subgraph with root node $\nodeu$ should be a POG representation of the formula
-$\simplify{\psi}{\passign}$.
-%% Having the
-%% routine add its target clause indicates that any total assignment
-%% $\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.
-
-The process for generating such a proof depends on the form of node $\nodeu$:
-\begin{enumerate}
-\item If $\nodeu$ is a literal $\lit'$, then the formula
-  $\simplify{\psi}{\passign}$ must consist of the single unit clause
-  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
-  Any of these can
-  serve as the target clause.
-\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
-  then, since the node originated from a d-DNNF graph, there must be
-  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
-  product node containing a literal node for $x$ as a child.  In either case, we
-  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
-  This will cause the addition of the target clause
-  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
-  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
-  causing the addition of the target clause
-  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-  These recursive results can be combined with the second and third defining clauses for $\nodeu$
-(see Table~\ref{tab:defining}(B))
-  to generate the target clause for $\nodeu$.
-\item If $\nodeu$ is a product node, then we can divide its children
-  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
-  \begin{enumerate}
-    \item For each literal
-  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
-  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
-  cases, this can be done by simple Boolean constraint propagation (BCP).
-  In other cases, we must prove that the formula
-  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
-  do so by writing the formula to a file, invoking a proof-generating
-  SAT solver, and then converting the generated unsatisfiability proof
-  into a sequence of clause additions in the CPOG file.
-\item For a single nonterminal child ($k = 1$), we recursively call
-  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$.
-\item For multiple nonterminal children ($k > 1$),
-  it must be the case that the clauses in
-  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
-  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
-  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
-  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
-  literals.
-  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
-  We recursively call
-  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \psi_i\right)$.
-\end{enumerate}
-  We then generate the target clause for node $\nodeu$,
-creating the hint by combining the results from the BCP and SAT calls for
-  the literals, the recursively computed target clauses, and all but
-  the first defining clause for node $\nodeu$
-(see Table~\ref{tab:defining}(A)).
-\end{enumerate}
-Most of these steps involve a polynomial number of
-operations per recursive call, with the exception of those that call
-a SAT solver to validate a literal.
-
-\subsection{Reverse Implication Proof}
-
-Completing the equivalence proof of input formula $\inputformula$ and its POG
-representation with root node $\noder$ requires showing that
-$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
-CPOG framework by first deleting all asserted clauses, except for the
-final unit clause for root literal $r$, and then deleting all of the
-input clauses.
-%%  Suppose at some point in this process that we want to
-%% delete clause $C$ such that the remaining set of clauses will form the
-%% CNF formula $\phi$.  Deleting $C$ requires proving that
-%% $\phi \imply C$.
-
-The asserted clauses can be deleted in reverse order, using the same
-hints that were used in their original assertions.  By reversing the
-order, those clauses that were used in the hint when a clause was
-added will still remain when it is deleted.
-
-Each input clause deletion can be done as a single RUP step.  The
-proof generator constructs the hint sequence from the defining
-clauses of the POG nodes via a single, bottom-up pass through the
-graph.  The RUP deletion proof for input clause $C$ effectively proves that any
-total assignment $\assign$ that satisfies the POG definition clauses
-but does not satisfy $C$ will yield
-$\assign(r) = 0$.  It starts with the set of literals
-$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
-assignment $\assign$ to falsify clause $C$.
-It then
-adds literals via unit propagation until a
-conflict arises.    Unit literal $r$ gets
-added right away, setting up a potential conflict.
-
-Working upward through the graph, node $\nodeu$ is {\em marked} when
-the collected set of literals force $u$ to evaluate to $0$.  When marking $\nodeu$, the
-program adds $\obar{u}$ to the RUP literals and adds the appropriate
-defining clause to the hint.  A literal node for
-$\lit$ will be marked if $\lit \in C$, with no hint required.  If
-product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
-$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
-added to the hint.  Marking sum node $\nodeu$ requires that its two children are marked.
-The first defining
-clause for this node (see Table~\ref{tab:defining}(B)) will then be added to the hint.  At the very end, the program
-(assuming the reverse implication holds) will attempt to mark root
-node $\noder$, which would require $\assign(r) = 0$, yielding a
-conflict.
-
-It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
-each clause deletion requires a single RUP step having a hint with length
-bounded by the number of POG nodes.
-
-\section{Optimizations}
-
-The performance of the structural proof generator for forward implication, both in its execution time and
-the size of the proof generated, can be improved by two optimizations
-described here.  A key feature is that they do not require any changes
-to the proof framework---they build on the power of extended
-resolution to create new logical structures.  They
-involve declaring new product nodes to encode products of literals.
-These nodes are not part of the POG
-representation of the formula; they serve only to enable the forward
-implication proof.  Here we summarize the two optimizations.
-Appendix~\ref{app:optimizations} provides more details.
-
-{\em Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
-having many---tens or even hundreds---of literals as children.  The
-earlier formulation of $\validate$ considers each literal $\lit \in L$
-separately, calling a SAT solver for every literal that cannot be
-validated with BCP\@.  Literal grouping handles
-all of these literals together.  It defines product node $\nodev$
-having the literals as children.  The goal then becomes to prove that
-any total assignment must yield 1 for extension variable $v$.  Calling
-a solver with $v$ set to 0 yields an unsatisfiability proof that can
-be mapped back to a sequence of clause additions in the CPOG file validating
-all of the literals.
-
-{\em Lemmas:} Our formulation of $\validate$ requires each call at a
-node $\nodeu$ to recursively validate all of its children.  This
-effectively expands the graph into a tree, potentially requiring an
-exponential number of recursive steps.  Instead, for each node
-$\nodeu$ having $\indegree(\nodeu) > 1$, the program can define and
-generate the proof of a {\em lemma} for $\nodeu$ when it is first reached by
-a call to $\validate$ and then apply this lemma for this and
-subsequent calls.  The lemma states that the POG with root node
-$\nodeu$ satisfies forward implication for a formula
-$\gamma_{\nodeu}$, where some of the clauses in this formula are
-input clauses from $\inputformula$, but others are simplified
-versions of input clauses.
-The key idea is to introduce product nodes to encode (via DeMorgan's
-Laws) the simplified clauses and have these serve as lemma arguments.
-
-The combination of these two optimization guarantees that i) each call
-to $\validate$ for a product node will cause at most one invocation of
-the SAT solver, and ii) each call to $\validate$ for any node $\nodeu$
-will cause further recursive calls only once.  Our experimental
-results (Appendix~\ref{app:experiments}) shows that these
-optimizations yield substantial benefits.
-
-\input{lean}
-
-\section{Implementations}
-We have implemented versions of the three programs that, along with
-the \dfour{} knowledge compiler, form the toolchain illustrated in
-Figure~\ref{fig:chain}\footnote{All tools, along with a set of 
-benchmark problems are available at \url{https://github.com/rebryant/cpog}}.  The proof generator is the same in both
-cases, since it need not be trusted.
-Our {\em verified}
-versions of the proof checker and ring evaluator have been formally
-verified within the \lean{} theorem prover.  Our long term goal is to
-rely on these.  Our {\em prototype} versions are written in C
-(checker) and Python (ring evaluator).  These are currently faster and
-more scalable, but we anticipate their need will diminish as the
-verified version is further optimized.
-%% Our long-term goal is to have a toolchain suitable for use in a
-%% knowledge compiler competition, along the lines of the annual
-%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
-%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
-%% Our experiments
-%% consider the benchmark problems and computing capabilities that would
-%% arise in such a competition and evaluate whether our implementationed toolchain
-%% could handle these problems.  As with the original proof-generating
-%% SAT competitions, we assume that the proofs at a competition could be
-%% checked with the unverified checker.  Having a verified checker
-%% that works on some subset of the benchmarks would provide an
-%% additional degree of certainty.
-
-Our proof generator is written in C/C++ and uses
-\cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
-proof steps back into hinted CPOG clause additions, the generator can
-use either its own RUP proof generator, or it can invoke
-\dtrim{}~\cite{RAT}.  The latter yields shorter proofs and scales well
-to large proofs, but each invocation has a high startup cost.  We
-therefore only use it when solving larger problems (currently ones
-with over 1000 clauses).
-
-The proof generator can optionally be instructed to generate a {\em
-one-sided} proof, providing only the reverse-implication portion of the proof via
-input clause deletion.  This can provide useful information---any
-assignment that is a model for the compiled representation
-must also be a model for the input formula---even when
-full validation is impractical.
-
-Our prototype ring evaluator can perform both standard and weighted
-model counting.  It performs arithmetic over a subset of the rationals
-we call $\drational$, consisting of numbers of the form $a \cdot 2^{b}
-\cdot 5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented
-to have arbitrary range.  Allowing scaling by powers of 2 enables the
-density computation and rescaling required for standard model
-counting.  Allowing scaling by powers of both 2 and 5 enables exact
-decimal arithmetic, handling the weights used in the weighted model
-counting competitions.  To give a sense of scale, the counter
-generated a result with 260,909 decimal digits
-for one of the weighted benchmarks.
-
-\section{Experimental Evaluation}
-\label{sect:experimental}
-
-We summarize the results of our experiments here.  
-Appendix~\ref{app:experiments} provides a more complete description. For our
-evaluation, we used the public benchmark problems from the 2022
-standard and weighted model competitions.  We found that there were
-180 unique CNF files among these, ranging in size from 250 to
-2,753,207 clauses.
-%%We ran our programs on a processor with 64~GB of
-%%memory and having an attached high-speed, solid-state
-%%disk.
-With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
-benchmark problems.  Our proof generator was able to convert all of
-these into POGs, with their declarations ranging from 304 to
-2,761,457,765 (median 848,784) defining clauses.
-%(The
-%maximum count overflowed the overflowed the 32-bit signed integers we
-%used to represent clause identifiers in both the proof generator and
-%checker.)
-We ran our proof generator with a time limit of 10,000 seconds.
-It was able to generate full proofs for 108 of the problems and
-one-sided proofs for an additional 9 of them, leaving just 7 with no
-verification.  The prototype checker successfully verified all of the generated
-proofs.  The longest runtime for the combination of proof generator
-and checker for a full proof was 13,145 seconds.
-It is worth noting that, to date, we have not found
-any errors in the d-DNNF graphs generated by \dfour{}.
-
-We found that the monolithic approach for generating the forward
-implication proof works well for smaller POGs, but it becomes
-unreliable and inefficient for larger ones.
-%% These experiments suggest
-%% a possible hybrid approach, stopping the recursion of the structural
-%% approach and shifting to monolithic mode once the subgraph size is below some
-%% threshold.
-We also found that our two optimizations: literal grouping and
-lemmas can provide substantial improvements in proof size and runtime.
-Running the verified proof checker in \lean{} showed the same scaling
-with respect to the proof size as did the prototype checker, but it
-required, on average, around 5.9 times longer.
-
-Finally, the runtimes for the MICE toolchain versus our CPOG toolchain
-showed little correlation, reflecting the fact that the two solve
-different problems and use different approaches.  In general,
-our CPOG toolchain showed better scaling, in part due to its ability
-to control the recursion through lemmas.
-
-\section{Conclusions}
-\label{sect:future}
-
-We are hopeful that having checkable proofs for knowledge compilers
-will allow them to be used in applications where high levels of trust
-are required, and that it will provide a useful tool for developers of
-knowledge compilers.
-%%Although our current implementation only handles
-%%the outputs of one particular program, it could be adapted
-Our experiments demonstrate that our toolchain can already
- handle problems nearly at the limits of current knowledge compilers.
-Further engineering and optimization of our proof
-generator and checker could improve their performance and capacity substantially.
-
-\newpage
-
-%% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
-%% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
-%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{1 \leq i \leq k}
-%% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
-%% such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
-%% \cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
-%% such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
-%% \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
-%% of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
-%% for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
-%% the sum of $t_1, \ldots, t_k$.
-
-\bibliography{references}
-
-\newpage
-
-\appendix
-
-\section{CPOG Example}
-\label{app:cpog:example}
-
-\begin{figure}
-\begin{minipage}{0.58\textwidth}
-(A)  Input Formula\\[1.2ex]
 \begin{tabular}{lll}
@@
 \midrule
-1 & \texttt{-1 3 -4} & \texttt{0} \\
-2 & \texttt{-1 -3 4} & \texttt{0} \\
-3 & \texttt{3 -4} & \texttt{0}\\
-4 & \texttt{1 -3 4} & \texttt{0} \\
-5 & \texttt{-1 -2} & \texttt{0} \\
+1 & \texttt{1 -4} & \texttt{0} \\
+2 & \texttt{2 -4} & \texttt{0} \\
+3 & \texttt{3  4} & \texttt{0}\\
 \bottomrule
 \end{tabular}
-\\[1.8ex]
-(C) POG Declaration\\[1.2ex]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
-\midrule
-6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
-9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
-12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
-15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\
-18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
-22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
- & \texttt{r 10} && Root $r = s_{10}$\\
-\bottomrule
-\end{tabular}
+\end{center}
+
+\begin{figure}
+  \begin{minipage}{0.48\textwidth}
+    (A) POG representation of $\inputformula$ \\[1.2ex]
+    \input{dd/eg-preproj}
   \end{minipage}
-\begin{minipage}{0.35\textwidth}
-(B) POG Representation \\
-\input{dd/eg4}
+  \begin{minipage}{0.48\textwidth}
   \end{minipage}
-%% Break
-\\[2.5ex]
-\begin{minipage}{0.42\textwidth}
-(D) Implicit Clauses\\[1.2ex]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
-\midrule
-\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
-\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
-\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
-\midrule
-\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
-\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
-\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
-\midrule
-\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
-\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
-\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
-\midrule
-\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
-\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
-\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
-\midrule
-\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
-\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
-\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
-\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
-\midrule
-\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
-\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
-\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
-\bottomrule
-\end{tabular}
+
+  \begin{minipage}{0.48\textwidth}
+    (B) After projecting variable $y$ \\[1.2ex]
+    \input{dd/eg-proj}
   \end{minipage}
-\begin{minipage}{0.49\textwidth}
-(E) CPOG Assertions\\[1.2ex]
-\begin{tabular}{llllll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
-\midrule
-\texttt{25} & \texttt{a 5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
-\texttt{26} & \texttt{a 6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
-\texttt{27} & \texttt{a 3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
-\texttt{28} & \texttt{a 7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
-\texttt{29} & \texttt{a 8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
-\texttt{30} & \texttt{a 5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
-\texttt{31} & \texttt{a 6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
-\texttt{32} & \texttt{a 3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
-\texttt{33} & \texttt{a 7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
-\texttt{34} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
-\texttt{35} & \texttt{a 1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
-\texttt{36} & \texttt{a 10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
-\bottomrule
-\end{tabular}
-\\[1.5ex]
-(F) Input Clause Deletions\\[1.2ex]
-\begin{tabular}{lllll}
-  \toprule
- \multicolumn{3}{l}{CPOG line} & Explanation\\
-\midrule
- \texttt{d 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
- \texttt{d 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
- \texttt{d 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
- \texttt{d 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
- \texttt{d 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
-\bottomrule
-\end{tabular}
+  \begin{minipage}{0.48\textwidth}
+    (C) Replace $y$ by Skolem formula $x_1 \land x_2$ \\[1.2ex]
+    \input{dd/eg-proj-skolemize}
   \end{minipage}
-\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F)}
-\label{fig:eg4:all}
+  \caption{POG Representation of formula (A), and two ways of representing projections (B) and (C).
+  A black dot on an edge denotes Boolean negation.}
+  \label{fig:eg-proj}
 \end{figure}
 
-Figure \ref{fig:eg4:all} illustrates an example formula and shows how
-the CPOG file declares its POG representation.  The input formula (A)
-consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
-$x_4$.  The generated POG (B) has six nonterminal nodes representing
-four products and two sums.  We name these by the node
-type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
-extension variable.
-% We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
-  The first part of the CPOG file (C) declares
-these nodes using clause IDs that increment by three or four,
-depending on whether the node has two children or three.  The last two
-nonzero values in the sum declarations are the hint providing the
-required mutual exclusion proof.
+Figure \ref{fig:eg-proj}(A) shows the POG representation of the
+example formula, derived from the d-DNNF representation produced by
+\dfour{}.  We can see that \dfour{} split on variable $y$ at the top level, with the representation of
+$\obar{y} \land \inputformula$ on the left and 
+$y \land \inputformula$ on the right.
 
-\subsection{Node Declarations}
+Figure \ref{fig:eg-proj}(B) shows the result of a method for projecting out variable
+$y$ by computing the intersection of the sets of data models for
+$\aassign(y) = 0$ and $\aassign(y) = 1$.  For this example, these are
+given by the formula $x_1 \land x_2 \land x_3$, represented by POG
+node $\nodep_5$.  Using DeMorgan's Laws, we can use a sum node to
+``subtract'' these models from those given by
+$\simplify{\inputformula}{\obar{y}}$, so that the top-level sum
+$\nodes_8$ satisfies the disjoint-model requirement.
 
-We step through portions of the file to provide a better understanding of the CPOG proof framework.
-Figure
-\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
-defined by the POG operation declarations.  These do not appear in the
-CPOG file.  Referring back to the declarations of the sum nodes in
-Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
-$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
-\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
-$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
-have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
-clause IDs 16 and 19 as the hint.  These form a RUP proof for the clause
-$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
-$\nodes_{10}$ have disjoint models.
+Figure \ref{fig:eg-proj}(C) shows the result of replacing $y$ by the
+Skolem formula
+$\skolem_y = x_1 \land x_2$.  In that case node
+$\nodep_6$ of (A) becomes simply $x_1 \land x_2$, and this can be used
+to replace $\obar{y}$ as a child of node $\nodep_5$.
 
-\subsection{Forward Implication Proof}
+Observe that both POGs (B) and (C) use an exclusion term so that the formula
+on the left encodes those assignments satisfying $x_3 \land \neg (x_1 \land x_2)$.
 
-Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
-leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $\nodes_{10}$ is implied by the input clauses, i.e.,
-any total assignment $\assign$
-satisfying the input clauses must have $\assign(s_{10}) = 1$.
-Working backward, we can see that
-clause 29 indicates that variable $p_8$ will be implied by the input
-clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
-be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
-hint for clause 36.
+\section{Research Questions}
 
-\subsection{Reverse Implication Proof}
-
-Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
-delete the input clauses.  Consider the first of these, deleting
-input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
-that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
-The proof proceeds by first assuming that the clause is false, requiring
-$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
-clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
-clauses that cause unit propagation.  Hint clauses 8 and 10 force the
-assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
-$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
-via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
-completing the RUP proof.  The deletion hints for the other input
-clauses follow similar patterns---they work from the bottom nodes of
-the POG upward, showing that any total assignment that falsifies the clause
-must assign $\assign(s_{10}) = 0$.
-
-Deleting the asserted clauses is so simple that we do not show it.  It
-involves simply deleting the clauses from clause number 35 down to
-clause number 25, with each deletion using the same hint as were used
-to add that clause.  In the end, therefore, only the defining clauses
-for the POG nodes and the unit clause asserting $s_{10}$ remain,
-completing a proof that the POG is logically equivalent to the input
-formula.
-
-Observe that the forward implication proof shown in Figure~\ref{fig:eg4:all}  must ``visit'' nodes
-$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
-(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).
-Section~\ref{app:lemma:eg} shows how to use a lemma such that these nodes are each only visited once.
-
-\section{Optimizations for Forward Implication Proofs}
-
-The two optimizations we have implemented exploit the power of our
-general resolution framework to define new structures within the
-proof.
-They create new product nodes that are not part
-of the POG representation.
-
-\label{app:optimizations}
-
-\subsection{Literal Grouping}
-
-A single recursive step of $\validate$ can encounter product nodes
-having many literals as children.  The naive formulation of $\validate$
-considers each literal $\lit \in L$ separately.
-Literal grouping allows all literals to be validated with a single call to a SAT solver.
-It collects those literals
-$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
-product node $\nodev$ having these literals as children.  The goal
-then becomes to prove that any total assignment must yield 1 for extension
-variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
-  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
-  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
-  input clauses and the defining clauses for $\nodev$ into the hints.
-
-\subsection{Lemmas}
-\label{app:lemma}
-
-As we have noted, the recursive calling of $\validate$ starting at
-root $\noder$ effectively expands the POG into a tree, and this can
-lead to an exponential number of calls.
-These shared subgraphs arise when the knowledge compiler employs {\em clause caching}
-to detect that the simplified set of
-clauses arising from one partial assignment to the literals matches that
-of a previous partial assignment~\cite{darwiche:aaai:2002}.
-When this d-DNNF node is translated into POG
-node $\nodeu$, the proof generator can assume (and also check), that
-there is a simplified set of clauses $\gamma_{\nodeu}$
-for which the subgraph with root $\nodeu$ is its POG representation.
-
-The proof generator can exploit the sharing of subgraphs
-by constructing and proving a {\em lemma} for each node
-$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
-total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ and the defining clauses for the POG must yield
-$\assign(u) = 1$.  This lemma is then invoked for every node having
-$\nodeu$ as a child.
-As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.
-
-The challenge for implementing this strategy is to find a way to
-represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CPOG file.  Some may be
-unaltered input clauses, and these can be used directly.  Others,
-however will be clauses that do not appear in the input formula.  We
-implement these by adding POG product nodes to the CPOG file to create
-the appropriate clauses.  Consider an {\em argument} clause
-$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
-define product node $\nodev$ with arguments
-$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, then by
-Table~\ref{tab:defining}(A), we will introduce a defining clause
-$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
-  synthetic} clause having $\obar{v}$ as the {\em guard literal}.
-That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
-  activate} the clause, causing it to represent argument clause $C$.  On the other
-hand, a partial assignment with $\passign(v) = 1$ will
-cause the clause to become a tautology and therefore have no effect.
-
-Suppose for every clause $C_j \in \gamma_{\nodeu}$ that does not correspond to
-an input clause, we generate a synthetic clause $C'_j$ with guard literal
-$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma'_{\nodeu}$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
-while input clauses in $\gamma_{\nodeu}$ are left unchanged.
-Let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
-Invoking $\validate(\nodeu, \lassign, \gamma'_{\nodeu})$
- will then prove a lemma, given by the target clause
- $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
- showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u) = 1$.
-
-Later, when node $\nodeu$ is encountered by a call to $\validate(\nodeu, \passign, \psi)$, we invoke the lemma
-by showing that each synthetic clause
-$C_j$ matches some simplified clause in $\simplify{\psi}{\passign}$.  More precisely,
-for $1 \leq j \leq m$,
-we use clause addition to assert the clause
-$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$,
-showing that synthetic clause $C_j$ will be activated.
-Combining the lemma with these activations provides a derivation of the target clause for the call to $\validate$.
-
-Observe that the lemma structure can be hierarchical, since a shared
-subgraph may contain nodes that are themselves roots of shared
-subgraphs.  Nonetheless, the principles described allow the
-definition, proof, and applications of a lemma for each shared node in
-the graph.  For any node $\nodeu$, the first call to
-$\validate(\nodeu, \passign, \psi)$ may require further recursion,
-but any subsequent call can simply reuse the lemma proved by the first call.
-
-\subsection{Lemma Example}
-\label{app:lemma:eg}
-
-\begin{figure}
-(A) Additional nodes\\[1.0em]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
-\midrule
-25 & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
-28 & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
-\bottomrule
-\end{tabular}
-\\[1.0em]
-(B) Implicit Clauses\\[1.2em]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
-\midrule
-\texttt{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
-\texttt{26} & \texttt{-11 -3} & \texttt{0} & \\
-\texttt{27} & \texttt{-11 4} & \texttt{0} & \\
-\midrule
-\texttt{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
-\texttt{29} & \texttt{-12 3} & \texttt{0} & \\
-\texttt{30} & \texttt{-12 -4} & \texttt{0} & \\
-\bottomrule
-\end{tabular}
-\\[1.0em]
-(C) CPOG Assertions\\[1.0em]
-\begin{tabular}{llllll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
-\midrule
-\multicolumn{6}{l}{Lemma Proof} \\
-\texttt{31} & \texttt{a 5 11 12 3} & \texttt{0} & \texttt{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
-\texttt{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \texttt{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
-\texttt{33} & \texttt{a 3 7 11 12} & \texttt{0} & \texttt{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
-\texttt{34} & \texttt{a 7 11 12} & \texttt{0} & \texttt{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
-\midrule
-\multicolumn{6}{l}{Lemma Application \#1} \\
-\texttt{35} & \texttt{a -11 1} & \texttt{0} & \texttt{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
-\texttt{36} & \texttt{a -12 1} & \texttt{0} & \texttt{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
-\texttt{37} & \texttt{a 7 1} & \texttt{0} & \texttt{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
-\midrule
-\texttt{38} & \texttt{a 8 1} & \texttt{0} & \texttt{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
-\midrule
-\multicolumn{6}{l}{Lemma Application \#2} \\
-\texttt{39} & \texttt{a -11 -1} & \texttt{0} & \texttt{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
-\texttt{40} & \texttt{a -12 -1} & \texttt{0} & \texttt{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
-\texttt{41} & \texttt{a 7 -1} & \texttt{0} & \texttt{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
-\midrule
-\texttt{42} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
-\texttt{43} & \texttt{a 1 10} & \texttt{0} & \texttt{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
-\texttt{44} & \texttt{a 10} & \texttt{0} & \texttt{43 24 42} & \texttt{0} & $s_{10}$ \\
-\bottomrule
-\end{tabular}
-\caption{Example of lemma definition, proof, and application}
-\label{fig:eg4:lemmas}
-\end{figure}
-
-Figure~\ref{fig:eg4:lemmas} shows an alternate forward implication
-proof for the example of Figure~\ref{fig:eg4:all} using a lemma to
-represent the shared node $\nodes_7$.  We can see that the POG with
-this node as root encodes the Boolean formula $x_3 \leftrightarrow x_4$, having a CNF representation consisting of the clauses
-$\{x_3 ,\, \obar{x}_4\}$ and $\{\obar{x}_3 ,\, {x}_4\}$.  The product node
-declarations shown in Figure~\ref{fig:eg4:lemmas}(A) create synthetic
-clauses 25 and 28 to encode these arguments with activating literals
-$\obar{v}_{11}$ and $\obar{v}_{12}$, respectively.  Clauses 31--34
-then provide a proof of the lemma, stating that any assignment
-$\assign$ that activates these clauses will  assign $1$ to $s_7$.
-Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
-will cause the first synthetic clause to activate due to input clause
-3, and it will cause the second synthetic clause to activate due to
-input clause 4.  From this, clause 37 can use the lemma to state that
-assigning $0$ to $x_1$ will cause $s_7$ to evaluate to $1$.  Similarly,
-clauses 39 and 40 serve to activate the synthetic clauses when
-$\assign(x_1) = 1$, due to input clauses 1 and 2, and clause 41 then
-uses the lemma to state that assigning $1$ to $x_1$ will cause $s_7$ to
-evaluate to $1$.
-
-In this example, adding the lemma increases the proof length, but that
-is only because it is such a simple formula.
-
-\input{leanappendix}
-
-\section{Detailed Experimental Results}
-\label{app:experiments}
-
-Our experiments are designed to evaluate the following:
 \begin{itemize}
-\item
-  whether our toolchain can handle challenging benchmark problems,
-\item
-the effectiveness of some design choices and optimizations,
-\item
-the comparative performance of the prototype and verified checkers, and
-\item
-how the performance of our toolchain compares to that of the MICE model counter verifier.
+\item Algorithm to generate projection of POG over set of auxiliary variables $\avarset$
+  \begin{itemize}
+  \item Assume have clausal representation $\gamma_{\nodeu}$ for each node $\nodeu$.
+  \item Assume can call \dfour{} as subroutine to derive POG representing shared models (could blow up)
+  \item Assume POG derived from \dfour{} in NNF
+  \end{itemize}
+\item Algorithm to find Skolem formulas
+  \begin{itemize}
+  \item Easy if $y$ encoded as Tseitin variable
+  \item Probably OK if $y$ encoded using some variant of Greenbaum-Plaisted
+  \item Unclear how to solve in general
+  \end{itemize}
+\item Algorithm for forward implication proof
+  \begin{itemize}
+  \item Can use existing proof infrastructure
+  \item Use SAT solver to generate proofs of mutual exclusion and implication
+  \end{itemize}
+\item Reverse implication proof
+  \begin{itemize}
+  \item Need new proof rule(s).  Projected formula allows arbitrary assignment $\aassign$.
+  \item Skolem formulas might be helpful
+  \item Need to show that these form valid Skolemization
+  \end{itemize}
 \end{itemize}
 
-\subsection{Experimental Setup}
-
-As a test machine, we used a 2021~MacBook~Pro laptop having a 3.2~Ghz
-Apple M1 processor and with 64~GB of physical memory.  We also used an
-external 3~TB solid-state disk drive with an advertised read/write
-speed of one gigabit per second.  Despite being a laptop, this
-configuration is somewhat more powerful than the nodes typically
-found in server clusters.  The fast access to storage is especially
-important due to the very large files generated and processed by the
-tools.  Our tools generated individual files with over 160 gigabytes
-of data.
-
-For benchmarks, we downloaded 100 files each from the 2022 standard and weighted model counting competitions:
-\begin{center}
-       \url{https://mccompetition.org/2022/mc_description.html}
-\end{center}
-We found that 20 of these were duplicates across the two tracks,
-yielding a total of 180 unique benchmark problems, ranging in size
-from 250 to 2,753,207 clauses.
-
-Our default configuration for the proof generator used the structural
-method for the forward implication proof, with the two optimizations
-of literal grouping and lemmas.
-Running with a time limit of 4000 seconds\footnote{All measured times
-in this document are actual elapsed times, not CPU times.}, \dfour{}
-was able to generate d-DNNF representations for 124 of these, and it timed out on the
-rest.  Our proof generator was able to convert all of the generated
-d-DNNF graphs into POGs and determine how many defining clauses they
-would generate.  A POG operation with $k$ arguments requires $k+1$
-defining clauses, and so the total number of defining clauses in POG $P$ equals
-the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$ in Section~\ref{sect:pog}.
-The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.
-
-\subsection{Performance of the Proof Generator and Checker}
-
-\input{figure-runtime-d4-verify}
-\input{figure-clauses-d4-verify}
-
-For each of the 124 d-DNNF graphs generated by \dfour{}, we ran
-our proof generator and prototype checker, limiting proof
-generation to 10,000 seconds.  We ran the programs to generate and check
-one-sided proofs for the graphs, again with a time limit of 10,000
-seconds for proof generation.
-
-Figure~\ref{fig:d4:cpog} summarizes our results in terms of the time
-required by \dfour{} (X axis) versus the sum of the times for the
-proof generator and checker (Y axis).
-We were able to complete a full validation for 108 of the 124
-benchmarks, with times ranging from fractions of a second to 13,144
-seconds, with a median of 71.6 seconds.
-
-Relative to the runtime for \dfour{}, two problems ran faster with the
-generator and checker.  These were for problems having small numbers
-of models (relative to the number of variables), and so most of the
-time spent by both \dfour{} and our proof generator was in running a
-CDCL solver to search the very sparse solution space.  \cadical{}
-generally outperforms the miniSAT-based solver used by \dfour{}.  At
-the other extreme, one problem that required only 19 seconds for
-\dfour{} required 8657 seconds to generate a proof and 45 seconds to
-check it.  This benchmark appears to stem from weak performance
-by our implementation of BCP\@.  Overall the ratios
-between the combined generation plus checking times versus the time
-for \dfour{} had a harmonic mean of 5.5.
-
-Of the 16 benchmarks that could not be fully validated, one had so
-many defining clauses that it overflowed the 32-bit signed integers
-our programs use for clause identifiers.  Another exited due to the
-virtual memory limit imposed by the operating system, and the other 14
-timed out.  Of the 15 that did not overflow the clause counter, we
-were able to complete a one-sided verification for 9, but the other 6
-timed out.  Overall, we were able to provide some level of
-verification for 94\% of the benchmark problems.
-
-Figure~\ref{fig:defining:total} compares the number of defining
-clauses (X axis) with the total number of clauses (Y axis) for the 108
-problems that were fully verified.  These totals include the defining
-clauses for the POG, the additional defining clauses introduced to
-support literal grouping and lemmas, and the clauses added by RUP
-steps in the forward implication proof.  These totals ranged from 554
-to 770,382,773, with a median of 1,719,245.  The ratios of the total
-clauses versus definining clauses ranged from 1.54 to 6073, with a
-harmonic mean of 3.13.  The high numbers were for problems having very
-few models, and so the proof clauses must encode large proofs of
-unsatisfiability.
-
-\subsection{Monolithic Forward Implication Proofs}
-\label{app:experiment:monolithic}
-
-To assess the relative merits of the two approaches to forward
-implication proof generation, we ran the proof generator in monolithic
-mode on the 92 problems for which the combination of structural proof generation
-and checking required at most 3000 seconds and then kept only those
-measurements for which at least one of the two approaches had combined
-times (generation plus checking) of at most 1000 seconds.  There were
-a total of 83 such problems.  Figures~\ref{fig:monolithic:time} and
-\ref{fig:monolithic:clauses} show comparisons between the two
-approaches in terms of time and total clauses.
-
-Examining Figure~\ref{fig:monolithic:time}, we can see that the
-monolithic approach ran faster than the structural approach for 23 of
-the problems, including a case where the generator and checker
-required only one second, versus 71 seconds for the structural
-approach.  On the other hand, the monolithic approach timed out for 13
-of the cases and was slower for the other 37.  In general, we can see
-the monolithic approach faring worse on larger problems.
-
-Figure~\ref{fig:monolithic:clauses} shows the total number of clauses
-for the two approaches for the 70 cases where both approaches completed within 1000 seconds.
-As can be seen, the monolithic approach consistently produces
-shorter proofs, perhaps due to the benefits of the proof trimming by
-\dtrim{}.
-
-\input{figure-runtime-monolithic}
-\input{figure-clauses-monolithic}
-
-These experiments suggest that a hybrid of the two approaches might
-yield the best results in terms of consistency, runtime, and proof
-size.  It would begin the recursive descent of $\validate$ according
-to a structural approach, but monolithically generate
-the proof once it encounters a sufficiently small subgraph.
-
-\subsection{Impact of Optimizations}
-\label{app:experiment:optimize}
-
-\input{figure-clauses-optimizations}
-
-To assess the impact of the two optimizations: literal grouping and
-lemmas, we ran the proof generator on the 80 problems for which the
-combination of the two optimizations yielded proofs totaling at most
-10~million clauses.  Compared to the problems used to assess monolithic
-proof generation (Section~\ref{app:experiment:monolithic}), this set included one that had a short proof but
-long runtime and excluded four that had short runtimes but long
-proofs.  We then ran the proof generator on all three combinations of
-the optimizations being partially or totally disabled.
-Figure~\ref{fig:optimized:lessoptimized} shows how disabling these optimizations affected the
-total number of clauses in the proofs, where the X axis indicates the
-total number with both optimizations enabled, while the Y axis
-indicates the number with one (left) or both (right) optimizations
-disabled.
-
-As the figures illustrated, each optimization on its own can shorten
-the proofs, sometimes substantially, and the two combine to have even
-greater effect.  Of the 80 benchmark problems evaluated, and comparing
-with both optimizations disabled, 47 had the number of proof clauses
-reduced by a factor of 2 by using lemmas, 14 by using literal merging,
-and 60 by enabling both optimizations.  In the extreme cases, a lack
-of lemmas caused one proof to grow by a factor of 52.5, while a lack
-of literal grouping causes another proof to grow by a factor of 39.6.
-
-The time performance of these optimizations is less dramatic, but
-still significant.  Lemmas sometimes made proof generation slower, but
-in one case ran $38\times$ faster.  Similarly, literal grouping could
-take longer, but in one case ran $3.9\times$ faster.
-
-Overall, it is clear that these two optimizations are worthwhile, and
-sometimes critical, to success for some benchmarks.
-
-\subsection{Evaluation of the Verified Checker}
-\label{app:lean:evaluation}
-
-\input{figure-check}
-
-We ran the \lean{} checker on the 80 benchmark problems for which the
-generated CPOG file had at most 10~million total clauses.  These are the
-same benchmarks used in evaluating the optimizations (Section
-\ref{app:experiment:optimize}).
-All of the checks completed and confirmed the outcome of the prototype checker.
-
-Figure~\ref{fig:lean} shows the
-performance of both the \lean{} checker and the prototype
-checker on the Y axis, as functions of the total number of clauses on the X axis.
-As would be expected, the \lean{} checker consistently runs slower
-than the prototype checker.
-The ratio of the runtime for \lean{} versus the runtime for
-the prototype checker had a harmonic mean of $5.94$.
-
-Importantly, however, it can be seen that both checkers show the same
-overall performance trends.  The ratio of the two runtimes was less
-than $8.0$ for all but three small benchmarks.  This seems like a
-reasonable price to pay for a rigorous guarantee of correctness, and
-we are confident that we can reduce this gap with more effort in
-optimizing the verified checker.
-
-\subsection{Comparison with the MICE Framework}
-
-To compare the performance of our toolchain versus that of
-MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
-generation, proof checking, and ring evaluation in the CPOG toolchain compares to the time to
-run their proof generator \progname{nnf2trace} and checker
-\progname{sharptrace}.  We add the time for ring evaluation on our
-side, since their checker provides a count, although only an
-unweighted one.  We used the prototype versions of the checker and ring evaluator.
-We ran their tools on the 92 benchmark problems for
-which our toolchain requires at most 3000 seconds and retained
-the 84 data points for which at least one of the toolchains completes in under 1000 seconds.
-
-We can summarize the results as:
-\begin{itemize}
-\item 84 total data points
-\item 76 were completed by both programs in under 1000 seconds
-\item Of those 21 were faster with MICE, 55 with CPOG
-\item 7 completed with CPOG under 1000 seconds but exceeded that threshold for MICE
-\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CPOG
-\end{itemize}
-
-\input{figure-mice-cpog} The detailed results are shown in
-Figure~\ref{fig:mice}.  As can be seen, the comparative runtimes are
-highly variable, reflecting the fact that the two toolchains differ
-fundamentally in their objectives and their approaches.
-
-One shortcoming of \progname{nnf2trace} can be seen in
-the near-vertical series of points in the upper-lefthand corner.
-These correlate with a similar set of points along the left in
-Figure~\ref{fig:optimized:lessoptimized}.  Like the naive
-implementation of $\validate$, their program recursively traverses the
-graph representation of a formula, effectively expanding it into a
-tree.  They lack an optimization analogous to lemmas.
-
-We have not tried the MICE toolchain on larger benchmarks, but we
-suspect it would encounter significant scaling limitations.
-
 \end{document}

### 2023-04-12 doc/mc-paper.tex
Found typo
Simplified notation

@@
 \item For a single nonterminal child ($k = 1$), we recursively call
-  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$.
+  $\validate \left(\nodeu_1, \passign \cup L, \psi\right)$.
 \item For multiple nonterminal children ($k > 1$),
@@
   We recursively call
-  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \psi_i\right)$.
+  $\validate \left(\nodeu_i, \passign \cup L, \psi_i\right)$.
 \end{enumerate}
@@
 to 770,382,773, with a median of 1,719,245.  The ratios of the total
-clauses versus definining clauses ranged from 1.54 to 6073, with a
+clauses versus defining clauses ranged from 1.54 to 6073, with a
 harmonic mean of 3.13.  The high numbers were for problems having very

### 2023-04-12 doc/mc-paper.tex
More consistent notation

@@
 \item If $\nodeu$ is a product node, then we can divide its children
-  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
+  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
   \begin{enumerate}
     \item For each literal
-  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
+  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
   $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
@@
 \item For a single nonterminal child ($k = 1$), we recursively call
-  $\validate \left(\nodeu_1, \passign \cup L, \psi\right)$.
+  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
 \item For multiple nonterminal children ($k > 1$),
@@
   We recursively call
-  $\validate \left(\nodeu_i, \passign \cup L, \psi_i\right)$.
+  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
 \end{enumerate}
@@
 having many---tens or even hundreds---of literals as children.  The
-earlier formulation of $\validate$ considers each literal $\lit \in L$
+earlier formulation of $\validate$ considers each literal $\lit \in \lambda$
 separately, calling a SAT solver for every literal that cannot be
@@
 having many literals as children.  The naive formulation of $\validate$
-considers each literal $\lit \in L$ separately.
+considers each literal $\lit \in \lambda$ separately.
 Literal grouping allows all literals to be validated with a single call to a SAT solver.

### 2023-04-28 doc/sat2023-rebuttal.txt
First version

We wish to thank the reviewers for their helpful comments.  We will first respond to the most important items overall and then to ones by specific reviewers

DETERMINISTIC vs. DECISION
We will be more careful about the distinctions between these two flavors of DNNFs.  Our proof framework supports DETERMINISTIC, although our current proof generator relies on the inputs being in DECISION form.  We will clarify this.

PRIOR WORK
It was a major oversight on our part not to have cited the work by Capelli et al. AAAI-21.  This is, of course, very relevant.  Looking in detail at the conditions they check, it can be seen that we do much of the same work in our proof generator.  The main difference is that we then emit information as lines in a well-defined, formally-verified clausal proof framework.

We have been unable to locate a copy of their checker online.  If we can find one, we’ll do a comparison of runtimes and proof sizes.  D4 can emit a DRAT version of its conflict clauses.  At the very least, we will measure the sizes and checking times for these.

GENERALITY
Although we present results only for results from D4, we have also successfully used it on the results from DSHARP.  We have found that DSHARP lags far behind D4 in terms of both time and the size of the generated result.  Our proof framework is sufficiently general to handle SDDs.  We will list that as future work.

Your reviews sparked us to consider whether we could deal with formulas that have been preprocessed using equivalence-preserving transformations.  We were able to modify the source code of D4 to get it to perform preprocessing (it was disabled) and are collecting data now.  The structured approach does not work, but the monolithic approach works fine.  We will present data on this.  We do not believe the CD4 program is able to generate end-to-end proofs of this form.

REVIEWER#1

Your suggestions on how to make the presentation more clear are helpful.  The flow is not ideal, in part because we had to move so much information to the appendix to stay within the page limit

REVIEWER#2

Thank you for notifying us about prior work on d-Ds.  These are exactly what we mean by POGs, and we will include citations.

Regarding a proof of correctness.  We will work to make it easier to read the portions of the Lean presentation.  We feel that the value of a formal proof in Lean far exceeds what we could express with an informal proof.

Regarding the question of whether the SAT solver could introduce extension variables.  We do not allow this, for exactly the reason you mention.  The only extension variables are those encoding the AND and OR operations.

Your observation about the reverse implication proofs being tractable is noted.  We see from the Capelli paper that Darwiche had already observed this, and we will cite that.  For a practitioner, we feel that a one-sided proof would be better than nothing.

### 2023-04-30 doc/mc-data.tex
Better visualization of clause data
Quantify effect of preprocessing

@@
 
-\input{figure-clauses-d4-verify}
+\input{figure-clauses-d4-result}
 
@@
 disabled individually and in combination.  Figures
-\ref{fig:optimized:partially-optimized}--\ref{fig:optimized:unoptimized} illustrates the effect of the two
+\ref{fig:optimized:partially-optimized}--\ref{fig:optimized:lessoptimized} illustrates the effect of the two
 optimizations.  Each set of points shows the number of clauses when
@@
 
-Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 82.  All completed in less than 1000 seconds with structured.
+Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 83.  All completed in less than 1000 seconds with structured.
 Compare both runtimes and total number of clauses for structured vs. monolithic:
@@
 \item Structured complete 83.
-\item Monolithic completed  70,timed out on 13.
-\item Of the 70 completed, 23 were faster with monolithic, 37 were slower (min ratio = 0.015, max ratio = 9.66, hmean = 0.40)
-\item Of the 70 completed, 60 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
+\item Monolithic completed  69,timed out on 14.
+\item Of the 69 completed, 19 were faster with monolithic, 50 were slower (min ratio = 0.013, max ratio = 18.97, hmean = 0.41)
+\item Of the 69 completed, 59 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
+\end{itemize}
+
+\section{Preprocessing}
+
+\input{figure-runtime-preprocess}
+\input{figure-clauses-preprocess}
+
+Use preprocessor in D4.  Enable its three preprocessing methods
+(backbone, vivification, occElimination), all of which preserve
+equivalence.  Cannot use structured mode for forward implication
+proof, since modified clauses do not decompose in the way anticipated.
+
+Run same 83 benchmarks used to compare monolithic vs. structured mode.
+Incorporate time to run D4 into evaluation, since differs according to
+how D4 runs.
+
+Runtime results:
+\begin{itemize}
+\item Monolithic without preprocessing completes 68  of then with an overall time less than 1000 seconds.  (One less than when only consider generation and checking).
+\item Monolithic with prepreprocessing completes these plus one more (not the same as previously excluded)
+\item Of the 68, 53 run faster with preprocessing and 15 run slower (mininum ratio 0.04, maximum ratio 1.65, harmonic mean 0.38)
+\item Shifts only slightly when don't count D4 time.  Both complete 69 benchmarks. 56 faster, 13 slower (minimum ratio 0.04, maximum ratio 1.49, harmonic mean 0.37)
+\end{itemize}
+
+Clause results:
+\begin{itemize}
+\item Consider the 68 cases that completed for both with overall time less than 1000 seconds.
+\item 27 had more clauses without preprocessing, 19 were identical, and 22 had less.
+\item Min = 0.43, Max = 4.57, Harmonic mean = 0.98
 \end{itemize}

### 2023-04-30 doc/sat2023-rebuttal.txt
Updated

@@
-We wish to thank the reviewers for their helpful comments.  We will first respond to the most important items overall and then to ones by specific reviewers
+We wish to thank the reviewers for their helpful comments.  We will first respond to the most important items overall and then to ones by specific reviewers.
 
 DETERMINISTIC vs. DECISION
-We will be more careful about the distinctions between these two flavors of DNNFs.
+We will be more careful about the distinction between these two flavors of DNNFs.
 Our proof framework supports DETERMINISTIC, although our current proof generator relies on the inputs being in DECISION form.
@@
 PRIOR WORK
-It was a major oversight on our part not to have cited the work by Capelli et al.
-AAAI-21.
+It was a major oversight on our part not to have cited the work by Capelli, Lagniez, and Marquis in AAAI-21.
 This is, of course, very relevant.
@@
 The main difference is that we then emit information as lines in a well-defined, formally-verified clausal proof framework.
+As an example, they check that every path to a shared node in the decision-DNNF graph yields an equivalent simplified set of input clauses.
+Our proof generator checks this too, but then uses this information to apply a lemma, encoded as a sequence of clausal assertions.
+Our checker need not have any understanding of simplified input clauses or lemmas---it simply performs a sequence of RUP checks.
 
 We have been unable to locate a copy of their checker online.
-If we can find one, we’ll do a comparison of runtimes and proof sizes.
+If we can find one, we'll do a comparison of runtimes and proof sizes.
 D4 can emit a DRAT version of its conflict clauses.
@@
 
-Your reviews sparked us to consider whether we could deal with formulas that have been preprocessed using equivalence-preserving transformations.
-We were able to modify the source code of D4 to get it to perform preprocessing (it was disabled) and are collecting data now.
-The structured approach does not work, but the monolithic approach works fine.
+Reviewer #2’s comment that “Basically the knowledge compiler could be a black-box” sparked us to consider whether we could deal with formulas that have been preprocessed using equivalence-preserving transformations.
+We were able to modify the source code of D4 to enable its optional preprocessing.
+(This had been disabled in the version on Github.) The structured approach does not work, but the monolithic approach works fine.
+We have run the same benchmarks as we did for testing monolithic generation.
+On average, performing the three supported preprocessing steps improves the overall runtime (D4+generator+checker), with a harmonic mean speedup of 2.7X.
 We will present data on this.
@@
 
-Your suggestions on how to make the presentation more clear are helpful.  The flow is not ideal, in part because we had to move so much information to the appendix to stay within the page limit
+Your suggestions on how to make the presentation more clear are helpful.  The flow is not ideal, in part because we had to move so much information to the appendix to stay within the page limit.
 

### 2023-05-08 doc/mc-data.tex
Update statistics

--- 2023-03-22 doc/mc-data.tex
@@
      \end{itemize}
-   \item Ran program {\sc d2p} to generate CPOG file for maximum of 10,000 seconds
+   \item Ran program {\sc d2p} to generate CPOG file for maximum of
+     10,000 seconds.  Verifies file, generates unweighted count. If
+     finds literal weights, also generates weighted count
      \begin{itemize}
@@
      \item Completed successfully for all 108 instances
-     \item Time generally faster than {\sc d2p}. (min = 0.005, max = 1.4, harmonic mean = 0.10)
+     \item Time generally faster than {\sc d2p}. (min = 0.006, max = 1.38, harmonic mean = 0.107)
      \end{itemize}
@@
 
-\section*{Time to Generate and Validate CPOG}
+\section*{Computing Counts and Weighted Counts}
 
 \begin{itemize}
-\item Min 0.01, Average 1318.3, Max 13,144.8, Median 71.64
-\item Ratio to time for D4: Min 0.50, Harmonic Mean 5.50, Max 457.69
+\item Implemented C package to represent numbers of form $a \cdot 2^b \cdot 5^c$.
+  \begin{itemize}
+  \item For unweighted counts, first compute density and then scale by $2^n$
+  \item For unweighted counts, perform exact decimal arithmetic
+  \item The set of numbers of this form is not closed under a reciprocal operation, but the weights in the competion are either of the form
+    $w(x) + w(\overline{x}) = 1.0$ or $w(x) = w(\overline{x}) = 1$.  The latter case sums to 2, with reciprocal $1/2$.
+  \end{itemize}
+\item Can implement having $a$ be integer of arbitrary size and $b$ and $c$ being 32-bit integers.
+  \begin{itemize}
+    \item Represent $a$ by ``digits,'' each ranging in value from $0$ to $10^9-1$.
+    \item Fits in 32-bit number
+    \item Easy scaling by 10
+    \item Easy conversion to/from decimal string
+  \end{itemize}
+\item Weighted counts ranged up to 260,909 decimal digits.
+\end{itemize}
+
+\section*{Time to Generate, Validate, and Count CPOG}
+
+\begin{itemize}
+\item Min 0.012, Average 1,329.8, Max 13,198.3, Median 75.4
+\item Ratio to time for D4: Min 0.51, Harmonic Mean 5.87, Max 465.8
 \end{itemize}
@@
 
-\input{figure-clauses-d4-verify}
-
-\section*{Computing Counts and Weighted Counts}
-
-\begin{itemize}
-\item Implemented package to represent numbers of form $a \cdot 2^b \cdot 5^c$.
-  \begin{itemize}
-  \item For unweighted counts, first compute density and then scale by $2^n$
-  \item For unweighted counts, perform exact decimal arithmetic
-  \item The set of numbers of this form is not closed under a reciprocal operation, but the weights in the competion are either of the form
-    $w(x) + w(\overline{x}) = 1.0$ or $w(x) = w(\overline{x}) = 1$.  The latter case sums to 2, with reciprocal $1/2$.
-  \end{itemize}
-\item Can implement having $a$ be integer of arbitrary size and $b$ and $c$ being 32-bit integers.
-\item Weighted counts ranged up to 260,909 digits.
-\end{itemize}
+\input{figure-clauses-d4-result}
 
@@
 \begin{description}
-\item[Without lemmas:] Min 0.85, Max 37.98, Harmonic mean 1.80
-\item[Without literal grouping:] Min 0.83, Max 3.92, Harmonic mean 1.12
-\item[Without either optimization:] Min 0.91, Max 41.97, Harmonic mean 2.19
+\item[Without lemmas:] Min 0.89, Max 39.2, Harmonic mean 1.71
+\item[Without literal grouping:] Min 0.80, Max 3.77, Harmonic mean 1.11
+\item[Without either optimization:] Min 0.96, Max 43.69, Harmonic mean 2.20
 \end{description}
@@
   optimizations disabled, 40 had the time reduced by at least a factor of 2
-  using lemmas, 7 by using literal merging, and 49 by enabling both
+  using lemmas, 6 by using literal merging, and 48 by enabling both
   optimizations.
@@
 \item Monolithic completed  70,timed out on 13.
-\item Of the 70 completed, 23 were faster with monolithic, 37 were slower (min ratio = 0.015, max ratio = 9.66, hmean = 0.40)
+\item Of the 70 completed, 23 were faster with monolithic, 47 were slower (min ratio = 0.013, max ratio = 9.60, hmean = 0.36)
 \item Of the 70 completed, 60 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
+\end{itemize}
+
+\section{Preprocessing}
+
+\input{figure-runtime-preprocess}
+\input{figure-clauses-preprocess}
+
+Use preprocessor in D4.  Enable its three preprocessing methods
+(backbone, vivification, occElimination), all of which preserve
+equivalence.  Cannot use structured mode for forward implication
+proof, since modified clauses do not decompose in the way anticipated.
+
+Consider same 83 benchmarks used to compare monolithic vs. structured mode.
+Incorporate time to run D4 into evaluation, since differs according to
+how D4 runs.
+
+Runtime results:
+\begin{itemize}
+\item Monolithic without preprocessing completes 70  of them with an overall time less than 1000 seconds.
+\item Monolithic with prepreprocessing completes 69 (the one due to too much time on preprocessing)
+\item Of the 69, 53 run faster with preprocessing and 15 run slower (mininum ratio 0.04, maximum ratio 1.65, harmonic mean 0.38)
+\end{itemize}
+
+Clause results:
+\begin{itemize}
+\item Consider the 69 cases that completed for both with overall time less than 1000 seconds.
+\item 31 had fewer clauses with preprocessing, 17 were identical, and 21 had more.
+\item Min = 0.21, Max = 2.34, Harmonic mean = 0.85
 \end{itemize}
@@
 runtimes of our toolchain versus theirs.  The X-axis shows the sum of
-the elapsed times for our generator, checker, and counter (since the MICE checker also provides a count), while the
+the elapsed times for our generator and checker/counter, while the
 Y-axis shows the sum of the elapsed times for \textsc{nnf2trace},
 generating a file in their trace format from a d-DNNF graph generated
-by D4, and for \textsc{sharptrace}, a checker for the trace format.
+by D4, and for \textsc{sharptrace}, a checker and unweighted counter for the trace format
 The plot shows data points for those benchmarks for at least one of
 the two tools completed in less than 1000 seconds.
-
-We encountered a significant number of errors with both
-\textsc{nnf2trace} and \textsc{sharptrace}.  Based on the error
-messages, these appear to be mostly due to features of some of the
-CNF files or d-DNNF graphs they had not anticipated.  We provided a bug report to
-the authors, but they were unable to fix them on short notice.
 

### 2023-05-17 tools/toolchain_cd4.py
Run CD4 certification

@@ -1,0 +1,190 @@
+#!/usr/bin/python3
+
+# Run cd4 certifying ddnnf compiler
+
+import getopt
+import sys
+import os.path
+import subprocess
+import datetime
+import time
+
+def usage(name):
+    print("Usage: %s [-h] [-f] [-t TIME] FILE.EXT ..." % name)
+    print("  -h       Print this message")
+    print("  -f       Force regeneration of all files")
+    print("  -t TIME  Limit time for all programs")
+    print("  EXT      Can be any extension for wild-card matching (e.g., cnf, nnf)")
+
+# Defaults
+standardTimeLimit = 1000
+
+# Pathnames
+homePath = "/Users/bryant/repos"
+d4Path = homePath + "/d4"
+d4Program = d4Path + "/d4"
+checkProgram = d4Path + "/certificator"
+dratPath = homePath + "/drat-trim"
+dtrimProgram = dratPath + "/drat-trim"
+
+timeLimits = { "D4" : 4000, "DTRIM" : 4000, "CHECK" : 4000 }
+
+commentChar = 'c'
+
+def setTimeLimit(t):
+    global timeLimits, standardTimeLimit
+    standardTimeLimt = min(standardTimeLimit, t)
+    for k in timeLimits.keys():
+        timeLimits[k] = min(timeLimits[k], t)
+
+def checkFile(prefix, fname, logFile):
+    f = open(fname, 'r')
+    bytes = 0
+    lines = 0
+    for line in f:
+        if len(line) > 0 and line[0] == commentChar:
+            continue
+        bytes += len(line)
+        lines += 1
+    print("%s LOG: size %s %d lines %d bytes" % (prefix, fname, lines, bytes))
+    logFile.write("%s LOG: size %s %d lines %d bytes\n" % (prefix, fname, lines, bytes))
+    f.close()
+
+def runProgram(prefix, root, commandList, logFile, returnCodes = None):
+    if returnCodes is None:
+        returnCodes = [0]
+    if prefix in timeLimits:
+        timeLimit = timeLimits[prefix]
+    else:
+        timeLimit = standardTimeLimit
+    result = ""
+    cstring = " ".join(commandList)
+    print("%s. %s: Running '%s' with time limit of %d seconds" % (root, prefix, cstring, timeLimit))
+    logFile.write("%s LOG: Running %s\n" % (prefix, cstring))
+    logFile.write("%s LOG: Time limit %d seconds\n" % (prefix, timeLimit))
+    start = datetime.datetime.now()
+    try:
[…]

### 2023-05-17 tools/toolchain.py
Print NNF file size

@@ -150,6 +150,8 @@ def runD4(root, home, logFile, force):
     if preprocess:
         cmd += ["-preproc=backbone+vivification+occElimination"]
     ok = runProgram("D4", root, cmd, logFile)
+    if ok:
+        checkFile(root + ". D4 NNF", nnfName, logFile)
     if not ok and os.path.exists(nnfName):
         os.remove(nnfName)
     return ok
@@ -183,7 +185,8 @@ def runGen(root, home, logFile, force):
         cmd += ['-s']
     cmd += ["-C", str(clauseLimit), "-L", extraLogName, cnfName, nnfName, cpogName]
     ok = runProgram("GEN", root, cmd, logFile, extraLogName = extraLogName)
-    checkFile("GEN", cpogName, logFile)
+    if ok:
+        checkFile(root + ". GEN", cpogName, logFile)
     if not ok and os.path.exists(cpogName):
         os.remove(cpogName)
     return ok

### 2023-05-17 doc/mc-paper.tex
Partial revisions

@@
 
-We present \emph{Partitioned-Operation Graphs} (POGs), a generalization
+We present \emph{Partitioned-Operation Graphs} (POGs), a general form that can
+encode all
 of the representations used by existing knowledge compilers.
@@
 We have developed a program that generates POG representations from
-the d-DNNF graphs produced by the state-of-the-art knowledge compiler
+the decision-DNNF graphs (a restricted form of d-DNNF)
+produced by the state-of-the-art knowledge compiler
 \dfour{}, as well as checkable CPOG proofs certifying that the output POGs
 are equivalent to the input CNF formulas.  Our toolchain
-for generating and verifying POGs scales to all but the largest d-DNNF
+for generating and verifying POGs scales to all but the largest
 graphs produced by \dfour{} for formulas from a recent model counting
@@
 transforms the formula into a structured form for which model counting
-is straightforward.  For example, the {\em deterministic,
+is straightforward.  For example,the {\em deterministic
 decomposable, negation normal form} (d-DNNF) introduced by
-Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
+Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} (as well as the more restrictive {\em decision-DNNF} representation~\cite{oztok:cp:2014})
+ represents a
 Boolean formula as a directed acyclic graph, with terminal nodes
@@
 This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
-a generalization of the representations produced by current knowledge
+a general form that can encode all of the representations produced by current knowledge
 compilers. The CPOG (for ``certified'' POG) file format then
@@
 input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
-generates a d-DNNF representation, and the {\em proof
+generates a decision-DNNF representation,
+and the {\em proof
   generator} uses this to generate a CPOG file.
@@
 formulas from the 2022 standard and weighted model competitions.  Our
-tools handle all but the largest d-DNNF graphs generated
+tools handle all but the largest decision-DNNF graphs generated
 by \dfour{}.  We also measure the benefits of several optimizations as
@@
 The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
-By contrast, our proofs construct a new propositional formula, and we must verify
+Our work builds on the well-established technology and tools associated with clausal proof systems,
+but we require features not found in proofs of unsatisfiability.
+Our proofs construct a new propositional formula, and we must verify
 %both
@@
 %satisfies a set of rules and that it
-is equivalent to the input formula.
-This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
-Sections~\ref{sect:cpog} and \ref{section:formally-verified-toolchain}.
+is equivalent to the input formula.  This requires verifying
+additional proof steps, including clause deletion steps, and subtle
+invariants, as described in Sections~\ref{sect:cpog} and
+\ref{section:formally-verified-toolchain}.
+
+Capelli et al.\ developed a version of the \dfour{} knowledge compiler
+that generates a certificate in a proof system that is itself based on
+decision-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their
+modified version of \dfour{} generates additional annotations to the
+compiled representation to enable checking for equivalence with the
+input formula.  They must also track cases where the knowledge
+compiler detects that it can reuse a previously generated subgraph,
+yielding a DAG rather than a tree representation.  Their proof checker
+then must perform a number of checks that we incorporate into our
+proof generator.  Our framework has the advantage of being independent
+from the means by which the proofs are generated with proof checking
+based on very elementary clausal reasoning steps.
 
@@
 proof framework for model counting programs.  Their proof rules are
-based on the algorithms commonly used by model counters.
-They present a series of lemmas and a theorem stating
-that their checking framework is sound.  They also present
-experimental results for cases where they modify existing model
-counters to generate proof traces, and where they generate proof
-traces from d-DNNF graphs.
+based on the algorithms commonly used by model counters.  They present
+a series of lemmas and a theorem stating that their checking framework
+is sound.  They also present experimental results for cases where they
+modify existing model counters to generate proof traces, and where
+they generate proof traces from decision-DNNF graphs.
 %% They have not formally verified their
@@
 
-We know of no other
-work on certifying the output of a knowledge compiler.
+Both the
 
@@
 
-POGs generalize the d-DNNF graphs in two ways:
+POGs generalize decision-DNNF graphs in two ways:
 \begin{itemize}
@@
 %%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
-\item They allow arbitrary arguments to a $\por$ operation, as long as
+\item As with d-DNNF~\cite{darwiche:aaai:2002,darwiche:jair:2002}, they allow arbitrary arguments to a $\por$ operation, as long as
   these have disjoint models.  By contrast, each
-  sum node $\nodeu$ in a d-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
+  sum node $\nodeu$ in a decision-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
   any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
@@
 \end{itemize}
-  These generalizations allow more flexibility in the POG
-  representation while maintaining the ability to efficiently perform ring evaluation.
+The generalizations encompassed by POGs have also been referred to as {\em deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
+Our current proof generator only works for knowledge compilers
+generating decision-DNNF representations, but these generalizations
+allow for future extensions, while maintaining the ability to
+efficiently perform ring evaluation.
 
@@
 %% \item The {\em proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
-%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
+%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a decision-DNNF representation of the formula.
 %% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
@@
   The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
-  Declaring a node $\nodeu$
-implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
+  Declaring a node $\nodeu$ implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
   encoding the product or sum operation.
@@
 clause deletions represents a systematic transformation of the input formula
-into a POG formula.  Validating all of these steps serves to prove that the
-POG is logically equivalent to the input formula.
+into a POG formula.  Validating all of these steps serves to prove that
+POG $P$ is logically equivalent to the input formula.
 At the completion of the proof, the following \textsc{final conditions} must hold:
@@
 in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
-of unit clause $\{r\}$ and the defining clauses of the POG nodes. Recognizing that
+of unit clause $\{r\}$ and the defining clauses for the nodes, providing a Tseitin encoding of $P$. Recognizing that
 any total assignment to the input variables implicitly defines the assignments to the extension variables,
@@
 
-\section{Generating CPOG from d-DNNF}
+\section{Generating CPOG from decision-DNNF}
 \label{section:generating:cpog}
@@
 %%but it would also work for other tools that use a top-down search
-%%strategy to generate a d-DNNF representation.
-A d-DNNF
+%%strategy to generate a decision-DNNF representation.
+A decision-DNNF
 graph can be directly translated  into a POG,
@@
 \item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
-  then, since the node originated from a d-DNNF graph, there must be
+  then, since the node originated from a decision-DNNF graph, there must be
   some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
@@
   into a sequence of clause additions in the CPOG file.
+  (The solver is constrained to only use RUP inference rules, preventing it from introducing extension variables.)
 \item For a single nonterminal child ($k = 1$), we recursively call
@@
 
-Each input clause deletion can be done as a single RUP step.  The
+Each input clause deletion can be done as a single RUP step, based
+on an algorithm to test for clausal entailment in d-DNNF graphs~\cite{capelli:sat:2019,darwiche:jair:2002}.  The
 proof generator constructs the hint sequence from the defining
@@
 It is worth noting that, to date, we have not found
-any errors in the d-DNNF graphs generated by \dfour{}.
+any errors in the decision-DNNF graphs generated by \dfour{}.
 
@@
 
+This paper demonstrates a method for certifying the equivalence of two
+different representations of a Boolean formula: an input formula
+represented in conjunctive normal form, and a compiled representation
+that can then be used to extract useful information about the formula,
+including its weighted and unweighted model counts.  It builds on the
+extensive techniques that have been developed for clausal
+proof systems, including extended resolution and reverse unit propagation, as well as established tools, such as
+proof-generating SAT solvers and \dtrim{}.
+
 We are hopeful that having checkable proofs for knowledge compilers
@@
 %%the outputs of one particular program, it could be adapted
-Our experiments demonstrate that our toolchain can already
- handle problems nearly at the limits of current knowledge compilers.
-Further engineering and optimization of our proof
-generator and checker could improve their performance and capacity substantially.
+Our experiments demonstrate that our toolchain can already handle
+problems nearly at the limits of current knowledge compilers.  Further
+engineering and optimization of our proof generator and checker could
+improve their performance and capacity substantially.  We are hopeful
+that our tool can be adapted to handle other knowledge compiler
+representations, such as sentential decision diagrams (SDDs)~\cite{darwiche:ijcai:2011}.
 
@@
 of a previous partial assignment~\cite{darwiche:aaai:2002}.
-When this d-DNNF node is translated into POG
+When this decision-DNNF node is translated into POG
 node $\nodeu$, the proof generator can assume (and also check), that
@@
 in this document are actual elapsed times, not CPU times.}, \dfour{}
-was able to generate d-DNNF representations for 124 of these, and it timed out on the
+was able to generate decision-DNNF representations for 124 of these, and it timed out on the
 rest.  Our proof generator was able to convert all of the generated
-d-DNNF graphs into POGs and determine how many defining clauses they
+decision-DNNF graphs into POGs and determine how many defining clauses they
 would generate.  A POG operation with $k$ arguments requires $k+1$
@@
 
-For each of the 124 d-DNNF graphs generated by \dfour{}, we ran
+For each of the 124 decision-DNNF graphs generated by \dfour{}, we ran
 our proof generator and prototype checker, limiting proof

### 2023-05-18 doc/mc-data.tex
More data
Update data
Include results from CD4

@@
        \begin{itemize}
-       \item Number of POG nodes around $0{.}90\times$ number of dDNNF nodes
+       \item Number of POG nodes around ranged between $0.35\times$
+         and $1.0\times$ number of dDNNF nodes (harmonic mean =
+         $0.81\times$).  Greater optimizations with smaller
+         benchmarks.
        \end{itemize}
@@
 
+\section{Benchmarking the \lean{} Proof Checker}
+
+Ran \lean{} checker on the benchmarks yielding at most $10^7$ clauses, with time limit of 1000 seconds.
+\begin{itemize}
+\item 80 total
+\item All ran to completion
+\item All but three ran within factor of $8\times$.  Outliers were small ones
+\item Harmonic mean = $5.43\times$.
+\end{itemize}
+
+\input{figure-check}
+
 \section{Comparison with MICE}
@@
 
-\section{Benchmarking the \lean{} Proof Checker}
+\section{Comparison with CD4 Toolchain}
 
-Ran \lean{} checker on the benchmarks yielding at most $10^7$ clauses, with time limit of 1000 seconds.
+\input{figure-runtime-cd4}
+
+Ran D4 with certificate logging enabled (by specifying DRAT file).
+Used 114 benchmarks for which regular D4 ran in under 1000 seconds.
+Measure time for entire toolchains (compilation, proof generation,
+checking).  Checking for CD4 includes running drat-trim.
 \begin{itemize}
-\item 80 total
-\item All ran to completion
-\item All but three ran within factor of $8\times$.  Outliers were small ones
-\item Harmonic mean = $5.94$.
+\item Completed 80 in under 1000 seconds for both
+  \begin{itemize}
+    \item 7 ran faster with POG framework, up to $2.8\times$; 73 ran slower, up to $119.8\times$.  Harmonic mean = $19.63\times$ slower.
   \end{itemize}
+\item 2 completed in under 1000 seconds  with POGs but timed out with CD4.  Both had very few models and with long unsatisfiability proofs
+\item 28 completed in under 1000 seconds with CD4 but either timed out or failed with POGs.
+\item Remaining 4 exceeded 1000 seconds for both toolchains
 
-\input{figure-check}
+\end{itemize}
 
@@
 
-\begin{description}
-\item[Lemmas:] The justification proof must show that the conjunction
-  of the input clauses implies the extension variable associated with
-  the root node.  It does so by showing that this property holds for
-  every node and for every path from the root that node, relative to
-  the partial assignment given by that path.  This could potentially
-  require enumerating exponential numbers of paths to some nodes,
-  effectively converting the graph into a tree.  Instead, the proof
-  can contain a {\em lemma} for each node having indegree greater than
-  one, showing that the root node is implied by a set of clauses.
-  Some of these are the original input clauses, while some are {\em
-    synthetic clauses} representing input clauses that have been
-  reduced by eliminating false literals.  Given such a proof, the
-  lemma can be instantiated for each parent node.  Proofs with lemmas
-  incur some overhead---the synthetic clauses are declared as POG
-  product nodes, and additional clauses must be added to show that,
-  under a given context, the synthetic clauses are implied by some
-  input or containing synthetic clause.  On the other hand, they
-  guarantee that no node in the graph is the root of more than one
-  justication proof.
-\item[Literal grouping:] The generated POG representation will
-  typically have a number of product nodes forming the conjunction of
-  tens, or even hundreds, of literals.  These literals represent ones
-  that we either implied by unit propagation or were added when
-  assigning the complement to the corresponding variable led to a
-  conflict.  Justifying this product node requires showing that each
-  of these literals must hold under the given context.  When this
-  cannot be done by unit propagation, the program must invoke a SAT
-  solver to prove that the complemented value yields a conflict.
-  Rather than doing so for each literal individually, the prover can
-  create a new product node consisting of the literals that have not
-  yet been validated and then prove that falsifying the extension
-  variable associated with the product node yields a conflict.  
 \end{description}

### 2023-05-18 doc/mc-paper.tex
More revisions

@@
 special form, such as the deterministic,
-decomposable, negation normal form (d-DNNF)\@.
+decomposable, negation normal form (det-DNNF)\@.
 \emph{Knowledge compilation} is the process of converting a formula
@@
 We have developed a program that generates POG representations from
-the decision-DNNF graphs (a restricted form of d-DNNF)
+graphs 
 produced by the state-of-the-art knowledge compiler
@@
 transforms the formula into a structured form for which model counting
-is straightforward.  For example,the {\em deterministic
-decomposable, negation normal form} (d-DNNF) introduced by
-Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} (as well as the more restrictive {\em decision-DNNF} representation~\cite{oztok:cp:2014})
+is straightforward.  For example, the {\em deterministic
+decomposable, negation normal form}
+(det-DNNF) introduced by
+Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}
 represents a
@@
 properties of Boolean functions that can be efficiently computed from
-a d-DNNF representation.  These include standard and weighted model
+a det-DNNF representation.  These include standard and weighted model
 counting, and much more.
@@
 input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
-generates a decision-DNNF representation,
+generates a {\em decision} DNNF (dec-DNNF) representation (a restricted form of dec-DNNF),
 and the {\em proof
@@
 formulas from the 2022 standard and weighted model competitions.  Our
-tools handle all but the largest decision-DNNF graphs generated
+tools handle all but the largest dec-DNNF graphs generated
 by \dfour{}.  We also measure the benefits of several optimizations as
@@
 that generates a certificate in a proof system that is itself based on
-decision-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their
+dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their
 modified version of \dfour{} generates additional annotations to the
-compiled representation to enable checking for equivalence with the
-input formula.  They must also track cases where the knowledge
-compiler detects that it can reuse a previously generated subgraph,
-yielding a DAG rather than a tree representation.  Their proof checker
-then must perform a number of checks that we incorporate into our
-proof generator.  Our framework has the advantage of being independent
-from the means by which the proofs are generated with proof checking
-based on very elementary clausal reasoning steps.
+compiled representation, as well as a file of clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
+These files must be processed by two different checkers.
 
@@
 proof framework for model counting programs.  Their proof rules are
-based on the algorithms commonly used by model counters.  They present
-a series of lemmas and a theorem stating that their checking framework
-is sound.  They also present experimental results for cases where they
-modify existing model counters to generate proof traces, and where
-they generate proof traces from decision-DNNF graphs.
-%% They have not formally verified their
-%%proof framework or created a formally verified checker.
-Our toolchain is not directly comparable to theirs---they directly
-certify the numeric value produced by a model counter, while we
-certify that the graph generated by a knowledge compiler is logically
-equivalent to the input formula.  We can then validate a standard or
-weighted count from that representation.
-%% Their tool
-%% chain can be applied to some programs that are pure model counters,
-%%while ours can be applied to pure knowledge compilers.
-Nonetheless,
-we compare the performance of our toolchain to theirs as part of our
-experimental evaluation in Appendix~\ref{app:experiments}.
-
-Both the
+based on the algorithms commonly used by model counters.  They
+developed a program that can generate proof traces from dec-DNNF
+graphs and a program to check adherence to their proof rules.  This
+framework is not directly comparable to ours, since it only certifies
+the unweighted model count, but it has similar goals.
+
+Both of these prior certification frameworks are strongly tied to the
+algorithms used by the knowledge compilers and model counters.  Some
+of the conditions to be checked are relevant only to specific
+implementations.  Although the authors make informal arguments
+regarding the soundness of their frameworks, these do not provide
+strong levels of assurance.  Indeed, as we discuss in
+Appendix~\ref{app:comparison}, we uncovered a weakness in the \dfour{}
+verification framework due to its invalid assumption about the
+guarantees provided by \dtrim{}, the program it uses to check their
+DRAT file.  Our framework is very general and is based on a small set
+of proof rules.  It builds on the highly developed
+concepts of clausal proof systems.  These factors were important in enabling formal verification.
+In Appendix~\ref{app:comparison}
+we also compare the performance of our toolchain to these two others.
 
@@
 
-POGs generalize decision-DNNF graphs in two ways:
+POGs generalize dec-DNNF graphs in two ways:
 \begin{itemize}
@@
 %%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
-\item As with d-DNNF~\cite{darwiche:aaai:2002,darwiche:jair:2002}, they allow arbitrary arguments to a $\por$ operation, as long as
+\item As with det-DNNF~\cite{darwiche:aaai:2002,darwiche:jair:2002}, they allow arbitrary arguments to a $\por$ operation, as long as
   these have disjoint models.  By contrast, each
-  sum node $\nodeu$ in a decision-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
+  sum node $\nodeu$ in a dec-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
   any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
@@
 Our current proof generator only works for knowledge compilers
-generating decision-DNNF representations, but these generalizations
+generating dec-DNNF representations, but these generalizations
 allow for future extensions, while maintaining the ability to
@@
 %% \item The {\em proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
-%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a decision-DNNF representation of the formula.
+%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a dec-DNNF representation of the formula.
 %% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
@@
 
-\section{Generating CPOG from decision-DNNF}
+\section{Generating CPOG from dec-DNNF}
 \label{section:generating:cpog}
@@
 %%but it would also work for other tools that use a top-down search
-%%strategy to generate a decision-DNNF representation.
-A decision-DNNF
+%%strategy to generate a dec-DNNF representation.
+A dec-DNNF
 graph can be directly translated  into a POG,
@@
 \item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
-  then, since the node originated from a decision-DNNF graph, there must be
+  then, since the node originated from a dec-DNNF graph, there must be
   some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
@@
 Each input clause deletion can be done as a single RUP step, based
-on an algorithm to test for clausal entailment in d-DNNF graphs~\cite{capelli:sat:2019,darwiche:jair:2002}.  The
+on an algorithm to test for clausal entailment in det-DNNF graphs~\cite{capelli:sat:2019,darwiche:jair:2002}.  The
 proof generator constructs the hint sequence from the defining
@@
 It is worth noting that, to date, we have not found
-any errors in the decision-DNNF graphs generated by \dfour{}.
+any errors in the dec-DNNF graphs generated by \dfour{}.
 
@@
 of a previous partial assignment~\cite{darwiche:aaai:2002}.
-When this decision-DNNF node is translated into POG
+When this dec-DNNF node is translated into POG
 node $\nodeu$, the proof generator can assume (and also check), that
@@
 in this document are actual elapsed times, not CPU times.}, \dfour{}
-was able to generate decision-DNNF representations for 124 of these, and it timed out on the
+was able to generate dec-DNNF representations for 124 of these, and it timed out on the
 rest.  Our proof generator was able to convert all of the generated
-decision-DNNF graphs into POGs and determine how many defining clauses they
+dec-DNNF graphs into POGs and determine how many defining clauses they
 would generate.  A POG operation with $k$ arguments requires $k+1$
@@
 
-For each of the 124 decision-DNNF graphs generated by \dfour{}, we ran
+For each of the 124 dec-DNNF graphs generated by \dfour{}, we ran
 our proof generator and prototype checker, limiting proof

### 2023-05-18 cpog/prototype/clean_cnf.py
Cleanup

@@ -7,10 +7,6 @@ import readwrite
 def usage(name):
     print("Usage: %s INFILE.cnf OUTFILE.cnf")
 
-
-def eprint(s):
-    sys.stderr.write(s + '\n')
-
 def process(iname, oname):
     icount = 0
     ncount = 0
@@ -18,12 +14,12 @@ def process(iname, oname):
     try:
         pass
     except Exception as ex:
-        eprint("Failed to read CNF file %s: %s" % (iname, str(ex)))
+        print("Failed to read CNF file %s: %s" % (iname, str(ex)))
         return
     try:
         cwriter = readwrite.CnfWriter(creader.nvar, oname)
     except Exception as ex:
-        eprint("Failed to open CNF output file %s: %s" % (oname, str(ex)))
+        print("Failed to open CNF output file %s: %s" % (oname, str(ex)))
         return
     for clause in creader.clauses:
         icount += 1
@@ -32,7 +28,7 @@ def process(iname, oname):
             ncount += 1
             cwriter.doClause(clause)
     cwriter.finish()
-    eprint("Processed %d clauses in file %s.  Wrote %d clauses to file %s" % (icount, iname, ncount, oname))
+    print("Processed %d clauses in file %s.  Wrote %d clauses to file %s" % (icount, iname, ncount, oname))
 
 def run(name, args):
     if len(args) != 2:

### 2023-05-18 cpog/prototype/readwrite.py

@@ -275,7 +275,6 @@ class CnfReader():
             self.file.close()
         
     def readCnf(self, check):
-        print("Reading CNF.  check=%s" % str(check))
         lineNumber = 0
         nclause = 0
         self.nvar = 0

### 2023-05-19 doc/mc-paper.tex
Updates

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{tikz}
@@
 
-\newcommand{\pand}{\mathbin{\land^{\sf p}}}
-\newcommand{\por}{\mathbin{\lor^{\sf p}}}
-\DeclareMathOperator*{\Pand}{\bigwedge^{\sf p}}
-\DeclareMathOperator*{\Por}{\bigvee^{\sf p}}
+\newcommand{\pand}{\mathbin{\land^\textsf{p}}}
+\newcommand{\por}{\mathbin{\lor^\textsf{p}}}
+\DeclareMathOperator*{\Pand}{\bigwedge^\textsf{p}}
+\DeclareMathOperator*{\Por}{\bigvee^\textsf{p}}
 \newcommand{\boolnot}{\neg}
@@
 \newcommand{\drational}{\textbf{Q}_{2,5}}
+\newcommand{\entail}{\vDash}
+\newcommand{\entaildrat}{\entail_{\textrm{DRAT}}}
 
@@
 \newcommand{\dfour}{\progname{D4}}
+\newcommand{\cdfour}{\progname{CD4}}
 \newcommand{\Dfour}{\progname{D4}}
@@
 \newcommand{\dtrim}{\progname{drat-trim}}
+\newcommand{\Dtrim}{\progname{Drat-trim}}
 
@@
 
-%\nolinenumbers %uncomment to disable line numbering
+\nolinenumbers %uncomment to disable line numbering
+
+\EventEditors{Meena Mahajan and Friedrich Slivovsky}
+\EventNoEds{2}
+\EventLongTitle{26th International Conference on Theory and Applications of Satisfiability Testing (SAT 2023)}
+\EventShortTitle{SAT 2023}
+\EventAcronym{SAT}
+\EventYear{2023}
+\EventDate{July 4--8, 2023}
+\EventLocation{Alghero, Italy}
+\EventLogo{}
+\SeriesVolume{271}
+\ArticleNo{6}
 
@@
 
-We have developed a program that generates POG representations from
+We have developed a program that generates POG representations from det-DNNF
 graphs 
@@
 security protocols.  Some applications, however, require going
-beyond Boolean satisfiability.  For example, the {\em model
+beyond Boolean satisfiability.  For example, the \emph{model
   counting problem} requires computing the number of satisfying
@@
 
-One approach to model counting, known as {\em knowledge compilation},
+One approach to model counting, known as \emph{knowledge compilation},
 transforms the formula into a structured form for which model counting
-is straightforward.  For example, the {\em deterministic
+is straightforward.  For example, the \emph{deterministic
 decomposable, negation normal form}
@@
 One shortcoming of existing knowledge compilers is that they have no
+generally accepted
 way to validate that their generated results are correct, i.e., that
@@
 based on an extended resolution proof system~\cite{Tseitin:1983}.
-We establish a set of conditions that, when satisified by a CPOG file, guarantee that it
+We establish a set of conditions that, when satisified by a CPOG file, guarantees that it
 encodes a well-formed POG and provides a valid equivalence proof.
@@
 input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
-generates a {\em decision} DNNF (dec-DNNF) representation (a restricted form of dec-DNNF),
-and the {\em proof
+generates a \emph{decision} DNNF (dec-DNNF) representation (a restricted form of det-DNNF),
+and the \emph{proof
   generator} uses this to generate a CPOG file.
-The {\em proof checker} verifies the equivalence of the CNF and CPOG representations.
-The {\em ring evaluator} computes
+The \emph{proof checker} verifies the equivalence of the CNF and CPOG representations.
+The \emph{ring evaluator} computes
 a standard or weighted model count from the POG\@ representation.
@@
 
-To ensure soundness of the abstract CPOG proof system, as well as correctness of its
-concrete implementation, we formally verified the proof system as well as versions
-of the proof checker and ring evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
-Running these two programs on a particular CPOG file gives assurance that the proof
-and the model count are correct. Our experience with developing a formally verified
-proof checker has shown that, even within the well-understood framework of extended
-resolution, it can be challenging to formulate a full set of requirements that guarantee soundness.
-In fact, subtle details in the partitioned sum rule were identified during this process.
-
-We evaluate our toolchain using benchmark
-formulas from the 2022 standard and weighted model competitions.  Our
-tools handle all but the largest dec-DNNF graphs generated
-by \dfour{}.  We also measure the benefits of several optimizations as
-well as the relative performance of the verified checker with one
-designed for high performance and capacity.
+To ensure soundness of the abstract CPOG proof system, as well as
+correctness of its concrete implementation, we formally verified the
+proof system as well as versions of the proof checker and ring
+evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
+Running these two programs on a particular CPOG file gives strong
+assurance that the proof and the model count are correct. Our
+experience with developing a formally verified proof checker has shown
+that, even within the well-understood framework of extended
+resolution, it can be challenging to formulate a full set of
+requirements that guarantee soundness.  In fact, our efforts to
+formally verify our proof framework exposed subtle conditions that we had
+to impose on our partitioned sum rule.
+
+We evaluate our toolchain using benchmark formulas from the 2022
+standard and weighted model competitions.  Our tools handle all but
+the largest dec-DNNF graphs generated by \dfour{}.  We also measure
+the benefits of several optimizations as well as the relative
+performance of the verified checker with one designed for high
+performance and capacity.  We also show that our tools can provide
+end-to-end verification of formulas that have been transformed by an
+equivalence-preserving preprocessor.
 
@@
 
-Capelli et al.\ developed a version of the \dfour{} knowledge compiler
+Capelli et al.\ developed a knowledge compiler
 that generates a certificate in a proof system that is itself based on
-dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their
-modified version of \dfour{} generates additional annotations to the
-compiled representation, as well as a file of clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
-These files must be processed by two different checkers.
+dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{} program,
+a modified version of \dfour{}, generates annotations to the
+compiled representation, providing information about how the compiled version relates to the input clauses.
+It also generates a file of clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
+Completing the certification involves running two different checkers on
+the annotated dec-DNNF graph and the DRAT file.
+Although the authors make informal arguments
+regarding the soundness of their frameworks, these do not provide
+strong levels of assurance.  Indeed, 
+we have identified a weakness in the their
+methodology due to an invalid assumption about the
+guarantees provided by \dtrim{}, the program it uses to check the
+DRAT file.
+
+In more detail, \cdfour{} emits a sequence of clauses $R$ that
+includes the conflict clauses that arose via a top-down processing of
+the input clauses.  Given input formula $\phi$, their first task is to
+check whether $\phi \imply R$, i.e., that any model of $\phi$ is also
+a model of $R$.  They then base other parts of their proof on that
+property and use a separate checker to perform a series of additional
+checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
+satisifies the \emph{reverse asymmetric tautology} (RAT) property with
+respect to the preceding
+clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
+clause $C$ to a set of clauses maintains satisfiability, but it may
+not preserve models.  As an example, consider the following formulas:
+\begin{center}
+  \begin{tabular}{lccc}
+    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
+    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_1 \lor \obar{x}_2)$\\    
+  \end{tabular}
+\end{center}
+Clearly, these two formulas are not equivalent---$\phi_1$ has six
+models, while $\phi_2$ has five.  In particular, $\phi_1$ allows
+arbitrary assignments to variable $x_2$.  Critically, however the
+second clause of $\phi_2$ is RAT with respect to the first---any
+satisfying assignment to $\phi_1$ can be transformed into one that
+also satisfies $\phi_2$ by setting $x_2 = 0$.
+
+This weakness would allow a buggy (or malicious) version
+of \cdfour{} to generate a compiled representation of $\phi_2$ when
+given $\phi_1$ as its input.  By adding the second clause to $R$,
+the check with \dtrim{} would pass, as would the other tests performed
+by their checker.  We have been able to confirm this possibility with
+their compiler and checker\footnote{Downloaded May 18, 2023 from
+\url{https://github.com/crillab/d4}}.
+
+This weakness can be corrected by restricting \dtrim{} to only add
+clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
+commandline argument to \dtrim{}\footnote{Available at
+\url{https://github.com/marijnheule/drat-trim/tree/develop}.} that enforces this
+restriction.  This weakness, however, illustrates the general challenge of
+developing a new proof framework.
+Without a formal verification, there are likely to be 
+conditions that make the framework unsound.
 
@@
 the unweighted model count, but it has similar goals.
+Again, they provide only  informal arguments 
+regarding the soundness of their frameworks.
 
@@
 of the conditions to be checked are relevant only to specific
-implementations.  Although the authors make informal arguments
-regarding the soundness of their frameworks, these do not provide
-strong levels of assurance.  Indeed, as we discuss in
-Appendix~\ref{app:comparison}, we uncovered a weakness in the \dfour{}
-verification framework due to its invalid assumption about the
-guarantees provided by \dtrim{}, the program it uses to check their
-DRAT file.  Our framework is very general and is based on a small set
+implementations.    Our framework is very general and is based on a small set
 of proof rules.  It builds on the highly developed
@@
   Let $\varset$ denote a set of Boolean variables, and let $\assign$
-  be an {\em assignment} of truth values to some subset of the
+  be an \emph{assignment} of truth values to some subset of the
   variables, where $0$ denotes false and $1$ denotes true, i.e.,
   $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
-  \subseteq \varset$.  We say the assignment is {\em total} when it
+  \subseteq \varset$.  We say the assignment is \emph{total} when it
   assigns a value to every variable ($\varset' = \varset$), and that
-  it is {\em partial} otherwise.
+  it is \emph{partial} otherwise.
   The set of all possible total assignments over
@@
 For each variable $x \in \varset$,
-  we define the {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
+  we define the \emph{literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
   negation of $x$. An
@@
   The set of Boolean formulas is defined recursively.  Each
-  formula $\phi$ has an associated {\em dependency set}
+  formula $\phi$ has an associated \emph{dependency set}
   $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
@@
     and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
-  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
+  \item For formula $\phi$, its \emph{negation}, written $\boolnot \phi$ is a Boolean formula,
     with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
       $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
       $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
+  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
       $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
@@
   We highlight two special classes of Boolean formulas.  A formula is
-  in {\em negation normal form} when negation is applied only to variables.  A
-  formula is in {\em conjunctive normal form} (CNF) when i) it is in
+  in \emph{negation normal form} when negation is applied only to variables.  A
+  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
   negation normal form, and ii) sum is applied only to literals.  A CNF
-  formula can be represented as a set of {\em clauses}, each of which is a
+  formula can be represented as a set of \emph{clauses}, each of which is a
   set of literals.  Each clause represents the sum of the
@@
   set notation to reference the clauses within a formula and the
-  literals within a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause and the literal as a {\em unit} literal.
+  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
 This literal must be assigned value $1$ by any satisfying assignment of the formula.
@@
 \begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
-  A {\em partitioned-operation formula}
+  A \emph{partitioned-operation formula}
  satisfies the following for all product and sum operations:
@@
 \begin{definition}[Commutative Ring]
-  A {\em commutative ring} $\ring$ is an algebraic structure
+  A \emph{commutative ring} $\ring$ is an algebraic structure
   $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
@@
   the multiplicative identity.  Every element $a \in \dset$ has an
-  {\em additive inverse} $-a$ such that $a + -a = \addident$.
+  \emph{additive inverse} $-a$ such that $a + -a = \addident$.
 \label{def:ring}
@@
 \label{def:ring_evaluation}
-  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
+  For commutative ring $\ring$, a \emph{ring weight function} associates a value $w(x) \in \dset$ with
   every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.
 
-  For Boolean formula $\phi$ and ring weight function $w$, the {\em ring evaluation problem} computes
+  For Boolean formula $\phi$ and ring weight function $w$, the \emph{ring evaluation problem} computes
   \begin{eqnarray}
@@
 expressed as ring evaluation problems.  The
-(standard) {\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
+(standard) \emph{model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
 It can be cast as a ring evaluation problem by having $\radd$ and
@@
 weight function $w(x) = 1/2$ for every variable $x$.
-Ring evaluation of formula $\phi$ gives the {\em density} of
+Ring evaluation of formula $\phi$ gives the \emph{density} of
 the formula, i.e., the fraction of all possible total assignments that are
@@
 
-The {\em weighted model counting}  problem is also defined over
+The \emph{weighted model counting}  problem is also defined over
 rational numbers.  Some formulations  allow
@@
 
-The {\em function hashing problem} provides a test
+The \emph{function hashing problem} provides a test
 of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
-finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
+finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, choose a value from $\dset$ at random for $w(x)$.  Two formulas
 $\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
@@
 
-A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
-with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
-When $(\nodeu,\nodev) \in E$,
-node $\nodev$ is said to be a {\em child} of node $\nodeu$.
-The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
-$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
-Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
-Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
-A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
-Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
-It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
-Every POG has a designated {\em root node} $\noder$.
+A \emph{partitioned-operation graph} (POG) is a directed, acyclic
+graph with nodes $N$ and edges $E \subseteq N \times N$.  We denote
+nodes with boldface symbols, such as $\nodeu$ and $\nodev$.  When
+$(\nodeu,\nodev) \in E$, node $\nodev$ is said to be a \emph{child} of
+node $\nodeu$.  The in- and out-degrees of node $\nodeu$ are defined
+as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
+$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.  Node
+$\nodeu$ is said to be \emph{terminal} if $\outdegree(\nodeu) = 0$.  A
+terminal node is labeled by a Boolean constant or variable.  Node
+$\nodeu$ is said to be \emph{nonterminal} if $\outdegree(\nodeu) > 0$.
+A nonterminal node is labeled by Boolean operation $\pand$ or $\por$.
+A node can be labeled with operation $\pand$ or $\por$ only if it
+satisfies the partitioning restriction for that operation.  Every POG
+has a designated \emph{root node} $\noder$.  Each edge has
+a \emph{polarity}, indicating whether (negative polarity) or not
+(positive polarity) the corresponding argument should be negated.
 
@@
 
-We define the {\em size} of POG $P$, written $|P|$, to be the
-the number of nodes labeled $\pand$ or $\por$ plus the number of edges from these nodes to their children.  Ring
+We define the \emph{size} of POG $P$, written $|P|$, to be the
+the number of nonterminal nodes plus the number of edges from these nodes to their children.  Ring
 evaluation of $P$ can be performed with at most $|P|$ ring
@@
   these have disjoint models.  By contrast, each
-  sum node $\nodeu$ in a dec-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
+  sum node $\nodeu$ in a dec-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a \emph{decision variable} $x$ such that
   any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
@@
 \end{itemize}
-The generalizations encompassed by POGs have also been referred to as {\em deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
+The generalizations encompassed by POGs have also been referred to as \emph{deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
 Our current proof generator only works for knowledge compilers
@@
 The status of the proof at any step is represented as
-a set of {\em active} clauses $\theta$, i.e., those that
+a set of \emph{active} clauses $\theta$, i.e., those that
 have been added but not yet deleted.
 Our framework is based
-on {\em extended} resolution~\cite{Tseitin:1983}, where proof
-steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
+on \emph{extended} resolution~\cite{Tseitin:1983}, where proof
+steps can introduce new \emph{extension variables} encoding Boolean formulas over input and prior extension variables.
 Let $Z$
@@
 Clauses can be added in two different ways.  One is when they serve as
-the {\em defining clauses} for an extension variable.  This form
+the \emph{defining clauses} for an extension variable.  This form
 occurs only when defining $\pand$ and $\por$ operations, as is
 described in Section~\ref{sect:cpog}.  Clauses can also be added or
-deleted based on {\em implication redundancy}.  That is, when clause
+deleted based on \emph{implication redundancy}.  That is, when clause
 $C$ satisfies $\theta \imply C$ for formula $\theta$, then it can either
@@
 
-We use {\em reverse unit propagation} (RUP) to certify
+We use \emph{reverse unit propagation} (RUP) to certify
 implication redundancy when adding or deleting
@@
 proved redundant with respect to formula $\theta$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
-{\em antecedent} clauses, such that each $D_i$ is in $\theta$.
+\emph{antecedent} clauses, such that each $D_i$ is in $\theta$.
 A RUP step
@@
 %% antecedent clauses from right to left will derive a clause $C'$ such
-%% that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
+%% that $C' \subseteq C$.  By \emph{subsumption} %~\cite{philipp:lpar:2017},
 %% we then have $C' \rightarrow C$.  Compared to listing each resolution
@@
 %% \begin{itemize}
-%% \item The {\em proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
+%% \item The \emph{proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
 %% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a dec-DNNF representation of the formula.
-%% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
+%% \item The \emph{proof checker} ensures that all of the proof conditions are satisfied.
 %% \end{itemize}
@@
   The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
-  Declaring a node $\nodeu$ implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
+  Declaring a node $\nodeu$ implicitly adds an \emph{extension} variable $u$ and a set of \emph{defining} clauses $\theta_{u}$
   encoding the product or sum operation.
@@
   A clause can only be added if it is logically implied by the existing clauses.
-  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
+  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
 \item
@@
   A clause can only be deleted if it is logically implied by the remaining clauses.
-  A sequence of clause identifiers must be listed as a {\em hint} providing a RUP verification of the implication.
+  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
 \item The checker must track the dependency set for every input and
@@
     \midrule
-    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause} \\
-     & {\tt d} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause \\
-    \midrule
-    $C$    & {\tt p} & $V \; L^{*}$ {\tt 0}    &                  & Declare $\pand$ operation \\
-    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\
-    \midrule
-     & {\tt r} & $L$             &            & Declare root literal\\
+    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\texttt{a}}  & \makebox[15mm][l]{$L^{*}$ \texttt{0}} & \makebox[15mm][l]{$C^{+}$ \texttt{0}}  & \makebox[20mm][l]{Add RUP clause} \\
+     & \texttt{d} & $C$             & $C^{+}$  \texttt{0} & Delete RUP clause \\
+    \midrule
+    $C$    & \texttt{p} & $V \; L^{*}$ \texttt{0}    &                  & Declare $\pand$ operation \\
+    $C$    & \texttt{s} & $V \; L \; L$    & $C^{+}$ \texttt{0}  & Declare $\por$ operation \\
+    \midrule
+     & \texttt{r} & $L$             &            & Declare root literal\\
     \bottomrule
@@
 Clauses can be added via RUP addition
-(command {\tt a}), with a sequence of antecedent clauses (the
+(command \texttt{a}), with a sequence of antecedent clauses (the
 ``hint'').
-Similarly for clause deletion (command {\tt d}).
+Similarly for clause deletion (command \texttt{d}).
 
@@
 
-The declaration of a {\em product} operation, creating a node with operation $\pand$,
+The declaration of a \emph{product} operation, creating a node with operation $\pand$,
  has the form:
@@
 \begin{tabular}{ccccccccc}
-  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
-  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\tt 0} \\
+  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
+  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\texttt{0}} \\
 \end{tabular}
@@
 
-The declaration of a {\em sum} operation, creating a node with operation $\por$, has the form:
+The declaration of a \emph{sum} operation, creating a node with operation $\por$, has the form:
 \begin{center}
 \begin{tabular}{ccccccc}
-  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$}
+  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$}
 \makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
@@
 Finally, the literal denoting the root of the POG is declared with the
-{\tt r} command.  It can occur anywhere in the file.  Except in degenerate cases, it
+\texttt{r} command.  It can occur anywhere in the file.  Except in degenerate cases, it
 will be the extension variable representing the root of a graph.
@@
 prove that any total assignment $\assign$ to the input variables which
-also satisfies the POG definitions will
+also satisfies the POG defining clauses will
 assign a value to $u$ such that $\assign(u) =
@@
 clause deletions represents a systematic transformation of the input formula
-into a POG formula.  Validating all of these steps serves to prove that
+into a POG\@.  Validating all of these steps serves to prove that
 POG $P$ is logically equivalent to the input formula.
@@
 Let $\inputformula$ denote the input formula.
-The sequence of clause addition steps provides a {\em forward implication} proof that
+The sequence of clause addition steps provides a \emph{forward implication} proof that
 $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
@@
 total assignment $\alpha$ that falsifies the clause must
-falsify $\phi_\noder$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
+falsify $\phi_\noder$.  Deleting all but the final asserted clause and all input clauses provides a \emph{reverse implication} proof
 that
@@
 A dec-DNNF
-graph can be directly translated  into a POG,
-although in doing this conversion,
-our program also performs simplifications to
+graph can be directly translated  into a POG.
+In doing this conversion,
+our program performs simplifications to
 eliminate Boolean constants.
@@
 In addition, negation is only
-applied to variables, and so we can merge the negation operations into the terminal nodes,
-viewing the POG as consisting
-of {\em literal} nodes corresponding to input variables and their negations, along with
-{\em nonterminal} nodes, which can be further classified as {\em product} and {\em sum} nodes.
+applied to variables, and so the only edges with negative polarity will have variables as children.
+We can therefore view
+view the POG as consisting
+of \emph{literal} nodes corresponding to input variables and their negations, along with
+\emph{nonterminal} nodes, which can be further classified as \emph{product} and \emph{sum} nodes.
 
@@
 $\{r\}$ as a unit clause.  We have devised two methods of generating this
-proof.  The {\em monolithic} approach makes just one call to a
+proof.  The \emph{monolithic} approach makes just one call to a
 proof-generating SAT solver and has it determine the relationship
-between the two representations.  The {\em structural} approach
+between the two representations.  The \emph{structural} approach
 recursively traverses the POG, generating proof obligations at each
@@
 which $\passign(\lit) = 0$, and iii) eliminating any duplicate clauses.
-In doing these simplifications, we also track the {\em provenance}
+In doing these simplifications, we also track the \emph{provenance}
 of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
@@
 The procedure adds a number of clauses to the proof, culminating with
-the addition of the {\em target} clause:
+the addition of the \emph{target} clause:
 $u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
@@
 
-The recursive call $\validate(u, \passign, \psi)$ assumes that we have
+The recursive call $\validate(\nodeu, \passign, \psi)$ assumes that we have
 traversed a path from the root node down to node $\nodeu$, with the
@@
 
-Working upward through the graph, node $\nodeu$ is {\em marked} when
-the collected set of literals force $u$ to evaluate to $0$.  When marking $\nodeu$, the
+Working upward through the graph, node $\nodeu$ is \emph{marked} when
+the collected set of literals forces $u$ to evaluate to $0$.  When marking $\nodeu$, the
 program adds $\obar{u}$ to the RUP literals and adds the appropriate
@@
 
-{\em Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
+\emph{Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
 having many---tens or even hundreds---of literals as children.  The
@@
 
-{\em Lemmas:} Our formulation of $\validate$ requires each call at a
+\emph{Lemmas:} Our formulation of $\validate$ requires each call at a
 node $\nodeu$ to recursively validate all of its children.  This
@@
 $\nodeu$ having $\indegree(\nodeu) > 1$, the program can define and
-generate the proof of a {\em lemma} for $\nodeu$ when it is first reached by
+generate the proof of a \emph{lemma} for $\nodeu$ when it is first reached by
 a call to $\validate$ and then apply this lemma for this and
@@
 \section{Implementations}
-We have implemented versions of the three programs that, along with
+We have implemented programs that, along with
 the \dfour{} knowledge compiler, form the toolchain illustrated in
@@
 cases, since it need not be trusted.
-Our {\em verified}
-versions of the proof checker and ring evaluator have been formally
+Our \emph{verified}
+version of the proof checker and ring evaluator have been formally
 verified within the \lean{} theorem prover.  Our long term goal is to
-rely on these.  Our {\em prototype} versions are written in C
-(checker) and Python (ring evaluator).  These are currently faster and
-more scalable, but we anticipate their need will diminish as the
+rely on these.  Our \emph{prototype} version is written in C\@.
+It is faster and
+more scalable, but we anticipate its need will diminish as the
 verified version is further optimized.
@@
 
-Our prototype ring evaluator can perform both standard and weighted
-model counting.  It performs arithmetic over a subset of the rationals
+We incorporated a ring evaluator into the prototype checker.  It
+can perform both standard and weighted
+model counting with full precision.  It performs arithmetic over a subset of the rationals
 we call $\drational$, consisting of numbers of the form $a \cdot 2^{b}
@@
 \begin{minipage}{0.42\textwidth}
-(D) Implicit Clauses\\[1.2ex]
+(D) Defining Clauses\\[1.2ex]
 \begin{tabular}{llll}
@@
 lead to an exponential number of calls.
-These shared subgraphs arise when the knowledge compiler employs {\em clause caching}
+These shared subgraphs arise when the knowledge compiler employs \emph{clause caching}
 to detect that the simplified set of
@@
 The proof generator can exploit the sharing of subgraphs
-by constructing and proving a {\em lemma} for each node
+by constructing and proving a \emph{lemma} for each node
 $\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
@@
 implement these by adding POG product nodes to the CPOG file to create
-the appropriate clauses.  Consider an {\em argument} clause
+the appropriate clauses.  Consider an \emph{argument} clause
 $C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
@@
 $v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
-  synthetic} clause having $\obar{v}$ as the {\em guard literal}.
+  synthetic} clause having $\obar{v}$ as the \emph{guard literal}.
 That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
@@
 
+\subsection{Certifying Preprocessed Formulas}
+
+\input{figure-runtime-preprocess}
+\input{figure-clauses-preprocess}
+
+The \dfour{} knowledge compiler can optionally perform three different
+forms of preprocessing on the input formula.  These transform the
+formula while preserving logical equivalence~\cite{lagniez:aaai:2014}.
+Proving that the combination of preprocessing plus knowledge
+compilation preserved logical equivalence would be a valuable
+capability.  However, proof generators that operate based on the
+structure of the input clauses, including our structural approach,
+will generally fail in this case, since the compilation was not based
+on the original input clauses.
+
+One strength of monolithic proof generation is that it makes no
+assumptions about how the compiled representation.  A complete
+proof-generating SAT solver could, in principle, always generate the
+forward implication proof, as long as forward implication holds.  To
+test this hypothesis, we ran our proof framework with preprocessing
+enabled on the same 83 benchmarks as in
+Section~\ref{app:experiment:monolithic}.  We measured the time for two
+entire toolchains: one with compilation, proof generation, and
+checking, and the other starting with preprocessing and then having
+the same steps as the other.  The results are summarized in
+Figures~\ref{fig:preprocess:time} and \ref{fig:preprocess:clauses}.
+Setting a time limit of 1000 seconds, 70 of the formulas completed
+without preprocessing, while 69 did so with preprocessing.  The one
+that timed out with preprocessing did so due to the excessive time
+required for preprocessing.  Of the 69, 53 ran faster with
+preprocessing, while 15 ran slower.  The sizes of the generated 
+CPOG proofs were more variable: 31 had fewer clauses with preprocessing, 21 had more, and 17 were identical.
+
 \subsection{Evaluation of the Verified Checker}
@@
 optimizing the verified checker.
+
+\section{Comparison with other Certification Frameworks}
+\label{app:comparison}
+
+This section provides more insights into how our proof framework (POG)
+compares with other work in certified knowledge compilation and model counting.
+
+\subsection{Comparison with \cdfour{} Framework}
+
+\input{figure-cd4-cpog}
+
+Figure~\ref{fig:cd4:cpog} shows the results of a set of experiments
+comparing the performance of our toolchain (CPOG) versus one based on
+\cdfour{}.  It shows the timings for the 110 benchmarks for which
+either the CPOG toolchain or the \cdfour{} toolchain completes in less
+than 1000 seconds.  Both toolchains include compilation, in addition
+to proof generation and checking.  Of these, 80 completed for both
+toolchains, 28 timed out or failed with the CPOG toolchain, and 2
+timed out for the \cdfour{} toolchain.  The data clearly show that the
+\cdfour{} toolchain can handle larger benchmarks and generally runs
+faster.  Only 7 of the 80 that completed with both toolchains ran
+faster with the CPOG toolchain.  For these 80, the toolchain based on
+\cdfour{} ran faster by a harmonic mean factor of $19.8\times$.
+
+On the other hand, the \cdfour{} proof generation and checking is
+tightly connected to the operation of \dfour{}, and every change to
+the program could require changes to the proof steps and possibly
+changes to the proof framework.  For example, we have found that
+\cdfour{} does not work properly when the compiler makes use of its
+internal preprocessor.
 
@@
 
-\input{figure-mice-cpog} The detailed results are shown in
+\input{figure-mice-cpog}
+
+The detailed results are shown in
 Figure~\ref{fig:mice}.  As can be seen, the comparative runtimes are

### 2023-05-19 tools/toolchain.py
Omit separate counter

@@ -1,7 +1,25 @@
 #!/usr/bin/python3
 
-# Run model counting program on benchmark file
-# Use newer versions of programs
+
+#####################################################################################
+# Copyright (c) 2023 Randal E. Bryant, Carnegie Mellon University
+# Last edit: May 19, 2022
+# 
+# Permission is hereby granted, free of charge, to any person obtaining a copy of this software and
+# associated documentation files (the "Software"), to deal in the Software without restriction,
+# including without limitation the rights to use, copy, modify, merge, publish, distribute,
+# sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is
+# furnished to do so, subject to the following conditions:
+# 
+# The above copyright notice and this permission notice shall be included in all copies or
+# substantial portions of the Software.
+# 
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT
+# NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+# NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM,
+# DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT
+# OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+########################################################################################
 
 import getopt
 import sys
@@ -11,10 +29,10 @@ import datetime
 import time
 
 def usage(name):
-    print("Usage: %s [-h] [-1] [-f] [-s n|g|c] [-m] [-p] [-L] [-G] [-F] [-t TIME] FILE.EXT ..." % name)
+    print("Usage: %s [-h] [-1] [-f] [-s n|g] [-m] [-p] [-L] [-G] [-F] [-t TIME] FILE.EXT ..." % name)
     print("  -h       Print this message")
     print("  -f       Force regeneration of all files")
-    print("  -s n|g|c Stop after NNF generation, CPOG generation (g) or proof check (c)")
+    print("  -s n|g   Stop after NNF generation or CPOG generation (g)")
[…]

### 2023-05-19 doc/mc-paper.tex
Incremental changes

@@
 
-%\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Randal E.
+Bryant, Wojciech Nawrocki, Jeremy Avigad, and Marijn J.
+H.
+Heule} %TODO mandatory, please use full first names.
+LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
   For Boolean formula $\phi$ and ring weight function $w$, the \emph{ring evaluation problem} computes
-  \begin{eqnarray}
+  \begin{equation}
+    \begin{array}{rcl}
     \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
-  \end{eqnarray}
+    \end{array}
+  \end{equation}
   In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
@@
 Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
-\begin{eqnarray}
-%% V1
+%%\begin{eqnarray}
+\begin{displaymath}
+\begin{array}{rcl}
+\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \\ %\label{eqn:ring:negation} \\
 \textstyle
-\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
+\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \\ %\label{eqn:ring:product} \\
 \textstyle
-\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
-\textstyle
-\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
-\end{eqnarray}
+\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \\ %\label{eqn:ring:sum}
+\end{array}
+\end{displaymath}
 \end{proposition}
@@
 ii) for the remaining clauses eliminating those literals $\lit$ for
-which $\passign(\lit) = 0$, and iii) eliminating any duplicate clauses.
+which $\passign(\lit) = 0$, and iii) eliminating any duplicate or tautological clauses.
 In doing these simplifications, we also track the \emph{provenance}

### 2023-05-20 doc/mc-paper.tex
Split appendix into supplement

@@
 
+We have developed an online supplement to this
+paper~\cite{bryant:sat:2023:supplement} containing additional
+information, including a worked example, details on the algorithms,
+and extensive experimental results.  The archive also includes copies
+of the programs and a set of benchmark files for testing.
+
 \section{Related Work}
@@
 concepts of clausal proof systems.  These factors were important in enabling formal verification.
-In Appendix~\ref{app:comparison}
+In the supplement~\cite{bryant:sat:2023:supplement},
 we also compare the performance of our toolchain to these two others.
@@
 
-Appendix~\ref{app:cpog:example} shows the CPOG description for
+The supplement~\cite{bryant:sat:2023:supplement},
+shows the CPOG description for
 an input formula with five clauses, yielding a POG with five
@@
 implication proof.  Here we summarize the two optimizations.
-Appendix~\ref{app:optimizations} provides more details.
+The supplement~\cite{bryant:sat:2023:supplement} provides more details.
 
@@
 will cause further recursive calls only once.  Our experimental
-results (Appendix~\ref{app:experiments}) show that these
+results~\cite{bryant:sat:2023:supplement} show that these
 optimizations yield substantial benefits.
@@
 We summarize the results of our experiments here.  
-Appendix~\ref{app:experiments} provides a more complete description. For our
+The supplement~\cite{bryant:sat:2023:supplement}
+provides a more complete description. For our
 evaluation, we used the public benchmark problems from the 2022
@@
 
-\newpage
-
 %% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
@@
 
+\newpage
+
 \bibliography{references}
 
-\newpage
-
-\appendix
-
-\section{CPOG Example}
-\label{app:cpog:example}
-
-\begin{figure}
-\begin{minipage}{0.58\textwidth}
-(A)  Input Formula\\[1.2ex]
-\begin{tabular}{lll}
-\toprule
-\makebox[5mm]{ID} & \makebox[15mm]{Clauses} & \\
-\midrule
-1 & \texttt{-1 3 -4} & \texttt{0} \\
-2 & \texttt{-1 -3 4} & \texttt{0} \\
-3 & \texttt{3 -4} & \texttt{0}\\
-4 & \texttt{1 -3 4} & \texttt{0} \\
-5 & \texttt{-1 -2} & \texttt{0} \\
-\bottomrule
-\end{tabular}
-\\[1.8ex]
-(C) POG Declaration\\[1.2ex]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
-\midrule
-6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
-9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
-12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
-15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\
-18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
-22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
- & \texttt{r 10} && Root $r = s_{10}$\\
-\bottomrule
-\end{tabular}
-\end{minipage}
-\begin{minipage}{0.35\textwidth}
-(B) POG Representation \\
-\input{dd/eg4}
-\end{minipage}
-%% Break
-\\[2.5ex]
-\begin{minipage}{0.42\textwidth}
-(D) Defining Clauses\\[1.2ex]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
-\midrule
-\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
-\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
-\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
-\midrule
-\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
-\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
-\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
-\midrule
-\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
-\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
-\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
-\midrule
-\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
-\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
-\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
-\midrule
-\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
-\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
-\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
-\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
-\midrule
-\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
-\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
-\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
-\bottomrule
-\end{tabular}
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-(E) CPOG Assertions\\[1.2ex]
-\begin{tabular}{llllll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
-\midrule
-\texttt{25} & \texttt{a 5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
-\texttt{26} & \texttt{a 6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
-\texttt{27} & \texttt{a 3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
-\texttt{28} & \texttt{a 7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
-\texttt{29} & \texttt{a 8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
-\texttt{30} & \texttt{a 5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
-\texttt{31} & \texttt{a 6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
-\texttt{32} & \texttt{a 3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
-\texttt{33} & \texttt{a 7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
-\texttt{34} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
-\texttt{35} & \texttt{a 1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
-\texttt{36} & \texttt{a 10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
-\bottomrule
-\end{tabular}
-\\[1.5ex]
-(F) Input Clause Deletions\\[1.2ex]
-\begin{tabular}{lllll}
-  \toprule
- \multicolumn{3}{l}{CPOG line} & Explanation\\
-\midrule
- \texttt{d 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
- \texttt{d 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
- \texttt{d 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
- \texttt{d 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
- \texttt{d 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
-\bottomrule
-\end{tabular}
-\end{minipage}
-\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F)}
-\label{fig:eg4:all}
-\end{figure}
-
-Figure \ref{fig:eg4:all} illustrates an example formula and shows how
-the CPOG file declares its POG representation.  The input formula (A)
-consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
-$x_4$.  The generated POG (B) has six nonterminal nodes representing
-four products and two sums.  We name these by the node
-type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
-extension variable.
-% We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
-  The first part of the CPOG file (C) declares
-these nodes using clause IDs that increment by three or four,
-depending on whether the node has two children or three.  The last two
-nonzero values in the sum declarations are the hint providing the
-required mutual exclusion proof.
-
-\subsection{Node Declarations}
-
-We step through portions of the file to provide a better understanding of the CPOG proof framework.
-Figure
-\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
-defined by the POG operation declarations.  These do not appear in the
-CPOG file.  Referring back to the declarations of the sum nodes in
-Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
-$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
-\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
-$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
-have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
-clause IDs 16 and 19 as the hint.  These form a RUP proof for the clause
-$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
-$\nodes_{10}$ have disjoint models.
-
-\subsection{Forward Implication Proof}
-
-Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
-leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $\nodes_{10}$ is implied by the input clauses, i.e.,
-any total assignment $\assign$
-satisfying the input clauses must have $\assign(s_{10}) = 1$.
-Working backward, we can see that
-clause 29 indicates that variable $p_8$ will be implied by the input
-clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
-be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
-hint for clause 36.
-
-\subsection{Reverse Implication Proof}
-
-Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
-delete the input clauses.  Consider the first of these, deleting
-input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
-that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
-The proof proceeds by first assuming that the clause is false, requiring
-$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
-clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
-clauses that cause unit propagation.  Hint clauses 8 and 10 force the
-assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
-$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
-via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
-completing the RUP proof.  The deletion hints for the other input
-clauses follow similar patterns---they work from the bottom nodes of
-the POG upward, showing that any total assignment that falsifies the clause
-must assign $\assign(s_{10}) = 0$.
-
-Deleting the asserted clauses is so simple that we do not show it.  It
-involves simply deleting the clauses from clause number 35 down to
-clause number 25, with each deletion using the same hint as were used
-to add that clause.  In the end, therefore, only the defining clauses
-for the POG nodes and the unit clause asserting $s_{10}$ remain,
-completing a proof that the POG is logically equivalent to the input
-formula.
-
-Observe that the forward implication proof shown in Figure~\ref{fig:eg4:all}  must ``visit'' nodes
-$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
-(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).
-Section~\ref{app:lemma:eg} shows how to use a lemma such that these nodes are each only visited once.
-
-\section{Optimizations for Forward Implication Proofs}
-
-The two optimizations we have implemented exploit the power of our
-general resolution framework to define new structures within the
-proof.
-They create new product nodes that are not part
-of the POG representation.
-
-\label{app:optimizations}
-
-\subsection{Literal Grouping}
-
-A single recursive step of $\validate$ can encounter product nodes
-having many literals as children.  The naive formulation of $\validate$
-considers each literal $\lit \in \lambda$ separately.
-Literal grouping allows all literals to be validated with a single call to a SAT solver.
-It collects those literals
-$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
-product node $\nodev$ having these literals as children.  The goal
-then becomes to prove that any total assignment must yield 1 for extension
-variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
-  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
-  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
-  input clauses and the defining clauses for $\nodev$ into the hints.
-
-\subsection{Lemmas}
-\label{app:lemma}
-
-As we have noted, the recursive calling of $\validate$ starting at
-root $\noder$ effectively expands the POG into a tree, and this can
-lead to an exponential number of calls.
-These shared subgraphs arise when the knowledge compiler employs \emph{clause caching}
-to detect that the simplified set of
-clauses arising from one partial assignment to the literals matches that
-of a previous partial assignment~\cite{darwiche:aaai:2002}.
-When this dec-DNNF node is translated into POG
-node $\nodeu$, the proof generator can assume (and also check), that
-there is a simplified set of clauses $\gamma_{\nodeu}$
-for which the subgraph with root $\nodeu$ is its POG representation.
-
-The proof generator can exploit the sharing of subgraphs
-by constructing and proving a \emph{lemma} for each node
-$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
-total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ and the defining clauses for the POG must yield
-$\assign(u) = 1$.  This lemma is then invoked for every node having
-$\nodeu$ as a child.
-As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.
-
-The challenge for implementing this strategy is to find a way to
-represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CPOG file.  Some may be
-unaltered input clauses, and these can be used directly.  Others,
-however will be clauses that do not appear in the input formula.  We
-implement these by adding POG product nodes to the CPOG file to create
-the appropriate clauses.  Consider an \emph{argument} clause
-$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
-define product node $\nodev$ with arguments
-$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, then by
-Table~\ref{tab:defining}(A), we will introduce a defining clause
-$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
-  synthetic} clause having $\obar{v}$ as the \emph{guard literal}.
-That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
-  activate} the clause, causing it to represent argument clause $C$.  On the other
-hand, a partial assignment with $\passign(v) = 1$ will
-cause the clause to become a tautology and therefore have no effect.
-
-Suppose for every clause $C_j \in \gamma_{\nodeu}$ that does not correspond to
-an input clause, we generate a synthetic clause $C'_j$ with guard literal
-$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma'_{\nodeu}$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
-while input clauses in $\gamma_{\nodeu}$ are left unchanged.
-Let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
-Invoking $\validate(\nodeu, \lassign, \gamma'_{\nodeu})$
- will then prove a lemma, given by the target clause
- $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
- showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u) = 1$.
-
-Later, when node $\nodeu$ is encountered by a call to $\validate(\nodeu, \passign, \psi)$, we invoke the lemma
-by showing that each synthetic clause
-$C_j$ matches some simplified clause in $\simplify{\psi}{\passign}$.  More precisely,
-for $1 \leq j \leq m$,
-we use clause addition to assert the clause
-$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$,
-showing that synthetic clause $C_j$ will be activated.
-Combining the lemma with these activations provides a derivation of the target clause for the call to $\validate$.
-
-Observe that the lemma structure can be hierarchical, since a shared
-subgraph may contain nodes that are themselves roots of shared
-subgraphs.  Nonetheless, the principles described allow the
-definition, proof, and applications of a lemma for each shared node in
-the graph.  For any node $\nodeu$, the first call to
-$\validate(\nodeu, \passign, \psi)$ may require further recursion,
-but any subsequent call can simply reuse the lemma proved by the first call.
-
-\subsection{Lemma Example}
-\label{app:lemma:eg}
-
-\begin{figure}
-(A) Additional nodes\\[1.0em]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
-\midrule
-25 & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
-28 & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
-\bottomrule
-\end{tabular}
-\\[1.0em]
-(B) Implicit Clauses\\[1.2em]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
-\midrule
-\texttt{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
-\texttt{26} & \texttt{-11 -3} & \texttt{0} & \\
-\texttt{27} & \texttt{-11 4} & \texttt{0} & \\
-\midrule
-\texttt{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
-\texttt{29} & \texttt{-12 3} & \texttt{0} & \\
-\texttt{30} & \texttt{-12 -4} & \texttt{0} & \\
-\bottomrule
-\end{tabular}
-\\[1.0em]
-(C) CPOG Assertions\\[1.0em]
-\begin{tabular}{llllll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
-\midrule
-\multicolumn{6}{l}{Lemma Proof} \\
-\texttt{31} & \texttt{a 5 11 12 3} & \texttt{0} & \texttt{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
-\texttt{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \texttt{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
-\texttt{33} & \texttt{a 3 7 11 12} & \texttt{0} & \texttt{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
-\texttt{34} & \texttt{a 7 11 12} & \texttt{0} & \texttt{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
-\midrule
-\multicolumn{6}{l}{Lemma Application \#1} \\
-\texttt{35} & \texttt{a -11 1} & \texttt{0} & \texttt{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
-\texttt{36} & \texttt{a -12 1} & \texttt{0} & \texttt{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
-\texttt{37} & \texttt{a 7 1} & \texttt{0} & \texttt{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
-\midrule
-\texttt{38} & \texttt{a 8 1} & \texttt{0} & \texttt{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
-\midrule
-\multicolumn{6}{l}{Lemma Application \#2} \\
-\texttt{39} & \texttt{a -11 -1} & \texttt{0} & \texttt{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
-\texttt{40} & \texttt{a -12 -1} & \texttt{0} & \texttt{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
-\texttt{41} & \texttt{a 7 -1} & \texttt{0} & \texttt{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
-\midrule
-\texttt{42} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
-\texttt{43} & \texttt{a 1 10} & \texttt{0} & \texttt{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
-\texttt{44} & \texttt{a 10} & \texttt{0} & \texttt{43 24 42} & \texttt{0} & $s_{10}$ \\
-\bottomrule
-\end{tabular}
-\caption{Example of lemma definition, proof, and application}
-\label{fig:eg4:lemmas}
-\end{figure}
-
-Figure~\ref{fig:eg4:lemmas} shows an alternate forward implication
-proof for the example of Figure~\ref{fig:eg4:all} using a lemma to
-represent the shared node $\nodes_7$.  We can see that the POG with
-this node as root encodes the Boolean formula $x_3 \leftrightarrow x_4$, having a CNF representation consisting of the clauses
-$\{x_3 ,\, \obar{x}_4\}$ and $\{\obar{x}_3 ,\, {x}_4\}$.  The product node
-declarations shown in Figure~\ref{fig:eg4:lemmas}(A) create synthetic
-clauses 25 and 28 to encode these arguments with activating literals
-$\obar{v}_{11}$ and $\obar{v}_{12}$, respectively.  Clauses 31--34
-then provide a proof of the lemma, stating that any assignment
-$\assign$ that activates these clauses will  assign $1$ to $s_7$.
-Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
-will cause the first synthetic clause to activate due to input clause
-3, and it will cause the second synthetic clause to activate due to
-input clause 4.  From this, clause 37 can use the lemma to state that
-assigning $0$ to $x_1$ will cause $s_7$ to evaluate to $1$.  Similarly,
-clauses 39 and 40 serve to activate the synthetic clauses when
-$\assign(x_1) = 1$, due to input clauses 1 and 2, and clause 41 then
-uses the lemma to state that assigning $1$ to $x_1$ will cause $s_7$ to
-evaluate to $1$.
-
-In this example, adding the lemma increases the proof length, but that
-is only because it is such a simple formula.
-
-\input{leanappendix}
-
-\section{Detailed Experimental Results}
-\label{app:experiments}
-
-Our experiments are designed to evaluate the following:
-\begin{itemize}
-\item
-  whether our toolchain can handle challenging benchmark problems,
-\item
-the effectiveness of some design choices and optimizations,
-\item
-the comparative performance of the prototype and verified checkers, and
-\item
-how the performance of our toolchain compares to that of the MICE model counter verifier.
-\end{itemize}
-
-\subsection{Experimental Setup}
-
-As a test machine, we used a 2021~MacBook~Pro laptop having a 3.2~Ghz
-Apple M1 processor and with 64~GB of physical memory.  We also used an
-external 3~TB solid-state disk drive with an advertised read/write
-speed of one gigabit per second.  Despite being a laptop, this
-configuration is somewhat more powerful than the nodes typically
-found in server clusters.  The fast access to storage is especially
-important due to the very large files generated and processed by the
-tools.  Our tools generated individual files with over 160 gigabytes
-of data.
-
-For benchmarks, we downloaded 100 files each from the 2022 standard and weighted model counting competitions:
-\begin{center}
-       \url{https://mccompetition.org/2022/mc_description.html}
-\end{center}
-We found that 20 of these were duplicates across the two tracks,
-yielding a total of 180 unique benchmark problems, ranging in size
-from 250 to 2,753,207 clauses.
-
-Our default configuration for the proof generator used the structural
-method for the forward implication proof, with the two optimizations
-of literal grouping and lemmas.
-Running with a time limit of 4000 seconds\footnote{All measured times
-in this document are actual elapsed times, not CPU times.}, \dfour{}
-was able to generate dec-DNNF representations for 124 of these, and it timed out on the
-rest.  Our proof generator was able to convert all of the generated
-dec-DNNF graphs into POGs and determine how many defining clauses they
-would generate.  A POG operation with $k$ arguments requires $k+1$
-defining clauses, and so the total number of defining clauses in POG $P$ equals
-the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$ in Section~\ref{sect:pog}.
-The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.
-
-\subsection{Performance of the Proof Generator and Checker}
-
-\input{figure-runtime-d4-verify}
-\input{figure-clauses-d4-verify}
-
-For each of the 124 dec-DNNF graphs generated by \dfour{}, we ran
-our proof generator and prototype checker, limiting proof
-generation to 10,000 seconds.  We ran the programs to generate and check
-one-sided proofs for the graphs, again with a time limit of 10,000
-seconds for proof generation.
-
-Figure~\ref{fig:d4:cpog} summarizes our results in terms of the time
-required by \dfour{} (X axis) versus the sum of the times for the
-proof generator and checker (Y axis).
-We were able to complete a full validation for 108 of the 124
-benchmarks, with times ranging from fractions of a second to 13,144
-seconds, with a median of 71.6 seconds.
-
-Relative to the runtime for \dfour{}, two problems ran faster with the
-generator and checker.  These were for problems having small numbers
-of models (relative to the number of variables), and so most of the
-time spent by both \dfour{} and our proof generator was in running a
-CDCL solver to search the very sparse solution space.  \cadical{}
-generally outperforms the miniSAT-based solver used by \dfour{}.  At
-the other extreme, one problem that required only 19 seconds for
-\dfour{} required 8657 seconds to generate a proof and 45 seconds to
-check it.  This benchmark appears to stem from weak performance
-by our implementation of BCP\@.  Overall the ratios
-between the combined generation plus checking times versus the time
-for \dfour{} had a harmonic mean of 5.5.
-
-Of the 16 benchmarks that could not be fully validated, one had so
-many defining clauses that it overflowed the 32-bit signed integers
-our programs use for clause identifiers.  Another exited due to the
-virtual memory limit imposed by the operating system, and the other 14
-timed out.  Of the 15 that did not overflow the clause counter, we
-were able to complete a one-sided verification for 9, but the other 6
-timed out.  Overall, we were able to provide some level of
-verification for 94\% of the benchmark problems.
-
-Figure~\ref{fig:defining:total} compares the number of defining
-clauses (X axis) with the total number of clauses (Y axis) for the 108
-problems that were fully verified.  These totals include the defining
-clauses for the POG, the additional defining clauses introduced to
-support literal grouping and lemmas, and the clauses added by RUP
-steps in the forward implication proof.  These totals ranged from 554
-to 770,382,773, with a median of 1,719,245.  The ratios of the total
-clauses versus defining clauses ranged from 1.54 to 6073, with a
-harmonic mean of 3.13.  The high numbers were for problems having very
-few models, and so the proof clauses must encode large proofs of
-unsatisfiability.
-
-\subsection{Monolithic Forward Implication Proofs}
-\label{app:experiment:monolithic}
-
-To assess the relative merits of the two approaches to forward
-implication proof generation, we ran the proof generator in monolithic
-mode on the 92 problems for which the combination of structural proof generation
-and checking required at most 3000 seconds and then kept only those
-measurements for which at least one of the two approaches had combined
-times (generation plus checking) of at most 1000 seconds.  There were
-a total of 83 such problems.  Figures~\ref{fig:monolithic:time} and
-\ref{fig:monolithic:clauses} show comparisons between the two
-approaches in terms of time and total clauses.
-
-Examining Figure~\ref{fig:monolithic:time}, we can see that the
-monolithic approach ran faster than the structural approach for 23 of
-the problems, including a case where the generator and checker
-required only one second, versus 71 seconds for the structural
-approach.  On the other hand, the monolithic approach timed out for 13
-of the cases and was slower for the other 37.  In general, we can see
-the monolithic approach faring worse on larger problems.
-
-Figure~\ref{fig:monolithic:clauses} shows the total number of clauses
-for the two approaches for the 70 cases where both approaches completed within 1000 seconds.
-As can be seen, the monolithic approach consistently produces
-shorter proofs, perhaps due to the benefits of the proof trimming by
-\dtrim{}.
-
-\input{figure-runtime-monolithic}
-\input{figure-clauses-monolithic}
-
-These experiments suggest that a hybrid of the two approaches might
-yield the best results in terms of consistency, runtime, and proof
-size.  It would begin the recursive descent of $\validate$ according
-to a structural approach, but monolithically generate
-the proof once it encounters a sufficiently small subgraph.
-
-\subsection{Impact of Optimizations}
-\label{app:experiment:optimize}
-
-\input{figure-clauses-optimizations}
-
-To assess the impact of the two optimizations: literal grouping and
-lemmas, we ran the proof generator on the 80 problems for which the
-combination of the two optimizations yielded proofs totaling at most
-10~million clauses.  Compared to the problems used to assess monolithic
-proof generation (Section~\ref{app:experiment:monolithic}), this set included one that had a short proof but
-long runtime and excluded four that had short runtimes but long
-proofs.  We then ran the proof generator on all three combinations of
-the optimizations being partially or totally disabled.
-Figure~\ref{fig:optimized:lessoptimized} shows how disabling these optimizations affected the
-total number of clauses in the proofs, where the X axis indicates the
-total number with both optimizations enabled, while the Y axis
-indicates the number with one (left) or both (right) optimizations
-disabled.
-
-As the figures illustrated, each optimization on its own can shorten
-the proofs, sometimes substantially, and the two combine to have even
-greater effect.  Of the 80 benchmark problems evaluated, and comparing
-with both optimizations disabled, 47 had the number of proof clauses
-reduced by a factor of 2 by using lemmas, 14 by using literal merging,
-and 60 by enabling both optimizations.  In the extreme cases, a lack
-of lemmas caused one proof to grow by a factor of 52.5, while a lack
-of literal grouping causes another proof to grow by a factor of 39.6.
-
-The time performance of these optimizations is less dramatic, but
-still significant.  Lemmas sometimes made proof generation slower, but
-in one case ran $38\times$ faster.  Similarly, literal grouping could
-take longer, but in one case ran $3.9\times$ faster.
-
-Overall, it is clear that these two optimizations are worthwhile, and
-sometimes critical, to success for some benchmarks.
-
-\subsection{Certifying Preprocessed Formulas}
-
-\input{figure-runtime-preprocess}
-\input{figure-clauses-preprocess}
-
-The \dfour{} knowledge compiler can optionally perform three different
-forms of preprocessing on the input formula.  These transform the
-formula while preserving logical equivalence~\cite{lagniez:aaai:2014}.
-Proving that the combination of preprocessing plus knowledge
-compilation preserved logical equivalence would be a valuable
-capability.  However, proof generators that operate based on the
-structure of the input clauses, including our structural approach,
-will generally fail in this case, since the compilation was not based
-on the original input clauses.
-
-One strength of monolithic proof generation is that it makes no
-assumptions about how the compiled representation.  A complete
-proof-generating SAT solver could, in principle, always generate the
-forward implication proof, as long as forward implication holds.  To
-test this hypothesis, we ran our proof framework with preprocessing
-enabled on the same 83 benchmarks as in
-Section~\ref{app:experiment:monolithic}.  We measured the time for two
-entire toolchains: one with compilation, proof generation, and
-checking, and the other starting with preprocessing and then having
-the same steps as the other.  The results are summarized in
-Figures~\ref{fig:preprocess:time} and \ref{fig:preprocess:clauses}.
-Setting a time limit of 1000 seconds, 70 of the formulas completed
-without preprocessing, while 69 did so with preprocessing.  The one
-that timed out with preprocessing did so due to the excessive time
-required for preprocessing.  Of the 69, 53 ran faster with
-preprocessing, while 15 ran slower.  The sizes of the generated 
-CPOG proofs were more variable: 31 had fewer clauses with preprocessing, 21 had more, and 17 were identical.
-
-\subsection{Evaluation of the Verified Checker}
-\label{app:lean:evaluation}
-
-\input{figure-check}
-
-We ran the \lean{} checker on the 80 benchmark problems for which the
-generated CPOG file had at most 10~million total clauses.  These are the
-same benchmarks used in evaluating the optimizations (Section
-\ref{app:experiment:optimize}).
-All of the checks completed and confirmed the outcome of the prototype checker.
-
-Figure~\ref{fig:lean} shows the
-performance of both the \lean{} checker and the prototype
-checker on the Y axis, as functions of the total number of clauses on the X axis.
-As would be expected, the \lean{} checker consistently runs slower
-than the prototype checker.
-The ratio of the runtime for \lean{} versus the runtime for
-the prototype checker had a harmonic mean of $5.94$.
-
-Importantly, however, it can be seen that both checkers show the same
-overall performance trends.  The ratio of the two runtimes was less
-than $8.0$ for all but three small benchmarks.  This seems like a
-reasonable price to pay for a rigorous guarantee of correctness, and
-we are confident that we can reduce this gap with more effort in
-optimizing the verified checker.
-
-\section{Comparison with other Certification Frameworks}
-\label{app:comparison}
-
-This section provides more insights into how our proof framework (POG)
-compares with other work in certified knowledge compilation and model counting.
-
-\subsection{Comparison with \cdfour{} Framework}
-
-\input{figure-cd4-cpog}
-
-Figure~\ref{fig:cd4:cpog} shows the results of a set of experiments
-comparing the performance of our toolchain (CPOG) versus one based on
-\cdfour{}.  It shows the timings for the 110 benchmarks for which
-either the CPOG toolchain or the \cdfour{} toolchain completes in less
-than 1000 seconds.  Both toolchains include compilation, in addition
-to proof generation and checking.  Of these, 80 completed for both
-toolchains, 28 timed out or failed with the CPOG toolchain, and 2
-timed out for the \cdfour{} toolchain.  The data clearly show that the
-\cdfour{} toolchain can handle larger benchmarks and generally runs
-faster.  Only 7 of the 80 that completed with both toolchains ran
-faster with the CPOG toolchain.  For these 80, the toolchain based on
-\cdfour{} ran faster by a harmonic mean factor of $19.8\times$.
-
-On the other hand, the \cdfour{} proof generation and checking is
-tightly connected to the operation of \dfour{}, and every change to
-the program could require changes to the proof steps and possibly
-changes to the proof framework.  For example, we have found that
-\cdfour{} does not work properly when the compiler makes use of its
-internal preprocessor.
-
-\subsection{Comparison with the MICE Framework}
-
-To compare the performance of our toolchain versus that of
-MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
-generation, proof checking, and ring evaluation in the CPOG toolchain compares to the time to
-run their proof generator \progname{nnf2trace} and checker
-\progname{sharptrace}.  We add the time for ring evaluation on our
-side, since their checker provides a count, although only an
-unweighted one.  We used the prototype versions of the checker and ring evaluator.
-We ran their tools on the 92 benchmark problems for
-which our toolchain requires at most 3000 seconds and retained
-the 84 data points for which at least one of the toolchains completes in under 1000 seconds.
-
-We can summarize the results as:
-\begin{itemize}
-\item 84 total data points
-\item 76 were completed by both programs in under 1000 seconds
-\item Of those 21 were faster with MICE, 55 with CPOG
-\item 7 completed with CPOG under 1000 seconds but exceeded that threshold for MICE
-\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CPOG
-\end{itemize}
-
-\input{figure-mice-cpog}
-
-The detailed results are shown in
-Figure~\ref{fig:mice}.  As can be seen, the comparative runtimes are
-highly variable, reflecting the fact that the two toolchains differ
-fundamentally in their objectives and their approaches.
-
-One shortcoming of \progname{nnf2trace} can be seen in
-the near-vertical series of points in the upper-lefthand corner.
-These correlate with a similar set of points along the left in
-Figure~\ref{fig:optimized:lessoptimized}.  Like the naive
-implementation of $\validate$, their program recursively traverses the
-graph representation of a formula, effectively expanding it into a
-tree.  They lack an optimization analogous to lemmas.
-
-We have not tried the MICE toolchain on larger benchmarks, but we
-suspect it would encounter significant scaling limitations.
-
 \end{document}

### 2023-05-20 doc/mc-supplement.tex
Split appendix into supplement

\documentclass[letterpaper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{hyperref}

\newcommand{\pand}{\mathbin{\land^\textsf{p}}}
\newcommand{\por}{\mathbin{\lor^\textsf{p}}}
\DeclareMathOperator*{\Pand}{\bigwedge^\textsf{p}}
\DeclareMathOperator*{\Por}{\bigvee^\textsf{p}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
%\newcommand{\drational}{\mathbb{Q}_{2,5}}
\newcommand{\drational}{\textbf{Q}_{2,5}}
\newcommand{\entail}{\vDash}
\newcommand{\entaildrat}{\entail_{\textrm{DRAT}}}

\newcommand{\assign}{\alpha}
\newcommand{\passign}{\rho}
\newcommand{\lassign}{\beta}
\newcommand{\uassign}{{\cal U}}
\newcommand{\modelset}{{\cal M}}

\newcommand{\indegree}{\textrm{indegree}}
\newcommand{\outdegree}{\textrm{outdegree}}
\newcommand{\validate}{\textsf{validate}}
\newcommand{\prov}{\textrm{Prov}}
\newcommand{\inputformula}{\phi_I}
\newcommand{\pogformula}{\theta_P}

\newcommand{\makenode}[1]{\mathbf{#1}}
\newcommand{\nodeu}{\makenode{u}}
\newcommand{\nodev}{\makenode{v}}
\newcommand{\nodes}{\makenode{s}}
\newcommand{\nodep}{\makenode{p}}
\newcommand{\noder}{\makenode{r}}

\newcommand{\simplify}[2]{#1|_{#2}}

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\cdfour}{\progname{CD4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}
\newcommand{\Dtrim}{\progname{Drat-trim}}

\definecolor{redorange}{rgb}{0.878431, 0.235294, 0.192157}
\definecolor{lightblue}{rgb}{0.552941, 0.72549, 0.792157}
\definecolor{clearyellow}{rgb}{0.964706, 0.745098, 0}
\definecolor{clearorange}{rgb}{0.917647, 0.462745, 0}
\definecolor{mildgray}{rgb}{0.54902, 0.509804, 0.47451}
\definecolor{softblue}{rgb}{0.643137, 0.858824, 0.909804}
\definecolor{bluegray}{rgb}{0.141176, 0.313725, 0.603922}
\definecolor{lightgreen}{rgb}{0.709804, 0.741176, 0}
\definecolor{darkgreen}{rgb}{0.152941, 0.576471, 0.172549}
\definecolor{redpurple}{rgb}{0.835294, 0, 0.196078}
\definecolor{midblue}{rgb}{0, 0.592157, 0.662745}
\definecolor{clearpurple}{rgb}{0.67451, 0.0784314, 0.352941}
\definecolor{browngreen}{rgb}{0.333333, 0.313725, 0.145098}
\definecolor{darkestpurple}{rgb}{0.396078, 0.113725, 0.196078}
\definecolor{greypurple}{rgb}{0.294118, 0.219608, 0.298039}
\definecolor{darkturquoise}{rgb}{0, 0.239216, 0.298039}
\definecolor{darkbrown}{rgb}{0.305882, 0.211765, 0.160784}
\definecolor{midgreen}{rgb}{0.560784, 0.6, 0.243137}
\definecolor{darkred}{rgb}{0.576471, 0.152941, 0.172549}
\definecolor{darkpurple}{rgb}{0.313725, 0.027451, 0.470588}
\definecolor{darkestblue}{rgb}{0, 0.156863, 0.333333}
\definecolor{lightpurple}{rgb}{0.776471, 0.690196, 0.737255}
\definecolor{softgreen}{rgb}{0.733333, 0.772549, 0.572549}
\definecolor{offwhite}{rgb}{0.839216, 0.823529, 0.768627}
\definecolor{medgreen}{rgb}{0.15, 0.6, 0.15}

% Lean code:
\usepackage{listings}
%\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
\definecolor{keywordcolor}{rgb}{0.0, 0.1, 0.6}   % blue
\definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{commentcolor}{rgb}{0.4, 0.4, 0.4}   % grey
\definecolor{symbolcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{sortcolor}{rgb}{0.1, 0.5, 0.1}      % green
\definecolor{attributecolor}{rgb}{0.7, 0.1, 0.1} % red
\def\lstlanguagefiles{lstlean.tex}
% set default language
\lstset{language=lean, xleftmargin=1em}
\lstset{backgroundcolor=\color{white}}

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{SUPPLEMENT to \\ Certified Knowledge Compilation \\ with Application to Verified Model Counting}
\hideLIPIcs

\titlerunning{Certified Knowledge Compilation}

\author{Randal E. Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{Randy.Bryant@cs.cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}
\author{Wojciech Nawrocki}{Department of Philosophy, Carnegie Mellon University}{wjnawrocki@cmu.edu}{https://orcid.org/0000-0002-8839-0618}{Hoskinson Center for Formal Mathematics}
\author{Jeremy Avigad}{Department of Philosophy, Carnegie Mellon University}{avigad@cmu.edu}{https://orcid.org/0000-0003-1275-315X}{Hoskinson Center for Formal Mathematics}
\author{Marijn J. H. Heule}{Computer Science Department, Carnegie Mellon University}{marijn@cmu.edu}{https://orcid.org/0000-0002-5587-8801}{Supported by NSF grant CCF-2108521}

\authorrunning{R. E. Bryant et al.} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Randal E. Bryant, Wojciech Nawrocki, Jeremy Avigad, and Marijn J. H. Heule} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10003794</concept_id>
       <concept_desc>Theory of computation~Automated reasoning</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Automated reasoning}

%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
\keywords{Propositional model counting, Proof checking}

%\category{} %optional, e.g. invited paper

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\nolinenumbers %uncomment to disable line numbering

\EventEditors{Meena Mahajan and Friedrich Slivovsky}
\EventNoEds{2}
\EventLongTitle{26th International Conference on Theory and Applications of Satisfiability Testing (SAT 2023)}
\EventShortTitle{SAT 2023}
\EventAcronym{SAT}
\EventYear{2023}
\EventDate{July 4--8, 2023}
\EventLocation{Alghero, Italy}
\EventLogo{}
\SeriesVolume{271}
\ArticleNo{6}

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\EventEditors{John Q. Open and Joan R. Access}
%\EventNoEds{2}
%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
%\EventShortTitle{CVIT 2016}
%\EventAcronym{CVIT}
%\EventYear{2016}
%\EventDate{December 24--27, 2016}
%\EventLocation{Little Whinging, United Kingdom}
%\EventLogo{}
%\SeriesVolume{42}
%\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\program}[1]{\textsc{#1}}
\newcommand{\lean}{Lean~4}

\begin{document}

\maketitle

This document is a supplement to the paper
``Certified Knowledge Compilation with Applications to Verified Model Counting,''
published at the 2023 Conference on Theory and Applications of SAT Testing (SAT).

\section{CPOG Example}
\label{app:cpog:example}

\begin{figure}
\begin{minipage}{0.58\textwidth}
(A)  Input Formula\\[1.2ex]
\begin{tabular}{lll}
\toprule
\makebox[5mm]{ID} & \makebox[15mm]{Clauses} & \\
\midrule
1 & \texttt{-1 3 -4} & \texttt{0} \\
2 & \texttt{-1 -3 4} & \texttt{0} \\
3 & \texttt{3 -4} & \texttt{0}\\
4 & \texttt{1 -3 4} & \texttt{0} \\
5 & \texttt{-1 -2} & \texttt{0} \\
\bottomrule
\end{tabular}
\\[1.8ex]
(C) POG Declaration\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
\midrule
6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\
18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
 & \texttt{r 10} && Root $r = s_{10}$\\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.35\textwidth}
(B) POG Representation \\
\input{dd/eg4}
\end{minipage}
%% Break
\\[2.5ex]
\begin{minipage}{0.42\textwidth}
(D) Defining Clauses\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
\midrule
\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
\midrule
\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
\midrule
\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
\midrule
\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
\midrule
\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
\midrule
\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.49\textwidth}
(E) CPOG Assertions\\[1.2ex]
\begin{tabular}{llllll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
\midrule
\texttt{25} & \texttt{a 5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
\texttt{26} & \texttt{a 6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
\texttt{27} & \texttt{a 3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
\texttt{28} & \texttt{a 7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
\texttt{29} & \texttt{a 8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
\texttt{30} & \texttt{a 5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
\texttt{31} & \texttt{a 6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
\texttt{32} & \texttt{a 3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
\texttt{33} & \texttt{a 7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
\texttt{34} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
\texttt{35} & \texttt{a 1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
\texttt{36} & \texttt{a 10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
\bottomrule
\end{tabular}
\\[1.5ex]
(F) Input Clause Deletions\\[1.2ex]
\begin{tabular}{lllll}
  \toprule
 \multicolumn{3}{l}{CPOG line} & Explanation\\
\midrule
 \texttt{d 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
 \texttt{d 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
 \texttt{d 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
 \texttt{d 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
 \texttt{d 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
\bottomrule
\end{tabular}
\end{minipage}
\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F)}
\label{fig:eg4:all}
\end{figure}

Figure \ref{fig:eg4:all} illustrates an example formula and shows how
the CPOG file declares its POG representation.  The input formula (A)
consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
$x_4$.  The generated POG (B) has six nonterminal nodes representing
four products and two sums.  We name these by the node
type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
extension variable.
% We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
  The first part of the CPOG file (C) declares
these nodes using clause IDs that increment by three or four,
depending on whether the node has two children or three.  The last two
nonzero values in the sum declarations are the hint providing the
required mutual exclusion proof.

\subsection{Node Declarations}

We step through portions of the file to provide a better understanding of the CPOG proof framework.
Figure
\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CPOG file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
clause IDs 16 and 19 as the hint.  These form a RUP proof for the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$\nodes_{10}$ have disjoint models.

\subsection{Forward Implication Proof}

Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $\nodes_{10}$ is implied by the input clauses, i.e.,
any total assignment $\assign$
satisfying the input clauses must have $\assign(s_{10}) = 1$.
Working backward, we can see that
clause 29 indicates that variable $p_8$ will be implied by the input
clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
hint for clause 36.

\subsection{Reverse Implication Proof}

Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
The proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any total assignment that falsifies the clause
must assign $\assign(s_{10}) = 0$.

Deleting the asserted clauses is so simple that we do not show it.  It
involves simply deleting the clauses from clause number 35 down to
clause number 25, with each deletion using the same hint as were used
to add that clause.  In the end, therefore, only the defining clauses
for the POG nodes and the unit clause asserting $s_{10}$ remain,
completing a proof that the POG is logically equivalent to the input
formula.

Observe that the forward implication proof shown in Figure~\ref{fig:eg4:all}  must ``visit'' nodes
$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).
Section~\ref{app:lemma:eg} shows how to use a lemma such that these nodes are each only visited once.

\section{Optimizations for Forward Implication Proofs}

The two optimizations we have implemented exploit the power of our
general resolution framework to define new structures within the
proof.
They create new product nodes that are not part
of the POG representation.

\label{app:optimizations}

\subsection{Literal Grouping}

A single recursive step of $\validate$ can encounter product nodes
having many literals as children.  The naive formulation of $\validate$
considers each literal $\lit \in \lambda$ separately.
Literal grouping allows all literals to be validated with a single call to a SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
  input clauses and the defining clauses for $\nodev$ into the hints.

\subsection{Lemmas}
\label{app:lemma}

As we have noted, the recursive calling of $\validate$ starting at
root $\noder$ effectively expands the POG into a tree, and this can
lead to an exponential number of calls.
These shared subgraphs arise when the knowledge compiler employs \emph{clause caching}
to detect that the simplified set of
clauses arising from one partial assignment to the literals matches that
of a previous partial assignment~\cite{darwiche:aaai:2002}.
When this dec-DNNF node is translated into POG
node $\nodeu$, the proof generator can assume (and also check), that
there is a simplified set of clauses $\gamma_{\nodeu}$
for which the subgraph with root $\nodeu$ is its POG representation.

The proof generator can exploit the sharing of subgraphs
by constructing and proving a \emph{lemma} for each node
$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ and the defining clauses for the POG must yield
$\assign(u) = 1$.  This lemma is then invoked for every node having
$\nodeu$ as a child.
As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.

The challenge for implementing this strategy is to find a way to
represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CPOG file.  Some may be
unaltered input clauses, and these can be used directly.  Others,
however will be clauses that do not appear in the input formula.  We
implement these by adding POG product nodes to the CPOG file to create
the appropriate clauses.  Consider an \emph{argument} clause
$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
define a product node $\nodev$ with arguments
$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, 
we will introduce a defining clause
$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
  synthetic} clause having $\obar{v}$ as the \emph{guard literal}.
That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
  activate} the clause, causing it to represent argument clause $C$.  On the other
hand, a partial assignment with $\passign(v) = 1$ will
cause the clause to become a tautology and therefore have no effect.

Suppose for every clause $C_j \in \gamma_{\nodeu}$ that does not correspond to
an input clause, we generate a synthetic clause $C'_j$ with guard literal
$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma'_{\nodeu}$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
while input clauses in $\gamma_{\nodeu}$ are left unchanged.
Let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
Invoking $\validate(\nodeu, \lassign, \gamma'_{\nodeu})$
 will then prove a lemma, given by the target clause
 $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
 showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u) = 1$.

Later, when node $\nodeu$ is encountered by a call to $\validate(\nodeu, \passign, \psi)$, we invoke the lemma
by showing that each synthetic clause
$C_j$ matches some simplified clause in $\simplify{\psi}{\passign}$.  More precisely,
for $1 \leq j \leq m$,
we use clause addition to assert the clause
$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$,
showing that synthetic clause $C_j$ will be activated.
Combining the lemma with these activations provides a derivation of the target clause for the call to $\validate$.

Observe that the lemma structure can be hierarchical, since a shared
subgraph may contain nodes that are themselves roots of shared
subgraphs.  Nonetheless, the principles described allow the
definition, proof, and applications of a lemma for each shared node in
the graph.  For any node $\nodeu$, the first call to
$\validate(\nodeu, \passign, \psi)$ may require further recursion,
but any subsequent call can simply reuse the lemma proved by the first call.

\subsection{Lemma Example}
\label{app:lemma:eg}

\begin{figure}
(A) Additional nodes\\[1.0em]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
\midrule
25 & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
28 & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
\bottomrule
\end{tabular}
\\[1.0em]
(B) Implicit Clauses\\[1.2em]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
\midrule
\texttt{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
\texttt{26} & \texttt{-11 -3} & \texttt{0} & \\
\texttt{27} & \texttt{-11 4} & \texttt{0} & \\
\midrule
\texttt{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
\texttt{29} & \texttt{-12 3} & \texttt{0} & \\
\texttt{30} & \texttt{-12 -4} & \texttt{0} & \\
\bottomrule
\end{tabular}
\\[1.0em]
(C) CPOG Assertions\\[1.0em]
\begin{tabular}{llllll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
\midrule
\multicolumn{6}{l}{Lemma Proof} \\
\texttt{31} & \texttt{a 5 11 12 3} & \texttt{0} & \texttt{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
\texttt{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \texttt{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
\texttt{33} & \texttt{a 3 7 11 12} & \texttt{0} & \texttt{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
\texttt{34} & \texttt{a 7 11 12} & \texttt{0} & \texttt{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
\midrule
\multicolumn{6}{l}{Lemma Application \#1} \\
\texttt{35} & \texttt{a -11 1} & \texttt{0} & \texttt{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
\texttt{36} & \texttt{a -12 1} & \texttt{0} & \texttt{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
\texttt{37} & \texttt{a 7 1} & \texttt{0} & \texttt{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
\midrule
\texttt{38} & \texttt{a 8 1} & \texttt{0} & \texttt{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
\midrule
\multicolumn{6}{l}{Lemma Application \#2} \\
\texttt{39} & \texttt{a -11 -1} & \texttt{0} & \texttt{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
\texttt{40} & \texttt{a -12 -1} & \texttt{0} & \texttt{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
\texttt{41} & \texttt{a 7 -1} & \texttt{0} & \texttt{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
\midrule
\texttt{42} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
\texttt{43} & \texttt{a 1 10} & \texttt{0} & \texttt{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
\texttt{44} & \texttt{a 10} & \texttt{0} & \texttt{43 24 42} & \texttt{0} & $s_{10}$ \\
\bottomrule
\end{tabular}
\caption{Example of lemma definition, proof, and application}
\label{fig:eg4:lemmas}
\end{figure}

Figure~\ref{fig:eg4:lemmas} shows an alternate forward implication
proof for the example of Figure~\ref{fig:eg4:all} using a lemma to
represent the shared node $\nodes_7$.  We can see that the POG with
this node as root encodes the Boolean formula $x_3 \leftrightarrow x_4$, having a CNF representation consisting of the clauses
$\{x_3 ,\, \obar{x}_4\}$ and $\{\obar{x}_3 ,\, {x}_4\}$.  The product node
declarations shown in Figure~\ref{fig:eg4:lemmas}(A) create synthetic
clauses 25 and 28 to encode these arguments with activating literals
$\obar{v}_{11}$ and $\obar{v}_{12}$, respectively.  Clauses 31--34
then provide a proof of the lemma, stating that any assignment
$\assign$ that activates these clauses will  assign $1$ to $s_7$.
Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
will cause the first synthetic clause to activate due to input clause
3, and it will cause the second synthetic clause to activate due to
input clause 4.  From this, clause 37 can use the lemma to state that
assigning $0$ to $x_1$ will cause $s_7$ to evaluate to $1$.  Similarly,
clauses 39 and 40 serve to activate the synthetic clauses when
$\assign(x_1) = 1$, due to input clauses 1 and 2, and clause 41 then
uses the lemma to state that assigning $1$ to $x_1$ will cause $s_7$ to
evaluate to $1$.

In this example, adding the lemma increases the proof length, but that
is only because it is such a simple formula.

\input{leanappendix}

\section{Detailed Experimental Results}
\label{app:experiments}

Our experiments are designed to evaluate the following:
\begin{itemize}
\item
  whether our toolchain can handle challenging benchmark problems,
\item
the effectiveness of some design choices and optimizations,
\item
the comparative performance of the prototype and verified checkers, and
\item
how the performance of our toolchain compares to that of the MICE model counter verifier.
\end{itemize}

\subsection{Experimental Setup}

As a test machine, we used a 2021~MacBook~Pro laptop having a 3.2~Ghz
Apple M1 processor and with 64~GB of physical memory.  We also used an
external 3~TB solid-state disk drive with an advertised read/write
speed of one gigabit per second.  Despite being a laptop, this
configuration is somewhat more powerful than the nodes typically
found in server clusters.  The fast access to storage is especially
important due to the very large files generated and processed by the
tools.  Our tools generated individual files with over 160 gigabytes
of data.

For benchmarks, we downloaded 100 files each from the 2022 standard and weighted model counting competitions:
\begin{center}
       \url{https://mccompetition.org/2022/mc_description.html}
\end{center}
We found that 20 of these were duplicates across the two tracks,
yielding a total of 180 unique benchmark problems, ranging in size
from 250 to 2,753,207 clauses.

Our default configuration for the proof generator used the structural
method for the forward implication proof, with the two optimizations
of literal grouping and lemmas.
Running with a time limit of 4000 seconds\footnote{All measured times
in this document are actual elapsed times, not CPU times.}, \dfour{}
was able to generate dec-DNNF representations for 124 of these, and it timed out on the
rest.  Our proof generator was able to convert all of the generated
dec-DNNF graphs into POGs and determine how many defining clauses they
would generate.  A POG operation with $k$ arguments requires $k+1$
defining clauses, and so the total number of defining clauses in POG $P$ equals
the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$.
The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.

\subsection{Performance of the Proof Generator and Checker}

\input{figure-runtime-d4-verify}
\input{figure-clauses-d4-verify}

For each of the 124 dec-DNNF graphs generated by \dfour{}, we ran
our proof generator and prototype checker, limiting proof
generation to 10,000 seconds.  We ran the programs to generate and check
one-sided proofs for the graphs, again with a time limit of 10,000
seconds for proof generation.

Figure~\ref{fig:d4:cpog} summarizes our results in terms of the time
required by \dfour{} (X axis) versus the sum of the times for the
proof generator and checker (Y axis).
We were able to complete a full validation for 108 of the 124
benchmarks, with times ranging from fractions of a second to 13,144
seconds, with a median of 71.6 seconds.

Relative to the runtime for \dfour{}, two problems ran faster with the
generator and checker.  These were for problems having small numbers
of models (relative to the number of variables), and so most of the
time spent by both \dfour{} and our proof generator was in running a
CDCL solver to search the very sparse solution space.  \cadical{}
generally outperforms the miniSAT-based solver used by \dfour{}.  At
the other extreme, one problem that required only 19 seconds for
\dfour{} required 8657 seconds to generate a proof and 45 seconds to
check it.  This benchmark appears to stem from weak performance
by our implementation of BCP\@.  Overall the ratios
between the combined generation plus checking times versus the time
for \dfour{} had a harmonic mean of 5.5.

Of the 16 benchmarks that could not be fully validated, one had so
many defining clauses that it overflowed the 32-bit signed integers
our programs use for clause identifiers.  Another exited due to the
virtual memory limit imposed by the operating system, and the other 14
timed out.  Of the 15 that did not overflow the clause counter, we
were able to complete a one-sided verification for 9, but the other 6
timed out.  Overall, we were able to provide some level of
verification for 94\% of the benchmark problems.

Figure~\ref{fig:defining:total} compares the number of defining
clauses (X axis) with the total number of clauses (Y axis) for the 108
problems that were fully verified.  These totals include the defining
clauses for the POG, the additional defining clauses introduced to
support literal grouping and lemmas, and the clauses added by RUP
steps in the forward implication proof.  These totals ranged from 554
to 770,382,773, with a median of 1,719,245.  The ratios of the total
clauses versus defining clauses ranged from 1.54 to 6073, with a
harmonic mean of 3.13.  The high numbers were for problems having very
few models, and so the proof clauses must encode large proofs of
unsatisfiability.

\subsection{Monolithic Forward Implication Proofs}
\label{app:experiment:monolithic}

To assess the relative merits of the two approaches to forward
implication proof generation, we ran the proof generator in monolithic
mode on the 92 problems for which the combination of structural proof generation
and checking required at most 3000 seconds and then kept only those
measurements for which at least one of the two approaches had combined
times (generation plus checking) of at most 1000 seconds.  There were
a total of 83 such problems.  Figures~\ref{fig:monolithic:time} and
\ref{fig:monolithic:clauses} show comparisons between the two
approaches in terms of time and total clauses.

Examining Figure~\ref{fig:monolithic:time}, we can see that the
monolithic approach ran faster than the structural approach for 23 of
the problems, including a case where the generator and checker
required only one second, versus 71 seconds for the structural
approach.  On the other hand, the monolithic approach timed out for 13
of the cases and was slower for the other 37.  In general, we can see
the monolithic approach faring worse on larger problems.

Figure~\ref{fig:monolithic:clauses} shows the total number of clauses
for the two approaches for the 70 cases where both approaches completed within 1000 seconds.
As can be seen, the monolithic approach consistently produces
shorter proofs, perhaps due to the benefits of the proof trimming by
\dtrim{}.

\input{figure-runtime-monolithic}
\input{figure-clauses-monolithic}

These experiments suggest that a hybrid of the two approaches might
yield the best results in terms of consistency, runtime, and proof
size.  It would begin the recursive descent of $\validate$ according
to a structural approach, but monolithically generate
the proof once it encounters a sufficiently small subgraph.

\subsection{Impact of Optimizations}
\label{app:experiment:optimize}

\input{figure-clauses-optimizations}

To assess the impact of the two optimizations: literal grouping and
lemmas, we ran the proof generator on the 80 problems for which the
combination of the two optimizations yielded proofs totaling at most
10~million clauses.  Compared to the problems used to assess monolithic
proof generation (Section~\ref{app:experiment:monolithic}), this set included one that had a short proof but
long runtime and excluded four that had short runtimes but long
proofs.  We then ran the proof generator on all three combinations of
the optimizations being partially or totally disabled.
Figure~\ref{fig:optimized:lessoptimized} shows how disabling these optimizations affected the
total number of clauses in the proofs, where the X axis indicates the
total number with both optimizations enabled, while the Y axis
indicates the number with one (left) or both (right) optimizations
disabled.

As the figures illustrated, each optimization on its own can shorten
the proofs, sometimes substantially, and the two combine to have even
greater effect.  Of the 80 benchmark problems evaluated, and comparing
with both optimizations disabled, 47 had the number of proof clauses
reduced by a factor of 2 by using lemmas, 14 by using literal merging,
and 60 by enabling both optimizations.  In the extreme cases, a lack
of lemmas caused one proof to grow by a factor of 52.5, while a lack
of literal grouping causes another proof to grow by a factor of 39.6.

The time performance of these optimizations is less dramatic, but
still significant.  Lemmas sometimes made proof generation slower, but
in one case ran $38\times$ faster.  Similarly, literal grouping could
take longer, but in one case ran $3.9\times$ faster.

Overall, it is clear that these two optimizations are worthwhile, and
sometimes critical, to success for some benchmarks.

\subsection{Certifying Preprocessed Formulas}

\input{figure-runtime-preprocess}
\input{figure-clauses-preprocess}

The \dfour{} knowledge compiler can optionally perform three different
forms of preprocessing on the input formula.  These transform the
formula while preserving logical equivalence~\cite{lagniez:aaai:2014}.
Proving that the combination of preprocessing plus knowledge
compilation preserved logical equivalence would be a valuable
capability.  However, proof generators that operate based on the
structure of the input clauses, including our structural approach,
will generally fail in this case, since the compilation was not based
on the original input clauses.

One strength of monolithic proof generation is that it makes no
assumptions about how the compiled representation.  A complete
proof-generating SAT solver could, in principle, always generate the
forward implication proof, as long as forward implication holds.  To
test this hypothesis, we ran our proof framework with preprocessing
enabled on the same 83 benchmarks as in
Section~\ref{app:experiment:monolithic}.  We measured the time for two
entire toolchains: one with compilation, proof generation, and
checking, and the other starting with preprocessing and then having
the same steps as the other.  The results are summarized in
Figures~\ref{fig:preprocess:time} and \ref{fig:preprocess:clauses}.
Setting a time limit of 1000 seconds, 70 of the formulas completed
without preprocessing, while 69 did so with preprocessing.  The one
that timed out with preprocessing did so due to the excessive time
required for preprocessing.  Of the 69, 53 ran faster with
preprocessing, while 15 ran slower.  The sizes of the generated 
CPOG proofs were more variable: 31 had fewer clauses with preprocessing, 21 had more, and 17 were identical.

\subsection{Evaluation of the Verified Checker}
\label{app:lean:evaluation}

\input{figure-check}

We ran the \lean{} checker on the 80 benchmark problems for which the
generated CPOG file had at most 10~million total clauses.  These are the
same benchmarks used in evaluating the optimizations (Section
\ref{app:experiment:optimize}).
All of the checks completed and confirmed the outcome of the prototype checker.

Figure~\ref{fig:lean} shows the
performance of both the \lean{} checker and the prototype
checker on the Y axis, as functions of the total number of clauses on the X axis.
As would be expected, the \lean{} checker consistently runs slower
than the prototype checker.
The ratio of the runtime for \lean{} versus the runtime for
the prototype checker had a harmonic mean of $5.94$.

Importantly, however, it can be seen that both checkers show the same
overall performance trends.  The ratio of the two runtimes was less
than $8.0$ for all but three small benchmarks.  This seems like a
reasonable price to pay for a rigorous guarantee of correctness, and
we are confident that we can reduce this gap with more effort in
optimizing the verified checker.

\section{Comparison with other Certification Frameworks}
\label{app:comparison}

This section provides more insights into how our proof framework (POG)
compares with other work in certified knowledge compilation and model counting.

\subsection{Comparison with \cdfour{} Framework}

\input{figure-cd4-cpog}

Figure~\ref{fig:cd4:cpog} shows the results of a set of experiments
comparing the performance of our toolchain (CPOG) versus one based on
\cdfour{}.  It shows the timings for the 110 benchmarks for which
either the CPOG toolchain or the \cdfour{} toolchain completes in less
than 1000 seconds.  Both toolchains include compilation, in addition
to proof generation and checking.  Of these, 80 completed for both
toolchains, 28 timed out or failed with the CPOG toolchain, and 2
timed out for the \cdfour{} toolchain.  The data clearly show that the
\cdfour{} toolchain can handle larger benchmarks and generally runs
faster.  Only 7 of the 80 that completed with both toolchains ran
faster with the CPOG toolchain.  For these 80, the toolchain based on
\cdfour{} ran faster by a harmonic mean factor of $19.8\times$.

On the other hand, the \cdfour{} proof generation and checking is
tightly connected to the operation of \dfour{}, and every change to
the program could require changes to the proof steps and possibly
changes to the proof framework.  For example, we have found that
\cdfour{} does not work properly when the compiler makes use of its
internal preprocessor.

\subsection{Comparison with the MICE Framework}

To compare the performance of our toolchain versus that of
MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
generation, proof checking, and ring evaluation in the CPOG toolchain compares to the time to
run their proof generator \progname{nnf2trace} and checker
\progname{sharptrace}.  We add the time for ring evaluation on our
side, since their checker provides a count, although only an
unweighted one.  We used the prototype versions of the checker and ring evaluator.
We ran their tools on the 92 benchmark problems for
which our toolchain requires at most 3000 seconds and retained
the 84 data points for which at least one of the toolchains completes in under 1000 seconds.

We can summarize the results as:
\begin{itemize}
\item 84 total data points
\item 76 were completed by both programs in under 1000 seconds
\item Of those 21 were faster with MICE, 55 with CPOG
\item 7 completed with CPOG under 1000 seconds but exceeded that threshold for MICE
\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CPOG
\end{itemize}

\input{figure-mice-cpog}

The detailed results are shown in
Figure~\ref{fig:mice}.  As can be seen, the comparative runtimes are
highly variable, reflecting the fact that the two toolchains differ
fundamentally in their objectives and their approaches.

One shortcoming of \progname{nnf2trace} can be seen in
the near-vertical series of points in the upper-lefthand corner.
These correlate with a similar set of points along the left in
Figure~\ref{fig:optimized:lessoptimized}.  Like the naive
implementation of $\validate$, their program recursively traverses the
graph representation of a formula, effectively expanding it into a
tree.  They lack an optimization analogous to lemmas.

We have not tried the MICE toolchain on larger benchmarks, but we
suspect it would encounter significant scaling limitations.

\bibliography{references}

\end{document}

### 2023-05-20 doc/mc-paper.tex
Fine tuning

@@
 the input clauses.  Given input formula $\phi$, their first task is to
-check whether $\phi \imply R$, i.e., that any model of $\phi$ is also
-a model of $R$.  They then base other parts of their proof on that
+check whether $\phi \imply R$, i.e., that any assignment that satisfies $\phi$ also satisfies each of the clauses in $R$.
+They then base other parts of their proof on that
 property and use a separate checker to perform a series of additional
@@
 by their checker.  We have been able to confirm this possibility with
-their compiler and checker\footnote{Downloaded May 18, 2023 from
-\url{https://github.com/crillab/d4}}.
+their compiler and checker\footnote{Downloaded May 18, 2023 as\\
+\url{https://github.com/crillab/d4/tree/333370cc1e843dd0749c1efe88516e72b5239174}.}
 
@@
 
-In essence, a POG represents a partitioned-operation
-formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
+A POG represents a partitioned-operation
+formula with a sharing of common subformulas.  Every node in the graph can be viewed as a partitioned-operation formula, and so we write
 $\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
@@
 
-\section{The CPOG POG Representation and Proof System}
+\section{The CPOG Representation and Proof System}
 \label{sect:cpog}
@@
 the \dfour{} knowledge compiler, form the toolchain illustrated in
-Figure~\ref{fig:chain}\footnote{All tools, along with a set of 
-benchmark problems are available at \url{https://github.com/rebryant/cpog}}.  The proof generator is the same in both
+Figure~\ref{fig:chain}\footnote{The source code for all tools, along with a set of 
+benchmark problems are available as part of the supplement\cite{bryant:sat:2023:supplement}}.  The proof generator is the same in both
 cases, since it need not be trusted.

### 2023-05-20 doc/mc-paper.tex
Included plot in main paper

@@
 
+\input{figure-combined-verify}
+
 We summarize the results of our experiments here.  
@@
 We ran our proof generator with a time limit of 10,000 seconds.
-It was able to generate full proofs for 108 of the problems and
+The results are shown in Figure~\ref{fig:dual:total}, with the plot of runtimes on the left and the plot of the total number of clauses on the right.
+The proof generator was able to generate full proofs for 108 of the problems and
 one-sided proofs for an additional 9 of them, leaving just 7 with no
@@
 and checker for a full proof was 13,145 seconds.
+Overall the ratio
+between the combined generation plus checking time versus the time
+for \dfour{} had a harmonic mean of 5.5.
+The ratio between the total number of clauses and the number of defining clauses had a harmonic mean of 3.13.
 It is worth noting that, to date, we have not found

### 2023-05-20 doc/mc-paper.tex
Include more experimental results

@@
 is intractable on general representations. It can become tractable when formulas are expressed in a
-special form, such as the deterministic,
-decomposable, negation normal form (det-DNNF)\@.
+special form, such as the decision-decomposable, negation normal form (dec-DNNF)\@.
 \emph{Knowledge compilation} is the process of converting a formula
@@
 
-We have developed a program that generates POG representations from det-DNNF
+We have developed a program that generates POG representations from dec-DNNF
 graphs 
@@
 transforms the formula into a structured form for which model counting
-is straightforward.  For example, the \emph{deterministic
-decomposable, negation normal form}
+is straightforward.  For example, the \emph{deterministic, decomposable, negation normal form}
 (det-DNNF) introduced by
-Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}
-represents a
+Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
+\emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
+represent a
 Boolean formula as a directed acyclic graph, with terminal nodes
@@
 input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
-generates a \emph{decision} DNNF (dec-DNNF) representation (a restricted form of det-DNNF),
+generates a dec-DNNF representation,
 and the \emph{proof
@@
     $\phi_1$: & $(x_1 \lor x_3)$ & & \\
-    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_1 \lor \obar{x}_2)$\\    
+    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_2 \lor \obar{x}_3)$\\    
   \end{tabular}
@@
 Clearly, these two formulas are not equivalent---$\phi_1$ has six
-models, while $\phi_2$ has five.  In particular, $\phi_1$ allows
+models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
 arbitrary assignments to variable $x_2$.  Critically, however the
-second clause of $\phi_2$ is RAT with respect to the first---any
+second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
 satisfying assignment to $\phi_1$ can be transformed into one that
@@
 evaluation, we used the public benchmark problems from the 2022
-standard and weighted model competitions.  We found that there were
+standard and weighted model competitions\footnote{Downloaded from \url{https://mccompetition.org/2022/mc_description.html}}.  We found that there were
 180 unique CNF files among these, ranging in size from 250 to
@@
 these into POGs, with their declarations ranging from 304 to
-2,761,457,765 (median 848,784) defining clauses.
+2,761,457,765 median 848,784) defining clauses.
 %(The
@@
 We found that the monolithic approach for generating the forward
-implication proof works well for smaller POGs, but it becomes
-unreliable and inefficient for larger ones.
-%% These experiments suggest
-%% a possible hybrid approach, stopping the recursion of the structural
-%% approach and shifting to monolithic mode once the subgraph size is below some
-%% threshold.
+implication proof works well for smaller POGs (up to one million
+defining clauses), but it becomes inefficient for larger ones.  These
+experiments suggest a possible hybrid approach, stopping the recursion
+of the structural approach and shifting to monolithic mode once the
+subgraph size is below some threshold.  By using monolithic mode, we
+were also able to perform end-to-end verification of all but one of the
+benchmarks that could be verified without preprocessing, with a total
+time limit (including preprocessing) of 1,000 seconds.
+
 We also found that our two optimizations: literal grouping and
 lemmas can provide substantial improvements in proof size and runtime.
-Running the verified proof checker in \lean{} showed the same scaling
-with respect to the proof size as did the prototype checker, but it
-required, on average, around 5.9 times longer.
-
-Finally, the runtimes for the MICE toolchain versus our CPOG toolchain
+In the extreme cases, a lack
+of lemmas caused one proof to grow by a factor of 52.5, while a lack
+of literal grouping caused another proof to grow by a factor of 39.6.
+Overall, it is clear that these two optimizations are worthwhile, and
+sometimes critical, to success for some benchmarks.
+
+Running the verified proof checker in \lean{} required, on average
+(harmonic mean), around 5.9 times longer than the prototype checker.
+Encouraging, the scaling trends were identical for the two solvers,
+indicating that the two checkers have similar asymptotic performance.
+We consider a factor of 5.9 to be acceptable for the assurance
+provided by formal verification.
+
+In comparing other proof frameworks, we found that the \cdfour{}
+toolchain ran very fast and could handle very large benchmarks.  Even
+with a total time limit of 1,000 seconds, including the time for
+knowledge compilation, the \cdfour{} toolchain completed for 106
+benchmarks, while the CPOG toolchain completed for just 82.  We found
+that the runtimes for the MICE toolchain versus our CPOG toolchain
 showed little correlation, reflecting the fact that the two solve
-different problems and use different approaches.  In general,
-our CPOG toolchain showed better scaling, in part due to its ability
-to control the recursion through lemmas.
+different problems and use different approaches.  In general, our CPOG
+toolchain showed better scaling, in part due to its ability to control
+the recursion through lemmas.  On the other hand, neither of these
+toolchains could perform end-to-end verification when the knowledge
+compilation was preceded by preprocessing.
 

### 2023-05-20 doc/mc-supplement.tex
Included plot in main paper
Include more experimental results

@@
 of lemmas caused one proof to grow by a factor of 52.5, while a lack
-of literal grouping causes another proof to grow by a factor of 39.6.
+of literal grouping caused another proof to grow by a factor of 39.6.
 

### 2023-05-20 doc/mc-paper.tex
Tweaks

@@
 nonterminal node labeled by a Boolean And or Or operation.  Restrictions
-are placed on the structure of the graph such that a count of the
+are placed on the structure of the graph (described in Section~\ref{sect:pog}) such that a count of the
 models can be computed by a single bottom-up traversal.
@@
 generally accepted
-way to validate that their generated results are correct, i.e., that
+way to validate that 
 the compiled representation is logically equivalent to the original
@@
 users of SAT solvers to fully trust their results.  Experience has also
-shown that proof production capabilites allow SAT solver developers to quickly
+shown that being able to generate proofs allow SAT solver developers to quickly
 detect and diagnose bugs in their programs. This, in turn, has led
@@
 includes the conflict clauses that arose via a top-down processing of
-the input clauses.  Given input formula $\phi$, their first task is to
-check whether $\phi \imply R$, i.e., that any assignment that satisfies $\phi$ also satisfies each of the clauses in $R$.
+the input clauses.  Given input formula $\phi_I$, their first task is to
+check whether $\phi_I \imply R$, i.e., that any assignment that satisfies $\phi_I$ also satisfies each of the clauses in $R$.
 They then base other parts of their proof on that
-property and use a separate checker to perform a series of additional
+property and use a separate program to perform a series of additional
 checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
@@
 satisfying assignment to $\phi_1$ can be transformed into one that
-also satisfies $\phi_2$ by setting $x_2 = 0$.
+also satisfies $\phi_2$ by setting $x_2$ to 1.
 
@@
 clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
-commandline argument to \dtrim{}\footnote{Available at
-\url{https://github.com/marijnheule/drat-trim/tree/develop}.} that enforces this
-restriction.  This weakness, however, illustrates the general challenge of
+commandline argument to \dtrim{} that enforces this
+restriction.\footnote{Available at
+\url{https://github.com/marijnheule/drat-trim/tree/develop}.}  This weakness, however, illustrates the general challenge of
 developing a new proof framework.

### 2023-05-21 doc/mc-paper.tex
Small reductions

@@
 
-We present \emph{Partitioned-Operation Graphs} (POGs), a general form that can
+We present \emph{Partitioned-Operation Graphs} (POGs), a form that can
 encode all
@@
 This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
-a general form that can encode all of the representations produced by current knowledge
+a form that can encode all of the representations produced by current knowledge
 compilers. The CPOG (for ``certified'' POG) file format then
@@
 evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
-Running these two programs on a particular CPOG file gives strong
+Running these two programs on a  CPOG file gives strong
 assurance that the proof and the model count are correct. Our
@@
 and extensive experimental results.  The archive also includes copies
-of the programs and a set of benchmark files for testing.
+of the programs and a set of benchmark files.
 
@@
 the check with \dtrim{} would pass, as would the other tests performed
-by their checker.  We have been able to confirm this possibility with
+by their checker.  We have confirmed this possibility with
 their compiler and checker\footnote{Downloaded May 18, 2023 as\\

### 2023-05-21 doc/mc-paper.tex
Minor edits

@@
 %checker.)
-We ran our proof generator with a time limit of 10,000 seconds.
-The results are shown in Figure~\ref{fig:dual:total}, with the plot of runtimes on the left and the plot of the total number of clauses on the right.
+
+We ran our proof generator with a time limit of 10,000 seconds.  The
+results are shown in Figure~\ref{fig:dual:total}, The left-hand plot
+shows the elapsed time for the combination of proof generation,
+checking, and counting versus the time for
+\dfour{}.  The right-hand side shows the total number of clauses in the CPOG file versus the number of defining clauses.
 The proof generator was able to generate full proofs for 108 of the problems and
@@
 The ratio between the total number of clauses and the number of defining clauses had a harmonic mean of 3.13.
-It is worth noting that, to date, we have not found
+To date, we have not found
 any errors in the dec-DNNF graphs generated by \dfour{}.
@@
 (harmonic mean), around 5.9 times longer than the prototype checker.
-Encouraging, the scaling trends were identical for the two solvers,
+Encouragingly, the scaling trends were identical for the two solvers,
 indicating that the two checkers have similar asymptotic performance.

### 2023-05-21 doc/lean.tex
Split appendix into supplement
minor adjustments to Lean section

@@
 In this section, we briefly describe the functionality we implemented and what we proved about it.
-More information is provided in \cref{appendix:lean}.
+More information is provided in the supplement~\cite{bryant:sat:2023:supplement}.
 
@@
 \textbf{Proof checking.} The goal of a CPOG proof is to construct a POG that is equivalent to the input CNF $\inputformula$.
-The checker begins by parsing the input formula, initializing the set of active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one.
+The POG being constructed, and the set of active clauses are stored in the checker state \texttt{st}.
+The checker begins by parsing the input formula, initializing the active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one.
 It then processes every step of the CPOG proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
 Afterwards, it carries out the \textsc{final conditions} check of \cref{subsection:semantics}.
-Throughout the process, we maintain invariants which ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root.
+Throughout the process, we maintain invariants which ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\cref{thm:lean:equiv}).
 
 The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \cref{section:logical:foundations}.
-Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm Var} of propositional formulas.
+Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas.
+\lstinline{PropForm} is generic over the type of variables, so we instantiate it with our type as \lstinline{PropForm Var}.
 Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}.
@@
 
-The invariants refer to the checker state \lstinline{st} with fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far.
+The invariants refer to the checker state \lstinline{st} with fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the clausal POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far.
 For any $\nodeu\in P$, \lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.
@@
 \end{lstlisting}
-The next one guarantees that the set of solutions over the original variables is preserved:
+The next invariant guarantees that the set of solutions over the original variables is preserved:
 \begin{lstlisting}
@@
 
-invariants.partitioned : ∀ u : ILit, (st.pog.toPropForm u).partitioned
-invariants.equivalent_lits : ∀ u : ILit, equivalentOver st.inputCnf.vars
+invariants.partitioned : ∀ (u : ILit), (st.pog.toPropForm u).partitioned
+invariants.equivalent_lits : ∀ (u : ILit), equivalentOver st.inputCnf.vars
     (u ∧ st.pogDefsForm) (st.pog.toPropForm x)
 \end{lstlisting}
-These are maintained by valid CPOG proofs. Together with a few additional invariants
+These invariants are maintained by valid CPOG proofs. Together with additional invariants
 that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
-node $\noder$: the \textsc{final conditions} ensure that $\theta$ is logically equivalent to
-$\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$
-(\lstinline{equivalent st.clauseDb (r ∧ st.pogDefsForm)}), and from the invariants we can prove
-$\inputformula \iff \phi_\noder$ (\lstinline{equivalent st.inputCnf (st.pog.toPropForm r)}).
-After certifying a proof, the checker can pass its in-memory POG representation to the ring evaluator.
-% JA: I cut a few words here to save a line.
+node $\noder$:
+
+\begin{theorem}
+\label{thm:lean:equiv}
+If the proof checker has assembled POG $P$ with root node $\noder$ starting from input formula $\inputformula$, and \textsc{final conditions} hold of the checker state, then $\inputformula$ is logically equivalent to $\phi_\noder$.
+\end{theorem}
+\begin{proof}
+\textsc{Final conditions} imply that the active clausal formula $\theta$ is exactly $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.
+The conclusion follows from this and the checker invariants.
+The full proof is formally verified in Lean.
+\end{proof}
+After certifying a CPOG proof, the checker can pass its in-memory POG representation to the ring evaluator. We can conclude that this is partitioned directly from \texttt{invariants.partitioned}.
 
@@
 corresponds to the standard model count. To do so, we defined a function that
-% JA: deleted the word "directly" to save a line
 carries out an integer calculation of the number of models over a set of variables

### 2023-05-22 doc/lean.tex
Added cross reference

@@
 \label{thm:lean:equiv}
-If the proof checker has assembled POG $P$ with root node $\noder$ starting from input formula $\inputformula$, and \textsc{final conditions} hold of the checker state, then $\inputformula$ is logically equivalent to $\phi_\noder$.
+If the proof checker has assembled POG $P$ with root node $\noder$ starting from input formula $\inputformula$, and \textsc{final conditions} (as stated in Section~\ref{subsection:semantics}) hold of the checker state, then $\inputformula$ is logically equivalent to $\phi_\noder$.
 \end{theorem}

### 2023-05-22 doc/mc-paper.tex
fix double reference
Reworded comparision to det-DNNF and dec-DNNF

@@
 generated and efficiently checked
-\cite{RAT,wetzler14_drattrim}.
+\cite{heule:cade:2013,wetzler14_drattrim}.
 A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
@@
 
-POGs generalize dec-DNNF graphs in two ways:
+We can now define and compare the two related representations:
 \begin{itemize}
-\item They allow negation of arbitrary nodes in the graph, not just
-  variables.  
-  %, and so we allow it in the interest of generality.
-%%  Negation could be useful in some contexts, such as when
-%%  converting binary decision diagrams (BDDs) having complemented
-%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
-\item As with det-DNNF~\cite{darwiche:aaai:2002,darwiche:jair:2002}, they allow arbitrary arguments to a $\por$ operation, as long as
-  these have disjoint models.  By contrast, each
-  sum node $\nodeu$ in a dec-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a \emph{decision variable} $x$ such that
+\item A det-DNNF graph is a POG with negation applied only to the variables.
+\item A dec-DNNF graph is det-DNNF graph with the further restriction that any
+  sum node $\nodeu$ has have exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
   any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
-%%Our generalization is provided to allow an
-%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
-%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
 \end{itemize}
@@
 is the core rule supported by standard
-proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
+proof checkers~\cite{heule:cade:2013,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
 way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
@@
 use either its own RUP proof generator, or it can invoke
-\dtrim{}~\cite{RAT}.  The latter yields shorter proofs and scales well
+\dtrim{}~\cite{heule:cade:2013}.  The latter yields shorter proofs and scales well
 to large proofs, but each invocation has a high startup cost.  We

### 2023-05-22 doc/mc-paper.tex
Final edits for REB

@@
 performance of the verified checker with one designed for high
-performance and capacity.  We also show that our tools can provide
-end-to-end verification of formulas that have been transformed by an
-equivalence-preserving preprocessor.
+performance and capacity.
+
+We also show that our tools can provide end-to-end verification of
+formulas that have been transformed by an equivalence-preserving
+preprocessor.  That is, verification is based on the original formula,
+and so proof checking certifies correct operation of the preprocessor,
+the knowledge compiler, and the proof generator.
 
@@
 
-Capelli et al.\ developed a knowledge compiler
+Capelli, Lagniez, and Marquis developed a knowledge compiler
 that generates a certificate in a proof system that is itself based on
@@
 by their checker.  We have confirmed this possibility with
-their compiler and checker\footnote{Downloaded May 18, 2023 as\\
+their compiler and checker.\footnote{Downloaded May 18, 2023 as\\
 \url{https://github.com/crillab/d4/tree/333370cc1e843dd0749c1efe88516e72b5239174}.}
@@
 restriction.\footnote{Available at
-\url{https://github.com/marijnheule/drat-trim/tree/develop}.}  This weakness, however, illustrates the general challenge of
+\url{https://github.com/marijnheule/drat-trim/releases/tag/v05.22.2023}.}  This weakness, however, illustrates the general challenge of
 developing a new proof framework.
@@
 concepts of clausal proof systems.  These factors were important in enabling formal verification.
-In the supplement~\cite{bryant:sat:2023:supplement},
-we also compare the performance of our toolchain to these two others.
+In Section~\ref{sect:experimental} and the supplement~\cite{bryant:sat:2023:supplement},
+we also compare the performance of our toolchain to these other two.
 
@@
   formula $\phi$ has an associated \emph{dependency set}
-  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
+  $\dependencyset(\phi)  \subseteq \varset$, and a set of \emph{models} $\modelset(\phi)$,
   consisting of total assignments that satisfy the formula:
@@
 
-  We highlight two special classes of Boolean formulas.  A formula is
+  We highlight some special classes of Boolean formulas.  A formula is
   in \emph{negation normal form} when negation is applied only to variables.  A
@@
 $\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
-if they are logically equivalent.
-%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
+if they are logically equivalent,
+and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
 If they are not equivalent, then
@@
 Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
-%%\begin{eqnarray}
 \begin{displaymath}
 \begin{array}{rcl}
-\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \\ %\label{eqn:ring:negation} \\
+\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \\ 
 \textstyle
-\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \\ %\label{eqn:ring:product} \\
+\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \\ 
 \textstyle
-\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \\ %\label{eqn:ring:sum}
+\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \\ 
 \end{array}
@@
 
-We define the \emph{size} of POG $P$, written $|P|$, to be the
-the number of nonterminal nodes plus the number of edges from these nodes to their children.  Ring
-evaluation of $P$ can be performed with at most $|P|$ ring
-operations by traversing the graph from the terminal nodes up to
-the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
-The final result is then $\rep(\noder, w)$.
-
-We can now define and compare the two related representations:
+We can now define and compare two related representations:
 \begin{itemize}
-\item A det-DNNF graph is a POG with negation applied only to the variables.
-\item A dec-DNNF graph is det-DNNF graph with the further restriction that any
-  sum node $\nodeu$ has have exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
+\item A det-DNNF graph can be viewed a POG with negation applied only to variables.
+\item A dec-DNNF graph is a det-DNNF graph with the further restriction that any
+  sum node $\nodeu$ has exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
   any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
@@
 efficiently perform ring evaluation.
+
+We define the \emph{size} of POG $P$, written $|P|$, to be the
+the number of nonterminal nodes plus the number of edges from these nodes to their children.  Ring
+evaluation of $P$ can be performed with at most $|P|$ ring
+operations by traversing the graph from the terminal nodes up to
+the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
+The final result is then $\rep(\noder, w)$.
 
@@
   encoding the product or sum operation.
-%  This is the only means for adding extension variables to the proof.
+  This is the only means for adding extension variables to the proof.
 \item Boolean negation is supported implicitly by allowing the
@@
 the unit literal $v$.  A reference to literal $-v$ then provides a way
-of representing constant $0$.
+to represent constant $0$.
 
@@
 extension variable $u$ associated with any POG node $\nodeu$, we can therefore
-prove that any total assignment $\assign$ to the input variables which
+prove that any total assignment $\assign$ to the input variables that
 also satisfies the POG defining clauses will
@@
 shows the CPOG description for
-an input formula with five clauses, yielding a POG with five
+an input formula with five clauses, yielding a POG with six
 nonterminal nodes.  It explains how the clause addition and deletion
@@
 Except in degenerate cases,
-%where the formula is unsatisfiable or a tautology,
+where the formula is unsatisfiable or a tautology,
 we can therefore assume
@@
 applied to variables, and so the only edges with negative polarity will have variables as children.
-We can therefore view
+We can therefore 
 view the POG as consisting
@@
 that this approach can be very effective and generates short proofs
-for smaller problems, but it does not scale well enough for routine
+for smaller problems, but it does not scale well enough for general
 use.
@@
 (see Table~\ref{tab:defining}(B))
-  to generate the target clause for $\nodeu$.
+  to generate the target clause for $\nodeu$, requiring at most two RUP steps.
 \item If $\nodeu$ is a product node, then we can divide its children
@@
 \end{enumerate}
-  We then generate the target clause for node $\nodeu$,
+  We then generate the target clause for node $\nodeu$ with a single RUP step,
 creating the hint by combining the results from the BCP and SAT calls for
@@
 \end{enumerate}
-Most of these steps involve a polynomial number of
+Observe that 
+all of these steps involve a polynomial number of
 operations per recursive call, with the exception of those that call
@@
 the \dfour{} knowledge compiler, form the toolchain illustrated in
-Figure~\ref{fig:chain}\footnote{The source code for all tools, along with a set of 
-benchmark problems are available as part of the supplement\cite{bryant:sat:2023:supplement}}.  The proof generator is the same in both
+Figure~\ref{fig:chain}.\footnote{The source code for all tools
+is available as part of the supplement\cite{bryant:sat:2023:supplement}}  The proof generator is the same in both
 cases, since it need not be trusted.
@@
 evaluation, we used the public benchmark problems from the 2022
-standard and weighted model competitions\footnote{Downloaded from \url{https://mccompetition.org/2022/mc_description.html}}.  We found that there were
+standard and weighted model competitions.\footnote{Downloaded from \url{https://mccompetition.org/2022/mc_description.html}}  We found that there were
 180 unique CNF files among these, ranging in size from 250 to
 2,753,207 clauses.
-%%We ran our programs on a processor with 64~GB of
-%%memory and having an attached high-speed, solid-state
-%%disk.
+We ran our programs on a processor with 64~GB of
+memory and having an attached high-speed, solid-state
+disk.
 With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
@@
 these into POGs, with their declarations ranging from 304 to
-2,761,457,765 median 848,784) defining clauses.
+2,761,457,765 (median 848,784) defining clauses.
 %(The
@@
 checking, and counting versus the time for
-\dfour{}.  The right-hand side shows the total number of clauses in the CPOG file versus the number of defining clauses.
+\dfour{}.  
 The proof generator was able to generate full proofs for 108 of the problems and
@@
 verification.  The prototype checker successfully verified all of the generated
-proofs.  The longest runtime for the combination of proof generator
-and checker for a full proof was 13,145 seconds.
+proofs.  The longest runtime for the combination of proof generation, checking, and counting
+for a full proof was 13,145 seconds.
 Overall the ratio
-between the combined generation plus checking time versus the time
+between the combined time for generation, checking, and counting versus the time
 for \dfour{} had a harmonic mean of 5.5.
+The right-hand plot shows the total number of clauses in the CPOG file versus the number of defining clauses for the problems having full proofs.
 The ratio between the total number of clauses and the number of defining clauses had a harmonic mean of 3.13.
@@
 with a total time limit of 1,000 seconds, including the time for
-knowledge compilation, the \cdfour{} toolchain completed for 106
-benchmarks, while the CPOG toolchain completed for just 82.  We found
+knowledge compilation, the \cdfour{} toolchain completed 106
+benchmarks, while the CPOG toolchain completed just 82.  We found
 that the runtimes for the MICE toolchain versus our CPOG toolchain

### 2023-05-22 doc/mc-supplement.tex
Minor edit
Final edits for REB

@@
 assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
-$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
-via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
+$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading,
+via clause 22, to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
 completing the RUP proof.  The deletion hints for the other input
@@
 \item
+whether our toolchain can perform end-to-end verification with preprocessed formulas,
+\item
 the comparative performance of the prototype and verified checkers, and
 \item
-how the performance of our toolchain compares to that of the MICE model counter verifier.
+how the performance of our toolchain compares to that of related verification toolchains.
 \end{itemize}
@@
 of literal grouping and lemmas.
-Running with a time limit of 4000 seconds\footnote{All measured times
-in this document are actual elapsed times, not CPU times.}, \dfour{}
+Running with a time limit of 4000 seconds,\footnote{All measured times
+in this document are actual elapsed times, not CPU times.} \dfour{}
 was able to generate dec-DNNF representations for 124 of these, and it timed out on the
@@
 Figure~\ref{fig:d4:cpog} summarizes our results in terms of the time
-required by \dfour{} (X axis) versus the sum of the times for the
-proof generator and checker (Y axis).
+required by \dfour{} (X axis) versus the sum of the times for proof generation, checking, and counting
+ (Y axis).
 We were able to complete a full validation for 108 of the 124
@@
 Relative to the runtime for \dfour{}, two problems ran faster with the
-generator and checker.  These were for problems having small numbers
+generator, checker, and counter.  These were for problems having small numbers
 of models (relative to the number of variables), and so most of the
@@
 \dfour{} required 8657 seconds to generate a proof and 45 seconds to
-check it.  This benchmark appears to stem from weak performance
+check it (and count).  This benchmark appears to stem from weak performance
 by our implementation of BCP\@.  Overall the ratios
-between the combined generation plus checking times versus the time
+between the combined time for generation, checking, and counting versus the time
 for \dfour{} had a harmonic mean of 5.5.
@@
 implication proof generation, we ran the proof generator in monolithic
-mode on the 92 problems for which the combination of structural proof generation
-and checking required at most 3000 seconds and then kept only those
+mode on the 92 problems for which the combination of structural proof generation, checking, and counting
+required at most 3000 seconds and then kept only those
 measurements for which at least one of the two approaches had combined
-times (generation plus checking) of at most 1000 seconds.  There were
+times of at most 1000 seconds.  There were
 a total of 83 such problems.  Figures~\ref{fig:monolithic:time} and
@@
 monolithic approach ran faster than the structural approach for 23 of
-the problems, including a case where the generator and checker
+the problems, including a case where the generator, checker, and counter
 required only one second, versus 71 seconds for the structural
@@
 with both optimizations disabled, 47 had the number of proof clauses
-reduced by a factor of 2 by using lemmas, 14 by using literal merging,
+reduced by at least a factor of 2 by using lemmas, 14 by using literal merging,
 and 60 by enabling both optimizations.  In the extreme cases, a lack
@@
 One strength of monolithic proof generation is that it makes no
-assumptions about how the compiled representation.  A complete
+assumptions about how the compiled representation was generated.  A complete
 proof-generating SAT solver could, in principle, always generate the
@@
 Section~\ref{app:experiment:monolithic}.  We measured the time for two
-entire toolchains: one with compilation, proof generation, and
-checking, and the other starting with preprocessing and then having
+entire toolchains: one with compilation, proof generation, 
+checking, and counting, and the other starting with preprocessing and then having
 the same steps as the other.  The results are summarized in
@@
 
-This section provides more insights into how our proof framework (POG)
+This section provides more insights into how our proof framework (CPOG)
 compares with other work in certified knowledge compilation and model counting.
@@
 MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
-generation, proof checking, and ring evaluation in the CPOG toolchain compares to the time to
+generation, proof checking, and counting in the CPOG toolchain compares to the time to
 run their proof generator \progname{nnf2trace} and checker
-\progname{sharptrace}.  We add the time for ring evaluation on our
-side, since their checker provides a count, although only an
-unweighted one.  We used the prototype versions of the checker and ring evaluator.
+\progname{sharptrace}.
 We ran their tools on the 92 benchmark problems for

### 2023-05-23 doc/mc-paper.tex
space

@@
 Figure~\ref{fig:chain}.\footnote{The source code for all tools
-is available as part of the supplement\cite{bryant:sat:2023:supplement}}  The proof generator is the same in both
+is available as part of the supplement~\cite{bryant:sat:2023:supplement}}  The proof generator is the same in both
 cases, since it need not be trusted.

### 2023-05-23 doc/lean.tex
make minor edits to paper

@@
 Afterwards, it carries out the \textsc{final conditions} check of \cref{subsection:semantics}.
-Throughout the process, we maintain invariants which ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\cref{thm:lean:equiv}).
+Throughout the process, we maintain invariants that ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\cref{thm:lean:equiv}).
 
@@
 \end{lstlisting}
+Note that in the definition of \lstinline{uniqueExt}, the arrows associate to the right,
+so the definition says that the three assumptions imply the conclusion.
 The next invariant guarantees that the set of solutions over the original variables is preserved:

### 2023-05-23 doc/mc-paper.tex
make minor edits to paper
More extensive coverage of (un)soundness

@@
 
-Capelli, Lagniez, and Marquis developed a knowledge compiler
-that generates a certificate in a proof system that is itself based on
-dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{} program,
-a modified version of \dfour{}, generates annotations to the
-compiled representation, providing information about how the compiled version relates to the input clauses.
-It also generates a file of clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
-Completing the certification involves running two different checkers on
-the annotated dec-DNNF graph and the DRAT file.
-Although the authors make informal arguments
-regarding the soundness of their frameworks, these do not provide
-strong levels of assurance.  Indeed, 
-we have identified a weakness in the their
-methodology due to an invalid assumption about the
-guarantees provided by \dtrim{}, the program it uses to check the
-DRAT file.
+Capelli, Lagniez, and Marquis developed a knowledge compiler that
+generates a certificate in a proof system that is itself based on
+dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{}
+program, a modified version of \dfour{}, generates annotations to the
+compiled representation, providing information about how the compiled
+version relates to the input clauses.  It also generates a file of
+clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
+Completing the certification involves running two different checkers
+on the annotated dec-DNNF graph and the DRAT file.  Although the
+authors make informal arguments regarding the soundness of their
+frameworks, these do not provide strong levels of assurance.  Indeed,
+we have identified a weakness in the their methodology due to an
+invalid assumption about the guarantees provided by \dtrim{}, the
+program it uses to check the DRAT file.  Moreover, this weakness is
+\emph{exploitable}: their framework can be ``spoofed'' into accepting
+an incorrect compilation.
 
@@
 models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
-arbitrary assignments to variable $x_2$.  Critically, however the
+arbitrary assignments to variable $x_2$.  Critically, however, the
 second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
 satisfying assignment to $\phi_1$ can be transformed into one that
-also satisfies $\phi_2$ by setting $x_2$ to 1.
-
-This weakness would allow a buggy (or malicious) version
-of \cdfour{} to generate a compiled representation of $\phi_2$ when
-given $\phi_1$ as its input.  By adding the second clause to $R$,
-the check with \dtrim{} would pass, as would the other tests performed
+also satisfies $\phi_2$ by setting $x_2$ to 1, while keeping the values for other variables fixed.
+
+This weakness would allow a buggy (or malicious) version of \cdfour{}
+to spoof the checking framework.  Given formula $\phi_1$ as input, it
+could produce a compiled result, including annotations, based on $\phi_2$ and also include the second clause of $\phi_2$ to $R$.
+The check with \dtrim{} would pass, as would the other tests performed
 by their checker.  We have confirmed this possibility with
@@
 clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
-commandline argument to \dtrim{} that enforces this
+command-line argument to \dtrim{} that enforces this
 restriction.\footnote{Available at
@@
 developing a new proof framework.
-Without a formal verification, there are likely to be 
+As we can attest,
+without engaging in the effort to formally verify the framework, there are likely to be
 conditions that make the framework unsound.
@@
 Again, they provide only  informal arguments
-regarding the soundness of their frameworks.
+regarding the soundness of their framework.
 
@@
 correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$
-integers representing literals of existing variables.
+integers, indicating the arguments as literals of existing variables.
 As Table~\ref{tab:defining}(A) shows,
-this declaration implicitly causes $k+1$ clauses to be added to the proof, defining extension variable $v$ to be the product of its arguments.
+this declaration implicitly causes $k+1$ clauses to be added to the proof, using a Tseitin encoding to define extension variable $v$ as the product of its arguments.
 
@@
 not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
-signed integers representing literals of existing variables.  Hint $H$
+signed integers, indicating the arguments as literals of existing variables.  Hint $H$
 consists of a
-sequence of clause IDs, all of which must be defining clauses for other POG operations.
+sequence of clause IDs, all of which must be defining clauses for other POG operations.\footnote{The restriction to defining clauses in the hint is critical to soundness.
+Allowing the hint to include the IDs of input clauses creates an exploitable weakness.  We discovered this weakness in the course of our efforts at formal verification.}
 As Table~\ref{tab:defining}(B) shows,
-this declaration implicitly causes three clauses to be added to the proof, defining extension variable $v$ to be the sum of its arguments.
+this declaration implicitly causes three clauses to be added to the proof, using a Tseitin encoding to define extension variable $v$ as be the sum of its arguments.
 The hint must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.
@@
 
-\newpage
+%%\newpage
 

### 2023-05-24 doc/mc-paper.tex
Minor tweaks

@@
 frameworks, these do not provide strong levels of assurance.  Indeed,
-we have identified a weakness in the their methodology due to an
+we have identified a weakness in their methodology due to an
 invalid assumption about the guarantees provided by \dtrim{}, the
-program it uses to check the DRAT file.  Moreover, this weakness is
+program it uses to check the DRAT file.  This weakness is
 \emph{exploitable}: their framework can be ``spoofed'' into accepting
@@
 clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
-clause $C$ to a set of clauses maintains satisfiability, but it may
-not preserve models.  As an example, consider the following formulas:
+clause $C$ to a set of clauses maintains satisfiability, but does not necessarily
+preserve models.  As an example, consider the following formulas:
 \begin{center}
@@
 As we can attest,
-without engaging in the effort to formally verify the framework, there are likely to be
+without engaging in an effort to formally verify the framework, there are likely to be
 conditions that make the framework unsound.
@@
 As Table~\ref{tab:defining}(A) shows,
-this declaration implicitly causes $k+1$ clauses to be added to the proof, using a Tseitin encoding to define extension variable $v$ as the product of its arguments.
+this declaration implicitly causes $k+1$ clauses to be added to the proof, providing a Tseitin encoding that defines extension variable $v$ as the product of its arguments.
 
@@
 As Table~\ref{tab:defining}(B) shows,
-this declaration implicitly causes three clauses to be added to the proof, using a Tseitin encoding to define extension variable $v$ as be the sum of its arguments.
+this declaration implicitly causes three clauses to be added to the proof, providing a Tseitin encoding that defines extension variable $v$ as the sum of its arguments.
 The hint must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.
@@
 toolchain showed better scaling, in part due to its ability to control
-the recursion through lemmas.  On the other hand, neither of these
+the recursion through lemmas.  Neither of these prior
 toolchains could perform end-to-end verification when the knowledge

### 2023-05-24 doc/mc-paper.tex
Fix typos

@@
 In more detail, \cdfour{} emits a sequence of clauses $R$ that
-includes the conflict clauses that arose via a top-down processing of
+includes the conflict clauses that arose during a top-down processing of
 the input clauses.  Given input formula $\phi_I$, their first task is to
@@
 clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
-clause $C$ to a set of clauses maintains satisfiability, but does not necessarily
+clause $C$ to a set of clauses maintains satisfiability, but it does not necessarily
 preserve models.  As an example, consider the following formulas:

### 2023-05-24 doc/mc-paper.tex
Consolidated paper into single file

@@
 
-\input{lean}
+%%\input{lean}
+
+\section{A Formally Verified Toolchain}
+\label{section:formally-verified-toolchain}
+
+We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
+sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \cref{fig:chain}).
+These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}.
+Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic.
+In this section, we briefly describe the functionality we implemented and what we proved about it.
+More information is provided in the supplement~\cite{bryant:sat:2023:supplement}.
+
+\vspace{1em}\noindent
+\textbf{Proof checking.} The goal of a CPOG proof is to construct a POG that is equivalent to the input CNF $\inputformula$.
+The POG being constructed, and the set of active clauses are stored in the checker state \texttt{st}.
+The checker begins by parsing the input formula, initializing the active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one.
+It then processes every step of the CPOG proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
+Afterwards, it carries out the \textsc{final conditions} check of \cref{subsection:semantics}.
+Throughout the process, we maintain invariants that ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\cref{thm:lean:equiv}).
+
+The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \cref{section:logical:foundations}.
+Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas.
+\lstinline{PropForm} is generic over the type of variables, so we instantiate it with our type as \lstinline{PropForm Var}.
+Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}.
+Requiring totality is not a limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$.
+We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.
+
+The invariants refer to the checker state \lstinline{st} with fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the clausal POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far.
+For any $\nodeu\in P$, \lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.
+The first two invariants state that assignments to original variables extend uniquely to extension variables defining the POG nodes.
+In the formalization, we split this into extension and uniqueness:
+\begin{lstlisting}
+/-- Any assignment satisfying φ₁ extends to φ₂ while preserving values on X. -/
+def extendsOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
+  ∀ (σ₁ : PropAssignment Var), σ₁ ⊨ φ₁ → ∃ σ₂, σ₁.agreeOn X σ₂ ∧ σ₂ ⊨ φ₂
+/-- Assignments satisfying φ are determined on Y by their values on X. -/
+def uniqueExt (X Y : Set Var) (φ : PropForm Var) :=
+  ∀ (σ₁ σ₂ : PropAssignment Var), σ₁ ⊨ φ → σ₂ ⊨ φ → σ₁.agreeOn X σ₂ →
+    σ₁.agreeOn Y σ₂
+
+invariants.extends_pogDefsForm : extendsOver st.inputCnf.vars ⊤ st.pogDefsForm
+invariants.uep_pogDefsForm : uniqueExt st.inputCnf.vars st.allVars st.pogDefsForm
+\end{lstlisting}
+Note that in the definition of \lstinline{uniqueExt}, the arrows associate to the right,
+so the definition says that the three assumptions imply the conclusion.
+The next invariant guarantees that the set of solutions over the original variables is preserved:
+\begin{lstlisting}
+def equivalentOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
+  extendsOver X φ₁ φ₂ ∧ extendsOver X φ₂ φ₁
+
+invariants.equivInput : equivalentOver st.inputCnf.vars st.inputCnf st.clauseDb
+\end{lstlisting}
+Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\cref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
+\begin{lstlisting}
+def partitioned : PropForm Var → Prop
+  | tr | fls | var _ => True
+  | neg φ    => φ.partitioned
+  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(τ ⊨ φ ∧ τ ⊨ ψ)
+  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ φ.vars ∩ ψ.vars = ∅
+
+invariants.partitioned : ∀ (u : ILit), (st.pog.toPropForm u).partitioned
+invariants.equivalent_lits : ∀ (u : ILit), equivalentOver st.inputCnf.vars
+    (u ∧ st.pogDefsForm) (st.pog.toPropForm x)
+\end{lstlisting}
+These invariants are maintained by valid CPOG proofs. Together with additional invariants
+that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
+node $\noder$:
+
+\begin{theorem}
+\label{thm:lean:equiv}
+If the proof checker has assembled POG $P$ with root node $\noder$ starting from input formula $\inputformula$, and \textsc{final conditions} (as stated in Section~\ref{subsection:semantics}) hold of the checker state, then $\inputformula$ is logically equivalent to $\phi_\noder$.
+\end{theorem}
+\begin{proof}
+\textsc{Final conditions} imply that the active clausal formula $\theta$ is exactly $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.
+The conclusion follows from this and the checker invariants.
+The full proof is formally verified in Lean.
+\end{proof}
+After certifying a CPOG proof, the checker can pass its in-memory POG representation to the ring evaluator. We can conclude that this is partitioned directly from \texttt{invariants.partitioned}.
+
+\vspace{1em}\noindent
+\textbf{Ring evaluation}. We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
+(\cref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
+\begin{lstlisting}
+def weightSum {R : Type} [CommRing R]
+    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
+  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
+\end{lstlisting}
+The rules for efficient ring evaluation of partitioned formulas are expressed as:
+\begin{lstlisting}
+def ringEval (weight : Var → R) : PropForm Var → R
+  | tr       => 1
+  | fls      => 0
+  | var x    => weight x
+  | neg φ    => 1 - ringEval weight φ
+  | disj φ ψ => ringEval weight φ + ringEval weight ψ
+  | conj φ ψ => ringEval weight φ * ringEval weight ψ
+\end{lstlisting}
+\Cref{prop:ring:eval} is then formalized as follows:
+\begin{lstlisting}
+theorem ringEval_eq_weightSum (weight : Var → R) {φ : PropForm Var} :
+    partitioned φ → ringEval weight φ = weightSum weight φ (vars φ)
+\end{lstlisting}
+To efficiently compute the ring evaluation of a formula represented by a POG node, we implemented
+\lstinline{Pog.ringEval} and then proved that it matches the specification above:
+\begin{lstlisting}
+theorem ringEval_eq_ringEval (pog : Pog) (weight : Var → R) (x : Var) :
+  pog.ringEval weight x = (pog.toPropForm x).ringEval weight
+\end{lstlisting}
+Applying this to the output of our verified CPOG proof checker, which we know to be partitioned
+and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes
+the correct ring evaluation of $\inputformula$.
+
+\vspace{1em}\noindent
+\textbf{Model counting.} Finally, we established that ring evaluation with the appropriate weights
+corresponds to the standard model count. To do so, we defined a function that
+carries out an integer calculation of the number of models over a set of variables
+of cardinality \lstinline{nVars}:
+\begin{lstlisting}
+def countModels (nVars : Nat) : PropForm Var → Nat
+  | tr       => 2^nVars
+  | fls      => 0
+  | var _    => 2^(nVars - 1)
+  | neg φ    => 2^nVars - countModels nVars φ
+  | disj φ ψ => countModels nVars φ + countModels nVars ψ
+  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
+\end{lstlisting}
+We then formally proved that for a partitioned formula whose variables are among a finite set
+\lstinline{s}, this computation really does count the number of models over \lstinline{s}:
+\begin{lstlisting}
+theorem countModels_eq_card_models {φ : PropForm Var} {s : Finset Var} :
+  vars φ ⊆ s → partitioned φ → countModels (card s) φ = card (models φ s)
+\end{lstlisting}
+In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
+we have that the number of models of \lstinline{φ} on its variables is
+\lstinline{countModels φ (card (vars φ))}.
 
@@
 
-\input{figure-combined-verify}
+%%\input{figure-combined-verify}
+
+\begin{figure}
+\begin{minipage}{0.49\textwidth}
+\centering{%
+\begin{tikzpicture}[scale = 0.50]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
+      legend style={at={(0.35,0.96)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.4,
+                              xmode=log,xmin=0.01,xmax=4000,
+                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
+                              ymin=0.01, ymax=29000,
+                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
+                              xlabel={D4 runtime (seconds)},
+                              %ylabel={CPOG generation and checking runtime (seconds)},
+%                              title={D4 Defining Clause Generation}
+            ]
+%    \input{data-formatted/time-d4-verify}
+
+\addplot [only marks, color=redorange, mark=*,  mark options={scale=0.8}, opacity=0.7] coordinates { (5.758,13.334) (0.017,0.022) (0.014,0.034) (0.179,1.754) (0.016,0.046) (0.012,0.012) (0.024,0.13) (0.928,18.877) (0.069,0.624) (0.166,18.378) (1.48,21.674999999999997) (3.356,116.882) (3.184,46.001000000000005) (1.676,46.439) (0.016,0.063) (0.019,0.11100000000000002) (0.472,6.808) (0.02,0.126) (5.048,182.556) (0.398,74.411) (0.685,21.603) (0.252,3.414) (30.753,2292.7690000000002) (0.643,9.902) (2.154,36.441) (1.158,4.7219999999999995) (2.436,83.753) (1.636,132.504) (1.491,59.292) (4.683,1499.441) (30.566,4136.219) (145.967,2177.9669999999996) (2.666,40.712) (0.105,1.923) (5.508,31.97) (244.952,7273.887000000001) (0.206,8.503) (606.378,13198.325) (23.234,39.160000000000004) (0.837,182.089) (49.574,52.467) (25.077,42.550000000000004) (63.691,12374.461) (4.544,147.119) (1.679,359.63) (22.322,4493.414000000001) (2.076,61.509) (70.355,1322.4) (90.915,2908.19) (123.037,6910.268) (112.248,9602.115) (178.14,4147.301) (394.71,517.646) (158.115,180.629) (69.122,315.91) (508.972,258.496) (2.033,84.352) (270.954,433.669) (1536.914,859.557) (19.013,8856.852) (2.938,272.899) (598.709,8218.087) (9.311,105.213) (0.277,2.226) (0.101,1.086) (30.577,394.01700000000005) (0.18,2.051) (0.016,0.064) (0.021,0.134) (0.52,14.541) (0.016,0.063) (0.237,11.949000000000002) (0.985,58.088) (2.278,33.982) (1.053,16.815) (1.326,20.087) (2.908,45.419) (3.441,45.275) (2.06,30.061999999999998) (1.483,108.72) (3.377,47.61) (0.017,0.073) (0.016,0.052000000000000005) (3.536,72.871) (1.337,19.372999999999998) (5.587,81.767) (193.799,9953.933) (89.858,3959.588) (0.023,0.06899999999999999) (49.796,2973.447) (0.718,76.449) (5.364,204.765) (3.21,138.828) (41.455,1955.3980000000001) (1.083,53.278999999999996) (1.695,279.8) (1.345,286.818) (3.111,411.545) (6.397,1495.882) (22.214,666.7180000000001) (65.322,110.378) (22.917,420.612) (272.642,3116.728) (1.929,85.946) (1358.165,5872.383) (3951.002,10810.879) (14.808,1124.911) (93.537,4307.451)};
+
+%    \input{data-formatted/time-d4-verify-onesided-only}
+
+\addplot [only marks, color=bluegray, mark=square*,  mark options={scale=0.8}, opacity=0.7] coordinates { (205.743,2420.6859999999997) (326.575,4433.625) (476.021,3973.2659999999996) (494.38,5879.952) (1046.603,10779.982) (351.999,2069.865) (64.21,479.92100000000005) (316.505,3074.084) (788.377,6710.351000000001)};
+
+%    \input{data-formatted/time-d4-failure}
+
+\addplot [only marks, color=darkgreen, mark=diamond*, mark options={scale=1.1}, opacity=0.7] coordinates { (2052.239,25000) (424.73,25000) (2636.562,25000) (1403.431,25000) (3070.214,25000) (2053.446,25000) (2790.169,25000)};
+
+    \legend{
+ \textsf{Full validation},
+ \textsf{One-sided validation},
+ \textsf{Resource limit exceeded},
+    }
+    \addplot[mark=none, dashed] coordinates{(0.01,25000) (4000, 25000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,20000) (4000,20000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (2000,20000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (200, 20000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (20, 20000)};
+    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
+    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
+    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
+    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
+    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\centering{%
+\begin{tikzpicture}[scale = 0.50]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10},
+                              x post scale=1.8, y post scale=2.15,
+                              xmode=log,xmin=100,xmax=1000000000, 
+                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
+                              ymode=log, ymin=100, ymax=1000000000, 
+                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
+                              xlabel={Defining Clauses},
+                              %ylabel={Proof Clauses},
+%                              title={D4 Defining Clause Generation}
+            ]
+    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
+%%    \input{data-formatted/defining+total}
+
+\addplot [only marks, color=redorange, mark=*,  mark options={scale=0.8}, opacity=0.7] coordinates { (18271,161794) (1295,2397) (1659,4427) (127284,262373) (2325,6077) (359,554) (5373,14373) (314359,1213826) (44178,81780) (493,33207) (605988,1117440) (1141369,3809317) (1197649,2211944) (1001050,2393131) (2947,7851) (4991,13207) (355617,721935) (4897,13101) (2437497,5273728) (345,46079) (701346,1593947) (3191,26378) (5985623,28392373) (467550,940819) (1318024,2813565) (4713,23892) (1682059,3160386) (1406192,4043887) (899701,1704811) (4936014,11840667) (8038928,68469623) (70651620,197339867) (9567,251890) (39501,104182) (18007,187551) (125319627,374762287) (82194,297768) (269645146,770382773) (304,427342) (574062,8448281) (887,478450) (1757,399856) (78873354,503117290) (190248,429409) (2064844,13434107) (13737124,141587875) (50722,820648) (45923418,82243431) (83111464,188103550) (40776253,104351646) (79106742,265990284) (3133905,23973053) (16147,1697847) (45214,1277037) (72258,4612352) (905,2500003) (401032,840020) (355,2155900) (63050,4692737) (3195277,6098685) (1125105,4417869) (325251378,736764776) (6074672,13104909) (132891,287958) (65522,136116) (15561156,38336288) (143191,298755) (2851,7851) (5243,13911) (334862,681444) (2659,7335) (82585,317789) (447979,1612224) (905899,1733678) (433226,802268) (551970,1006319) (1891045,3387533) (1210098,2232426) (755273,1440924) (470934,1914084) (1178453,2188886) (3131,8575) (2515,6607) (1903735,4312279) (1107098,1868005) (1801878,3324418) (47556240,206213289) (24087642,83491057) (3109,8227) (16038150,101261480) (312728,642675) (3175308,7847036) (142059,747752) (20382514,69650789) (453769,1563728) (498246,1154507) (797866,6840229) (3835114,17412807) (5035130,10668725) (647011,9905453) (5189,1234644) (740235,2672407) (219979530,417455256) (402460,844395) (2295530,81147729) (1772352,119113610) (3975194,16763326) (8493275,41873955)};
+
+    \node[left] at (axis cs: 900,1200) {$1\times$};
+    \node[left] at (axis cs: 900,12000) {$10\times$};
+    \node[left] at (axis cs: 900,120000) {$100\times$};
+    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
+    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\end{minipage}
+\caption{Runtime (left) and proof size (right) for CPOG proofs.  The runtime includes proof generation, checking, and model counting relative to the runtime for D4.
+  The proof size is measured as total clauses relative to the number of defining clauses.}
+\label{fig:dual:total}
+\end{figure}
 

### 2023-05-24 doc/mc-paper.tex
Final version?

@@
 \usepackage{pgfplots}
+\pgfplotsset{compat=1.18}
 \usepackage{booktabs}
@@
 We have developed an online supplement to this
-paper~\cite{bryant:sat:2023:supplement} containing additional
-information, including a worked example, details on the algorithms,
-and extensive experimental results.  The archive also includes copies
-of the programs and a set of benchmark files.
+paper~\cite{bryant:sat:2023:supplement} that includes
+a worked example, more details on the algorithms,
+and extensive experimental results.
 
@@
 to spoof the checking framework.  Given formula $\phi_1$ as input, it
-could produce a compiled result, including annotations, based on $\phi_2$ and also include the second clause of $\phi_2$ to $R$.
+could produce a compiled result, including annotations, based on $\phi_2$ and also include the second clause of $\phi_2$ in $R$.
 The check with \dtrim{} would pass, as would the other tests performed
@@
 \end{proof}
-After certifying a CPOG proof, the checker can pass its in-memory POG representation to the ring evaluator. We can conclude that this is partitioned directly from \texttt{invariants.partitioned}.
+After certifying a CPOG proof, the checker can pass its in-memory POG representation to the ring evaluator, along with the partitioning guarantee provided by \texttt{invariants.partitioned}.
 
@@
 Figure~\ref{fig:chain}.\footnote{The source code for all tools
-is available as part of the supplement~\cite{bryant:sat:2023:supplement}}  The proof generator is the same in both
+is available at \url{https://github.com/rebryant/cpog/releases/tag/v1.0.0}.}  The proof generator is the same in both
 cases, since it need not be trusted.

### 2023-06-12 doc/dd/dd.py
Allow sizing of bubbles on edges

@@ -27,7 +27,7 @@ import math
 #       Position options: above, below, left, right
 
 # Edge
-#    e FID TID value=neutral mark=none [only=XX]
+#    e FID TID value=neutral mark=none radius=10 [only=XX]
 #       FID and TID are Ids of source and destination nodes
 #       value options: neutral, high, low, path
 #       mark options: none, bubble   
@@ -349,14 +349,16 @@ class Edge(Only):
     mtype = EdgeType.none
     edgeSpacing = 12.0
     edgeFraction = 0.65
+    radius = NodeType.radius[NodeType.root]
 
-    def __init__(self, fromId=None, toId=None, etype=None):
+    def __init__(self, fromId=None, toId=None, etype=None, eradius = None):
         Only.__init__(self)
         self.fromId = fromId
         self.fromNode = None
         self.toId = toId
         self.toNode = None
         self.etype = EdgeType.neutral if etype is None else etype
+        self.radius = NodeType.radius[NodeType.root] if eradius is None else eradius
 
     def parse(self, line):
         fields = line.split()
@@ -370,7 +372,7 @@ class Edge(Only):
                 raise ParseException(line, "Invalid key=value syntax")
             prefix, value = info
             try:
-                key = parseLabel(prefix, ["type", "mark", "only"])
+                key = parseLabel(prefix, ["type", "mark", "only", "radius"])
             except NoMatchException:
                 raise ParseException(line, "Invalid prefix '%s'" % prefix)
             except MultiMatchException:
@@ -391,6 +393,11 @@ class Edge(Only):
                     raise ParseException(line, "Ambiguous mark '%s'" % value)
             elif key == "only":
                 self.setOnly(value)
+            elif key == "radius":
+                try:
+                    self.radius = int(value)
+                except:
+                    raise ParseException(line, "Invalid radius")
         return self
         
     def renderTikz(self, outfile, lowerLeft, scale=1.0, scheme=Scheme.blackWhite):
@@ -417,7 +424,7 @@ class Edge(Only):
         if self.mtype == EdgeType.bubble:
             cx = (sx+fx)/2
             cy = (sy+fy)/2
-            r = scale * NodeType.radius[NodeType.root]
+            r = scale * self.radius
             outfile.write("%s\\draw [fill=%s,draw=%s] (%.2f,%.2f) circle [radius=%.2f];\n" % (self.indent, edge, edge, cx,cy,r))
         self.onlySuffix(outfile)
 

### 2023-06-15 doc/dd/dd.py
Add arcs

@@ -27,10 +27,11 @@ import math
 #       Position options: above, below, left, right
 
 # Edge
-#    e FID TID value=neutral mark=none radius=10 [only=XX]
+#    e FID TID value=neutral mark=none radius=10 arrow=none[only=XX]
 #       FID and TID are Ids of source and destination nodes
 #       value options: neutral, high, low, path
 #       mark options: none, bubble   
+#       arrow options: none, to, from, both
 
 # Graphic primitives
 #
@@ -333,6 +334,9 @@ class EdgeType:
     names = ["neutral", "high", "low", "path"]
     none, bubble = list(range(2))
     markNames = ["none", "bubble"]
+    arrowNone, arrowFrom, arrowTo, arrowBidirectional  = list(range(4))
+    arrowNames = ["none", "from", "to", "bidirectional"]
+    arrowFormat = ["", "[latex-]", "[-latex]", "[-latex-]"]
 
     def parse(self, prefix):
         return getIndex(prefix, self.names)
@@ -340,25 +344,33 @@ class EdgeType:
     def parseMark(self, prefix):
         return getIndex(prefix, self.markNames)
 
+    def parseArrow(self, prefix):
+        return getIndex(prefix, self.arrowNames)
+
 class Edge(Only):
     fromId = None
     fromNode = None
     toId = None
     toNode = None
-    etype = None
+    etype = EdgeType.neutral
     mtype = EdgeType.none
     edgeSpacing = 12.0
     edgeFraction = 0.65
     radius = NodeType.radius[NodeType.root]
+    arrow = EdgeType.arrowNone
 
-    def __init__(self, fromId=None, toId=None, etype=None, eradius = None):
+    def __init__(self, fromId=None, toId=None, etype=None, eradius = None, earrow = None):
         Only.__init__(self)
         self.fromId = fromId
         self.fromNode = None
         self.toId = toId
         self.toNode = None
-        self.etype = EdgeType.neutral if etype is None else etype
-        self.radius = NodeType.radius[NodeType.root] if eradius is None else eradius
+        if etype is not None:
+            self.etype = etype
+        if eradius is not None:
+            self.radius = eradius
+        if earrow is not None:
+            self.arrow = earrow
 
     def parse(self, line):
         fields = line.split()
@@ -372,7 +384,7 @@ class Edge(Only):
                 raise ParseException(line, "Invalid key=value syntax")
[…]

### 2023-06-20 cpog/generator/clausal.cpp
Streamlined presentation, based on Zoom call

@@ -1156,6 +1156,159 @@ bool Cnf_reasoner::enable_pog(Pog_writer *pw) {
     return true;
 }
 
+// BCP support.  Perform unit propagation
+// For given clause, have four possibilities:
+// 1. The clause is satisfied.  Return false
+// 2. The clause has >= 2 unassigned literals.  Set up watch pointers.  Return false
+// 3. The clause has 1 unassigned literal.  Add this to unit_literals.  Return false
+// 4. The clause has a conflict.  Return true
+
+// The behavior is a bit different if this is the initial setup (first_pass = true) vs. regular BCP:
+// In the first pass, ALL unassigned literals are moved to the beginning, and both watch pointers are assigned
+// In other passes, the loop can exit when it's found two unassigned literals, and only the second watch pointer need be assigned
+bool Cnf_reasoner::bcp_unit_propagate(int cid, bool first_pass) {
+    Clause *cp  = get_clause(cid);
+    int unassigned_count = 0;
+    for (int i = 0; i < cp->length; i++) {
+	int clit = (*cp)[idx];
+	if (unit_literals.find(clit) != unit_literals.end())
+	    // Clause satisfied
+	    return false;
+	else if (unit_literals.find(-clit) == unit.literals.end()) {
+	    // Literal unassigned.  Swap into place for next unassigned literal
+	    (*cp)[idx] = (*cp)[unassigned_count];
+	    (*cp)[unassigned_count++] = clit;
+	    if (!first_pass && unassigned_count >= 2)
+		break;
+	}
+    }
+    if (unsassigned_count == 0)
+	// Conflict
+	return true;
+    else if (unsassigned_count == 1)
+	// Unit literal
+	unit_queue.push_back((*cp)[0]);
+    else {
+	// Set up watch pointers
+	if (first_pass)
+	    // Need to set up both watch pointers
+	    watches.insert({(*cp)[0], cid});
+	watches.insert({(*cp)[1], cid});
+    }
+    return false;
+}
+
+// Perform Boolean constraint propagation
+// Return ID of any generated conflict clause (or 0)
+// Update set of active clauses
+int Cnf_reasoner::bcp(bool bounded) {
+    bool conflict = false;
+    int unit_count = 0; 
+    int ncid = 0;
+    int pcount = 0;
+
+
+    watches.clear();
+    unit_queue.clear();
[…]

### 2023-07-07 cpog/generator/clausal.cpp
Enable different solvers for monolithic generation

@@ -2912,6 +2912,10 @@ int Cnf_reasoner::monolithic_validate_root(int root_literal) {
     double start = tod();
 #if SOLVER == CADICAL
     snprintf(cmd, 350, "cadical --no-binary --unsat -q %s - | drat-trim %s -L %s > /dev/null", cnf_name, cnf_name, lrat_name);
+#elif SOLVER == LCADICAL
+    snprintf(cmd, 350, "cadical --no-binary --unsat -q --lrat=1 %s %s", cnf_name, lrat_name);
+#elif SOLVER == TCADICAL
+    snprintf(cmd, 350, "cadical --no-binary --unsat -q --lrat=1 %s - | lrat-trim --no-binary -q - %s", cnf_name, lrat_name);
 #else
     snprintf(cmd, 350, "kissat --no-binary --unsat -q %s - | drat-trim %s -L %s > /dev/null", cnf_name, cnf_name, lrat_name);
 #endif

### 2023-07-25 doc/JAIR/jair-data.tex
Start of JAIR paper

--- 2023-05-18 doc/mc-data.tex
@@
-\documentclass[letterpaper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-%This is a template for producing LIPIcs articles. 
-%See lipics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
+\documentclass[twoside,11pt]{article}
+\usepackage{jair, theapa, rawfonts}
 
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+\jairheading{1}{2024}{1--16}{10/22}{03/24}
 
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+\ShortHeadings{Certified Knowledge Compilation}{Bryant, Nawrocki, Avigad, \& Heule}
 
@@
 
-\bibliographystyle{plainurl}% the mandatory bibstyle
+\bibliographystyle{theapa}
 
-\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting}
+\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting \\ Experimental Details }
 
-\titlerunning{Certified Knowledge Compilation}
-
-\author{Randal E.
-Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{rebryant@cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}
-\author{Wojciech Nawrocki}{Department of Philosophy, Carnegie Mellon University}{wjnawrock@cmu.edu}{https://orcid.org/0000-0002-8839-0618}{}
-\author{Jeremy Avigad}{Department of Philosophy, Carnegie Mellon University}{avigad@cmu.edu}{https://orcid.org/0000-0003-1275-315X}{}
-\author{Marijn J. H. Heule}{Computer Science Department, Carnegie Mellon University}{marijn@cmu.edu}{https://orcid.org/0000-0002-5587-8801}{Supported by NSF grant CCF-2108521}
-
-\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-%\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-
-%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
-
-%\category{} %optional, e.g. invited paper
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-%\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\EventEditors{John Q. Open and Joan R. Access}
-%\EventNoEds{2}
-%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-%\EventShortTitle{CVIT 2016}
-%\EventAcronym{CVIT}
-%\EventYear{2016}
-%\EventDate{December 24--27, 2016}
-%\EventLocation{Little Whinging, United Kingdom}
-%\EventLogo{}
-%\SeriesVolume{42}
-%\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\author{\name Randal E. Bryant \email rebryant@cmu.edu \\
+  \name Wojciech Nawrocki \email wjnawrock@cmu.edu \\
+  \name Jeremy Avigad \email avigad@cmu.edu \\
+  \name Marijn J. H. Heule \email marijn@cmu.edu \\
+  \addr Carnegie Mellon University \\ Pittsburgh, PA  USA}
 
@@
 
+\begin{abstract}
+  This document gives details on the experiments
+\end{abstract}
+
 \section*{Benchmark machine}
@@
      \begin{itemize}
-    \item  20 instances were duplicated in the two tracks
-     \item Reduces set of benchmarks to 180
+    \item  20 instances were duplicated in the two tracks.  
+     \item Reduced set of benchmarks to 180 by including some from Track 1 and some from Track 2
      \end{itemize}
@@
      \begin{itemize}
-     \item Completed for 124 instances
+     \item Completed for 123 instances
      \end{itemize}
@@
      \item Used heuristics to identify root node, and then put in topological order with root node at end.
+     \item Successfully generated POG representation for 122 benchmarks
+       \begin{itemize}
+         \item One failed with too many defining clauses (2,761,457,765).  That count exceeded the 32-bit signed integer used to represent clause identifiers.
+       \end{itemize}
+
      \item Measure complexity by number of defining clauses $D$
@@
        \item $D$ also equals sum of number of edges and number of nodes
-       \item For 124 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
+       \item For 122 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
+       \end{itemize}
+
+     \item Also measure \emph{tree ratio}: $D_t / D$.
+       \begin{itemize}
+       \item $D_t$ Number of defining clauses if graph were expanded to have no shared subgraphs (i.e., a tree)
+       \item $D$ Number of defining clauses for the POG representation
+       \item Provides an indication of the importance of exploiting shared subgraphs in the POG generation
+       \end{itemize}
+
+     \item Performance (122 instances):
+       \begin{itemize}
+       \item Times: Min = 0.012 secs, Max = 2804.5 secs, Median = 3.46 secs
+       \item Defining clauses: Min = 304, Max = 1,765,743,261, Median = 774,883
+       \item Tree ratio: Min = 1.0, Max = 52,409.7, Median = 11.61, Harmonic Mean = 3.21
+       \item Tree Sizes ($D_t$): Min = 304, max = 14,924,670,455,484, median = 7,669,582.
        \end{itemize}
      \end{itemize}
-   \item Ran program {\sc d2p} to generate CPOG file for maximum of
-     10,000 seconds.  Verifies file, generates unweighted count. If
-     finds literal weights, also generates weighted count
+\end{enumerate}
+
+\section*{Selecting Reduced Benchmark Set}
+
+To make the experiments tractable, want to select set of benchmarks that are more tractable.
+Limit set to the 92 with at most $10^7$ defining clauses.
+
+D4 Performance (92 instances):
        \begin{itemize}
-     \item Completed for 108 instances
-     \item Exceeded clause limit for 1 instance (Our programs represent a clause ID as a 32-bit signed value)
-     \item Ran out of memory for 1 instance
-     \item Timed out for 14 instances
-     \item Was able to do one-sided verification for 9 of these.
+       \item Times: Min = 0.012 secs, Max = 1672.9 secs, Median = 1.85 secs
+       \item Defining clauses: Min = 304, Max = 8,493,275, Median = 378,325
+       \item Tree ratio: Min = 1.0, Max = 363.9, Median = 9.37, Harmonic Mean = 2.87
+       \item Tree Sizes ($D_t$): Min = 304, max = 1,205,743,897, median = 2,690,776
        \end{itemize}
-   \item For completed CPOG files, ran {\sc cpog-check} CPOG checker
-     \begin{itemize}
-     \item Tests all conditions for correct CPOG
-     \item Most time spent performing RUP checks for clause addition and deletion
-     \item Completed successfully for all 108 instances
-     \item Time generally faster than {\sc d2p}. (min = 0.006, max = 1.38, harmonic mean = 0.107)
-     \end{itemize}
-\end{enumerate}
+
+\section*{CPOG Generation}
+
+We have devised two different approaches to generating the forward implication proof from the dec-DNNF generated by {\sc D4}:
+\begin{description}
+\item[Monolithic:] Make a single call to a proof generating SAT solver.
+\item[Structured:] Traverse the POG top-down and recursively generate a forward implication proof for each node encountered.
+Lemmas can be introduced to avoid an exponential expansion of the POG into a tree.
+\end{description}
+
+Our implementation also allows a hybrid of these approaches: at any
+point in the traversal, it can generate a proof for a node via a
+single call to a SAT solver.  This capability supports both full
+monolithic generation (perform monolithic generation at the POG root),
+full structured generation (continue traversing to the POG leaves), or
+a balance between the two.
+The challenge is then to devise a
+heuristic criterion for deciding whether to continue the traversal below
+some node or to generate a proof for that node monolithically.
+
+Our first set of experiments applies full monolithic and full structured generation to the reduced benchmark set.  We set a time limit of 1000 seconds for the proof generation.
+
+\end{document}
 
@@
 
-\input{figure-runtime-cd4}
+\input{figure-cd4-cpog}
 
@@
 \item Remaining 4 exceeded 1000 seconds for both toolchains
+\end{itemize}
 
-\end{itemize}
+Found that CD4 cannot handle files that are first preprocessed with D4's built-in preprocessors.
 

### 2023-07-25 doc/JAIR/jair-paper.tex
Start of JAIR paper

--- 2023-05-24 doc/mc-paper.tex
@@
-\documentclass[letterpaper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-%This is a template for producing LIPIcs articles.
-%See lipics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
-
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+\documentclass[twoside,11pt]{article}
+\usepackage{jair, theapa, rawfonts}
+
+\ShortHeadings{Certified Knowledge Compilation}{Bryant, Nawrocki, Avigad, \& Heule}
+
+\jairheading{1}{2024}{1--16}{10/22}{03/24}
+
+\ShortHeadings{Certified Knowledge Compilation}{Bryant, Nawrocki, Avigad, \& Heule}
 
@@
 \usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage{tikz}
@@
 \usepackage{booktabs}
-\usepackage{hyperref}
+%%\usepackage{hyperref}
+\newcommand{\url}[1]{\texttt{#1}}
 
@@
 
-\bibliographystyle{plainurl}% the mandatory bibstyle
+\newtheorem{dfn}{Definition}
+\newtheorem{prop}{Proposition}
+\newtheorem{thm}{Theorem}
 
@@
 
-\titlerunning{Certified Knowledge Compilation}
-
-\author{Randal E.
-Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{Randy.Bryant@cs.cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}
-\author{Wojciech Nawrocki}{Department of Philosophy, Carnegie Mellon University}{wjnawrocki@cmu.edu}{https://orcid.org/0000-0002-8839-0618}{Hoskinson Center for Formal Mathematics}
-\author{Jeremy Avigad}{Department of Philosophy, Carnegie Mellon University}{avigad@cmu.edu}{https://orcid.org/0000-0003-1275-315X}{Hoskinson Center for Formal Mathematics}
-\author{Marijn J. H. Heule}{Computer Science Department, Carnegie Mellon University}{marijn@cmu.edu}{https://orcid.org/0000-0002-5587-8801}{Supported by NSF grant CCF-2108521}
-
-\authorrunning{R. E. Bryant et al.} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Randal E.
-Bryant, Wojciech Nawrocki, Jeremy Avigad, and Marijn J.
-H.
-Heule} %TODO mandatory, please use full first names.
-LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-\begin{CCSXML}
-<ccs2012>
-   <concept>
-       <concept_id>10003752.10003790.10003794</concept_id>
-       <concept_desc>Theory of computation~Automated reasoning</concept_desc>
-       <concept_significance>500</concept_significance>
-       </concept>
- </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Theory of computation~Automated reasoning}
-
-%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
-\keywords{Propositional model counting, Proof checking}
-
-%\category{} %optional, e.g. invited paper
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-\nolinenumbers %uncomment to disable line numbering
-
-\EventEditors{Meena Mahajan and Friedrich Slivovsky}
-\EventNoEds{2}
-\EventLongTitle{26th International Conference on Theory and Applications of Satisfiability Testing (SAT 2023)}
-\EventShortTitle{SAT 2023}
-\EventAcronym{SAT}
-\EventYear{2023}
-\EventDate{July 4--8, 2023}
-\EventLocation{Alghero, Italy}
-\EventLogo{}
-\SeriesVolume{271}
-\ArticleNo{6}
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\EventEditors{John Q. Open and Joan R. Access}
-%\EventNoEds{2}
-%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-%\EventShortTitle{CVIT 2016}
-%\EventAcronym{CVIT}
-%\EventYear{2016}
-%\EventDate{December 24--27, 2016}
-%\EventLocation{Little Whinging, United Kingdom}
-%\EventLogo{}
-%\SeriesVolume{42}
-%\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\author{\name Randal E. Bryant \email rebryant@cmu.edu \\
+  \name Wojciech Nawrocki \email wjnawrock@cmu.edu \\
+  \name Jeremy Avigad \email avigad@cmu.edu \\
+  \name Marijn J. H. Heule \email marijn@cmu.edu \\
+  \addr Carnegie Mellon University \\ Pittsburgh, PA  USA}
 
@@
 
-\begin{definition}[Boolean Formulas]
+\begin{dfn}[Boolean Formulas]
   The set of Boolean formulas is defined recursively.  Each
@@
 \label{def:boolean}
-\end{definition}
+\end{dfn}
 
@@
 
-\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
+\begin{dfn}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
   A \emph{partitioned-operation formula}
@@
       \end{enumerate}
-\end{definition}
+\end{dfn}
      We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.
@@
 
-\begin{definition}[Commutative Ring]
+\begin{dfn}[Commutative Ring]
   A \emph{commutative ring} $\ring$ is an algebraic structure
@@
 \label{def:ring}
-\end{definition}
+\end{dfn}
 We write $a - b$ as a shorthand for $a + -b$.
 
-\begin{definition}[Ring Evaluation Problem]
+\begin{dfn}[Ring Evaluation Problem]
 \label{def:ring_evaluation}
@@
 \label{def:weight}
-\end{definition}
+\end{dfn}
 
@@
 Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
-\begin{proposition}
+\begin{prop}
 \label{prop:ring:eval}
@@
 \end{displaymath}
-\end{proposition}
-As is described in \cref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.
+\end{prop}
+As is described in \ref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.
 
@@
 We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
-sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \cref{fig:chain}).
+sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \ref{fig:chain}).
 These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}.
@@
 It then processes every step of the CPOG proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
-Afterwards, it carries out the \textsc{final conditions} check of \cref{subsection:semantics}.
-Throughout the process, we maintain invariants that ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\cref{thm:lean:equiv}).
-
-The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \cref{section:logical:foundations}.
+Afterwards, it carries out the \textsc{final conditions} check of \ref{subsection:semantics}.
+Throughout the process, we maintain invariants that ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\ref{thm:lean:equiv}).
+
+The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \ref{section:logical:foundations}.
 Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas.
@@
 \end{lstlisting}
-Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\cref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
+Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\ref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
 \begin{lstlisting}
@@
 
-\begin{theorem}
+\begin{thm}
 \label{thm:lean:equiv}
 If the proof checker has assembled POG $P$ with root node $\noder$ starting from input formula $\inputformula$, and \textsc{final conditions} (as stated in Section~\ref{subsection:semantics}) hold of the checker state, then $\inputformula$ is logically equivalent to $\phi_\noder$.
-\end{theorem}
+\end{thm}
 \begin{proof}
@@
 \textbf{Ring evaluation}. We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
-(\cref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
+(\ref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
 \begin{lstlisting}
@@
 \end{lstlisting}
-\Cref{prop:ring:eval} is then formalized as follows:
+\ref{prop:ring:eval} is then formalized as follows:
 \begin{lstlisting}
@@
 evaluation, we used the public benchmark problems from the 2022
-standard and weighted model competitions.\footnote{Downloaded from \url{https://mccompetition.org/2022/mc_description.html}}  We found that there were
+standard and weighted model competitions.\footnote{Downloaded from \url{https://mccompetition.org/2022/mc\_description.html}}  We found that there were
 180 unique CNF files among these, ranging in size from 250 to
@@
 \bibliography{references}
+\bibliographystyle{theapa}
 

### 2023-07-31 doc/pkc-notes.tex
Write up of PKC algorithm

\documentclass[letterpaper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\nolinenumbers %uncomment to disable line numbering

\usepackage{amsmath}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{hyperref}

\newcommand{\pand}{\mathbin{\land^{\sf p}}}
\newcommand{\por}{\mathbin{\lor^{\sf p}}}
\DeclareMathOperator*{\Pand}{\bigwedge^{\sf p}}
\DeclareMathOperator*{\Por}{\bigvee^{\sf p}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

\newcommand{\dvarset}{D}
\newcommand{\varset}{X}
\newcommand{\pvarset}{P}
\newcommand{\dependencyset}{{\cal V}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\drep}{\textbf{R}_D}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
%\newcommand{\drational}{\mathbb{Q}_{2,5}}
\newcommand{\drational}{\textbf{Q}_{2,5}}

\newcommand{\assign}{\rho}
\newcommand{\uassign}{{\cal U}}
\newcommand{\modelset}{{\cal M}}
\newcommand{\dassign}{\alpha}
\newcommand{\passign}{\beta}
\newcommand{\acombine}{\!\cdot\!}
\newcommand{\dmodelset}{{\cal M_{D}}}
\newcommand{\pure}{\mathit{Pure}}

\newcommand{\simplify}[2]{#1|_{#2}}
\newcommand{\ureduce}[2]{\forall #2\, #1}
\newcommand{\partition}[2]{\mathit{Part}(#1, #2)}

\newcommand{\makenode}[1]{\mathbf{#1}}
\newcommand{\nodeu}{\makenode{u}}
\newcommand{\nodev}{\makenode{v}}
\newcommand{\nodes}{\makenode{s}}
\newcommand{\nodep}{\makenode{p}}
\newcommand{\noder}{\makenode{r}}

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}
\newcommand{\lean}{Lean~4}

\newcommand{\algo}[1]{\textsf{#1}}
\newcommand{\kc}{\algo{KC}}
\newcommand{\pkc}{\algo{PKC}}
\newcommand{\traverse}{\algo{Traverse}}
\newcommand{\posop}{\algo{Pos}}
\newcommand{\negop}{\algo{Neg}}
\newcommand{\bcp}{\algo{BCP}}
\newcommand{\uprop}{\algo{Uprop}}
\newcommand{\simpl}{\also{Simpl}}

\definecolor{redorange}{rgb}{0.878431, 0.235294, 0.192157}
\definecolor{lightblue}{rgb}{0.552941, 0.72549, 0.792157}
\definecolor{clearyellow}{rgb}{0.964706, 0.745098, 0}
\definecolor{clearorange}{rgb}{0.917647, 0.462745, 0}
\definecolor{mildgray}{rgb}{0.54902, 0.509804, 0.47451}
\definecolor{softblue}{rgb}{0.643137, 0.858824, 0.909804}
\definecolor{bluegray}{rgb}{0.141176, 0.313725, 0.603922}
\definecolor{lightgreen}{rgb}{0.709804, 0.741176, 0}
\definecolor{darkgreen}{rgb}{0.152941, 0.576471, 0.172549}
\definecolor{redpurple}{rgb}{0.835294, 0, 0.196078}
\definecolor{midblue}{rgb}{0, 0.592157, 0.662745}
\definecolor{clearpurple}{rgb}{0.67451, 0.0784314, 0.352941}
\definecolor{browngreen}{rgb}{0.333333, 0.313725, 0.145098}
\definecolor{darkestpurple}{rgb}{0.396078, 0.113725, 0.196078}
\definecolor{greypurple}{rgb}{0.294118, 0.219608, 0.298039}
\definecolor{darkturquoise}{rgb}{0, 0.239216, 0.298039}
\definecolor{darkbrown}{rgb}{0.305882, 0.211765, 0.160784}
\definecolor{midgreen}{rgb}{0.560784, 0.6, 0.243137}
\definecolor{darkred}{rgb}{0.576471, 0.152941, 0.172549}
\definecolor{darkpurple}{rgb}{0.313725, 0.027451, 0.470588}
\definecolor{darkestblue}{rgb}{0, 0.156863, 0.333333}
\definecolor{lightpurple}{rgb}{0.776471, 0.690196, 0.737255}
\definecolor{softgreen}{rgb}{0.733333, 0.772549, 0.572549}
\definecolor{offwhite}{rgb}{0.839216, 0.823529, 0.768627}
\definecolor{medgreen}{rgb}{0.15, 0.6, 0.15}

% Lean code:
\usepackage{listings}
%\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
\definecolor{keywordcolor}{rgb}{0.0, 0.1, 0.6}   % blue
\definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{commentcolor}{rgb}{0.4, 0.4, 0.4}   % grey
\definecolor{symbolcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{sortcolor}{rgb}{0.1, 0.5, 0.1}      % green
\definecolor{attributecolor}{rgb}{0.7, 0.1, 0.1} % red
\def\lstlanguagefiles{lstlean.tex}
% set default language
\lstset{language=lean, xleftmargin=1em}
\lstset{backgroundcolor=\color{white}}

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Notes on \\ Projected Knowledge Compilation \\ \today }

\titlerunning{Projected Knowledge Compilation}

\author{Randal E. Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{Randy.Bryant@cs.cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}

\authorrunning{R. E. Bryant} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

%\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10003794</concept_id>
       <concept_desc>Theory of computation~Automated reasoning</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Automated reasoning}

%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
\keywords{knowledge compilation, model counting, projected model counting}

%\category{} %optional, e.g. invited paper

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\EventEditors{John Q. Open and Joan R. Access}
%\EventNoEds{2}
%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
%\EventShortTitle{CVIT 2016}
%\EventAcronym{CVIT}
%\EventYear{2016}
%\EventDate{December 24--27, 2016}
%\EventLocation{Little Whinging, United Kingdom}
%\EventLogo{}
%\SeriesVolume{42}
%\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\section{Logical Foundations}

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an \emph{assignment} of truth values to some subset of the
  variables, where $0$ denotes false and $1$ denotes true, i.e.,
  $\assign \colon Y \rightarrow \{0,1\}$ for some $Y  \subseteq \varset$.
We say the assignment is \emph{total with respect to a set of variables} $Y$, when it assigns a value to every $x \in Y$, and that it is simply \emph{total} when it is total with respect to $\varset$.
  The set of all possible total assignments with respect to a set of variables $Y$ is denoted $\uassign(Y)$.

For each variable $x \in \varset$,
  we define the \emph{literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, with
   $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\theta$ has an associated \emph{dependency set}
  $\dependencyset(\theta)  \subseteq \varset$, and a set of \emph{models} $\modelset(\theta)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Constant formulas $\nil$ (logical falsehood) and $\tautology$ (tautology) are Boolean formulas,
    with $\dependencyset(\nil) = \dependencyset(\tautology) = \emptyset$, with $\modelset(\nil) = \emptyset$, and with $\modelset(\tautology) = \uassign(\varset)$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign(\varset) \mid  \assign(x)=1\}$.
  \item For formula $\theta$, its \emph{negation}, written $\boolnot \theta$ is a Boolean formula,
    with $\dependencyset(\boolnot \theta) = \dependencyset(\theta)$ and $\modelset(\boolnot \theta) = \uassign(\varset) - \modelset(\theta)$.
  \item For formulas $\theta_1, \theta_2, \ldots, \theta_k$, their \emph{product} $\theta = \bigwedge_{1 \leq i \leq k} \theta_i$ is a Boolean formula, with
      $\dependencyset(\theta) = \bigcup_{1 \leq i \leq k} \dependencyset(\theta_i)$ and
      $\modelset(\theta) = \bigcap_{1 \leq i \leq k} \modelset(\theta_i)$.
  \item For formulas $\theta_1, \theta_2, \ldots, \theta_k$, their \emph{sum} $\theta = \bigvee_{1 \leq i \leq k} \theta_i$ is a Boolean formula, with
      $\dependencyset(\theta) = \bigcup_{1 \leq i \leq k} \dependencyset(\theta_i)$ and
      $\modelset(\theta) = \bigcup_{1 \leq i \leq k} \modelset(\theta_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

\subsection{Conjunctive Normal Form (CNF)}

  Many tools for reasoning about Boolean formulas operate on formulas in \emph{conjunctive normal form}, defined as follows.
  Formula $\theta$ is in  \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.

  A CNF
  formula can be represented as a set of \emph{clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.  We the symbol $\phi$ to denote a CNF formula.

\subsection{Partitioned-Operation Formulas}

Computing many important properties of Boolean formulas, including
whether they are satisfiable or how many models that have, is
intractable for general Boolean formulas, as well as for those in
conjunctive normal form.  By imposing sufficient restrictions, on the
other hand, these properties can be computed in polynomial time (with
respect to the formula size).

The set of partitioned-operation operation formulas (POFs) is defined to be those Boolean formulas $\psi$ obeying the following recursively defined restrictions.
\begin{description}
\item[Constants:] 
$\psi = \tautology$ and $\psi = \nil$.
 \item[Literals:]
   $\psi = x$ and $\psi = \obar{x}$ for $x \in \varset$
 \item[Product operations:] $\psi = P(\psi_1, \psi_2, \ldots, \psi_k)$ where each $\psi_i$ is a POF\@.
   This product represents the Boolean formula
   $\psi = \bigwedge_{1 \leq i \leq k} \psi_i$.
 The operation arguments must be defined over disjoint sets of variables.
 That is, $\dependencyset(\psi_i) \cap \dependencyset(\psi_j) = \emptyset$ for all $1 \leq i < j \leq k$.
 \item[Sum operations:] $\psi = S(\psi_1, \psi_2, \ldots, \psi_k)$ where each $\psi_i$ is a POF\@.
   This sum represents the Boolean formula
   $\psi = \bigvee_{1 \leq i \leq k} \psi_i$.
The operation arguments must have disjoint models.
That is, $\modelset(\psi_i) \cap \modelset(\psi_j) = \emptyset$ for $1 \leq i < j \leq k$.
\item[Negations:] $\neg \psi'$, where $\psi'$ is a sum or product operation.
\end{description}

\subsection{Projected Formulas}

The set of Boolean variables $\varset$ is divided into two sets: {\em data} variables
$\dvarset$ and {\em auxiliary} variables $\pvarset$.  We use separate
assignments for the two sets of variables:
$\dassign \colon \dvarset \rightarrow \{0,1\}$
 and
 $\passign \colon \pvarset \rightarrow \{0,1\}$, and the symbol $\acombine$ to combine the two assignments:
 $\dassign\acombine\passign \colon \varset \rightarrow \{0,1\}$.
Sets $\uassign(\dvarset)$ and $\uassign(\pvarset)$ therefore denote the set of all assignments that are total with respect to $\dvarset$ and $\pvarset$, respectively.

For Boolean formula $\theta$,
its set of {\em data models}, written $\dmodelset(\theta)$ is defined as:
\begin{eqnarray}
\dmodelset(\theta) & = & \{ \dassign \in \uassign(\dvarset) \mid  \exists \passign \in \uassign(\pvarset) \; \dassign\acombine\passign \in \modelset(\theta) \} \label{eqn:dmodelset}
\end{eqnarray}

\subsection{Data Ring Evaluation}

  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
  every data variable $x \in \dvarset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.

  For Boolean formula $\theta$ and ring weight function $w$, its {\em data ring evaluation} computes
  \begin{eqnarray}
    \drep(\theta, w) & = & \sum_{\alpha \in \dmodelset(\theta)} \;\; \prod_{\lit \in \alpha} w(\lit) \label{eqn:rep}
  \end{eqnarray}
where $\sum$ and $\prod$ represent the ring sum $\radd$ and ring product $\rmul$ operations, respectively.

For a partitioned-operation formula $\psi$ such that $\dependencyset(\psi) \subseteq \dvarset$, its data ring evaluation can be computed based on the recursive structure of the formula:
\begin{itemize}
\item Constants: $\drep(\tautology) = \mulident$ and $\drep(\nil) = \addident$
\item Literals: $\drep(x) = w(x)$ and $\drep(\obar{x}) = \mulident - x$
\item Product operations: 
$\drep(P(\psi_1, \psi_2, \ldots, \psi_k)) = \prod_{1 \leq i \leq k} \drep(\psi_i)$
\item Sum Operations
$\drep(S(\psi_1, \psi_2, \ldots, \psi_k)) = \sum_{1 \leq i \leq k} \drep(\psi_i)$
\item Negations: $\drep(\neg \psi) = \mulident -\psi$
\end{itemize}

The task of \emph{projected knowledge compilation} is then defined as follows: Given clausal formula $\phi$, generate a partitioned-operation formula $\psi$ having $\dependencyset(\psi) \subseteq \dvarset$, such that
$\dmodelset(\psi) = \dmodelset(\phi)$.

\section{Compilation Algorithm}

We can build on existing knowledge compilers to implement a projected knowledge compiler.

\subsection{Operations on Clausal Representations}

Let $U(\phi)$ denote the set of \emph{unit literals} in $\phi$, i.e., the set of all literals $\lit$ such that clause $\{ \lit \} \in \phi$.  In the following $L$ denotes a set of literals.

We define the following operations for transforming clausal formula $\phi$ into a new clausal formula:
\begin{description}
\item[$\posop(\phi, L)$:] Remove tautologies, as well as clauses satisfied by the literals: $\posop(\phi, L) = \{ C \in \phi \mid  C \cap L \not = \emptyset\}$.

\item[$\negop(\phi, L)$:] Remove the negations of the literals from all clauses.  Let $\negop(C, L) = \{ \lit \in C \mid  \obar{\lit} \not \in L\}$.
  Then $\negop(\phi, L) = \{ \negop(C, L) \mid  C \in \phi \}$.  If this process generates an empty clause, then the result is $\nil$.

\item[$\simplify{\phi}{L}$:] Simplify the clauses by assigning the literals, removing all occurrences of any literal for the variables in $\dependencyset(L)$:
$\simplify{\phi}{L} = \posop(\negop(\phi, L), L)$.

\item[$\uprop(\phi)$:]  Perform unit propagation based on the unit literals of $\phi$.  This involves simplifying the clauses according to the unit literals and then adding back the unit clauses.
$\uprop(\phi) = \simplify{\phi}{U(\phi)} \cup \{ \{\lit \} \mid  \lit \in U(\phi) \}$.

\item[$\bcp(\phi)$:] Perform Boolean constraint propagation (BCP).
  Repeatedly compute $\phi \leftarrow \uprop(\phi)$ until either $\phi$ becomes $\nil$ or a fixed point is reached.

\item[$\ureduce{\phi}{x}$:] Express the universal quantification of $\phi$ with respect to variable $x$ in clausal form:
$\ureduce{\phi}{x} = \negop(\phi, \{x, \obar{x}\})$.

\item[$\partition{\phi}{Y}$:] Find the subset of clauses that depend only on the variables in $Y$:
$\partition{\phi}{Y} = \{ C \in \phi \mid \dependencyset(C) \subseteq Y\}$.

\end{description}

\subsection{Operations on POFs}

When constructing POFs via a bottom-up process, we can perform a set of simplifications, as follows:

\begin{itemize}
\item
The use of negation can be optimized as follows:
\begin{itemize}
\item Constants: $\neg \tautology = \nil$ and $\neg \nil = \tautology$
\item Literals: $\neg x = \obar{x}$ and $\neg \obar{x} = x$
\item Negations: $\neg \neg \psi = \psi$
\end{itemize}

\item 
When forming a product $P(\psi_1, \psi_2, \ldots, \psi_k)$ we can perform the following simplifications:
\begin{itemize}
\item If $\psi_i = \nil$ for any argument $\psi_i$, then the result is $\nil$
\item Eliminate any argument $\psi_i$ for which $\psi_i = \tautology$.
\item If the reduced set of arguments consists of a single formula, then this formula becomes the result.
\item Otherwise, form a product operation over the reduced set of arguments.
\end{itemize}

\item
Similarly, when forming a sum $S(\psi_1, \psi_2, \ldots, \psi_k)$ we can perform the following simplifications:
\begin{itemize}
\item If $\psi_i = \tautology$ for any argument $\psi_i$, then the result is $\tautology$
\item Eliminate any argument $\psi_i$ for which $\psi_i = \nil$.
\item If the reduced set of arguments consists of a single formula, then this formula becomes the result.
\item Otherwise, form a sum operation over the reduced set of arguments.
\end{itemize}
\end{itemize}

\subsection{The Supporting Knowledge Compiler}

We assume that we have a knowledge compiler that can convert a clausal formula $\phi$ into a logically equivalent POF $\psi$.  Furthermore, we assume that the generated POF obeys the following restrictions:
\begin{itemize}
\item The POF is in normal form, i.e., there are no negated product or sum operations.
\item The sum operations are \emph{decision operations}.  That is, each sum operation is of the form $S(\psi_1, \psi_2)$, and it has an associated decision variable $x$,
  such that 1) either $\psi_1 = x$ or $\psi_1$ is a product operation having $x$ as an argument, and 2)
either $\psi_2 = \obar{x}$ or $\psi_2$ is a product operation having $\obar{x}$ as an argument.
\end{itemize}
We let $\kc(\phi)$ denote the output of the knowledge compiler for input formula $\phi$.

\subsection{Algorithm}

Our goal is to implement a projecting knowledge compiler that produces a POF $\pkc(\phi)$ when given input formula $\phi$.
We define our algorithm in terms of a recursive function $\traverse(\psi, \phi)$ that simultaneously operates on a POF $\psi$ and a clausal formula $\phi$.  Projected knowledge compilation is then defined
as $\pkc(\phi) = \traverse(\kc(\phi), \bcp(\phi))$.

The traversal is defined based on the recursive structure of its first argument:
\begin{description}
\item[Constants:]
  $\traverse(\tautology, \phi) = \tautology$ and 
  $\traverse(\nil, \phi) = \nil$.

\item[Literals:]
  $\traverse(\lit, \phi) = \lit$ if $\dependencyset(\lit) \subseteq \dvarset$, and $\traverse(\lit, \phi) = \tautology$ if $\dependencyset(\lit) \in \pvarset$.
\item[Product Operations:]
  Let the product operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
  \begin{enumerate}
  \item Let $L = \{\lit_1, \lit_2, \ldots, \lit_k\}$.
  \item Define clausal formula $\phi'$ to be $\bcp(\simplify{\phi}{L})$
  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.
  \item Return the product (with simplifications) for the set of arguments $\{\lit \in L \mid  \dependencyset(\lit) \subseteq \dvarset\} \cup \{ \psi'_i \mid  1 \leq i \leq m \}$.
  \end{enumerate}
\item[Sum Operations:]
  Let the sum operation be of the form $S(\lit_1, \lit_2)$ and having decision variable $x$.
  \begin{enumerate}
  \item For $1 \leq i \leq 2$, recursively compute $\psi'_i = \traverse(\psi_i, \phi)$.
  \item If $x \in \dvarset$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$.
  \item Otherwise, let $\psi' = \pkc(\ureduce{\phi}{x})$
    \begin{enumerate}
    \item If $\psi' = \nil$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$
    \item Otherwise return nested formula $S(\neg S(\neg \psi'_1, \psi'), \psi'_2)$.
    \end{enumerate}
  \end{enumerate}
\end{description}

Of these steps, the most interesting ones involve the handling of sum
operations when the decision variable $x$ is to be projected.  The
recursive calls yielded subformulas $\psi'_1$ and $\psi'_2$, representing the projected compilation of $\phi$ when $x$ is assigned $1$ and $0$, respectively.
These
formulas potentially have overlapping models and so care must be taken
in forming their sum.  Formula $\ureduce{\phi}{x}$ is the clausal
representation of the universal quantification of $\phi$ with respect
to $x$, and so $\psi' = \pkc(\ureduce{\phi}{x})$ will be a POF whose
data models are the intersection of those for $\psi_1'$ and $\psi_2'$.  In
the event that $\psi' = \nil$, the two formulas are disjoint.  This
will occur, for example, when $x$ is a Tseitin variable encoding some
Boolean formula over the data variables.  In that case, the sum can be
formed directly.  If $\psi' \not = \tautology$, we want to
``subtract'' the overlapping models from one of the arguments before
forming the sum with the other.  We can do so using DeMorgan's Laws.  
That is formula $\neg S(\neg \psi'_1, \psi')$ will have as data models $\dmodelset(\psi'_1) - \dmodelset(\psi')$.  Furthermore, the arguments to this sum operation will have disjoint models
since $\dmodelset(\psi') \subseteq \dmodelset(\psi'_1)$, and therefore
$\dmodelset(\psi') \cap \dmodelset(\neg \psi'_1) = \emptyset$.
As can be seen, we could also reverse the role of the two arguments, computing the result
$S(\psi'_1, \neg S(\neg \psi'_2, \psi'))$.

\subsection{Optimizations}

We can improve the efficiency of the traversal by applying the following optimizations:
\begin{itemize}
\item Only data variables in $\psi$.  If $\dependencyset(\psi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \psi$.
\item Only data variables in $\phi$.  If $\dependencyset(\phi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \kc(\phi)$.
\item Only projection variables in $\phi$.  If $\dependencyset(\phi) \subseteq \pvarset$, then $\traverse(\psi, \phi) = \tautology$ if formula $\phi$ is satisfiable and $\nil$ otherwise.
\item Cached result. If $\psi$ is a sum operation and some previous traversal had the same clausal formula $\phi$ as its second argument, then the previously computed result can be reused.\footnote{
Caching for product operations is not, in general valid, since the two recursive calls made when processing a sum operation can be for product operations with the same clausal argument.}
\item Pure literal reduction.
Literal $\lit$ is said to be \emph{pure} in $\phi$ if some clause in $\phi$ contains $\lit$, and no clause in $\phi$ contains  $\obar{\lit}$.
Let $\pure(\phi)$ denote the set of all pure literals in $\phi$, and $\pure_{\pvarset}(\phi) = \{ \lit \in \pure(\phi) \mid  \dependencyset(\lit) \subseteq \pvarset \}$.
If $\pure_{\pvarset}(\phi) \not = \emptyset$, then $\phi$ can be reduced as
$\bcp(\simplify{\phi}{\pure_{\pvarset}(\phi)})$.
\end{itemize}

\section{Partitioned-Operation Graphs (POGs)}

To allow a sharing of subformulas, we represent partitioned-operation
formulas as a class of directed acyclic graphs, known as POGs.  These
graphs have nodes denoting product and sum operations and edges
denoting arguments, where an edge can be directed to a variable, to
$\tautology$, or to another POG node.  Each edge has a \emph{phase}
indicating whether  the argument is to be used directly
(phase = $1$) or negated (phase = $0$).  The top-level POF is represented by a \emph{root edge}.

\section{Example Formula}

Consider an encoding of the formula $(x_1 \land x_2) \lor x_3$ using
encoding variable $y$.  A full {\em Tseitin} encoding of $y
\leftrightarrow (x_1 \land x_2)$ requires three clauses, but we can
use a Plaisted-Greenbaum encoding in this context, encoding just $y
\rightarrow (x_1 \land x_2)$, represented by the clauses
$x_1 \lor \obar{y}$ and $x_2 \lor \obar{y}$.  We then encode the disjunction
with the clause $x_3 \lor y$.  The DIMACS representation of the input formula $\phi$ is then as follows (using variable 4 to represent $y$):

\begin{center}
\begin{tabular}{ll}
\toprule
\makebox[10mm]{Clause} & \\
\midrule
 \texttt{1 -4} & \texttt{0} \\
 \texttt{2 -4} & \texttt{0} \\
 \texttt{3  4} & \texttt{0}\\
\bottomrule
\end{tabular}
\end{center}

\begin{figure}
  \begin{minipage}{0.48\textwidth}
    (A) POG representation of input formula \\[1.2ex]
    \input{dd/eg-preproj}
  \end{minipage}
  \begin{minipage}{0.48\textwidth}
  \end{minipage}
  \begin{minipage}{0.48\textwidth}
    (B) After projecting variable $y$ \\[1.2ex]
    \input{dd/eg-proj}
  \end{minipage}
  \caption{POG Representation of formula (A), and its projection (B)
  A black dot on an edge denotes Boolean negation.}
  \label{fig:eg-proj}
\end{figure}

Figure \ref{fig:eg-proj}(A) shows the POG representation of the
example formula, derived from the dec-DNNF representation produced by
\dfour{}.  In this graph, product nodes are labeled with the operation $\pand$, and sum nodes are labeled with the operation $\por$, emphasizing that these operations satisfy their respective partitioning constraints.
We can see that \dfour{} used $y$ as the decision variable at the top level, creating sum node $\nodes_7$ having edges to
node $\nodep_5$ representing product operation $P(x_3, \obar{y})$ on the left and  node $\nodep_6$
representing product operation $P(x_1, x_2, y)$ on the right.

Figure \ref{fig:eg-proj}(B) shows the result of our algorithm, based on the following recursive steps.
Computing $\traverse(P(x_3, \obar{y}), \phi)$ yields the literal $x_3$, while
computing $\traverse(P(x_1, x_2, y), \phi)$ yields node $\nodep_{10}$ representing the product operation $P(x_1, x_2)$.
Computing $\ureduce{\phi}{y}$ yields the clausal formula $\phi'$ given as
\begin{center}
\begin{tabular}{ll}
\toprule
\makebox[10mm]{Clause} & \\
\midrule
\texttt{1} & \texttt{0} \\
\texttt{2} & \texttt{0} \\
 \texttt{3} & \texttt{0}\\
\bottomrule
\end{tabular}
\end{center}
and this is compiled to node $\nodep_{8}$ representing product operation $P(x_1, x_2, x_3)$.
Finally, the nodes are assembled using our DeMorgan's Law transformations as shown with root node $\nodes_{11}$.

We can see in the POG of (B) that node $\nodep_{8}$ encodes $x_1 \land x_2
\land x_3$, and this formula's only model assigns value $1$ to
variables $x_1$, $x_2$, and $x_3$.  This is the only assignment
satisfying both $x_3$ and $x_1 \land x_2$, and so our construction
subtracts this model from those of $x_3$ before forming a sum with
$x_1 \land x_2$.

\end{document}

### 2023-07-31 cpog/pkc/pkc.py
Print more info

@@ -60,6 +60,9 @@ def run(name, args):
     delta = datetime.datetime.now() - start
     seconds = delta.seconds + 1e-6 * delta.microseconds
     rtime = seconds - (cnf.satTime + cnf.d4Time)
+    print("GEN: Calls: %d  D4, %d SAT solver" % (cnf.d4Calls, cnf.satCalls))
+    print("GEN: Node traversals: %d product, %d show variable sum, %d already mutex sum, %d added exclusion sum" % (
+        project.productTraversals, project.showSumTraversals, project.mutexSumTraversals, project.excludingSumTraversals))
     print("GEN: Elapsed time for generation: %.2f seconds" % seconds)
     print("GEN:    Time for SAT:  %.2f seconds" %  cnf.satTime)
     print("GEN:    Time for D4:   %.2f seconds" %  cnf.d4Time)

### 2023-07-31 cpog/pkc/project.py

@@ -17,6 +17,13 @@ class ProjectionException(Exception):
     def __str__(self):
         return "Projection Exception: " + str(self.value)
 
+# Data gathering
+productTraversals = 0
+excludingSumTraversals = 0
+mutexSumTraversals = 0
+showSumTraversals = 0
+
+
 # Create cache of previous solutions
 # Indexed by both packed CNF representation and by previous POG
 
@@ -65,11 +72,13 @@ class Projector:
         self.pog = self.pcnf.makePog(self.verbLevel, trySimple = False)
 
     def traverseProduct(self, node, pcnf):
+        global productTraversals
         if self.verbLevel >= 3:
             print("Product traversal of node %s" % node.string())
         contextLiterals = set([])
         nchildren = []
         nodeLits = []
+        productTraversals += 1
         for child in node.children:
             if self.pog.isNode(child):
                 nodeLits.append(child)
@@ -92,6 +101,7 @@ class Projector:
         return nlit
 
     def traverseSum(self, node, pcnf):
+        global showSumTraversals, mutexSumTraversals, excludingSumTraversals
         if self.verbLevel >= 3:
             print("Sum traversal of node %s" % node.string())
         nchildren = [self.traverse(child, pcnf) for child in node.children]
@@ -99,11 +109,14 @@ class Projector:
         if dvar is None:
             raise ProjectionException("Can't find decision variable for node %s" % node.string())
         nlit = None
-        if dvar not in self.showVariables:
+        if dvar in self.showVariables:
+            showSumTraversals += 1
+        else:
             if self.verbLevel >= 3:
                 print("  Factoring decision variable %d" % dvar)
             mpcnf = pcnf.reduce(ignoredVariables = set([dvar]))
             if mpcnf.isSat():
+                excludingSumTraversals += 1
                 xpog = pcnf2ppog(mpcnf, self.showVariables, self.verbLevel)
                 if (self.verbLevel >= 3):
                     print("  Exclusionary term")
@@ -113,8 +126,10 @@ class Projector:
                 nlit = self.pog.addSum([-mlit, nchildren[1]])
                 if self.verbLevel >= 3:
                     print("   Traversing %s.  Integrated exclusionary term %d" % (node.string(), xlit))
-            elif self.verbLevel >= 3:
-                print("  Traversing %s.  Exclusionary term unsatisfiable" % (node.string()))
+            else:
+                mutexSumTraversals += 1
[…]

### 2023-07-31 tools/dup_finder.py
Create unique copies

@@ -4,10 +4,18 @@
 # Does syntactic match of clauses
 
 import sys
+import shutil
+
 import readwrite
 
 def usage(name):
-    print("Usage: %s [-v] fname1 fname2 ..." % name)
+    print("Usage: %s [-h]  [-v] [-u PATH] fname1 fname2 ..." % name)
+    print("  -h      Print this message")
+    print("  -v      Verbose")
+    print("  -u PATH  Make a copy of the first representative from each class in PATH")
+
+# Global variables
+verbose = False
 
 def match(fname1, fname2):
     try:
@@ -28,6 +36,10 @@ def match(fname1, fname2):
             if lit1 != lit2:
                 return False
     return True
+
+def stripPath(path):
+    fields = path.split('/')
+    return fields[-1]
 
 class Eclass:
     nvar = 0
@@ -56,7 +68,16 @@ class Eclass:
     def show(self):
         return "{%s}" % ", ".join(self.fnames)
 
-def build(fnames, verbose = False):
+    def uniqueCopy(self, path):
+        src = self.fnames[0]
+        dest = path + '/' + stripPath(src)
+        if len(self.fnames) == 1:
+                print("%s --> %s (Unique)" % (src, dest))
+        else:
+            print("%s --> %s (Duplicated by %s)" % (src, dest, ", ".join(self.fnames[1:])))
+        shutil.copy(src, dest)
+
+def build(fnames):
     classes = []
     for fname in fnames:
         found = False
@@ -82,19 +103,31 @@ def build(fnames, verbose = False):
     return classes
 
 def run(name, args):
+    global verbose
+    verbose = False
+    path = None
     if len(args) == 0 or args[0] == '-h':
         usage(name)
         return
     verbose = False
-    if args[0] == '-v':
-        verbose = True
-        args = args[1:]
-    classes = build(args, verbose)
+    while args[0][0] == '-':
+        if args[0] == '-v':
+            verbose = True
+            args = args[1:]
+        if args[0] == '-u':
+            path = args[1]
+            args = args[2:]
+        else:
+            print("Unknown flag '%s'" % args[0])
+            return
+    classes = build(args)
     for eclass in classes:
         if len(eclass.fnames) > 1:
-            print(" ".join(eclass.fnames))
+            print("Class: " + " ".join(eclass.fnames))
+        if path is not None:
+            eclass.uniqueCopy(path)
 
 if __name__ == "__main__":
     run(sys.argv[0], sys.argv[1:])
-            
+    sys.exit(0)
     

### 2023-08-01 doc/pkc-notes.tex
Minor corrections

@@
   $\assign \colon Y \rightarrow \{0,1\}$ for some $Y  \subseteq \varset$.
-We say the assignment is \emph{total with respect to a set of variables} $Y$, when it assigns a value to every $x \in Y$, and that it is simply \emph{total} when it is total with respect to $\varset$.
+We say the assignment is \emph{total with respect to a set of variables} $Y$, when it assigns a value to every $x \in Y$.
   The set of all possible total assignments with respect to a set of variables $Y$ is denoted $\uassign(Y)$.
@@
   $\dependencyset(\theta)  \subseteq \varset$, and a set of \emph{models} $\modelset(\theta)$,
-  consisting of total assignments that satisfy the formula:
+  consisting of assignments that are total with respect to $\varset$ and that satisfy the formula:
   \begin{enumerate}
@@
   Many tools for reasoning about Boolean formulas operate on formulas in \emph{conjunctive normal form}, defined as follows.
-  Formula $\theta$ is in  \emph{negation normal form} when negation is applied only to variables.  A
+  A formula is in  \emph{negation normal form} when negation is applied only to variables.  A
   formula is in \emph{conjunctive normal form} (CNF) when i) it is in
-  negation normal form, and ii) sum is applied only to literals.
+  negation normal form, and ii) sum is applied only to literals.  We use symbol $\phi$ to denote CNF formulas.
 
@@
   literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
-This literal must be assigned value $1$ by any satisfying assignment of the formula.  We the symbol $\phi$ to denote a CNF formula.
+This literal must be assigned value $1$ by any satisfying assignment of the formula.  We use symbol $\phi$ to denote a CNF formula.
 
@@
 \begin{description}
-\item[$\posop(\phi, L)$:] Remove tautologies, as well as clauses satisfied by the literals: $\posop(\phi, L) = \{ C \in \phi \mid  C \cap L \not = \emptyset\}$.
+\item[$\posop(\phi, L)$:] Remove tautologies, as well as clauses satisfied by the literals: $\posop(\phi, L) = \{ C \in \phi \mid  C \cap L = \emptyset\}$.
 
@@
 \item Literals: $\neg x = \obar{x}$ and $\neg \obar{x} = x$
-\item Negations: $\neg \neg \psi = \psi$
+\item Double negations: $\neg \neg \psi = \psi$
 \end{itemize}
@@
 \end{itemize}
+\end{itemize}
+
+With these simplifications, we can assume that the POFs we generate will satisfy the following:
+\begin{itemize}
+\item Constant values $\tautology$ and $\nil$ only occur when the POF is itself
+  $\tautology$ or $\nil$.
+\item Negation is only applied to product and sum operations.
+\item Every product and sum operation will have at least two arguments.
 \end{itemize}
@@
   \item Define clausal formula $\phi'$ to be $\bcp(\simplify{\phi}{L})$
-  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.
+  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.  (Partitioning is only required when $m > 1$.)
   \item Return the product (with simplifications) for the set of arguments $\{\lit \in L \mid  \dependencyset(\lit) \subseteq \dvarset\} \cup \{ \psi'_i \mid  1 \leq i \leq m \}$.
@@
 operations when the decision variable $x$ is to be projected.  The
-recursive calls yielded subformulas $\psi'_1$ and $\psi'_2$, representing the projected compilation of $\phi$ when $x$ is assigned $1$ and $0$, respectively.
+recursive calls yield subformulas $\psi'_1$ and $\psi'_2$, representing the projected compilation of $\phi$ when $x$ is assigned $1$ and $0$, respectively.
 These
@@
 If $\pure_{\pvarset}(\phi) \not = \emptyset$, then $\phi$ can be reduced as
-$\bcp(\simplify{\phi}{\pure_{\pvarset}(\phi)})$.
+$\phi \leftarrow \bcp(\simplify{\phi}{\pure_{\pvarset}(\phi)})$.
 \end{itemize}

### 2023-08-01 doc/pkc-notes.tex
Refinements

@@
 \begin{description}
-\item[$\posop(\phi, L)$:] Remove tautologies, as well as clauses satisfied by the literals: $\posop(\phi, L) = \{ C \in \phi \mid  C \cap L = \emptyset\}$.
+\item[$\posop(\phi, L)$:] Remove clauses satisfied by the literals in $L$: $\posop(\phi, L) = \{ C \in \phi \mid C \cap L = \emptyset\}$.
 
-\item[$\negop(\phi, L)$:] Remove the negations of the literals from all clauses.  Let $\negop(C, L) = \{ \lit \in C \mid  \obar{\lit} \not \in L\}$.
+\item[$\negop(\phi, L)$:] Remove the negations of the literals in $L$ from all clauses.  Let $\negop(C, L) = \{ \lit \in C \mid  \obar{\lit} \not \in L\}$.
   Then $\negop(\phi, L) = \{ \negop(C, L) \mid  C \in \phi \}$.  If this process generates an empty clause, then the result is $\nil$.
 
-\item[$\simplify{\phi}{L}$:] Simplify the clauses by assigning the literals, removing all occurrences of any literal for the variables in $\dependencyset(L)$:
+\item[$\simplify{\phi}{L}$:] Simplify the clauses by assigning the literals in $L$, removing all occurrences of any literal for the variables in $\dependencyset(L)$:
 $\simplify{\phi}{L} = \posop(\negop(\phi, L), L)$.
@@
 
-When constructing POFs via a bottom-up process, we can perform a set of simplifications, as follows:
-
+When constructing a POF via a bottom-up process, we can perform simplifications as the formula is constructed, as follows:
 \begin{itemize}
@@
 Our goal is to implement a projecting knowledge compiler that produces a POF $\pkc(\phi)$ when given input formula $\phi$.
-We define our algorithm in terms of a recursive function $\traverse(\psi, \phi)$ that simultaneously operates on a POF $\psi$ and a clausal formula $\phi$.
+We define our algorithm in terms of a recursive function $\traverse(\psi, \phi)$ that traverses a POF $\psi$ while maintaining its clausal formula $\phi$.
 Projected knowledge compilation is then defined
@@
 \item[Product Operations:]
-  Let the product operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
+  Let the operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
   \begin{enumerate}
@@
 \item[Sum Operations:]
-  Let the sum operation be of the form $S(\lit_1, \lit_2)$ and having decision variable $x$.
+  Let the operation be of the form $S(\lit_1, \lit_2)$ and having decision variable $x$.
   \begin{enumerate}
@@
 $\dmodelset(\psi') \cap \dmodelset(\neg \psi'_1) = \emptyset$.
+The two arguments to the top-level sum operation are disjoint, because the overlapping models have been subtracted from the first argument.
 As can be seen, we could also reverse the role of the two arguments, computing the result
@@
 \end{itemize}
+
+The first two termination rules have some subtle difference.  For the
+initial formulation of $\traverse$, the arguments will always satisfy
+$\dependencyset(\psi) \subseteq \dependencyset(\phi)$.  However, the
+pure literal reduction rule can eliminate some literals from $\phi$
+that occur in $\psi$, and so the two sets of variables can become
+incomparable.  When both rules are applicable, it is more efficient to use the first rule, since it avoids an invocation of $\kc$.
 
@@
 example formula, derived from the dec-DNNF representation produced by
-\dfour{}.
-In this graph, product nodes are labeled with the operation $\pand$, and sum nodes are labeled with the operation $\por$, emphasizing that these operations satisfy their respective partitioning constraints.
+\dfour{}.  In this graph, product nodes are labeled $P$, and sum nodes are labeled  $S$.
 We can see that \dfour{} used $y$ as the decision variable at the top level, creating sum node $\nodes_7$ having edges to
@@
 and this is compiled to node $\nodep_{8}$ representing product operation $P(x_1, x_2, x_3)$.
-Finally, the nodes are assembled using our DeMorgan's Law transformations as shown with root node $\nodes_{11}$.
+Finally, the nodes are assembled using our DeMorgan's Law transformations yielding a root edge to node $\nodes_{11}$.
 

### 2023-08-03 cpog/prototype/cpog_checker.py
Fixed bug in density computation

@@ -998,7 +998,7 @@ class OperationManager:
                 if arg < 0:
                     val = val.oneminus()
                 wts.append(val)
-            result = P52(1)
+            result = P52(1) if op == self.conjunction else P52(0)
             for w in wts:
                 result = result.mul(w) if op == self.conjunction else result.add(w)
             weights[outVar] = result

### 2023-08-03 cpog/generator/cpog-generate.cpp
Fix comment

@@ -1,7 +1,5 @@
 // Translate DNNF representation into POG
 
-// Usage: d2p [VLEVEL] file.cnf file.d4nnf
-
 #include <stdio.h>
 #include <stdlib.h>
 #include <unistd.h>

### 2023-08-03 doc/pkc-notes.tex
Polishing

@@
 \newcommand{\traverse}{\algo{Traverse}}
-\newcommand{\posop}{\algo{Pos}}
-\newcommand{\negop}{\algo{Neg}}
+\newcommand{\cfilter}{\algo{FilterClause}}
+\newcommand{\lfilter}{\algo{FilterLiteral}}
 \newcommand{\bcp}{\algo{BCP}}
@@
   formula is in \emph{conjunctive normal form} (CNF) when i) it is in
-  negation normal form, and ii) sum is applied only to literals.  We use symbol $\phi$ to denote CNF formulas.
+  negation normal form, ii) the sum operation is applied only to literals, and iii) the product is applied only to sums.  We use symbol $\phi$ to denote CNF formulas.
 
@@
 
+  A \emph{commutative ring} $\ring$ is an algebraic structure
+  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
+  with elements in the set $\dset$ and with commutative and
+  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
+  such that multiplication distributes
+  over addition.  $\addident$ is the additive identity and $\mulident$ is
+  the multiplicative identity.  Every element $a \in \dset$ has an
+  \emph{additive inverse} $-a$ such that $a + -a = \addident$.
+We write $a - b$ as a shorthand for $a + -b$.
+
   For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
@@
   \begin{eqnarray}
-    \drep(\theta, w) & = & \sum_{\alpha \in \dmodelset(\theta)} \;\; \prod_{\lit \in \alpha} w(\lit) \label{eqn:rep}
+    \drep[\theta, w] & = & \sum_{\alpha \in \dmodelset(\theta)} \;\; \prod_{\lit \in \alpha} w(\lit) \label{eqn:rep}
   \end{eqnarray}
@@
 \begin{itemize}
-\item Constants: $\drep(\tautology) = \mulident$ and $\drep(\nil) = \addident$
-\item Literals: $\drep(x) = w(x)$ and $\drep(\obar{x}) = \mulident - x$
+\item Constants: $\drep[\tautology] = \mulident$ and $\drep[\nil] = \addident$
+\item Literals: $\drep[x] = w(x)$ and $\drep[\obar{x}] = \mulident - x$
 \item Product operations: 
-$\drep(P(\psi_1, \psi_2, \ldots, \psi_k)) = \prod_{1 \leq i \leq k} \drep(\psi_i)$
+$\drep[P(\psi_1, \psi_2, \ldots, \psi_k)] = \prod_{1 \leq i \leq k} \drep(\psi_i)$
 \item Sum Operations
-$\drep(S(\psi_1, \psi_2, \ldots, \psi_k)) = \sum_{1 \leq i \leq k} \drep(\psi_i)$
-\item Negations: $\drep(\neg \psi) = \mulident -\psi$
+$\drep[S(\psi_1, \psi_2, \ldots, \psi_k)] = \sum_{1 \leq i \leq k} \drep(\psi_i)$
+\item Negations: $\drep[\neg \psi] = \mulident -\drep[\psi]$
 \end{itemize}
@@
 
-We can build on existing knowledge compilers to implement a projected knowledge compiler.
+We can implement a projected knowledge compiler that operates by making multiple invocations of a standard knowledge compiler.
 
@@
 
-Let $U(\phi)$ denote the set of \emph{unit literals} in $\phi$, i.e., the set of all literals $\lit$ such that clause $\{ \lit \} \in \phi$.  In the following $L$ denotes a set of literals.
+Let $U(\phi)$ denote the set of \emph{unit literals} in $\phi$, i.e., the set of all literals $\lit$ such that clause $\{ \lit \} \in \phi$.  
+We define the following operations for transforming clausal formula $\phi$ into a new clausal formula.  In the following $L$ denotes a set of literals.
+\begin{description}
+\item[$\cfilter(\phi, L)$:] Remove clauses satisfied by the literals in $L$: $\cfilter(\phi, L) = \{ C \in \phi \mid C \cap L = \emptyset\}$.
 
-We define the following operations for transforming clausal formula $\phi$ into a new clausal formula:
-\begin{description}
-\item[$\posop(\phi, L)$:] Remove clauses satisfied by the literals in $L$: $\posop(\phi, L) = \{ C \in \phi \mid C \cap L = \emptyset\}$.
-
-\item[$\negop(\phi, L)$:] Remove the negations of the literals in $L$ from all clauses.  Let $\negop(C, L) = \{ \lit \in C \mid  \obar{\lit} \not \in L\}$.
-  Then $\negop(\phi, L) = \{ \negop(C, L) \mid  C \in \phi \}$.  If this process generates an empty clause, then the result is $\nil$.
+\item[$\lfilter(\phi, L)$:] Remove the literals in $L$ from all clauses.  
+   $\lfilter(\phi, L) = \{ C - L \mid  C \in \phi \}$.  If this process generates an empty clause, then the result is $\nil$.
 
 \item[$\simplify{\phi}{L}$:] Simplify the clauses by assigning the literals in $L$, removing all occurrences of any literal for the variables in $\dependencyset(L)$:
-$\simplify{\phi}{L} = \posop(\negop(\phi, L), L)$.
+$\simplify{\phi}{L} = \lfilter(\cfilter(\phi, L), \obar{L})$, where $\obar{L} = \{ \obar{\lit} \mid \lit \in L \}$.
 
@@
 
-\item[$\ureduce{\phi}{x}$:] Express the universal quantification of $\phi$ with respect to variable $x$ in clausal form:
-$\ureduce{\phi}{x} = \negop(\phi, \{x, \obar{x}\})$.
+\item[$\ureduce{\phi}{x}$:] Express the universal quantification of $\phi$ with respect to variable $x$ in clausal form, by removing all occurrences of $x$ and $\obar{x}$:
+$\ureduce{\phi}{x} = \lfilter(\phi, \{x, \obar{x}\})$.
 
@@
 
-When constructing a POF via a bottom-up process, we can perform simplifications as the formula is constructed, as follows:
+When constructing a POF via a bottom-up process, we can perform simplifications as the formula is constructed:
 \begin{itemize}
 \item
-The use of negation can be optimized as follows:
+Optimized negation:
 \begin{itemize}
@@
 \item 
-When forming a product $P(\psi_1, \psi_2, \ldots, \psi_k)$ we can perform the following simplifications:
+Optimizing
+product $P(\psi_1, \psi_2, \ldots, \psi_k)$:
 \begin{itemize}
@@
 \item
-Similarly, when forming a sum $S(\psi_1, \psi_2, \ldots, \psi_k)$ we can perform the following simplifications:
+Optimizing sum $S(\psi_1, \psi_2, \ldots, \psi_k)$:
 \begin{itemize}
@@
 
-We assume that we have a knowledge compiler that can convert a clausal formula $\phi$ into a logically equivalent POF $\psi$.
-Furthermore, we assume that the generated POF obeys the following restrictions:
+A standard knowledge compiler converts a clausal formula $\phi$ into a logically equivalent POF $\psi$, denoted
+$\psi = \kc(\phi)$.
+We assume that the generated POF obeys the following restrictions:
 \begin{itemize}
@@
 \item The sum operations are \emph{decision operations}.  That is, each sum operation is of the form $S(\psi_1, \psi_2)$, and it has an associated decision variable $x$,
-  such that 1) either $\psi_1 = x$ or $\psi_1$ is a product operation having $x$ as an argument, and 2)
+  such that i) either $\psi_1 = x$ or $\psi_1$ is a product operation having $x$ as an argument, and ii)
 either $\psi_2 = \obar{x}$ or $\psi_2$ is a product operation having $\obar{x}$ as an argument.
 \end{itemize}
-We let $\kc(\phi)$ denote the output of the knowledge compiler for input formula $\phi$.
+These restrictions hold for a knowledge compiler that produces a dec-DNNF representation of the input formula, with this representation converted to a POF\@.
 
@@
 
-Our goal is to implement a projecting knowledge compiler that produces a POF $\pkc(\phi)$ when given input formula $\phi$.
-We define our algorithm in terms of a recursive function $\traverse(\psi, \phi)$ that traverses a POF $\psi$ while maintaining its clausal formula $\phi$.
+Our goal is to implement a projecting knowledge compiler that produces POF $\psi = \pkc(\phi)$ when given input formula $\phi$.
+We define our algorithm in terms of a recursive function $\traverse(\psi, \phi)$ that traverses POF $\psi$ while maintaining its clausal representation $\phi$.
 Projected knowledge compilation is then defined
@@
   \item If $x \in \dvarset$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$.
-  \item Otherwise, let $\psi' = \pkc(\ureduce{\phi}{x})$
+  \item Otherwise, let $\psi_{\forall} = \pkc(\ureduce{\phi}{x})$
     \begin{enumerate}
-    \item If $\psi' = \nil$, then return the sum (with simplifications) $S(\psi'_1, \psi'_2)$
-    \item Otherwise return nested formula $S(\neg S(\neg \psi'_1, \psi'), \psi'_2)$.
+    \item If $\psi_{\forall} = \nil$, then return the sum (with simplifications) $S(\psi_1, \psi_2)$
+    \item Otherwise return nested formula $S(\neg S(\neg \psi'_1, \psi_{\forall}), \psi'_2)$.
     \end{enumerate}
@@
 representation of the universal quantification of $\phi$ with respect
-to $x$, and so $\psi' = \pkc(\ureduce{\phi}{x})$ will be a POF whose
+to $x$, and so $\psi_{\forall} = \pkc(\ureduce{\phi}{x})$ will be a POF whose
 data models are the intersection of those for $\psi_1'$ and $\psi_2'$.  In
@@
 Boolean formula over the data variables.  In that case, the sum can be
-formed directly.  If $\psi' \not = \tautology$, we want to
+formed directly.  If $\psi_{\forall} \not = \nil$, we want to
 ``subtract'' the overlapping models from one of the arguments before
 forming the sum with the other.  We can do so using DeMorgan's Laws.  
-That is formula $\neg S(\neg \psi'_1, \psi')$ will have as data models $\dmodelset(\psi'_1) - \dmodelset(\psi')$.  Furthermore, the arguments to this sum operation will have disjoint models
-since $\dmodelset(\psi') \subseteq \dmodelset(\psi'_1)$, and therefore
-$\dmodelset(\psi') \cap \dmodelset(\neg \psi'_1) = \emptyset$.
+That is formula $\neg S(\neg \psi'_1, \psi_{\forall})$ is logically equivalent to $\psi'_1 \land \neg \psi_{\forall}$,
+and so it
+will have as data models $\dmodelset(\psi'_1) - \dmodelset(\psi_{\forall})$.  Furthermore, the arguments to this sum operation will have disjoint models
+since $\dmodelset(\psi_{\forall}) \subseteq \dmodelset(\psi'_1)$, and therefore
+$\dmodelset(\psi_{\forall}) \cap \dmodelset(\neg \psi'_1) = \emptyset$.
 The two arguments to the top-level sum operation are disjoint, because the overlapping models have been subtracted from the first argument.
 As can be seen, we could also reverse the role of the two arguments, computing the result
-$S(\psi'_1, \neg S(\neg \psi'_2, \psi'))$.
+$S(\psi'_1, \neg S(\neg \psi'_2, \psi_{\forall}))$.
 
@@
 If $\pure_{\pvarset}(\phi) \not = \emptyset$, then $\phi$ can be reduced as
-$\phi \leftarrow \bcp(\simplify{\phi}{\pure_{\pvarset}(\phi)})$.
+ $\phi \leftarrow \cfilter(\phi, \pure_{\pvarset})$.  Eliminating clauses can cause other literals to become pure, and so this reduction should be iterated until no pure literals remain.
 \end{itemize}
 
-The first two termination rules have some subtle difference.  For the
-initial formulation of $\traverse$, the arguments will always satisfy
-$\dependencyset(\psi) \subseteq \dependencyset(\phi)$.  However, the
-pure literal reduction rule can eliminate some literals from $\phi$
-that occur in $\psi$, and so the two sets of variables can become
-incomparable.  When both rules are applicable, it is more efficient to use the first rule, since it avoids an invocation of $\kc$.
+The first two termination rules have some subtle differences.  For the
+unoptimized version of $\traverse$, the arguments will always satisfy
+$\dependencyset(\psi) \subseteq \dependencyset(\phi)$.  However, 
+pure literal reduction can eliminate some literals from $\phi$
+that remain in $\psi$, and so the two sets of variables can become
+incomparable.  When both rules are applicable, it is more efficient to use the first, since it avoids an invocation of $\kc$.
 

### 2023-08-03 cpog/pkc/clause.cpp
Start of performant PKC

@@ -1,0 +1,440 @@
+/*========================================================================
+  Copyright (c) 2022, 2023 Randal E. Bryant, Carnegie Mellon University
+  
+  Permission is hereby granted, free of charge, to any person
+  obtaining a copy of this software and associated documentation files
+  (the "Software"), to deal in the Software without restriction,
+  including without limitation the rights to use, copy, modify, merge,
+  publish, distribute, sublicense, and/or sell copies of the Software,
+  and to permit persons to whom the Software is furnished to do so,
+  subject to the following conditions:
+  
+  The above copyright notice and this permission notice shall be
+  included in all copies or substantial portions of the Software.
+  
+  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
+  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
+  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
+  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
+  SOFTWARE.
+  ========================================================================*/
+
+#include <cstdlib>
+#include <cstdio>
+#include <ctype.h>
+#include <cstring>
+#include "clause.hh"
+#include "report.h"
+#include "counters.h"
+
+global_data global;
+
+static int skip_line(FILE *infile) {
+    int c;
+    while ((c = getc(infile)) != EOF) {
+	if (c == '\n')
+	    return c;
+    }
+    return c;
+}
+
+// Skip over comment lines, spaces, newlines, etc., until find something interesting
+// Return false if EOF encountered without finding anything
+static bool find_token(FILE *infile) {
+    int c;
+    while ((c = getc(infile)) != EOF) {
+	if (c == 'c') {
+	    c = skip_line(infile);
+	    ungetc(c, infile);
+	} else if (!isspace(c)) {
[…]

### 2023-08-03 cpog/pkc/counters.h

@@ -1,0 +1,70 @@
+/*========================================================================
+  Copyright (c) 2023 Randal E. Bryant, Carnegie Mellon University
+  
+  Permission is hereby granted, free of
+  charge, to any person obtaining a copy of this software and
+  associated documentation files (the "Software"), to deal in the
+  Software without restriction, including without limitation the
+  rights to use, copy, modify, merge, publish, distribute, sublicense,
+  and/or sell copies of the Software, and to permit persons to whom
+  the Software is furnished to do so, subject to the following
+  conditions:
+  
+  The above copyright notice and this permission notice shall be
+  included in all copies or substantial portions of the Software.
+  
+  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
+  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
+  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
+  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
+  SOFTWARE.
+========================================================================*/
+
+
+#pragma once
+
+#include "ilist.h"
+
+// Count all the interesting stuff
+
+typedef enum { 
+    COUNT_INPUT_VAR, COUNT_INPUT_CLAUSE, COUNT_KC_POG_PRODUCT, COUNT_KC_POG_SUM, COUNT_KC_POG_SIZE, 
+    COUNT_POG_PRODUCT, COUNT_POG_SUM, COUNT_POG_SIZE,
+    COUNT_VISIT_AND, COUNT_VISIT_DATA_SUM, COUNT_VISIT_MUTEX_SUM, COUNT_VISIT_EXCLUDING_SUM,
+    COUNT_SAT_CALL, COUNT_KC_CALL,
+    COUNT_NUM
+} counter_t;
+
+typedef enum { TIME_TOTAL, TIME_SAT, TIME_KC, TIME_NUM } timer_t;
+
[…]

### 2023-08-03 cpog/pkc/pkc.cpp

@@ -1,0 +1,72 @@
+#include <cstdio>
+#include <cstdlib>
+#include <unistd.h>
+
+
+#include "report.h"
+#include "counters.h"
+#include "clause.hh"
+#include "pog.hh"
+
+void usage(const char *name) {
+    lprintf("Usage: %s [-h] [-v VLEVEL] FORMULA.cnf FORMULA.pog\n", name);
+    lprintf("  -h          Print this information\n");
+    lprintf("  -v VERB     Set verbosity level\n");
+}
+
+static int run(FILE *cnf_file) {
+    Cnf cnf;
+    cnf.import_file(cnf_file);
+    fprintf(stdout, "Initial file:\n");
+    if (cnf.is_loaded()) {
+	cnf.write(stdout);
+    } else {
+	err(false, "Failed to load CNF file\n");
+	return 1;
+    }
+    Clausal_reasoner cr(&cnf);
+    bool conflict = cr.bcp(100);
+    cnf_archive_t arx = cr.extract();
+    cnf.deallocate();
+    Cnf ncnf;
+    ncnf.import_archive(arx);
+    fprintf(stdout, "After BCP (conflict = %s):\n", conflict ? "true" : "false");    
+    ncnf.write(stdout);
+    ncnf.deallocate();
+    return 0;
+}
+
+int main(int argc, char *const argv[]) {
+    FILE *cnf_file = NULL;
+    verblevel = 1;
+    int c;
+    while ((c = getopt(argc, argv, "hv:")) != -1) {
+	switch (c) {
+	case 'h':
+	    usage(argv[0]);
+	    return 0;
+	case 'v':
+	    verblevel = atoi(optarg);
+	    break;
+	default:
+	    lprintf("Unknown comandline option '%c'\n", c);
+	    usage(argv[0]);
+	    break;
+	}
+
+    }
+    int argi = optind;
+    if (argi >= argc) {
+	lprintf("Name of input CNF file required\n");
+	usage(argv[0]);
+	return 1;
+    }
+    cnf_file = fopen(argv[argi], "r");
+    if (cnf_file == NULL) {
+	lprintf("Can't open '%s'\n", argv[argi]);
+	return 1;
+    }
+    int result = run(cnf_file);
+    fclose(cnf_file);
+    return result;
+}

### 2023-08-03 cpog/pkc/pog.cpp

@@ -1,0 +1,495 @@
+/*========================================================================
+  Copyright (c) 2022, 2023 Randal E. Bryant, Carnegie Mellon University
+  
+  Permission is hereby granted, free of charge, to any person
+  obtaining a copy of this software and associated documentation files
+  (the "Software"), to deal in the Software without restriction,
+  including without limitation the rights to use, copy, modify, merge,
+  publish, distribute, sublicense, and/or sell copies of the Software,
+  and to permit persons to whom the Software is furnished to do so,
+  subject to the following conditions:
+  
+  The above copyright notice and this permission notice shall be
+  included in all copies or substantial portions of the Software.
+  
+  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+  EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+  MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+  NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS
+  BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN
+  ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN
+  CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
+  SOFTWARE.
+  ========================================================================*/
+
+#include <cstdlib>
+#include <ctype.h>
+#include <cstring>
+#include <map>
+#include "report.h"
+#include "counters.h"
+#include "pog.hh"
+
+// Support for computing hash function over POG arguments
+// Represent by signature over modular field
+
+#define STATE_BYTES 256
+#define CHUNK_SIZE 1024
+static const unsigned hash_modulus = 2147483647U;
+static char hash_state[STATE_BYTES];
+
+static std::vector<unsigned> var_hash;
+
+// Hash signatures for POG operations
+
+static unsigned pog_hash[POG_NUM];
+
+static void init_hash(int val) {
+    int var = IABS(val);
+    if (var_hash.size() == 0) {
+	// Initialization
[…]

### 2023-08-03 cpog/pkc/clause.cpp
Change handling of input variable count

@@ -30,8 +30,6 @@
 #include "report.h"
 #include "counters.h"
 
-global_data global;
-
 static int skip_line(FILE *infile) {
     int c;
     while ((c = getc(infile)) != EOF) {
@@ -59,7 +57,8 @@ static bool find_token(FILE *infile) {
 }
 
 // Tools to enable constructing CNF representations
-static void start_build(std::vector<int> &varx, int nclause) {
+static void start_build(std::vector<int> &varx, int nvar, int nclause) {
+    varx.push_back(nvar);
     varx.push_back(nclause);
     for (int cid = 1; cid <= nclause+1; cid++)
 	// Place holder for starting indices
@@ -67,7 +66,7 @@ static void start_build(std::vector<int> &varx, int nclause) {
 }
 
 static void new_clause(std::vector<int> &varx, int cid) {
-    varx[cid] = varx.size();
+    varx[cid+1] = varx.size();
 }
 
 static void add_literal(std::vector<int> &varx, int lit) {
@@ -75,8 +74,8 @@ static void add_literal(std::vector<int> &varx, int lit) {
 }
 
 static cnf_archive_t finish_build(std::vector<int> &varx) {
-    int nclause = varx[0];
-    varx[nclause+1] = varx.size();
+    int nclause = varx[1];
+    varx[nclause+2] = varx.size();
     cnf_archive_t arx = (cnf_archive_t) calloc(varx.size(), sizeof(int));
     if (!arx) {
 	err(false, "Couldn't allocate space for clauses\n");
@@ -97,7 +96,6 @@ void Cnf::import_file(FILE *infile) {
     int expectedNclause = 0;
     int nclause = 0;
     bool read_failed = false;
-    global.nvar = 0;
     bool got_header = false;
     int c;
     if (arx) {
@@ -130,8 +128,7 @@ void Cnf::import_file(FILE *infile) {
 		err(false, "Invalid CNF header\n");
 		return;
 	    } 
-	    global.nvar = expectedNvar;
-	    start_build(varx, expectedNclause);
+	    start_build(varx, expectedNvar, expectedNclause);
 	    c = skip_line(infile);
 	    got_header = true;
 	    break;
@@ -168,7 +165,7 @@ void Cnf::import_file(FILE *infile) {
     }	
     arx = finish_build(varx);
[…]

### 2023-08-03 cpog/pkc/pog.cpp

@@ -30,153 +30,6 @@
 #include "counters.h"
 #include "pog.hh"
 
-// Support for computing hash function over POG arguments
-// Represent by signature over modular field
-
-#define STATE_BYTES 256
-#define CHUNK_SIZE 1024
-static const unsigned hash_modulus = 2147483647U;
-static char hash_state[STATE_BYTES];
-
-static std::vector<unsigned> var_hash;
-
-// Hash signatures for POG operations
-
-static unsigned pog_hash[POG_NUM];
-
-static void init_hash(int val) {
-    int var = IABS(val);
-    if (var_hash.size() == 0) {
-	// Initialization
-	initstate(1, hash_state, STATE_BYTES);
-	for (int i = 0; i <= POG_NUM; i++) {
-	    pog_hash[i] = random() % hash_modulus;
-	}
-    }
-    if (var >= var_hash.size()) {
-	// Add more random values
-	size_t osize = var_hash.size();
-	size_t nsize = osize + (1 + (var - osize)/CHUNK_SIZE) * CHUNK_SIZE;
-	var_hash.resize(nsize);
-	const char *ostate = setstate(hash_state);
-	for (unsigned i = osize; i < nsize; i++)
-	    var_hash[i] = random() % hash_modulus;
-	setstate(ostate);
-    }
-}
-
-static unsigned next_hash_int(unsigned sofar, int val) {
-    init_hash(val);
-    int var = IABS(val);
-    unsigned vval = var_hash[var];
-    unsigned long  lval = val < 0 ? hash_modulus - vval : vval;
-    return (lval * sofar) % hash_modulus;
-}
-
-unsigned Pog::node_hash(int var) {
-    int idx = node_index(var);
-    if (idx < 0)
-	return 0;
-    init_hash(var);
-    unsigned sofar = pog_hash[(int) nodes[idx].type];
-    int offset = nodes[idx].offset;
-    int degree = nodes[idx].degree;
-    for (int i = 0; i < degree; i++) {
-	sofar = next_hash_int(sofar, arguments[offset + i]);
-    }
-    return sofar;
-}
-
-bool Pog::node_equal(int var1, int var2) {
-    int idx1 = node_index(var1);
-    int idx2 = node_index(var2);
-    if (idx1 == idx2)
-	return true;
-    if (idx1 < 0 || idx2 < 0)
-	return false;
-    if (nodes[idx1].type != nodes[idx2].type)
-	return false;
[…]

### 2023-08-03 cpog/pkc/pog_test.cpp

@@ -1,0 +1,68 @@
+#include <cstdio>
+#include <cstdlib>
+#include <unistd.h>
+
+
+#include "report.h"
+#include "counters.h"
+#include "pog.hh"
+
+void usage(const char *name) {
+    lprintf("Usage: %s [-h] [-n NVAR] [-v VLEVEL] FORMULA.nnf ...\n", name);
+    lprintf("  -h          Print this information\n");
+    lprintf("  -v VERB     Set verbosity level\n");
+    lprintf("  -n NVAR     Set number of input variables\n");
+}
+
+Pog *pog;
+
+static int run(FILE *nnf_file) {
+    int root = pog->load_nnf(nnf_file);
+    printf("Got root %d\n", root);
+    pog->show(stdout);
+    return 0;
+}
+
+int main(int argc, char *const argv[]) {
+    FILE *nnf_file = NULL;
+    verblevel = 1;
+    int nvar = 100;
+    int c;
+    while ((c = getopt(argc, argv, "hn:v:")) != -1) {
+	switch (c) {
+	case 'h':
+	    usage(argv[0]);
+	    return 0;
+	case 'v':
+	    verblevel = atoi(optarg);
+	    break;
+	case 'n':
+	    nvar = atoi(optarg);
+	    break;
+	default:
+	    lprintf("Unknown comandline option '%c'\n", c);
+	    usage(argv[0]);
+	    break;
+	}
+
+    }
+    pog = new Pog(nvar);
+    int argi = optind;
+    if (argi >= argc) {
+	lprintf("Name of input NNF file required\n");
+	usage(argv[0]);
+	return 1;
+    }
+    int result = 0;
+    while (result == 0 && argi < argc) {
+	nnf_file = fopen(argv[argi], "r");
+	if (nnf_file == NULL) {
+	    lprintf("Can't open '%s'\n", argv[argi]);
+	    return 1;
+	}
+	result = run(nnf_file);
+	fclose(nnf_file);
+	argi++;
+    }
+    return result;
+}

### 2023-08-08 cpog/pkc/clause.cpp
PKC working on simple examples

@@ -71,6 +71,12 @@ void File_manager::flush() {
     }
 }
 
+// Put literals in ascending order of the variables
+static bool abs_less(int x, int y) {
+    return IABS(x) < IABS(y);
+}
+
+
 static int skip_line(FILE *infile) {
     int c;
     while ((c = getc(infile)) != EOF) {
@@ -154,9 +160,9 @@ Cnf::Cnf() {
     arx = NULL;
 }
 
-// Process comment, looking additional show variables
+// Process comment, looking additional data variables
 // Return last character
-static void process_comment(FILE *infile, std::unordered_set<int> &show_variables) {
+static void process_comment(FILE *infile, std::unordered_set<int> &data_variables) {
     char buf[50];
     int len;
     if (find_string_token(infile, buf, 50, &len) && len == 1 && strncmp(buf, "p", 1) == 0
@@ -164,10 +170,10 @@ static void process_comment(FILE *infile, std::unordered_set<int> &show_variable
 	int var = -1;
 	while (var != 0) {
 	    if (fscanf(infile, "%d", &var) != 1) {
-		err(false, "Couldn't read show variable\n");
+		err(false, "Couldn't read data variable\n");
 		break;
 	    } else if (var != 0) {
-		show_variables.insert(var);
+		data_variables.insert(var);
 	    }
 	}
     }
@@ -188,14 +194,14 @@ void Cnf::import_file(FILE *infile) {
     }
     std::vector<int> varx;
     bool eof = false;
-    // Clear set of show variables
-    show_variables.clear();
+    // Clear set of data variables
+    data_variables.clear();
     // Look for CNF header
     while ((c = getc(infile)) != EOF) {
 	if (isspace(c)) 
 	    continue;
 	if (c == 'c') {
-	    process_comment(infile, show_variables);
+	    process_comment(infile, data_variables);
 	    continue;
 	}
 	if (c == EOF) {
@@ -243,7 +249,7 @@ void Cnf::import_file(FILE *infile) {
 		return;
 	    } else if (c == 'c' && starting_clause) {
 		c = getc(infile);
-		process_comment(infile, show_variables);
+		process_comment(infile, data_variables);
 		continue;
 	    }
 	    else if (fscanf(infile, "%d", &lit) != 1) {
@@ -261,7 +267,7 @@ void Cnf::import_file(FILE *infile) {
 	if (isspace(c)) 
 	    continue;
 	if (c == 'c') 
[…]

### 2023-08-08 cpog/pkc/counters.h

@@ -32,15 +32,15 @@
 
 typedef enum { 
     COUNT_INPUT_VAR, COUNT_INPUT_CLAUSE, COUNT_KC_POG_PRODUCT, COUNT_KC_POG_SUM, COUNT_KC_POG_SIZE, 
-    COUNT_POG_PRODUCT, COUNT_POG_SUM, COUNT_POG_SIZE,
-    COUNT_VISIT_AND, COUNT_VISIT_DATA_SUM, COUNT_VISIT_MUTEX_SUM, COUNT_VISIT_EXCLUDING_SUM,
+    COUNT_POG_PRODUCT, COUNT_POG_SUM,
+    COUNT_VISIT_PRODUCT, COUNT_VISIT_DATA_SUM, COUNT_VISIT_MUTEX_SUM, COUNT_VISIT_EXCLUDING_SUM,
     COUNT_SAT_CALL, COUNT_KC_CALL,
     COUNT_NUM
 } counter_t;
 
 typedef enum { TIME_TOTAL, TIME_SAT, TIME_KC, TIME_NUM } timer_t;
 
-typedef enum { HISTO_CLAUSAL, HISTO_POG, HISTO_NUM } histogram_t;
+typedef enum { HISTO_SAT_CLAUSES, HISTO_KC_CLAUSES, HISTO_POG_NODES, HISTO_NUM } histogram_t;
 
 /* Allow this headerfile to define C++ constructs if requested */
 #ifdef __cplusplus

### 2023-08-08 cpog/pkc/pkc.cpp

@@ -9,28 +9,43 @@
 
 
 void usage(const char *name) {
-    lprintf("Usage: %s [-h] [-v VLEVEL] FORMULA.cnf FORMULA.pog\n", name);
+    lprintf("Usage: %s [-h] [-k] [-v VLEVEL] FORMULA.cnf FORMULA.pog\n", name);
     lprintf("  -h          Print this information\n");
+    lprintf("  -k          Keep intermdiate files\n");
     lprintf("  -v VERB     Set verbosity level\n");
 }
 
-static int run(const char *cnf_name) {
+static int run(const char *cnf_name, const char *pog_name) {
     Project proj(cnf_name);
+    if (verblevel >= 5) {
+	printf("Initial POG:\n");
+	proj.show(stdout);
+    }
+    proj.projecting_compile();
+    if (verblevel >= 5) {
+	printf("Projected POG:\n");
+	proj.show(stdout);
+    }
+    proj.write(pog_name);
     return 0;
 }
 
 int main(int argc, char *const argv[]) {
     FILE *cnf_file = NULL;
     verblevel = 1;
+    bool keep = false;
     int c;
     double start = tod();
-    while ((c = getopt(argc, argv, "hv:")) != -1) {
+    while ((c = getopt(argc, argv, "hkv:")) != -1) {
 	switch (c) {
 	case 'h':
 	    usage(argv[0]);
 	    return 0;
 	case 'v':
 	    verblevel = atoi(optarg);
+	    break;
+	case 'k':
+	    keep = true;
 	    break;
 	default:
 	    lprintf("Unknown comandline option '%c'\n", c);
@@ -45,8 +60,23 @@ int main(int argc, char *const argv[]) {
 	usage(argv[0]);
 	return 1;
     }
-    int result = run(argv[argi]);
+    const char *cnf_name = argv[argi++];
+    if (argi >= argc) {
+	lprintf("Name of output POG file required\n");
+	usage(argv[0]);
+	return 1;
+    }
+    const char *pog_name = argv[argi++];
+    if (argi < argc) {
+	lprintf("Unknown argument '%s'", argv[argi]);
+	usage(argv[0]);
+	return 1;
+    }
+    int result = run(cnf_name, pog_name);
     double elapsed = tod() - start;
+    
     lprintf("Total execution %.2f seconds\n", elapsed);
+    if (!keep)
+	fmgr.flush();
     return result;
 }

### 2023-08-08 cpog/pkc/pog.cpp

@@ -32,6 +32,11 @@
 #include "pog.hh"
 
 
+// Put literals in ascending order of the variables
+static bool abs_less(int x, int y) {
+    return IABS(x) < IABS(y);
+}
+
 /// Support for NNF reading
 
 // Try to read single alphabetic character from line
@@ -157,7 +162,7 @@ Nnf::Nnf(int n, FILE *infile) {
 		std::vector<int> *node = new std::vector<int>;
 		nodes[nid] = node;
 		node->push_back((int) ntype);
-		report(5, "Line #%d.  Created NNF type %s node %d from NNF node %d\n",
+		report(6, "Line #%d.  Created NNF type %s node %d from NNF node %d\n",
 		       line_number, nnf_type_name[ntype], nid, largs[0]); 
 	    }
 	} else {
@@ -191,13 +196,13 @@ Nnf::Nnf(int n, FILE *infile) {
 		for (int i = 2; i < largs.size()-1; i++)
 		    xnode->push_back(largs[i]);
 		xnode->push_back(cnid);
-		report(5, "Line #%d. Created node %d to hold literals between nodes %d and %d\n",
+		report(6, "Line #%d. Created node %d to hold literals between nodes %d and %d\n",
 		       line_number, xid, pnid, cnid);
 		cnid = xid;
 	    }
 	    parent->push_back(cnid);
 	    node_with_parent.insert(cnid);
-	    report(5, "Line #%d.  Adding edge between nodes %d and %d\n", line_number, pnid, cnid);
+	    report(6, "Line #%d.  Adding edge between nodes %d and %d\n", line_number, pnid, cnid);
 	}
     }
     for (auto kv : nodes) {
@@ -207,7 +212,7 @@ Nnf::Nnf(int n, FILE *infile) {
 	if ((*node)[0] == NNF_OR && node->size() == 2 && root_id == 0
 	    && node_with_parent.find(nid) == node_with_parent.end()) {
 	    root_id = nid;
-	    report(5, "Setting root to %d\n", nid);
+	    report(6, "Setting root to %d\n", nid);
 	}
     }
     if (root_id == 0)
@@ -344,12 +349,39 @@ bool Pog::node_equal(int var1, int var2) {
     return true;
 }
 
+int Pog::get_decision_variable(int edge) {
+    if (get_type(edge) != POG_SUM)
+	return 0;
+    int edge1 = get_argument(edge, 0);
+    int n1 = 1;
[…]

### 2023-08-08 cpog/pkc/project.cpp

@@ -9,11 +9,15 @@ Project::Project(const char *cnf_name) {
 	err(true, "Couldn't open file '%s'\n", cnf_name);
     Cnf *cnf = new Cnf();
     cnf->import_file(cnf_file);
+    fclose(cnf_file);
     if (!cnf->is_loaded())
 	err(true, "Failed to load CNF from file '%s'\n", cnf_name);
-    report(1, "CNF file loaded %d variables, %d clauses, %d show variables\n", cnf->variable_count(), cnf->clause_count(), cnf->show_variables.size());
+    report(1, "CNF file loaded %d variables, %d clauses, %d data variables\n", cnf->variable_count(), cnf->clause_count(), cnf->data_variables.size());
+    incr_count_by(COUNT_INPUT_VAR, cnf->variable_count());
+    incr_count_by(COUNT_INPUT_CLAUSE, cnf->clause_count());
     fmgr.set_root(cnf_name);
     cr = new Clausal_reasoner(cnf);
+    cr->bcp(false);
     pog = new Pog(cnf->variable_count());
     root_literal = compile();
     report(1, "Initial POG created.  %d node, %d edges  POG size %d  Root literal = %d\n", 
@@ -37,21 +41,136 @@ int Project::compile() {
     cr->write(cnf_file);
     fclose(cnf_file);
     snprintf(cmd, 200, "d4 %s -dDNNF -out=%s > /dev/null", cnf_name, nnf_name);
-    report(3, "Running '%s'\n", cmd);
+    report(4, "Running '%s'\n", cmd);
     FILE *pipe = popen(cmd, "w");
     int rc = pclose(pipe);
     double elapsed = tod()-start;
     incr_timer(TIME_KC, elapsed);
-    report(2, "Running D4 on %s required %.3f seconds.  Return code = %d\n", cnf_name, elapsed, rc);
+    incr_count(COUNT_KC_CALL);
+    incr_histo(HISTO_KC_CLAUSES, cr->active_clause_count());
+    report(3, "Running D4 on %s required %.3f seconds.  Return code = %d\n", cnf_name, elapsed, rc);
 
     FILE *nnf_file = fopen(nnf_name, "r");
     if (!nnf_file)
 	err(true, "Couldn't open NNF file '%s'\n", nnf_name);
+    int osize = pog->node_count();
     int root = pog->load_nnf(nnf_file);
[…]

### 2023-08-08 cpog/prototype/cpog_checker.py
More lenient parsing in count mode

@@ -1082,6 +1082,7 @@ class Prover:
 
     # Find zero-terminated list of integers in fields (or single field consisting of '*').  Return (list, rest)
     # Flag error if something goes wrong
+    # In count-only mode, don't require terminator
     def findList(self, fields, starOk = False):
         ls = []
         rest = fields
@@ -1101,7 +1102,8 @@ class Prover:
                 return (ls, rest)
             ls.append(val)
             starOk = False
-        self.flagError("No terminating 0 found")
+        if not self.countMode:
+            self.flagError("No terminating 0 found")
         return (ls, rest)
 
     def prove(self, fname):
@@ -1164,7 +1166,7 @@ class Prover:
         if root is None:
             print("CHECKER: Can't determine count.  Don't know root")
             return P52()
-        return self.omgr.count(self.cmgr.actualRoot, weights)
+        return self.omgr.count(root, weights)
 
     def invalidCommand(self, cmd):
         self.flagError("Invalid command '%s' in proof" % cmd)
@@ -1259,10 +1261,11 @@ class Prover:
         except:
             self.flagError("Couldn't add operation with clause #%d: Non-integer arguments" % (id))
             return
-        if args[-1] != 0:
+        if args[-1] == 0:
+            args = args[:-1]
+        elif not self.countMode:
             self.flagError("Couldn't add operation with clause #%d: No terminating 0 found" % (id))
             return
-        args = args[:-1]
         (ok, msg) = self.omgr.addOperation(self.omgr.conjunction, args[0], args[1:], id)
         if not ok:
             self.flagError("Couldn't add operation with clause #%d: %s" % (id, msg))

### 2023-08-08 doc/pkc-notes.tex
Fix detail on partitioning

@@
 \newcommand{\ureduce}[2]{\forall #2\, #1}
-\newcommand{\partition}[2]{\mathit{Part}(#1, #2)}
 
@@
 \newcommand{\uprop}{\algo{Uprop}}
-\newcommand{\simpl}{\also{Simpl}}
+\newcommand{\partition}{\algo{Partition}}
+\newcommand{\occur}{\algo{Occur}}
 
@@
 
-\item[$\partition{\phi}{Y}$:] Find the subset of clauses that depend only on the variables in $Y$:
-$\partition{\phi}{Y} = \{ C \in \phi \mid \dependencyset(C) \subseteq Y\}$.
+\item[$\occur(\phi, Y)$:] Iteratively expand the set of variables $Y$ to include all those variables that co-occur in one or more clauses in $\phi$.
+  That is, starting with $Y' = Y$,
+  repeatedly compute $Y' \leftarrow Y' \cup \dependencyset(C)$
+  for every clause $C$ such that $Y' \cap \dependencyset(C) \not = \emptyset$,
+  until a fixed point is reached, and then return $Y'$.
+
+\item[$\partition(\phi, Y)$:] Find the subset of clauses that depend only on the variables in $\occur(Y)$:
+$\partition(\phi, Y) = \{ C \in \phi \mid \dependencyset(C) \subseteq \occur(Y)\}$.
 
@@
   \item Define clausal formula $\phi'$ to be $\bcp(\simplify{\phi}{L})$
-  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition{\phi'}{\dependencyset(\psi_i)})$.  (Partitioning is only required when $m > 1$.)
+  \item For $1 \leq i \leq m$, recursively compute $\psi'_i = \traverse(\psi_i, \partition(\phi', \dependencyset(\psi_i)))$.  (Partitioning is only required when $m > 1$.)
   \item Return the product (with simplifications) for the set of arguments $\{\lit \in L \mid  \dependencyset(\lit) \subseteq \dvarset\} \cup \{ \psi'_i \mid  1 \leq i \leq m \}$.

### 2023-08-08 doc/JAIR/jair-paper.tex
Updates

@@
 (det-DNNF) introduced by
-Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
+\citeA{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
 \emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
@@
 
-\begin{dfn}[Boolean Formulas]
+\begin{dfn}
   The set of Boolean formulas is defined recursively.  Each
@@
 
-\begin{dfn}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
+\begin{dfn}
+  \label{def:partitioned-operation-formula}
   A \emph{partitioned-operation formula}
@@
 
-\begin{dfn}[Commutative Ring]
+\begin{dfn}
   A \emph{commutative ring} $\ring$ is an algebraic structure
@@
 
-We describe the structural approach to proof generation as a recursive procedure
+The structural approach to proof generation takes the form of a recursive procedure
 $\validate(\nodeu, \passign, \psi)$ taking as arguments POG

### 2023-08-08 doc/pkc-notes.tex
Minor tweaks

@@
 
-\item[$\ureduce{\phi}{x}$:] Express the universal quantification of $\phi$ with respect to variable $x$ in clausal form, by removing all occurrences of $x$ and $\obar{x}$:
+\item[$\ureduce{\phi}{x}$:] Express the universal quantification of $\phi$ with respect to variable $x$ in clausal form.  This involves removing all occurrences of $x$ and $\obar{x}$:
 $\ureduce{\phi}{x} = \lfilter(\phi, \{x, \obar{x}\})$.

### 2023-08-11 cpog/prototype/cpog_checker.py
Cleanup

@@ -38,6 +38,14 @@ def usage(name):
 
 
 ######################################################################################
+# Configuration parameters
+######################################################################################
+
+## Need to render numbers with lots of digits
+sys.set_int_max_str_digits(10 * sys.get_int_max_str_digits())
+
+
+######################################################################################
 # Design options
 ######################################################################################
 

### 2023-08-11 pkc/q25.c
Small optimization

@@ -152,7 +152,7 @@ static uint32_t q25_div_word(int id, uint32_t divisor) {
 }
 
 /* Take out multiples of n, where n = 2^p2 * 5^p5, and n <= RADIX */
-static void q25_reduce_multiple(int id, uint32_t p2, uint32_t p5, uint32_t n) {
+static void old_q25_reduce_multiple(int id, uint32_t p2, uint32_t p5, uint32_t n) {
     uint32_t word;
     while ((word = digit_buffer[id][0])  % n == 0) {
 	int pwr = 0;
@@ -168,6 +168,32 @@ static void q25_reduce_multiple(int id, uint32_t p2, uint32_t p5, uint32_t n) {
 	working_val[id].pwr5 += p5*pwr;
     }
 }
+
+/* Take out multiples of n, where n = 2^p2 * 5^p5, and n <= RADIX */
+static void q25_reduce_multiple(int id, uint32_t p2, uint32_t p5, uint32_t n) {
+    uint32_t word;
+    while ((word = digit_buffer[id][0])  % n == 0) {
+	int pwr = 0;
+	uint64_t scale = 1;
+	uint64_t rradix = Q25_RADIX;
+	uint64_t rword = word;
+	// Try expanding to two words.  Allows extracting more powers of two
+	if (working_val[id].dcount > 1) {
+	    rradix *= Q25_RADIX;
+	    rword += (uint64_t) Q25_RADIX * digit_buffer[id][1];
+	}
+	uint64_t nscale = scale * n;
+	while (nscale <= Q25_RADIX && rradix % nscale == 0 && rword % nscale == 0) {
+	    pwr ++;
+	    scale = nscale;
+	    nscale *= n;
+	}
+	q25_div_word(id, scale);
+	working_val[id].pwr2 += p2*pwr;
+	working_val[id].pwr5 += p5*pwr;
+    }
+}
+
 
 /* Take out as many multiples of 10 as possible.  Assume nonzero */
 static void q25_reduce10(int id) {

### 2023-08-11 pkc/counters.h
More granular data

@@ -33,7 +33,8 @@ typedef enum {
     COUNT_POG_INITIAL_PRODUCT, COUNT_POG_INITIAL_SUM, COUNT_POG_INITIAL_EDGES,
     COUNT_POG_FINAL_PRODUCT, COUNT_POG_FINAL_SUM, COUNT_POG_FINAL_EDGES,
     COUNT_POG_PRODUCT, COUNT_POG_SUM, COUNT_POG_EDGES,
-    COUNT_VISIT_PRODUCT, COUNT_VISIT_DATA_SUM, COUNT_VISIT_MUTEX_SUM, COUNT_VISIT_EXCLUDING_SUM,
+    COUNT_VISIT_PARTITION_PRODUCT, COUNT_VISIT_LITERAL_PRODUCT, COUNT_VISIT_MIXED_PRODUCT,
+    COUNT_VISIT_DATA_SUM, COUNT_VISIT_MUTEX_SUM, COUNT_VISIT_EXCLUDING_SUM,
     COUNT_SAT_CALL, COUNT_SIMPLE_KC, COUNT_KC_CALL,
     COUNT_PKC_DATA_ONLY, COUNT_PKC_PROJECT_ONLY, COUNT_PKC_REUSE,
     COUNT_NUM

### 2023-08-11 pkc/pkc.cpp

@@ -86,13 +86,17 @@ static void stat_report(double elapsed) {
     lprintf("%s    KC POG AVG             : %.2f\n", prefix, pog_avg);
     lprintf("%s    KC POG MAX             : %d\n", prefix, pog_max);
 
-    lprintf("%s Node Traversal:\n", prefix);
-    int vp, vd, vm, ve;
-    lprintf("%s    Traverse Product       : %d\n", prefix, vp = get_count(COUNT_VISIT_PRODUCT));
-    lprintf("%s    Traverse Data Sum      : %d\n", prefix, vd = get_count(COUNT_VISIT_DATA_SUM));
-    lprintf("%s    Traverse Mutex Sum     : %d\n", prefix, vm = get_count(COUNT_VISIT_MUTEX_SUM));    
-    lprintf("%s    Traverse Excluding Sum : %d\n", prefix, ve = get_count(COUNT_VISIT_EXCLUDING_SUM));    
-    lprintf("%s    Traverse TOTAL         : %d\n", prefix, vp+vd+vm+ve);
+    lprintf("%s Node Traversals:\n", prefix);
+    int vpp, vpl, vpm, vp, vsd, vsm, vse, vs;
+    lprintf("%s         Literal Prod      : %d\n", prefix, vpl = get_count(COUNT_VISIT_LITERAL_PRODUCT));
+    lprintf("%s         Partition Prod    : %d\n", prefix, vpp = get_count(COUNT_VISIT_PARTITION_PRODUCT));
+    lprintf("%s         Mixed  Prod       : %d\n", prefix, vpm = get_count(COUNT_VISIT_MIXED_PRODUCT));
+    lprintf("%s       Total Product       : %d\n", prefix, vp = vpl+vpp+vpm);
+    lprintf("%s         Data Sum          : %d\n", prefix, vsd = get_count(COUNT_VISIT_DATA_SUM));
+    lprintf("%s         Mutex Sum         : %d\n", prefix, vsm = get_count(COUNT_VISIT_MUTEX_SUM));    
+    lprintf("%s         Excluding Su      : %d\n", prefix, vse = get_count(COUNT_VISIT_EXCLUDING_SUM));    
+    lprintf("%s       Total Sum           : %d\n", prefix, vs = vsd + vsm + vse);
+    lprintf("%s    Traverse TOTAL         : %d\n", prefix, vp+vs);
 
     lprintf("%s PKC Optimizations:\n", prefix);
[…]

### 2023-08-11 pkc/project.cpp

@@ -261,9 +261,10 @@ int Project::traverse_product(int edge) {
 	}
     }
     if (cedges.size() > 0) {
-	if (cedges.size() == 1) 
+	if (cedges.size() == 1) {
 	    nargs.push_back(traverse(cedges[0]));
-	else {
+	    incr_count(COUNT_VISIT_MIXED_PRODUCT);
+	} else {
 	    std::unordered_set<int> vset;
 	    cr->bcp(true);
 	    for (int cedge : cedges) {
@@ -274,14 +275,16 @@ int Project::traverse_product(int edge) {
 		nargs.push_back(traverse(cedge));
 		cr->pop_context();
 	    }
+	    incr_count(COUNT_VISIT_PARTITION_PRODUCT);
 	}
-    }
+    } else {
+	incr_count(COUNT_VISIT_LITERAL_PRODUCT);
+    }	
     cr->pop_context();
     pog->start_node(POG_PRODUCT);
     for (int cedge : nargs)
 	pog->add_argument(cedge);
     int nedge = pog->finish_node();
     report(5, "Traversal of Product node %d yielded edge %d.\n", edge, nedge);
-    incr_count(COUNT_VISIT_PRODUCT);
     return nedge;
 }

### 2023-08-11 doc/pkc-notes.tex
Update optimization info

@@
 \item[Product Operations:]
-  Let the operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\psi_i$ is either a sum or a  product operation.
+  Let the operation be of the form $P(\lit_1, \lit_2, \ldots, \lit_k, \psi_1, \psi_2, \ldots, \psi_m)$, where each $\lit_i$ is a literal,
+  and each $\psi_i$ is either a sum or a  product operation.
   \begin{enumerate}
@@
 \item Only data variables in $\psi$.  If $\dependencyset(\psi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \psi$.
-\item Only data variables in $\phi$.  If $\dependencyset(\phi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \kc(\phi)$.
-\item Only projection variables in $\phi$.  If $\dependencyset(\phi) \subseteq \pvarset$, then $\traverse(\psi, \phi) = \tautology$ if formula $\phi$ is satisfiable and $\nil$ otherwise.
-\item Cached result. If $\psi$ is a sum operation and some previous traversal had the same clausal formula $\phi$ as its second argument, then the previously computed result can be reused.\footnote{
-Caching for product operations is not, in general valid, since the two recursive calls made when processing a sum operation can be for product operations with the same clausal argument.}
-\item Pure literal reduction.
-Literal $\lit$ is said to be \emph{pure} in $\phi$ if some clause in $\phi$ contains $\lit$, and no clause in $\phi$ contains  $\obar{\lit}$.
-Let $\pure(\phi)$ denote the set of all pure literals in $\phi$, and $\pure_{\pvarset}(\phi) = \{ \lit \in \pure(\phi) \mid  \dependencyset(\lit) \subseteq \pvarset \}$.
-If $\pure_{\pvarset}(\phi) \not = \emptyset$, then $\phi$ can be reduced as
- $\phi \leftarrow \cfilter(\phi, \pure_{\pvarset})$.  Eliminating clauses can cause other literals to become pure, and so this reduction should be iterated until no pure literals remain.
+%%\item Only data variables in $\phi$.  If $\dependencyset(\phi) \subseteq \dvarset$, then $\traverse(\psi, \phi) = \kc(\phi)$.
+\item Only projection variables in $\psi$.  If $\dependencyset(\psi) \subseteq \pvarset$, then $\traverse(\psi, \phi) = \tautology$ if formula $\phi$ is satisfiable and $\nil$ otherwise.
+\item Cached result. If $\psi$ was an argument to a previous call to $\traverse$, then the previously computed result can be reused.
+%%\item Pure literal reduction.
+%%Literal $\lit$ is said to be \emph{pure} in $\phi$ if some clause in $\phi$ contains $\lit$, and no clause in $\phi$ contains  $\obar{\lit}$.
+%%Let $\pure(\phi)$ denote the set of all pure literals in $\phi$, and $\pure_{\pvarset}(\phi) = \{ \lit \in \pure(\phi) \mid  \dependencyset(\lit) \subsete%%q \pvarset \}$.
+%%If $\pure_{\pvarset}(\phi) \not = \emptyset$, then $\phi$ can be reduced as
+%% $\phi \leftarrow \cfilter(\phi, \pure_{\pvarset})$.  Eliminating clauses can cause other literals to become pure, and so this reduction should be iterate%%d until no pure literals remain.
 \end{itemize}
 
-The first two termination rules have some subtle differences.  For the
-unoptimized version of $\traverse$, the arguments will always satisfy
-$\dependencyset(\psi) \subseteq \dependencyset(\phi)$.  However, 
-pure literal reduction can eliminate some literals from $\phi$
-that remain in $\psi$, and so the two sets of variables can become
-incomparable.  When both rules are applicable, it is more efficient to use the first, since it avoids an invocation of $\kc$.
+%% The first two termination rules have some subtle differences.  For the
+%% unoptimized version of $\traverse$, the arguments will always satisfy
+%% $\dependencyset(\psi) \subseteq \dependencyset(\phi)$.  However, 
+%% pure literal reduction can eliminate some literals from $\phi$
+%% that remain in $\psi$, and so the two sets of variables can become
+%% incomparable.  When both rules are applicable, it is more efficient to use the first, since it avoids an invocation of $\kc$.
 

### 2023-08-11 pkc/pkc.cpp
Implement driver program

@@ -9,10 +9,11 @@
 
 
 void usage(const char *name) {
-    lprintf("Usage: %s [-h] [-k] [-v VLEVEL] [-O OPT] [-b BLIM] FORMULA.cnf [FORMULA.pog]\n", name);
+    lprintf("Usage: %s [-h] [-k] [-v VLEVEL] [-L LOG] [-O OPT] [-b BLIM] FORMULA.cnf [FORMULA.pog]\n", name);
     lprintf("  -h          Print this information\n");
     lprintf("  -k          Keep intermdiate files\n");
     lprintf("  -v VERB     Set verbosity level\n");
+    lprintf("  -L LOG      Record all results to file LOG\n");
     lprintf("  -O OPT      Select optimization level: 0 None, 1 +Simple KC, 2 +Reuse, 3 +Analyze vars, 4 +Pure literal\n");
     lprintf("  -b BLIM     Limit iterations of Boolean constraint propagation\n");
 }
@@ -146,7 +147,7 @@ int main(int argc, char *const argv[]) {
     int bcp_limit = 0;
     bool keep = false;
     int c;
-    while ((c = getopt(argc, argv, "hkv:O:b:")) != -1) {
+    while ((c = getopt(argc, argv, "hkv:L:O:b:")) != -1) {
 	switch (c) {
 	case 'h':
 	    usage(argv[0]);
@@ -156,6 +157,9 @@ int main(int argc, char *const argv[]) {
 	    break;
 	case 'k':
 	    keep = true;
+	    break;
+	case 'L':
+	    set_logname(optarg);
 	    break;
 	case 'O':
 	    optlevel = atoi(optarg);

### 2023-08-11 pkc/pkc_drive.py

@@ -1,0 +1,180 @@
+#!/usr/bin/python3
+
+
+#####################################################################################
+# Copyright (c) 2023 Randal E. Bryant, Carnegie Mellon University
+# Last edit: May 19, 2022
+# 
+# Permission is hereby granted, free of charge, to any person obtaining a copy of this software and
+# associated documentation files (the "Software"), to deal in the Software without restriction,
+# including without limitation the rights to use, copy, modify, merge, publish, distribute,
+# sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is
+# furnished to do so, subject to the following conditions:
+# 
+# The above copyright notice and this permission notice shall be included in all copies or
+# substantial portions of the Software.
+# 
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT
+# NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+# NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM,
+# DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT
+# OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+########################################################################################
+
+import getopt
+import sys
+import os.path
+import subprocess
+import datetime
+import time
+
+def usage(name):
+    print("Usage: %s [-h] [-f] [-v VERB] [-t TIME] [-l NFILE] [FILE.EXT ...]" % name)
+    print("  -h       Print this message")
+    print("  -f       Force regeneration of all files")
+    print("  -v       Set verbosity level")
+    print("  -t TIME  Limit time for each of the programs")
+    print("  -l NFILE Specify file containing root names")
+    print("  EXT can be any extension for wild-card matching (e.g., cnf, nnf)")
+
+# Defaults
+standardTimeLimit = 60
+
+verbLevel = 1
+force = False
+
+nameFile = None
+
+# Pathnames
+homePath = "/Users/bryant/repos"
+
+genHome = homePath + "/model-counting/pkc"
+genProgram = genHome + "/pkc"
+
+timeLimit = 4000
+
+commentChar = 'c'
+
+def trim(s):
[…]

### 2023-08-20 doc/JAIR/jair-paper.tex
Incorporated example + optimizations into paper

@@
 
-\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting}
+\title{Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting}
 
@@
 Our proofs construct a new propositional formula, and we must verify
-%both
+both
 that it
-%satisfies a set of rules and that it
+satisfies a set of rules and that it
 is equivalent to the input formula.  This requires verifying
@@
 models.  For $n = |\varset|$, scaling the density by $2^n$
-yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
-in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.
+yields the number of models.
 
@@
 either from different compilers or from a single compiler with different configuration parameters.
-
-%We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.
 
@@
 The final result is then $\rep(\noder, w)$.
+
+Kimmig's formulation of algebraic model
+counting~\cite{kimmig:jal:2017} is more general of ours.  It allows
+the algebraic structure to be a \emph{semiring}, i.e., there need not
+be additive inverses.  For example, it can have the sum operation be
+$\max$ and the product operation be addition over the real numbers plus $\infty$.  This more general form
+requires that the representation generated by the knowledge compiler
+obey a property known as \emph{smoothness}~\cite{darwiche:jair:2002}.
+Within our formulation, a partitioned-operation formula is smooth when
+all arguments to each sum operation have identical dependency sets.
+That is, every sum operation $\bigvee_{1 \leq i \leq k} \phi_i$ has
+$\dependencyset(\phi_i) = \dependencyset(\phi_j)$ for $1 \leq i < j \leq k$.
+Smoothness can be ensured by adding redundant formulas to
+artificially introduce variables.  For example, if subformula $\phi_i$
+lacks having variable $x$ in its dependency set, it can be replaced by
+$(x \por \obar{x}) \pand \phi_i$.  The consideration of smoothness is
+orthogonal to the issues addressed by this paper.  Given a smoothed
+dec-DNNF graph generated by a knowledge compiler, our toolchain will
+convert this into a smooth POG and verify its equivalence to the input
+formula.
 
@@
 
-The supplement~\cite{bryant:sat:2023:supplement},
-shows the CPOG description for
-an input formula with five clauses, yielding a POG with six
-nonterminal nodes.  It explains how the clause addition and deletion
-steps yield a proof of equivalence between the input formula and its POG
-representation.
+\subsection{CPOG Example}
+\label{sect:cpog:example}
+
+\begin{figure}
+\begin{minipage}{0.58\textwidth}
+(A)  Input Formula\\[1.2ex]
+\begin{tabular}{lll}
+\toprule
+\makebox[5mm]{ID} & \makebox[15mm]{Clauses} & \\
+\midrule
+1 & \texttt{-1 3 -4} & \texttt{0} \\
+2 & \texttt{-1 -3 4} & \texttt{0} \\
+3 & \texttt{3 -4} & \texttt{0}\\
+4 & \texttt{1 -3 4} & \texttt{0} \\
+5 & \texttt{-1 -2} & \texttt{0} \\
+\bottomrule
+\end{tabular}
+\\[1.8ex]
+(C) POG Declaration\\[1.2ex]
+\begin{tabular}{llll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
+\midrule
+6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
+9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
+12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
+15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\
+18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
+22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
+ & \texttt{r 10} && Root $r = s_{10}$\\
+\bottomrule
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.35\textwidth}
+(B) POG Representation \\
+\input{dd/eg4}
+\end{minipage}
+%% Break
+\\[2.5ex]
+\begin{minipage}{0.42\textwidth}
+(D) Defining Clauses\\[1.2ex]
+\begin{tabular}{llll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
+\midrule
+\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
+\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
+\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
+\midrule
+\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
+\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
+\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
+\midrule
+\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
+\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
+\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
+\midrule
+\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
+\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
+\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
+\midrule
+\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
+\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
+\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
+\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
+\midrule
+\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
+\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
+\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
+\bottomrule
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+(E) CPOG Assertions\\[1.2ex]
+\begin{tabular}{llllll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
+\midrule
+\texttt{25} & \texttt{a 5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
+\texttt{26} & \texttt{a 6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
+\texttt{27} & \texttt{a 3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
+\texttt{28} & \texttt{a 7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
+\texttt{29} & \texttt{a 8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
+\texttt{30} & \texttt{a 5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
+\texttt{31} & \texttt{a 6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
+\texttt{32} & \texttt{a 3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
+\texttt{33} & \texttt{a 7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
+\texttt{34} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
+\texttt{35} & \texttt{a 1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
+\texttt{36} & \texttt{a 10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
+\bottomrule
+\end{tabular}
+\\[1.5ex]
+(F) Input Clause Deletions\\[1.2ex]
+\begin{tabular}{lllll}
+  \toprule
+ \multicolumn{3}{l}{CPOG line} & Explanation\\
+\midrule
+ \texttt{d 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
+ \texttt{d 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
+ \texttt{d 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
+ \texttt{d 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
+ \texttt{d 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
+\bottomrule
+\end{tabular}
+\end{minipage}
+\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F)}
+\label{fig:eg4:all}
+\end{figure}
+
+Figure \ref{fig:eg4:all} illustrates an example formula and shows how
+the CPOG file declares its POG representation.  The input formula (A)
+consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
+$x_4$.  The generated POG (B) has six nonterminal nodes representing
+four products and two sums.  We name these by the node
+type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
+extension variable.
+% We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
+  The first part of the CPOG file (C) declares
+these nodes using clause IDs that increment by three or four,
+depending on whether the node has two children or three.  The last two
+nonzero values in the sum declarations are the hint providing the
+required mutual exclusion proof.
+
+\subsection{Node Declarations}
+
+We step through portions of the file to provide a better understanding of the CPOG proof framework.
+Figure
+\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
+defined by the POG operation declarations.  These do not appear in the
+CPOG file.  Referring back to the declarations of the sum nodes in
+Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
+$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
+\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
+$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
+have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
+clause IDs 16 and 19 as the hint.  These form a RUP proof for the clause
+$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
+$\nodes_{10}$ have disjoint models.
+
+\subsection{Forward Implication Proof}
+
+Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
+leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $\nodes_{10}$ is implied by the input clauses, i.e.,
+any total assignment $\assign$
+satisfying the input clauses must have $\assign(s_{10}) = 1$.
+Working backward, we can see that
+clause 29 indicates that variable $p_8$ will be implied by the input
+clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
+be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
+hint for clause 36.
+
+\subsection{Reverse Implication Proof}
+
+Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
+delete the input clauses.  Consider the first of these, deleting
+input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
+that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
+The proof proceeds by first assuming that the clause is false, requiring
+$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
+clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
+clauses that cause unit propagation.  Hint clauses 8 and 10 force the
+assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
+$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading,
+via clause 22, to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
+completing the RUP proof.  The deletion hints for the other input
+clauses follow similar patterns---they work from the bottom nodes of
+the POG upward, showing that any total assignment that falsifies the clause
+must assign $\assign(s_{10}) = 0$.
+
+Deleting the asserted clauses is so simple that we do not show it.  It
+involves simply deleting the clauses from clause number 35 down to
+clause number 25, with each deletion using the same hint as were used
+to add that clause.  In the end, therefore, only the defining clauses
+for the POG nodes and the unit clause asserting $s_{10}$ remain,
+completing a proof that the POG is logically equivalent to the input
+formula.
 
@@
 proof consists of a series of clause assertions leading to one adding
-$\{r\}$ as a unit clause.  We have devised two methods of generating this
+$\{r\}$ as a unit clause.  We have devised two methods for generating this
 proof.  The \emph{monolithic} approach makes just one call to a
@@
 
+The forward implication proof of Figure~\ref{fig:eg4:all}(E) was
+generated by the structural approach.  Working from step 36 backward,
+we can see that steps 35 and 36 complete the call to
+$\validate(\nodes_{10}, \emptyset, \phi_I)$.  This call used $x_1$ as
+the splitting variable, first calling $\validate(\nodep_{8},
+\{\obar{x}_1\}, \emptyset, \phi_I)$, which completed with step 29, and
+$\validate(\nodep_{9}, \{x_1\}, \phi_I)$, which completed with step
+34.  We see that each of these calls required separate traversals of
+nodes $\nodes_7$, $\nodep_6$, and $\nodep_5$, with the former yielding
+proof steps 25--27 and the latter yielding proof steps 30--32.  This
+demonstrates how our simple formulation of $\validate$ effectively
+expands the graph into a tree.  This shortcoming is avoided by the use of lemmas, as is described in Section~\ref{sect:lemma}.
+
 \subsection{Reverse Implication Proof}
@@
 representation of the formula; they serve only to enable the forward
-implication proof.  Here we summarize the two optimizations.
-The supplement~\cite{bryant:sat:2023:supplement} provides more details.
-
-\emph{Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
-having many---tens or even hundreds---of literals as children.  The
-earlier formulation of $\validate$ considers each literal $\lit \in \lambda$
-separately, calling a SAT solver for every literal that cannot be
-validated with BCP\@.  Literal grouping handles
-all of these literals together.  It defines product node $\nodev$
-having the literals as children.  The goal then becomes to prove that
-any total assignment must yield 1 for extension variable $v$.  Calling
-a solver with $v$ set to 0 yields an unsatisfiability proof that can
-be mapped back to a sequence of clause additions in the CPOG file validating
-all of the literals.
-
-\emph{Lemmas:} Our formulation of $\validate$ requires each call at a
-node $\nodeu$ to recursively validate all of its children.  This
-effectively expands the graph into a tree, potentially requiring an
-exponential number of recursive steps.  Instead, for each node
-$\nodeu$ having $\indegree(\nodeu) > 1$, the program can define and
-generate the proof of a \emph{lemma} for $\nodeu$ when it is first reached by
-a call to $\validate$ and then apply this lemma for this and
-subsequent calls.  The lemma states that the POG with root node
-$\nodeu$ satisfies forward implication for a formula
-$\gamma_{\nodeu}$, where some of the clauses in this formula are
-input clauses from $\inputformula$, but others are simplified
-versions of input clauses.
-The key idea is to introduce product nodes to encode (via DeMorgan's
-Laws) the simplified clauses and have these serve as lemma arguments.
+implication proof.
 
@@
 optimizations yield substantial benefits.
+
+\subsection{Literal Grouping}
+\label{sect:grouping}
+
+A single recursive step of $\validate$ can encounter product nodes
+having many literals as children.  The naive formulation of $\validate$
+considers each literal $\lit \in \lambda$ separately.
+Literal grouping allows all literals to be validated with a single call to a SAT solver.
+It collects those literals
+$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
+product node $\nodev$ having these literals as children.  The goal
+then becomes to prove that any total assignment must yield 1 for extension
+variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
+  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
+  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
+  input clauses and the defining clauses for $\nodev$ into the hints.
+
+\subsection{Lemmas}
+\label{sect:lemma}
+
+As we have noted, the recursive calling of $\validate$ starting at
+root $\noder$ effectively expands the POG into a tree, and this can
+lead to an exponential number of calls.
+These shared subgraphs arise when the knowledge compiler employs \emph{clause caching}
+to detect that the simplified set of
+clauses arising from one partial assignment to the literals matches that
+of a previous partial assignment~\cite{darwiche:aaai:2002}.
+When this dec-DNNF node is translated into POG
+node $\nodeu$, the proof generator can assume (and also check), that
+there is a simplified set of clauses $\gamma_{\nodeu}$
+for which the subgraph with root $\nodeu$ is its POG representation.
+
+The proof generator can exploit the sharing of subgraphs
+by constructing and proving a \emph{lemma} for each node
+$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
+total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ and the defining clauses for the POG must yield
+$\assign(u) = 1$.  This lemma is then invoked for every node having
+$\nodeu$ as a child.
+As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.
+
+The challenge for implementing this strategy is to find a way to
+represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CPOG file.  Some may be
+unaltered input clauses, and these can be used directly.  Others,
+however will be clauses that do not appear in the input formula.  We
+implement these by adding POG product nodes to the CPOG file to create
+the appropriate clauses.  Consider an \emph{argument} clause
+$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
+define a product node $\nodev$ with arguments
+$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, 
+we will introduce a defining clause
+$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
+  synthetic} clause having $\obar{v}$ as the \emph{guard literal}.
+That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
+  activate} the clause, causing it to represent argument clause $C$.  On the other
+hand, a partial assignment with $\passign(v) = 1$ will
+cause the clause to become a tautology and therefore have no effect.
+
+Suppose for every clause $C_j \in \gamma_{\nodeu}$ that does not correspond to
+an input clause, we generate a synthetic clause $C'_j$ with guard literal
+$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma'_{\nodeu}$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
+while input clauses in $\gamma_{\nodeu}$ are left unchanged.
+Let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
+Invoking $\validate(\nodeu, \lassign, \gamma'_{\nodeu})$
+ will then prove a lemma, given by the target clause
+ $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
+ showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u) = 1$.
+
+Later, when node $\nodeu$ is encountered by a call to $\validate(\nodeu, \passign, \psi)$, we invoke the lemma
+by showing that each synthetic clause
+$C_j$ matches some simplified clause in $\simplify{\psi}{\passign}$.  More precisely,
+for $1 \leq j \leq m$,
+we use clause addition to assert the clause
+$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$,
+showing that synthetic clause $C_j$ will be activated.
+Combining the lemma with these activations provides a derivation of the target clause for the call to $\validate$.
+
+Observe that the lemma structure can be hierarchical, since a shared
+subgraph may contain nodes that are themselves roots of shared
+subgraphs.  Nonetheless, the principles described allow the
+definition, proof, and applications of a lemma for each shared node in
+the graph.  For any node $\nodeu$, the first call to
+$\validate(\nodeu, \passign, \psi)$ may require further recursion,
+but any subsequent call can simply reuse the lemma proved by the first call.
+
+\subsection{Lemma Example}
+\label{app:lemma:eg}
+
+\begin{figure}
+(A) Additional nodes\\[1.0em]
+\begin{tabular}{llll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
+\midrule
+25 & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
+28 & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
+\bottomrule
+\end{tabular}
+\\[1.0em]
+(B) Implicit Clauses\\[1.2em]
+\begin{tabular}{llll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
+\midrule
+\texttt{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
+\texttt{26} & \texttt{-11 -3} & \texttt{0} & \\
+\texttt{27} & \texttt{-11 4} & \texttt{0} & \\
+\midrule
+\texttt{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
+\texttt{29} & \texttt{-12 3} & \texttt{0} & \\
+\texttt{30} & \texttt{-12 -4} & \texttt{0} & \\
+\bottomrule
+\end{tabular}
+\\[1.0em]
+(C) CPOG Assertions\\[1.0em]
+\begin{tabular}{llllll}
+\toprule
+\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
+\midrule
+\multicolumn{6}{l}{Lemma Proof} \\
+\texttt{31} & \texttt{a 5 11 12 3} & \texttt{0} & \texttt{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
+\texttt{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \texttt{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
+\texttt{33} & \texttt{a 3 7 11 12} & \texttt{0} & \texttt{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
+\texttt{34} & \texttt{a 7 11 12} & \texttt{0} & \texttt{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
+\midrule
+\multicolumn{6}{l}{Lemma Application \#1} \\
+\texttt{35} & \texttt{a -11 1} & \texttt{0} & \texttt{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
+\texttt{36} & \texttt{a -12 1} & \texttt{0} & \texttt{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
+\texttt{37} & \texttt{a 7 1} & \texttt{0} & \texttt{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
+\midrule
+\texttt{38} & \texttt{a 8 1} & \texttt{0} & \texttt{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
+\midrule
+\multicolumn{6}{l}{Lemma Application \#2} \\
+\texttt{39} & \texttt{a -11 -1} & \texttt{0} & \texttt{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
+\texttt{40} & \texttt{a -12 -1} & \texttt{0} & \texttt{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
+\texttt{41} & \texttt{a 7 -1} & \texttt{0} & \texttt{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
+\midrule
+\texttt{42} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
+\texttt{43} & \texttt{a 1 10} & \texttt{0} & \texttt{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
+\texttt{44} & \texttt{a 10} & \texttt{0} & \texttt{43 24 42} & \texttt{0} & $s_{10}$ \\
+\bottomrule
+\end{tabular}
+\caption{Example of lemma definition, proof, and application}
+\label{fig:eg4:lemmas}
+\end{figure}
+
+Figure~\ref{fig:eg4:lemmas} shows an alternate forward implication
+proof for the example of Figure~\ref{fig:eg4:all} using a lemma to
+represent the shared node $\nodes_7$.  We can see that the POG with
+this node as root encodes the Boolean formula $x_3 \leftrightarrow x_4$, having a CNF representation consisting of the clauses
+$\{x_3 ,\, \obar{x}_4\}$ and $\{\obar{x}_3 ,\, {x}_4\}$.  The product node
+declarations shown in Figure~\ref{fig:eg4:lemmas}(A) create synthetic
+clauses 25 and 28 to encode these arguments with activating literals
+$\obar{v}_{11}$ and $\obar{v}_{12}$, respectively.  Clauses 31--34
+then provide a proof of the lemma, stating that any assignment
+$\assign$ that activates these clauses will  assign $1$ to $s_7$.
+Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
+will cause the first synthetic clause to activate due to input clause
+3, and it will cause the second synthetic clause to activate due to
+input clause 4.  From this, clause 37 can use the lemma to state that
+assigning $0$ to $x_1$ will cause $s_7$ to evaluate to $1$.  Similarly,
+clauses 39 and 40 serve to activate the synthetic clauses when
+$\assign(x_1) = 1$, due to input clauses 1 and 2, and clause 41 then
+uses the lemma to state that assigning $1$ to $x_1$ will cause $s_7$ to
+evaluate to $1$.
+
+In this example, adding the lemma increases the proof length, but that
+is only because it is such a simple formula.
 

### 2023-08-20 doc/JAIR/dd/dd.py

@@ -1,0 +1,728 @@
+# Support generation of decision diagram images
+
+import math
+
+# Process files containing basic drawing commands.
+# Each command has combination of positional and keyword arguments
+
+# DDs drawn on grid with (0,0) at lower, lefthand corner
+# Nominal distance between two levels = 100
+# Default radius of branch or leaf node is 25.  Root node=5
+
+# General rule:
+#   Only need to give as much of a keyword or value as required to disambiguate it
+
+# Commands
+# Node
+#    n ID X Y label=TTT type=branch radius=25/5 fill=background edge=neutral [only=XX]
+#       ID is unique alphanumeric identifier
+#       TTT is something in TeX (with no spaces)
+#       type options: branch, root, leaf, empty
+
+
+# Tag
+#    t ID TTT position=above offset=10 [only=XX]
+#       ID identifies node to tag
+#       TTT is something in TeX.
+#       Position options: above, below, left, right
+
+# Edge
+#    e FID TID value=neutral mark=none radius=10 arrow=none[only=XX]
+#       FID and TID are Ids of source and destination nodes
+#       value options: neutral, high, low, path
+#       mark options: none, bubble   
+#       arrow options: none, to, from, both
+
+# Graphic primitives
+#
+# Rectangle
+#    r LLX LLY URX URY options=""  [only=XX]
+#       Options can be any text enclosed in square brackets in Tikz.  Use ~ instead of space
+
+# Colors
+#    Either alter existing color definition or define new one
+#    Existing (prefixes are sufficient):
+#      neutralcolor: outline for nodes, default for edges, fill for roots
+#      highcolor: High edge
+#      lowcolor: Low edge
+#      pathcolor: User-defined
+#      fillcolor: Fill in leaf & branch nodes
+#      backgroundcolor: Fill for rectangles
+#    VALUE: R:G:B values (each integer from 0 to 255)
+
+def trim(s):
+    while len(s) > 0 and s[-1] == '\n':
+        s = s[:-1]
+    return s
+
+class NoMatchException(Exception):
+    
+    prefix = ""
+
+    def __init__(self, prefix):
+        self.prefix = prefix
+
+    def __str__(self):
+        msg = "Match exception.  No keyword matching prefix '%s'" % self.prefix
+        return msg
+
+class MultiMatchException(Exception):
+
+    prefix = ""
+    def __init__(self, prefix):
+        self.prefix = prefix
+
+    def __str__(self):
[…]

### 2023-08-20 doc/JAIR/jair-paper.tex
More progress

@@
 
-We have developed an online supplement to this
-paper~\cite{bryant:sat:2023:supplement} that includes
-a worked example, more details on the algorithms,
-and extensive experimental results.
+This paper is an extended version of one published at the
+2023 Conference on the Theory and Application of Boolean Satisfiability~\cite{bryant:sat:2023}.
 

### 2023-08-22 doc/JAIR/jair-paper.tex
Tighten up notation

@@
 \newcommand{\ring}{{\cal R}}
+\newcommand{\semiring}{{\cal S}}
 \newcommand{\dset}{{\cal A}}
 \newcommand{\rep}{\textbf{R}}
+\newcommand{\srep}{\textbf{S}}
 \newcommand{\radd}{+}
@@
 \newcommand{\ifandonlyif}{\Leftrightarrow}
+\newcommand{\eequiv}{\ifandonlyif_{\varset}}
 %\newcommand{\drational}{\mathbb{Q}_{2,5}}
@@
 
+\newcommand{\extend}[1]{#1^{*}}
 \newcommand{\assign}{\alpha}
+\newcommand{\eassign}{\extend{\alpha}}
 \newcommand{\passign}{\rho}
@@
 concepts of clausal proof systems.  These factors were important in enabling formal verification.
-In Section~\ref{sect:experimental} and the supplement~\cite{bryant:sat:2023:supplement},
+In Section~\ref{sect:experimental},
 we also compare the performance of our toolchain to these other two.
@@
 Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
-\begin{displaymath}
-\begin{array}{rcl}
-\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \\
+\begin{eqnarray}
+\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation}\\
 \textstyle
-\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \\
+\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product}\\
 \textstyle
-\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \\
-\end{array}
-\end{displaymath}
+\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum} 
+\end{eqnarray}
 \end{prop}
@@
 $\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
-Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.
+Each such formula has a set of models $\modelset(\phi_{\nodeu})$.
 
@@
   sum node $\nodeu$ has exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
-  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
+  any total assignment $\assign \in  \modelset(\phi_{\nodeu_b})$ has $\assign(x)=b$, for $b \in \{0,1\}$.
 \end{itemize}
@@
 
-Kimmig's formulation of algebraic model
-counting~\cite{kimmig:jal:2017} is more general of ours.  It allows
-the algebraic structure to be a \emph{semiring}, i.e., there need not
-be additive inverses.  For example, it can have the sum operation be
-$\max$ and the product operation be addition over the real numbers plus $\infty$.  This more general form
-requires that the representation generated by the knowledge compiler
-obey a property known as \emph{smoothness}~\cite{darwiche:jair:2002}.
-Within our formulation, a partitioned-operation formula is smooth when
-all arguments to each sum operation have identical dependency sets.
-That is, every sum operation $\bigvee_{1 \leq i \leq k} \phi_i$ has
-$\dependencyset(\phi_i) = \dependencyset(\phi_j)$ for $1 \leq i < j \leq k$.
-Smoothness can be ensured by adding redundant formulas to
-artificially introduce variables.  For example, if subformula $\phi_i$
-lacks having variable $x$ in its dependency set, it can be replaced by
-$(x \por \obar{x}) \pand \phi_i$.  The consideration of smoothness is
-orthogonal to the issues addressed by this paper.  Given a smoothed
-dec-DNNF graph generated by a knowledge compiler, our toolchain will
-convert this into a smooth POG and verify its equivalence to the input
-formula.
-
 \section{Clausal Proof Framework}
-
-We write (possibly subscripted) $\theta$ for
-formulas encoded as clauses, possibly with extension variables.
-We write (possibly subscripted)
+\label{sect:clausal}
+
+A proof in our framework consists of a sequence of clause addition and
+deletion steps, with each step preserving the set of solutions to the
+original formula.  The status of the proof at any step is represented
+as a set of \emph{active} clauses $\theta$, i.e., those that have been
+added but not yet deleted.  Our framework is based on \emph{extended}
+resolution~\cite{Tseitin:1983}, where proof steps can introduce new
+\emph{extension variables} encoding Boolean formulas over input and
+prior extension variables.  That is, each extension variable $z$ is
+introduced via a set of \emph{defining clauses} that encode a formula
+$z \ifandonlyif F$, where $F$ is a Boolean formula over a subset of the input
+variables $\varset$ and previously defined extension variables $\exvarset$.
+We write $\theta$ for
+formulas encoded as clauses, possibly with extension variables, and
 $\phi$ for formulas
@@
 
-A proof in our framework consists of a sequence of clause addition
-and deletion steps, with each step preserving the set of solutions to
-the original formula.
-The status of the proof at any step is represented as
-a set of \emph{active} clauses $\theta$, i.e., those that
-have been added but not yet deleted.
-Our framework is based
-on \emph{extended} resolution~\cite{Tseitin:1983}, where proof
-steps can introduce new \emph{extension variables} encoding Boolean formulas over input and prior extension variables.
-Let $Z$
+Let $\exvarset(\theta)$
 denote the set of extension variables occuring in formula $\theta$.
+For any assignment $\assign$ to the variables in $\varset$, the defining
+clauses induce a unique assignment $\eassign$ to the variables in $\varset
+\cup \exvarset(\theta)$.  For Boolean formula $\phi$ over variables
+$\varset$ and clausal
+formula $\theta$ over the variables $\varset \cup \exvarset(\theta)$, we say that $\phi$ is {\em
+  effectively equivalent} to $\theta$, written
+$\phi \eequiv \theta$,
+when for any assignment
+$\assign$ to the variables in $\varset$, assignment $\assign$ is a model of $\phi$ if
+and only if its extension $\eassign$ is a model of $\theta$.
 Starting with  $\theta$ equal to input formula $\inputformula$,
 the proof must maintain the invariant that
-$\inputformula \ifandonlyif \exists Z\,\theta$.
+$\inputformula \eequiv \theta$.
 
 Clauses can be added in two different ways.  One is when they serve as
-the \emph{defining clauses} for an extension variable.  This form
+the defining clauses for an extension variable.  This form
 occurs only when defining $\pand$ and $\por$ operations, as is
@@
   The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
-  Declaring a node $\nodeu$ implicitly adds an \emph{extension} variable $u$ and a set of \emph{defining} clauses $\theta_{u}$
+  Declaring a node $\nodeu$ implicitly adds an extension variable $u$ and a set of defining clauses $\theta_{u}$
   encoding the product or sum operation.
@@
 
-The defining clauses for a product or sum
-operation uniquely define the value of its extension variable for any assignment of values to the argument variables.
-For the
-extension variable $u$ associated with any POG node $\nodeu$, we can therefore
-prove that any total assignment $\assign$ to the input variables that
-also satisfies the POG defining clauses will
-assign a value to $u$ such that $\assign(u) =
-1$ if and only if $\assign \in \modelset(\nodeu)$.
+As was described in Section~\ref{sect:clausal}, the defining clauses
+for the product and sum operations uniquely define the values of their
+extension variables for any assignment of values to the argument
+variables.  That is, for assignment $\assign$ to the variables in
+$\varset$, the defining clauses induce a unique assignment $\eassign$
+to all data and extension variables.
+Every POG node $\nodeu$ represents POG formula $\phi_{\nodeu}$ and has an associated extension variable $u$.
+We can prove that
+for any total assignment $\assign$ to the input variables, we will have
+$\eassign(u) = 1$ if and only if $\assign \in \modelset(\phi_\nodeu)$.
 
@@
 The sequence of operator declarations, asserted clauses, and
-clause deletions represents a systematic transformation of the input formula
+clause deletions represents a systematic transformation of the input formula $\inputformula$
 into a POG\@.  Validating all of these steps serves to prove that
@@
 \end{enumerate}
-In other words, at the end of the proof it must hold that the active clauses be exactly those
-in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
-of unit clause $\{r\}$ and the defining clauses for the nodes, providing a Tseitin encoding of $P$. Recognizing that
-any total assignment to the input variables implicitly defines the assignments to the extension variables,
-we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
-Let $\inputformula$ denote the input formula.
+In other words, at the end of the proof it must hold that the active
+clauses be exactly those in $\pogformula \doteq \{\{r\}\} \cup \;
+\bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting of unit
+clause $\{r\}$ and the defining clauses for the nodes, providing a
+Tseitin encoding of $P$.  By our invariant, we are guaranteed that
+$\inputformula \eequiv \pogformula$.  That is, for any total
+assignment $\assign$ to the input variables, its unique extension 
+$\eassign$ to the POG variables will satisfy $\eassign(r) = 1$ if and only if
+$\assign \in \modelset(\inputformula)$.
+
 The sequence of clause addition steps provides a \emph{forward implication} proof that
-$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
-assignment $\assign$ satisfying the input formula must also satisfy
+$\assign \in \modelset(\inputformula) \imply \eassign(r) = 1$.  That is, any total
+assignment $\assign$ satisfying the input formula must, when extended, also satisfy
 the formula represented by the POG\@.
 Conversely,
-each proof step that deletes an input clause proves that any
-total assignment $\alpha$ that falsifies the clause must
-falsify $\phi_\noder$.  Deleting all but the final asserted clause and all input clauses provides a \emph{reverse implication} proof
-that
-$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.
+the sequence of clause deletion steps that delete all intermediate added clauses and all input clauses
+provides a \emph{reverse implication} proof:
+$\eassign(r) = 1 \imply \assign \in \modelset(\inputformula)$.
+It does so by contradiction, proving that when $\eassign(r) = 0$, we must have $\assign \not \in \modelset(\inputformula)$.
 
@@
 any total assignment $\assign$
-satisfying the input clauses must have $\assign(s_{10}) = 1$.
+satisfying the input clauses must have its extension to $\eassign$ yield $\eassign(s_{10}) = 1$.
 Working backward, we can see that
@@
 input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
-that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
+that there is no total assignment $\assign$ that falsifies this clause but extends to $\eassign$ such that  $\assign(s_{10}) = 1$.
 The proof proceeds by first assuming that the clause is false, requiring
-$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
-clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
+$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consists of unit
+clauses (e.g., clause 36 asserting that $\eassign(p_{10}) = 1$) or
 clauses that cause unit propagation.  Hint clauses 8 and 10 force the
-assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
-$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading,
-via clause 22, to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
+assignments $\eassign(p_5) = \eassign(p_6) = 0$.  These, plus hint clause 12 force
+$\eassign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\eassign(p_8) = \eassign(p_9) = 0$, leading,
+via clause 22, to $\eassign(s_{10}) = 0$.  But this contradicts clause 36,
 completing the RUP proof.  The deletion hints for the other input
@@
 the POG upward, showing that any total assignment that falsifies the clause
-must assign $\assign(s_{10}) = 0$.
+must, when extended, have $\eassign(s_{10}) = 0$.
 
@@
 show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
-that any total assignment $\assign$ that is a model of $\inputformula$
-and the POG definition clauses
-will yield $\assign(r) = 1$, for root literal $r$.  This part of the
+that any total assignment $\assign$ that is a model of $\inputformula$ will extend to assignment $\eassign$
+such that $\eassign(r) = 1$, for root literal $r$.  This part of the
 proof consists of a series of clause assertions leading to one adding
@@
 assignment $\assign$ such that $\passign \subseteq \assign$
-will assign $\assign(u) = 1$.
+will extend to assignment $\eassign$ such that $\eassign(u) = 1$.
 The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
@@
 graph.  The RUP deletion proof for input clause $C$ effectively proves that any
-total assignment $\assign$ that satisfies the POG definition clauses
-but does not satisfy $C$ will yield
-$\assign(r) = 0$.  It starts with the set of literals
+total assignment $\assign$ that does not
+satisfy $C$ will extend to assignment $\eassign$ such that
+$\eassign(r) = 0$.  It starts with the set of literals
 $\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
@@
 Working upward through the graph, node $\nodeu$ is \emph{marked} when
-the collected set of literals forces $u$ to evaluate to $0$.  When marking $\nodeu$, the
+the collected set of literals forces $\eassign(u) = 0$.  When marking $\nodeu$, the
 program adds $\obar{u}$ to the RUP literals and adds the appropriate
@@
 (assuming the reverse implication holds) will attempt to mark root
-node $\noder$, which would require $\assign(r) = 0$, yielding a
+node $\noder$, which would require $\eassign(r) = 0$, yielding a
 conflict.
@@
 will cause further recursive calls only once.  Our experimental
-results~\cite{bryant:sat:2023:supplement} show that these
+results show that these
 optimizations yield substantial benefits.
@@
 product node $\nodev$ having these literals as children.  The goal
-then becomes to prove that any total assignment must yield 1 for extension
-variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
+then becomes to prove that any total assignment $\assign$ consistent with the partial assignment $\passign$,
+must, when extended to $\eassign$, yield $\eassign(v) = 1$.
+A single call to the solver can generate this proof by invoking it on the formula
   $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
@@
 $\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
-total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ and the defining clauses for the POG must yield
-$\assign(u) = 1$.  This lemma is then invoked for every node having
+total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ must extend to assignment $\eassign$ such that
+$\eassign(u) = 1$.  This lemma is then invoked for every node having
 $\nodeu$ as a child.
@@
  $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
- showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u) = 1$.
+ showing that any total assignment $\assign$ that activates the synthetic clauses will cause $u$ to be assigned $1$.
+ More precisely, given assignment $\assign$ and its extension $\eassign$, if $\eassign(v_j) = 0$ for every guard literal $\obar{v}_j$, then $\eassign(u) = 1$.
 
@@
 
-We summarize the results of our experiments here.
-The supplement~\cite{bryant:sat:2023:supplement}
-provides a more complete description. For our
+For our
 evaluation, we used the public benchmark problems from the 2022
@@
 
-\section{Conclusions}
+\section{Concluding Remarks}
 \label{sect:future}
@@
 
+\subsection{Generalizing to Semirings}
+
+Kimmig's formulation of algebraic model
+counting~\cite{kimmig:jal:2017} is more general than ours.  It allows
+the algebraic structure to be a \emph{semiring}.  A commutative
+semiring $\semiring$ obeys all properties of a commutative ring, except that the
+elements of the set need not have additive inverses.
+We can then define the {\em semiring evaluation problem} as computing
+  \begin{equation}
+    \begin{array}{rcl}
+    \srep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:srep}
+    \end{array}
+  \end{equation}
+  where sum $\sum$ is computed according to the semiring addition operation $\radd$ and product $\prod$
+  is computed according to the semiring product operation $\rmul$.
+
+  As an example, for $|X| = n$, consider the algebraic structure
+  $\langle \{0, 1, \ldots, n\}, \max, +, 0, 0 \rangle$.
+For each variable $x$, assign weights
+$w(x) = 1$ and $w(\obar{x}) = 0$.  Then the
+weight of an assignment is the number of variables set to 1, and an
+evaluation of (\ref{eqn:srep}) gives the maximum weight of all
+satisfying assignments.
+
+Semiring evaluation can be performed via knowledge compilation
+by requiring 
+that the representation generated by
+the compiler obeys a property known as
+\emph{smoothness}~\cite{darwiche:jair:2002,shih:nips:2019}.  Within our formulation,
+a partitioned-operation formula is smooth when all arguments to each
+sum operation have identical dependency sets.  That is, every sum
+operation $\bigvee_{1 \leq i \leq k} \phi_i$ has
+$\dependencyset(\phi_i) = \dependencyset(\phi_j)$ for $1 \leq i < j
+\leq k$.  Smoothness can be ensured by adding redundant formulas to
+artificially introduce variables.  For example, if subformula $\phi_i$
+lacks having variable $x$ in its dependency set, it can be replaced by
+$(x \por \obar{x}) \pand \phi_i$.  When a knowledge compiler generates
+a representation that is both in normal form and is smooth, then a semiring evaluation of the formula
+can proceed by first assigning each literal $\lit$ the value $w(\lit)$.
+Then the product and sum operations are evaluated in manners analogous to
+(\ref{eqn:ring:product}) and
+(\ref{eqn:ring:sum}).
+
+Our POG representation can support evaluation of semiring formulas by
+imposing the restriction that the POG is in normal form and that is smooth.
+Given a smoothed
+dec-DNNF graph generated by a knowledge compiler, our toolchain will
+convert this into a smooth POG in normal form and verify its
+equivalence to the input formula.  Full verification would also require checking that the POG is smooth.
+
 %% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and

### 2023-08-22 doc/JAIR/jair-data.tex
Add experimental data

@@
      \begin{itemize}
-     \item Simplifications via constant propagation
-       \begin{itemize}
-       \item Number of POG nodes around ranged between $0.35\times$
-         and $1.0\times$ number of dDNNF nodes (harmonic mean =
-         $0.81\times$).  Greater optimizations with smaller
-         benchmarks.
-       \end{itemize}
      \item Used heuristics to identify root node, and then put in topological order with root node at end.
@@
        \item $D$ also equals sum of number of edges and number of nodes
-       \item For 122 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
+       \item For 122 instances, Min = 304, Median = 774,883, Mean = 72,389,514, Max = 1,765,743,261
        \end{itemize}
@@
        \item Defining clauses: Min = 304, Max = 1,765,743,261, Median = 774,883
+       \item Tree Sizes ($D_t$): Min = 304, max = 14,924,670,455,484, median = 7,669,582.
        \item Tree ratio: Min = 1.0, Max = 52,409.7, Median = 11.61, Harmonic Mean = 3.21
-       \item Tree Sizes ($D_t$): Min = 304, max = 14,924,670,455,484, median = 7,669,582.
        \end{itemize}
@@
 
+\section*{Summary Result}
+
+Paper describes two ways to generate the forward implication proof:
+structured and monolithic.  Can also combine the two: at any point in
+the traversal during structured generation, the forward implication
+proof for the entire subgraph can be generated by a single call to a proof-generating
+SAT solver.  We'll refer to this as a \emph{hybrid} approach.
+
+Ran some experiments on all but the largest benchmarks (presented
+later) seeing under what conditions was it best to use a monolithic,
+structured, or hybrid approach, and where the hybrid approach should
+shift from structured to monolithic.
+
+General findings:
+\begin{itemize}
+\item The tree ratio provides a reasonable discriminator for monolithic vs.~structured.  Based on experiments, set threshold at 5.0. Use monolithic below this and structured above.
+\item We could not find any way to exploit the hybrid approach for these benchmarks.
+\item On the other hand, the monolithic approach becomes impractical beyond around $10^7$ defining clauses.  Even the best SAT solvers become unstable beyond this.
+\item Therefore, use hybrid approach for very large problems with tree ratios below 5.0, shift to monolithic once the tree size is at most $10^7$.
+\end{itemize}
+
+Our summary data is based on this set of rules.
+
+Summary:
+
+\begin{itemize}
+\item Of the 122 benchmarks, 100 were completed using the hybrid approach within 10,000 seconds.
+\end{itemize}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.70]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
+      legend style={at={(0.30,0.96)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.4,
+                              xmode=log,
+                              xmin=0.01,xmax=4000,
+                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
+                              ymin=0.01, ymax=29000,
+                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
+                              xlabel={D4 runtime (seconds)}, ylabel={CPOG generation and checking runtime (seconds)},
+%                              title={D4 Defining Clause Generation}
+            ]
+    \input{data-formatted/seconds-summary}
+%    \input{data-formatted/seconds-summary-onesided}
+    \input{data-formatted/seconds-summary-failures}
+    \legend{
+      \scriptsize \textsf{Full validation},
+%      \scriptsize \textsf{One-sided validation},
+      \scriptsize \textsf{Resource limit exceeded},
+    }
+    \addplot[mark=none, dashed] coordinates{(0.01,25000) (4000, 25000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,20000) (4000,20000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (2000,20000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (200, 20000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (20, 20000)};
+    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
+    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
+    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
+    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
+    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
+       \end{axis}
+
+\end{tikzpicture}
+} % centering
+\caption{Combined runtime for CPOG proof generation and checking  as function of D4 runtime.  Timeouts are shown as points on the dashed line.}
+\label{fig:d4:cpog}
+\end{figure}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.70]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10},
+      legend style={at={(0.92,0.28)}},
+      legend cell align={left},
+                              x post scale=1.9, y post scale=2.10,
+                              xmode=log,xmin=100,xmax=1e10, 
+                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, 
+                              xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$, $10^{10}$},
+                              ymode=log, ymin=100, ymax=1800000000, 
+                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000},
+                              yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
+                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
+            ]
+    \input{data-formatted/clauses-summary}
+%%    \input{data-formatted/clauses-defining-onesided}
+    \input{data-formatted/clauses-summary-failures}
+
+    \legend{
+      \scriptsize \textsf{Full validation},
+%      \scriptsize \textsf{One-sided validation},
+      \scriptsize \textsf{No validation},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,1e9) (1e10,1e9)};
+    \addplot[color=black,dashed] coordinates{(1e2,1.5e9) (1e10,1.5e9)};
+    \node[left] at (axis cs: 900,1200) {$1\times$};
+    \node[left] at (axis cs: 900,12000) {$10\times$};
+    \node[left] at (axis cs: 900,120000) {$100\times$};
+    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
+    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Total number of clauses in CPOG file as function of number of defining clauses}
+\label{fig:defining:total}
+\end{figure}
+
 \section*{Selecting Reduced Benchmark Set}
@@
 To make the experiments tractable, want to select set of benchmarks that are more tractable.
-Limit set to the 92 with at most $10^7$ defining clauses.
+Limit set to the 90 with at most $10^7$ defining clauses, and for which D4 runs in at most 1000 seconds.
 
-D4 Performance (92 instances):
+D4 Performance (90 instances):
        \begin{itemize}
-       \item Times: Min = 0.012 secs, Max = 1672.9 secs, Median = 1.85 secs
+       \item Times: Min = 0.012 secs, Max = 660.2 secs, Median = 1.72 secs
        \item Defining clauses: Min = 304, Max = 8,493,275, Median = 378,325
-       \item Tree ratio: Min = 1.0, Max = 363.9, Median = 9.37, Harmonic Mean = 2.87
        \item Tree Sizes ($D_t$): Min = 304, max = 1,205,743,897, median = 2,690,776
+       \item Tree ratio: Min = 1.0, Max = 363.9, Median = 9.98, Harmonic Mean = 2.94
        \end{itemize}
@@
 Our implementation also allows a hybrid of these approaches: at any
-point in the traversal, it can generate a proof for a node via a
+point in the structured traversal, it can generate a proof for a node via a
 single call to a SAT solver.  This capability supports both full
@@
 
-Our first set of experiments applies full monolithic and full structured generation to the reduced benchmark set.  We set a time limit of 1000 seconds for the proof generation.
+Our first set of experiments applies full monolithic and full structured generation to the reduced benchmark set.  
+After some experimenting, we devised the following ``hybrid'' method: when the tree ratio is at most $5.0$, use the monolithic approach, otherwise use the structured approach.
+We set a time limit of 1000 seconds for the proof generation and checking.
+Overall, our results on the 92 benchmarks were:
+\begin{itemize}
+\item Monolithic: 83 completed
+\item Structured: 82 completed
+\item Hybrid: 85 completed
+\end{itemize}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.70]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
+      legend style={at={(0.30,0.96)}},
+      legend cell align={left},
+                              x post scale=1.8, y post scale=2.0,
+                              xmode=log,xmin=0.01,xmax=1000,
+                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
+                              ymin=0.01, ymax=2000,
+                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
+                              xlabel={D4 runtime (seconds)}, ylabel={Toolchain runtime (seconds)},
+            ]
+
+    \input{data-formatted/sub10M-seconds-hybrid}
+    \legend{
+      \scriptsize \textsf{Hybrid},
+    }
+
+    \input{data-formatted/sub10M-seconds-failures-hybrid}
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Total runtime for toolchain using three different approaches to forward-implication proof generation, as function of D4 runtime}
+\label{fig:d4:cpog:compare}
+\end{figure}
 

### 2023-08-22 tools/merge_csv.py
More features

@@ -16,10 +16,12 @@ def eprint(s):
 
 
 def usage(name):
-    eprint("Usage: %s [-h] [-f]  [-s I1:I2:...:Ik] [-r I1:I2] [-l L0,L1,L2,...] FILE1.csv FILE2.csv ... FILEn.csv" % name)
+    eprint("Usage: %s [-h] [-f]  [-s I1:I2:...:Ik] [-r I1:I2] [-t I:T1:T2] [-o I:T] [-l L0,L1,L2,...] FILE1.csv FILE2.csv ... FILEn.csv" % name)
     eprint("  -f            Filter out lines that have at least one field missing")
-    eprint("  -s I1:I2:...:Ik Sum the values from specified files 1..n and add as new column")
-    eprint("  -r I1:I2      Form the ratio between the values from specified files 1 and 2 and add as new column")
+    eprint("  -s I1:I2:...:Ik Sum the values from specified columns 1..k and add as new column")
+    eprint("  -r I1:I2      Form the ratio between the values from specified columns 1 and 2 and add as new column")
+    eprint("  -t I:T1:T2    Compare value from column with value T1.  If greater, set to T2. Add as new column")
+    eprint("  -o I:T        Compare value from column with value T1.  If greater, omit entire line.  If I negative, invert sense of test")
     eprint("  -h            Print this message")
     eprint("  -l LABELS     Provide comma-separated set of heading labels")
     eprint("  FILE1.csv ... Source files")
@@ -136,8 +138,52 @@ def divideEntries(ratioList):
         fields.append(rval)
     globalCount += 1
 
+def omitEntries(omitTuple):
+    global globalEntries, globalCount
+    idx, thresh = omitTuple
+    gt = True
+    if idx < 0:
+        gt = False
+        idx = -idx
+    omitList = []
+    for k in globalEntries.keys():
+        fields = globalEntries[k]
+        try:
+            sfield = fields[idx-1]
+        except:
+            print("Couldn't get field #%d from line with key %s.  Fields: %s" % (idx, k, str(fields)))
+            sys.exit(1)
+        try:
+            vfield = float(sfield)
+        except:
+            print("Couldn't convert value '%s' to float.  Fields: %s" % (sfield, str(fields)))
+            sys.exit(1)
+        if gt and vfield > thresh or not gt and vfield <= thresh:
+            omitList.append(k)
+    for k in omitList:
+        del globalEntries[k]
 
[…]

### 2023-08-24 doc/JAIR/jair-paper.tex

@@
 \newcommand{\nodev}{\makenode{v}}
+\newcommand{\nodew}{\makenode{w}}
 \newcommand{\nodes}{\makenode{s}}
@@
 \newcommand{\dtrim}{\progname{drat-trim}}
+\newcommand{\ltrim}{\progname{lrat-trim}}
 \newcommand{\Dtrim}{\progname{Drat-trim}}
@@
 respect to the preceding
-clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
+clauses~\cite{jarvisalo:ijcar:2012,heule:cade:2013}.  Adding a RAT
 clause $C$ to a set of clauses maintains satisfiability, but it does not necessarily
@@
 Each input clause deletion can be done as a single RUP step, based
-on an algorithm to test for clausal entailment in det-DNNF graphs~\cite{capelli:sat:2019,darwiche:jair:2002}.  The
+on an algorithm to test for clausal entailment in det-DNNF graphs~\cite{darwiche:jair:2002,capelli:sat:2019}.  The
 proof generator constructs the hint sequence from the defining
@@
 In this section, we briefly describe the functionality we implemented and what we proved about it.
-More information is provided in the supplement~\cite{bryant:sat:2023:supplement}.
 
@@
 
-\section{Implementations}
+\section{Implementation}
 We have implemented programs that, along with
@@
 
-Our proof generator is written in C/C++ and uses
-\cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
-proof steps back into hinted CPOG clause additions, the generator can
-use either its own RUP proof generator, or it can invoke
-\dtrim{}~\cite{heule:cade:2013}.  The latter yields shorter proofs and scales well
-to large proofs, but each invocation has a high startup cost.  We
-therefore only use it when solving larger problems (currently ones
-with over 1000 clauses).
+Our proof generator is written in C/C++ and uses a recent version of
+the \cadical{} SAT solver that directly generates hinted proofs in
+LRAT format~\cite{biere:sat:2023}.  It also uses their tool \ltrim{}
+to reduce the length of the generated proofs.
 
@@
 
-We incorporated a ring evaluator into the prototype checker.  It
-can perform both standard and weighted
-model counting with full precision.  It performs arithmetic over a subset of the rationals
-we call $\drational$, consisting of numbers of the form $a \cdot 2^{b}
-\cdot 5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented
-to have arbitrary range.  Allowing scaling by powers of 2 enables the
+We incorporated a ring evaluator into the prototype checker.  It can
+perform both standard and weighted model counting with full precision.
+It performs arithmetic over a subset of the rationals we call
+$\drational$, consisting of numbers of the form $a \cdot 2^{b} \cdot
+5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented to
+have arbitrary range.  Allowing scaling by powers of 2 enables the
 density computation and rescaling required for standard model
@@
 counting competitions.  To give a sense of scale, the counter
-generated a result with 260,909 decimal digits
-for one of the weighted benchmarks.
+generated a result with 260,909 decimal digits for one of the weighted
+benchmarks.  Our implementation of arbitrary-range integers represents
+a number as a sequence of ``digits'' with each digit ranging from $0$
+to $10^9-1$, and with the digits stored as four-byte blocks.  This
+allows easy conversion to and from a decimal representation of the
+number.
 
@@
 
-\section{Concluding Remarks}
-\label{sect:future}
-
-This paper demonstrates a method for certifying the equivalence of two
-different representations of a Boolean formula: an input formula
-represented in conjunctive normal form, and a compiled representation
-that can then be used to extract useful information about the formula,
-including its weighted and unweighted model counts.  It builds on the
-extensive techniques that have been developed for clausal
-proof systems, including extended resolution and reverse unit propagation, as well as established tools, such as
-proof-generating SAT solvers and \dtrim{}.
+\section{Extensions}
 
@@
 are required, and that it will provide a useful tool for developers of
-knowledge compilers.
-%%Although our current implementation only handles
-%%the outputs of one particular program, it could be adapted
-Our experiments demonstrate that our toolchain can already handle
-problems nearly at the limits of current knowledge compilers.  Further
-engineering and optimization of our proof generator and checker could
-improve their performance and capacity substantially.  We are hopeful
-that our tool can be adapted to handle other knowledge compiler
-representations, such as sentential decision diagrams (SDDs)~\cite{darwiche:ijcai:2011}.
+knowledge compilers.  Our current implementation only handles the
+outputs of knowledge compilers that generate dec-DNNF graphs, and it
+supports only queries that can be computed via ring evaluation.  Here
+we discuss ways to extend both capabilities.
+
+\subsection{Validating Sentential Decision Diagrams}
+\label{sect:sdd}
+
+\begin{figure}
+  \begin{center}
+    \input{dd/sdd}
+  \end{center}
+\caption{Translation of SDD node $\nodeu$ into POG node $\hat{\nodeu}$.  An SDD node with $2k$ children expands into $k$ product nodes and $k-1$ sum nodes.}
+\label{fig:sdd:pog}
+\end{figure}
+
+\citeA{darwiche:ijcai:2011} introduced \emph{Sentential
+Decision Diagrams} (SDDs) as a generalization of 
+Ordered Binary Decision Diagrams (OBDDs)~\cite{Bryant:1986}.  Like
+OBDDs, SDDs are a canonical representation and support bottom-up
+construction.
+
+As with other representations, each node $\nodeu$ in an SDD denotes a
+Boolean formula $\phi_{\nodeu}$.  Leaf nodes correspond to $\tautology$, $\nil$, and literals.
+Each nonterminal node has $2k$ children,
+organized as pairs $\nodep_i$ and $\nodes_i$ for $1\leq i \leq k$.
+The dependency set for the node must be partitioned into disjoint sets $X_p$ and $X_s$ such that
+$\dependencyset(\phi_{\nodep_i}) \subseteq X_p$ and
+$\dependencyset(\phi_{\nodes_i}) \subseteq X_s$ for all $1 \leq i \leq k$.
+Children $\nodep_1, \nodep_2, \ldots, \nodep_k$ serve as a set of
+\emph{selector} functions with mutually disjoint models.  That is,
+$\modelset(\phi_{\nodep_i}) \cap \modelset(\phi_{\nodep_j}) =
+\emptyset$ for $1 \leq i < j \leq k$.
+The Boolean formula associated with node $\nodeu$ is then
+\begin{eqnarray}
+\phi_{\nodeu} & = & \bigvee_{1 \leq i \leq k} \left[\phi_{\nodep_i} \land \phi_{\nodes_i}\right] \label{eqn:sdd}
+\end{eqnarray}
+
+An SDD can be translated into a POG, where an SDD node with $2k$
+children expands into $k$ product nodes and $k-1$ sum nodes.
+In particular, assume that each nonterminal SDD node $\nodev$ can be represented by a
+subgraph in the POG with root node $\hat{\nodev}$.  For leaf node $\nodev$, its counterpart $\hat{\nodev}$ is
+the associated constant or literal, rather than a POG node.  Figure~\ref{fig:sdd:pog} then
+illustrates how node $\nodeu$ with $2k$ children would be expanded
+into a subgraph with root node $\hat{\nodeu}$.  We can see that each
+of the product nodes has arguments with disjoint dependency sets, and
+each of the sum nodes has disjoint sets of models.
+
+Several different methods could be used to generate a CPOG
+certification from an SDD\@.  For the forward implication proof, the
+monolithic approach we have used could generate the proof via a single
+call to a proof-generating SAT solver.  For an SDD generated by the
+bottom up approach originally proposed by \citeA{darwiche:ijcai:2011},
+the generator could be modified to generate proof steps for each
+recursive call during the construction, as is done to generate proofs
+of unsatisfiability with proof-generating, BDD-based SAT
+solvers~\cite{Jussila:2006,ebddres,bryant:fmcad:2022,bryant:tocl:2023}.
+For an SDD generated by a top-down process, a structural method
+similar to the one we have described could be applied.  A further
+requirement would be to generate mutual exclusion proofs for the $k-1$
+sum nodes illustrated in Figure~\ref{fig:sdd:pog}.  These would be
+based on the disjoint model property of the $k$ nodes of the form
+$\hat{\nodep}_i$.  One approach would be to simply make $k-1$ calls a
+proof-generating SAT solver.
 
 \subsection{Generalizing to Semirings}
+\label{sect:semiring}
 
@@
 
-%% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
-%% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
-%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{1 \leq i \leq k}
-%% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
-%% such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
-%% \cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
-%% such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
-%% \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
-%% of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
-%% for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
-%% the sum of $t_1, \ldots, t_k$.
+\section{Concluding Remarks}
+\label{sect:future}
+
+This paper demonstrates a method for certifying the equivalence of two
+different representations of a Boolean formula: an input formula
+represented in conjunctive normal form, and a compiled representation
+that can then be used to extract useful information about the formula,
+including its weighted and unweighted model counts.  It builds on the
+extensive techniques that have been developed for clausal
+proof systems, including extended resolution and reverse unit propagation, as well as established tools, such as
+proof-generating SAT solvers and \dtrim{}.
+
+Our experiments demonstrate that our toolchain can already handle
+problems nearly at the limits of current knowledge compilers.  Further
+engineering and optimization of our proof generator and checker could
+improve their performance and capacity substantially.
 

### 2023-08-27 doc/JAIR/jair-data.tex
Include one-sided data
Completed overall data

@@
        \item $D$ also equals sum of number of edges and number of nodes
-       \item For 122 instances, Min = 304, Median = 774,883, Mean = 72,389,514, Max = 1,765,743,261
+       \item For 123 instances, Min = 304, Median = 774,883, Mean = 72,389,514, Max = 2,761,457,765 (1,765,743,261 for those that didn't overflow).
        \end{itemize}
@@
 
-     \item Performance (122 instances):
+     \item Performance (122 instances.  Don't include the one that overflowed):
        \begin{itemize}
@@
 \item The tree ratio provides a reasonable discriminator for monolithic vs.~structured.  Based on experiments, set threshold at 5.0. Use monolithic below this and structured above.
-\item We could not find any way to exploit the hybrid approach for these benchmarks.
-\item On the other hand, the monolithic approach becomes impractical beyond around $10^7$ defining clauses.  Even the best SAT solvers become unstable beyond this.
-\item Therefore, use hybrid approach for very large problems with tree ratios below 5.0, shift to monolithic once the tree size is at most $10^7$.
+\item We could not find any way to exploit the hybrid approach for the smaller benchmarks.
+\item On the other hand, the monolithic approach becomes impractical beyond around $10^6$ defining clauses.  Even the best SAT solvers become unstable beyond this.
+\item Therefore, use hybrid approach for very large problems with tree ratios below 5.0, shift to monolithic once the tree size is at most $10^6$.
+\item Also tried with limit of $10^5$ clauses.  Generally did worse in terms of completions.
 \end{itemize}
@@
 \begin{itemize}
-\item Of the 122 benchmarks, 100 were completed using the hybrid approach within 10,000 seconds.
+\item Of the 123 benchmarks, 110 were completed using the hybrid approach within 10,000 seconds.
+\item One additional one completed within 10,000 for structural (and for hybrid with a threshold of $10^5$ clauses.
+\item For seven others, was able to do one-sided proof
+\item Only one had more than $10^9$ defining clauses.
+\item Total of 5 (including 1 overflow case) for which no form of validation succeeded.
+  \begin{itemize}
+  \item Three of these had more than $10^9$ defining clauses
+  \item Smallest had 405 million
+
+  \end{itemize}
 \end{itemize}
@@
     \input{data-formatted/seconds-summary}
-%    \input{data-formatted/seconds-summary-onesided}
+    \input{data-formatted/seconds-summary-structured-only}
+    \input{data-formatted/seconds-summary-onesided}
     \input{data-formatted/seconds-summary-failures}
     \legend{
-      \scriptsize \textsf{Full validation},
-%      \scriptsize \textsf{One-sided validation},
+      \scriptsize \textsf{Full validation, Hybrid},
+      \scriptsize \textsf{Full validation, Structural},
+      \scriptsize \textsf{One-sided validation},
       \scriptsize \textsf{Resource limit exceeded},
@@
     \input{data-formatted/clauses-summary}
-%%    \input{data-formatted/clauses-defining-onesided}
+    \input{data-formatted/clauses-summary-structured-only}
+    \input{data-formatted/clauses-summary-onesided}
     \input{data-formatted/clauses-summary-failures}
@@
     \legend{
-      \scriptsize \textsf{Full validation},
-%      \scriptsize \textsf{One-sided validation},
+      \scriptsize \textsf{Full validation, Hybrid},
+      \scriptsize \textsf{Full validation, Structural},
+      \scriptsize \textsf{One-sided validation},
       \scriptsize \textsf{No validation},

### 2023-08-27 doc/JAIR/jair-data.tex
Summary statistics

@@
     }
-    \addplot[mark=none, dashed] coordinates{(0.01,25000) (4000, 25000)};
+    \addplot[mark=none, dashed] coordinates{(0.01,25000) (4000, 25000)};) 
     \addplot[mark=none, color=lightblue] coordinates{(0.01,20000) (4000,20000)};
@@
 \end{figure}
+
+Figure~\ref{fig:d4:cpog} compares the running time for the combination
+of generating and checking the CPOG file (Y axis), versus the running
+time for C4 (X Axis).  If we compare the ratio between the two we find the following:
+\begin{itemize}
+\item Min = 0.27 (i.e., Verification chain was $3.64\times$ faster
+\item Max = 177.0
+\item Median = 12.5
+\end{itemize}
 
@@
 \end{figure}
+
+Figure~\ref{fig:defining:total} compares the total number of clauses in the CPOG representation (Y axis) versus the number of defining clauses (X axis).
+Since the former include the latter, the ratio between these cannot be less than 1.0.
+Comparing the ratios gives:
+\begin{itemize}
+\item Min = 1.02
+\item Max = 9460.2
+\item Median = 2.29
+\end{itemize}
 

### 2023-08-28 doc/JAIR/jair-data.tex
Completed data for structured vs. monolithic

@@
 
+\section*{Selecting Reduced Benchmark Set}
+
+To make the experiments tractable, want to select set of benchmarks that are more tractable.
+Limit set to the 90 with at most $10^7$ defining clauses, and for which D4 runs in at most 1000 seconds.
+
+D4 Performance (90 instances):
+       \begin{itemize}
+       \item Times: Min = 0.012 secs, Max = 660.2 secs, Median = 1.72 secs
+       \item Defining clauses: Min = 304, Max = 8,493,275, Median = 378,325
+       \item Tree Sizes ($D_t$): Min = 304, max = 1,205,743,897, median = 2,690,776
+       \item Tree ratio: Min = 1.0, Max = 363.9, Median = 9.98, Harmonic Mean = 2.94
+       \end{itemize}
+
+\section*{CPOG Generation}
+
+We have devised two different approaches to generating the forward implication proof from the dec-DNNF generated by {\sc D4}:
+\begin{description}
+\item[Monolithic:] Make a single call to a proof generating SAT solver.
+\item[Structured:] Traverse the POG top-down and recursively generate a forward implication proof for each node encountered.
+Lemmas can be introduced to avoid an exponential expansion of the POG into a tree.
+\end{description}
+
+Our implementation also allows a hybrid of these approaches: at any
+point in the structured traversal, it can generate a proof for a node via a
+single call to a SAT solver.  This capability supports both full
+monolithic generation (perform monolithic generation at the POG root),
+full structured generation (continue traversing to the POG leaves), or
+a balance between the two.
+The challenge is then to devise a
+heuristic criterion for deciding whether to continue the traversal below
+some node or to generate a proof for that node monolithically.
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.70]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
+      legend style={at={(0.30,0.96)}},
+      legend cell align={left},
+                              x post scale=2.0, y post scale=2.0,
+                              xmode=log,xmin=0.01,xmax=1800,
+                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
+                              ymode=log, ymin=0.01, ymax=1800,
+                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
+                              xlabel={Monolithic generation + check (seconds)}, ylabel={Structural generation + check (seconds)},
+            ]
+
+    \input{data-formatted/sub10M-seconds-subthresh}
+    \input{data-formatted/sub10M-seconds-supthresh}
+    \legend{
+      \scriptsize \textsf{Tree ratio $\leq \; 5{.}0$},
+      \scriptsize \textsf{Tree ratio $> \; 5{.}0$},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (1000,1000)};
+    \addplot[mark=none, color=lightblue] coordinates{(0.01,1000) (1000,1000)};
+    \addplot[mark=none, color=lightblue] coordinates{(1000,0.01) (1000,1000)};
+    \addplot[color=black,dashed] coordinates{(1e-2,1500) (1500,1500)};
+    \addplot[color=black,dashed] coordinates{(1500,1e-2) (1500,1500)};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Structural (Y-axis) versus monolithic (X-axis) execution times.  Monolithic generally performed better for lower tree ratios.}
+\label{fig:seconds:mono:structured}
+\end{figure}
+
+Our first set of experiments applies full monolithic and full
+structured generation to the reduced benchmark set.  Figure
+~\ref{fig:seconds:mono:structured} shows a plot comparing the two
+approaches.  Each axis shows the number of seconds to generate and
+check a CPOG file.  Data points to the left of the diagonal line ran
+faster with the structural method, while those to the right ran faster
+with the monolithic method.  Five of the benchmarks exceeded the 1000-second threshold for both approaches.
+The data are divided into two groups:
+those having tree ratios below 5.0 and those having tree ratios above
+5.0.  
+As can be seen there is some correlation.  Of the 85 benchmarks for which at least one completed within 1000 seconds:
+\begin{itemize}
+\item For those with tree ratios below 5.0, 12 ran faster with monolithic generation and 23 with structural
+\item For those with tree ratios above 5.0, 37 ran faster with monolithic generation and 13 with structural
+\end{itemize}
+Based on this, we devised the following selection rule: when the
+tree ratio is at most $5.0$, use the monolithic approach, otherwise
+use the structured approach.  That would yield the better choice for
+60 of the 85 cases.  In terms of failures, there is one benchmark with
+a high tree ratio for which the monolithic approach requires 835
+seconds, and the structural approach requires 1388. This is therefore
+marked as a failure by following our rule.  On the other hand, the rule makes the better choice for four of the cases where one approach times out while the other completes.
+Our data set was too sparse to do more tuning, including the optimal threshold selection.
+
+Overall, our results on the 90 benchmarks were:
+\begin{itemize}
+\item Monolithic: 83 completed
+\item Structured: 82 completed
+\item Threshold of 5.0: 85 completed
+\end{itemize}
+
+We tried various experiments where the proof generator starts at the
+top using a structural approach and then switches to a monolithic
+approach once the tree size for a node drops below some threshold.  We
+could not find any selection procedure that improved the overall
+performance for the 90 smaller benchmarks.  On the other hand, it
+proved useful for the very large benchmarks, because calling a SAT
+solver on a problem with more than $10^6$ clauses often causes the
+solver to fail.  We therefore refined the rule into a \emph{hybrid} approach to operate as follows:
+\begin{enumerate}
+\item Label all nodes by the tree size of the subgraph for which it is the root.
+\item Compute the total size of the graph and the tree ratio of the root.
+\item Proceed with proof generation with the following rules
+  \begin{enumerate}
+  \item If the tree ratio is below 5.0 and the tree size of the root is below $10^6$, do the entire proof generation with a monolithic approach
+  \item If the tree ratio is below 5.0 and the tree size of the root is above $10^6$, start with a structural approach and shift to a monolithic approach once the tree size at a node is below $10^6$.
+  \item If the tree ratio is above 5.0, do the entire proof generation
+    with a structural approach.
+  \end{enumerate}
+\end{enumerate}
+
+\begin{figure}
+\centering{%
+\begin{tikzpicture}[scale = 0.70]
+  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
+      legend style={at={(0.30,0.96)}},
+      legend cell align={left},
+                              x post scale=2.0, y post scale=2.0,
+                              xmode=log,xmin=1e2,xmax=3e7,
+                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, 
+                              xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$, $10^{10}$},
+                              ymode=log,ymin=1e2,ymax=3e7,
+                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, 
+                              yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$, $10^{10}$},
+                              xlabel={Monolithic proof (clauses)}, ylabel={Structural proof + check (clauses)},
+            ]
+
+    \input{data-formatted/sub10M-clauses-subthresh}
+    \input{data-formatted/sub10M-clauses-supthresh}
+    \legend{
+      \scriptsize \textsf{Tree ratio $\leq \; 5{.}0$},
+      \scriptsize \textsf{Tree ratio $> \; 5{.}0$},
+    }
+    \addplot[mark=none, color=lightblue] coordinates{(100,100) (2e7,2e7)};
+    \addplot[mark=none, color=lightblue] coordinates{(100,2e7) (2e7,2e7)};
+    \addplot[mark=none, color=lightblue] coordinates{(2e7,100) (2e7,2e7)};
+    \addplot[color=black,dashed] coordinates{(100,2.5e7) (2.5e7,2.5e7)};
+    \addplot[color=black,dashed] coordinates{(2.5e7,100) (2.5e7,2.5e7)};
+
+          \end{axis}
+\end{tikzpicture}
+} % centering
+\caption{Structural (Y-axis) versus monolithic (X-axis) proof sizes.  Monolithic generally performed better for lower tree ratios.}
+\label{fig:clauses:mono:structured}
+\end{figure}
+
+Figure~\ref{fig:clauses:mono:structured} shows the comparative proof
+sizes (in clauses) for the two approaches.  As can be seen, the
+monolithic approach tends to generate shorter proofs.  Of the 80 cases
+for which both methods completed within 1000 seconds, 18 had shorter
+proofs with the structural approach and 62 with the monolithic
+approach.  In setting our tree ratio threshold to 5.0, we get the shorter proofs for 45 cases and suboptimal proof sizes for 40.
+
 \section*{Summary Result}
 
-Paper describes two ways to generate the forward implication proof:
-structured and monolithic.  Can also combine the two: at any point in
-the traversal during structured generation, the forward implication
-proof for the entire subgraph can be generated by a single call to a proof-generating
-SAT solver.  We'll refer to this as a \emph{hybrid} approach.
-
-Ran some experiments on all but the largest benchmarks (presented
-later) seeing under what conditions was it best to use a monolithic,
-structured, or hybrid approach, and where the hybrid approach should
-shift from structured to monolithic.
-
-General findings:
-\begin{itemize}
-\item The tree ratio provides a reasonable discriminator for monolithic vs.~structured.  Based on experiments, set threshold at 5.0. Use monolithic below this and structured above.
-\item We could not find any way to exploit the hybrid approach for the smaller benchmarks.
-\item On the other hand, the monolithic approach becomes impractical beyond around $10^6$ defining clauses.  Even the best SAT solvers become unstable beyond this.
-\item Therefore, use hybrid approach for very large problems with tree ratios below 5.0, shift to monolithic once the tree size is at most $10^6$.
-\item Also tried with limit of $10^5$ clauses.  Generally did worse in terms of completions.
-\end{itemize}
-
-Our summary data is based on this set of rules.
+Our summary data is based on the hybrid approach.
 
@@
 \item Of the 123 benchmarks, 110 were completed using the hybrid approach within 10,000 seconds.
-\item One additional one completed within 10,000 for structural (and for hybrid with a threshold of $10^5$ clauses.
+\item One additional one completed within 10,000 for structural (and for hybrid with a threshold of $10^5$ clauses).
 \item For seven others, was able to do one-sided proof
@@
 \begin{itemize}
-\item Min = 0.27 (i.e., Verification chain was $3.64\times$ faster
+\item Min = 0.27 (i.e., verification chain was $3.64\times$ faster)
 \item Max = 177.0
@@
 \end{itemize}
-
-\section*{Selecting Reduced Benchmark Set}
-
-To make the experiments tractable, want to select set of benchmarks that are more tractable.
-Limit set to the 90 with at most $10^7$ defining clauses, and for which D4 runs in at most 1000 seconds.
-
-D4 Performance (90 instances):
-       \begin{itemize}
-       \item Times: Min = 0.012 secs, Max = 660.2 secs, Median = 1.72 secs
-       \item Defining clauses: Min = 304, Max = 8,493,275, Median = 378,325
-       \item Tree Sizes ($D_t$): Min = 304, max = 1,205,743,897, median = 2,690,776
-       \item Tree ratio: Min = 1.0, Max = 363.9, Median = 9.98, Harmonic Mean = 2.94
-       \end{itemize}
-
-\section*{CPOG Generation}
-
-We have devised two different approaches to generating the forward implication proof from the dec-DNNF generated by {\sc D4}:
-\begin{description}
-\item[Monolithic:] Make a single call to a proof generating SAT solver.
-\item[Structured:] Traverse the POG top-down and recursively generate a forward implication proof for each node encountered.
-Lemmas can be introduced to avoid an exponential expansion of the POG into a tree.
-\end{description}
-
-Our implementation also allows a hybrid of these approaches: at any
-point in the structured traversal, it can generate a proof for a node via a
-single call to a SAT solver.  This capability supports both full
-monolithic generation (perform monolithic generation at the POG root),
-full structured generation (continue traversing to the POG leaves), or
-a balance between the two.
-The challenge is then to devise a
-heuristic criterion for deciding whether to continue the traversal below
-some node or to generate a proof for that node monolithically.
-
-Our first set of experiments applies full monolithic and full structured generation to the reduced benchmark set.  
-After some experimenting, we devised the following ``hybrid'' method: when the tree ratio is at most $5.0$, use the monolithic approach, otherwise use the structured approach.
-We set a time limit of 1000 seconds for the proof generation and checking.
-Overall, our results on the 92 benchmarks were:
-\begin{itemize}
-\item Monolithic: 83 completed
-\item Structured: 82 completed
-\item Hybrid: 85 completed
-\end{itemize}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.70]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
-      legend style={at={(0.30,0.96)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=2.0,
-                              xmode=log,xmin=0.01,xmax=1000,
-                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
-                              ymin=0.01, ymax=2000,
-                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
-                              xlabel={D4 runtime (seconds)}, ylabel={Toolchain runtime (seconds)},
-            ]
-
-    \input{data-formatted/sub10M-seconds-hybrid}
-    \legend{
-      \scriptsize \textsf{Hybrid},
-    }
-
-    \input{data-formatted/sub10M-seconds-failures-hybrid}
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Total runtime for toolchain using three different approaches to forward-implication proof generation, as function of D4 runtime}
-\label{fig:d4:cpog:compare}
-\end{figure}
 
@@
 
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
-                              x post scale=1.8, y post scale=1.8,
-                              xmin=0,xmax=1000,
-                              xtick = {0,100,200,300,400,500,600,700,800,900,1000},
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \input{data-formatted/time+defining-success}
-    \input{data-formatted/time+defining-failure}
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Number of defining clauses in generated POG as function of D4 runtime (X linear scale)}
-\label{fig:generate:d4}
-\end{figure}
-
 \end{description}

### 2023-08-29 doc/JAIR/jair-data.tex
Tweaks

@@
 top using a structural approach and then switches to a monolithic
-approach once the tree size for a node drops below some threshold.  We
-could not find any selection procedure that improved the overall
-performance for the 90 smaller benchmarks.  On the other hand, it
-proved useful for the very large benchmarks, because calling a SAT
-solver on a problem with more than $10^6$ clauses often causes the
-solver to fail.  We therefore refined the rule into a \emph{hybrid} approach to operate as follows:
+approach once the tree size for a node drops below some threshold.
+This was helpful for very large benchmarks, using high thresholds.
+We also found that SAT solvers could not reliably handle problems with $10^7$ or more clauses.
+We therefore refined the rule into a \emph{hybrid} approach that operates as follows:
 \begin{enumerate}
-\item Label all nodes by the tree size of the subgraph for which it is the root.
+\item With a bottom-up traversal of the graph, Label each node by the tree size of the subgraph for which it is the root.
 \item Compute the total size of the graph and the tree ratio of the root.
@@
 \end{enumerate}
+Unless noted otherwise, The remainder of our experimental data is based on this approach.
 

### 2023-11-14 doc/JAIR/jair-paper.tex
Some improvements

@@
 \definecolor{medgreen}{rgb}{0.15, 0.6, 0.15}
+\definecolor{midred}{rgb}{0.80,0.3,0.3}
 
@@
 \lstset{backgroundcolor=\color{white}}
+
+\newcommand{\rtext}[1]{\textcolor{midred}{\texttt{#1}}}
+\newcommand{\btext}[1]{\textcolor{midblue}{\texttt{#1}}}
 
@@
   providing a RUP proof that the arguments are mutually exclusive.
-  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness
+  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness.
 %  Generalizing to $k$ arguments would require
@@
 \midrule
-1 & \texttt{-1 3 -4} & \texttt{0} \\
-2 & \texttt{-1 -3 4} & \texttt{0} \\
-3 & \texttt{3 -4} & \texttt{0}\\
-4 & \texttt{1 -3 4} & \texttt{0} \\
-5 & \texttt{-1 -2} & \texttt{0} \\
+\rtext{1} & \texttt{-1 3 -4} & \texttt{0} \\
+\rtext{2} & \texttt{-1 -3 4} & \texttt{0} \\
+\rtext{3} & \texttt{3 -4} & \texttt{0}\\
+\rtext{4} & \texttt{1 -3 4} & \texttt{0} \\
+\rtext{5} & \texttt{-1 -2} & \texttt{0} \\
 \bottomrule
@@
 \midrule
-6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
-9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
-12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
-15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\
-18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
-22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
+\rtext{6} & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
+\rtext{9} & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
+\rtext{12} & \texttt{s 7 5 6} \; \rtext{7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
+\rtext{15} & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\
+\rtext{18} & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
+\rtext{22} & \texttt{s 10 8 9} \; \rtext{16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
  & \texttt{r 10} && Root $r = s_{10}$\\
@@
 \midrule
-\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
-\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
-\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
-\midrule
-\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
-\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
-\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
-\midrule
-\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
-\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
-\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
-\midrule
-\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
-\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
-\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
-\midrule
-\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
-\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
-\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
-\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
-\midrule
-\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
-\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
-\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
+\rtext{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
+\rtext{7} & \texttt{-5 -3} & \texttt{0} & \\
+\rtext{8} & \texttt{-5 -4} & \texttt{0} & \\
+\midrule
+\rtext{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
+\rtext{10} & \texttt{-6 3} & \texttt{0} & \\
+\rtext{11} & \texttt{-6 4} & \texttt{0} & \\
+\midrule
+\rtext{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
+\rtext{13} & \texttt{7 -5} & \texttt{0} & \\
+\rtext{14} & \texttt{7 -6} & \texttt{0} & \\
+\midrule
+\rtext{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
+\rtext{16} & \texttt{-8 -1} & \texttt{0} & \\
+\rtext{17} & \texttt{-8 7} & \texttt{0} & \\
+\midrule
+\rtext{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
+\rtext{19} & \texttt{-9 1} & \texttt{0} & \\
+\rtext{20} & \texttt{-9 -2} & \texttt{0} & \\
+\rtext{21} & \texttt{-9 7} & \texttt{0} & \\
+\midrule
+\rtext{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
+\rtext{23} & \texttt{10 -8} & \texttt{0} & \\
+\rtext{24} & \texttt{10 -9} & \texttt{0} & \\
 \bottomrule
@@
 \midrule
-\texttt{25} & \texttt{a 5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
-\texttt{26} & \texttt{a 6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
-\texttt{27} & \texttt{a 3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
-\texttt{28} & \texttt{a 7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
-\texttt{29} & \texttt{a 8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
-\texttt{30} & \texttt{a 5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
-\texttt{31} & \texttt{a 6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
-\texttt{32} & \texttt{a 3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
-\texttt{33} & \texttt{a 7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
-\texttt{34} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
-\texttt{35} & \texttt{a 1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
-\texttt{36} & \texttt{a 10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
+\rtext{25} & \texttt{a 5 1 3} & \texttt{0} & \rtext{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
+\rtext{26} & \texttt{a 6 1 -3} & \texttt{0} & \rtext{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
+\rtext{27} & \texttt{a 3 7 1} & \texttt{0} & \rtext{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
+\rtext{28} & \texttt{a 7 1} & \texttt{0} & \rtext{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
+\rtext{29} & \texttt{a 8 1} & \texttt{0} & \rtext{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
+\rtext{30} & \texttt{a 5 -1 3} & \texttt{0} & \rtext{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
+\rtext{31} & \texttt{a 6 -1 -3} & \texttt{0} & \rtext{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
+\rtext{32} & \texttt{a 3 7 -1} & \texttt{0} & \rtext{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
+\rtext{33} & \texttt{a 7 -1} & \texttt{0} & \rtext{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
+\rtext{34} & \texttt{a 9 -1} & \texttt{0} & \rtext{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
+\rtext{35} & \texttt{a 1 10} & \texttt{0} & \rtext{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
+\rtext{36} & \texttt{a 10} & \texttt{0} & \rtext{35 24 34} & \texttt{0} & $s_{10}$ \\
 \bottomrule
@@
 \midrule
- \texttt{d 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
- \texttt{d 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
- \texttt{d 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
- \texttt{d 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
- \texttt{d 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
+ \texttt{d 1} & \rtext{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
+ \texttt{d 2} & \rtext{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
+ \texttt{d 3} & \rtext{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
+ \texttt{d 4} & \rtext{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
+ \texttt{d 5} & \rtext{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
 \bottomrule
@@
 \end{minipage}
-\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F)}
+\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F).  The defining clauses (D) are implicitly
+defined by the POG declaration (C).}
 \label{fig:eg4:all}
@@
 Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
-$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
+$\nodes_7$ has clause IDs 7 and 10 as the hint.  We can see in Figure
 \ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
@@
 Working backward, we can see that
-clause 29 indicates that variable $p_8$ will be implied by the input
-clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
-be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
+clause 35 indicates that variable $s_{10}$ will be implied by the input
+clauses when $\assign(x_1) = 0$. Clause 34 indicates that node $p_9$ will
+be implied by the input clauses when $\assign(x_1) = 1$, while defining clause 24 shows that node $s_{10}$ will be implied
+by the input clauses when $\eassign{p_9} = 1$.  These three clauses
+serve as the
 hint for clause 36.
@@
 involves simply deleting the clauses from clause number 35 down to
-clause number 25, with each deletion using the same hint as were used
+clause number 25, with each deletion using the same hint as was used
 to add that clause.  In the end, therefore, only the defining clauses
@@
 to the proof framework---they build on the power of extended
-resolution to enable new logical structures.  They
+resolution to enable the construction of new logical structures.  They
 involve declaring new product nodes to encode products of literals.
@@
 \midrule
-25 & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
-28 & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
+\rtext{25} & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
+\rtext{28} & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
 \bottomrule
@@
 \midrule
-\texttt{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
-\texttt{26} & \texttt{-11 -3} & \texttt{0} & \\
-\texttt{27} & \texttt{-11 4} & \texttt{0} & \\
-\midrule
-\texttt{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
-\texttt{29} & \texttt{-12 3} & \texttt{0} & \\
-\texttt{30} & \texttt{-12 -4} & \texttt{0} & \\
+\rtext{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
+\rtext{26} & \texttt{-11 -3} & \texttt{0} & \\
+\rtext{27} & \texttt{-11 4} & \texttt{0} & \\
+\midrule
+\rtext{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
+\rtext{29} & \texttt{-12 3} & \texttt{0} & \\
+\rtext{30} & \texttt{-12 -4} & \texttt{0} & \\
 \bottomrule
@@
 \multicolumn{6}{l}{Lemma Proof} \\
-\texttt{31} & \texttt{a 5 11 12 3} & \texttt{0} & \texttt{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
-\texttt{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \texttt{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
-\texttt{33} & \texttt{a 3 7 11 12} & \texttt{0} & \texttt{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
-\texttt{34} & \texttt{a 7 11 12} & \texttt{0} & \texttt{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
+\rtext{31} & \texttt{a 5 11 12 3} & \texttt{0} & \rtext{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
+\rtext{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \rtext{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
+\rtext{33} & \texttt{a 3 7 11 12} & \texttt{0} & \rtext{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
+\rtext{34} & \texttt{a 7 11 12} & \texttt{0} & \rtext{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
 \midrule
 \multicolumn{6}{l}{Lemma Application \#1} \\
-\texttt{35} & \texttt{a -11 1} & \texttt{0} & \texttt{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
-\texttt{36} & \texttt{a -12 1} & \texttt{0} & \texttt{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
-\texttt{37} & \texttt{a 7 1} & \texttt{0} & \texttt{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
-\midrule
-\texttt{38} & \texttt{a 8 1} & \texttt{0} & \texttt{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
+\rtext{35} & \texttt{a -11 1} & \texttt{0} & \rtext{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
+\rtext{36} & \texttt{a -12 1} & \texttt{0} & \rtext{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
+\rtext{37} & \texttt{a 7 1} & \texttt{0} & \rtext{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
+\midrule
+\rtext{38} & \texttt{a 8 1} & \texttt{0} & \rtext{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
 \midrule
 \multicolumn{6}{l}{Lemma Application \#2} \\
-\texttt{39} & \texttt{a -11 -1} & \texttt{0} & \texttt{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
-\texttt{40} & \texttt{a -12 -1} & \texttt{0} & \texttt{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
-\texttt{41} & \texttt{a 7 -1} & \texttt{0} & \texttt{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
-\midrule
-\texttt{42} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
-\texttt{43} & \texttt{a 1 10} & \texttt{0} & \texttt{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
-\texttt{44} & \texttt{a 10} & \texttt{0} & \texttt{43 24 42} & \texttt{0} & $s_{10}$ \\
+\rtext{39} & \texttt{a -11 -1} & \texttt{0} & \rtext{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
+\rtext{40} & \texttt{a -12 -1} & \texttt{0} & \rtext{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
+\rtext{41} & \texttt{a 7 -1} & \texttt{0} & \rtext{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
+\midrule
+\rtext{42} & \texttt{a 9 -1} & \texttt{0} & \rtext{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
+\rtext{43} & \texttt{a 1 10} & \texttt{0} & \rtext{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
+\rtext{44} & \texttt{a 10} & \texttt{0} & \rtext{43 24 42} & \texttt{0} & $s_{10}$ \\
 \bottomrule
@@
 then provide a proof of the lemma, stating that any assignment
-$\assign$ that activates these clauses will  assign $1$ to $s_7$.
+$\assign$ that activates these clauses will, when extended,  assign $1$ to $s_7$.
 Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
-will cause the first synthetic clause to activate due to input clause
+will, when extended, cause the first synthetic clause to activate due to input clause
 3, and it will cause the second synthetic clause to activate due to

### 2023-11-14 tools/merge_csv.py
More features

@@ -16,8 +16,9 @@ def eprint(s):
 
 
 def usage(name):
-    eprint("Usage: %s [-h] [-f]  [-s I1:I2:...:Ik] [-r I1:I2] [-t I:T1:T2] [-o I:T] [-l L0,L1,L2,...] FILE1.csv FILE2.csv ... FILEn.csv" % name)
+    eprint("Usage: %s [-h] [-f] [-d VAL] [-s I1:I2:...:Ik] [-r I1:I2] [-t I:T1:T2] [-o I:T] [-l L0,L1,L2,...] FILE1.csv FILE2.csv ... FILEn.csv" % name)
     eprint("  -f            Filter out lines that have at least one field missing")
+    eprint("  -d VAL        Use default of VAL when field missing")
     eprint("  -s I1:I2:...:Ik Sum the values from specified columns 1..k and add as new column")
     eprint("  -r I1:I2      Form the ratio between the values from specified columns 1 and 2 and add as new column")
     eprint("  -t I:T1:T2    Compare value from column with value T1.  If greater, set to T2.  If I negative, invert sense of threshold.  Add as new column")
@@ -60,21 +61,22 @@ def processFile(fname):
     
 # Merge two sets of entries.
 # When they both don't have the same keys, then either form superset or subset
-def merge(entries1, count1, entries2, count2, subset = True):
+def merge(entries1, count1, entries2, count2, subset = True, default = None):
     entries = {}
+    dfill = "" if default is None else default
     for k in entries1.keys():
         entry1 = entries1[k]
         if k in entries2:
             entry2 = entries2[k]
             entries[k] = entry1 + entry2
         elif not subset:
-            entry2 = [""] * count2
+            entry2 = [dfill] * count2
             entries[k] = entry1 + entry2
     if not subset:
         for k in entries2.keys():
             if k in entries1:
                 continue
-            entry1 = [""] * count1
+            entry1 = [dfill] * count1
             entry2 = entries2[k]
             entries[k] = entry1 + entry2
     return entries
@@ -86,7 +88,7 @@ def mergeConstant(entries1, value):
         entries[k] = entry1 + [str(value)]
     return entries
 
-def nextFile(fname, first, subset):
+def nextFile(fname, first, subset, default):
     global globalEntries, globalCount
     value = None
     try:
@@ -103,7 +105,7 @@ def nextFile(fname, first, subset):
         globalEntries = entries
         globalCount = ccount
     else:
[…]

### 2023-11-16 doc/JAIR/jair-data.tex
Work on optimization data

@@
                               yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$, $10^{10}$},
-                              xlabel={Monolithic proof (clauses)}, ylabel={Structural proof + check (clauses)},
+                              xlabel={Monolithic proof (clauses)}, ylabel={Structural proof  (clauses)},
             ]
@@
 
-\section*{Computing Counts and Weighted Counts}
-
-\begin{itemize}
-\item Implemented C package to represent numbers of form $a \cdot 2^b \cdot 5^c$.
-  \begin{itemize}
-  \item For unweighted counts, first compute density and then scale by $2^n$
-  \item For unweighted counts, perform exact decimal arithmetic
-  \item The set of numbers of this form is not closed under a reciprocal operation, but the weights in the competion are either of the form
-    $w(x) + w(\overline{x}) = 1.0$ or $w(x) = w(\overline{x}) = 1$.  The latter case sums to 2, with reciprocal $1/2$.
-  \end{itemize}
-\item Can implement having $a$ be integer of arbitrary size and $b$ and $c$ being 32-bit integers.
-  \begin{itemize}
-    \item Represent $a$ by ``digits,'' each ranging in value from $0$ to $10^9-1$.
-    \item Fits in 32-bit number
-    \item Easy scaling by 10
-    \item Easy conversion to/from decimal string
-  \end{itemize}
-\item Weighted counts ranged up to 260,909 decimal digits.
-\end{itemize}
-
-\section*{Time to Generate, Validate, and Count CPOG}
-
-\begin{itemize}
-\item Min 0.012, Average 1,329.8, Max 13,198.3, Median 75.4
-\item Ratio to time for D4: Min 0.51, Harmonic Mean 5.87, Max 465.8
-\end{itemize}
-
-\input{figure-runtime-d4-verify}
-
-\begin{figure}
-\centering{%
-\begin{tikzpicture}[scale = 0.90]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=3000000000,
-      legend style={at={(0.30,0.98)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=1.8,
-                              xmode=log, xmin=0.01,xmax=4000,
-                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$,$10^{10}$},
-                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \input{data-formatted/time+defining-success}
-    \input{data-formatted/time+defining-onesided-only}
-    \input{data-formatted/time+defining-failure}
-    \legend{
-      \scriptsize \textsf{Full validation},
-      \scriptsize \textsf{One-sided validation},
-      \scriptsize \textsf{Resource limit exceeded},
-    }
- \end{axis}
-\end{tikzpicture}
-} % centering
-\caption{Number of defining clauses in generated POG as function of D4 runtime}
-\label{fig:generate:d4:log}
-\end{figure}
-
-\section*{Generated CPOG Files}
-
-\begin{itemize}
-\item Proof Clauses: Min 554, Median 1,719,245, Average 45,001,945, Max 770,382,773
-\item Ratio to number of defining clauses: Min 1.54, Harmonic Mean 3.13, Max 6072.96
-\end{itemize}
-
-\input{figure-clauses-d4-result}
-
 \section*{Benefits of Optimizations}
@@
 \end{description}
+These optimizations are only applicable when the forward implication
+proof is generated by the structural method, and so we concentrate on
+the 50 benchmarks (from the smaller set of 90 benchmarks having at
+most $10^7$ defining clauses) having tree ratios greater than $5.0$.
 

### 2023-12-22 doc/JAIR/jair-paper.tex
Correct bug in structured proof description

@@
 Figure~\ref{fig:chain} illustrates our certifying knowledge
-compilation and model counting toolchain.  Starting with
-input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
-generates a dec-DNNF representation,
-and the \emph{proof
-  generator} uses this to generate a CPOG file.
-The \emph{proof checker} verifies the equivalence of the CNF and CPOG representations.
-The \emph{ring evaluator} computes
-a standard or weighted model count from the POG\@ representation.
-As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
-moves the root of trust away from the complex and highly
-optimized knowledge compiler to a relatively simple checker and
-evaluator.  Importantly, the proof generator need not be
-trusted---its errors will be caught by the proof checker.
+compilation and model counting toolchain.  Starting with input formula
+$\inputformula$, the \dfour{} knowledge
+compiler~\cite{lagniez:ijcai:2017} generates a dec-DNNF
+representation, and the \emph{proof generator} uses this to generate a
+CPOG file.  The \emph{proof checker} verifies the equivalence of the
+CNF and CPOG representations.  The \emph{ring evaluator} computes a
+standard or weighted model count from the POG\@ representation.  As
+the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
+moves the root of trust away from the complex and highly optimized
+knowledge compiler to a relatively simple checker and evaluator.
+Importantly, the proof generator need not be trusted---its errors will
+be caught by the proof checker.
 
@@
 
-This paper is an extended version of one published at the
-2023 Conference on the Theory and Application of Boolean Satisfiability~\cite{bryant:sat:2023}.
+This paper is an extended version of one published at the 2023
+Conference on the Theory and Application of Boolean
+Satisfiability~\cite{bryant:sat:2023}.  It provides greater detail of
+the algorithms, the formal verification, and the experimental results.
 
@@
   We highlight some special classes of Boolean formulas.  A formula is
-  in \emph{negation normal form} when negation is applied only to variables.  A
-  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
-  negation normal form, and ii) sum is applied only to literals.  A CNF
-  formula can be represented as a set of \emph{clauses}, each of which is a
-  set of literals.  Each clause represents the sum of the
-  literals, and the formula is the product of its clauses.  We use
-  set notation to reference the clauses within a formula and the
-  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
-This literal must be assigned value $1$ by any satisfying assignment of the formula.
+  in \emph{negation normal form} when negation is applied only to
+  variables.  A formula is in \emph{conjunctive normal form} (CNF)
+  when (i) it is in negation normal form, and (ii) sum is applied only
+  to literals, and (iii) there is a single product operation over all
+  of the sums.  A CNF formula can be represented as a set of
+  \emph{clauses}, each of which is a set of literals.  Each clause
+  represents the sum of the literals, and the formula is the product
+  of its clauses.  We use set notation to reference the clauses within
+  a formula and the literals within a clause.  A clause consisting of
+  a single literal is referred to as a \emph{unit} clause and the
+  literal as a \emph{unit} literal.  This literal must be assigned
+  value $1$ by any satisfying assignment of the formula.
 
@@
 $\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
-obtained from $\psi$ by: i) eliminating any
+obtained from $\psi$ by: (i) eliminating any
 clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
-ii) for the remaining clauses eliminating those literals $\lit$ for
-which $\passign(\lit) = 0$, and iii) eliminating any duplicate or tautological clauses.
+(ii) for the remaining clauses eliminating those literals $\lit$ for
+which $\passign(\lit) = 0$, and (iii) eliminating any duplicate or tautological clauses.
 In doing these simplifications, we also track the \emph{provenance}
@@
   it must be the case that the clauses in
-  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
+  $\gamma = \simplify{\psi}{\passign \cup \lambda}$ can be partitioned into $k$ subsets
   $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
@@
 
-The combination of these two optimization guarantees that i) each call
+The combination of these two optimization guarantees that (i) each call
 to $\validate$ for a product node will cause at most one invocation of
-the SAT solver, and ii) each call to $\validate$ for any node $\nodeu$
+the SAT solver, and (ii) each call to $\validate$ for any node $\nodeu$
 will cause further recursive calls only once.  Our experimental

### 2024-01-05 doc/JAIR/jair-paper.tex
Minor polishing

@@
 labeled by Boolean variables and their complements, and with each
-nonterminal node labeled by a Boolean And or Or operation.  Restrictions
+nonterminal node labeled by a Boolean \textsc{and} or \textsc{or} operation.  Restrictions
 are placed on the structure of the graph (described in Section~\ref{sect:pog}) such that a count of the
@@
 Kimmig et al.\ present a very general {\em
-  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
+  algebraic model counting} framework~\cite{kimmig:jal:2017} describing
 properties of Boolean functions that can be efficiently computed from
@@
 resolution, it can be challenging to formulate a full set of
-requirements that guarantee soundness.  In fact, our efforts to
+requirements that guarantee soundness.  In fact, as described in Section~\ref{sect:lean:subtle-condition}, our efforts to
 formally verify our proof framework exposed subtle conditions that we had
@@
 Conference on the Theory and Application of Boolean
-Satisfiability~\cite{bryant:sat:2023}.  It provides greater detail of
+Satisfiability~\cite{bryant:sat:2023}.  It provides much greater detail about
 the algorithms, the formal verification, and the experimental results.
@@
 but we require features not found in proofs of unsatisfiability.
-Our proofs construct a new propositional formula, and we must verify
+In particular, our proofs construct a new propositional formula, and we must verify
 both
@@
 invariants, as described in Sections~\ref{sect:cpog} and
-\ref{section:formally-verified-toolchain}.
+\ref{sect:formally-verified-toolchain}.
 
@@
 \section{Logical Foundations}
-\label{section:logical:foundations}
+\label{sect:logical:foundations}
 
@@
 \end{prop}
-As is described in \ref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.
+As is described in \ref{sect:formally-verified-toolchain}, we have proved these three equations using \lean{}.
 
@@
 \item A det-DNNF graph can be viewed a POG with negation applied only to variables.
-\item A dec-DNNF graph is a det-DNNF graph with the further restriction that any
-  sum node $\nodeu$ has exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
-  any total assignment $\assign \in  \modelset(\phi_{\nodeu_b})$ has $\assign(x)=b$, for $b \in \{0,1\}$.
+\item A dec-DNNF graph is a det-DNNF graph with the further
+  restriction that any sum node $\nodeu$ has exactly two children
+  $\nodeu_1$ and $\nodeu_0$, and it has an associated \emph{decision variable} $x$.
+  For $b \in \{0,1\}$, node $\nodeu_b$ can be a terminal node with variable $x$, where the polarity of the edge from $\nodeu$ to $\nodeu_b$ is
+  negative for $b=0$ and positive for $b=1$.
+  Alternatively, $\nodeu_b$ can be a product node having either literal $\obar{x}$ ($b=0$), or literal $x$ ($b=1$) as one of its arguments.
+  Either form implies that any total assignment $\assign \in \modelset(\phi_{\nodeu_b})$
+  has $\assign(x)=b$, for $b \in \{0,1\}$.
 \end{itemize}
@@
 denote the set of extension variables occuring in formula $\theta$.
-For any assignment $\assign$ to the variables in $\varset$, the defining
+For any total assignment $\assign$ to the variables in $\varset$, the defining
 clauses induce a unique assignment $\eassign$ to the variables in $\varset
@@
 identifier $C$, with the first ones being the IDs of the input clauses and successive
-ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
-such that any clause identifier $C'$ given in the hint when adding clause $C$ must have $C' < C$.
+ones being the IDs of added clauses.  Clause identifiers must be defined in order,
+with any clause identifier $C'$ given in the hint when adding clause $C$ having $C' < C$.
 
@@
 \section{Generating CPOG from dec-DNNF}
-\label{section:generating:cpog}
+\label{sect:generating:cpog}
 
@@
 \section{A Formally Verified Toolchain}
-\label{section:formally-verified-toolchain}
+\label{sect:formally-verified-toolchain}
+\label{sect:lean:subtle-condition}
 
@@
 
-The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \ref{section:logical:foundations}.
+The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \ref{sect:logical:foundations}.
 Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas.

### 2024-01-05 doc/JAIR/jair-paper.tex
Minor polishing

@@
 
-The forward implication proof of Figure~\ref{fig:eg4:all}(E) was
+As examples, the forward implication proof of Figure~\ref{fig:eg4:all}(E) was
 generated by the structural approach.  Working from step 36 backward,
@@
  $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
- showing that any total assignment $\assign$ that activates the synthetic clauses will cause $u$ to be assigned $1$.
+ showing that any total assignment $\assign$ that activates the synthetic clauses will, when cause $u$ to be assigned $1$.
  More precisely, given assignment $\assign$ and its extension $\eassign$, if $\eassign(v_j) = 0$ for every guard literal $\obar{v}_j$, then $\eassign(u) = 1$.
@@
 subgraph may contain nodes that are themselves roots of shared
-subgraphs.  Nonetheless, the principles described allow the
+subgraphs.  Even then, the principles described allow the
 definition, proof, and applications of a lemma for each shared node in

### 2024-01-05 doc/JAIR/jair-paper.tex
Minor polishing
Fix up Extensions section

@@
 As with other representations, each node $\nodeu$ in an SDD denotes a
-Boolean formula $\phi_{\nodeu}$.  Leaf nodes correspond to $\tautology$, $\nil$, and literals.
+Boolean formula $\phi_{\nodeu}$.  Leaf nodes correspond to Boolean constants and literals.
 Each nonterminal node has $2k$ children,
@@
 subgraph in the POG with root node $\hat{\nodev}$.  For leaf node $\nodev$, its counterpart $\hat{\nodev}$ is
-the associated constant or literal, rather than a POG node.  Figure~\ref{fig:sdd:pog} then
+the associated terminal node.  Figure~\ref{fig:sdd:pog} then
 illustrates how node $\nodeu$ with $2k$ children would be expanded
@@
 based on the disjoint model property of the $k$ nodes of the form
-$\hat{\nodep}_i$.  One approach would be to simply make $k-1$ calls a
+$\hat{\nodep}_i$.  One approach would be to simply make $k-1$ calls to a
 proof-generating SAT solver.
@@
 
-Kimmig's formulation of algebraic model
-counting~\cite{kimmig:jal:2017} is more general than ours.  It allows
+The formulation of algebraic model
+counting by \citeA{kimmig:jal:2017} is more general than ours.  It allows
 the algebraic structure to be a \emph{semiring}.  A commutative
@@
 $(x \por \obar{x}) \pand \phi_i$.  When a knowledge compiler generates
-a representation that is both in normal form and is smooth, then a semiring evaluation of the formula
+a representation in negation normal form that is smooth, then a semiring evaluation of the formula
 can proceed by first assigning each literal $\lit$ the value $w(\lit)$.
@@
 Our POG representation can support evaluation of semiring formulas by
-imposing the restriction that the POG is in normal form and that is smooth.
+imposing the restriction that the POG is in negation normal form and that it is smooth.
 Given a smoothed
@@
 convert this into a smooth POG in normal form and verify its
-equivalence to the input formula.  Full verification would also require checking that the POG is smooth.
+equivalence to the input formula.  Of course, full verification would also require checking that the POG is smooth.
 

### 2024-01-05 doc/JAIR/jair-paper.tex
Split Lean section as separate file

@@
 
-\section{A Formally Verified Toolchain}
-\label{sect:formally-verified-toolchain}
-\label{sect:lean:subtle-condition}
-
-We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
-sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \ref{fig:chain}).
-These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}.
-Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic.
-In this section, we briefly describe the functionality we implemented and what we proved about it.
-
-\vspace{1em}\noindent
-\textbf{Proof checking.} The goal of a CPOG proof is to construct a POG that is equivalent to the input CNF $\inputformula$.
-The POG being constructed, and the set of active clauses are stored in the checker state \texttt{st}.
-The checker begins by parsing the input formula, initializing the active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one.
-It then processes every step of the CPOG proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
-Afterwards, it carries out the \textsc{final conditions} check of \ref{subsection:semantics}.
-Throughout the process, we maintain invariants that ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\ref{thm:lean:equiv}).
-
-The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \ref{sect:logical:foundations}.
-Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas.
-\lstinline{PropForm} is generic over the type of variables, so we instantiate it with our type as \lstinline{PropForm Var}.
-Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}.
-Requiring totality is not a limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$.
-We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.
-
-The invariants refer to the checker state \lstinline{st} with fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the clausal POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far.
-For any $\nodeu\in P$, \lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.
-The first two invariants state that assignments to original variables extend uniquely to extension variables defining the POG nodes.
-In the formalization, we split this into extension and uniqueness:
-\begin{lstlisting}
-/-- Any assignment satisfying φ₁ extends to φ₂ while preserving values on X. -/
-def extendsOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
-  ∀ (σ₁ : PropAssignment Var), σ₁ ⊨ φ₁ → ∃ σ₂, σ₁.agreeOn X σ₂ ∧ σ₂ ⊨ φ₂
-/-- Assignments satisfying φ are determined on Y by their values on X. -/
-def uniqueExt (X Y : Set Var) (φ : PropForm Var) :=
-  ∀ (σ₁ σ₂ : PropAssignment Var), σ₁ ⊨ φ → σ₂ ⊨ φ → σ₁.agreeOn X σ₂ →
-    σ₁.agreeOn Y σ₂
-
-invariants.extends_pogDefsForm : extendsOver st.inputCnf.vars ⊤ st.pogDefsForm
-invariants.uep_pogDefsForm : uniqueExt st.inputCnf.vars st.allVars st.pogDefsForm
-\end{lstlisting}
-Note that in the definition of \lstinline{uniqueExt}, the arrows associate to the right,
-so the definition says that the three assumptions imply the conclusion.
-The next invariant guarantees that the set of solutions over the original variables is preserved:
-\begin{lstlisting}
-def equivalentOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
-  extendsOver X φ₁ φ₂ ∧ extendsOver X φ₂ φ₁
-
-invariants.equivInput : equivalentOver st.inputCnf.vars st.inputCnf st.clauseDb
-\end{lstlisting}
-Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\ref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
-\begin{lstlisting}
-def partitioned : PropForm Var → Prop
-  | tr | fls | var _ => True
-  | neg φ    => φ.partitioned
-  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(τ ⊨ φ ∧ τ ⊨ ψ)
-  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ φ.vars ∩ ψ.vars = ∅
-
-invariants.partitioned : ∀ (u : ILit), (st.pog.toPropForm u).partitioned
-invariants.equivalent_lits : ∀ (u : ILit), equivalentOver st.inputCnf.vars
-    (u ∧ st.pogDefsForm) (st.pog.toPropForm x)
-\end{lstlisting}
-These invariants are maintained by valid CPOG proofs. Together with additional invariants
-that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
-node $\noder$:
-
-\begin{thm}
-\label{thm:lean:equiv}
-If the proof checker has assembled POG $P$ with root node $\noder$ starting from input formula $\inputformula$, and \textsc{final conditions} (as stated in Section~\ref{subsection:semantics}) hold of the checker state, then $\inputformula$ is logically equivalent to $\phi_\noder$.
-\end{thm}
-\begin{proof}
-\textsc{Final conditions} imply that the active clausal formula $\theta$ is exactly $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.
-The conclusion follows from this and the checker invariants.
-The full proof is formally verified in Lean.
-\end{proof}
-After certifying a CPOG proof, the checker can pass its in-memory POG representation to the ring evaluator, along with the partitioning guarantee provided by \texttt{invariants.partitioned}.
-
-\vspace{1em}\noindent
-\textbf{Ring evaluation}. We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
-(\ref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
-\begin{lstlisting}
-def weightSum {R : Type} [CommRing R]
-    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
-  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
-\end{lstlisting}
-The rules for efficient ring evaluation of partitioned formulas are expressed as:
-\begin{lstlisting}
-def ringEval (weight : Var → R) : PropForm Var → R
-  | tr       => 1
-  | fls      => 0
-  | var x    => weight x
-  | neg φ    => 1 - ringEval weight φ
-  | disj φ ψ => ringEval weight φ + ringEval weight ψ
-  | conj φ ψ => ringEval weight φ * ringEval weight ψ
-\end{lstlisting}
-\ref{prop:ring:eval} is then formalized as follows:
-\begin{lstlisting}
-theorem ringEval_eq_weightSum (weight : Var → R) {φ : PropForm Var} :
-    partitioned φ → ringEval weight φ = weightSum weight φ (vars φ)
-\end{lstlisting}
-To efficiently compute the ring evaluation of a formula represented by a POG node, we implemented
-\lstinline{Pog.ringEval} and then proved that it matches the specification above:
-\begin{lstlisting}
-theorem ringEval_eq_ringEval (pog : Pog) (weight : Var → R) (x : Var) :
-  pog.ringEval weight x = (pog.toPropForm x).ringEval weight
-\end{lstlisting}
-Applying this to the output of our verified CPOG proof checker, which we know to be partitioned
-and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes
-the correct ring evaluation of $\inputformula$.
-
-\vspace{1em}\noindent
-\textbf{Model counting.} Finally, we established that ring evaluation with the appropriate weights
-corresponds to the standard model count. To do so, we defined a function that
-carries out an integer calculation of the number of models over a set of variables
-of cardinality \lstinline{nVars}:
-\begin{lstlisting}
-def countModels (nVars : Nat) : PropForm Var → Nat
-  | tr       => 2^nVars
-  | fls      => 0
-  | var _    => 2^(nVars - 1)
-  | neg φ    => 2^nVars - countModels nVars φ
-  | disj φ ψ => countModels nVars φ + countModels nVars ψ
-  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
-\end{lstlisting}
-We then formally proved that for a partitioned formula whose variables are among a finite set
-\lstinline{s}, this computation really does count the number of models over \lstinline{s}:
-\begin{lstlisting}
-theorem countModels_eq_card_models {φ : PropForm Var} {s : Finset Var} :
-  vars φ ⊆ s → partitioned φ → countModels (card s) φ = card (models φ s)
-\end{lstlisting}
-In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
-we have that the number of models of \lstinline{φ} on its variables is
-\lstinline{countModels φ (card (vars φ))}.
+\input{lean-section}
 

### 2024-01-05 doc/JAIR/lean-section.tex
Split Lean section as separate file

--- 2023-05-24 doc/mc-paper.tex
@@
-\documentclass[letterpaper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-%This is a template for producing LIPIcs articles.
-%See lipics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
+%% Separate file for Lean presentation
 
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+\section{A Formally Verified Toolchain}
+\label{sect:formally-verified-toolchain}
+\label{sect:lean:subtle-condition}
 
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
+sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \ref{fig:chain}).
+These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}.
+Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic.
+In this section, we briefly describe the functionality we implemented and what we proved about it.
 
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{tikz}
-\usepackage{pgfplots}
-\usepackage{booktabs}
-\usepackage{hyperref}
+\vspace{1em}\noindent
+\textbf{Proof checking.} The goal of a CPOG proof is to construct a POG that is equivalent to the input CNF $\inputformula$.
+The POG being constructed, and the set of active clauses are stored in the checker state \texttt{st}.
+The checker begins by parsing the input formula, initializing the active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one.
+It then processes every step of the CPOG proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
+Afterwards, it carries out the \textsc{final conditions} check of \ref{subsection:semantics}.
+Throughout the process, we maintain invariants that ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\ref{thm:lean:equiv}).
 
-\newcommand{\pand}{\mathbin{\land^\textsf{p}}}
-\newcommand{\por}{\mathbin{\lor^\textsf{p}}}
-\DeclareMathOperator*{\Pand}{\bigwedge^\textsf{p}}
-\DeclareMathOperator*{\Por}{\bigvee^\textsf{p}}
-\newcommand{\boolnot}{\neg}
-\newcommand{\tautology}{\top}
-\newcommand{\nil}{\bot}
-\newcommand{\obar}[1]{\overline{#1}}
-\newcommand{\oneminus}{{\sim}}
-\newcommand{\lit}{\ell}
+The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \ref{sect:logical:foundations}.
+Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas.
+\lstinline{PropForm} is generic over the type of variables, so we instantiate it with our type as \lstinline{PropForm Var}.
+Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}.
+Requiring totality is not a limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$.
+We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.
 
-\newcommand{\varset}{X}
-\newcommand{\exvarset}{Z}
-\newcommand{\dependencyset}{{\cal D}}
-\newcommand{\litset}{{\cal L}}
-\newcommand{\ring}{{\cal R}}
-\newcommand{\dset}{{\cal A}}
-\newcommand{\rep}{\textbf{R}}
-\newcommand{\radd}{+}
-\newcommand{\rmul}{\times}
-\newcommand{\addident}{\textbf{0}}
-\newcommand{\mulident}{\textbf{1}}
-\newcommand{\imply}{\Rightarrow}
-\newcommand{\ifandonlyif}{\Leftrightarrow}
-%\newcommand{\drational}{\mathbb{Q}_{2,5}}
-\newcommand{\drational}{\textbf{Q}_{2,5}}
-\newcommand{\entail}{\vDash}
-\newcommand{\entaildrat}{\entail_{\textrm{DRAT}}}
+The invariants refer to the checker state \lstinline{st} with fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the clausal POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far.
+For any $\nodeu\in P$, \lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.
+The first two invariants state that assignments to original variables extend uniquely to extension variables defining the POG nodes.
+In the formalization, we split this into extension and uniqueness:
+\begin{lstlisting}
+/-- Any assignment satisfying φ₁ extends to φ₂ while preserving values on X. -/
+def extendsOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
+  ∀ (σ₁ : PropAssignment Var), σ₁ ⊨ φ₁ → ∃ σ₂, σ₁.agreeOn X σ₂ ∧ σ₂ ⊨ φ₂
+/-- Assignments satisfying φ are determined on Y by their values on X. -/
+def uniqueExt (X Y : Set Var) (φ : PropForm Var) :=
+  ∀ (σ₁ σ₂ : PropAssignment Var), σ₁ ⊨ φ → σ₂ ⊨ φ → σ₁.agreeOn X σ₂ →
+    σ₁.agreeOn Y σ₂
 
-\newcommand{\assign}{\alpha}
-\newcommand{\passign}{\rho}
-\newcommand{\lassign}{\beta}
-\newcommand{\uassign}{{\cal U}}
-\newcommand{\modelset}{{\cal M}}
+invariants.extends_pogDefsForm : extendsOver st.inputCnf.vars ⊤ st.pogDefsForm
+invariants.uep_pogDefsForm : uniqueExt st.inputCnf.vars st.allVars st.pogDefsForm
+\end{lstlisting}
+Note that in the definition of \lstinline{uniqueExt}, the arrows associate to the right,
+so the definition says that the three assumptions imply the conclusion.
+The next invariant guarantees that the set of solutions over the original variables is preserved:
+\begin{lstlisting}
+def equivalentOver (X : Set Var) (φ₁ φ₂ : PropForm Var) :=
+  extendsOver X φ₁ φ₂ ∧ extendsOver X φ₂ φ₁
 
-\newcommand{\indegree}{\textrm{indegree}}
-\newcommand{\outdegree}{\textrm{outdegree}}
-\newcommand{\validate}{\textsf{validate}}
-\newcommand{\prov}{\textrm{Prov}}
-\newcommand{\inputformula}{\phi_I}
-\newcommand{\pogformula}{\theta_P}
+invariants.equivInput : equivalentOver st.inputCnf.vars st.inputCnf st.clauseDb
+\end{lstlisting}
+Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\ref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
+\begin{lstlisting}
+def partitioned : PropForm Var → Prop
+  | tr | fls | var _ => True
+  | neg φ    => φ.partitioned
+  | disj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ ∀ τ, ¬(τ ⊨ φ ∧ τ ⊨ ψ)
+  | conj φ ψ => φ.partitioned ∧ ψ.partitioned ∧ φ.vars ∩ ψ.vars = ∅
 
-\newcommand{\makenode}[1]{\mathbf{#1}}
-\newcommand{\nodeu}{\makenode{u}}
-\newcommand{\nodev}{\makenode{v}}
-\newcommand{\nodes}{\makenode{s}}
-\newcommand{\nodep}{\makenode{p}}
-\newcommand{\noder}{\makenode{r}}
+invariants.partitioned : ∀ (u : ILit), (st.pog.toPropForm u).partitioned
+invariants.equivalent_lits : ∀ (u : ILit), equivalentOver st.inputCnf.vars
+    (u ∧ st.pogDefsForm) (st.pog.toPropForm x)
+\end{lstlisting}
+These invariants are maintained by valid CPOG proofs. Together with additional invariants
+that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
+node $\noder$:
 
-\newcommand{\simplify}[2]{#1|_{#2}}
+\begin{thm}
+\label{thm:lean:equiv}
+If the proof checker has assembled POG $P$ with root node $\noder$ starting from input formula $\inputformula$, and \textsc{final conditions} (as stated in Section~\ref{subsection:semantics}) hold of the checker state, then $\inputformula$ is logically equivalent to $\phi_\noder$.
+\end{thm}
+\begin{proof}
+\textsc{Final conditions} imply that the active clausal formula $\theta$ is exactly $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.
+The conclusion follows from this and the checker invariants.
+The full proof is formally verified in Lean.
+\end{proof}
+After certifying a CPOG proof, the checker can pass its in-memory POG representation to the ring evaluator, along with the partitioning guarantee provided by \texttt{invariants.partitioned}.
 
-\newcommand{\progname}[1]{\textsc{#1}}
-\newcommand{\dfour}{\progname{D4}}
-\newcommand{\cdfour}{\progname{CD4}}
-\newcommand{\Dfour}{\progname{D4}}
-\newcommand{\cadical}{\progname{CaDiCal}}
-\newcommand{\dtrim}{\progname{drat-trim}}
-\newcommand{\Dtrim}{\progname{Drat-trim}}
+\vspace{1em}\noindent
+\textbf{Ring evaluation}. We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
+(\ref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
+\begin{lstlisting}
+def weightSum {R : Type} [CommRing R]
+    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
+  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
+\end{lstlisting}
+The rules for efficient ring evaluation of partitioned formulas are expressed as:
+\begin{lstlisting}
+def ringEval (weight : Var → R) : PropForm Var → R
+  | tr       => 1
+  | fls      => 0
+  | var x    => weight x
+  | neg φ    => 1 - ringEval weight φ
+  | disj φ ψ => ringEval weight φ + ringEval weight ψ
+  | conj φ ψ => ringEval weight φ * ringEval weight ψ
+\end{lstlisting}
+\ref{prop:ring:eval} is then formalized as follows:
+\begin{lstlisting}
+theorem ringEval_eq_weightSum (weight : Var → R) {φ : PropForm Var} :
+    partitioned φ → ringEval weight φ = weightSum weight φ (vars φ)
+\end{lstlisting}
+To efficiently compute the ring evaluation of a formula represented by a POG node, we implemented
+\lstinline{Pog.ringEval} and then proved that it matches the specification above:
+\begin{lstlisting}
+theorem ringEval_eq_ringEval (pog : Pog) (weight : Var → R) (x : Var) :
+  pog.ringEval weight x = (pog.toPropForm x).ringEval weight
+\end{lstlisting}
+Applying this to the output of our verified CPOG proof checker, which we know to be partitioned
+and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes
+the correct ring evaluation of $\inputformula$.
 
-\definecolor{redorange}{rgb}{0.878431, 0.235294, 0.192157}
-\definecolor{lightblue}{rgb}{0.552941, 0.72549, 0.792157}
-\definecolor{clearyellow}{rgb}{0.964706, 0.745098, 0}
-\definecolor{clearorange}{rgb}{0.917647, 0.462745, 0}
-\definecolor{mildgray}{rgb}{0.54902, 0.509804, 0.47451}
-\definecolor{softblue}{rgb}{0.643137, 0.858824, 0.909804}
-\definecolor{bluegray}{rgb}{0.141176, 0.313725, 0.603922}
-\definecolor{lightgreen}{rgb}{0.709804, 0.741176, 0}
-\definecolor{darkgreen}{rgb}{0.152941, 0.576471, 0.172549}
-\definecolor{redpurple}{rgb}{0.835294, 0, 0.196078}
-\definecolor{midblue}{rgb}{0, 0.592157, 0.662745}
-\definecolor{clearpurple}{rgb}{0.67451, 0.0784314, 0.352941}
-\definecolor{browngreen}{rgb}{0.333333, 0.313725, 0.145098}
-\definecolor{darkestpurple}{rgb}{0.396078, 0.113725, 0.196078}
-\definecolor{greypurple}{rgb}{0.294118, 0.219608, 0.298039}
-\definecolor{darkturquoise}{rgb}{0, 0.239216, 0.298039}
-\definecolor{darkbrown}{rgb}{0.305882, 0.211765, 0.160784}
-\definecolor{midgreen}{rgb}{0.560784, 0.6, 0.243137}
-\definecolor{darkred}{rgb}{0.576471, 0.152941, 0.172549}
-\definecolor{darkpurple}{rgb}{0.313725, 0.027451, 0.470588}
-\definecolor{darkestblue}{rgb}{0, 0.156863, 0.333333}
-\definecolor{lightpurple}{rgb}{0.776471, 0.690196, 0.737255}
-\definecolor{softgreen}{rgb}{0.733333, 0.772549, 0.572549}
-\definecolor{offwhite}{rgb}{0.839216, 0.823529, 0.768627}
-\definecolor{medgreen}{rgb}{0.15, 0.6, 0.15}
-
-% Lean code:
-\usepackage{listings}
-%\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
-\definecolor{keywordcolor}{rgb}{0.0, 0.1, 0.6}   % blue
-\definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
-\definecolor{commentcolor}{rgb}{0.4, 0.4, 0.4}   % grey
-\definecolor{symbolcolor}{rgb}{0.0, 0.1, 0.6}    % blue
-\definecolor{sortcolor}{rgb}{0.1, 0.5, 0.1}      % green
-\definecolor{attributecolor}{rgb}{0.7, 0.1, 0.1} % red
-\def\lstlanguagefiles{lstlean.tex}
-% set default language
-\lstset{language=lean, xleftmargin=1em}
-\lstset{backgroundcolor=\color{white}}
-
-\bibliographystyle{plainurl}% the mandatory bibstyle
-
-\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting}
-
-\titlerunning{Certified Knowledge Compilation}
-
-\author{Randal E.
-Bryant}{Computer Science Department, Carnegie Mellon University, Pittsburgh, PA 15213 USA}{Randy.Bryant@cs.cmu.edu}{https://orcid.org/0000-0001-5024-6613}{Supported by NSF grant CCF-2108521}
-\author{Wojciech Nawrocki}{Department of Philosophy, Carnegie Mellon University}{wjnawrocki@cmu.edu}{https://orcid.org/0000-0002-8839-0618}{Hoskinson Center for Formal Mathematics}
-\author{Jeremy Avigad}{Department of Philosophy, Carnegie Mellon University}{avigad@cmu.edu}{https://orcid.org/0000-0003-1275-315X}{Hoskinson Center for Formal Mathematics}
-\author{Marijn J. H. Heule}{Computer Science Department, Carnegie Mellon University}{marijn@cmu.edu}{https://orcid.org/0000-0002-5587-8801}{Supported by NSF grant CCF-2108521}
-
-\authorrunning{R. E. Bryant et al.} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Randal E.
-Bryant, Wojciech Nawrocki, Jeremy Avigad, and Marijn J.
-H.
-Heule} %TODO mandatory, please use full first names.
-LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-\begin{CCSXML}
-<ccs2012>
-   <concept>
-       <concept_id>10003752.10003790.10003794</concept_id>
-       <concept_desc>Theory of computation~Automated reasoning</concept_desc>
-       <concept_significance>500</concept_significance>
-       </concept>
- </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Theory of computation~Automated reasoning}
-
-%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
-\keywords{Propositional model counting, Proof checking}
-
-%\category{} %optional, e.g. invited paper
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-\nolinenumbers %uncomment to disable line numbering
-
-\EventEditors{Meena Mahajan and Friedrich Slivovsky}
-\EventNoEds{2}
-\EventLongTitle{26th International Conference on Theory and Applications of Satisfiability Testing (SAT 2023)}
-\EventShortTitle{SAT 2023}
-\EventAcronym{SAT}
-\EventYear{2023}
-\EventDate{July 4--8, 2023}
-\EventLocation{Alghero, Italy}
-\EventLogo{}
-\SeriesVolume{271}
-\ArticleNo{6}
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\EventEditors{John Q. Open and Joan R. Access}
-%\EventNoEds{2}
-%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-%\EventShortTitle{CVIT 2016}
-%\EventAcronym{CVIT}
-%\EventYear{2016}
-%\EventDate{December 24--27, 2016}
-%\EventLocation{Little Whinging, United Kingdom}
-%\EventLogo{}
-%\SeriesVolume{42}
-%\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\program}[1]{\textsc{#1}}
-\newcommand{\lean}{Lean~4}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-Computing many useful properties of Boolean formulas, such as their weighted or unweighted model count,
-is intractable on general representations. It can become tractable when formulas are expressed in a
-special form, such as the decision-decomposable, negation normal form (dec-DNNF)\@.
-\emph{Knowledge compilation} is the process of converting a formula
-into such a form. Unfortunately existing knowledge compilers provide no guarantee that their output correctly
-represents the original formula, and therefore they cannot validate a model count, or any other computed value.
-
-We present \emph{Partitioned-Operation Graphs} (POGs), a form that can
-encode all
-of the representations used by existing knowledge compilers.
-We have designed  CPOG, a framework that can express proofs of equivalence between a
-POG  and a Boolean formula in conjunctive normal form (CNF).
-
-We have developed a program that generates POG representations from dec-DNNF
-graphs
-produced by the state-of-the-art knowledge compiler
-\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
-are equivalent to the input CNF formulas.  Our toolchain
-for generating and verifying POGs scales to all but the largest
-graphs produced by \dfour{} for formulas from a recent model counting
-competition. Additionally, we have developed a formally verified CPOG
-checker and model counter for POGs in the \lean{} proof assistant.
-In doing so, we proved the soundness of our proof framework. These programs
-comprise the first formally verified toolchain for weighted and unweighted
-model counting.
-\end{abstract}
-
-\section{Introduction}
-
-Given a Boolean formula $\phi$, modern Boolean satisfiability (SAT) solvers can
-find an assignment satisfying $\phi$ or generate a proof that no
-such assignment exists.  They have applications across a variety of
-domains including computational mathematics, combinatorial
-optimization, and the formal verification of hardware, software, and
-security protocols.  Some applications, however, require going
-beyond Boolean satisfiability.  For example, the \emph{model
-  counting problem} requires computing the number of satisfying
-assignments of a formula, including in cases where there are far too many
-to enumerate individually.  Model counting has
-applications in artificial intelligence, computer security, and
-statistical sampling.  There are also many useful extensions of standard model counting,
-including {\em
-  weighted model counting}, where a weight is defined for
-each possible assignment, and the goal becomes to compute the sum of the weights
-of the satisfying assignments.
-
-Model counting is a challenging problem---more challenging than the
-already NP-hard Boolean satisfiability.  Several
-tractable variants of Boolean satisfiability, including 2-SAT, become
-intractable when the goal is to count models and not just determine
-satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
-model counters that scale to very large formulas have been developed, as
-witnessed by the progress in recent model counting competitions.
-
-One approach to model counting, known as \emph{knowledge compilation},
-transforms the formula into a structured form for which model counting
-is straightforward.  For example, the \emph{deterministic, decomposable, negation normal form}
-(det-DNNF) introduced by
-Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
-\emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
-represent a
-Boolean formula as a directed acyclic graph, with terminal nodes
-labeled by Boolean variables and their complements, and with each
-nonterminal node labeled by a Boolean And or Or operation.  Restrictions
-are placed on the structure of the graph (described in Section~\ref{sect:pog}) such that a count of the
-models can be computed by a single bottom-up traversal.
-Kimmig et al.\ present a very general {\em
-  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
-properties of Boolean functions that can be efficiently computed from
-a det-DNNF representation.  These include standard and weighted model
-counting, and much more.
-
-One shortcoming of existing knowledge compilers is that they have no
-generally accepted
-way to validate that
-the compiled representation is logically equivalent to the original
-formula.  By contrast, all modern SAT solvers can generate
-checkable proofs when they encounter unsatisfiable formulas.  The
-guarantee provided by a checkable certificate of correctness enables
-users of SAT solvers to fully trust their results.  Experience has also
-shown that being able to generate proofs allow SAT solver developers to quickly
-detect and diagnose bugs in their programs. This, in turn, has led
-to more reliable SAT solvers.
-
-This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
-a form that can encode all of the representations produced by current knowledge
-compilers. The CPOG (for ``certified'' POG) file format then
-captures both the structure of a POG
-and a checkable proof of its logical equivalence to a Boolean formula in
-conjunctive normal form (CNF).  A CPOG
-proof consists of a sequence of clause addition and deletion steps,
-based on an extended resolution proof system~\cite{Tseitin:1983}.
-We establish a set of conditions that, when satisified by a CPOG file, guarantees that it
-encodes a well-formed POG and provides a valid equivalence proof.
-
-\begin{figure}
-\centering{
-\begin{tikzpicture}[scale=0.17]
-  \draw (8,10) rectangle (20,18);
-  \node at (14,15.2) {Knowledge};
-  \node at (14,12.8) {Compiler};
-
-  \draw (28,10) rectangle (40,18);
-  \node at (34,15.2) {Proof};
-  \node at (34,12.8) {Generator};
-
-  \draw (50,10) rectangle (62,18);
-  \node at (56,15.2) {Proof};
-  \node at (56,12.8) {Checker};
-
-  \draw (50,0) rectangle (62,8);
-  \node at (56,5.2) {Ring};
-  \node at (56,2.8) {Evaluator};
-
-  \draw[dashed] (48,-2) rectangle (64,22);
-  \node at (56,20) {Trusted Code};
-
-  %% CNF
-  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (50,16);
-  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
-  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
-  \node [left] at (-1,20) {\large {$\inputformula$}};
-  \node [above] at (2,20) {\texttt{.cnf}};
-
-  %% NNF
-  \draw[thick] (20,12) [-latex] -- (28,12);
-  \node [above] at (24,12) {\texttt{.ddnnf}};
-
-  %% CPOG
-  \draw[thick] (40,12) [-latex] -- (50,12);
-  \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
-  \node [above] at (44,12) {\texttt{.cpog}};
-
-  %% OK/Not
-  \draw[thick] (62,14) [-latex] -- (66,14) ;
-  \node [right] at (66,15.2) { OK /} ;
-  \node [right] at (66,12.8) { Not OK} ;
-  \draw[thick] (62,4) [-latex] -- (66,4) ;
-  \node [right] at (66,4) {\large {$\textbf{R}(\inputformula)$}};
-
-\end{tikzpicture}
-}
-\caption{Certifying toolchain.
-  The output of a standard knowledge compiler is converted into a combined graph/proof (CPOG)
-  which can be independently checked and evaluated}
-\label{fig:chain}
-\end{figure}
-
-Figure~\ref{fig:chain} illustrates our certifying knowledge
-compilation and model counting toolchain.  Starting with
-input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
-generates a dec-DNNF representation,
-and the \emph{proof
-  generator} uses this to generate a CPOG file.
-The \emph{proof checker} verifies the equivalence of the CNF and CPOG representations.
-The \emph{ring evaluator} computes
-a standard or weighted model count from the POG\@ representation.
-As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
-moves the root of trust away from the complex and highly
-optimized knowledge compiler to a relatively simple checker and
-evaluator.  Importantly, the proof generator need not be
-trusted---its errors will be caught by the proof checker.
-
-To ensure soundness of the abstract CPOG proof system, as well as
-correctness of its concrete implementation, we formally verified the
-proof system as well as versions of the proof checker and ring
-evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
-Running these two programs on a  CPOG file gives strong
-assurance that the proof and the model count are correct. Our
-experience with developing a formally verified proof checker has shown
-that, even within the well-understood framework of extended
-resolution, it can be challenging to formulate a full set of
-requirements that guarantee soundness.  In fact, our efforts to
-formally verify our proof framework exposed subtle conditions that we had
-to impose on our partitioned sum rule.
-
-We evaluate our toolchain using benchmark formulas from the 2022
-standard and weighted model competitions.  Our tools handle all but
-the largest dec-DNNF graphs generated by \dfour{}.  We also measure
-the benefits of several optimizations as well as the relative
-performance of the verified checker with one designed for high
-performance and capacity.
-
-We also show that our tools can provide end-to-end verification of
-formulas that have been transformed by an equivalence-preserving
-preprocessor.  That is, verification is based on the original formula,
-and so proof checking certifies correct operation of the preprocessor,
-the knowledge compiler, and the proof generator.
-
-We have developed an online supplement to this
-paper~\cite{bryant:sat:2023:supplement} containing additional
-information, including a worked example, details on the algorithms,
-and extensive experimental results.  The archive also includes copies
-of the programs and a set of benchmark files.
-
-\section{Related Work}
-
-Generating proofs of unsatisfiability in SAT solvers has a long
-tradition~\cite{ZhangMalik} and has become widely accepted due to the
-formulation of clausal proof systems for which proofs can readily be
-generated and efficiently checked
-\cite{heule:cade:2013,wetzler14_drattrim}.
-A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
-The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
-Our work builds on the well-established technology and tools associated with clausal proof systems,
-but we require features not found in proofs of unsatisfiability.
-Our proofs construct a new propositional formula, and we must verify
-%both
-that it
-%satisfies a set of rules and that it
-is equivalent to the input formula.  This requires verifying
-additional proof steps, including clause deletion steps, and subtle
-invariants, as described in Sections~\ref{sect:cpog} and
-\ref{section:formally-verified-toolchain}.
-
-Capelli, Lagniez, and Marquis developed a knowledge compiler that
-generates a certificate in a proof system that is itself based on
-dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{}
-program, a modified version of \dfour{}, generates annotations to the
-compiled representation, providing information about how the compiled
-version relates to the input clauses.  It also generates a file of
-clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
-Completing the certification involves running two different checkers
-on the annotated dec-DNNF graph and the DRAT file.  Although the
-authors make informal arguments regarding the soundness of their
-frameworks, these do not provide strong levels of assurance.  Indeed,
-we have identified a weakness in their methodology due to an
-invalid assumption about the guarantees provided by \dtrim{}, the
-program it uses to check the DRAT file.  This weakness is
-\emph{exploitable}: their framework can be ``spoofed'' into accepting
-an incorrect compilation.
-
-In more detail, \cdfour{} emits a sequence of clauses $R$ that
-includes the conflict clauses that arose during a top-down processing of
-the input clauses.  Given input formula $\phi_I$, their first task is to
-check whether $\phi_I \imply R$, i.e., that any assignment that satisfies $\phi_I$ also satisfies each of the clauses in $R$.
-They then base other parts of their proof on that
-property and use a separate program to perform a series of additional
-checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
-satisifies the \emph{reverse asymmetric tautology} (RAT) property with
-respect to the preceding
-clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
-clause $C$ to a set of clauses maintains satisfiability, but it does not necessarily
-preserve models.  As an example, consider the following formulas:
-\begin{center}
-  \begin{tabular}{lccc}
-    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
-    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_2 \lor \obar{x}_3)$\\
-  \end{tabular}
-\end{center}
-Clearly, these two formulas are not equivalent---$\phi_1$ has six
-models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
-arbitrary assignments to variable $x_2$.  Critically, however, the
-second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
-satisfying assignment to $\phi_1$ can be transformed into one that
-also satisfies $\phi_2$ by setting $x_2$ to 1, while keeping the values for other variables fixed.
-
-This weakness would allow a buggy (or malicious) version of \cdfour{}
-to spoof the checking framework.  Given formula $\phi_1$ as input, it
-could produce a compiled result, including annotations, based on $\phi_2$ and also include the second clause of $\phi_2$ to $R$.
-The check with \dtrim{} would pass, as would the other tests performed
-by their checker.  We have confirmed this possibility with
-their compiler and checker.\footnote{Downloaded May 18, 2023 as\\
-\url{https://github.com/crillab/d4/tree/333370cc1e843dd0749c1efe88516e72b5239174}.}
-
-This weakness can be corrected by restricting \dtrim{} to only add
-clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
-command-line argument to \dtrim{} that enforces this
-restriction.\footnote{Available at
-\url{https://github.com/marijnheule/drat-trim/releases/tag/v05.22.2023}.}  This weakness, however, illustrates the general challenge of
-developing a new proof framework.
-As we can attest,
-without engaging in an effort to formally verify the framework, there are likely to be
-conditions that make the framework unsound.
-
-Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
-proof framework for model counting programs.  Their proof rules are
-based on the algorithms commonly used by model counters.  They
-developed a program that can generate proof traces from dec-DNNF
-graphs and a program to check adherence to their proof rules.  This
-framework is not directly comparable to ours, since it only certifies
-the unweighted model count, but it has similar goals.
-Again, they provide only  informal arguments
-regarding the soundness of their framework.
-
-Both of these prior certification frameworks are strongly tied to the
-algorithms used by the knowledge compilers and model counters.  Some
-of the conditions to be checked are relevant only to specific
-implementations.    Our framework is very general and is based on a small set
-of proof rules.  It builds on the highly developed
-concepts of clausal proof systems.  These factors were important in enabling formal verification.
-In Section~\ref{sect:experimental} and the supplement~\cite{bryant:sat:2023:supplement},
-we also compare the performance of our toolchain to these other two.
-
-\section{Logical Foundations}
-\label{section:logical:foundations}
-
-  Let $\varset$ denote a set of Boolean variables, and let $\assign$
-  be an \emph{assignment} of truth values to some subset of the
-  variables, where $0$ denotes false and $1$ denotes true, i.e.,
-  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
-  \subseteq \varset$.  We say the assignment is \emph{total} when it
-  assigns a value to every variable ($\varset' = \varset$), and that
-  it is \emph{partial} otherwise.
-  The set of all possible total assignments over
-  $\varset$ is denoted $\uassign$.
-
-For each variable $x \in \varset$,
-  we define the \emph{literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
-  negation of $x$. An
-  assignment $\assign$ can be viewed as a set of literals, where
-  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
-  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
-$\obar{\lit} = x$ when $\lit = \obar{x}$.
-
-\begin{definition}[Boolean Formulas]
-  The set of Boolean formulas is defined recursively.  Each
-  formula $\phi$ has an associated \emph{dependency set}
-  $\dependencyset(\phi)  \subseteq \varset$, and a set of \emph{models} $\modelset(\phi)$,
-  consisting of total assignments that satisfy the formula:
-  \begin{enumerate}
-  \item Boolean constants $0$ and $1$ are Boolean formulas,
-    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
-  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
-    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
-  \item For formula $\phi$, its \emph{negation}, written $\boolnot \phi$ is a Boolean formula,
-    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
-  \end{enumerate}
-\label{def:boolean}
-\end{definition}
-
-  We highlight some special classes of Boolean formulas.  A formula is
-  in \emph{negation normal form} when negation is applied only to variables.  A
-  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
-  negation normal form, and ii) sum is applied only to literals.  A CNF
-  formula can be represented as a set of \emph{clauses}, each of which is a
-  set of literals.  Each clause represents the sum of the
-  literals, and the formula is the product of its clauses.  We use
-  set notation to reference the clauses within a formula and the
-  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
-This literal must be assigned value $1$ by any satisfying assignment of the formula.
-
-\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
-  A \emph{partitioned-operation formula}
- satisfies the following for all product and sum operations:
-      \begin{enumerate}
-      \item The arguments to each product must have disjoint dependency sets.  That is, operation
-        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \item The arguments to each sum must have disjoint models.  That is, operation
-        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \end{enumerate}
-\end{definition}
-     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.
-
-  \section{Ring Evaluation of a Boolean Formula}
-
-We propose a general framework for summarizing properties of Boolean
-formulas along the lines of algebraic model counting~\cite{kimmig:jal:2017}.
-
-\begin{definition}[Commutative Ring]
-  A \emph{commutative ring} $\ring$ is an algebraic structure
-  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
-  with elements in the set $\dset$ and with commutative and
-  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
-  such that multiplication distributes
-  over addition.  $\addident$ is the additive identity and $\mulident$ is
-  the multiplicative identity.  Every element $a \in \dset$ has an
-  \emph{additive inverse} $-a$ such that $a + -a = \addident$.
-\label{def:ring}
-\end{definition}
-We write $a - b$ as a shorthand for $a + -b$.
-
-\begin{definition}[Ring Evaluation Problem]
-\label{def:ring_evaluation}
-  For commutative ring $\ring$, a \emph{ring weight function} associates a value $w(x) \in \dset$ with
-  every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.
-
-  For Boolean formula $\phi$ and ring weight function $w$, the \emph{ring evaluation problem} computes
-  \begin{equation}
-    \begin{array}{rcl}
-    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
-    \end{array}
-  \end{equation}
-  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
-\label{def:weight}
-\end{definition}
-
-Many important properties of Boolean formulas can be
-expressed as ring evaluation problems.  The
-(standard) \emph{model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
-It can be cast as a ring evaluation problem by having $\radd$ and
-$\rmul$ be addition and multiplication over rational numbers and using
-weight function $w(x) = 1/2$ for every variable $x$.
-Ring evaluation of formula $\phi$ gives the \emph{density} of
-the formula, i.e., the fraction of all possible total assignments that are
-models.  For $n = |\varset|$, scaling the density by $2^n$
-yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
-in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.
-
-The \emph{weighted model counting}  problem is also defined over
-rational numbers.  Some formulations  allow
-independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
-$W(x) + W(\obar{x}) \not = 1$.
-We can cast this as a
-ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
-performing ring evaluation with weight function $w(x) = W(x)/r(x)$ for each
-variable $x$, and computing the weighted count
-as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
-Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.
-
-The \emph{function hashing problem} provides a test
-of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
-finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, choose a value from $\dset$ at random for $w(x)$.  Two formulas
-$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
-if they are logically equivalent,
-and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
-If they are not equivalent, then
-the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
-least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
-Function hashing can therefore be used as part of a
-randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
-For example, it can compare different runs on a single formula,
-either from different compilers or from a single compiler with different configuration parameters.
-
-%We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.
-
-\section{Partitioned-Operation Graphs (POGs)}
-\label{sect:pog}
-
-Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
-\begin{proposition}
-\label{prop:ring:eval}
-Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
-\begin{displaymath}
-\begin{array}{rcl}
-\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \\
-\textstyle
-\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \\
-\textstyle
-\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \\
-\end{array}
-\end{displaymath}
-\end{proposition}
-As is described in \cref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.
-
-A \emph{partitioned-operation graph} (POG) is a directed, acyclic
-graph with nodes $N$ and edges $E \subseteq N \times N$.  We denote
-nodes with boldface symbols, such as $\nodeu$ and $\nodev$.  When
-$(\nodeu,\nodev) \in E$, node $\nodev$ is said to be a \emph{child} of
-node $\nodeu$.  The in- and out-degrees of node $\nodeu$ are defined
-as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
-$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.  Node
-$\nodeu$ is said to be \emph{terminal} if $\outdegree(\nodeu) = 0$.  A
-terminal node is labeled by a Boolean constant or variable.  Node
-$\nodeu$ is said to be \emph{nonterminal} if $\outdegree(\nodeu) > 0$.
-A nonterminal node is labeled by Boolean operation $\pand$ or $\por$.
-A node can be labeled with operation $\pand$ or $\por$ only if it
-satisfies the partitioning restriction for that operation.  Every POG
-has a designated \emph{root node} $\noder$.  Each edge has
-a \emph{polarity}, indicating whether (negative polarity) or not
-(positive polarity) the corresponding argument should be negated.
-
-A POG represents a partitioned-operation
-formula with a sharing of common subformulas.  Every node in the graph can be viewed as a partitioned-operation formula, and so we write
-$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
-Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.
-
-We can now define and compare two related representations:
-\begin{itemize}
-\item A det-DNNF graph can be viewed a POG with negation applied only to variables.
-\item A dec-DNNF graph is a det-DNNF graph with the further restriction that any
-  sum node $\nodeu$ has exactly two children $\nodeu_1$ and $\nodeu_0$, and for these there is a \emph{decision variable} $x$ such that
-  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
-\end{itemize}
-The generalizations encompassed by POGs have also been referred to as \emph{deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
-Our current proof generator only works for knowledge compilers
-generating dec-DNNF representations, but these generalizations
-allow for future extensions, while maintaining the ability to
-efficiently perform ring evaluation.
-
-We define the \emph{size} of POG $P$, written $|P|$, to be the
-the number of nonterminal nodes plus the number of edges from these nodes to their children.  Ring
-evaluation of $P$ can be performed with at most $|P|$ ring
-operations by traversing the graph from the terminal nodes up to
-the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
-The final result is then $\rep(\noder, w)$.
-
-\section{Clausal Proof Framework}
-
-We write (possibly subscripted) $\theta$ for
-formulas encoded as clauses, possibly with extension variables.
-We write (possibly subscripted)
-$\phi$ for formulas
-that use no extension variables.
-
-A proof in our framework consists of a sequence of clause addition
-and deletion steps, with each step preserving the set of solutions to
-the original formula.
-The status of the proof at any step is represented as
-a set of \emph{active} clauses $\theta$, i.e., those that
-have been added but not yet deleted.
-Our framework is based
-on \emph{extended} resolution~\cite{Tseitin:1983}, where proof
-steps can introduce new \emph{extension variables} encoding Boolean formulas over input and prior extension variables.
-Let $Z$
-denote the set of extension variables occuring in formula $\theta$.
-Starting with  $\theta$ equal to input formula $\inputformula$,
-the proof must maintain the invariant that
-$\inputformula \ifandonlyif \exists Z\,\theta$.
-
-Clauses can be added in two different ways.  One is when they serve as
-the \emph{defining clauses} for an extension variable.  This form
-occurs only when defining $\pand$ and $\por$ operations, as is
-described in Section~\ref{sect:cpog}.  Clauses can also be added or
-deleted based on \emph{implication redundancy}.  That is, when clause
-$C$ satisfies $\theta \imply C$ for formula $\theta$, then it can either
-be added to $\theta$ to create the formula $\theta \cup \{C\}$ or it can be deleted
-from $\theta \cup \{C\}$ to create $\theta$.
-
-We use \emph{reverse unit propagation} (RUP) to certify
-implication redundancy when adding or deleting
-clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
-RUP
-is the core rule supported by standard
-proof checkers~\cite{heule:cade:2013,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
-way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
-Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
-proved redundant with respect to formula $\theta$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
-\emph{antecedent} clauses, such that each $D_i$ is in $\theta$.
-A RUP step
-proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
-that the combination of the antecedents plus the negation of $C$ leads
-to a contradiction.  The negation of $C$ is the formula
-$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
-\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
-clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
-check processes the clauses of the antecedent in sequence, inferring
-additional unit clauses.  In processing clause $D_i$, if all but one
-of the literals in the clause is the negation of one of the
-accumulated unit clauses, then we can add this literal to the
-accumulated set.  That is, all but this literal have been falsified,
-and so it must be set to true for the clause to be satisfied.  The
-final step with clause $D_k$ must cause a contradiction, i.e., all of
-its literals are falsified by the accumulated unit clauses.
-
-Compared to the proofs of unsatisfiability generated by SAT solvers,
-ours have important differences.  Most
-significantly, each proof step must preserve the set of solutions with respect to the input variables;
-our proofs must therefore justify both clause deletions and additions.
-By contrast, an unsatisfiability proof need only guarantee that
-no proof step causes a satisfiable set of clauses to become
-unsatisfiable, and therefore it need only justify clause additions.
-
-\section{The CPOG Representation and Proof System}
-\label{sect:cpog}
-
-%% The proof system is based on extended
-%% resolution~\cite{Tseitin:1983} with reverse unit propagation
-%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
-%% proving that a set of clauses logically implies another clause.
-%% Alternate version
-%% knowledge compilation.  The proof system is based on extended
-%% resolution~\cite{Tseitin:1983} with reverse unit propagation
-%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
-%% proving that a set of clauses logically implies another clause.
-%%
-%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
-%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
-%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
-%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
-%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
-%%
-%% \begin{center}
-%%   \begin{tabular}{lcccc}
-%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
-%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
-%%   \midrule
-%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
-%%   \end{tabular}
-%% \end{center}
-%%
-%% RUP is an alternative formulation of resolution.  For target clause
-%% $C$, it can be seen that applying resolution operations to the
-%% antecedent clauses from right to left will derive a clause $C'$ such
-%% that $C' \subseteq C$.  By \emph{subsumption} %~\cite{philipp:lpar:2017},
-%% we then have $C' \rightarrow C$.  Compared to listing each resolution
-%% operation as a separate step, using RUP as the basic proof step makes
-%% the proofs more compact.
-%%
-%% As is shown in Figure~\ref{fig:chain}
-%% two programs are involved in producing a certified compilation of a formula:
-%% \begin{itemize}
-%% \item The \emph{proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
-%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a dec-DNNF representation of the formula.
-%% \item The \emph{proof checker} ensures that all of the proof conditions are satisfied.
-%% \end{itemize}
-%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b
-
-A CPOG file provides both a declaration of a POG, as well as a checkable
-proof that a Boolean formula, given in conjunctive normal
-form, is logically equivalent to the POG\@.
-The proof format draws its inspiration from the LRAT~\cite{lrat} and
-QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
-Key properties include:
-\begin{itemize}
-  \item
-  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
-  Declaring a node $\nodeu$ implicitly adds an \emph{extension} variable $u$ and a set of \emph{defining} clauses $\theta_{u}$
-  encoding the product or sum operation.
-  This is the only means for adding extension variables to the proof.
-\item Boolean negation is supported implicitly by allowing the
-  arguments of the $\por$ and $\pand$ operations to be literals and not just
-  variables.
-\item
-  The file contains explicit clause addition steps.
-  A clause can only be added if it is logically implied by the existing clauses.
-  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
-\item
-  The file contains explicit clause deletion steps.
-  A clause can only be deleted if it is logically implied by the remaining clauses.
-  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
-\item The checker must track the dependency set for every input and
-  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
-  The associated extension variable has a dependency set equal to the union of those of its arguments.
-\item Declaring a $\por$ operation requires a sequence of clauses
-  providing a RUP proof that the arguments are mutually exclusive.
-  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness
-%  Generalizing to $k$ arguments would require
-%  $k\,(k-1)/2$ proofs of disjointness.
-\end{itemize}
-
-\subsection{Syntax}
-\label{subsection:syntax}
-
-\begin{table}
-  \caption{CPOG Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
-  \label{tab:cpog:syntax}
-\centering{
-  \begin{tabular}{lllll}
-    \toprule
-    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
-    \midrule
-    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\texttt{a}}  & \makebox[15mm][l]{$L^{*}$ \texttt{0}} & \makebox[15mm][l]{$C^{+}$ \texttt{0}}  & \makebox[20mm][l]{Add RUP clause} \\
-     & \texttt{d} & $C$             & $C^{+}$  \texttt{0} & Delete RUP clause \\
-    \midrule
-    $C$    & \texttt{p} & $V \; L^{*}$ \texttt{0}    &                  & Declare $\pand$ operation \\
-    $C$    & \texttt{s} & $V \; L \; L$    & $C^{+}$ \texttt{0}  & Declare $\por$ operation \\
-    \midrule
-     & \texttt{r} & $L$             &            & Declare root literal\\
-    \bottomrule
-  \end{tabular}
-  }
-\end{table}
-
-Table~\ref{tab:cpog:syntax} shows the declarations that can occur in a CPOG file.
-%% The checker is provided with the input formula as a separate file.
-As with other clausal proof formats, a variable is
-represented by a positive integer $v$, with the first ones being input
-variables and successive ones being extension variables.  Literal $\lit$
-is represented by a signed integer, with $-v$ being the logical negation of
-variable $v$.  Each clause is indicated by a positive integer
-identifier $C$, with the first ones being the IDs of the input clauses and successive
-ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
-such that any clause identifier $C'$ given in the hint when adding clause $C$ must have $C' < C$.
-
-The first set of proof rules are similar to those in other clausal
-proofs.
-Clauses can be added via RUP addition
-(command \texttt{a}), with a sequence of antecedent clauses (the
-``hint'').
-Similarly for clause deletion (command \texttt{d}).
-
-\begin{table}
-\caption{Defining Clauses for Product (A) and Sum (B) Operations}
-\begin{minipage}{0.54\textwidth}
-\begin{center}
-\begin{tabular}{cccccc}
-\multicolumn{6}{c}{(A).  Product Operation $\pand$}\\
-\toprule
-\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
-\midrule
-  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
-  $i\!+\!1$ & $-v$ & $\lit_1$  \\
-  $i\!+\!2$ & $-v$ & $\lit_2$  \\
-  & $\ldots$ \\
-  $i\!+\!k$ & $-v$ & $\lit_k$  \\
-\bottomrule
-\end{tabular}
-\end{center}
-\end{minipage}
-\begin{minipage}{0.44\textwidth}
-\begin{center}
-\begin{tabular}{cccc}
-\multicolumn{4}{c}{(B).  Sum Operation $\por$}\\
-\toprule
-\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
-\midrule
-  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
-  $i\!+\!1$ & $v$ & $-\lit_1$ \\
-  $i\!+\!2$ & $v$ & $-\lit_2$ \\
-\bottomrule
-$\;$ \\
-$\;$ \\
-\end{tabular}
-\end{center}
-\end{minipage}
-\label{tab:defining}
-\end{table}
-
-The declaration of a \emph{product} operation, creating a node with operation $\pand$,
- has the form:
-\begin{center}
-\begin{tabular}{ccccccccc}
-  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
-  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\texttt{0}} \\
-\end{tabular}
-\end{center}
-Integer $i$ is a new clause ID, $v$ is a positive integer that does not
-correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$
-integers, indicating the arguments as literals of existing variables.
-As Table~\ref{tab:defining}(A) shows,
-this declaration implicitly causes $k+1$ clauses to be added to the proof, providing a Tseitin encoding that defines extension variable $v$ as the product of its arguments.
-
-The dependency sets for the arguments represented by each pair of
-literals $\lit_i$
-and $\lit_{j}$ must
-be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
-representing Boolean constant $1$.  The only clause added to the proof will be
-the unit literal $v$.  A reference to literal $-v$ then provides a way
-to represent constant $0$.
-
-The declaration of a \emph{sum} operation, creating a node with operation $\por$, has the form:
-\begin{center}
-\begin{tabular}{ccccccc}
-  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$}
-\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
-\end{tabular}
-\end{center}
-Integer $i$ is a new clause ID, $v$ is a positive integer that does
-not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
-signed integers, indicating the arguments as literals of existing variables.  Hint $H$
-consists of a
-sequence of clause IDs, all of which must be defining clauses for other POG operations.\footnote{The restriction to defining clauses in the hint is critical to soundness.
-Allowing the hint to include the IDs of input clauses creates an exploitable weakness.  We discovered this weakness in the course of our efforts at formal verification.}
-As Table~\ref{tab:defining}(B) shows,
-this declaration implicitly causes three clauses to be added to the proof, providing a Tseitin encoding that defines extension variable $v$ as the sum of its arguments.
-The hint must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.
-
-Finally, the literal denoting the root of the POG is declared with the
-\texttt{r} command.  It can occur anywhere in the file.  Except in degenerate cases, it
-will be the extension variable representing the root of a graph.
-%, but in
-%degenerate cases it can be the positive or negative literal of an
-%input variable.
-
-\subsection{Semantics}
-\label{subsection:semantics}
-
-The defining clauses for a product or sum
-operation uniquely define the value of its extension variable for any assignment of values to the argument variables.
-For the
-extension variable $u$ associated with any POG node $\nodeu$, we can therefore
-prove that any total assignment $\assign$ to the input variables that
-also satisfies the POG defining clauses will
-assign a value to $u$ such that $\assign(u) =
-1$ if and only if $\assign \in \modelset(\nodeu)$.
-
-%% A CPOG proof follows the same general form as a QRAT dual
-%% proof~\cite{bryant:cade:2021}---one that ensures that each clause
-%% addition and each clause deletion preserves equivalence.  With CPOG,
-%% however, clauses are defined both explicitly and implicitly.  Starting
-%% with the set of input clauses, the proof consists of a sequence of
-%% steps that both add and delete clauses.  Each addition must be truth
-%% preserving, that is, any satisfying total assignment for the set of clauses
-%% before clause addition should still be a satisfying assignment afterwards.
-%% Each deletion must be falsehood preserving.  That is, there can be no
-%% new satisfying assignments when the clause is deleted.
-
-The sequence of operator declarations, asserted clauses, and
-clause deletions represents a systematic transformation of the input formula
-into a POG\@.  Validating all of these steps serves to prove that
-POG $P$ is logically equivalent to the input formula.
-At the completion of the proof, the following \textsc{final conditions} must hold:
-\begin{enumerate}
-\item There is exactly one remaining clause that was added via RUP
-  addition, and this is a unit clause consisting of root literal $r$.
-\item All of the input clauses have been deleted.
-\end{enumerate}
-In other words, at the end of the proof it must hold that the active clauses be exactly those
-in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
-of unit clause $\{r\}$ and the defining clauses for the nodes, providing a Tseitin encoding of $P$. Recognizing that
-any total assignment to the input variables implicitly defines the assignments to the extension variables,
-we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
-Let $\inputformula$ denote the input formula.
-The sequence of clause addition steps provides a \emph{forward implication} proof that
-$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
-assignment $\assign$ satisfying the input formula must also satisfy
-the formula represented by the POG\@.
-Conversely,
-each proof step that deletes an input clause proves that any
-total assignment $\alpha$ that falsifies the clause must
-falsify $\phi_\noder$.  Deleting all but the final asserted clause and all input clauses provides a \emph{reverse implication} proof
-that
-$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.
-
-The supplement~\cite{bryant:sat:2023:supplement},
-shows the CPOG description for
-an input formula with five clauses, yielding a POG with six
-nonterminal nodes.  It explains how the clause addition and deletion
-steps yield a proof of equivalence between the input formula and its POG
-representation.
-
-\section{Generating CPOG from dec-DNNF}
-\label{section:generating:cpog}
-
-%%Figure~\ref{fig:chain} illustrates one method of generating a CPOG
-%%file from an input formula in conjunctive normal form.  Our toolchain
-%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
-%%but it would also work for other tools that use a top-down search
-%%strategy to generate a dec-DNNF representation.
-A dec-DNNF
-graph can be directly translated  into a POG.
-In doing this conversion,
-our program performs simplifications to
-eliminate Boolean constants.
-Except in degenerate cases,
-where the formula is unsatisfiable or a tautology,
-we can therefore assume
-that the POG does not contain any constant nodes.
-In addition, negation is only
-applied to variables, and so the only edges with negative polarity will have variables as children.
-We can therefore
-view the POG as consisting
-of \emph{literal} nodes corresponding to input variables and their negations, along with
-\emph{nonterminal} nodes, which can be further classified as \emph{product} and \emph{sum} nodes.
-
-\subsection{Forward Implication Proof}
-
-For input formula $\inputformula$ and its translation into a POG $P$
-with root node $\noder$, the most challenging part of the proof is to
-show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
-that any total assignment $\assign$ that is a model of $\inputformula$
-and the POG definition clauses
-will yield $\assign(r) = 1$, for root literal $r$.  This part of the
-proof consists of a series of clause assertions leading to one adding
-$\{r\}$ as a unit clause.  We have devised two methods of generating this
-proof.  The \emph{monolithic} approach makes just one call to a
-proof-generating SAT solver and has it determine the relationship
-between the two representations.  The \emph{structural} approach
-recursively traverses the POG, generating proof obligations at each
-node encountered.  It may require multiple calls to a proof-generating SAT
-solver.
-
-As notation,
-let $\psi$ be a subset of the clauses in $\inputformula$.
-For partial assignment
-$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
-obtained from $\psi$ by: i) eliminating any
-clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
-ii) for the remaining clauses eliminating those literals $\lit$ for
-which $\passign(\lit) = 0$, and iii) eliminating any duplicate or tautological clauses.
-In doing these simplifications, we also track the \emph{provenance}
-of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
-More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov_{\passign}(C, \psi)$ denote
-those clauses $C' \in \psi$, such that
-$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
-We then extend the definition of $\prov$ to any simplified formula
-$\gamma$ as $\prov_{\passign}(\gamma, \psi) = \bigcup_{C \in \gamma} \prov_{\passign}(C, \psi)$.
-
-The monolithic approach
-takes advantage of the clausal representations of
-the input formula $\inputformula$ and the POG formula $\phi_\noder$.
-We can express the negation of $\phi_\noder$ in clausal form as
-$\theta_{\obar{\noder}} \doteq \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
-Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
-when the formula $\inputformula \land \theta_{\obar{\noder}}$
-is unsatisfiable, where the
-conjunction can be expressed as the union
-of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
-For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CPOG proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
-Our experimental results show
-that this approach can be very effective and generates short proofs
-for smaller problems, but it does not scale well enough for general
-use.
-
-We describe the structural approach to proof generation as a recursive procedure
-$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
-node $\nodeu$, partial assignment
-$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
-The procedure adds a number of clauses to the proof, culminating with
-the addition of the \emph{target} clause:
-$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
-that $(\bigwedge_{\lit \in \passign} \lit) \imply u$, i.e.,
-that any total
-assignment $\assign$ such that $\passign \subseteq \assign$
-will assign $\assign(u) = 1$.
-The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
-The result will therefore be to add unit clause $\{r\}$ to the proof.
-Here we present a correct, but somewhat inefficient formulation of
-$\validate$.  We then refine it with some optimizations.
-
-The recursive call $\validate(\nodeu, \passign, \psi)$ assumes that we have
-traversed a path from the root node down to node $\nodeu$, with the
-literals encountered in the product nodes forming the partial
-assignment $\passign$.  The set of clauses $\psi$ can be a proper
-subset of the input clauses $\inputformula$ when a product node has caused
-a splitting into clauses containing disjoint variables.
-The subgraph with root node $\nodeu$ should be a POG representation of the formula
-$\simplify{\psi}{\passign}$.
-%% Having the
-%% routine add its target clause indicates that any total assignment
-%% $\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.
-
-The process for generating such a proof depends on the form of node $\nodeu$:
-\begin{enumerate}
-\item If $\nodeu$ is a literal $\lit'$, then the formula
-  $\simplify{\psi}{\passign}$ must consist of the single unit clause
-  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
-  Any of these can
-  serve as the target clause.
-\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
-  then, since the node originated from a dec-DNNF graph, there must be
-  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
-  product node containing a literal node for $x$ as a child.  In either case, we
-  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
-  This will cause the addition of the target clause
-  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
-  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
-  causing the addition of the target clause
-  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-  These recursive results can be combined with the second and third defining clauses for $\nodeu$
-(see Table~\ref{tab:defining}(B))
-  to generate the target clause for $\nodeu$, requiring at most two RUP steps.
-\item If $\nodeu$ is a product node, then we can divide its children
-  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
-  \begin{enumerate}
-    \item For each literal
-  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
-  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
-  cases, this can be done by simple Boolean constraint propagation (BCP).
-  In other cases, we must prove that the formula
-  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
-  do so by writing the formula to a file, invoking a proof-generating
-  SAT solver, and then converting the generated unsatisfiability proof
-  into a sequence of clause additions in the CPOG file.
-  (The solver is constrained to only use RUP inference rules, preventing it from introducing extension variables.)
-\item For a single nonterminal child ($k = 1$), we recursively call
-  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
-\item For multiple nonterminal children ($k > 1$),
-  it must be the case that the clauses in
-  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
-  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
-  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
-  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
-  literals.
-  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
-  We recursively call
-  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
-\end{enumerate}
-  We then generate the target clause for node $\nodeu$ with a single RUP step,
-creating the hint by combining the results from the BCP and SAT calls for
-  the literals, the recursively computed target clauses, and all but
-  the first defining clause for node $\nodeu$
-(see Table~\ref{tab:defining}(A)).
-\end{enumerate}
-Observe that
-all of these steps involve a polynomial number of
-operations per recursive call, with the exception of those that call
-a SAT solver to validate a literal.
-
-\subsection{Reverse Implication Proof}
-
-Completing the equivalence proof of input formula $\inputformula$ and its POG
-representation with root node $\noder$ requires showing that
-$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
-CPOG framework by first deleting all asserted clauses, except for the
-final unit clause for root literal $r$, and then deleting all of the
-input clauses.
-%%  Suppose at some point in this process that we want to
-%% delete clause $C$ such that the remaining set of clauses will form the
-%% CNF formula $\phi$.  Deleting $C$ requires proving that
-%% $\phi \imply C$.
-
-The asserted clauses can be deleted in reverse order, using the same
-hints that were used in their original assertions.  By reversing the
-order, those clauses that were used in the hint when a clause was
-added will still remain when it is deleted.
-
-Each input clause deletion can be done as a single RUP step, based
-on an algorithm to test for clausal entailment in det-DNNF graphs~\cite{capelli:sat:2019,darwiche:jair:2002}.  The
-proof generator constructs the hint sequence from the defining
-clauses of the POG nodes via a single, bottom-up pass through the
-graph.  The RUP deletion proof for input clause $C$ effectively proves that any
-total assignment $\assign$ that satisfies the POG definition clauses
-but does not satisfy $C$ will yield
-$\assign(r) = 0$.  It starts with the set of literals
-$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
-assignment $\assign$ to falsify clause $C$.
-It then
-adds literals via unit propagation until a
-conflict arises.    Unit literal $r$ gets
-added right away, setting up a potential conflict.
-
-Working upward through the graph, node $\nodeu$ is \emph{marked} when
-the collected set of literals forces $u$ to evaluate to $0$.  When marking $\nodeu$, the
-program adds $\obar{u}$ to the RUP literals and adds the appropriate
-defining clause to the hint.  A literal node for
-$\lit$ will be marked if $\lit \in C$, with no hint required.  If
-product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
-$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
-added to the hint.  Marking sum node $\nodeu$ requires that its two children are marked.
-The first defining
-clause for this node (see Table~\ref{tab:defining}(B)) will then be added to the hint.  At the very end, the program
-(assuming the reverse implication holds) will attempt to mark root
-node $\noder$, which would require $\assign(r) = 0$, yielding a
-conflict.
-
-It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
-each clause deletion requires a single RUP step having a hint with length
-bounded by the number of POG nodes.
-
-\section{Optimizations}
-
-The performance of the structural proof generator for forward implication, both in its execution time and
-the size of the proof generated, can be improved by two optimizations
-described here.  A key feature is that they do not require any changes
-to the proof framework---they build on the power of extended
-resolution to enable new logical structures.  They
-involve declaring new product nodes to encode products of literals.
-These nodes are not part of the POG
-representation of the formula; they serve only to enable the forward
-implication proof.  Here we summarize the two optimizations.
-The supplement~\cite{bryant:sat:2023:supplement} provides more details.
-
-\emph{Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
-having many---tens or even hundreds---of literals as children.  The
-earlier formulation of $\validate$ considers each literal $\lit \in \lambda$
-separately, calling a SAT solver for every literal that cannot be
-validated with BCP\@.  Literal grouping handles
-all of these literals together.  It defines product node $\nodev$
-having the literals as children.  The goal then becomes to prove that
-any total assignment must yield 1 for extension variable $v$.  Calling
-a solver with $v$ set to 0 yields an unsatisfiability proof that can
-be mapped back to a sequence of clause additions in the CPOG file validating
-all of the literals.
-
-\emph{Lemmas:} Our formulation of $\validate$ requires each call at a
-node $\nodeu$ to recursively validate all of its children.  This
-effectively expands the graph into a tree, potentially requiring an
-exponential number of recursive steps.  Instead, for each node
-$\nodeu$ having $\indegree(\nodeu) > 1$, the program can define and
-generate the proof of a \emph{lemma} for $\nodeu$ when it is first reached by
-a call to $\validate$ and then apply this lemma for this and
-subsequent calls.  The lemma states that the POG with root node
-$\nodeu$ satisfies forward implication for a formula
-$\gamma_{\nodeu}$, where some of the clauses in this formula are
-input clauses from $\inputformula$, but others are simplified
-versions of input clauses.
-The key idea is to introduce product nodes to encode (via DeMorgan's
-Laws) the simplified clauses and have these serve as lemma arguments.
-
-The combination of these two optimization guarantees that i) each call
-to $\validate$ for a product node will cause at most one invocation of
-the SAT solver, and ii) each call to $\validate$ for any node $\nodeu$
-will cause further recursive calls only once.  Our experimental
-results~\cite{bryant:sat:2023:supplement} show that these
-optimizations yield substantial benefits.
-
-\input{lean}
-
-\section{Implementations}
-We have implemented programs that, along with
-the \dfour{} knowledge compiler, form the toolchain illustrated in
-Figure~\ref{fig:chain}.\footnote{The source code for all tools
-is available as part of the supplement~\cite{bryant:sat:2023:supplement}}  The proof generator is the same in both
-cases, since it need not be trusted.
-Our \emph{verified}
-version of the proof checker and ring evaluator have been formally
-verified within the \lean{} theorem prover.  Our long term goal is to
-rely on these.  Our \emph{prototype} version is written in C\@.
-It is faster and
-more scalable, but we anticipate its need will diminish as the
-verified version is further optimized.
-%% Our long-term goal is to have a toolchain suitable for use in a
-%% knowledge compiler competition, along the lines of the annual
-%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
-%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
-%% Our experiments
-%% consider the benchmark problems and computing capabilities that would
-%% arise in such a competition and evaluate whether our implementationed toolchain
-%% could handle these problems.  As with the original proof-generating
-%% SAT competitions, we assume that the proofs at a competition could be
-%% checked with the unverified checker.  Having a verified checker
-%% that works on some subset of the benchmarks would provide an
-%% additional degree of certainty.
-
-Our proof generator is written in C/C++ and uses
-\cadical{}~\cite{biere-cadical-2019} as its SAT solver.  To convert
-proof steps back into hinted CPOG clause additions, the generator can
-use either its own RUP proof generator, or it can invoke
-\dtrim{}~\cite{heule:cade:2013}.  The latter yields shorter proofs and scales well
-to large proofs, but each invocation has a high startup cost.  We
-therefore only use it when solving larger problems (currently ones
-with over 1000 clauses).
-
-The proof generator can optionally be instructed to generate a {\em
-one-sided} proof, providing only the reverse-implication portion of the proof via
-input clause deletion.  This can provide useful information---any
-assignment that is a model for the compiled representation
-must also be a model for the input formula---even when
-full validation is impractical.
-
-We incorporated a ring evaluator into the prototype checker.  It
-can perform both standard and weighted
-model counting with full precision.  It performs arithmetic over a subset of the rationals
-we call $\drational$, consisting of numbers of the form $a \cdot 2^{b}
-\cdot 5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented
-to have arbitrary range.  Allowing scaling by powers of 2 enables the
-density computation and rescaling required for standard model
-counting.  Allowing scaling by powers of both 2 and 5 enables exact
-decimal arithmetic, handling the weights used in the weighted model
-counting competitions.  To give a sense of scale, the counter
-generated a result with 260,909 decimal digits
-for one of the weighted benchmarks.
-
-\section{Experimental Evaluation}
-\label{sect:experimental}
-
-\input{figure-combined-verify}
-
-We summarize the results of our experiments here.
-The supplement~\cite{bryant:sat:2023:supplement}
-provides a more complete description. For our
-evaluation, we used the public benchmark problems from the 2022
-standard and weighted model competitions.\footnote{Downloaded from \url{https://mccompetition.org/2022/mc_description.html}}  We found that there were
-180 unique CNF files among these, ranging in size from 250 to
-2,753,207 clauses.
-We ran our programs on a processor with 64~GB of
-memory and having an attached high-speed, solid-state
-disk.
-With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
-benchmark problems.  Our proof generator was able to convert all of
-these into POGs, with their declarations ranging from 304 to
-2,761,457,765 (median 848,784) defining clauses.
-%(The
-%maximum count overflowed the overflowed the 32-bit signed integers we
-%used to represent clause identifiers in both the proof generator and
-%checker.)
-
-We ran our proof generator with a time limit of 10,000 seconds.  The
-results are shown in Figure~\ref{fig:dual:total}, The left-hand plot
-shows the elapsed time for the combination of proof generation,
-checking, and counting versus the time for
-\dfour{}.
-The proof generator was able to generate full proofs for 108 of the problems and
-one-sided proofs for an additional 9 of them, leaving just 7 with no
-verification.  The prototype checker successfully verified all of the generated
-proofs.  The longest runtime for the combination of proof generation, checking, and counting
-for a full proof was 13,145 seconds.
-Overall the ratio
-between the combined time for generation, checking, and counting versus the time
-for \dfour{} had a harmonic mean of 5.5.
-The right-hand plot shows the total number of clauses in the CPOG file versus the number of defining clauses for the problems having full proofs.
-The ratio between the total number of clauses and the number of defining clauses had a harmonic mean of 3.13.
-To date, we have not found
-any errors in the dec-DNNF graphs generated by \dfour{}.
-
-We found that the monolithic approach for generating the forward
-implication proof works well for smaller POGs (up to one million
-defining clauses), but it becomes inefficient for larger ones.  These
-experiments suggest a possible hybrid approach, stopping the recursion
-of the structural approach and shifting to monolithic mode once the
-subgraph size is below some threshold.  By using monolithic mode, we
-were also able to perform end-to-end verification of all but one of the
-benchmarks that could be verified without preprocessing, with a total
-time limit (including preprocessing) of 1,000 seconds.
-
-We also found that our two optimizations: literal grouping and
-lemmas can provide substantial improvements in proof size and runtime.
-In the extreme cases, a lack
-of lemmas caused one proof to grow by a factor of 52.5, while a lack
-of literal grouping caused another proof to grow by a factor of 39.6.
-Overall, it is clear that these two optimizations are worthwhile, and
-sometimes critical, to success for some benchmarks.
-
-Running the verified proof checker in \lean{} required, on average
-(harmonic mean), around 5.9 times longer than the prototype checker.
-Encouragingly, the scaling trends were identical for the two solvers,
-indicating that the two checkers have similar asymptotic performance.
-We consider a factor of 5.9 to be acceptable for the assurance
-provided by formal verification.
-
-In comparing other proof frameworks, we found that the \cdfour{}
-toolchain ran very fast and could handle very large benchmarks.  Even
-with a total time limit of 1,000 seconds, including the time for
-knowledge compilation, the \cdfour{} toolchain completed 106
-benchmarks, while the CPOG toolchain completed just 82.  We found
-that the runtimes for the MICE toolchain versus our CPOG toolchain
-showed little correlation, reflecting the fact that the two solve
-different problems and use different approaches.  In general, our CPOG
-toolchain showed better scaling, in part due to its ability to control
-the recursion through lemmas.  Neither of these prior
-toolchains could perform end-to-end verification when the knowledge
-compilation was preceded by preprocessing.
-
-\section{Conclusions}
-\label{sect:future}
-
-This paper demonstrates a method for certifying the equivalence of two
-different representations of a Boolean formula: an input formula
-represented in conjunctive normal form, and a compiled representation
-that can then be used to extract useful information about the formula,
-including its weighted and unweighted model counts.  It builds on the
-extensive techniques that have been developed for clausal
-proof systems, including extended resolution and reverse unit propagation, as well as established tools, such as
-proof-generating SAT solvers and \dtrim{}.
-
-We are hopeful that having checkable proofs for knowledge compilers
-will allow them to be used in applications where high levels of trust
-are required, and that it will provide a useful tool for developers of
-knowledge compilers.
-%%Although our current implementation only handles
-%%the outputs of one particular program, it could be adapted
-Our experiments demonstrate that our toolchain can already handle
-problems nearly at the limits of current knowledge compilers.  Further
-engineering and optimization of our proof generator and checker could
-improve their performance and capacity substantially.  We are hopeful
-that our tool can be adapted to handle other knowledge compiler
-representations, such as sentential decision diagrams (SDDs)~\cite{darwiche:ijcai:2011}.
-
-%% Each node $\nodeu$ in an $SDD$ has children $p_1, p_2, \ldots, p_k$ and
-%% $s_1, s_2, \ldots p_k$, where the Boolean formula $\phi(\nodeu)$ associated
-%% with the node is defined recursively as $\phi(\nodeu) = \bigwedge_{1 \leq i \leq k}
-%% [\phi(p_i) \land \phi(s_i)]$.  Moreover, each children $p_i$ and $p_j$
-%% such that $1 \leq i < j \leq k$ must satisfy $\modelset(\phi(p_i))
-%% \cap \modelset(\phi(p_j)) = \emptyset$, and for every $p_i$ and $s_j$
-%% such that $1 \leq i,j \leq k$, we must have $\dependencyset(\phi(p_i))
-%% \cap \dependencyset(\phi(s_j)) = \emptyset$.  The POG representation
-%% of such a node could have $k$ product nodes: $t_i = p_i \pand s_i$
-%% for $1 \leq i \leq k$, and then $k-1$ binary sum nodes to form
-%% the sum of $t_1, \ldots, t_k$.
-
-%%\newpage
-
-\bibliography{references}
-
-\end{document}
+\vspace{1em}\noindent
+\textbf{Model counting.} Finally, we established that ring evaluation with the appropriate weights
+corresponds to the standard model count. To do so, we defined a function that
+carries out an integer calculation of the number of models over a set of variables
+of cardinality \lstinline{nVars}:
+\begin{lstlisting}
+def countModels (nVars : Nat) : PropForm Var → Nat
+  | tr       => 2^nVars
+  | fls      => 0
+  | var _    => 2^(nVars - 1)
+  | neg φ    => 2^nVars - countModels nVars φ
+  | disj φ ψ => countModels nVars φ + countModels nVars ψ
+  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
+\end{lstlisting}
+We then formally proved that for a partitioned formula whose variables are among a finite set
+\lstinline{s}, this computation really does count the number of models over \lstinline{s}:
+\begin{lstlisting}
+theorem countModels_eq_card_models {φ : PropForm Var} {s : Finset Var} :
+  vars φ ⊆ s → partitioned φ → countModels (card s) φ = card (models φ s)
+\end{lstlisting}
+In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
+we have that the number of models of \lstinline{φ} on its variables is
+\lstinline{countModels φ (card (vars φ))}.

### 2024-01-05 doc/JAIR/experimental-section.tex
Split Experimental results as separate file

--- 2024-01-05 doc/JAIR/jair-paper.tex
@@
-\documentclass[twoside,11pt]{article}
-\usepackage{jair, theapa, rawfonts}
-
-\ShortHeadings{Certified Knowledge Compilation}{Bryant, Nawrocki, Avigad, \& Heule}
-
-\jairheading{1}{2024}{1--16}{10/22}{03/24}
-
-\ShortHeadings{Certified Knowledge Compilation}{Bryant, Nawrocki, Avigad, \& Heule}
-
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{tikz}
-\usepackage{pgfplots}
-\pgfplotsset{compat=1.18}
-\usepackage{booktabs}
-%%\usepackage{hyperref}
-\newcommand{\url}[1]{\texttt{#1}}
-
-\newcommand{\pand}{\mathbin{\land^\textsf{p}}}
-\newcommand{\por}{\mathbin{\lor^\textsf{p}}}
-\DeclareMathOperator*{\Pand}{\bigwedge^\textsf{p}}
-\DeclareMathOperator*{\Por}{\bigvee^\textsf{p}}
-\newcommand{\boolnot}{\neg}
-\newcommand{\tautology}{\top}
-\newcommand{\nil}{\bot}
-\newcommand{\obar}[1]{\overline{#1}}
-\newcommand{\oneminus}{{\sim}}
-\newcommand{\lit}{\ell}
-
-\newcommand{\varset}{X}
-\newcommand{\exvarset}{Z}
-\newcommand{\dependencyset}{{\cal D}}
-\newcommand{\litset}{{\cal L}}
-\newcommand{\ring}{{\cal R}}
-\newcommand{\semiring}{{\cal S}}
-\newcommand{\dset}{{\cal A}}
-\newcommand{\rep}{\textbf{R}}
-\newcommand{\srep}{\textbf{S}}
-\newcommand{\radd}{+}
-\newcommand{\rmul}{\times}
-\newcommand{\addident}{\textbf{0}}
-\newcommand{\mulident}{\textbf{1}}
-\newcommand{\imply}{\Rightarrow}
-\newcommand{\ifandonlyif}{\Leftrightarrow}
-\newcommand{\eequiv}{\ifandonlyif_{\varset}}
-%\newcommand{\drational}{\mathbb{Q}_{2,5}}
-\newcommand{\drational}{\textbf{Q}_{2,5}}
-\newcommand{\entail}{\vDash}
-\newcommand{\entaildrat}{\entail_{\textrm{DRAT}}}
-
-\newcommand{\extend}[1]{#1^{*}}
-\newcommand{\assign}{\alpha}
-\newcommand{\eassign}{\extend{\alpha}}
-\newcommand{\passign}{\rho}
-\newcommand{\lassign}{\beta}
-\newcommand{\uassign}{{\cal U}}
-\newcommand{\modelset}{{\cal M}}
-
-\newcommand{\indegree}{\textrm{indegree}}
-\newcommand{\outdegree}{\textrm{outdegree}}
-\newcommand{\validate}{\textsf{validate}}
-\newcommand{\prov}{\textrm{Prov}}
-\newcommand{\inputformula}{\phi_I}
-\newcommand{\pogformula}{\theta_P}
-
-\newcommand{\makenode}[1]{\mathbf{#1}}
-\newcommand{\nodeu}{\makenode{u}}
-\newcommand{\nodev}{\makenode{v}}
-\newcommand{\nodew}{\makenode{w}}
-\newcommand{\nodes}{\makenode{s}}
-\newcommand{\nodep}{\makenode{p}}
-\newcommand{\noder}{\makenode{r}}
-
-\newcommand{\simplify}[2]{#1|_{#2}}
-
-\newcommand{\progname}[1]{\textsc{#1}}
-\newcommand{\dfour}{\progname{D4}}
-\newcommand{\cdfour}{\progname{CD4}}
-\newcommand{\Dfour}{\progname{D4}}
-\newcommand{\cadical}{\progname{CaDiCal}}
-\newcommand{\dtrim}{\progname{drat-trim}}
-\newcommand{\ltrim}{\progname{lrat-trim}}
-\newcommand{\Dtrim}{\progname{Drat-trim}}
-
-\definecolor{redorange}{rgb}{0.878431, 0.235294, 0.192157}
-\definecolor{lightblue}{rgb}{0.552941, 0.72549, 0.792157}
-\definecolor{clearyellow}{rgb}{0.964706, 0.745098, 0}
-\definecolor{clearorange}{rgb}{0.917647, 0.462745, 0}
-\definecolor{mildgray}{rgb}{0.54902, 0.509804, 0.47451}
-\definecolor{softblue}{rgb}{0.643137, 0.858824, 0.909804}
-\definecolor{bluegray}{rgb}{0.141176, 0.313725, 0.603922}
-\definecolor{lightgreen}{rgb}{0.709804, 0.741176, 0}
-\definecolor{darkgreen}{rgb}{0.152941, 0.576471, 0.172549}
-\definecolor{redpurple}{rgb}{0.835294, 0, 0.196078}
-\definecolor{midblue}{rgb}{0, 0.592157, 0.662745}
-\definecolor{clearpurple}{rgb}{0.67451, 0.0784314, 0.352941}
-\definecolor{browngreen}{rgb}{0.333333, 0.313725, 0.145098}
-\definecolor{darkestpurple}{rgb}{0.396078, 0.113725, 0.196078}
-\definecolor{greypurple}{rgb}{0.294118, 0.219608, 0.298039}
-\definecolor{darkturquoise}{rgb}{0, 0.239216, 0.298039}
-\definecolor{darkbrown}{rgb}{0.305882, 0.211765, 0.160784}
-\definecolor{midgreen}{rgb}{0.560784, 0.6, 0.243137}
-\definecolor{darkred}{rgb}{0.576471, 0.152941, 0.172549}
-\definecolor{darkpurple}{rgb}{0.313725, 0.027451, 0.470588}
-\definecolor{darkestblue}{rgb}{0, 0.156863, 0.333333}
-\definecolor{lightpurple}{rgb}{0.776471, 0.690196, 0.737255}
-\definecolor{softgreen}{rgb}{0.733333, 0.772549, 0.572549}
-\definecolor{offwhite}{rgb}{0.839216, 0.823529, 0.768627}
-\definecolor{medgreen}{rgb}{0.15, 0.6, 0.15}
-\definecolor{midred}{rgb}{0.80,0.3,0.3}
-
-% Lean code:
-\usepackage{listings}
-%\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
-\definecolor{keywordcolor}{rgb}{0.0, 0.1, 0.6}   % blue
-\definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
-\definecolor{commentcolor}{rgb}{0.4, 0.4, 0.4}   % grey
-\definecolor{symbolcolor}{rgb}{0.0, 0.1, 0.6}    % blue
-\definecolor{sortcolor}{rgb}{0.1, 0.5, 0.1}      % green
-\definecolor{attributecolor}{rgb}{0.7, 0.1, 0.1} % red
-\def\lstlanguagefiles{lstlean.tex}
-% set default language
-\lstset{language=lean, xleftmargin=1em}
-\lstset{backgroundcolor=\color{white}}
-
-\newcommand{\rtext}[1]{\textcolor{midred}{\texttt{#1}}}
-\newcommand{\btext}[1]{\textcolor{midblue}{\texttt{#1}}}
-
-\newtheorem{dfn}{Definition}
-\newtheorem{prop}{Proposition}
-\newtheorem{thm}{Theorem}
-
-\title{Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting}
-
-\author{\name Randal E. Bryant \email rebryant@cmu.edu \\
-  \name Wojciech Nawrocki \email wjnawrock@cmu.edu \\
-  \name Jeremy Avigad \email avigad@cmu.edu \\
-  \name Marijn J. H. Heule \email marijn@cmu.edu \\
-  \addr Carnegie Mellon University \\ Pittsburgh, PA  USA}
-
-\newcommand{\program}[1]{\textsc{#1}}
-\newcommand{\lean}{Lean~4}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-Computing many useful properties of Boolean formulas, such as their weighted or unweighted model count,
-is intractable on general representations. It can become tractable when formulas are expressed in a
-special form, such as the decision-decomposable, negation normal form (dec-DNNF)\@.
-\emph{Knowledge compilation} is the process of converting a formula
-into such a form. Unfortunately existing knowledge compilers provide no guarantee that their output correctly
-represents the original formula, and therefore they cannot validate a model count, or any other computed value.
-
-We present \emph{Partitioned-Operation Graphs} (POGs), a form that can
-encode all
-of the representations used by existing knowledge compilers.
-We have designed  CPOG, a framework that can express proofs of equivalence between a
-POG  and a Boolean formula in conjunctive normal form (CNF).
-
-We have developed a program that generates POG representations from dec-DNNF
-graphs
-produced by the state-of-the-art knowledge compiler
-\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
-are equivalent to the input CNF formulas.  Our toolchain
-for generating and verifying POGs scales to all but the largest
-graphs produced by \dfour{} for formulas from a recent model counting
-competition. Additionally, we have developed a formally verified CPOG
-checker and model counter for POGs in the \lean{} proof assistant.
-In doing so, we proved the soundness of our proof framework. These programs
-comprise the first formally verified toolchain for weighted and unweighted
-model counting.
-\end{abstract}
-
-\section{Introduction}
-
-Given a Boolean formula $\phi$, modern Boolean satisfiability (SAT) solvers can
-find an assignment satisfying $\phi$ or generate a proof that no
-such assignment exists.  They have applications across a variety of
-domains including computational mathematics, combinatorial
-optimization, and the formal verification of hardware, software, and
-security protocols.  Some applications, however, require going
-beyond Boolean satisfiability.  For example, the \emph{model
-  counting problem} requires computing the number of satisfying
-assignments of a formula, including in cases where there are far too many
-to enumerate individually.  Model counting has
-applications in artificial intelligence, computer security, and
-statistical sampling.  There are also many useful extensions of standard model counting,
-including {\em
-  weighted model counting}, where a weight is defined for
-each possible assignment, and the goal becomes to compute the sum of the weights
-of the satisfying assignments.
-
-Model counting is a challenging problem---more challenging than the
-already NP-hard Boolean satisfiability.  Several
-tractable variants of Boolean satisfiability, including 2-SAT, become
-intractable when the goal is to count models and not just determine
-satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
-model counters that scale to very large formulas have been developed, as
-witnessed by the progress in recent model counting competitions.
-
-One approach to model counting, known as \emph{knowledge compilation},
-transforms the formula into a structured form for which model counting
-is straightforward.  For example, the \emph{deterministic, decomposable, negation normal form}
-(det-DNNF) introduced by
-\citeA{darwiche:aaai:2002,darwiche:ecai:2004}, as well as the more restricted
-\emph{decision-DNNF} (dec-DNNF)~\cite{huang:jair:2007,oztok:cp:2014},
-represent a
-Boolean formula as a directed acyclic graph, with terminal nodes
-labeled by Boolean variables and their complements, and with each
-nonterminal node labeled by a Boolean \textsc{and} or \textsc{or} operation.  Restrictions
-are placed on the structure of the graph (described in Section~\ref{sect:pog}) such that a count of the
-models can be computed by a single bottom-up traversal.
-Kimmig et al.\ present a very general {\em
-  algebraic model counting} framework~\cite{kimmig:jal:2017} describing
-properties of Boolean functions that can be efficiently computed from
-a det-DNNF representation.  These include standard and weighted model
-counting, and much more.
-
-One shortcoming of existing knowledge compilers is that they have no
-generally accepted
-way to validate that
-the compiled representation is logically equivalent to the original
-formula.  By contrast, all modern SAT solvers can generate
-checkable proofs when they encounter unsatisfiable formulas.  The
-guarantee provided by a checkable certificate of correctness enables
-users of SAT solvers to fully trust their results.  Experience has also
-shown that being able to generate proofs allow SAT solver developers to quickly
-detect and diagnose bugs in their programs. This, in turn, has led
-to more reliable SAT solvers.
-
-This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
-a form that can encode all of the representations produced by current knowledge
-compilers. The CPOG (for ``certified'' POG) file format then
-captures both the structure of a POG
-and a checkable proof of its logical equivalence to a Boolean formula in
-conjunctive normal form (CNF).  A CPOG
-proof consists of a sequence of clause addition and deletion steps,
-based on an extended resolution proof system~\cite{Tseitin:1983}.
-We establish a set of conditions that, when satisified by a CPOG file, guarantees that it
-encodes a well-formed POG and provides a valid equivalence proof.
-
-\begin{figure}
-\centering{
-\begin{tikzpicture}[scale=0.17]
-  \draw (8,10) rectangle (20,18);
-  \node at (14,15.2) {Knowledge};
-  \node at (14,12.8) {Compiler};
-
-  \draw (28,10) rectangle (40,18);
-  \node at (34,15.2) {Proof};
-  \node at (34,12.8) {Generator};
-
-  \draw (50,10) rectangle (62,18);
-  \node at (56,15.2) {Proof};
-  \node at (56,12.8) {Checker};
-
-  \draw (50,0) rectangle (62,8);
-  \node at (56,5.2) {Ring};
-  \node at (56,2.8) {Evaluator};
-
-  \draw[dashed] (48,-2) rectangle (64,22);
-  \node at (56,20) {Trusted Code};
-
-  %% CNF
-  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (50,16);
-  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
-  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
-  \node [left] at (-1,20) {\large {$\inputformula$}};
-  \node [above] at (2,20) {\texttt{.cnf}};
-
-  %% NNF
-  \draw[thick] (20,12) [-latex] -- (28,12);
-  \node [above] at (24,12) {\texttt{.ddnnf}};
-
-  %% CPOG
-  \draw[thick] (40,12) [-latex] -- (50,12);
-  \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
-  \node [above] at (44,12) {\texttt{.cpog}};
-
-  %% OK/Not
-  \draw[thick] (62,14) [-latex] -- (66,14) ;
-  \node [right] at (66,15.2) { OK /} ;
-  \node [right] at (66,12.8) { Not OK} ;
-  \draw[thick] (62,4) [-latex] -- (66,4) ;
-  \node [right] at (66,4) {\large {$\textbf{R}(\inputformula)$}};
-
-\end{tikzpicture}
-}
-\caption{Certifying toolchain.
-  The output of a standard knowledge compiler is converted into a combined graph/proof (CPOG)
-  which can be independently checked and evaluated}
-\label{fig:chain}
-\end{figure}
-
-Figure~\ref{fig:chain} illustrates our certifying knowledge
-compilation and model counting toolchain.  Starting with input formula
-$\inputformula$, the \dfour{} knowledge
-compiler~\cite{lagniez:ijcai:2017} generates a dec-DNNF
-representation, and the \emph{proof generator} uses this to generate a
-CPOG file.  The \emph{proof checker} verifies the equivalence of the
-CNF and CPOG representations.  The \emph{ring evaluator} computes a
-standard or weighted model count from the POG\@ representation.  As
-the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
-moves the root of trust away from the complex and highly optimized
-knowledge compiler to a relatively simple checker and evaluator.
-Importantly, the proof generator need not be trusted---its errors will
-be caught by the proof checker.
-
-To ensure soundness of the abstract CPOG proof system, as well as
-correctness of its concrete implementation, we formally verified the
-proof system as well as versions of the proof checker and ring
-evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
-Running these two programs on a  CPOG file gives strong
-assurance that the proof and the model count are correct. Our
-experience with developing a formally verified proof checker has shown
-that, even within the well-understood framework of extended
-resolution, it can be challenging to formulate a full set of
-requirements that guarantee soundness.  In fact, as described in Section~\ref{sect:lean:subtle-condition}, our efforts to
-formally verify our proof framework exposed subtle conditions that we had
-to impose on our partitioned sum rule.
-
-We evaluate our toolchain using benchmark formulas from the 2022
-standard and weighted model competitions.  Our tools handle all but
-the largest dec-DNNF graphs generated by \dfour{}.  We also measure
-the benefits of several optimizations as well as the relative
-performance of the verified checker with one designed for high
-performance and capacity.
-
-We also show that our tools can provide end-to-end verification of
-formulas that have been transformed by an equivalence-preserving
-preprocessor.  That is, verification is based on the original formula,
-and so proof checking certifies correct operation of the preprocessor,
-the knowledge compiler, and the proof generator.
-
-This paper is an extended version of one published at the 2023
-Conference on the Theory and Application of Boolean
-Satisfiability~\cite{bryant:sat:2023}.  It provides much greater detail about
-the algorithms, the formal verification, and the experimental results.
-
-\section{Related Work}
-
-Generating proofs of unsatisfiability in SAT solvers has a long
-tradition~\cite{ZhangMalik} and has become widely accepted due to the
-formulation of clausal proof systems for which proofs can readily be
-generated and efficiently checked
-\cite{heule:cade:2013,wetzler14_drattrim}.
-A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
-The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
-Our work builds on the well-established technology and tools associated with clausal proof systems,
-but we require features not found in proofs of unsatisfiability.
-In particular, our proofs construct a new propositional formula, and we must verify
-both
-that it
-satisfies a set of rules and that it
-is equivalent to the input formula.  This requires verifying
-additional proof steps, including clause deletion steps, and subtle
-invariants, as described in Sections~\ref{sect:cpog} and
-\ref{sect:formally-verified-toolchain}.
-
-Capelli, Lagniez, and Marquis developed a knowledge compiler that
-generates a certificate in a proof system that is itself based on
-dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{}
-program, a modified version of \dfour{}, generates annotations to the
-compiled representation, providing information about how the compiled
-version relates to the input clauses.  It also generates a file of
-clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
-Completing the certification involves running two different checkers
-on the annotated dec-DNNF graph and the DRAT file.  Although the
-authors make informal arguments regarding the soundness of their
-frameworks, these do not provide strong levels of assurance.  Indeed,
-we have identified a weakness in their methodology due to an
-invalid assumption about the guarantees provided by \dtrim{}, the
-program it uses to check the DRAT file.  This weakness is
-\emph{exploitable}: their framework can be ``spoofed'' into accepting
-an incorrect compilation.
-
-In more detail, \cdfour{} emits a sequence of clauses $R$ that
-includes the conflict clauses that arose during a top-down processing of
-the input clauses.  Given input formula $\phi_I$, their first task is to
-check whether $\phi_I \imply R$, i.e., that any assignment that satisfies $\phi_I$ also satisfies each of the clauses in $R$.
-They then base other parts of their proof on that
-property and use a separate program to perform a series of additional
-checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
-satisifies the \emph{reverse asymmetric tautology} (RAT) property with
-respect to the preceding
-clauses~\cite{jarvisalo:ijcar:2012,heule:cade:2013}.  Adding a RAT
-clause $C$ to a set of clauses maintains satisfiability, but it does not necessarily
-preserve models.  As an example, consider the following formulas:
-\begin{center}
-  \begin{tabular}{lccc}
-    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
-    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_2 \lor \obar{x}_3)$\\
-  \end{tabular}
-\end{center}
-Clearly, these two formulas are not equivalent---$\phi_1$ has six
-models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
-arbitrary assignments to variable $x_2$.  Critically, however, the
-second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
-satisfying assignment to $\phi_1$ can be transformed into one that
-also satisfies $\phi_2$ by setting $x_2$ to 1, while keeping the values for other variables fixed.
-
-This weakness would allow a buggy (or malicious) version of \cdfour{}
-to spoof the checking framework.  Given formula $\phi_1$ as input, it
-could produce a compiled result, including annotations, based on $\phi_2$ and also include the second clause of $\phi_2$ in $R$.
-The check with \dtrim{} would pass, as would the other tests performed
-by their checker.  We have confirmed this possibility with
-their compiler and checker.\footnote{Downloaded May 18, 2023 as\\
-\url{https://github.com/crillab/d4/tree/333370cc1e843dd0749c1efe88516e72b5239174}.}
-
-This weakness can be corrected by restricting \dtrim{} to only add
-clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
-command-line argument to \dtrim{} that enforces this
-restriction.\footnote{Available at
-\url{https://github.com/marijnheule/drat-trim/releases/tag/v05.22.2023}.}  This weakness, however, illustrates the general challenge of
-developing a new proof framework.
-As we can attest,
-without engaging in an effort to formally verify the framework, there are likely to be
-conditions that make the framework unsound.
-
-Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
-proof framework for model counting programs.  Their proof rules are
-based on the algorithms commonly used by model counters.  They
-developed a program that can generate proof traces from dec-DNNF
-graphs and a program to check adherence to their proof rules.  This
-framework is not directly comparable to ours, since it only certifies
-the unweighted model count, but it has similar goals.
-Again, they provide only  informal arguments
-regarding the soundness of their framework.
-
-Both of these prior certification frameworks are strongly tied to the
-algorithms used by the knowledge compilers and model counters.  Some
-of the conditions to be checked are relevant only to specific
-implementations.    Our framework is very general and is based on a small set
-of proof rules.  It builds on the highly developed
-concepts of clausal proof systems.  These factors were important in enabling formal verification.
-In Section~\ref{sect:experimental},
-we also compare the performance of our toolchain to these other two.
-
-\section{Logical Foundations}
-\label{sect:logical:foundations}
-
-  Let $\varset$ denote a set of Boolean variables, and let $\assign$
-  be an \emph{assignment} of truth values to some subset of the
-  variables, where $0$ denotes false and $1$ denotes true, i.e.,
-  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
-  \subseteq \varset$.  We say the assignment is \emph{total} when it
-  assigns a value to every variable ($\varset' = \varset$), and that
-  it is \emph{partial} otherwise.
-  The set of all possible total assignments over
-  $\varset$ is denoted $\uassign$.
-
-For each variable $x \in \varset$,
-  we define the \emph{literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
-  negation of $x$. An
-  assignment $\assign$ can be viewed as a set of literals, where
-  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
-  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
-$\obar{\lit} = x$ when $\lit = \obar{x}$.
-
-\begin{dfn}
-  The set of Boolean formulas is defined recursively.  Each
-  formula $\phi$ has an associated \emph{dependency set}
-  $\dependencyset(\phi)  \subseteq \varset$, and a set of \emph{models} $\modelset(\phi)$,
-  consisting of total assignments that satisfy the formula:
-  \begin{enumerate}
-  \item Boolean constants $0$ and $1$ are Boolean formulas,
-    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
-  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
-    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
-  \item For formula $\phi$, its \emph{negation}, written $\boolnot \phi$ is a Boolean formula,
-    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
-  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
-      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
-      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
-  \end{enumerate}
-\label{def:boolean}
-\end{dfn}
-
-  We highlight some special classes of Boolean formulas.  A formula is
-  in \emph{negation normal form} when negation is applied only to
-  variables.  A formula is in \emph{conjunctive normal form} (CNF)
-  when (i) it is in negation normal form, and (ii) sum is applied only
-  to literals, and (iii) there is a single product operation over all
-  of the sums.  A CNF formula can be represented as a set of
-  \emph{clauses}, each of which is a set of literals.  Each clause
-  represents the sum of the literals, and the formula is the product
-  of its clauses.  We use set notation to reference the clauses within
-  a formula and the literals within a clause.  A clause consisting of
-  a single literal is referred to as a \emph{unit} clause and the
-  literal as a \emph{unit} literal.  This literal must be assigned
-  value $1$ by any satisfying assignment of the formula.
-
-\begin{dfn}
-  \label{def:partitioned-operation-formula}
-  A \emph{partitioned-operation formula}
- satisfies the following for all product and sum operations:
-      \begin{enumerate}
-      \item The arguments to each product must have disjoint dependency sets.  That is, operation
-        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \item The arguments to each sum must have disjoint models.  That is, operation
-        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
-      \end{enumerate}
-\end{dfn}
-     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.
-
-  \section{Ring Evaluation of a Boolean Formula}
-
-We propose a general framework for summarizing properties of Boolean
-formulas along the lines of algebraic model counting~\cite{kimmig:jal:2017}.
-
-\begin{dfn}
-  A \emph{commutative ring} $\ring$ is an algebraic structure
-  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
-  with elements in the set $\dset$ and with commutative and
-  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
-  such that multiplication distributes
-  over addition.  $\addident$ is the additive identity and $\mulident$ is
-  the multiplicative identity.  Every element $a \in \dset$ has an
-  \emph{additive inverse} $-a$ such that $a + -a = \addident$.
-\label{def:ring}
-\end{dfn}
-We write $a - b$ as a shorthand for $a + -b$.
-
-\begin{dfn}[Ring Evaluation Problem]
-\label{def:ring_evaluation}
-  For commutative ring $\ring$, a \emph{ring weight function} associates a value $w(x) \in \dset$ with
-  every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.
-
-  For Boolean formula $\phi$ and ring weight function $w$, the \emph{ring evaluation problem} computes
-  \begin{equation}
-    \begin{array}{rcl}
-    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
-    \end{array}
-  \end{equation}
-  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
-\label{def:weight}
-\end{dfn}
-
-Many important properties of Boolean formulas can be
-expressed as ring evaluation problems.  The
-(standard) \emph{model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
-It can be cast as a ring evaluation problem by having $\radd$ and
-$\rmul$ be addition and multiplication over rational numbers and using
-weight function $w(x) = 1/2$ for every variable $x$.
-Ring evaluation of formula $\phi$ gives the \emph{density} of
-the formula, i.e., the fraction of all possible total assignments that are
-models.  For $n = |\varset|$, scaling the density by $2^n$
-yields the number of models.
-
-The \emph{weighted model counting}  problem is also defined over
-rational numbers.  Some formulations  allow
-independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
-$W(x) + W(\obar{x}) \not = 1$.
-We can cast this as a
-ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
-performing ring evaluation with weight function $w(x) = W(x)/r(x)$ for each
-variable $x$, and computing the weighted count
-as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
-Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.
-
-The \emph{function hashing problem} provides a test
-of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
-finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, choose a value from $\dset$ at random for $w(x)$.  Two formulas
-$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
-if they are logically equivalent,
-and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
-If they are not equivalent, then
-the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
-least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
-Function hashing can therefore be used as part of a
-randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
-For example, it can compare different runs on a single formula,
-either from different compilers or from a single compiler with different configuration parameters.
-
-\section{Partitioned-Operation Graphs (POGs)}
-\label{sect:pog}
-
-Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
-\begin{prop}
-\label{prop:ring:eval}
-Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
-\begin{eqnarray}
-\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation}\\
-\textstyle
-\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product}\\
-\textstyle
-\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum} 
-\end{eqnarray}
-\end{prop}
-As is described in \ref{sect:formally-verified-toolchain}, we have proved these three equations using \lean{}.
-
-A \emph{partitioned-operation graph} (POG) is a directed, acyclic
-graph with nodes $N$ and edges $E \subseteq N \times N$.  We denote
-nodes with boldface symbols, such as $\nodeu$ and $\nodev$.  When
-$(\nodeu,\nodev) \in E$, node $\nodev$ is said to be a \emph{child} of
-node $\nodeu$.  The in- and out-degrees of node $\nodeu$ are defined
-as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
-$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.  Node
-$\nodeu$ is said to be \emph{terminal} if $\outdegree(\nodeu) = 0$.  A
-terminal node is labeled by a Boolean constant or variable.  Node
-$\nodeu$ is said to be \emph{nonterminal} if $\outdegree(\nodeu) > 0$.
-A nonterminal node is labeled by Boolean operation $\pand$ or $\por$.
-A node can be labeled with operation $\pand$ or $\por$ only if it
-satisfies the partitioning restriction for that operation.  Every POG
-has a designated \emph{root node} $\noder$.  Each edge has
-a \emph{polarity}, indicating whether (negative polarity) or not
-(positive polarity) the corresponding argument should be negated.
-
-A POG represents a partitioned-operation
-formula with a sharing of common subformulas.  Every node in the graph can be viewed as a partitioned-operation formula, and so we write
-$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
-Each such formula has a set of models $\modelset(\phi_{\nodeu})$.
-
-We can now define and compare two related representations:
-\begin{itemize}
-\item A det-DNNF graph can be viewed a POG with negation applied only to variables.
-\item A dec-DNNF graph is a det-DNNF graph with the further
-  restriction that any sum node $\nodeu$ has exactly two children
-  $\nodeu_1$ and $\nodeu_0$, and it has an associated \emph{decision variable} $x$.
-  For $b \in \{0,1\}$, node $\nodeu_b$ can be a terminal node with variable $x$, where the polarity of the edge from $\nodeu$ to $\nodeu_b$ is
-  negative for $b=0$ and positive for $b=1$.
-  Alternatively, $\nodeu_b$ can be a product node having either literal $\obar{x}$ ($b=0$), or literal $x$ ($b=1$) as one of its arguments.
-  Either form implies that any total assignment $\assign \in \modelset(\phi_{\nodeu_b})$
-  has $\assign(x)=b$, for $b \in \{0,1\}$.
-\end{itemize}
-The generalizations encompassed by POGs have also been referred to as \emph{deterministic decomposable circuits} (d-Ds)~\cite{monet:amw:2018}.
-Our current proof generator only works for knowledge compilers
-generating dec-DNNF representations, but these generalizations
-allow for future extensions, while maintaining the ability to
-efficiently perform ring evaluation.
-
-We define the \emph{size} of POG $P$, written $|P|$, to be the
-the number of nonterminal nodes plus the number of edges from these nodes to their children.  Ring
-evaluation of $P$ can be performed with at most $|P|$ ring
-operations by traversing the graph from the terminal nodes up to
-the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
-The final result is then $\rep(\noder, w)$.
-
-\section{Clausal Proof Framework}
-\label{sect:clausal}
-
-A proof in our framework consists of a sequence of clause addition and
-deletion steps, with each step preserving the set of solutions to the
-original formula.  The status of the proof at any step is represented
-as a set of \emph{active} clauses $\theta$, i.e., those that have been
-added but not yet deleted.  Our framework is based on \emph{extended}
-resolution~\cite{Tseitin:1983}, where proof steps can introduce new
-\emph{extension variables} encoding Boolean formulas over input and
-prior extension variables.  That is, each extension variable $z$ is
-introduced via a set of \emph{defining clauses} that encode a formula
-$z \ifandonlyif F$, where $F$ is a Boolean formula over a subset of the input
-variables $\varset$ and previously defined extension variables $\exvarset$.
-We write $\theta$ for
-formulas encoded as clauses, possibly with extension variables, and
-$\phi$ for formulas
-that use no extension variables.
-
-Let $\exvarset(\theta)$
-denote the set of extension variables occuring in formula $\theta$.
-For any total assignment $\assign$ to the variables in $\varset$, the defining
-clauses induce a unique assignment $\eassign$ to the variables in $\varset
-\cup \exvarset(\theta)$.  For Boolean formula $\phi$ over variables
-$\varset$ and clausal
-formula $\theta$ over the variables $\varset \cup \exvarset(\theta)$, we say that $\phi$ is {\em
-  effectively equivalent} to $\theta$, written
-$\phi \eequiv \theta$,
-when for any assignment
-$\assign$ to the variables in $\varset$, assignment $\assign$ is a model of $\phi$ if
-and only if its extension $\eassign$ is a model of $\theta$.
-Starting with  $\theta$ equal to input formula $\inputformula$,
-the proof must maintain the invariant that
-$\inputformula \eequiv \theta$.
-
-Clauses can be added in two different ways.  One is when they serve as
-the defining clauses for an extension variable.  This form
-occurs only when defining $\pand$ and $\por$ operations, as is
-described in Section~\ref{sect:cpog}.  Clauses can also be added or
-deleted based on \emph{implication redundancy}.  That is, when clause
-$C$ satisfies $\theta \imply C$ for formula $\theta$, then it can either
-be added to $\theta$ to create the formula $\theta \cup \{C\}$ or it can be deleted
-from $\theta \cup \{C\}$ to create $\theta$.
-
-We use \emph{reverse unit propagation} (RUP) to certify
-implication redundancy when adding or deleting
-clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
-RUP
-is the core rule supported by standard
-proof checkers~\cite{heule:cade:2013,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
-way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
-Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
-proved redundant with respect to formula $\theta$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
-\emph{antecedent} clauses, such that each $D_i$ is in $\theta$.
-A RUP step
-proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
-that the combination of the antecedents plus the negation of $C$ leads
-to a contradiction.  The negation of $C$ is the formula
-$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
-\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
-clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
-check processes the clauses of the antecedent in sequence, inferring
-additional unit clauses.  In processing clause $D_i$, if all but one
-of the literals in the clause is the negation of one of the
-accumulated unit clauses, then we can add this literal to the
-accumulated set.  That is, all but this literal have been falsified,
-and so it must be set to true for the clause to be satisfied.  The
-final step with clause $D_k$ must cause a contradiction, i.e., all of
-its literals are falsified by the accumulated unit clauses.
-
-Compared to the proofs of unsatisfiability generated by SAT solvers,
-ours have important differences.  Most
-significantly, each proof step must preserve the set of solutions with respect to the input variables;
-our proofs must therefore justify both clause deletions and additions.
-By contrast, an unsatisfiability proof need only guarantee that
-no proof step causes a satisfiable set of clauses to become
-unsatisfiable, and therefore it need only justify clause additions.
-
-\section{The CPOG Representation and Proof System}
-\label{sect:cpog}
-
-%% The proof system is based on extended
-%% resolution~\cite{Tseitin:1983} with reverse unit propagation
-%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
-%% proving that a set of clauses logically implies another clause.
-%% Alternate version
-%% knowledge compilation.  The proof system is based on extended
-%% resolution~\cite{Tseitin:1983} with reverse unit propagation
-%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
-%% proving that a set of clauses logically implies another clause.
-%%
-%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
-%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
-%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
-%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
-%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
-%%
-%% \begin{center}
-%%   \begin{tabular}{lcccc}
-%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
-%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
-%%   \midrule
-%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
-%%   \end{tabular}
-%% \end{center}
-%%
-%% RUP is an alternative formulation of resolution.  For target clause
-%% $C$, it can be seen that applying resolution operations to the
-%% antecedent clauses from right to left will derive a clause $C'$ such
-%% that $C' \subseteq C$.  By \emph{subsumption} %~\cite{philipp:lpar:2017},
-%% we then have $C' \rightarrow C$.  Compared to listing each resolution
-%% operation as a separate step, using RUP as the basic proof step makes
-%% the proofs more compact.
-%%
-%% As is shown in Figure~\ref{fig:chain}
-%% two programs are involved in producing a certified compilation of a formula:
-%% \begin{itemize}
-%% \item The \emph{proof generator} produces a CPOG file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
-%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a dec-DNNF representation of the formula.
-%% \item The \emph{proof checker} ensures that all of the proof conditions are satisfied.
-%% \end{itemize}
-%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b
-
-A CPOG file provides both a declaration of a POG, as well as a checkable
-proof that a Boolean formula, given in conjunctive normal
-form, is logically equivalent to the POG\@.
-The proof format draws its inspiration from the LRAT~\cite{lrat} and
-QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
-Key properties include:
-\begin{itemize}
-  \item
-  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
-  Declaring a node $\nodeu$ implicitly adds an extension variable $u$ and a set of defining clauses $\theta_{u}$
-  encoding the product or sum operation.
-  This is the only means for adding extension variables to the proof.
-\item Boolean negation is supported implicitly by allowing the
-  arguments of the $\por$ and $\pand$ operations to be literals and not just
-  variables.
-\item
-  The file contains explicit clause addition steps.
-  A clause can only be added if it is logically implied by the existing clauses.
-  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
-\item
-  The file contains explicit clause deletion steps.
-  A clause can only be deleted if it is logically implied by the remaining clauses.
-  A sequence of clause identifiers must be listed as a \emph{hint} providing a RUP verification of the implication.
-\item The checker must track the dependency set for every input and
-  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
-  The associated extension variable has a dependency set equal to the union of those of its arguments.
-\item Declaring a $\por$ operation requires a sequence of clauses
-  providing a RUP proof that the arguments are mutually exclusive.
-  Only binary $\por$ operations are allowed to avoid requiring multiple proofs of disjointness.
-%  Generalizing to $k$ arguments would require
-%  $k\,(k-1)/2$ proofs of disjointness.
-\end{itemize}
-
-\subsection{Syntax}
-\label{subsection:syntax}
-
-\begin{table}
-  \caption{CPOG Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
-  \label{tab:cpog:syntax}
-\centering{
-  \begin{tabular}{lllll}
-    \toprule
-    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
-    \midrule
-    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\texttt{a}}  & \makebox[15mm][l]{$L^{*}$ \texttt{0}} & \makebox[15mm][l]{$C^{+}$ \texttt{0}}  & \makebox[20mm][l]{Add RUP clause} \\
-     & \texttt{d} & $C$             & $C^{+}$  \texttt{0} & Delete RUP clause \\
-    \midrule
-    $C$    & \texttt{p} & $V \; L^{*}$ \texttt{0}    &                  & Declare $\pand$ operation \\
-    $C$    & \texttt{s} & $V \; L \; L$    & $C^{+}$ \texttt{0}  & Declare $\por$ operation \\
-    \midrule
-     & \texttt{r} & $L$             &            & Declare root literal\\
-    \bottomrule
-  \end{tabular}
-  }
-\end{table}
-
-Table~\ref{tab:cpog:syntax} shows the declarations that can occur in a CPOG file.
-%% The checker is provided with the input formula as a separate file.
-As with other clausal proof formats, a variable is
-represented by a positive integer $v$, with the first ones being input
-variables and successive ones being extension variables.  Literal $\lit$
-is represented by a signed integer, with $-v$ being the logical negation of
-variable $v$.  Each clause is indicated by a positive integer
-identifier $C$, with the first ones being the IDs of the input clauses and successive
-ones being the IDs of added clauses.  Clause identifiers must be defined in order,
-with any clause identifier $C'$ given in the hint when adding clause $C$ having $C' < C$.
-
-The first set of proof rules are similar to those in other clausal
-proofs.
-Clauses can be added via RUP addition
-(command \texttt{a}), with a sequence of antecedent clauses (the
-``hint'').
-Similarly for clause deletion (command \texttt{d}).
-
-\begin{table}
-\caption{Defining Clauses for Product (A) and Sum (B) Operations}
-\begin{minipage}{0.54\textwidth}
-\begin{center}
-\begin{tabular}{cccccc}
-\multicolumn{6}{c}{(A).  Product Operation $\pand$}\\
-\toprule
-\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
-\midrule
-  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
-  $i\!+\!1$ & $-v$ & $\lit_1$  \\
-  $i\!+\!2$ & $-v$ & $\lit_2$  \\
-  & $\ldots$ \\
-  $i\!+\!k$ & $-v$ & $\lit_k$  \\
-\bottomrule
-\end{tabular}
-\end{center}
-\end{minipage}
-\begin{minipage}{0.44\textwidth}
-\begin{center}
-\begin{tabular}{cccc}
-\multicolumn{4}{c}{(B).  Sum Operation $\por$}\\
-\toprule
-\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
-\midrule
-  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
-  $i\!+\!1$ & $v$ & $-\lit_1$ \\
-  $i\!+\!2$ & $v$ & $-\lit_2$ \\
-\bottomrule
-$\;$ \\
-$\;$ \\
-\end{tabular}
-\end{center}
-\end{minipage}
-\label{tab:defining}
-\end{table}
-
-The declaration of a \emph{product} operation, creating a node with operation $\pand$,
- has the form:
-\begin{center}
-\begin{tabular}{ccccccccc}
-  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
-  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\texttt{0}} \\
-\end{tabular}
-\end{center}
-Integer $i$ is a new clause ID, $v$ is a positive integer that does not
-correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$
-integers, indicating the arguments as literals of existing variables.
-As Table~\ref{tab:defining}(A) shows,
-this declaration implicitly causes $k+1$ clauses to be added to the proof, providing a Tseitin encoding that defines extension variable $v$ as the product of its arguments.
-
-The dependency sets for the arguments represented by each pair of
-literals $\lit_i$
-and $\lit_{j}$ must
-be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
-representing Boolean constant $1$.  The only clause added to the proof will be
-the unit literal $v$.  A reference to literal $-v$ then provides a way
-to represent constant $0$.
-
-The declaration of a \emph{sum} operation, creating a node with operation $\por$, has the form:
-\begin{center}
-\begin{tabular}{ccccccc}
-  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$}
-\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
-\end{tabular}
-\end{center}
-Integer $i$ is a new clause ID, $v$ is a positive integer that does
-not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
-signed integers, indicating the arguments as literals of existing variables.  Hint $H$
-consists of a
-sequence of clause IDs, all of which must be defining clauses for other POG operations.\footnote{The restriction to defining clauses in the hint is critical to soundness.
-Allowing the hint to include the IDs of input clauses creates an exploitable weakness.  We discovered this weakness in the course of our efforts at formal verification.}
-As Table~\ref{tab:defining}(B) shows,
-this declaration implicitly causes three clauses to be added to the proof, providing a Tseitin encoding that defines extension variable $v$ as the sum of its arguments.
-The hint must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.
-
-Finally, the literal denoting the root of the POG is declared with the
-\texttt{r} command.  It can occur anywhere in the file.  Except in degenerate cases, it
-will be the extension variable representing the root of a graph.
-%, but in
-%degenerate cases it can be the positive or negative literal of an
-%input variable.
-
-\subsection{Semantics}
-\label{subsection:semantics}
-
-As was described in Section~\ref{sect:clausal}, the defining clauses
-for the product and sum operations uniquely define the values of their
-extension variables for any assignment of values to the argument
-variables.  That is, for assignment $\assign$ to the variables in
-$\varset$, the defining clauses induce a unique assignment $\eassign$
-to all data and extension variables.
-Every POG node $\nodeu$ represents POG formula $\phi_{\nodeu}$ and has an associated extension variable $u$.
-We can prove that
-for any total assignment $\assign$ to the input variables, we will have
-$\eassign(u) = 1$ if and only if $\assign \in \modelset(\phi_\nodeu)$.
-
-%% A CPOG proof follows the same general form as a QRAT dual
-%% proof~\cite{bryant:cade:2021}---one that ensures that each clause
-%% addition and each clause deletion preserves equivalence.  With CPOG,
-%% however, clauses are defined both explicitly and implicitly.  Starting
-%% with the set of input clauses, the proof consists of a sequence of
-%% steps that both add and delete clauses.  Each addition must be truth
-%% preserving, that is, any satisfying total assignment for the set of clauses
-%% before clause addition should still be a satisfying assignment afterwards.
-%% Each deletion must be falsehood preserving.  That is, there can be no
-%% new satisfying assignments when the clause is deleted.
-
-The sequence of operator declarations, asserted clauses, and
-clause deletions represents a systematic transformation of the input formula $\inputformula$
-into a POG\@.  Validating all of these steps serves to prove that
-POG $P$ is logically equivalent to the input formula.
-At the completion of the proof, the following \textsc{final conditions} must hold:
-\begin{enumerate}
-\item There is exactly one remaining clause that was added via RUP
-  addition, and this is a unit clause consisting of root literal $r$.
-\item All of the input clauses have been deleted.
-\end{enumerate}
-In other words, at the end of the proof it must hold that the active
-clauses be exactly those in $\pogformula \doteq \{\{r\}\} \cup \;
-\bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting of unit
-clause $\{r\}$ and the defining clauses for the nodes, providing a
-Tseitin encoding of $P$.  By our invariant, we are guaranteed that
-$\inputformula \eequiv \pogformula$.  That is, for any total
-assignment $\assign$ to the input variables, its unique extension 
-$\eassign$ to the POG variables will satisfy $\eassign(r) = 1$ if and only if
-$\assign \in \modelset(\inputformula)$.
-
-The sequence of clause addition steps provides a \emph{forward implication} proof that
-$\assign \in \modelset(\inputformula) \imply \eassign(r) = 1$.  That is, any total
-assignment $\assign$ satisfying the input formula must, when extended, also satisfy
-the formula represented by the POG\@.
-Conversely,
-the sequence of clause deletion steps that delete all intermediate added clauses and all input clauses
-provides a \emph{reverse implication} proof:
-$\eassign(r) = 1 \imply \assign \in \modelset(\inputformula)$.
-It does so by contradiction, proving that when $\eassign(r) = 0$, we must have $\assign \not \in \modelset(\inputformula)$.
-
-\subsection{CPOG Example}
-\label{sect:cpog:example}
-
-\begin{figure}
-\begin{minipage}{0.58\textwidth}
-(A)  Input Formula\\[1.2ex]
-\begin{tabular}{lll}
-\toprule
-\makebox[5mm]{ID} & \makebox[15mm]{Clauses} & \\
-\midrule
-\rtext{1} & \texttt{-1 3 -4} & \texttt{0} \\
-\rtext{2} & \texttt{-1 -3 4} & \texttt{0} \\
-\rtext{3} & \texttt{3 -4} & \texttt{0}\\
-\rtext{4} & \texttt{1 -3 4} & \texttt{0} \\
-\rtext{5} & \texttt{-1 -2} & \texttt{0} \\
-\bottomrule
-\end{tabular}
-\\[1.8ex]
-(C) POG Declaration\\[1.2ex]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
-\midrule
-\rtext{6} & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
-\rtext{9} & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
-\rtext{12} & \texttt{s 7 5 6} \; \rtext{7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
-\rtext{15} & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\
-\rtext{18} & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
-\rtext{22} & \texttt{s 10 8 9} \; \rtext{16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
- & \texttt{r 10} && Root $r = s_{10}$\\
-\bottomrule
-\end{tabular}
-\end{minipage}
-\begin{minipage}{0.35\textwidth}
-(B) POG Representation \\
-\input{dd/eg4}
-\end{minipage}
-%% Break
-\\[2.5ex]
-\begin{minipage}{0.42\textwidth}
-(D) Defining Clauses\\[1.2ex]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
-\midrule
-\rtext{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
-\rtext{7} & \texttt{-5 -3} & \texttt{0} & \\
-\rtext{8} & \texttt{-5 -4} & \texttt{0} & \\
-\midrule
-\rtext{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
-\rtext{10} & \texttt{-6 3} & \texttt{0} & \\
-\rtext{11} & \texttt{-6 4} & \texttt{0} & \\
-\midrule
-\rtext{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
-\rtext{13} & \texttt{7 -5} & \texttt{0} & \\
-\rtext{14} & \texttt{7 -6} & \texttt{0} & \\
-\midrule
-\rtext{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
-\rtext{16} & \texttt{-8 -1} & \texttt{0} & \\
-\rtext{17} & \texttt{-8 7} & \texttt{0} & \\
-\midrule
-\rtext{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
-\rtext{19} & \texttt{-9 1} & \texttt{0} & \\
-\rtext{20} & \texttt{-9 -2} & \texttt{0} & \\
-\rtext{21} & \texttt{-9 7} & \texttt{0} & \\
-\midrule
-\rtext{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
-\rtext{23} & \texttt{10 -8} & \texttt{0} & \\
-\rtext{24} & \texttt{10 -9} & \texttt{0} & \\
-\bottomrule
-\end{tabular}
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-(E) CPOG Assertions\\[1.2ex]
-\begin{tabular}{llllll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
-\midrule
-\rtext{25} & \texttt{a 5 1 3} & \texttt{0} & \rtext{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
-\rtext{26} & \texttt{a 6 1 -3} & \texttt{0} & \rtext{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
-\rtext{27} & \texttt{a 3 7 1} & \texttt{0} & \rtext{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
-\rtext{28} & \texttt{a 7 1} & \texttt{0} & \rtext{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
-\rtext{29} & \texttt{a 8 1} & \texttt{0} & \rtext{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
-\rtext{30} & \texttt{a 5 -1 3} & \texttt{0} & \rtext{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
-\rtext{31} & \texttt{a 6 -1 -3} & \texttt{0} & \rtext{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
-\rtext{32} & \texttt{a 3 7 -1} & \texttt{0} & \rtext{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
-\rtext{33} & \texttt{a 7 -1} & \texttt{0} & \rtext{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
-\rtext{34} & \texttt{a 9 -1} & \texttt{0} & \rtext{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
-\rtext{35} & \texttt{a 1 10} & \texttt{0} & \rtext{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
-\rtext{36} & \texttt{a 10} & \texttt{0} & \rtext{35 24 34} & \texttt{0} & $s_{10}$ \\
-\bottomrule
-\end{tabular}
-\\[1.5ex]
-(F) Input Clause Deletions\\[1.2ex]
-\begin{tabular}{lllll}
-  \toprule
- \multicolumn{3}{l}{CPOG line} & Explanation\\
-\midrule
- \texttt{d 1} & \rtext{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
- \texttt{d 2} & \rtext{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
- \texttt{d 3} & \rtext{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
- \texttt{d 4} & \rtext{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
- \texttt{d 5} & \rtext{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
-\bottomrule
-\end{tabular}
-\end{minipage}
-\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F).  The defining clauses (D) are implicitly
-defined by the POG declaration (C).}
-\label{fig:eg4:all}
-\end{figure}
-
-Figure \ref{fig:eg4:all} illustrates an example formula and shows how
-the CPOG file declares its POG representation.  The input formula (A)
-consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
-$x_4$.  The generated POG (B) has six nonterminal nodes representing
-four products and two sums.  We name these by the node
-type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
-extension variable.
-% We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
-  The first part of the CPOG file (C) declares
-these nodes using clause IDs that increment by three or four,
-depending on whether the node has two children or three.  The last two
-nonzero values in the sum declarations are the hint providing the
-required mutual exclusion proof.
-
-\subsection{Node Declarations}
-
-We step through portions of the file to provide a better understanding of the CPOG proof framework.
-Figure
-\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
-defined by the POG operation declarations.  These do not appear in the
-CPOG file.  Referring back to the declarations of the sum nodes in
-Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
-$\nodes_7$ has clause IDs 7 and 10 as the hint.  We can see in Figure
-\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
-$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
-have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
-clause IDs 16 and 19 as the hint.  These form a RUP proof for the clause
-$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
-$\nodes_{10}$ have disjoint models.
-
-\subsection{Forward Implication Proof}
-
-Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
-leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $\nodes_{10}$ is implied by the input clauses, i.e.,
-any total assignment $\assign$
-satisfying the input clauses must have its extension to $\eassign$ yield $\eassign(s_{10}) = 1$.
-Working backward, we can see that
-clause 35 indicates that variable $s_{10}$ will be implied by the input
-clauses when $\assign(x_1) = 0$. Clause 34 indicates that node $p_9$ will
-be implied by the input clauses when $\assign(x_1) = 1$, while defining clause 24 shows that node $s_{10}$ will be implied
-by the input clauses when $\eassign{p_9} = 1$.  These three clauses
-serve as the
-hint for clause 36.
-
-\subsection{Reverse Implication Proof}
-
-Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
-delete the input clauses.  Consider the first of these, deleting
-input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
-that there is no total assignment $\assign$ that falsifies this clause but extends to $\eassign$ such that  $\assign(s_{10}) = 1$.
-The proof proceeds by first assuming that the clause is false, requiring
-$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consists of unit
-clauses (e.g., clause 36 asserting that $\eassign(p_{10}) = 1$) or
-clauses that cause unit propagation.  Hint clauses 8 and 10 force the
-assignments $\eassign(p_5) = \eassign(p_6) = 0$.  These, plus hint clause 12 force
-$\eassign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\eassign(p_8) = \eassign(p_9) = 0$, leading,
-via clause 22, to $\eassign(s_{10}) = 0$.  But this contradicts clause 36,
-completing the RUP proof.  The deletion hints for the other input
-clauses follow similar patterns---they work from the bottom nodes of
-the POG upward, showing that any total assignment that falsifies the clause
-must, when extended, have $\eassign(s_{10}) = 0$.
-
-Deleting the asserted clauses is so simple that we do not show it.  It
-involves simply deleting the clauses from clause number 35 down to
-clause number 25, with each deletion using the same hint as was used
-to add that clause.  In the end, therefore, only the defining clauses
-for the POG nodes and the unit clause asserting $s_{10}$ remain,
-completing a proof that the POG is logically equivalent to the input
-formula.
-
-\section{Generating CPOG from dec-DNNF}
-\label{sect:generating:cpog}
-
-%%Figure~\ref{fig:chain} illustrates one method of generating a CPOG
-%%file from an input formula in conjunctive normal form.  Our toolchain
-%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
-%%but it would also work for other tools that use a top-down search
-%%strategy to generate a dec-DNNF representation.
-A dec-DNNF
-graph can be directly translated  into a POG.
-In doing this conversion,
-our program performs simplifications to
-eliminate Boolean constants.
-Except in degenerate cases,
-where the formula is unsatisfiable or a tautology,
-we can therefore assume
-that the POG does not contain any constant nodes.
-In addition, negation is only
-applied to variables, and so the only edges with negative polarity will have variables as children.
-We can therefore
-view the POG as consisting
-of \emph{literal} nodes corresponding to input variables and their negations, along with
-\emph{nonterminal} nodes, which can be further classified as \emph{product} and \emph{sum} nodes.
-
-\subsection{Forward Implication Proof}
-
-For input formula $\inputformula$ and its translation into a POG $P$
-with root node $\noder$, the most challenging part of the proof is to
-show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
-that any total assignment $\assign$ that is a model of $\inputformula$ will extend to assignment $\eassign$
-such that $\eassign(r) = 1$, for root literal $r$.  This part of the
-proof consists of a series of clause assertions leading to one adding
-$\{r\}$ as a unit clause.  We have devised two methods for generating this
-proof.  The \emph{monolithic} approach makes just one call to a
-proof-generating SAT solver and has it determine the relationship
-between the two representations.  The \emph{structural} approach
-recursively traverses the POG, generating proof obligations at each
-node encountered.  It may require multiple calls to a proof-generating SAT
-solver.
-
-As notation,
-let $\psi$ be a subset of the clauses in $\inputformula$.
-For partial assignment
-$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
-obtained from $\psi$ by: (i) eliminating any
-clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
-(ii) for the remaining clauses eliminating those literals $\lit$ for
-which $\passign(\lit) = 0$, and (iii) eliminating any duplicate or tautological clauses.
-In doing these simplifications, we also track the \emph{provenance}
-of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
-More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov_{\passign}(C, \psi)$ denote
-those clauses $C' \in \psi$, such that
-$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
-We then extend the definition of $\prov$ to any simplified formula
-$\gamma$ as $\prov_{\passign}(\gamma, \psi) = \bigcup_{C \in \gamma} \prov_{\passign}(C, \psi)$.
-
-The monolithic approach
-takes advantage of the clausal representations of
-the input formula $\inputformula$ and the POG formula $\phi_\noder$.
-We can express the negation of $\phi_\noder$ in clausal form as
-$\theta_{\obar{\noder}} \doteq \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
-Forward implication will hold when $\inputformula \imply \phi_\noder$, or  equivalently
-when the formula $\inputformula \land \theta_{\obar{\noder}}$
-is unsatisfiable, where the
-conjunction can be expressed as the union
-of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
-For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CPOG proof, and so the empty clause in the proof becomes the unit clause $\{r\}$.
-Our experimental results show
-that this approach can be very effective and generates short proofs
-for smaller problems, but it does not scale well enough for general
-use.
-
-The structural approach to proof generation takes the form of a recursive procedure
-$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
-node $\nodeu$, partial assignment
-$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
-The procedure adds a number of clauses to the proof, culminating with
-the addition of the \emph{target} clause:
-$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
-that $(\bigwedge_{\lit \in \passign} \lit) \imply u$, i.e.,
-that any total
-assignment $\assign$ such that $\passign \subseteq \assign$
-will extend to assignment $\eassign$ such that $\eassign(u) = 1$.
-The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
-The result will therefore be to add unit clause $\{r\}$ to the proof.
-Here we present a correct, but somewhat inefficient formulation of
-$\validate$.  We then refine it with some optimizations.
-
-The recursive call $\validate(\nodeu, \passign, \psi)$ assumes that we have
-traversed a path from the root node down to node $\nodeu$, with the
-literals encountered in the product nodes forming the partial
-assignment $\passign$.  The set of clauses $\psi$ can be a proper
-subset of the input clauses $\inputformula$ when a product node has caused
-a splitting into clauses containing disjoint variables.
-The subgraph with root node $\nodeu$ should be a POG representation of the formula
-$\simplify{\psi}{\passign}$.
-%% Having the
-%% routine add its target clause indicates that any total assignment
-%% $\assign$ such that $\passign \subseteq \assign$ will yield $\alpha(u) = 1$.
-
-The process for generating such a proof depends on the form of node $\nodeu$:
-\begin{enumerate}
-\item If $\nodeu$ is a literal $\lit'$, then the formula
-  $\simplify{\psi}{\passign}$ must consist of the single unit clause
-  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
-  Any of these can
-  serve as the target clause.
-\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
-  then, since the node originated from a dec-DNNF graph, there must be
-  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
-  product node containing a literal node for $x$ as a child.  In either case, we
-  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
-  This will cause the addition of the target clause
-  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
-  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
-  causing the addition of the target clause
-  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
-  These recursive results can be combined with the second and third defining clauses for $\nodeu$
-(see Table~\ref{tab:defining}(B))
-  to generate the target clause for $\nodeu$, requiring at most two RUP steps.
-\item If $\nodeu$ is a product node, then we can divide its children
-  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
-  \begin{enumerate}
-    \item For each literal
-  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
-  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
-  cases, this can be done by simple Boolean constraint propagation (BCP).
-  In other cases, we must prove that the formula
-  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
-  do so by writing the formula to a file, invoking a proof-generating
-  SAT solver, and then converting the generated unsatisfiability proof
-  into a sequence of clause additions in the CPOG file.
-  (The solver is constrained to only use RUP inference rules, preventing it from introducing extension variables.)
-\item For a single nonterminal child ($k = 1$), we recursively call
-  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
-\item For multiple nonterminal children ($k > 1$),
-  it must be the case that the clauses in
-  $\gamma = \simplify{\psi}{\passign \cup \lambda}$ can be partitioned into $k$ subsets
-  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
-  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
-  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
-  literals.
-  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
-  We recursively call
-  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
-\end{enumerate}
-  We then generate the target clause for node $\nodeu$ with a single RUP step,
-creating the hint by combining the results from the BCP and SAT calls for
-  the literals, the recursively computed target clauses, and all but
-  the first defining clause for node $\nodeu$
-(see Table~\ref{tab:defining}(A)).
-\end{enumerate}
-Observe that
-all of these steps involve a polynomial number of
-operations per recursive call, with the exception of those that call
-a SAT solver to validate a literal.
-
-As examples, the forward implication proof of Figure~\ref{fig:eg4:all}(E) was
-generated by the structural approach.  Working from step 36 backward,
-we can see that steps 35 and 36 complete the call to
-$\validate(\nodes_{10}, \emptyset, \phi_I)$.  This call used $x_1$ as
-the splitting variable, first calling $\validate(\nodep_{8},
-\{\obar{x}_1\}, \emptyset, \phi_I)$, which completed with step 29, and
-$\validate(\nodep_{9}, \{x_1\}, \phi_I)$, which completed with step
-34.  We see that each of these calls required separate traversals of
-nodes $\nodes_7$, $\nodep_6$, and $\nodep_5$, with the former yielding
-proof steps 25--27 and the latter yielding proof steps 30--32.  This
-demonstrates how our simple formulation of $\validate$ effectively
-expands the graph into a tree.  This shortcoming is avoided by the use of lemmas, as is described in Section~\ref{sect:lemma}.
-
-\subsection{Reverse Implication Proof}
-
-Completing the equivalence proof of input formula $\inputformula$ and its POG
-representation with root node $\noder$ requires showing that
-$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
-CPOG framework by first deleting all asserted clauses, except for the
-final unit clause for root literal $r$, and then deleting all of the
-input clauses.
-%%  Suppose at some point in this process that we want to
-%% delete clause $C$ such that the remaining set of clauses will form the
-%% CNF formula $\phi$.  Deleting $C$ requires proving that
-%% $\phi \imply C$.
-
-The asserted clauses can be deleted in reverse order, using the same
-hints that were used in their original assertions.  By reversing the
-order, those clauses that were used in the hint when a clause was
-added will still remain when it is deleted.
-
-Each input clause deletion can be done as a single RUP step, based
-on an algorithm to test for clausal entailment in det-DNNF graphs~\cite{darwiche:jair:2002,capelli:sat:2019}.  The
-proof generator constructs the hint sequence from the defining
-clauses of the POG nodes via a single, bottom-up pass through the
-graph.  The RUP deletion proof for input clause $C$ effectively proves that any
-total assignment $\assign$ that does not
-satisfy $C$ will extend to assignment $\eassign$ such that
-$\eassign(r) = 0$.  It starts with the set of literals
-$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
-assignment $\assign$ to falsify clause $C$.
-It then
-adds literals via unit propagation until a
-conflict arises.    Unit literal $r$ gets
-added right away, setting up a potential conflict.
-
-Working upward through the graph, node $\nodeu$ is \emph{marked} when
-the collected set of literals forces $\eassign(u) = 0$.  When marking $\nodeu$, the
-program adds $\obar{u}$ to the RUP literals and adds the appropriate
-defining clause to the hint.  A literal node for
-$\lit$ will be marked if $\lit \in C$, with no hint required.  If
-product node $\nodeu$ has some child $\nodeu_i$ that is marked, then
-$\nodeu$ is marked and clause $i+1$ from among its defining clauses (see Table~\ref{tab:defining}(A)) is
-added to the hint.  Marking sum node $\nodeu$ requires that its two children are marked.
-The first defining
-clause for this node (see Table~\ref{tab:defining}(B)) will then be added to the hint.  At the very end, the program
-(assuming the reverse implication holds) will attempt to mark root
-node $\noder$, which would require $\eassign(r) = 0$, yielding a
-conflict.
-
-It can be seen that the reverse implication proof will be polynomial in the size of the POG\@, because
-each clause deletion requires a single RUP step having a hint with length
-bounded by the number of POG nodes.
-
-\section{Optimizations}
-
-The performance of the structural proof generator for forward implication, both in its execution time and
-the size of the proof generated, can be improved by two optimizations
-described here.  A key feature is that they do not require any changes
-to the proof framework---they build on the power of extended
-resolution to enable the construction of new logical structures.  They
-involve declaring new product nodes to encode products of literals.
-These nodes are not part of the POG
-representation of the formula; they serve only to enable the forward
-implication proof.
-
-The combination of these two optimization guarantees that (i) each call
-to $\validate$ for a product node will cause at most one invocation of
-the SAT solver, and (ii) each call to $\validate$ for any node $\nodeu$
-will cause further recursive calls only once.  Our experimental
-results show that these
-optimizations yield substantial benefits.
-
-\subsection{Literal Grouping}
-\label{sect:grouping}
-
-A single recursive step of $\validate$ can encounter product nodes
-having many literals as children.  The naive formulation of $\validate$
-considers each literal $\lit \in \lambda$ separately.
-Literal grouping allows all literals to be validated with a single call to a SAT solver.
-It collects those literals
-$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
-product node $\nodev$ having these literals as children.  The goal
-then becomes to prove that any total assignment $\assign$ consistent with the partial assignment $\passign$,
-must, when extended to $\eassign$, yield $\eassign(v) = 1$.
-A single call to the solver can generate this proof by invoking it on the formula
-  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
-  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
-  input clauses and the defining clauses for $\nodev$ into the hints.
-
-\subsection{Lemmas}
-\label{sect:lemma}
-
-As we have noted, the recursive calling of $\validate$ starting at
-root $\noder$ effectively expands the POG into a tree, and this can
-lead to an exponential number of calls.
-These shared subgraphs arise when the knowledge compiler employs \emph{clause caching}
-to detect that the simplified set of
-clauses arising from one partial assignment to the literals matches that
-of a previous partial assignment~\cite{darwiche:aaai:2002}.
-When this dec-DNNF node is translated into POG
-node $\nodeu$, the proof generator can assume (and also check), that
-there is a simplified set of clauses $\gamma_{\nodeu}$
-for which the subgraph with root $\nodeu$ is its POG representation.
-
-The proof generator can exploit the sharing of subgraphs
-by constructing and proving a \emph{lemma} for each node
-$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
-total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ must extend to assignment $\eassign$ such that
-$\eassign(u) = 1$.  This lemma is then invoked for every node having
-$\nodeu$ as a child.
-As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.
-
-The challenge for implementing this strategy is to find a way to
-represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CPOG file.  Some may be
-unaltered input clauses, and these can be used directly.  Others,
-however will be clauses that do not appear in the input formula.  We
-implement these by adding POG product nodes to the CPOG file to create
-the appropriate clauses.  Consider an \emph{argument} clause
-$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
-define a product node $\nodev$ with arguments
-$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, 
-we will introduce a defining clause
-$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
-  synthetic} clause having $\obar{v}$ as the \emph{guard literal}.
-That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
-  activate} the clause, causing it to represent argument clause $C$.  On the other
-hand, a partial assignment with $\passign(v) = 1$ will
-cause the clause to become a tautology and therefore have no effect.
-
-Suppose for every clause $C_j \in \gamma_{\nodeu}$ that does not correspond to
-an input clause, we generate a synthetic clause $C'_j$ with guard literal
-$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma'_{\nodeu}$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
-while input clauses in $\gamma_{\nodeu}$ are left unchanged.
-Let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
-Invoking $\validate(\nodeu, \lassign, \gamma'_{\nodeu})$
- will then prove a lemma, given by the target clause
- $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
- showing that any total assignment $\assign$ that activates the synthetic clauses will, when cause $u$ to be assigned $1$.
- More precisely, given assignment $\assign$ and its extension $\eassign$, if $\eassign(v_j) = 0$ for every guard literal $\obar{v}_j$, then $\eassign(u) = 1$.
-
-Later, when node $\nodeu$ is encountered by a call to $\validate(\nodeu, \passign, \psi)$, we invoke the lemma
-by showing that each synthetic clause
-$C_j$ matches some simplified clause in $\simplify{\psi}{\passign}$.  More precisely,
-for $1 \leq j \leq m$,
-we use clause addition to assert the clause
-$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$,
-showing that synthetic clause $C_j$ will be activated.
-Combining the lemma with these activations provides a derivation of the target clause for the call to $\validate$.
-
-Observe that the lemma structure can be hierarchical, since a shared
-subgraph may contain nodes that are themselves roots of shared
-subgraphs.  Even then, the principles described allow the
-definition, proof, and applications of a lemma for each shared node in
-the graph.  For any node $\nodeu$, the first call to
-$\validate(\nodeu, \passign, \psi)$ may require further recursion,
-but any subsequent call can simply reuse the lemma proved by the first call.
-
-\subsection{Lemma Example}
-\label{app:lemma:eg}
-
-\begin{figure}
-(A) Additional nodes\\[1.0em]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
-\midrule
-\rtext{25} & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
-\rtext{28} & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
-\bottomrule
-\end{tabular}
-\\[1.0em]
-(B) Implicit Clauses\\[1.2em]
-\begin{tabular}{llll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
-\midrule
-\rtext{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
-\rtext{26} & \texttt{-11 -3} & \texttt{0} & \\
-\rtext{27} & \texttt{-11 4} & \texttt{0} & \\
-\midrule
-\rtext{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
-\rtext{29} & \texttt{-12 3} & \texttt{0} & \\
-\rtext{30} & \texttt{-12 -4} & \texttt{0} & \\
-\bottomrule
-\end{tabular}
-\\[1.0em]
-(C) CPOG Assertions\\[1.0em]
-\begin{tabular}{llllll}
-\toprule
-\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
-\midrule
-\multicolumn{6}{l}{Lemma Proof} \\
-\rtext{31} & \texttt{a 5 11 12 3} & \texttt{0} & \rtext{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
-\rtext{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \rtext{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
-\rtext{33} & \texttt{a 3 7 11 12} & \texttt{0} & \rtext{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
-\rtext{34} & \texttt{a 7 11 12} & \texttt{0} & \rtext{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
-\midrule
-\multicolumn{6}{l}{Lemma Application \#1} \\
-\rtext{35} & \texttt{a -11 1} & \texttt{0} & \rtext{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
-\rtext{36} & \texttt{a -12 1} & \texttt{0} & \rtext{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
-\rtext{37} & \texttt{a 7 1} & \texttt{0} & \rtext{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
-\midrule
-\rtext{38} & \texttt{a 8 1} & \texttt{0} & \rtext{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
-\midrule
-\multicolumn{6}{l}{Lemma Application \#2} \\
-\rtext{39} & \texttt{a -11 -1} & \texttt{0} & \rtext{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
-\rtext{40} & \texttt{a -12 -1} & \texttt{0} & \rtext{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
-\rtext{41} & \texttt{a 7 -1} & \texttt{0} & \rtext{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
-\midrule
-\rtext{42} & \texttt{a 9 -1} & \texttt{0} & \rtext{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
-\rtext{43} & \texttt{a 1 10} & \texttt{0} & \rtext{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
-\rtext{44} & \texttt{a 10} & \texttt{0} & \rtext{43 24 42} & \texttt{0} & $s_{10}$ \\
-\bottomrule
-\end{tabular}
-\caption{Example of lemma definition, proof, and application}
-\label{fig:eg4:lemmas}
-\end{figure}
-
-Figure~\ref{fig:eg4:lemmas} shows an alternate forward implication
-proof for the example of Figure~\ref{fig:eg4:all} using a lemma to
-represent the shared node $\nodes_7$.  We can see that the POG with
-this node as root encodes the Boolean formula $x_3 \leftrightarrow x_4$, having a CNF representation consisting of the clauses
-$\{x_3 ,\, \obar{x}_4\}$ and $\{\obar{x}_3 ,\, {x}_4\}$.  The product node
-declarations shown in Figure~\ref{fig:eg4:lemmas}(A) create synthetic
-clauses 25 and 28 to encode these arguments with activating literals
-$\obar{v}_{11}$ and $\obar{v}_{12}$, respectively.  Clauses 31--34
-then provide a proof of the lemma, stating that any assignment
-$\assign$ that activates these clauses will, when extended,  assign $1$ to $s_7$.
-Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
-will, when extended, cause the first synthetic clause to activate due to input clause
-3, and it will cause the second synthetic clause to activate due to
-input clause 4.  From this, clause 37 can use the lemma to state that
-assigning $0$ to $x_1$ will cause $s_7$ to evaluate to $1$.  Similarly,
-clauses 39 and 40 serve to activate the synthetic clauses when
-$\assign(x_1) = 1$, due to input clauses 1 and 2, and clause 41 then
-uses the lemma to state that assigning $1$ to $x_1$ will cause $s_7$ to
-evaluate to $1$.
-
-In this example, adding the lemma increases the proof length, but that
-is only because it is such a simple formula.
-
-%%\input{lean}
-
-\input{lean-section}
-
-\section{Implementation}
-We have implemented programs that, along with
-the \dfour{} knowledge compiler, form the toolchain illustrated in
-Figure~\ref{fig:chain}.\footnote{The source code for all tools
-is available at \url{https://github.com/rebryant/cpog/releases/tag/v1.0.0}.}  The proof generator is the same in both
-cases, since it need not be trusted.
-Our \emph{verified}
-version of the proof checker and ring evaluator have been formally
-verified within the \lean{} theorem prover.  Our long term goal is to
-rely on these.  Our \emph{prototype} version is written in C\@.
-It is faster and
-more scalable, but we anticipate its need will diminish as the
-verified version is further optimized.
-%% Our long-term goal is to have a toolchain suitable for use in a
-%% knowledge compiler competition, along the lines of the annual
-%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
-%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
-%% Our experiments
-%% consider the benchmark problems and computing capabilities that would
-%% arise in such a competition and evaluate whether our implementationed toolchain
-%% could handle these problems.  As with the original proof-generating
-%% SAT competitions, we assume that the proofs at a competition could be
-%% checked with the unverified checker.  Having a verified checker
-%% that works on some subset of the benchmarks would provide an
-%% additional degree of certainty.
-
-Our proof generator is written in C/C++ and uses a recent version of
-the \cadical{} SAT solver that directly generates hinted proofs in
-LRAT format~\cite{biere:sat:2023}.  It also uses their tool \ltrim{}
-to reduce the length of the generated proofs.
-
-The proof generator can optionally be instructed to generate a {\em
-one-sided} proof, providing only the reverse-implication portion of the proof via
-input clause deletion.  This can provide useful information---any
-assignment that is a model for the compiled representation
-must also be a model for the input formula---even when
-full validation is impractical.
-
-We incorporated a ring evaluator into the prototype checker.  It can
-perform both standard and weighted model counting with full precision.
-It performs arithmetic over a subset of the rationals we call
-$\drational$, consisting of numbers of the form $a \cdot 2^{b} \cdot
-5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented to
-have arbitrary range.  Allowing scaling by powers of 2 enables the
-density computation and rescaling required for standard model
-counting.  Allowing scaling by powers of both 2 and 5 enables exact
-decimal arithmetic, handling the weights used in the weighted model
-counting competitions.  To give a sense of scale, the counter
-generated a result with 260,909 decimal digits for one of the weighted
-benchmarks.  Our implementation of arbitrary-range integers represents
-a number as a sequence of ``digits'' with each digit ranging from $0$
-to $10^9-1$, and with the digits stored as four-byte blocks.  This
-allows easy conversion to and from a decimal representation of the
-number.
-
 \section{Experimental Evaluation}
@@
 compilation was preceded by preprocessing.
-
-\section{Extensions}
-
-We are hopeful that having checkable proofs for knowledge compilers
-will allow them to be used in applications where high levels of trust
-are required, and that it will provide a useful tool for developers of
-knowledge compilers.  Our current implementation only handles the
-outputs of knowledge compilers that generate dec-DNNF graphs, and it
-supports only queries that can be computed via ring evaluation.  Here
-we discuss ways to extend both capabilities.
-
-\subsection{Validating Sentential Decision Diagrams}
-\label{sect:sdd}
-
-\begin{figure}
-  \begin{center}
-    \input{dd/sdd}
-  \end{center}
-\caption{Translation of SDD node $\nodeu$ into POG node $\hat{\nodeu}$.  An SDD node with $2k$ children expands into $k$ product nodes and $k-1$ sum nodes.}
-\label{fig:sdd:pog}
-\end{figure}
-
-\citeA{darwiche:ijcai:2011} introduced \emph{Sentential
-Decision Diagrams} (SDDs) as a generalization of 
-Ordered Binary Decision Diagrams (OBDDs)~\cite{Bryant:1986}.  Like
-OBDDs, SDDs are a canonical representation and support bottom-up
-construction.
-
-As with other representations, each node $\nodeu$ in an SDD denotes a
-Boolean formula $\phi_{\nodeu}$.  Leaf nodes correspond to Boolean constants and literals.
-Each nonterminal node has $2k$ children,
-organized as pairs $\nodep_i$ and $\nodes_i$ for $1\leq i \leq k$.
-The dependency set for the node must be partitioned into disjoint sets $X_p$ and $X_s$ such that
-$\dependencyset(\phi_{\nodep_i}) \subseteq X_p$ and
-$\dependencyset(\phi_{\nodes_i}) \subseteq X_s$ for all $1 \leq i \leq k$.
-Children $\nodep_1, \nodep_2, \ldots, \nodep_k$ serve as a set of
-\emph{selector} functions with mutually disjoint models.  That is,
-$\modelset(\phi_{\nodep_i}) \cap \modelset(\phi_{\nodep_j}) =
-\emptyset$ for $1 \leq i < j \leq k$.
-The Boolean formula associated with node $\nodeu$ is then
-\begin{eqnarray}
-\phi_{\nodeu} & = & \bigvee_{1 \leq i \leq k} \left[\phi_{\nodep_i} \land \phi_{\nodes_i}\right] \label{eqn:sdd}
-\end{eqnarray}
-
-An SDD can be translated into a POG, where an SDD node with $2k$
-children expands into $k$ product nodes and $k-1$ sum nodes.
-In particular, assume that each nonterminal SDD node $\nodev$ can be represented by a
-subgraph in the POG with root node $\hat{\nodev}$.  For leaf node $\nodev$, its counterpart $\hat{\nodev}$ is
-the associated terminal node.  Figure~\ref{fig:sdd:pog} then
-illustrates how node $\nodeu$ with $2k$ children would be expanded
-into a subgraph with root node $\hat{\nodeu}$.  We can see that each
-of the product nodes has arguments with disjoint dependency sets, and
-each of the sum nodes has disjoint sets of models.
-
-Several different methods could be used to generate a CPOG
-certification from an SDD\@.  For the forward implication proof, the
-monolithic approach we have used could generate the proof via a single
-call to a proof-generating SAT solver.  For an SDD generated by the
-bottom up approach originally proposed by \citeA{darwiche:ijcai:2011},
-the generator could be modified to generate proof steps for each
-recursive call during the construction, as is done to generate proofs
-of unsatisfiability with proof-generating, BDD-based SAT
-solvers~\cite{Jussila:2006,ebddres,bryant:fmcad:2022,bryant:tocl:2023}.
-For an SDD generated by a top-down process, a structural method
-similar to the one we have described could be applied.  A further
-requirement would be to generate mutual exclusion proofs for the $k-1$
-sum nodes illustrated in Figure~\ref{fig:sdd:pog}.  These would be
-based on the disjoint model property of the $k$ nodes of the form
-$\hat{\nodep}_i$.  One approach would be to simply make $k-1$ calls to a
-proof-generating SAT solver.
-
-\subsection{Generalizing to Semirings}
-\label{sect:semiring}
-
-The formulation of algebraic model
-counting by \citeA{kimmig:jal:2017} is more general than ours.  It allows
-the algebraic structure to be a \emph{semiring}.  A commutative
-semiring $\semiring$ obeys all properties of a commutative ring, except that the
-elements of the set need not have additive inverses.
-We can then define the {\em semiring evaluation problem} as computing
-  \begin{equation}
-    \begin{array}{rcl}
-    \srep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:srep}
-    \end{array}
-  \end{equation}
-  where sum $\sum$ is computed according to the semiring addition operation $\radd$ and product $\prod$
-  is computed according to the semiring product operation $\rmul$.
-
-  As an example, for $|X| = n$, consider the algebraic structure
-  $\langle \{0, 1, \ldots, n\}, \max, +, 0, 0 \rangle$.
-For each variable $x$, assign weights
-$w(x) = 1$ and $w(\obar{x}) = 0$.  Then the
-weight of an assignment is the number of variables set to 1, and an
-evaluation of (\ref{eqn:srep}) gives the maximum weight of all
-satisfying assignments.
-
-Semiring evaluation can be performed via knowledge compilation
-by requiring 
-that the representation generated by
-the compiler obeys a property known as
-\emph{smoothness}~\cite{darwiche:jair:2002,shih:nips:2019}.  Within our formulation,
-a partitioned-operation formula is smooth when all arguments to each
-sum operation have identical dependency sets.  That is, every sum
-operation $\bigvee_{1 \leq i \leq k} \phi_i$ has
-$\dependencyset(\phi_i) = \dependencyset(\phi_j)$ for $1 \leq i < j
-\leq k$.  Smoothness can be ensured by adding redundant formulas to
-artificially introduce variables.  For example, if subformula $\phi_i$
-lacks having variable $x$ in its dependency set, it can be replaced by
-$(x \por \obar{x}) \pand \phi_i$.  When a knowledge compiler generates
-a representation in negation normal form that is smooth, then a semiring evaluation of the formula
-can proceed by first assigning each literal $\lit$ the value $w(\lit)$.
-Then the product and sum operations are evaluated in manners analogous to
-(\ref{eqn:ring:product}) and
-(\ref{eqn:ring:sum}).
-
-Our POG representation can support evaluation of semiring formulas by
-imposing the restriction that the POG is in negation normal form and that it is smooth.
-Given a smoothed
-dec-DNNF graph generated by a knowledge compiler, our toolchain will
-convert this into a smooth POG in normal form and verify its
-equivalence to the input formula.  Of course, full verification would also require checking that the POG is smooth.
-
-\section{Concluding Remarks}
-\label{sect:future}
-
-This paper demonstrates a method for certifying the equivalence of two
-different representations of a Boolean formula: an input formula
-represented in conjunctive normal form, and a compiled representation
-that can then be used to extract useful information about the formula,
-including its weighted and unweighted model counts.  It builds on the
-extensive techniques that have been developed for clausal
-proof systems, including extended resolution and reverse unit propagation, as well as established tools, such as
-proof-generating SAT solvers and \dtrim{}.
-
-Our experiments demonstrate that our toolchain can already handle
-problems nearly at the limits of current knowledge compilers.  Further
-engineering and optimization of our proof generator and checker could
-improve their performance and capacity substantially.
-
-%%\newpage
-
-\bibliography{references}
-\bibliographystyle{theapa}
-
-\end{document}

### 2024-01-05 doc/JAIR/jair-paper.tex
Split Experimental results as separate file

@@
 
-\section{Experimental Evaluation}
-\label{sect:experimental}
-
-%%\input{figure-combined-verify}
-
-\begin{figure}
-\begin{minipage}{0.49\textwidth}
-\centering{%
-\begin{tikzpicture}[scale = 0.50]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
-      legend style={at={(0.35,0.96)}},
-      legend cell align={left},
-                              x post scale=1.8, y post scale=2.4,
-                              xmode=log,xmin=0.01,xmax=4000,
-                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
-                              ymin=0.01, ymax=29000,
-                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
-                              xlabel={D4 runtime (seconds)},
-                              %ylabel={CPOG generation and checking runtime (seconds)},
-%                              title={D4 Defining Clause Generation}
-            ]
-%    \input{data-formatted/time-d4-verify}
-
-\addplot [only marks, color=redorange, mark=*,  mark options={scale=0.8}, opacity=0.7] coordinates { (5.758,13.334) (0.017,0.022) (0.014,0.034) (0.179,1.754) (0.016,0.046) (0.012,0.012) (0.024,0.13) (0.928,18.877) (0.069,0.624) (0.166,18.378) (1.48,21.674999999999997) (3.356,116.882) (3.184,46.001000000000005) (1.676,46.439) (0.016,0.063) (0.019,0.11100000000000002) (0.472,6.808) (0.02,0.126) (5.048,182.556) (0.398,74.411) (0.685,21.603) (0.252,3.414) (30.753,2292.7690000000002) (0.643,9.902) (2.154,36.441) (1.158,4.7219999999999995) (2.436,83.753) (1.636,132.504) (1.491,59.292) (4.683,1499.441) (30.566,4136.219) (145.967,2177.9669999999996) (2.666,40.712) (0.105,1.923) (5.508,31.97) (244.952,7273.887000000001) (0.206,8.503) (606.378,13198.325) (23.234,39.160000000000004) (0.837,182.089) (49.574,52.467) (25.077,42.550000000000004) (63.691,12374.461) (4.544,147.119) (1.679,359.63) (22.322,4493.414000000001) (2.076,61.509) (70.355,1322.4) (90.915,2908.19) (123.037,6910.268) (112.248,9602.115) (178.14,4147.301) (394.71,517.646) (158.115,180.629) (69.122,315.91) (508.972,258.496) (2.033,84.352) (270.954,433.669) (1536.914,859.557) (19.013,8856.852) (2.938,272.899) (598.709,8218.087) (9.311,105.213) (0.277,2.226) (0.101,1.086) (30.577,394.01700000000005) (0.18,2.051) (0.016,0.064) (0.021,0.134) (0.52,14.541) (0.016,0.063) (0.237,11.949000000000002) (0.985,58.088) (2.278,33.982) (1.053,16.815) (1.326,20.087) (2.908,45.419) (3.441,45.275) (2.06,30.061999999999998) (1.483,108.72) (3.377,47.61) (0.017,0.073) (0.016,0.052000000000000005) (3.536,72.871) (1.337,19.372999999999998) (5.587,81.767) (193.799,9953.933) (89.858,3959.588) (0.023,0.06899999999999999) (49.796,2973.447) (0.718,76.449) (5.364,204.765) (3.21,138.828) (41.455,1955.3980000000001) (1.083,53.278999999999996) (1.695,279.8) (1.345,286.818) (3.111,411.545) (6.397,1495.882) (22.214,666.7180000000001) (65.322,110.378) (22.917,420.612) (272.642,3116.728) (1.929,85.946) (1358.165,5872.383) (3951.002,10810.879) (14.808,1124.911) (93.537,4307.451)};
-
-%    \input{data-formatted/time-d4-verify-onesided-only}
-
-\addplot [only marks, color=bluegray, mark=square*,  mark options={scale=0.8}, opacity=0.7] coordinates { (205.743,2420.6859999999997) (326.575,4433.625) (476.021,3973.2659999999996) (494.38,5879.952) (1046.603,10779.982) (351.999,2069.865) (64.21,479.92100000000005) (316.505,3074.084) (788.377,6710.351000000001)};
-
-%    \input{data-formatted/time-d4-failure}
-
-\addplot [only marks, color=darkgreen, mark=diamond*, mark options={scale=1.1}, opacity=0.7] coordinates { (2052.239,25000) (424.73,25000) (2636.562,25000) (1403.431,25000) (3070.214,25000) (2053.446,25000) (2790.169,25000)};
-
-    \legend{
- \textsf{Full validation},
- \textsf{One-sided validation},
- \textsf{Resource limit exceeded},
-    }
-    \addplot[mark=none, dashed] coordinates{(0.01,25000) (4000, 25000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,20000) (4000,20000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (2000,20000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (200, 20000)};
-    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (20, 20000)};
-    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
-    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
-    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
-    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
-    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-\centering{%
-\begin{tikzpicture}[scale = 0.50]
-  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10},
-                              x post scale=1.8, y post scale=2.15,
-                              xmode=log,xmin=100,xmax=1000000000, 
-                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              ymode=log, ymin=100, ymax=1000000000, 
-                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
-                              xlabel={Defining Clauses},
-                              %ylabel={Proof Clauses},
-%                              title={D4 Defining Clause Generation}
-            ]
-    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
-    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
-%%    \input{data-formatted/defining+total}
-
-\addplot [only marks, color=redorange, mark=*,  mark options={scale=0.8}, opacity=0.7] coordinates { (18271,161794) (1295,2397) (1659,4427) (127284,262373) (2325,6077) (359,554) (5373,14373) (314359,1213826) (44178,81780) (493,33207) (605988,1117440) (1141369,3809317) (1197649,2211944) (1001050,2393131) (2947,7851) (4991,13207) (355617,721935) (4897,13101) (2437497,5273728) (345,46079) (701346,1593947) (3191,26378) (5985623,28392373) (467550,940819) (1318024,2813565) (4713,23892) (1682059,3160386) (1406192,4043887) (899701,1704811) (4936014,11840667) (8038928,68469623) (70651620,197339867) (9567,251890) (39501,104182) (18007,187551) (125319627,374762287) (82194,297768) (269645146,770382773) (304,427342) (574062,8448281) (887,478450) (1757,399856) (78873354,503117290) (190248,429409) (2064844,13434107) (13737124,141587875) (50722,820648) (45923418,82243431) (83111464,188103550) (40776253,104351646) (79106742,265990284) (3133905,23973053) (16147,1697847) (45214,1277037) (72258,4612352) (905,2500003) (401032,840020) (355,2155900) (63050,4692737) (3195277,6098685) (1125105,4417869) (325251378,736764776) (6074672,13104909) (132891,287958) (65522,136116) (15561156,38336288) (143191,298755) (2851,7851) (5243,13911) (334862,681444) (2659,7335) (82585,317789) (447979,1612224) (905899,1733678) (433226,802268) (551970,1006319) (1891045,3387533) (1210098,2232426) (755273,1440924) (470934,1914084) (1178453,2188886) (3131,8575) (2515,6607) (1903735,4312279) (1107098,1868005) (1801878,3324418) (47556240,206213289) (24087642,83491057) (3109,8227) (16038150,101261480) (312728,642675) (3175308,7847036) (142059,747752) (20382514,69650789) (453769,1563728) (498246,1154507) (797866,6840229) (3835114,17412807) (5035130,10668725) (647011,9905453) (5189,1234644) (740235,2672407) (219979530,417455256) (402460,844395) (2295530,81147729) (1772352,119113610) (3975194,16763326) (8493275,41873955)};
-
-    \node[left] at (axis cs: 900,1200) {$1\times$};
-    \node[left] at (axis cs: 900,12000) {$10\times$};
-    \node[left] at (axis cs: 900,120000) {$100\times$};
-    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
-    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};
-
-          \end{axis}
-\end{tikzpicture}
-} % centering
-\end{minipage}
-\caption{Runtime (left) and proof size (right) for CPOG proofs.  The runtime includes proof generation, checking, and model counting relative to the runtime for D4.
-  The proof size is measured as total clauses relative to the number of defining clauses.}
-\label{fig:dual:total}
-\end{figure}
-
-For our
-evaluation, we used the public benchmark problems from the 2022
-standard and weighted model competitions.\footnote{Downloaded from \url{https://mccompetition.org/2022/mc\_description.html}}  We found that there were
-180 unique CNF files among these, ranging in size from 250 to
-2,753,207 clauses.
-We ran our programs on a processor with 64~GB of
-memory and having an attached high-speed, solid-state
-disk.
-With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
-benchmark problems.  Our proof generator was able to convert all of
-these into POGs, with their declarations ranging from 304 to
-2,761,457,765 (median 848,784) defining clauses.
-%(The
-%maximum count overflowed the overflowed the 32-bit signed integers we
-%used to represent clause identifiers in both the proof generator and
-%checker.)
-
-We ran our proof generator with a time limit of 10,000 seconds.  The
-results are shown in Figure~\ref{fig:dual:total}, The left-hand plot
-shows the elapsed time for the combination of proof generation,
-checking, and counting versus the time for
-\dfour{}.
-The proof generator was able to generate full proofs for 108 of the problems and
-one-sided proofs for an additional 9 of them, leaving just 7 with no
-verification.  The prototype checker successfully verified all of the generated
-proofs.  The longest runtime for the combination of proof generation, checking, and counting
-for a full proof was 13,145 seconds.
-Overall the ratio
-between the combined time for generation, checking, and counting versus the time
-for \dfour{} had a harmonic mean of 5.5.
-The right-hand plot shows the total number of clauses in the CPOG file versus the number of defining clauses for the problems having full proofs.
-The ratio between the total number of clauses and the number of defining clauses had a harmonic mean of 3.13.
-To date, we have not found
-any errors in the dec-DNNF graphs generated by \dfour{}.
-
-We found that the monolithic approach for generating the forward
-implication proof works well for smaller POGs (up to one million
-defining clauses), but it becomes inefficient for larger ones.  These
-experiments suggest a possible hybrid approach, stopping the recursion
-of the structural approach and shifting to monolithic mode once the
-subgraph size is below some threshold.  By using monolithic mode, we
-were also able to perform end-to-end verification of all but one of the
-benchmarks that could be verified without preprocessing, with a total
-time limit (including preprocessing) of 1,000 seconds.
-
-We also found that our two optimizations: literal grouping and
-lemmas can provide substantial improvements in proof size and runtime.
-In the extreme cases, a lack
-of lemmas caused one proof to grow by a factor of 52.5, while a lack
-of literal grouping caused another proof to grow by a factor of 39.6.
-Overall, it is clear that these two optimizations are worthwhile, and
-sometimes critical, to success for some benchmarks.
-
-Running the verified proof checker in \lean{} required, on average
-(harmonic mean), around 5.9 times longer than the prototype checker.
-Encouragingly, the scaling trends were identical for the two solvers,
-indicating that the two checkers have similar asymptotic performance.
-We consider a factor of 5.9 to be acceptable for the assurance
-provided by formal verification.
-
-In comparing other proof frameworks, we found that the \cdfour{}
-toolchain ran very fast and could handle very large benchmarks.  Even
-with a total time limit of 1,000 seconds, including the time for
-knowledge compilation, the \cdfour{} toolchain completed 106
-benchmarks, while the CPOG toolchain completed just 82.  We found
-that the runtimes for the MICE toolchain versus our CPOG toolchain
-showed little correlation, reflecting the fact that the two solve
-different problems and use different approaches.  In general, our CPOG
-toolchain showed better scaling, in part due to its ability to control
-the recursion through lemmas.  Neither of these prior
-toolchains could perform end-to-end verification when the knowledge
-compilation was preceded by preprocessing.
+\input{experimental-section}
 

### 2024-01-06 doc/JAIR/jair-paper.tex
fix: typos

@@
 $\nodes_7$ has clause IDs 7 and 10 as the hint.  We can see in Figure
-\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
+\ref{fig:eg4:all}(D) that these two clauses form a RUP proof for the clause
 $\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
@@
 be implied by the input clauses when $\assign(x_1) = 1$, while defining clause 24 shows that node $s_{10}$ will be implied
-by the input clauses when $\eassign{p_9} = 1$.  These three clauses
+by the input clauses when $\eassign(p_9) = 1$.  These three clauses
 serve as the
@@
 input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
-that there is no total assignment $\assign$ that falsifies this clause but extends to $\eassign$ such that  $\assign(s_{10}) = 1$.
+that there is no total assignment $\assign$ that falsifies this clause but extends to $\eassign$ such that $\eassign(s_{10}) = 1$.
 The proof proceeds by first assuming that the clause is false, requiring
 $\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consists of unit
-clauses (e.g., clause 36 asserting that $\eassign(p_{10}) = 1$) or
+clauses (e.g., clause 36 asserting that $\eassign(s_{10}) = 1$) or
 clauses that cause unit propagation.  Hint clauses 8 and 10 force the

### 2024-01-06 doc/JAIR/jair-paper.tex
fix: typo

@@
  $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
- showing that any total assignment $\assign$ that activates the synthetic clauses will, when cause $u$ to be assigned $1$.
+ showing that any total assignment $\assign$ that activates the synthetic clauses will cause $u$ to be assigned $1$.
  More precisely, given assignment $\assign$ and its extension $\eassign$, if $\eassign(v_j) = 0$ for every guard literal $\obar{v}_j$, then $\eassign(u) = 1$.

### 2024-01-06 doc/JAIR/jair-paper.tex
Change equivalence statement

@@
 formula $\theta$ over the variables $\varset \cup \exvarset(\theta)$, we say that $\phi$ is {\em
-  effectively equivalent} to $\theta$, written
+  equivalent over $\varset$} to $\theta$, written
 $\phi \eequiv \theta$,

### 2024-01-07 doc/JAIR/lean-section.tex
feat: Lean section

@@
 
-We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
-sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \ref{fig:chain}).
-These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}.
-Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic.
-In this section, we briefly describe the functionality we implemented and what we proved about it.
+We set out to formally verify the system with two goals in mind:
+first, to ensure that the CPOG framework is mathematically sound;
+and second, to implement correct-by-construction proof checking
+and ring evaluation (the ``Trusted Code'' components of \ref{fig:chain}).
+These two goals are achieved with a single proof development
+in the \lean{} programming language~\cite{demoura:cade:2021}.
+Verification was greatly aided
+by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean}
+automated proof search tactic.
+\lean{} is based on a logical foundation
+in which expressions have a computational interpretation.
+As in other proof assistants such as Isabelle and Coq,
+functions defined in the formal system
+can be compiled to machine code.
+At the same time,
+we can state and prove claims about them within the same system,
+thereby verifying that our functions compute the intended results.
+In this section,
+we describe the functionality we implemented,
+what we proved about it,
+and the assumptions we made.
 
-\vspace{1em}\noindent
-\textbf{Proof checking.} The goal of a CPOG proof is to construct a POG that is equivalent to the input CNF $\inputformula$.
-The POG being constructed, and the set of active clauses are stored in the checker state \texttt{st}.
-The checker begins by parsing the input formula, initializing the active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one.
-It then processes every step of the CPOG proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect.
-Afterwards, it carries out the \textsc{final conditions} check of \ref{subsection:semantics}.
-Throughout the process, we maintain invariants that ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root (\ref{thm:lean:equiv}).
+\paragraph{Data structures and mathematical model.}
+When thinking about formal verification,
+it is helpful to distinguish between data structures
+that play a role in the code being executed,
+and \emph{ghost} definitions that serve as a mathematical model,
+allowing us to state and prove specifications,
+but are erased during compilation and not executed.
+We generally store definitions in the two classes
+under {\tt Data/} and {\tt Model/},
+respectively.
 
-The specifications we use to state these invariants are built on a general theory of propositional logic, mirroring \ref{sect:logical:foundations}.
-Following the DIMACS CNF convention, we define the data types \lstinline{Var} of variables being positive natural numbers, \lstinline{ILit} of literals being non-zero integers, and \lstinline{PropForm} of propositional formulas.
-\lstinline{PropForm} is generic over the type of variables, so we instantiate it with our type as \lstinline{PropForm Var}.
+Among the former is our representation of CNF formulas.
+Following the DIMACS CNF convention,
+a variable is represented as a positive natural number,
+a literal is a non-zero integer,
+a clause is an array of literals,
+and a CNF formula is an array of clauses.
+\begin{lstlisting}
+def Var := { x : Nat // 0 < x }
+def ILit := { i : Int // i ≠ 0 }
+abbrev IClause := Array ILit
+abbrev ICnf := Array IClause
+\end{lstlisting}
+
+A POG is represented as a flat array of elements.
+Each element {\tt PogElt} of a POG is either a variable,
+a binary disjunction (sum),
+or an arbitrary conjunction (product).
+\begin{lstlisting}
+inductive PogElt where
+  | var (x : Var) : PogElt
+  | disj (x : Var) (l r : ILit) : PogElt
+  | conj (x : Var) (args : Array ILit) : PogElt
+\end{lstlisting}
+In the first case,
+the argument \lstinline{x} is the index of an input variable;
+in disjunctions and conjunctions,
+it is an extension variable appearing in the CPOG file.
+A \lstinline{Pog} is then an array of \lstinline{PogElt}s that is well-founded
+in the sense that each element depends only on prior elements in the array.
+Note that representing edges as literals
+allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
+
+On the mathematical side,
+our specifications rely on a general theory of propositional logic
+mirroring \ref{sect:logical:foundations}.
+The type \lstinline{PropForm} describes the syntax of propositional formulas.
+It is generic over the type of variables,
+so we instantiate it with numeric variables as \lstinline{PropForm Var}.
 Assignments of truth values are taken to be total functions \lstinline{PropAssignment Var := Var → Bool}.
-Requiring totality is not a limitation: instead of talking about two equal, partial assignments to a subset $X' \subseteq X$ of variables, we can more conveniently talk about two total assignments that agree on $X'$.
+Requiring totality is not a limitation:
+instead of talking about two equal,
+partial assignments to a subset $X' \subseteq X$ of variables,
+we can more conveniently talk about two total assignments that agree on $X'$.
 We write \lstinline{σ ⊨ φ} when \lstinline{σ : PropAssignment Var} satisfies \lstinline{φ : PropForm Var}.
 
-The invariants refer to the checker state \lstinline{st} with fields \lstinline{st.inputCnf} for $\inputformula$, \lstinline{st.clauseDb} for $\theta$, \lstinline{st.pog} for $P$, \lstinline{st.pogDefsForm} for the clausal POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$, and \lstinline{st.allVars} for all variables (original and extension) added so far.
-For any $\nodeu\in P$, \lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.
-The first two invariants state that assignments to original variables extend uniquely to extension variables defining the POG nodes.
+Functions \lstinline{ILit.toPropForm},
+\lstinline{IClause.toPropForm},
+\lstinline{ICnf.toPropForm},
+and \lstinline{Pog.toPropForm}
+relate data structures to the formulas they encode.
+For example, given a literal \lstinline{u},
+\lstinline{P.toPropForm u} denotes the interpretation
+of the node $\nodeu$ corresponding to \lstinline{u} in the POG \lstinline{P}
+as a propositional formula $\phi_\nodeu/\neg\phi_\nodeu$
+over the input variables.
+It is negated if \lstinline{u} has negative polarity.
+Lean provides a convenient ``anonymous projection'' notation
+that allows writing \lstinline{P.toPropForm u} instead of \lstinline{Pog.toPropForm P u}
+when \lstinline{P} has type \lstinline{Pog},
+\lstinline{C.toPropForm} instead of \lstinline{IClause.toPropForm C}
+when \lstinline{C} has type \lstinline{IClause},
+etc.
+
+In order to reason about composite formulas,
+we found it easier to work with propositional formulas modulo logical equivalence,
+a structure known in logic as the \emph{Lindenbaum--Tarski algebra},
+rather than using \lstinline{PropForm} directly.
+Its advantage is that equivalent but not syntactically equal formulas
+(such as $x \vee \neg x$ and $\top$)
+give rise to equal elements in the algebra,
+and equality has a privileged position in proof assistants based on type theory:
+equals can be substituted for equals in any context.
+In this way, forgetting syntactic detail is helpful.
+On the other hand,
+using the algebra gives rise to some challenges.
+The algebra, called {\tt PropFun}, is defined as a quotient,
+with Boolean operations and the entailment relation
+lifted from the syntax of formulas to the new type.
+It is no longer straightforward to say
+when an element of the quotient ``depends'' on a variable
+since equivalent formulas can refer to different sets of variables.
+Instead, we use a semantic notion of dependence
+in which an element $\phi$ of the quotient depends on a variable $x$
+if and only if there is a truth assignment that satisfies $\phi$,
+but falsifies $\phi$ after $x$ is flipped.
+\begin{lstlisting}
+/-- The *semantic variables* of `φ`
+are those it is sensitive to as a Boolean function.
+Unlike `vars`, this set is stable under equivalence of formulas. -/
+def semVars (φ : PropFun ν) : Set ν :=
+  { x | ∃ (τ : PropAssignment ν), τ ⊨ φ ∧ τ.set x (!τ x) ⊭ φ }
+\end{lstlisting}
+
+\paragraph{Proof checking.}
+The goal of a CPOG proof is to construct a POG
+that is equivalent to the input CNF $\inputformula$.
+The database of active clauses,
+the POG being constructed,
+and its root literal,
+are stored in a checker state structure {\tt PreState}.
+The checker begins by parsing the input formula,
+initializing the active clauses to $\theta \leftarrow \inputformula$,
+and initializing the POG $P$ to an empty one.
+It then processes every step of the CPOG proof,
+either modifying its state by adding/deleting clauses in $\theta$
+and adding nodes to $P$,
+or throwing an exception if a step is incorrect.
+Afterwards, it carries out the \textsc{final conditions} check of \ref{subsection:semantics}.
+
+Throughout the process,
+we maintain invariants needed to establish the final result.
+These ensure that $P$ is partitioned
+and that a successful final check entails the logical equivalence
+of $\inputformula$ and $\phi_\noder$,
+where $\noder$ is the final POG root (\ref{thm:lean:equiv}).
+Formally, we define a type {\tt State}
+consisting of those {\tt PreState}s
+that satisfy all the invariants.
+A {\tt State} is a structure
+combining {\tt PreState} fields with additional ones
+storing computationally irrelevant \emph{ghost state}
+that asserts the invariants.
+The fields of \lstinline{st : PreState} include
+\lstinline{st.inputCnf} for $\inputformula$,
+\lstinline{st.clauseDb} for $\theta$,
+and \lstinline{st.pog} for $P$.
+We write \lstinline{st.pogDefsForm} for the clausal POG definitions formula $\bigwedge_{\nodeu\in P}\theta_u$,
+and \lstinline{st.allVars} for all variables (original and extension) added so far.
+For any $\nodeu\in P$,
+\lstinline{st.pog.toPropForm u} computes $\phi_\nodeu$.
+
+The first invariant states that assignments to input variables
+extend uniquely to extension variables defining the POG nodes.
 In the formalization, we split this into extension and uniqueness:
@@
 so the definition says that the three assumptions imply the conclusion.
-The next invariant guarantees that the set of solutions over the original variables is preserved:
+The next invariant guarantees that the set of solutions over the input variables is preserved:
 \begin{lstlisting}
@@
 \end{lstlisting}
-These invariants are maintained by valid CPOG proofs. Together with additional invariants
-that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
-node $\noder$:
+
+The bulk of our work involved showing
+that these invariants are indeed maintained by the checker
+when going through a valid CPOG proof,
+modifying the active clause database and the POG.
+Together with additional, technical invariants
+about the correctness of cached computations,
+they imply the soundness theorem for $P$ with root node $\noder$:
 
@@
 
-\vspace{1em}\noindent
-\textbf{Ring evaluation}. We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
+\paragraph{Ring evaluation.} We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
 (\ref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
@@
 
-\vspace{1em}\noindent
-\textbf{Model counting.} Finally, we established that ring evaluation with the appropriate weights
+\paragraph{Model counting.} Finally, we established that ring evaluation with the appropriate weights
 corresponds to the standard model count. To do so, we defined a function that
@@
 \end{lstlisting}
-In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
-we have that the number of models of \lstinline{φ} on its variables is
+In particular, taking \lstinline{s} to be exactly the variables appearing in \lstinline{φ},
+we have that the number of models of \lstinline{φ} over its variables is
 \lstinline{countModels φ (card (vars φ))}.
+
+\subsection{Trust}
+In this subsection, we clarify what has been verified and what has to be trusted.
+Recall that our first step is to parse CNF and CPOG files in order to read in the initial
+formula and the proof. We do not verify this step. Instead, the verified checker exposes
+flags \verb+--print-cnf+ and \verb+--print-cpog+ which reprint the consumed formula or proof,
+respectively. Comparing this to the actual files using {\tt diff} provides an easy way of ensuring
+that what was parsed matches their contents. This involves trusting only the correctness of the
+print procedure and {\tt diff}. Similarly, if one wants to establish the correctness of the POG
+contained in the CPOG file, one can print out the POG that is constructed by the checker and compare.
+
+Lean's code extraction replaces calculations on natural numbers and integers with
+efficient but unverified arbitrary precision versions.
+Lean also uses an efficient implementation of arrays; within the
+formal system, these are defined in terms of lists, but code extraction replaces them
+with dynamic arrays and uses reference counting to allow destructive updates when it is safe
+to do so \cite{Ullrich:de:Moura:19}.
+Finally, Lean's standard library implements hashmaps in terms of arrays.
+Many of the basic properties of hashmaps have been formally verified, but not all.
+In particular, we make use of a fold operation whose verification is not yet complete.
+Thus our proofs depend on the assumption that the fold operation has the
+expected properties.
+
+In summary, in addition to trusting Lean's foundation and kernel checker,
+we also have to trust that code extraction respects that foundation,
+that the implementation of the fold operation for hashmaps satisfies its description,
+and that, after parsing, the computation uses the correct input formula.
+All of our specifications have been completely proven and verified relative to these assumptions.
+
+%P: clause db
+% We define clause databases in {\tt ClauseDb} as hashmaps
+% that store each clause together with a flag
+% indicating whether it has been deleted.
+
+%P: files and directories
+% Our code is contained in the directory {\tt ProofChecker}.
+% The main checker loop is implemented in {\tt Checker/CheckerCore.lean}.
+% The ring evaluator and model counter are contained in the
+% {\tt Count} directory.

### 2024-01-07 doc/JAIR/lean-section.tex
Minor typo

@@
 and second, to implement correct-by-construction proof checking
-and ring evaluation (the ``Trusted Code'' components of \ref{fig:chain}).
+and ring evaluation (the ``Trusted Code'' components of Figure~\ref{fig:chain}).
 These two goals are achieved with a single proof development

### 2024-01-07 doc/JAIR/lean-section.tex
add citations and fix references in Lean section

@@
 in which expressions have a computational interpretation.
-As in other proof assistants such as Isabelle and Coq,
+As in other proof assistants such as Isabelle~\cite{nipkow:et:al:02} and Coq~\cite{coq},
 functions defined in the formal system
@@
 our specifications rely on a general theory of propositional logic
-mirroring \ref{sect:logical:foundations}.
+mirroring Section~\ref{sect:logical:foundations}.
 The type \lstinline{PropForm} describes the syntax of propositional formulas.
@@
 \begin{lstlisting}
-/-- The *semantic variables* of `φ`
-are those it is sensitive to as a Boolean function.
-Unlike `vars`, this set is stable under equivalence of formulas. -/
+/-- The *semantic variables* of `φ` are those it is sensitive to as a Boolean
+function. Unlike `vars`, this set is stable under equivalence of formulas. -/
 def semVars (φ : PropFun ν) : Set ν :=
@@
 or throwing an exception if a step is incorrect.
-Afterwards, it carries out the \textsc{final conditions} check of \ref{subsection:semantics}.
+Afterwards, it carries out the \textsc{final conditions} check of Section~\ref{subsection:semantics}.
 
@@
 of $\inputformula$ and $\phi_\noder$,
-where $\noder$ is the final POG root (\ref{thm:lean:equiv}).
+where $\noder$ is the final POG root (Theorem~\ref{thm:lean:equiv}).
 Formally, we define a type {\tt State}
@@
 \end{lstlisting}
-Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (\ref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
+Finally, for every node $\nodeu\in P$ with corresponding literal $u$ we ensure that $\phi_\nodeu$ is partitioned (Definition~\ref{def:partitioned-operation-formula}) and relate $\phi_\nodeu$ to its clausal encoding $\theta_u \doteq u \wedge \bigwedge_{\nodev\in P}\theta_v$:
 \begin{lstlisting}
@@
 \paragraph{Ring evaluation.} We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
-(\ref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
+(Definition \ref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
 \begin{lstlisting}
@@
 \end{lstlisting}
-\ref{prop:ring:eval} is then formalized as follows:
+Proposition~\ref{prop:ring:eval} is then formalized as follows:
 \begin{lstlisting}
